### 2018-09-13 notes/powell_vs.tex
Add some notes

\documentclass{article}

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,subfigure}

\def\x{{\mathbf{x}}}
\def\a{{\mathbf{a}}}
\def\s{{\mathbf{s}}}
\def\r{{\mathbf{r}}}
\def\I{{\mathbf{I}}}
\def\eps{{\boldsymbol{\varepsilon}}}
\def\m{{\boldsymbol{\mu}}}
\def\param{{\boldsymbol{\theta}}}

\title{Powell's method revisited using variational sampling?} 
\author{Alexis Roche} 
%\date{}

\begin{document}

\maketitle

\section{Introduction}

Some thoughts I had during the May 1st break... I came across the concept of variational sampling more than 8 years ago. I was always convinced that it was an interesting idea, but the reality is that I haven't been able to find any clearly useful application of it so far.

My latest attempt to use VS as a building block for the Expectation Propagation algorithm ended up showing in practice that VS was not really better than the good old Laplace method. Before that, I had considered a brute-force implementation of VS to solve the problem of evaluating a KL-optimal approximation at a very general level, and the conclusion was that VS was not appropriate in large dimension.

One path I haven't followed up to now is the natural continuation of the one that took me to the VS idea. I was initially trying to perform gradient-free quadratic approximations to a cost function in the context of running Powell's optimization method. The standard implementation of Powell's method performs successive line minimizations using Brent's method, a fairly simple scheme that relies on one-dimensional quadratic approximations using three points. Beside the problem that Brent's method requires a fair number of function evaluations, it doesn't provide a global quadratic approximation to the objective function, unlike Newton's method -- and such an approximation would be nice to evaluate things like the posterior variance (if the cost is some kind of log-likelihood function).

\section{Factorial fit}

So here is the problem to solve. Consider an unnormalized distribution $p(\x)$. We want to approximate it by a factorized form $q(\x) = \prod_i g_i(\x_i)$, where $(\x_1,\x_2,\ldots)$ are coordinate blocks and each factor is parameterized independently, 
$$
g_i(\x_i) = \exp [\param_i^\top \phi(\x_i)]
.
$$

This is not even a restriction of the general fitting problem if the coordinate blocks are allowed to overlap. In fact, for now, the ``big idea'' is to perform an alternate minimization in the parameter space as opposed to a full minimization. Does it make our life easier? Of course, it does, because each $\param_i$ lives in a smaller dimensional space than all parameters altogether, so we will be performing many simple minimizations instead of a single very difficult one. VS can be used for each such minimization, obviously yielding a sequence of convex problems...

I cannot believe that I haven't thought about that earlier! This is actually not true, but it is true that I got lost on that  track, for reasons that I don't really know... I probably was blinded by two things: one was the EP algorithm, the other was some sort of belief that alternate minimization is irrelevant in convex problems. This conception is clearly wrong when considering the benefit of dimension reduction. On the other hand, the key difference with EP is that we don't assume here that the target distribution has a factorial form -- only the fitting distribution is factorial. So we are not trying to solve the same problem as EP, in fact our problem is (much) more general.

Each optimization step boils down to fitting $p(\x)$ with a form $c(\x)g_i(\x_i)$, where $c(\x)=\prod_{j\not= i}g_j(\x_j)$ is the very same thing as the context in EP, by minimizing the KL divergence $D(p\|cg_i)$ wrt $g_i$.

\section{Variational sampling idea}

The generalized KL divergence reads:
$$
D(p\|q) = \int  p \log \frac{p}{q} - p + q .
$$

If $p$ is our target distribution, the objective is minimized wrt $q$ in some approximation space. This is obviously equivalent to maximizing the ``log-likelihood'':
$$
L(q) = \int p \log q - q,
$$
or, more generally, if $q$ decomposes as $q=c g$ with $c$ fixed:
$$
L_c(f) = \int p \log g - c g
$$

The key idea of VS is to approximate the generally intractable objective via a sampling average. For this, we need to choose some sampling distribution $s(\x)$ and apply an associated quadrature rule:
$$
\int s(\x) f(\x) d\x \approx \sum_k w_k f(\x_k),
$$
leading to the surrogate objective:
$$
\tilde{L}_{c,s}(g)
= 
\sum_k 
w_k
\left[
a(\x_k) \log g(\x_k)
- b(\x_k) g(\x_k)
\right],
\qquad
a = \frac{p}{s},
\quad
b = \frac{c}{s}
.
$$

When the approximation space is an exponential family, $f(\x) = \exp[\param^\top \phi(\x)]$, we have:
$$
\tilde{L}_{c,s}(\param)
= 
\sum_k 
w_k \left[
a(\x_k) \param^\top \phi(\x_k)
- b(\x_k) e^{\param^\top \phi(\x_k)}
\right]
$$

$$
\nabla \tilde{L}_{c,s}(\param)
= 
\sum_k 
w_k \left[
a(\x_k) 
- b(\x_k) e^{\param^\top \phi(\x_k)}
\right]
\phi(\x_k)
.
$$

$$
\nabla \nabla^\top \tilde{L}_{c,s}(\param)
= 
- \sum_k 
w_k 
b(\x_k) e^{\param^\top \phi(\x_k)}
\phi(\x_k) \phi(\x_k)^\top 
.
$$

There are two arbitrary distributions involved in VS:
\begin{itemize}
\item the reference or cavity or context distribution $c$, which can be thought of as some window function defining the area of interest;
\item the sampling distribution $s$, which impacts the VS approximation accuracy; it's not obvious how to optimally choose $s$ (see appendix~\ref{app:opt_sampling}), but intuition suggests that sampling from~$c$ should do.
\end{itemize}

\subsection{Importance of context}

In my initial take on VS \cite{ijasp:13}, there was no context, meaning $c\equiv 1$. Choosing an appropriate sampling distribution was then fairly painful, as one just cannot sample from a uniform distribution... As an educated guess, one can try and sample from a distribution that approximates the target, yielding a typical ``chicken and egg" problem.

In my work on combining VS with EP \cite{rr:16}, I was using the context as the sampling distribution, which makes sense if the context is narrow enough compared to the variations of the target distribution -- something we expect to be verified in the late EP iterations. Arguably, a better choice would be the context multiplied by the current approximation to the factor being updated... By construction of EP, the context is also included in the target, i.e. $p=cf$, yielding an alternative interpretation that the factor $f$ is being approximated by another factor $g$. Choosing $s=cg_0$, we have $a=f/g_0$ and $b=1/g_0$.

\subsection{Iterative VS}

To date, I have only studied one-step versions of VS, but we can think of the following iterative scheme in case the sampling distribution is completely off as the context is too broad.

Given a target distribution $p$ and a context $c$, consider $f=p/c$ and do the following:
\begin{enumerate}
\item Pick some initial narrow context $c$ and some initial factor fit $g$ and set $s\leftarrow c g$.
\item Solve the VS problem for $s$, yielding a factor update $g$.
\item Widen context, reset $s\leftarrow c g$, and return to 2 until the context reaches its shape.
\end{enumerate}

If some sort of convergence occurs, we are left with both a parametric approximation to the target and an approximate independence sampler... Sounds a bit naive, though. In low dimension, there is some hope that the above iteration may work, but I wouldn't be too confident in high dimension... Intuition says that we need to take very small steps to update the context and that could be very time consuming. But, anyway, forget about the brute-force version of VS in large dimension... just too heavy.

\section{Dimension reduction scheme}

So far we haven't made any use of the fact that the factor depends on a subset of coordinates only. The 1000 dollars question is: do we really need to sample the whole $\x$ space if the factor involves a few coordinates? Needless to say, this could be time consuming.

Let us take a step back and note that, if $f_i$ is a function of $\x_i$ only, the objective (not the surrogate one) amounts to an integral in the relevant subspace:
$$
L_\pi(f) = \int [ p_i(\x_i) \log f(\x_i)  - \pi_i(\x_i) f(\x_i) ] d\x_i,
$$
with:
$$
p_i(\x_i) = \int p(\x) d\x_{-i},
\qquad
\pi_i(\x_i) = \int \pi(\x) d\x_{-i}
,
$$
but this is still cumbersome because we don't have the marginal $p_i(\x_i)$ and would need some extra magic to somehow evaluate it.

What if we approximate $p_i(\x_i)$ proportionally to a coordinate application for some fixed values of the coordinates not included in $\x_i$... This would be verified, for instance, if we replace the actual target with:
$$
\tilde{p}(\x) = \lambda c(\x) p(\a_{-i}, \x_i),
$$
where $\lambda$ is some constant and $\a$ is some current estimate of the target mode and $\a_{-i}$ denoting the ``fixed" coordinates of $\a$. This way, the factor to be adjusted is essentially the coordinate application at the mode estimate.

Doesn't that sound familiar? Yes, it reminds us of the EP trick! The {\em only} difference, or say, addition, is that the factor is independent from any context in EP because the target is assumed to be factorial, while here no such assumption is made. And what I am finding out here, is that there is a very simple way to generalize the EP trick, which is to do this $p(\a_{-i}, \x_i)$.

How shall we choose $\lambda$? Let us assume {\em non-overlapping coordinate blocks}. We are approximating each coordinate application by some parametric form and a particular iteration of our algorithm is concerned with a particular coordinate application. Defining:
$$
g_{-i}(\x)\equiv \prod_{j\not=i} g_j(\x_j),
\qquad
c(\x) \equiv \frac{g_{-i}(\x)}{g_{-i}(\a)},
$$
where it is natural to choose $\a$ as the current approximation maximizer, the coordinates of which respectively maximize each coordinate fit, a natural approximation to the target distribution is:
$$
p(\x)
\approx
\tilde {p}(\x)
= c(\x) p(\a_{-i},\x_i)
\approx c(\x) g_i(\x_i),
$$
which verifies the property $\tilde{p}(\a)=p(\a)$. The quantity $c(\x)$ can be seen as the {\em context} in which the coordinate fit is performed, so here the context is normalized (it's not the mere product of the factors).

In summary, our scheme is similar to EP up to two modifications:
\begin{enumerate}
     \item The context is normalized.
     \item The target factor is contextual (depends on the context). 
\end{enumerate}

\section{Generalized EP}

The goal is to describe some target distribution as a set of {\em factors}. All it takes is a parametric approximation space for each factor and a {\em contextual} approximation scheme. The contextual scheme is a method to approximate the target depending on the said factors. More specifically, there is one scheme for each factor, which determines an approximation to the target given the factor under consideration and all the other factors fitted.

We thus have two approximation levels: one structural (the description stage) and one numerical (the fitting stage). The algorithm is fully determined by the description stage.

In the classical EP, the description is exact in the sense that it recovers exactly the target if all factors are exact.

In the coordinate-wise EP, the description is inexact unless the target is separable, i.e. is the product of its coordinate applications up to a constant. But we could not care less.

The two things that these situations have in common is the factorial nature of the approximating distribution and the fact that each factor may involve a reduced set of coordinates. In fact, if they don't, one may question the benefit of the factor decomposition.

\subsection{Motivation}

The above insight helps understanding the very nature of EP: it is a dimension reduction trick! Minka never gave a clear motivation for the structural approximation underlying EP \cite{Minka-01,Minka-05}.

When factors correspond to data items, one advantage is that we don't need to load the whole data in memory -- but is it decisive? Our premise is that we want an approximation of the form: 
$$
p(\x) 
\approx
c(\x) g_i(\x_i)
$$

Another potential motivation, which is more or less what Minka seemed to have in mind, is that the fitting problem may enjoy a closed-form solution with the approximation while being intractable without. This is of course not always true, as in the logistic and hinge regression problems considered in my EP/VS paper. Reason why VS is an alternative to EP {\em per se} in moderate dimension.

Dimension reduction appears as a stronger rational: if the factors happen to depend on subsets of variables, then the EP trick leaves us with smaller dimensional fitting problems. I don't remember Minka mentioning this. In the classical online learning setting, this does not happen because each factor is the model response to a single data item and thus basically depends on all the model parameters...

\subsection{Global fit}

As the algorithm hopefully converges, do the different factor approximations $c_{-i}g_i$ converge to the same distribution? If so, it would yield a global fit to the target, which may be needed sometimes.

Open question: how to design the contextual scheme to guarantee this convergence in general? It is straightfoward in the classical EP. However, in the coordinate-wise EP, it is unlikely to occur because, even if $\a$ converges, all fitted factors may not have exactly the same value at $\a$.

To address this, we can simply define a global approximation scheme to be used in the end, which does not need to be consistent with the contextual approximation scheme.

The contextual scheme applies when trying to best approximate the target while able to evaluate each factor independently, and only one exactly.

The global scheme applies when trying to best approximate the target while able to evaluate each fitted factor independently.

It is about setting constraints to ourselves in order to gain freedom. As often in computational statistics.

\subsection{Machine learning}

It sounds a bit like I was not using EP at its best for machine learning. Instead of having factors associated with data items, they should be associated with model parameters. But... wait... the target distribution does not factorize across model parameters, so EP is not applicable here. It's not, but we have seen how we can extend EP to satisfy our need.

If we still wish to do online learning, we have the opportunity to keep the factorization across batches, and sub-factorize each batch factor across model parameters. This means that each iteration of the algorithm will update a specific parameter subset based on a specific batch.

\section{Epilogue}

Il n'est pas n\'ecessaire d'imposer la factorisation par coordonn\'ees de la forme approximante. Celle-ci d\'ecoule directement de l'approximation contextuelle ``g\'en\'eralis\'ee'' dans laquelle le facteur \`a ajuster est remplac\'e par le produit de ses applications coordonn\'ees. Avec \c ca et la forme param\'etrique factorielle, choisie {\em a priori}, non seulement l'approximation est factorielle mais, et c'est le point cl\'e, l'ajustement des facteurs peut se faire coordonn\'ee par coordonn\'ee, c-\`a-d, grosso modo, par des ajustements unidimensionnels peu co\^uteux.

La seule innovation n\'ecessaire, par rapport au FF-EP de base, c'est l'approximation du facteur lui-m\^eme dans l'approximation contextuelle. On a vu en quoi cette id\'ee permet de r\'eduire la dimensionalit\'e du probl\`eme d'ajustement par divergence KL et de faire de l'approximation variationnelle {\em low cost}.

Une subtilit\'e qui m'avait \'echapp\'e lorsque je travaillais sur FF-EP, c'est que la m\'ethode de Laplace (qu'on peut voir comme une alternative \`a l'ajustement via KL) garantit elle aussi la s\'eparabilit\'e de l'ajustement des applications coordonn\'ees (\`a condition de faire le DL au centre de la cavit\'e). En revanche, les m\'ethodes de quadrature n'ont pas cette propri\'et\'e. On le voit ais\'ement avec la quadrature gaussienne: chaque moment est estim\'e par une somme pond\'er\'ee de $2d+1$ points o\`u toutes les dimensions varient \`a tour de r\^ole.

Pour \'eviter cette lourdeur, il suffit de recourir \`a notre petite astuce: imposer la forme s\'eparable au facteur et remarquer que, d\`es lors, l'ajustement via KL est lui-m\^eme s\'eparable. Le gain en efficacit\'e est proportionnel \`a la dimension, donc \'enorme en grande dimension!

Intuitivement, on ne perd pas grand chose \`a ajuster chaque application coordonn\'ee l'une apr\`es l'autre. C'est aussi ce que fait la m\'ethode de Laplace ``gratuitement'' au sens o\`u elle donne une approximation quadratique {\em non-s\'eparable} dont on d\'eduit naturellement une approximation factorielle en ignorant les termes non-diagonaux du hessien. C'est ``\'evident'' mais je me rends compte que je ne m'\'etais pas fatigu\'e \`a justifier cela dans mon c\'el\`ebre article.

Une justification, celle que j'avais en t\^ete je crois, c'est de dire qu'on applique la formule de mise \`a jour de FF-EP \`a l'approximation de Laplace du facteur, ce qui a pour seul effet d'annuler les termes non-diagonaux du hessien. Pourquoi? Parce que c'est une propri\'et\'e de la gaussienne: en annulant les moments crois\'es, on ne modifie pas les moments non-crois\'es.

Une autre justification, c'est d'appliquer la m\'ethode de Laplace \`a l'approximation contextuelle, ce qui produit le m\^eme r\'esultat mais est plus coh\'erent avec l'id\'ee que la m\'ethode de Laplace remplace la minimisation de la divergence KL.

En d'autres termes, la g\'en\'eralisation de l'approximation contextuelle EP \'etait d\'ej\`a plus ou moins sous-jacente \`a mon article de 2016...

\appendix

\section{Optimal sampling distribution}
\label{app:opt_sampling}

For random sampling, it can be shown that the VS solution achieves a KL divergence larger than the actual minimum by a quantity vanishing at asymptotic rate $\nu/N$ with:
$$
\nu = \frac{1}{2}\int \frac{[p(\x)-q_\star(\x)]^2}{s(\x)} \psi(\x)^2 d\x,
\qquad 
\psi(\x)^2 = \sum_j \phi_j(\x)^2,
$$
where $q_\star$ is the optimal factor, i.e. $q_\star=\arg\min_q D(p\|q)$, and $\phi_j$ is an orthornomal basis for the dot product induced by~$q_\star$:
$$
\int q_\star \phi_j \phi_k  = \delta_{jk} 
\qquad \Rightarrow \quad
\int cg_\star \psi^2 = n,
$$
where $n$ is the number of free parameters.

Minimizing $\nu$ wrt $s$ is akin to optimal design in importance sampling. The optimal sampling distribution is: $s_\star \propto |p-q_\star|\psi$. As can be seen, it is not simply $p$ as could perhaps be expected or hoped. Note that the equivalent IS problem is to compute the integral:
$$
\int (p-q_\star)\psi,
$$
but this is purely theoretical as we don't know $q_\star$ (yet we do know that this integral is zero).

Now it is time to remember that we may have a context, meaning that $p=cf$ and $q=cg$. We can rewrite the same story using the minimizer $g_\star$ of $D(cf\|cg)$ and the same kind of orthonormal basis built from $cg_\star$:
$$
\nu = \frac{1}{2}\int \frac{c^2}{s} (f-g_\star)^2 \psi^2,
$$
and the optimal sampling distribution is $s_\star \propto c|f-g_\star|\psi$. If the functions under consideration vary slowly compared to $c$, then $c$ should be a fair approximation to $s_\star$. Using $c$ as a sampling distribution, we have:
$$
\nu = \frac{1}{2}\int c (f-g_\star)^2 \psi^2
$$

Letting $\omega=cg_\star\psi^2/n$, which is a density, i.e. $\int \omega = 1$, we see that the variance factor is proportional to the chi-square divergence between the factors in the context $\omega$: 
$$
\nu 
= \frac{n}{2} \int \omega \frac{(f-g_\star)^2}{g_\star}
= \frac{n}{2} D_2(\omega f\| \omega g_\star)
.
$$

So we have an interesting outcome here: the variance relates to the contextual divergence between factors, but this is neither the same divergence nor the same context as the initial ones (those considered for minimization): $\chi_2$ instead of KL and $\omega$ instead of $c$... Yet, we can say loosely that $\nu$ should be smaller as the context gets narrower, hence VS should be faster to converge.

If we are not in ``narrow context'' regime, then we are kinda screwed, right? But we can try and generalize the above argument:
$$
\nu = \frac{n}{2}\int \omega \frac{c}{s} \frac{(f-g_\star)^2}{g_\star}
.
$$

So, now, we see that this is the same story as before by redefining the context as $\omega'=\omega c / s$, which may no longer be a density. If we set $s=c g_\star/z$ with the constant $z$ so that $\int s = 1$, then the context simplifies to $\omega'=(z/n) c \psi^2$, hence:
$$
\nu = \frac{z}{2} D_2(c' f\| c' g_\star),
\qquad
c' \equiv c\psi^2
.
$$

It now looks like we have a similar interpretation as above, with the exception that $n$ is replaced by $z$, so the variance does no longer depend on the family dimension! On the other hand, the chi-square divergence involved here may not be as small as one would like since we assume a possibly broad context.

Those are just some insights on the effect of the sampling distribution. In summary: if the context is narrow, it is a good choice as the sampling distribution; if not, then multiply it by some initial guess of the factor fit... and pray cause there is no warranty.

The conclusion is that VS may be inefficient unless used in a certain {\em context}. A context is a region wider than an infinitesimal neighborhood but can't be too wide. If very narrow, VS will most likely work very well but will ressemble the Laplace method. If broader, VS efficiency will heavily rely on the availability of a good approximation to the target.

\bibliographystyle{abbrv}
\bibliography{cvis,stat,alexis}

\end{document}

### 2018-09-13 notes/variational_point_matching.tex
Add some notes

\documentclass{article}

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,subfigure}

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\def\s{{\mathbf{s}}}
\def\r{{\mathbf{r}}}
\def\I{{\mathbf{I}}}
\def\eps{{\boldsymbol{\varepsilon}}}
\def\m{{\boldsymbol{\mu}}}
\def\param{{\boldsymbol{\theta}}}

\title{Point matching notes} 
\author{Alexis Roche} 
%\date{}

\begin{document}

\maketitle

\section{Introduction}

This is where I say some bs.

\section{Generative model}

\subsection{Gaussian point pair model}

Assume that corresponding points $(\x,\y)$ in $\mathbb{R}^d$ are random variables defined by:
\begin{eqnarray*}
\x & = & \s + \eps\\
\y & = & \s + \eps'
\end{eqnarray*}
with $\s\sim N(\m, v \I_d)$, $\eps\sim N(0, w \I_d)$, $\eps'\sim N(0, w \I_d)$, $\eps\perp \s$, $\eps'\perp \s$, $\eps'\perp \eps$. It follows that $(\x,\y)$ is distributed like a $2d$ Gaussian variable with mean $(\m,\m)$ and variance matrix:
$$
V = w \I_{2d} + v E,
\qquad
E \equiv
\left(
\begin{array}{cc}
\I_d & \I_d\\
\I_d & \I_d
\end{array}
\right)
.
$$

The determinant of $V$ is: 
$$
|V| = \left[ (v+w)^2 - v^2 \right]^d
$$

Remarking that $E^2=2E$, we easily find the inverse variance matrix to be:
$$
V^{-1} 
= 
\frac{1}{w} \left(\I_{2d} - \frac{v}{w+2v} E \right)
.
$$

Introducing the variance ratio $\lambda = v/w$, these expressions read:
$$
|V| = w^{2d} (1 + 2\lambda)^d,
\qquad
V^{-1} = 
w^{-1} \left[ 
\I_{2d} - \lambda (1+2\lambda)^{-1} E
\right] 
.
$$

Letting now $\nu=\lambda (1+2\lambda)^{-1}$ so that $\lambda=\nu(1-2\nu)^{-1}$, these read:
$$
|V| = w^{2d} (1 - 2\nu)^{-d},
\qquad
V^{-1} = 
w^{-1} \left[ 
\I_{2d} - \nu E
\right] 
.
$$

Note that $(\x\ \y)^\top E (\x\ \y)= \|\x+\y\|^2$. Therefore, the negated log-likelihood $\ell(\param) = -2\log p(\x,\y)$ associated with a single pair is, up to an additive constant,
\begin{eqnarray*}
\ell(\param)
& = & 
2d \log w - d \log(1-2\nu) 
+ w^{-1} \left[
\|\x-\m\|^2 + \|\y-\m\|^2 - \nu \|\x+\y-2\m\|^2
\right] \\
& = & 
2d \log w - d \log(1-2\nu)
+ w^{-1} \left[
(1-\nu) (\|\x-\m\|^2 + \|\y-\m\|^2) 
- 2 \nu (\x-\m)^\top(\y-\m)
\right]
,
\end{eqnarray*}
where $\param=(\m, w,\nu)$ stands for the aggregated parameter vector. Derivatives are computed as follows.

\begin{eqnarray*}
\frac{\partial \ell}{\partial \m}
& = & 
w^{-1} \left[
2(1-\nu)(\m - \x + \m - \y) 
- 2\nu (2\m - \x - \y) \right] \\
& = & 
4 w^{-1} (1-2\nu) 
\left(\m - \frac{\x + \y}{2} \right)
.
\end{eqnarray*}

Denoting $R=(\|\x-\m\|^2+\|\y-\m\|^2)/2$ and $C=(\x-\m)^\top(\y-\m)$,

$$
\frac{\partial \ell}{\partial w}
=
2d w^{-1} - 2w^{-2}
[(1-\nu) R - \nu C]
.
$$

$$
\frac{\partial \ell}{\partial \nu}
=
2d (1-2\nu)^{-1} - 2w^{-1}(R + C)
.
$$

\subsection{Parameter estimation}

Now, assume that we are given a set of i.i.d. pairs $(\x_k,\y_k)$, $k=1,\ldots,n$. We want to estimate the $\param$ by maximum likelihood, which leads to minimizing:
$$
L(\param) 
= \sum_k \ell_k(\param)
.
$$

Let us denote $\bar{\x}$, $\bar{\y}$, $\bar{C}$ and $\bar{R}$ the sample means of $\x$, $\y$, $C$ and $R$, respectively. From the above derivative expressions, we clearly have:
$$
\frac{\partial L}{\partial \m} = 0 
\quad \Rightarrow \quad
\m = \frac{1}{2} (\bar{\x} + \bar{\y})
.
$$

Furthermore, 
$$
\frac{\partial L}{\partial w}=0
\quad \Rightarrow \quad
dw = (1-\nu)\bar{R} -\nu\bar{C}.
$$

$$
\frac{\partial L}{\partial \nu}=0
\quad \Rightarrow \quad
dw = (1-2\nu)(\bar{R}+\bar{C}).
$$

By equating these two last expressions, we find that:
$$
\nu = \frac{\bar{C}}{\bar{R}+\bar{C}}
$$

It follows that:
$$
1-2\nu = \frac{\bar{R}-\bar{C}}{\bar{R}+\bar{C}}
\quad
\Rightarrow
\quad
\lambda = \nu(1-2\nu)^{-1} = \frac{\bar{C}}{\bar{R}-\bar{C}}
$$

Therefore, 
$$
w = \frac{1}{d} (\bar{R} -\bar{C}),
\qquad
v = \frac{1}{d} \bar{C}
.
$$

\subsection{Estimation with fixed parameters}

If $\mu$ is fixed, $\bar{R}$ and $\bar{C}$ are computed using the fixed $\mu$ rather than the ML estimate.

If $w$ is fixed and $v$ is not, we need to estimate $v$ via $\nu$ from the equation $\partial L / \partial \nu = 0$, yielding:
$$
(1-2\nu)^{-1}
= (1+2\lambda)
= \frac{\bar{R} + \bar{C}}{dw}
\quad\Rightarrow\quad
v 
= \frac{1}{d}(\bar{R} + \bar{C}) - \frac{w}{2} 
= v_{ML} + \frac{w_{ML} - w}{2}.
$$

If $v$ is fixed and $w$ is not, things look more tricky but this case is not of practical interest...

\subsection{Marginal and conditional distributions}

Under the above model, the marginal distributions of $\x$ and $\y$ are the same Gaussian with mean $\m$ and variance $(v+w)\I_d$.

For the conditionals, it is clear that $p(\x|\y)$ and $p(\y|\x)$ are the same Gaussian distribution up to the permutation of $\x$ and $\y$. The mean and variance are given by the generic formula:
$$
\m_{x|y} = \m_x + \Sigma_{xy}\Sigma_{yy}^{-1}(\y-\m_y),
\qquad
\Sigma_{x|y} = \Sigma_{xx} - \Sigma_{xy}\Sigma_{yy}^{-1}\Sigma_{yx}
.
$$

In this case, 
$$
\m_x=\m_y=\m,
\qquad
\Sigma_{xy}=\Sigma_{yx}=v\I_d,
\qquad
\Sigma_{xx}=\Sigma_{yy}=(v+w)\I_d,
$$
therefore:
$$
\m_{x|y} = \alpha\y + (1-\alpha)\m,
\qquad
\alpha = \frac{v}{v+w},
$$
and:
$$
\Sigma_{x|y}
=
[v+w - v^2 (v+w)^{-1}]\I_d
= 
(v+w) (1-\alpha^2) \I_d
.
$$

\subsection{Null model}

Consider a trivial model variant whereby points stem from independent seeds:
\begin{eqnarray*}
\x & = & \s + \eps\\
\y & = & \s' + \eps'
\end{eqnarray*}
with $\s'\sim \s$ and $\s\perp s'$, implying that $\x\perp \y$ with the same marginal distribution $N(\m, (v+w)\I_d)$. The twice negated log-likelihood then simplifies to:
$$
\ell(\param)
=
2d \log (v+w) 
+ (v+w)^{-1} \left[
\|\x-\m\|^2 + \|\y-\m\|^2 
\right]
.
$$

Note that $v$ and $w$ are not separately identifiable, but the sum of the two is. The maximum likelihood parameter estimates are easily seen to be:
$$
\m = \frac{1}{2} (\bar{\x} + \bar{\y}),
\qquad
v+w = \frac{1}{d} \bar{R} 
.
$$

Therefore, the ML distribution can be alternatively computed by taking the ML distribution corresponding to the point pair model, and then replacing it by the product of its marginals.

\section{Matching algorithm}

Let us turn to assuming that two point sets $(\x_1,\ldots,\x_n)$ and $(\y_1,\ldots,\y_m)$ are related by a correspondence function $q:\Omega_q\subset\{1,\ldots,n\} \to \{1,\ldots,m\}$ and a spatial transform $T:\mathbb{R}^2\to \mathbb{R}^2$, so that, roughly speaking, $T(\x_k)\approx \y_{q(k)}$. The subset $\Omega_q$ defines the source points with a correspondence, meaning that missing correspondences are accounted for.

Let us assume that, for given $q$ and $T$, the set of matched point pairs $(T(\x_k),\y_{q(k)})_{k\in\Omega}$, for is independently and identically distributed according to the above joint Gaussian model, which assumes, in particular, that $q$ is a one-to-one mapping. This generative model can be justified as ``minimally discriminative'' in the sense that it minimizes the KL divergence to the set of distributions assuming statistical independence between the two point sets. The distribution of each point pair is thus:
$$
p_T(\x_k, \y_{q(k)}) = |J_T(\x_k)| p(T(\x_k), \y_{q(k)})
,
$$
where $J_T(\x)$ is the transformation Jacobian matrix.

Using the PDF projection theorem (or the super composite likelihood framework), we can define a likelihood function for $(q,T)$:
$$
L(q,T) = \prod_{k\in\Omega_q} \frac{p(T(\x_k), \y_{q(k)})}{\pi(T(\x_k), \y_{q(k)})},
$$
where $\pi$ is some reference distribution representing a null hypothesis. Note that the Jacobian terms in the numerator and denominator cancel out. Accouting for the fact that distribution parameters are unknown, the (composite) log-likelihood function is a function of multiple variables:
$$
L(q,T,\param,\param_0) 
= 
\prod_{k\in\Omega_q} 
\frac{p_\param(T(\x_k), \y_{q(k)})}{\pi_{\param_0}(T(\x_k),\y_{q(k)})}
.
$$

In this story, $\param_0$ plays a bit of an unusual role because the composite likelihood is to be {\em minimized} along $\param_0$. The rational behind this is the elimination of nuisance parameters in super composite likelihood models. The joint estimation of all parameters can be seen in terms of Game theory. It is a four-player game, where each parameter is associated with a ``player''. The payoff for the players associated with $q$, $T$, and $\param$ is the log-likelihood, and the negated log-likelihood for the player associated with $\param_0$.

This special player accounts for the fact that the null model is not fully specified. As a null hypothesis, we assume that the point pairs are statistically independent consistently with the minimum discrimination property of the generative model. Plugging in the likelihood the minimal $\param_0$ while fixing all other parameters boils down to taking the product of the marginals:
$$
\min_{\param_0}\max_\param
L(q,T,\param,\param_0) 
= 
\prod_{k\in\Omega_q} 
\frac{p_{\hat{\param}}(T(\x_k), \y_{q(k)})}{p_{\hat{\param}}(T(\x_k))p_{\hat{\param}}(\y_{q(k)})}
.
$$

Let us redefine the log-likelihood ratio:
$$
R(q,T,\param) = \prod_{k\in\Omega_q} 
\frac{p_{\param}(T(\x_k), \y_{q(k)})}{p_{\param}(T(\x_k))p_{\param}(\y_{q(k)})}
$$

The above shows that the game can alternatively be seen as a three-player game with players respectively associated with $q$ and $T$ have $R(q,T,\param)$ as a payoff, and a third player associated with $\param$, which seeks to maximize the numerator only, leading to the maximum likelihood parameter estimates computed above.

\section{Registration}

Let us examine the likelihood ratio maximization along $T$, which amounts to maximizing:
$$
C(T) = 
\prod_{k\in\Omega_q} 
p_{\param}(\y_{q(k)}|T(\x_k))
.
$$

Recall that $p(\y|\x)$ is a Gaussian distribution with mean $\mu_{\y|\x} = \alpha\x + (1-\alpha)\m$ and isotropic variance matrix. By noting the identity:
$$
\alpha \x + (1-\alpha) \m - \y 
= \alpha \{ \x - \alpha^{-1}[\y - (1-\alpha) \m)]\},
$$
we conclude that the maximization of $T$ amounts to a least square problem where the target points are corrected via:
$$
\y' = \alpha^{-1} [ \y - (1-\alpha) \m ]
.
$$

Hence the optimal transformation $T$ is found by minimizing least squares:
$$
C(T) = 
\sum_k
\|T(\x_k)-\y_k'\|^2
$$

Assuming an affine transformation $T(\x)=A\x + t$, the Jacobian matrix is constant equal to $A$. We can ignore the translation in the variational problem by centering the points, that is, computing $\tilde{\x}_k = \x_k - \bar{\x}$ and $\tilde{\y}'_k = \y'_k - \bar{\y}'$ with:
$$
\bar{\x} = \frac{1}{n} \sum_k \x_k,
\qquad
\bar{\y}' = \frac{1}{n} \sum_k \y'_k
.
$$

Once the optimal $A$ is found, the optimal translation is then given by $t = \bar{\y}'- A \bar{\x}$.

\subsection{Rigid transform}

If $T$ is searched among rigid transforms, the solution is found by computing the SVD of:
$$
K = \frac{1}{n}\sum_k \tilde{\y}'_k \tilde{\x}_k^\top
$$

Specifically, the SVD gives orthogonal matrices $U$ and $V$ and a diagonal matrix $S$  with positive elements such that $K = U \Lambda V$. The optimal rotation is then $R = U {\rm diag}(1,\ldots,1,|U||V|) V$.

\subsection{Scaling transform}

In the case of a scaling transform, we can decompose the unknown matrix as $A=sR$. The optimal rotation is, again, found by standard least sqaures and is easily seen to be the same as in the rigid case (see Pennec). For the optimal scale $s$, we need to minimize:
$$
C(s) = 
\sum_k \|s R\tilde{\x}_k-\tilde{\y}_k'\|^2
.
$$

By developing the squared Euclidean distance term, we see that, up to a constant,
$$
C(s)
= 
n(\sigma_x^2 s^2 - 2\gamma s),
$$
with:
$$
\sigma_x^2 = \frac{1}{n} \sum_k \|\tilde{\x}_k\|^2,
\qquad
\gamma = {\rm trace}(RK^\top)
,
$$
implying that the optimal scale is given by:
$$
s = \frac{\gamma}{\sigma^2_x}
.
$$

\subsection{Affine transform}

In the case of an unconstrained affine transformation, the optimality condition is readily obtained by applying rules of derivation w.r.t. matrices:
$$
A \Sigma_x - K = 0
$$
with:
$$
\Sigma_x = \frac{1}{n} \sum_k \tilde{\x}_k \tilde{\x}_k^\top
.
$$

It follows that the optimal matrix $A$ is given by:
$$
A = K \Sigma_x^{-1}
.
$$

\section{MaxEnt}

Assume we have a pseudo generative model that minimizes the discrimination $D(p_\theta \| \pi)$ on a certain subspace $P_\theta$ for a given $\theta$. Under linear moment constraints, the solution has the exponential form:
$$
p_\pi (y|\theta,\lambda) = \frac{1}{Z_\pi(\theta,\lambda)} \pi(y) e^{-\lambda t(y,\theta)}
$$

This can be seen using the Lagrangian:
$$
{\cal L}(p,\lambda) = 
D(p\|\pi) + \lambda [\int p(y)t(y,\theta)dy - \tau] dy
,
$$
which is minimized at fixed $\lambda$ (and $\theta$) by the above distribution. The achieved Lagrangian minimum defines the dual function at fixed $\theta$:
\begin{eqnarray*}
\psi(\theta,\lambda) = 
& = & 
\min_p {\cal L}(p,\lambda) \\
& = & 
-\log Z_\pi(\theta,\lambda) - \lambda \tau \\
& = & 
\log\frac{p_\pi(y|\theta,\lambda)}{\pi(y)} + \lambda [t(y,\theta)-\tau],
\quad \forall y.
\end{eqnarray*}

This is true for any $y$, and in particular for the one observed. The dual function amounts to the log-likelihood if we equate $\tau$ with its empirical value (meaning that $\tau$ depends on $\theta$). In this story, $\theta$ is fixed but we see that the dual function, as a function of $(\theta,\lambda)$, amounts to the joint log-likelihood:
$$
\psi(\theta,\lambda) 
= 
\log\frac{p_\pi(y|\theta,\lambda)}{\pi(y)}
.
$$

For instance, we can apply an EM algorithm with $\theta$ as a latent variable. This boils down to maximizing a soft version of the dual function: 
$$
\Psi(q, \lambda) = \int q(\theta) \left[
\psi(\theta,\lambda)- \log q(\theta)
\right]
d\theta
.
$$

How does this reasoning generalize when $\pi$ is searched in a certain set of distributions $P_0$? The dual function then becomes a function of $(\theta,\lambda)$ and $\pi$. At fixed $\theta$, $\lambda$ and $\pi$ are concurrently optimized, which ends up defining two parametric families $p(y|\theta)$ and $\pi(y|\theta)$. Under mild regularity conditions, we have:
$$
\max_\lambda \psi(\theta,\lambda,\pi) 
= \min_{p\in P_\theta} \psi(\theta,\lambda,\pi)
.
$$

Therefore, the optimal $\pi$ corresponds to a minmax problem:
$$
\min_{\pi\in P_0}
\min_{p\in P_\theta} 
D(p\|\pi) 
= 
\min_\pi \max_\lambda \psi(\theta,\lambda,\pi)
.
$$

We might be able to replace $\pi$ with a parameter $\lambda_0$ so that we have a $\psi(\theta,\lambda,\lambda_0)$. Under

\section{Variational point matching}

$\theta=(M,T)$ represents the pointwise correspondence function and the spatial transform, both of which are unknown. If we focus on matching, $M$ is the variable of main interest and $T$ is more like a nuisance parameter. We also have distribution parameters $(\lambda,\lambda_0)$. The lack of a fixed reference comes from the fact that the null hypothesis is that of statistical independence between the two point sets, which does not specify a distribution.

When updating the distribution parameters, we take the

We can have an EM using probablistic correspondences but pointwise transform estimates. Why? Because the correspondences are the bottleneck in terms of optimization since they turn the problem into a discrete optimization one. It's a way to smooth it.

\section{Rubbish}

$$
A \Sigma_x A^{\top} - K A^{\top} -w^{-1} (1-\nu) \I_d  = 0
$$

Note that:
$$
\frac{\nu}{1-\nu}
=
\frac{v}{v+w}
$$

\appendix

\section{Multivariate Gaussian}

The $n$-dimensional normal distribution $N(\m,V)$ is defined by:
$$
-2\log p(\x)
= 
n\log(2\pi)
+\log |V|
+ (\x-\m)^\top V^{-1} (\x-\m),
$$

\section{Derivatives wrt a matrix}

Consider the squared Euclidean distance: 
$$
C = \| T(\x) - \y \|^2
$$

Assume a pure linear transform to start.
$$
T(\x) = A\x
$$

$$
C = 
\x^\top A^\top A \x - 2 \y^\top A \x
+ \y^\top \y 
$$

Formula can be found in Pennec's thesis.
$$
\frac{\partial (\log |A|)}{\partial A}
= A^{-\top}
$$

$$
\frac{\partial (\y^\top A \x)}{\partial A}
= \y \x^\top,
\qquad
\frac{\partial (\y^\top A^\top B A \x)}{\partial A}
= B A \x \y^\top + B^\top A \y \x^\top
$$

This implies:
$$
\frac{\partial C}{\partial A}
=
2 A \x \x^\top - 2 \y \x^\top
$$

Least square optimality condition.

\end{document}

### 2018-10-27 notes/composite_inference.tex
Convert text file to TEX

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\title{Inf\'erence composite bay\'esienne (ou pas)}
\author{Alexis Roche}

\begin{document}

\maketitle

\section{Introduction}

Le concept de mod\`ele composite... Il faut le voir comme une m\'ethode d'inf\'erence qui proc\`ede par aggr\'egation d'inf\'erences binaires. Chaque inf\'erence binaire est r\'ealis\'ee par un "expert" utilisant {\em sa propre repr\'esentation} et {\em ind\'ependant} au sens o\`u ses r\'esultats ne d\'ependent pas des autres experts.

Chaque expert "vit" dans un monde des possibles r\'eduit \`a deux hypoth\`eses, dont l'une est commune \`a tous les experts. Plus concr\`etement, sa mission est de calculer un rapport de probabilit\'es entre l'hypoth\`ese dont il est charg\'e et l'hypoth\`ese de r\'ef\'erence commune \`a tous les experts. C'est ensuite au "coordinateur" de construire une fonction de probabilit\'e multi-classes en collectant les r\'eponses des experts.

La base de donn\'ees d'apprentissage, si elle existe, est centralis\'ee. Tous les experts y ont acc\`es, mais formattent les donn\'ees \`a leur fa\c{c}on, c-\`a-d selon leur repr\'esentation. Chaque expert effectue son apprentissage de fa\c{c}on classique, dans son monde propre, avec sa repr\'esentation propre voire son propre sous-ensemble de donn\'ees.

Dans le cas de l'apprentissage "\`a la vol\'ee" (qui n'a de sens que dans le cas o\`u chaque expert utilise un mod\`ele g\'en\'eratif), c'est la m\^eme id\'ee de base: chaque expert \'elimine les param\`etres de son mod\`ele selon sa propre repr\'esentation et \'evalue son propre rapport de probabilit\'es que le coordinateur utilise exactement comme dans le cas de l'apprentissage s\'epar\'e.

En d'autres termes, le coordinateur ne sait pas comment les param\`etres du mod\`ele sont \'elimin\'es. Il en d\'el\`egue la responsabilit\'e aux experts. Sa responsabilit\'e \`a lui est de combiner les rapports de probabilit\'es fournis par les experts.

\section{Lien avec PDF projection}

Il faut d\'ej\`a comprendre en quoi ce paradigme certes intuitif diff\`ere du paradigme classique pour l'inf\'erence multi-classes. Supposons qu'on ait un mod\`ele g\'en\'eratif de la forme:

$$
p(y|x) = a(y) b(z_x, x),
$$ 
o\`u $zx = z(x, y)$ est un "attribut contextuel", on prouve facilement que cette forme se ram\`ene \`a celle du PDF projection theorem: 
\begin{equation}
\label{eq:pdf_proj}
p(y|x) = \pi(y) \frac{p(z_x|x)}{\pi(z_x)}
\end{equation}

Une telle structure peut apparaitre naturellement dans un mod\`ele ou r\'esulter de l'application du principe d'entropie maximale si on s\'electionne le mod\`ele en supposant connues les distributions de l'attribut contextuel, $p(z_x|x)$. Notons que dans ce dernier cas, la distribution de r\'ef\'erence $\pi(y)$ est arbitraire.

Cette strat\'egie d'entropie maximale d\'ecoule en fait d'un probl\`eme de th\'eorie des jeux d\'ecrit par Gr\"unwald, 2005: il s'agit de simuler des donn\'ees en l'absence d'un mod\`ele g\'en\'eratif complet. Le PDF projection n'est donc qu'une solution optimale pour un probl\`eme de synth\`ese de donn\'ees.

Mais est-ce qu'on peut invoquer le PDF projection pour faire de l'inf\'erence? Dans son article de 2005 (je crois), Baggenstoss observe que le PDF projection est exact si la repr\'esentation est exhaustive. Autrement dit, on peut se passer du principe de maximum d'entropie si on suppose l'exhaustivit\'e de la repr\'esentation. C'est une justification alternative mais plus fondamentale dans le contexte de l'inf\'erence.

\section{Maximum de vraisemblance compl\`ete: exemple}

Supposons qu'on ait la forme (\ref{eq:pdf_proj}). Si on traite p(y|x) de fa\c{c}on classique, sans tenir compte de cette structure, on voit que l'estimation des param\`etres du mod\`ele par maximum de vraisemblance (par exemple) conduit \`a:

$$
\max_{\alpha, \beta} \left[
\sum_k \log \pi_\alpha(y_k)/\pi_\alpha(z_{x_k}) + \sum_k \log p_\beta(z_{x_k}|x_k)
\right]
$$

En supposant que les param\`etres alpha de pi et beta des autres lois g\'en\'eratives sont disjoints et que pi est la loi de la classe 0, cela revient \`a maximiser chaque terme ind\'ependamment, donc notamment:

$$
\max_\alpha \left[
\sum_k f_k(\alpha)
\right]
$$ 
avec: $ f_k(\alpha) = \log \pi_\alpha(y_k)$  si $x_k = 0$ et $f_k(\alpha) = \log \pi_\alpha(y_k)/\pi_\alpha(z_{x_k})$ si $x_k > 0$.

Supposons par exemple qu'on a 3 classes et que $y = (z_1, z_2)$ sont ind\'ependants dans la classe 0, {\em i.e.} $\pi(z_1, z_2) = \pi(z_1)p(z_2)$. On a alors: $f_k(\alpha) = \log \pi_\alpha(z_{1k})\pi_\alpha(z_{2k})$ si $x_k=0$, $f_k(\alpha) = \log \pi_\alpha(z_{2k})$ si $x_k=1$,  et $f_k(\alpha) = \log \pi_\alpha(z_{1k})$ si $x_k=2$.

Si on suppose en outre que pi(z1) et pi(z2) correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution "nulle" de z1 (l'attribut "exhaustif" pour la classe 1) sur les donn\'ees issues des classes 0 et 2...

Que vient faire la classe 2 dans cette histoire??? N'est-il pas absurde d'utiliser des donn\'ees de la classe 2 pour ajuster la loi de la classe 0?

Non car on peut se rendre compte que le mod\`ele implique que z1 est distribu\'e de la m\^eme fa\c{c}on dans les classes 0 et 2:

$$p2(z1, z2) = pi(z1) p2(z2) ==> p2(z1) = pi(z1)$$

Il est judicieux, dans ce cas, de ne pas se restreindre aux donn\'ees de la classe 0 pour ajuster sa distribution. Sans surprise, la vraisemblance donne lieu \`a une estimation statistiquement efficace \`a d\'efaut d'\^etre calculatoirement simple. Bien s\^ur, si le mod\`ele n'est pas r\'ealiste (en d'autres termes, si l'attribut contextuel n'est pas exhaustif pour chaque classe), alors ce n'est pas une super id\'ee de m\'elanger les donn\'ees, mais ce n'est peut-\^etre pas non plus une super id\'ee d'utiliser l'approche composite...

\section{Limitations de la vraisemblance compl\`ete}

Est-ce \`a dire que le maximum de vraisemblance "compl\`ete" est toujours la meilleure approche? On a d\'ej\`a compris que son application emp\^eche les experts d'\^etre ind\'ependants. En effet, l'ajustement de la loi nulle doit \^etre centralis\'e car il fait intervenir des informations que les experts n'ont pas. Dans l'exemple ci-dessus, l'expert pour la classe 1 ne peut ajuster pi(z1) lui-m\^eme car il a besoin pour \c{c}a des valeurs de z2.

Pour avoir un mod\`ele complet, on doit avoir une repr\'esentation "originelle" des donn\'ees, c-\`a-d qu'on doit d\'efinir la variable y alors que celle-ci n'a n'a pas lieu d'\^etre dans l'approche composite. L'ajustement des lois $\pi(z_x)$ via la vraisemblance compl\`ete est-il ind\'ependant de la repr\'esentation "originelle"? La r\'eponse est non. Si on inclut dans y un attribut z3 qui a, par exemple, la m\^eme variance que z1 sous l'hypoth\`ese de r\'ef\'erence alors, clairement, l'estimation par maximum de vraisemblance "compl\`ete" de pi(z1) va d\'ependre de z3... c-\`a-d qu'on devra mesurer z3 pour pouvoir ajuster cette loi.

Ce probl\`eme n'est pas sp\'ecifique \`a notre sujet d'\'etude. On sait bien que toute inf\'erence statistique d\'epend du choix d'une repr\'esentation. Si les maximums de vraisemblance marginale et jointe donnaient les m\^emes r\'esultats, \c{c}a se saurait. Et il est bien naturel que ce ne soit pas le cas car l'int\'er\^et de choisir une repr\'esentation plus riche est justement d'incorporer plus d'information - encore faut-il acqu\'erir les donn\'ees qui vont avec.

Imaginons maintenant que la repr\'esentation choisie soit de grande dimension et implique des tas de param\`etres additionnels... Est-ce que \c{c}a peut conduire \`a un mauvais ajustement des lois $\pi(z_x)$ \`a cause du curse of dimensionality? Je ne pense pas.

Reste une difficult\'e conceptuelle avec la vraisemblance compl\`ete: le mod\`ele stipule que tel attribut est suffisant pour tester telle hypoth\`ese. Mais l'estimation des param\`etres remet en cause ce principe.

\section{Vraisemblance r\'eduite}

Mes r\'eflexions de la fin de l'\'et\'e 2017 m'avaient conduit \`a r\'ealiser que l'approche composite \'equivaut \`a une m\'ethode d'\'elimination de param\`etres alternative applicable \`a tout mod\`ele g\'en\'eratif complet d\`es lors qu'il admet une repr\'esentation adaptive r\'eduite (ce qui est toujours le cas, en fait!). Appelons-la vraisemblance r\'eduite. Il y a quelques similitudes avec le concept de vraisemblance restreinte mais ce n'est pas tout-\`a-fait la m\^eme chose.

La vraisemblance r\'eduite garantit l'exhaustivit\'e EN UN SENS FORT: en clair, elle laisse chaque expert \'eliminer les param\`etres par ses propres moyens. Contrairement \`a la vraisemblance compl\`ete, elle ne fournit pas un estimateur unique des param\`etres de la distribution de r\'ef\'erence (chaque expert y va de son estimateur...). Si les param\`etres des distributions de classe sont li\'es, c'est encore plus le bordel.

Mais quelle importance si le but assum\'e est d'estimer les rapports de vraisemblance:

$$R(x) = p(y|x) / pi(y) = pi(z_x|x) / pi(z_x)$$

La vraisemblance r\'eduite permet de le faire sans exploiter toute l'information contenue dans les donn\'ees mais avec plusieurs avantages:

\begin{itemize}
\item exhaustivit\'e forte 
\item coh\'erence asymptotique (et oui, c'est maintenu car la vraisemblance   restreinte reste la vraisemblance...) 
\item ind\'ependance vis-\`a-vis de toute repr\'esentation "universelle" 
\item flexibilit\'e: ind\'epedendance des experts et donc possibililt\'e de   combiner leurs r\'esultats sans connaitre les donn\'ees utilis\'ees (et   donc notamment s'ils ont utilis\'e des donn\'ees diff\'erentes) 
\item simplicit\'e calculatoire: on n'a m\^eme pas besoin d'un mod\`ele complet,   il suffit de mod\'eliser les attributs sous chaque hypoth\`ese d'int\'er\^et 
\end{itemize}

\section{Le cas du recalage}

En recalage, l'exhaustivit\'e forte signifie que la mesure de similarit\'e ne d\'epend que de ce qui se passe dans la zone de recouvrement entre les images. Propri\'et\'e on ne peut plus naturelle, mais qu'on n'obtient pas avec la vraisemblance compl\`ete.

Je suis d\'esormais convaincu de la validit\'e du mod\`ele g\'en\'eratif "i.i.d" pr\'esent\'e dans ma th\`ese pour le recalage: il est justifi\'e par le th\'eor\`eme d'\'echangeabilit\'e de de Finetti dans la mesure o\`u on suppose que les pixels/voxels sont m\'elang\'es au pr\'ealable, ce qui revient \`a dire que la repr\'esentation "universelle" est un ensemble de pixels (donc sans notion d'ordre).

On peut soit appliquer le th\'eor\`eme dans la r\'egion de recouvrement des images et obtenir ainsi un mod\`ele incomplet, ou l'appliquer globalement en introduisant des distributions pour les points non-appari\'es et obtenir un mod\`ele qui a l'apparence d'une PDF projection sans en \^etre une. Dans le premier cas, on ne peut pas utiliser la vraisemblance compl\`ete car elle n'existe pas mais on peut utiliser la vraisemblance r\'eduite!

Cela dit, autant avoir un mod\`ele complet. La vraisemblance compl\`ete est alors applicable et conduit \`a estimer les distributions des points non-appari\'es pr\'ecis\'ement sur les points non-appari\'es. Mais des propri\'et\'es bizarres apparaissent:

\begin{itemize}
\item La mesure de similarit\'e d\'epend de pixels ext\'erieurs \`a la zone de   recouvrement... 
\item Le nombre de points cibles non-appari\'es est potentiellement infini si l'image cible doit \^etre interpol\'ee. La distribution associ\'ee est alors hyper stable vis-\`a-vis de la transformation mais a aussi un poids infini dans la mesure de recalage... 
\end{itemize}

L'intuition dit qu'il est bien plus naturel d'utiliser la vraisemblance r\'eduite. Tellement plus naturel que c'est une \'evidence. Et c'est ce qui me g\^ene encore un peu dans cette histoire. C'est \'evident mais je n'ai pas d'argument th\'eorique pour rejeter la vraisemblance compl\`ete.

\section{Elimination de param\`etres}

On peut voir l'\'elimination de param\`etres comme un probl\`eme d'approximation ou, si on pr\'ef\`ere, de DECISION, c-\`a-d un probl\`eme de choix d'une valeur unique dans un ensemble. Il s'agit ici de repr\'esenter un mod\`ele param\'etrique par une distribution unique afin d'\'evaluer "l'\'evidence" du mod\`ele, c-\`a-d la probabilit\'e qu'il conf\`ere aux donn\'ees.

Notons la diff\'erence de perspective avec l'estimation ponctuelle, qui cherche directement \`a estimer la variable d'int\'er\^et, par exemple, dans un cadre bay\'esien, par son mode ou sa moyenne a posteriori. Dans ce cas, les param\`etres de nuisance sont \'elimin\'es par marginalisation car ils n'apparaissent pas dans la fonction de perte associ\'ee au probl\`eme de d\'ecision.

Ici, l'objectif est en soi d'estimer l'\'evidence de mani\`ere \`a obtenir une FONCTION DE VRAISEMBLANCE de la variable d'int\'er\^et x. On ne se contente pas d'un estimateur ponctuel de x. Celui-ci sera obtenu comme sous-produit de l'\'elimination de param\`etres, mais d'autres sous-produits seront disponibles, par exemple l'incertidude sur x.

Pour chaque $x$, on repr\'esente donc la famille de distributions sur $y$, $p_\theta(y|x)$, par une distribution unique index\'ee par $x$ qui d\'efinit donc une loi conditionnelle $q(y|x)$ libre de tout param\`etre. Comme on travaille \`a $x$ fix\'e, on peut omettre $x$ dans les notations suivantes par souci de simplicit\'e. Comment choisir cette loi?

\subsection{Elimination bay\'esienne}

Une id\'ee naturelle est de choisir une sorte de centre de la famille, c'est-\`a-dire un \'el\'ement situ\'e \`a faible "distance" de tout autre \'el\'ement. La "distance" en question d\'efinit la fonction de perte et peut \^etre prise, par exemple, au hasard, comme la divergence KL:

$$L(theta, q) = D(p_theta\|q)$$

Dans une perspective bay\'esienne, on minimise le risque int\'egr\'e par rapport \`a la distribution jointe des donn\'ees et des param\`etres, ce qui suppose la d\'efinition d'un a priori pi(theta). On note que la d\'ecision q est ici ind\'ependante des donn\'ees par contrainte du probl\`eme. Ceci donne:

$$R(q) = int pi(theta) p_theta(y) log [p_theta(y) / q(y)] dy dtheta$$

$$==> argmin R(q) = int pi(theta) p_theta(y) dtheta$$

On tombe nez \`a nez avec la marginalisation. Le bon vieux barycentre, quoi... (qui n'appartient pas forc\'ement \`a la famille). Pas besoin d'une grande th\'eorie pour \c{c}a.

On peut cependant obtenir d'autres strat\'egies tout aussi bay\'esiennes en choisissant d'autres divergences; par exemple, la divergence KL invers\'ee:

$$L(theta, q) = D(q||p_theta)$$

$$
R(q) = \int pi(\theta)q(y) log[q(y) / p_\theta(y)] dy d\theta = int q(y) log q(y) dy - \int \pi(\theta) q(y) log p_\theta(y) dy d\theta
$$

$$ dR / dq = \log q(y) + 1 - \int \pi(\theta) \log p_\theta(y) dy ==> \arg\min R(q) = K \exp[\int \pi(\theta) \log p_\theta(y) dy ]$$

La solution est cette fois la moyenne g\'eom\'etrique. Tout \c{c}a pour dire que la marginalisation n'est PAS NECESSAIREMENT la m\'ethode bay\'esienne de choix pour l'\'elimination des param\`etres de nuisance. Tout d\'epend du probl\`eme de d\'ecision et de la fonction de perte associ\'ee.

\subsection{NML}

Il existe une alternative non-bay\'esienne \`a l'\'elimination de param\`etres issue de la communaut\'e minimum description length (MDL): la normalized maximum likelihood distribution (NML). Sa justification classique (cf. Gr\"unwald ou Rissanen) repose sur la notion de "regret", c-\`a-d la diff\'erence de longueur de code entre la distribution caract\'eristique candidate et la distribution qui compresse le plus l'observation:

$$L(y, q) = - log q(y) - min_theta [-log p_theta(y)] = -log q(y) + log ML(y)$$

avec: $ML(y) = max_theta p_theta(y)$

Le maximum de vraisemblance apparait naturellement dans cette formulation en tant que "meilleur mod\`ele" mais c'est un mod\`ele "na\"if" car ML(y) ne d\'efinit pas une distribution pr\'edictive valide \`a cause du couplage entre l'observation et l'estim\'ee MV.

Une diff\'erence essentielle avec l'approche pr\'ec\'edente est que la fonction de perte d\'epend de l'observation. Tout se passe en fait comme si l'observation rempla\c{c}ait theta comme "\'etat inconnu du monde". Cela a du sens dans la mesure o\`u la distribution caract\'eristique doit \^etre choisie ind\'ependamment de toute observation. En fait, on peut consid\'erer que l'\'etat du monde est donn\'e par le couple $(y, theta)$, ce qui permet de d\'efinir le regret pour tout theta:

$$L(y, theta; q) = log p_theta(y) / q(y)$$

D\`es lors, il est naturel de chercher \`a minimiser le "pire" regret:

$$R(q) = \max_{y, \theta} L(y, \theta; q)$$

en notant au passage qu'on ne fait pas l'hypoth\`ese que $y$ est distribu\'e selon un certain $p_\theta(y)$. Cette approche \'evite de choisir des a priori sur $y$ et $\theta$. Comme c'est classique dans ce type de probl\`emes minimax, on probabilise $y$ en introduisant le risque moyen:

$$L(p, theta; q) = E_p [L(y, theta; q)] = int p(y) log p_theta(y) / log q(y) dy = D(p||q) - D(p||p_theta) $$

On a \'evidemment:

$$R(q) = max_p max_theta L(p, theta, q) = max_p L(p, q)$$

o\`u:

$$L(p, q) = int p(y) log ML(y) / q(y) dy$$

Il suffit alors de montrer que les op\'erateurs min et max commutent:

$$min_q R(q) = min_q max_p L(p, q) = max_p min_q L(p, q)$$

On voit alors que quelque soit p, la distribution de Shtarkov (1987) dite NML minimise TOUJOURS $L(p, q)$:

$$\arg\min_q L(p, q) = ML(y) / Z_{nml}$$

avec en plus la propri\'et\'e que le minimum atteint est ind\'ependant de p:

$$min_q L(p, q) = log Znml$$

Ainsi, la distribution NML est solution de notre probl\`eme!

Au-del\`a de la nature "Bayes robuste" de cette formulation, on voit qu'elle consiste plus fondamentalement \`a remplacer $D(p_theta\|q)$ dans la 1\`ere formulation par $D(p\|q) - D(p\|p_theta)$ avec l'id\'ee que $(p,\theta)$ repr\'esente l'\'etat du monde.

\subsection{Crit\`ere d'Akaike}

Une autre approche bien connue est celle d'Akaike, 1974.

L'id\'ee de base est de d\'efinir l'ad\'equation d'un mod\`ele via la divergence KL ou, de fa\c{c}on \'equivalente, la log-vraisemblance moyenne:

$$S(p, p_theta) = int p(y) log p_theta(y) dy$$

plut\^ot que via la notion empirique d'\'evidence. Dans cette perspective, la distribution r\'ealisant le maximum de S sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure "optimiste" qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:

$$S* = max_theta S(p, p_theta)$$

Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la "vraie" distribution des donn\'ees, et que S* est donc inconnu. Mais Akaike justifie l'approximation:

$$S* ~ log ML(y) - k$$

o\`u k est la dimension de theta. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement...

J'ai l'impression qu'il suffit d'appliquer le fameux th\'eor\`eme de Wilks:

$$2 log [ML(y) / p(y)] ~ \chi_2(k)$$

valable asymptotiquement SI $p(y)$ APPARTIENT A LA FAMILLE et sous les conditions usuelles d'\'echantillonnage simple. Wilks nous dit alors que la log-vraisemblance profile sur-estime le score du mod\`ele par la moiti\'e d'un $\chi_2$ \`a $k$ degr\'es de libert\'e. En prenant l'esp\'erance de l'expression ci-dessus, les choses deviennent tr\`es claires:

$$S_\star = E[log ML(y)] - k/2$$

Ainsi, la quantit\'e:

$$log ML(y) - k/2$$

fournit un estimateur sans biais de S*. C'est presque la correction d'Akaike... \`a un facteur 2 pr\`es. Akaike se serait-il plant\'e?

Il semble que cette incoh\'erence vienne du fait qu'Akaike cherche en fait \`a estimer:

$$S** = int p(y) log p_theta_ML(y) dy$$

et non $S* = max_theta S(p, p_theta)$, ce qui me semble plus naturel mais c'est moi.

Quoiqu'il en soit, tout se passe comme si on d\'efinissait l'\'evidence du mod\`ele comme:

$$q(y) = f(k) ML(y) avec f(k) = exp(-k/2) ou exp(-k)$$

mais c'est justifi\'e sous l'hypoth\`ese ABSURDE DANS LE CONTEXTE DE LA SELECTION DE MODELE que p(y) est dans la famille. Si tous les mod\`eles test\'es sont corrects, il suffit de prendre le plus parcimonieux!

Peut-on g\'en\'eraliser Wilks au cas d'un mod\`ele mal sp\'ecifi\'e? Pour cela, on peut se reporter au papier de Harts, 1982 et, en recollant les morceaux, voir que:

$$E {2 log [ML(y) / p*(y)]} = trace(B A^-1)$$

o\`u $p*$ est la projection KL de p sur la famille, avec:

$$A = E[ d^2 log p_\theta / d\theta^2 ] (Hessien de la log-vraisemblance moyenne) B = E[ (d\log p_\theta / d\theta) (d\log p_\theta / d\theta)' ]$$

Dans le cas o\`u $p=p_\star$ appartient \`a la famille, ces deux matrices (calcul\'ees en $theta_\star$) sont \'egales (\`a l'information de Fisher), d'o\`u:

$$trace(B A^-1) = trace(I_k) = k$$

On retrouve bien le r\'esultat de Wilks. Mais on voit aussi qu'il ne se g\'en\'eralise pas trivialement. Le biais d\'epend de la "vraie" distribution des donn\'ees; et s'il est grand, on pourrait se faire berner par un mod\`ele incorrect. On peut l'estimer empiriquement, ce biais, si \c{c}a rentre dans le budget calculatoire, ou essayer de le majorer...

On peut voir que $B = A + G$ avec:

$$G = E[(d^2 p_theta / dtheta^2) / p_theta]$$

$$==> B A^-1 = I_k + G A^-1$$

$$==> trace(A B^-1) = k + trace(G A^-1)$$

Facile de montrer que $G=0$ si $p=p_\star$ mais sinon... ?????

Cela \'etant, le r\'esultat montre que le biais sur la log-vraisemblance MOYENNE est en O(1/n), o\`u n est le nombre de points. Ainsi, lorsqu'on compare des mod\`eles sur un "grand" \'echantillon, on peut raisonnablement supposer que le biais est n\'egligeable.

En conclusion, l'int\'er\^et majeur de l'approche d'Akaike est, \`a mon sens, de justifier la vraisemblance profile comme mesure d'\'evidence \`a la limite des grands \'echantillons. Sous r\'eserve que n soit constant \`a travers les mod\`eles (ce qui n'est pas le cas dans le recalage).

\section{LE PARADOXE RECALAGE/APPARIEMENTS}

J'ai but\'e pendant des ann\'ees sur une difficult\'e conceptuelle de ma formulation statistique du recalage: par simple raffinement de l'\'echantillonnage, on peut rendre une transformation infiniment plus vraisemblable qu'une autre et donc rendre l'incertitude bay\'esienne sur le recalage aussi petite qu'on veut.

Ce ph\'enom\`ene vient du fait que la log-vraisemblance d'une transformation est proportionnelle au nombre de points appari\'es. Or ce nombre est d\'etermin\'e par l'\'echantillonnage et est donc arbitraire. Certes, on s'attend \`a ce que la vraisemblance d\'epende de la repr\'esentation des donn\'ees. Mais \`a ce point? Cette observation m'a fait longtemps douter de l'hypoth\`ese d'ind\'ependance des pixels.

Elle semble en tous cas contradictoire avec un principe \'el\'ementaire de th\'eorie de l'information. Souviens-toi de l'in\'egalit\'e fondamentale du traitement des donn\'ees. Notre recalage statistique repose sur un pr\'e-traitement des donn\'ees pour le moins radical qui consiste \`a r\'eduire chaque image \`a un histogramme (ou, de fa\c{c}on \'equivalente, une liste des valeurs d'intensit\'e dans un ordre al\'eatoire et ignorant la position des points).

En notant X les param\`etres de transformation, Y=(I, J) le couple d'images et Z=(fI, fJ) sa repr\'esentation du pauvre, on a une belle cha\^ine de Markov:

$X \to Y \to Z$

Le fait que X soit "aussi pr\'ecis que l'on veut" connaissant Z signifie que I(X,Z) est "aussi grand que l'on veut". Or la th\'eorie de l'information affirme que I(X,Z) <= I(X,Y) donc I(X,Y) est infinie. Toute proc\'edure bay\'esienne de recalage aurait donc n\'ecessairement une incertitude nulle (sauf \'eventuellement \`a reposer sur une repr\'esentation encore plus pauvre que la n\^otre). N'est-ce pas la preuve irr\'efutable qu'il y a une faille dans notre raisonnement? Comment l'algorithme pourrait-il \^etre certain de ne pas se tromper alors que chacun sait qu'il y a toujours des erreurs de recalage?

J'avais not\'e en 2007 que le paradoxe "disparait" si on s'int\'eresse \`a des transformations libres. Les param\`etres recherch\'es sont alors des appariements et sont d\'ependants de l'\'echantillonnage. Le fait que la vraisemblance augmente par raffinement de l'\'echantillonnage est alors un simple effet de nombre car les distributions marginales des appariements sont inchang\'ees. On ne peut \^etre certain des appariements \`a quelque r\'esolution que ce soit.

Contrairement \`a l'intuition qui a parfois fauss\'e mon jugement, l'ind\'ependance a posteriori des appariements n'est pas le signe d'un "exc\`es de confiance" de l'algorithme. Bien au contraire! L'ind\'ependance a posteriori, en plus du fait que les lois marginales a posteriori sont ind\'ependantes du contexte (c-\`a-d, n'utilisent pas les pixels voisins du point cible orrespondant) illustre bien l'information limit\'ee fournie par la repr\'esentation. La distribution a posteriori pourrait dfficilement \^etre plus "entropique"!

Ouf... l'in\'egalit\'e fondamentale du traitement des donn\'ees n'est pas bafou\'ee, au contraire elle est confirm\'ee dans toute sa gloire.  Mais, alors, si l'incertitude sur des transformations libres est "grande", comment peut-elle \^etre nulle pour des transformations param\'etriques?

Le probl\`eme n'a rien \`a voir avec notre mod\`ele particulier de repr\'esentation des images. Le simple fait d'imaginer qu'on puisse d\'efinir une vraisemblance sur des appariements denses conduit au "paradoxe".

Si la vraisemblance en question est not\'ee L(Q)=p(Y|Q), que vaut p(Y|X), la vraisemblance sur les param\`etres de transformation? On suppose une chaine de Markov:

$X \to Q \to D$

autrement dit, que les donn\'ees sont ind\'ependantes de la transformation CONNAISSANT les appariements, ce qui semble \'evident. On a donc:

$p(Y|X) = \int p(Y|Q)p(Q|X) dQ$

En supposant (pour fixer les id\'ees) une d\'ependance d\'eterministe X |-> Q, on a donc, tout simplement:

$L(X) = L(Q(X))$

Ce qui g\'en\'eralise notre "curse of dimensionality": plus la grille sous-jacente \`a Q est grande, plus l'ordre de grandeur de la vraisemblance sur X est grande, et donc plus grands sont les rapports de vraisemblances entre transformations test\'ees.

D'une certaine mani\`ere, on dit juste ici que si on a une infinit\'e de points de contr\^ole, on peut estimer exactement (au sens d'une variance a posteriori nulle) les param\`etres d'une transformation param\'etrique. On ne dit pas que les appariements sont exacts. S'ils l'\'etaient, un nombre fini suffirait.

Ceci semble paradoxal mais c'est un effet "m\'ecanique" de la dimension. Consid\'erons le cas similaire de la r\'egression aux moindres carr\'es, pour laquelle il est bien connu que la variance d'estimation est donn\'ee par:

$V = s^2 (X^\top X)^{-1}$

La matrice X'X \'etant proportionnelle \`a la fr\'equence d'\'echantillonnage, V est une fonction d\'ecroissante de celle-ci, sauf \`a consid\'erer que la variance du bruit augmente avec la fr\'equence d'\'echantillonnage, ce qui a du sens physique si le bruit est moyenn\'e dans la p\'eriode d'\'echantillonnage. Ca n'en a pas pour nous car le "bruit" est ici l'incertitude marginale sur chaque appariement et n'a donc aucune raison "physique" de varier en fonction de l'\'echantillonnage.

Force est de conclure que la variance a posteriori sur nos param\`etres de recalage est virtuellement nulle. Est-ce \`a dire que le recalage est exact? Bien s\^ur que non!

D'abord, il y a toujours du biais car le mod\`ele de transformation ne peut \^etre parfait. Un cerveau n'est pas rigide. La technique d'imagerie introduit des distorsions g\'eom\'etriques. Il y aura toujours des d\'eformations r\'esiduelles.

Ensuite, notre mod\`ele suppose connues les position des points dans les images, que ce soit les pixels "natifs" ou les points interpol\'es. Quand j'attribue \`a un pixel la position de son centre, n'est-ce pas arbitraire? La vraie position n'est-elle pas "quelque part dans le pixel"?

Imaginons l'exp\'erience suivante. On prend une fonction constante par morceaux, on l'\'echantillonne (en ajoutant du bruit et en moyennant le signal par pixel), puis on interpole le signal \'echantillonn\'e. Est-ce qu'on r\'ecup\`ere exactement la fonction originale? Est-ce que les points "saillants" (points de contour, coins...) peuvent \^etre localis\'es avec certitude? Evidemment pas.

D'une part, cela signifie qu'il est impossible \`a un op\'erateur humain d'\'etablir une v\'erit\'e terrain exacte \`a partir d'amers rep\'er\'es \`a l'oeil. D'autre part, l'algorithme de recalage op\`ere sur des images qui pr\'esentent la r\'ealit\'e d'une mani\`ere "floue" non prise en compte par notre mod\`ele statistique.

Comment pourrait-il l'\^etre? En introduisant une incertitude sur la localisation des points? Question ouverte.

===========================================================

L'intuition qui est derri\`ere \c{c}a est que le max de vraisemblance donne une bonne estim\'ee de theta mais celle-ci \'etant d\'ependante des donn\'ees, elle ne d\'efinit pas un "code" valide. Le regret exprime la diff\'erence entre un code universel valide et le code "na\"if" fourni par MV. L'autre id\'ee sous-jacente \`a NML est qu'on ne suppose pas que la "vraie" distribution appartient \`a la famille. D\`es lors, il est naturel de consid\'erer le regret maximal comme crit\`ere, d'o\`u un probl\`eme minimax:

$$Lnml(q) = max_p int p(y) log ML(y)/q(y) dy$$

$$==> argmin Lnml(q) = ML(y) / int ML(y) dy$$

Cette approche fournit une version corrig\'ee de la "vraisemblance profile" comme estimation de l'\'evidence du mod\`ele. Le terme correctif, qui exprime la complexit\'e du mod\`ele, ne peut \^etre ignor\'e que s'il s'av\`ere ind\'ependant de x.

Dans le cas de la vraisemblance r\'eduite, ces deux codes sont disponibles en version r\'eduite, c-\`a-d en rempla\c{c}ant $y$ par $z_x$. Ils permettent alors de calculer les rapports d'\'evidence pertinents pour estimer $p(y|x)/\pi(y)$, sans d'ailleurs connaitre $\pi(y)$. C'est toute la substantifique mo\"elle de cette approche qu'est la vraisemblance r\'eduite.

\end{document}

### 2018-10-27 notes/txt2tex.py

@@ -1,0 +1,78 @@
+import sys
+import os
+
+CODE = 'ansi'
+
+def latex_encoding(s):
+    out = s.replace('à', '\\`a')
+    out = out.replace('é', "\\'e")
+    out = out.replace('è', '\\`e')
+    out = out.replace('ê', '\\^e')
+    out = out.replace('ë', '\\"e')
+    out = out.replace('ï', '\\"i')
+    out = out.replace('î', '\\^i')
+    out = out.replace('ô', '\\^o')
+    out = out.replace('û', '\\^u')
+    out = out.replace('ù', '\\`u')
+    out = out.replace('ü', '\\"u')
+    out = out.replace('ç', '\\c{c}')
+    return out
+    
+	
+class LatexGenerator(object):
+
+    def __init__(self, text_file, author='Alexis Roche'):
+	
+        self.name = os.path.splitext(text_file)[0] + '.tex'
+        
+        f = open(text_file)
+        self._text = [l.encode(CODE).decode() for l in f.readlines()]		
+        text = [latex_encoding(l) for l in self._text]
+        f.close()
+        title = text[0].replace('\n','')
+        self.preamble = [
+            '\\documentclass{article}\n', 
+	    '\\usepackage{fullpage}\n', 
+            '\\usepackage{amsmath,amssymb}\n',
+            '\\title{%s}\n' % title,
+            '\\author{%s}\n' % author,
+            '\\begin{document}\n',
+            '\\maketitle\n']
+        self.end = ['\n', '\\end{document}\n']
+        self.text = text[1:]
+        self.find_sections()
+		
+    def find_sections(self):
+        out = []
+        sec, sub = False, False		
+        for l in self.text:
+            if '***' in l:
+                print('section')
+                sec = True 
+            elif '---' in l:
+                print('subsection')
+                sub = True
+            else:
+                ls = l.strip()
+                if ls != '':
+                    if sec:
+                        out.append('\\section{%s}\n' % ls)
+                        sec = False
+                    elif sub:
+                        out.append('\\subsection{%s}\n' % ls)
+                        sub = False
+                    else:
+                        out.append(l.replace('\n', ' '))
+                else:
+                    out.append('\n\n')
+        self.text = out
+
+    def save(self):
+        f = open(self.name, 'w')
+        f.writelines(self.preamble + self.text + self.end)
+        f.close()
+	    
+
+g = LatexGenerator(sys.argv[1])
+g.save()
+

### 2018-10-29 notes/composite_inference.tex
Latex conversion

@@
 
-Il faut d\'ej\`a comprendre en quoi ce paradigme certes intuitif diff\`ere du paradigme classique pour l'inf\'erence multi-classes. Supposons qu'on ait un mod\`ele g\'en\'eratif de la forme:
-
+Expliquons d\'ej\`a en quoi ce paradigme certes intuitif diff\`ere du paradigme classique pour l'inf\'erence multi-classes. Supposons qu'on ait un mod\`ele g\'en\'eratif de la forme: 
 $$
@@
 $$ 
-o\`u $zx = z(x, y)$ est un "attribut contextuel", on prouve facilement que cette forme se ram\`ene \`a celle du PDF projection theorem: 
+o\`u $zx = z(x, y)$ est un attribut ``contextuel'' ou ``adaptatif'' (notion plus g\'en\'erale que celle de statistique), on prouve facilement que cette forme se ram\`ene \`a celle du PDF projection theorem:
 \begin{equation}
@@
 
-Une telle structure peut apparaitre naturellement dans un mod\`ele ou r\'esulter de l'application du principe d'entropie maximale si on s\'electionne le mod\`ele en supposant connues les distributions de l'attribut contextuel, $p(z_x|x)$.
+Une telle structure peut apparaitre naturellement dans un mod\`ele ou r\'esulter de l'application du principe d'entropie relative maximale si on s\'electionne le mod\`ele en supposant connues les distributions de l'attribut adaptatif, $p(z_x|x)$.
 Notons que dans ce dernier cas, la distribution de r\'ef\'erence $\pi(y)$ est arbitraire.
@@
 Cette strat\'egie d'entropie maximale d\'ecoule en fait d'un probl\`eme de th\'eorie des jeux d\'ecrit par Gr\"unwald, 2005: il s'agit de simuler des donn\'ees en l'absence d'un mod\`ele g\'en\'eratif complet.
-Le PDF projection n'est donc qu'une solution optimale pour un probl\`eme de synth\`ese de donn\'ees.
+Le PDF projection est donc une solution optimale \`a un probl\`eme de synth\`ese de donn\'ees.
 
@@
 Autrement dit, on peut se passer du principe de maximum d'entropie si on suppose l'exhaustivit\'e de la repr\'esentation.
-C'est une justification alternative mais plus fondamentale dans le contexte de l'inf\'erence.
+C'est une justification alternative mais plus forte dans le contexte de l'inf\'erence.
 
@@
 Supposons qu'on ait la forme (\ref{eq:pdf_proj}).
-Si on traite p(y|x) de fa\c{c}on classique, sans tenir compte de cette structure, on voit que l'estimation des param\`etres du mod\`ele par maximum de vraisemblance (par exemple) conduit \`a:
-
+Si on traite $p(y|x)$ de fa\c{c}on classique, sans tenir compte de cette structure, on voit que l'estimation des param\`etres du mod\`ele par maximum de vraisemblance (par exemple) conduit \`a:
 $$
@@
 
-En supposant que les param\`etres alpha de pi et beta des autres lois g\'en\'eratives sont disjoints et que pi est la loi de la classe 0, cela revient \`a maximiser chaque terme ind\'ependamment, donc notamment:
-
+En supposant que les param\`etres $\alpha$ de $\pi$ et $\beta$ des autres lois g\'en\'eratives sont disjoints et que pi est la loi de la classe~0, cela revient \`a maximiser chaque terme ind\'ependamment, donc notamment:
 $$
@@
 
-Supposons par exemple qu'on a 3 classes et que $y = (z_1, z_2)$ sont ind\'ependants dans la classe 0, {\em i.e.} $\pi(z_1, z_2) = \pi(z_1)p(z_2)$.
+Supposons par exemple qu'on ait 3 classes et que $y = (z_1, z_2)$ sont ind\'ependants dans la classe~0, {\em i.e.} $\pi(z_1, z_2) = \pi(z_1)p(z_2)$.
 On a alors: $f_k(\alpha) = \log \pi_\alpha(z_{1k})\pi_\alpha(z_{2k})$ si $x_k=0$, $f_k(\alpha) = \log \pi_\alpha(z_{2k})$ si $x_k=1$,  et $f_k(\alpha) = \log \pi_\alpha(z_{1k})$ si $x_k=2$.
 
-Si on suppose en outre que pi(z1) et pi(z2) correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution "nulle" de z1 (l'attribut "exhaustif" pour la classe 1) sur les donn\'ees issues des classes 0 et 2...
+Si on suppose en outre que $\pi(z1)$ et $\pi(z2)$ correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution ``nulle'' de $z_1$ (l'attribut ``exhaustif'' pour la classe~1) sur les donn\'ees issues des classes~0 et~2...
 
-Que vient faire la classe 2 dans cette histoire??? N'est-il pas absurde d'utiliser des donn\'ees de la classe 2 pour ajuster la loi de la classe 0?
+Que vient faire la classe~2 dans cette histoire??? N'est-il pas absurde d'utiliser des donn\'ees de la classe~2 pour ajuster la loi de la classe~0?
 
-Non car on peut se rendre compte que le mod\`ele implique que z1 est distribu\'e de la m\^eme fa\c{c}on dans les classes 0 et 2:
-
-$$p2(z1, z2) = pi(z1) p2(z2) ==> p2(z1) = pi(z1)$$
+Non car on peut se rendre compte que le mod\`ele implique que $z_1$ est distribu\'e de la m\^eme fa\c{c}on dans les classes~0 et~2: 
+$$
+p_2(z_1, z_2) = \pi(z_1) p_2(z_2) 
+\Rightarrow 
+p_2(z_1) = \pi(z_1)
+$$
 
@@
 
-Est-ce \`a dire que le maximum de vraisemblance "compl\`ete" est toujours la meilleure approche?
+Est-ce \`a dire que le maximum de vraisemblance compl\`ete est toujours la meilleure approche?
 On a d\'ej\`a compris que son application emp\^eche les experts d'\^etre ind\'ependants.
 En effet, l'ajustement de la loi nulle doit \^etre centralis\'e car il fait intervenir des informations que les experts n'ont pas.
-Dans l'exemple ci-dessus, l'expert pour la classe 1 ne peut ajuster pi(z1) lui-m\^eme car il a besoin pour \c{c}a des valeurs de z2.
+Dans l'exemple ci-dessus, l'expert pour la classe~1 ne peut ajuster $\pi(z_1)$ lui-m\^eme car il a besoin pour \c{c}a des valeurs de $z_2$.
 
-Pour avoir un mod\`ele complet, on doit avoir une repr\'esentation "originelle" des donn\'ees, c-\`a-d qu'on doit d\'efinir la variable y alors que celle-ci n'a n'a pas lieu d'\^etre dans l'approche composite.
-L'ajustement des lois $\pi(z_x)$ via la vraisemblance compl\`ete est-il ind\'ependant de la repr\'esentation "originelle"?
+Pour avoir un mod\`ele complet, on doit avoir une repr\'esentation {\em originelle} des donn\'ees, c-\`a-d qu'on doit d\'efinir la variable $y$ alors que celle-ci n'a n'a pas lieu d'\^etre dans l'approche composite.
+L'ajustement des lois $\pi(z_x)$ via la vraisemblance compl\`ete est-il ind\'ependant de la repr\'esentation originelle?
 La r\'eponse est non.
-Si on inclut dans y un attribut z3 qui a, par exemple, la m\^eme variance que z1 sous l'hypoth\`ese de r\'ef\'erence alors, clairement, l'estimation par maximum de vraisemblance "compl\`ete" de pi(z1) va d\'ependre de z3...
-c-\`a-d qu'on devra mesurer z3 pour pouvoir ajuster cette loi.
+Si on inclut dans y un attribut z3 qui a, par exemple, la m\^eme variance que $z_1$ sous l'hypoth\`ese de r\'ef\'erence alors, clairement, l'estimation par maximum de vraisemblance compl\`ete de $\pi(z_1)$ va d\'ependre de $z_3$...
+c-\`a-d qu'on devra mesurer $z_3$ pour pouvoir ajuster cette loi.
 
@@
 
-La vraisemblance r\'eduite garantit l'exhaustivit\'e EN UN SENS FORT: en clair, elle laisse chaque expert \'eliminer les param\`etres par ses propres moyens.
+La vraisemblance r\'eduite garantit l'exhaustivit\'e {\em en un sens fort}: en clair, elle laisse chaque expert \'eliminer les param\`etres par ses propres moyens.
 Contrairement \`a la vraisemblance compl\`ete, elle ne fournit pas un estimateur unique des param\`etres de la distribution de r\'ef\'erence (chaque expert y va de son estimateur...).
@@
 Mais quelle importance si le but assum\'e est d'estimer les rapports de vraisemblance: 
-
-$$R(x) = p(y|x) / pi(y) = pi(z_x|x) / pi(z_x)$$
+$$
+R(x) 
+= \frac{p(y|x)}{\pi(y)}
+= \frac{p(z_x|x)}{\pi(z_x)}
+$$
 
 La vraisemblance r\'eduite permet de le faire sans exploiter toute l'information contenue dans les donn\'ees mais avec plusieurs avantages: 
-
 \begin{itemize}
@@
 \item coh\'erence asymptotique (et oui, c'est maintenu car la vraisemblance   restreinte reste la vraisemblance...) 
-\item ind\'ependance vis-\`a-vis de toute repr\'esentation "universelle" 
+\item ind\'ependance vis-\`a-vis de toute repr\'esentation ``universelle'' 
 \item flexibilit\'e: ind\'epedendance des experts et donc possibililt\'e de   combiner leurs r\'esultats sans connaitre les donn\'ees utilis\'ees (et   donc notamment s'ils ont utilis\'e des donn\'ees diff\'erentes)
@@
 
-Je suis d\'esormais convaincu de la validit\'e du mod\`ele g\'en\'eratif "i.i.d" pr\'esent\'e dans ma th\`ese pour le recalage: il est justifi\'e par le th\'eor\`eme d'\'echangeabilit\'e de de Finetti dans la mesure o\`u on suppose que les pixels/voxels sont m\'elang\'es au pr\'ealable, ce qui revient \`a dire que la repr\'esentation "universelle" est un ensemble de pixels (donc sans notion d'ordre).
+Je suis d\'esormais convaincu de la validit\'e du mod\`ele g\'en\'eratif i.i.d.
+pr\'esent\'e dans ma th\`ese pour le recalage: il est justifi\'e par le th\'eor\`eme d'\'echangeabilit\'e de de Finetti dans la mesure o\`u on suppose que les pixels/voxels sont m\'elang\'es au pr\'ealable, ce qui revient \`a dire que la repr\'esentation ``universelle'' est un ensemble de pixels (donc sans notion d'ordre).
 
@@
 Mais des propri\'et\'es bizarres apparaissent:
-
 \begin{itemize}
@@
 
-On peut voir l'\'elimination de param\`etres comme un probl\`eme d'approximation ou, si on pr\'ef\`ere, de DECISION, c-\`a-d un probl\`eme de choix d'une valeur unique dans un ensemble.
-Il s'agit ici de repr\'esenter un mod\`ele param\'etrique par une distribution unique afin d'\'evaluer "l'\'evidence" du mod\`ele, c-\`a-d la probabilit\'e qu'il conf\`ere aux donn\'ees.
+On peut voir l'\'elimination de param\`etres comme un probl\`eme d'approximation ou, si on pr\'ef\`ere, de {\em d\'ecision}, c-\`a-d un probl\`eme de choix d'une valeur unique dans un ensemble.
+Il s'agit ici de repr\'esenter un mod\`ele param\'etrique par une distribution unique afin d'\'evaluer {\em l'\'evidence du mod\`ele}, c-\`a-d la probabilit\'e qu'il conf\`ere aux donn\'ees.
 
@@
 
-Ici, l'objectif est en soi d'estimer l'\'evidence de mani\`ere \`a obtenir une FONCTION DE VRAISEMBLANCE de la variable d'int\'er\^et x.
+Ici, l'objectif est en soi d'estimer l'\'evidence de mani\`ere \`a obtenir une {\em fonction} DE VRAISEMBLANCE de la variable d'int\'er\^et x.
 On ne se contente pas d'un estimateur ponctuel de x.
@@
 
-Une id\'ee naturelle est de choisir une sorte de centre de la famille, c'est-\`a-dire un \'el\'ement situ\'e \`a faible "distance" de tout autre \'el\'ement.
-La "distance" en question d\'efinit la fonction de perte et peut \^etre prise, par exemple, au hasard, comme la divergence KL:
-
-$$L(theta, q) = D(p_theta\|q)$$
+Une id\'ee naturelle est de choisir une sorte de centre de la famille, c'est-\`a-dire un \'el\'ement situ\'e \`a faible ``distance'' de tout autre \'el\'ement.
+La ``distance'' en question d\'efinit la fonction de perte et peut \^etre prise, par exemple, au hasard, comme la divergence KL:
+$$L(\theta, q) = D(p_\theta\|q)$$
 
@@
 Ceci donne:
+$$R(q) = \int \pi(\theta) p_\theta(y) \log [p_\theta(y) / q(y)] dy d\theta$$
 
-$$R(q) = int pi(theta) p_theta(y) log [p_theta(y) / q(y)] dy dtheta$$
-
-$$==> argmin R(q) = int pi(theta) p_theta(y) dtheta$$
+$$\Rightarrow \arg\min R(q) = \int \pi(\theta) p_\theta(y) d\theta$$
 
@@
 
-On peut cependant obtenir d'autres strat\'egies tout aussi bay\'esiennes en choisissant d'autres divergences; par exemple, la divergence KL invers\'ee:
-
-$$L(theta, q) = D(q||p_theta)$$
-
+On peut cependant obtenir d'autres strat\'egies tout aussi bay\'esiennes en choisissant d'autres divergences; par exemple, la divergence KL invers\'ee: $L(\theta, q) = D(q\|p_\theta)$,
+\begin{eqnarray*}
+R(q) 
+& = & \int pi(\theta)q(y) \log \frac{q(y)}{p_\theta(y)} dy d\theta \\
+& = & \int q(y) \log q(y) dy  - \int \pi(\theta) q(y) \log p_\theta(y) dy d\theta
+\end{eqnarray*}
+ce qui donne:
 $$ 
-R(q) = \int pi(\theta)q(y) log[q(y) / p_\theta(y)] dy d\theta = int q(y) log q(y) dy - \int \pi(\theta) q(y) log p_\theta(y) dy d\theta
+\frac{\partial R}{\partial q}
+= \log q(y) + 1 - \int \pi(\theta) \log p_\theta(y) dy 
+\quad \Rightarrow 
+\arg\min R(q) = K \exp[\int \pi(\theta) \log p_\theta(y) dy ]
 $$
 
-$$ dR / dq = \log q(y) + 1 - \int \pi(\theta) \log p_\theta(y) dy ==> \arg\min R(q) = K \exp[\int \pi(\theta) \log p_\theta(y) dy ]$$
-
 La solution est cette fois la moyenne g\'eom\'etrique.
-Tout \c{c}a pour dire que la marginalisation n'est PAS NECESSAIREMENT la m\'ethode bay\'esienne de choix pour l'\'elimination des param\`etres de nuisance.
+Tout \c{c}a pour dire que la marginalisation n'est {\em pas n\'ecessairement} la m\'ethode bay\'esienne de choix pour l'\'elimination des param\`etres de nuisance.
 Tout d\'epend du probl\`eme de d\'ecision et de la fonction de perte associ\'ee.
@@
 Sa justification classique (cf.
-Gr\"unwald ou Rissanen) repose sur la notion de "regret", c-\`a-d la diff\'erence de longueur de code entre la distribution caract\'eristique candidate et la distribution qui compresse le plus l'observation:
+Gr\"unwald ou Rissanen) repose sur la notion de {\em regret}, c-\`a-d la diff\'erence de longueur de code entre la distribution caract\'eristique candidate et la distribution qui compresse le plus l'observation:
+$$
+L(y, q) = - \log q(y) - \min_\theta [-\log p_\theta(y)] = -\log q(y) + \log \ell_\star(y)
+$$ 
+avec: $\ell_\star(y) = \max_\theta p_\theta(y)$, la {\em vraisemblance profile} vue comme une fonction de~$y$.
 
-$$L(y, q) = - log q(y) - min_theta [-log p_theta(y)] = -log q(y) + log ML(y)$$
-
-avec: $ML(y) = max_theta p_theta(y)$
-
-Le maximum de vraisemblance apparait naturellement dans cette formulation en tant que "meilleur mod\`ele" mais c'est un mod\`ele "na\"if" car ML(y) ne d\'efinit pas une distribution pr\'edictive valide \`a cause du couplage entre l'observation et l'estim\'ee MV.
+Le maximum de vraisemblance apparait naturellement dans cette formulation en tant que ``meilleur mod\`ele'' mais c'est un mod\`ele na\"if car $\ell_\star(y)$ ne d\'efinit pas une distribution pr\'edictive valide \`a cause du couplage entre l'observation et l'estim\'ee MV.
 
 Une diff\'erence essentielle avec l'approche pr\'ec\'edente est que la fonction de perte d\'epend de l'observation.
-Tout se passe en fait comme si l'observation rempla\c{c}ait theta comme "\'etat inconnu du monde".
+Tout se passe en fait comme si l'observation rempla\c{c}ait theta comme ``\'etat inconnu du monde''.
 Cela a du sens dans la mesure o\`u la distribution caract\'eristique doit \^etre choisie ind\'ependamment de toute observation.
-En fait, on peut consid\'erer que l'\'etat du monde est donn\'e par le couple $(y, theta)$, ce qui permet de d\'efinir le regret pour tout theta:
+En fait, on peut consid\'erer que l'\'etat du monde est donn\'e par le couple $(y, \theta)$, ce qui permet de d\'efinir le regret pour tout $\theta$:
+$$
+L(y, \theta; q) = \log \frac{p_\theta(y)}{q(y)}
+$$
 
-$$L(y, theta; q) = log p_theta(y) / q(y)$$
-
-D\`es lors, il est naturel de chercher \`a minimiser le "pire" regret:
-
-$$R(q) = \max_{y, \theta} L(y, \theta; q)$$
-
+D\`es lors, il est naturel de chercher \`a minimiser le {\em pire regret}: 
+$$
+R(q) = \max_{y, \theta} L(y, \theta; q)
+$$ 
 en notant au passage qu'on ne fait pas l'hypoth\`ese que $y$ est distribu\'e selon un certain $p_\theta(y)$.
@@
 Comme c'est classique dans ce type de probl\`emes minimax, on probabilise $y$ en introduisant le risque moyen:
-
-$$L(p, theta; q) = E_p [L(y, theta; q)] = int p(y) log p_theta(y) / log q(y) dy = D(p||q) - D(p||p_theta) $$
+$$
+L(p, \theta; q) 
+= E_p [L(y, \theta; q)] 
+= \int p(y) \log \frac{p_\theta(y)}{q(y)} dy 
+= D(p\|q) - D(p\|p_theta) 
+$$
 
 On a \'evidemment: 
-
-$$R(q) = max_p max_theta L(p, theta, q) = max_p L(p, q)$$
-
+$$R(q) = \max_p \max_\theta L(p, \theta, q) = \max_p L(p, q)$$ 
 o\`u: 
-
-$$L(p, q) = int p(y) log ML(y) / q(y) dy$$
+$$L(p, q) = \int p(y) \log \frac{\ell_\star(y)}{q(y)} dy$$
 
 Il suffit alors de montrer que les op\'erateurs min et max commutent:  
+$$\min_q R(q) = \min_q \max_p L(p, q) = \max_p \min_q L(p, q)$$
 
-$$min_q R(q) = min_q max_p L(p, q) = max_p min_q L(p, q)$$
-
-On voit alors que quelque soit p, la distribution de Shtarkov (1987) dite NML minimise TOUJOURS $L(p, q)$:
-
-$$\arg\min_q L(p, q) = ML(y) / Z_{nml}$$
-
+On voit alors que quelque soit $p$, la distribution de Shtarkov (1987) dite NML minimise {\em toujours} $L(p, q)$: 
+$$\arg\min_q L(p, q) = \frac{\ell_\star(y)}{Z_{nml}}$$ 
 avec en plus la propri\'et\'e que le minimum atteint est ind\'ependant de p: 
-
-$$min_q L(p, q) = log Znml$$
+$$\min_q L(p, q) = \log Z_{nml}$$
 
@@
 
-Au-del\`a de la nature "Bayes robuste" de cette formulation, on voit qu'elle consiste plus fondamentalement \`a remplacer $D(p_theta\|q)$ dans la 1\`ere formulation par $D(p\|q) - D(p\|p_theta)$ avec l'id\'ee que $(p,\theta)$ repr\'esente l'\'etat du monde.
+Au-del\`a de la nature {\em Bayes robuste} de cette formulation, on voit qu'elle consiste plus fondamentalement \`a remplacer $D(p_\theta\|q)$ dans la 1\`ere formulation par $D(p\|q) - D(p\|p_\theta)$ avec l'id\'ee que $(p,\theta)$ repr\'esente l'\'etat du monde.
 
@@
 L'id\'ee de base est de d\'efinir l'ad\'equation d'un mod\`ele via la divergence KL ou, de fa\c{c}on \'equivalente, la log-vraisemblance moyenne: 
+$$S(p, p_\theta) = \int p(y) \log p_\theta(y) dy$$ 
+plut\^ot que via la notion empirique d'\'evidence.
+Dans cette perspective, la distribution r\'ealisant le maximum de S sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure {\em optimiste} qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:
+$$S_\star = \max_\theta S(p, p_\theta)$$
 
-$$S(p, p_theta) = int p(y) log p_theta(y) dy$$
-
-plut\^ot que via la notion empirique d'\'evidence.
-Dans cette perspective, la distribution r\'ealisant le maximum de S sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure "optimiste" qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:
-
-$$S* = max_theta S(p, p_theta)$$
-
-Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la "vraie" distribution des donn\'ees, et que S* est donc inconnu. Mais Akaike justifie l'approximation:
-
-$$S* ~ log ML(y) - k$$
-
-o\`u k est la dimension de theta. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement...
+Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la ``vraie'' distribution des donn\'ees, et que S* est donc inconnu. Mais Akaike justifie l'approximation: 
+$$S_\star \approx log \ell_\star(y) - k$$ 
+o\`u k est la dimension de $\theta$. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement...
 
 J'ai l'impression qu'il suffit d'appliquer le fameux th\'eor\`eme de Wilks: 
-
-$$2 log [ML(y) / p(y)] ~ \chi_2(k)$$
-
+$$2 \log \frac{\ell_\star(y)}{p(y)} \sim \chi_2(k)$$ 
 valable asymptotiquement SI $p(y)$ APPARTIENT A LA FAMILLE et sous les conditions usuelles d'\'echantillonnage simple.
@@
 En prenant l'esp\'erance de l'expression ci-dessus, les choses deviennent tr\`es claires:
-
-$$S_\star = E[log ML(y)] - k/2$$
+$$S_\star = E[log \ell_\star(y)] - \frac{k}{2}$$
 
 Ainsi, la quantit\'e: 
-
-$$log ML(y) - k/2$$
-
-fournit un estimateur sans biais de S*. C'est presque la correction d'Akaike... \`a un facteur 2 pr\`es. Akaike se serait-il plant\'e?
+$$log \ell_\star(y) - k/2$$ 
+fournit un estimateur sans biais de $S_\star$. C'est presque la correction d'Akaike... \`a un facteur 2 pr\`es. Akaike se serait-il plant\'e?
 
 Il semble que cette incoh\'erence vienne du fait qu'Akaike cherche en fait \`a estimer: 
+$$S_{\star\star} = \int p(y) \log ML(y) dy$$ 
+et non $S_\star = \max_\theta S(p, p_\theta)$, ce qui me semble plus naturel mais c'est moi.
 
-$$S** = int p(y) log p_theta_ML(y) dy$$
-
-et non $S* = max_theta S(p, p_theta)$, ce qui me semble plus naturel mais c'est moi.
-
-Quoiqu'il en soit, tout se passe comme si on d\'efinissait l'\'evidence du mod\`ele comme:
-
-$$q(y) = f(k) ML(y) avec f(k) = exp(-k/2) ou exp(-k)$$
-
-mais c'est justifi\'e sous l'hypoth\`ese ABSURDE DANS LE CONTEXTE DE LA SELECTION DE MODELE que p(y) est dans la famille.
+Quoiqu'il en soit, tout se passe comme si on d\'efinissait l'\'evidence du mod\`ele comme: $q(y) = f(k) \ell_\star(y)$ avec $f(k) = \exp(-k/2)$ ou $\exp(-k)$ mais c'est justifi\'e sous l'hypoth\`ese {\em absurde dans le contexte de la s\'election de mod\`ele} que $p(y)$ est dans la famille.
 Si tous les mod\`eles test\'es sont corrects, il suffit de prendre le plus parcimonieux!
@@
 Peut-on g\'en\'eraliser Wilks au cas d'un mod\`ele mal sp\'ecifi\'e? Pour cela, on peut se reporter au papier de Harts, 1982 et, en recollant les morceaux, voir que: 
+$$
+E \big[2 \log \frac{\ell_\star(y)}{p_\star(y)} \big] = {\rm trace}(B A^{-1})
+$$ 
+o\`u $p_\star$ est la projection KL de p sur la famille, avec:
+$$
+A = -E\left[ \frac{\partial^2 \log p_\theta}{\partial\theta^2} \right],
+\quad 
+B = E\left[ \frac{\partial \log p_\theta}{\partial\theta} \frac{\partial \log p_\theta}{\partial\theta}^\top \right]
+$$
 
-$$E {2 log [ML(y) / p*(y)]} = trace(B A^-1)$$
-
-o\`u $p*$ est la projection KL de p sur la famille, avec:
-
-$$A = E[ d^2 log p_\theta / d\theta^2 ] (Hessien de la log-vraisemblance moyenne) B = E[ (d\log p_\theta / d\theta) (d\log p_\theta / d\theta)' ]$$
-
-Dans le cas o\`u $p=p_\star$ appartient \`a la famille, ces deux matrices (calcul\'ees en $theta_\star$) sont \'egales (\`a l'information de Fisher), d'o\`u:
-
-$$trace(B A^-1) = trace(I_k) = k$$
+Dans le cas o\`u $p=p_\star$ appartient \`a la famille, ces deux matrices (calcul\'ees en $\theta_\star$) sont \'egales (\`a l'information de Fisher), d'o\`u: 
+$$
+{\rm trace}(B A^{-1}) = {\rm trace}({\bf I}_k) = k
+$$
 
@@
 Mais on voit aussi qu'il ne se g\'en\'eralise pas trivialement.
-Le biais d\'epend de la "vraie" distribution des donn\'ees; et s'il est grand, on pourrait se faire berner par un mod\`ele incorrect.
+Le biais d\'epend de la ``vraie'' distribution des donn\'ees; et s'il est grand, on pourrait se faire berner par un mod\`ele incorrect.
 On peut l'estimer empiriquement, ce biais, si \c{c}a rentre dans le budget calculatoire, ou essayer de le majorer...
@@
 On peut voir que $B = A + G$ avec:
+$$
+G = E[\frac{\partial^2 p_\theta}{\partial\theta^2}\big / p_\theta]
+$$
 
-$$G = E[(d^2 p_theta / dtheta^2) / p_theta]$$
-
-$$==> B A^-1 = I_k + G A^-1$$
-
-$$==> trace(A B^-1) = k + trace(G A^-1)$$
+Donc:
+$$B A^{-1} = {\bf I}_k + G A^{-1}
+\Rightarrow
+{\rm trace}(A B^{-1}) = k + {\rm trace}(G A^{-1})
+$$
 
@@
 
-Cela \'etant, le r\'esultat montre que le biais sur la log-vraisemblance MOYENNE est en O(1/n), o\`u n est le nombre de points.
-Ainsi, lorsqu'on compare des mod\`eles sur un "grand" \'echantillon, on peut raisonnablement supposer que le biais est n\'egligeable.
+Cela \'etant, le r\'esultat montre que le biais sur la log-vraisemblance {\em moyenne} est en $O(1/n)$, o\`u $n$ est le nombre de points.
+Ainsi, lorsqu'on compare des mod\`eles sur un ``grand'' \'echantillon, on peut raisonnablement supposer que le biais est n\'egligeable.
 
@@
 
-\section{LE PARADOXE RECALAGE/APPARIEMENTS}
+\section{Le paradoxe recalage/appariements}
 
@@
 
-En notant X les param\`etres de transformation, Y=(I, J) le couple d'images et Z=(fI, fJ) sa repr\'esentation du pauvre, on a une belle cha\^ine de Markov:
+En notant $X$ les param\`etres de transformation, $Y=(I, J)$ le couple d'images et $Z=(f_I, f_J)$ sa repr\'esentation du pauvre, on a une belle cha\^ine de Markov: $X \to Y \to Z$.
 
-$X \to Y \to Z$
-
-Le fait que X soit "aussi pr\'ecis que l'on veut" connaissant Z signifie que I(X,Z) est "aussi grand que l'on veut".
-Or la th\'eorie de l'information affirme que I(X,Z) <= I(X,Y) donc I(X,Y) est infinie.
+Le fait que $X$ soit {\em aussi pr\'ecis que l'on veut} connaissant $Z$ signifie que $I(X,Z)$ est {\em aussi grand que l'on veut}.
+Or la th\'eorie de l'information affirme que $I(X,Z) \leq I(X,Y)$ donc $I(X,Y)$ est infinie.
 Toute proc\'edure bay\'esienne de recalage aurait donc n\'ecessairement une incertitude nulle (sauf \'eventuellement \`a reposer sur une repr\'esentation encore plus pauvre que la n\^otre).
@@
 
-J'avais not\'e en 2007 que le paradoxe "disparait" si on s'int\'eresse \`a des transformations libres.
+J'avais not\'e en 2007 que le paradoxe ``disparait'' si on s'int\'eresse \`a des transformations libres.
 Les param\`etres recherch\'es sont alors des appariements et sont d\'ependants de l'\'echantillonnage.
@@
 
-Contrairement \`a l'intuition qui a parfois fauss\'e mon jugement, l'ind\'ependance a posteriori des appariements n'est pas le signe d'un "exc\`es de confiance" de l'algorithme.
+Contrairement \`a l'intuition qui a parfois fauss\'e mon jugement, l'ind\'ependance a posteriori des appariements n'est pas le signe d'un ``exc\`es de confiance'' de l'algorithme.
 Bien au contraire!
 L'ind\'ependance a posteriori, en plus du fait que les lois marginales a posteriori sont ind\'ependantes du contexte (c-\`a-d, n'utilisent pas les pixels voisins du point cible orrespondant) illustre bien l'information limit\'ee fournie par la repr\'esentation.
-La distribution a posteriori pourrait dfficilement \^etre plus "entropique"!
+La distribution a posteriori pourrait dfficilement \^etre plus {\em entropique}!
 
@@
 l'in\'egalit\'e fondamentale du traitement des donn\'ees n'est pas bafou\'ee, au contraire elle est confirm\'ee dans toute sa gloire.
-Mais, alors, si l'incertitude sur des transformations libres est "grande", comment peut-elle \^etre nulle pour des transformations param\'etriques?
+Mais, alors, si l'incertitude sur des transformations libres est ``grande'', comment peut-elle \^etre nulle pour des transformations param\'etriques?
 
 Le probl\`eme n'a rien \`a voir avec notre mod\`ele particulier de repr\'esentation des images.
-Le simple fait d'imaginer qu'on puisse d\'efinir une vraisemblance sur des appariements denses conduit au "paradoxe".
+Le simple fait d'imaginer qu'on puisse d\'efinir une vraisemblance sur des appariements denses conduit au ``paradoxe''.
 
-Si la vraisemblance en question est not\'ee L(Q)=p(Y|Q), que vaut p(Y|X), la vraisemblance sur les param\`etres de transformation? On suppose une chaine de Markov:
+Si la vraisemblance en question est not\'ee L(Q)=p(Y|Q), que vaut p(Y|X), la vraisemblance sur les param\`etres de transformation?
+On suppose une chaine de Markov: $X \to Q \to D$, autrement dit, que les donn\'ees sont ind\'ependantes de la transformation CONNAISSANT les appariements, ce qui semble \'evident.
+On a donc:
+$$
+p(Y|X) = \int p(Y|Q)p(Q|X) dQ
+$$
 
-$X \to Q \to D$
-
-autrement dit, que les donn\'ees sont ind\'ependantes de la transformation CONNAISSANT les appariements, ce qui semble \'evident. On a donc:
-
-$p(Y|X) = \int p(Y|Q)p(Q|X) dQ$
-
-En supposant (pour fixer les id\'ees) une d\'ependance d\'eterministe X |-> Q, on a donc, tout simplement:
-
-$L(X) = L(Q(X))$
-
-Ce qui g\'en\'eralise notre "curse of dimensionality": plus la grille sous-jacente \`a Q est grande, plus l'ordre de grandeur de la vraisemblance sur X est grande, et donc plus grands sont les rapports de vraisemblances entre transformations test\'ees.
+En supposant (pour fixer les id\'ees) une d\'ependance d\'eterministe $X \mapsto Q$, on a donc, tout simplement: 
+$$
+L(X) = L(Q(X)),
+$$
+ce qui g\'en\'eralise notre {\em curse of dimensionality}: plus la grille sous-jacente \`a $Q$ est grande, plus l'ordre de grandeur de la vraisemblance sur $X$ est grande, et donc plus grands sont les rapports de vraisemblances entre transformations test\'ees.
 
@@
 
-Ceci semble paradoxal mais c'est un effet "m\'ecanique" de la dimension.
+Ceci semble paradoxal mais c'est un effet ``m\'ecanique'' de la dimension.
 Consid\'erons le cas similaire de la r\'egression aux moindres carr\'es, pour laquelle il est bien connu que la variance d'estimation est donn\'ee par:
+$$V = \sigma^2 (X^\top X)^{-1}$$
 
-$V = s^2 (X^\top X)^{-1}$
-
-La matrice X'X \'etant proportionnelle \`a la fr\'equence d'\'echantillonnage, V est une fonction d\'ecroissante de celle-ci, sauf \`a consid\'erer que la variance du bruit augmente avec la fr\'equence d'\'echantillonnage, ce qui a du sens physique si le bruit est moyenn\'e dans la p\'eriode d'\'echantillonnage.
-Ca n'en a pas pour nous car le "bruit" est ici l'incertitude marginale sur chaque appariement et n'a donc aucune raison "physique" de varier en fonction de l'\'echantillonnage.
+La matrice $X^\top X$ \'etant proportionnelle \`a la fr\'equence d'\'echantillonnage, V est une fonction d\'ecroissante de celle-ci, sauf \`a consid\'erer que la variance du bruit augmente avec la fr\'equence d'\'echantillonnage, ce qui a du sens physique si le bruit est moyenn\'e dans la p\'eriode d'\'echantillonnage.
+Ca n'en a pas pour nous car le ``bruit'' est ici l'incertitude marginale sur chaque appariement et n'a donc aucune raison ``physique'' de varier en fonction de l'\'echantillonnage.
 
@@
 
-Ensuite, notre mod\`ele suppose connues les position des points dans les images, que ce soit les pixels "natifs" ou les points interpol\'es.
+Ensuite, notre mod\`ele suppose connues les position des points dans les images, que ce soit les pixels ``natifs'' ou les points interpol\'es.
 Quand j'attribue \`a un pixel la position de son centre, n'est-ce pas arbitraire?
-La vraie position n'est-elle pas "quelque part dans le pixel"?
+La vraie position n'est-elle pas ``quelque part dans le pixel''?
 
@@
 Est-ce qu'on r\'ecup\`ere exactement la fonction originale?
-Est-ce que les points "saillants" (points de contour, coins...) peuvent \^etre localis\'es avec certitude?
-Evidemment pas.
+Est-ce que les {\em points saillants} (points de contour, coins...) peuvent \^etre localis\'es avec certitude?
+Evidemment pas
 
 D'une part, cela signifie qu'il est impossible \`a un op\'erateur humain d'\'etablir une v\'erit\'e terrain exacte \`a partir d'amers rep\'er\'es \`a l'oeil.
-D'autre part, l'algorithme de recalage op\`ere sur des images qui pr\'esentent la r\'ealit\'e d'une mani\`ere "floue" non prise en compte par notre mod\`ele statistique.
+D'autre part, l'algorithme de recalage op\`ere sur des images qui pr\'esentent la r\'ealit\'e d'une mani\`ere ``floue'' non prise en compte par notre mod\`ele statistique.
 
-Comment pourrait-il l'\^etre? En introduisant une incertitude sur la localisation des points? Question ouverte.
+Comment pourrait-il l'\^etre?
+La mani\`ere la plus intuitive, \`a mon sens, est de faire d\'ependre les appariements de la transformation de fa\c con non-d\'eterministe en introduisant une incertitude de localisation r\'esiduelle.
+C'est d\'ej\`a ce que Maes faisait implicitement en 1997 avec l'id\'ee de l'interpolation PV.
+Cela revient ici \`a modifier $P(Q|X)$ de fa\c{c}on simple.
+Pour calculer la vraisemblance $L(X)$, on peut utiliser un algorithme~EM dans lequel $Q$ est consid\'er\'e comme donn\'ee manquante.
+On voit alors que le calcul de la vraisemblance fait apparaitre des poids analogues \`a l'interpolation PV mais {\em modul\'es it\'erativement par la distribution conjointe d'intensitit\'e}.
 

### 2018-10-29 notes/txt2tex.py

@@ -21,11 +21,9 @@ def latex_encoding(s):
 	
 class LatexGenerator(object):
 
-    def __init__(self, text_file, author='Alexis Roche'):
-	
-        self.name = os.path.splitext(text_file)[0] + '.tex'
-        
-        f = open(text_file)
+    def __init__(self, input_file, author='Alexis Roche'):
+       
+        f = open(input_file)
         self._text = [l.encode(CODE).decode() for l in f.readlines()]		
         text = [latex_encoding(l) for l in self._text]
         f.close()
@@ -74,5 +72,5 @@ class LatexGenerator(object):
 	    
 
 g = LatexGenerator(sys.argv[1])
-g.save()
+g.save(sys.argv[2])
 

### 2018-11-09 notes/composite_inference.tex
Add revised composite inference paper project

@@
 $$ 
-o\`u $zx = z(x, y)$ est un attribut ``contextuel'' ou ``adaptatif'' (notion plus g\'en\'erale que celle de statistique), on prouve facilement que cette forme se ram\`ene \`a celle du PDF projection theorem:
+o\`u $zx = z(x, y)$ est un attribut ``contextuel" ou ``adaptatif'' (notion plus g\'en\'erale que celle de statistique), on prouve facilement que cette forme se ram\`ene \`a celle du PDF projection theorem:
 \begin{equation}
@@
 
-Si on suppose en outre que $\pi(z1)$ et $\pi(z2)$ correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution ``nulle'' de $z_1$ (l'attribut ``exhaustif'' pour la classe~1) sur les donn\'ees issues des classes~0 et~2...
+Si on suppose en outre que $\pi(z1)$ et $\pi(z2)$ correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution "nulle" de $z_1$ (l'attribut "exhaustif" pour la classe~1) sur les donn\'ees issues des classes~0 et~2...
 
@@
 
-Est-ce \`a dire que le maximum de vraisemblance compl\`ete est toujours la meilleure approche?
+Est-ce \`a dire que le maximum de vraisemblance "compl\`ete" est toujours la meilleure approche?
 On a d\'ej\`a compris que son application emp\^eche les experts d'\^etre ind\'ependants.
@@
 
-Pour avoir un mod\`ele complet, on doit avoir une repr\'esentation {\em originelle} des donn\'ees, c-\`a-d qu'on doit d\'efinir la variable $y$ alors que celle-ci n'a n'a pas lieu d'\^etre dans l'approche composite.
-L'ajustement des lois $\pi(z_x)$ via la vraisemblance compl\`ete est-il ind\'ependant de la repr\'esentation originelle?
+Pour avoir un mod\`ele complet, on doit avoir une repr\'esentation "originelle" des donn\'ees, c-\`a-d qu'on doit d\'efinir la variable y alors que celle-ci n'a n'a pas lieu d'\^etre dans l'approche composite.
+L'ajustement des lois $\pi(z_x)$ via la vraisemblance compl\`ete est-il ind\'ependant de la repr\'esentation "originelle"?
 La r\'eponse est non.
-Si on inclut dans y un attribut z3 qui a, par exemple, la m\^eme variance que $z_1$ sous l'hypoth\`ese de r\'ef\'erence alors, clairement, l'estimation par maximum de vraisemblance compl\`ete de $\pi(z_1)$ va d\'ependre de $z_3$...
+Si on inclut dans y un attribut z3 qui a, par exemple, la m\^eme variance que $z_1$ sous l'hypoth\`ese de r\'ef\'erence alors, clairement, l'estimation par maximum de vraisemblance "compl\`ete" de $\pi(z_1)$ va d\'ependre de $z_3$...
 c-\`a-d qu'on devra mesurer $z_3$ pour pouvoir ajuster cette loi.
@@
 \item coh\'erence asymptotique (et oui, c'est maintenu car la vraisemblance   restreinte reste la vraisemblance...) 
-\item ind\'ependance vis-\`a-vis de toute repr\'esentation ``universelle'' 
+\item ind\'ependance vis-\`a-vis de toute repr\'esentation "universelle" 
 \item flexibilit\'e: ind\'epedendance des experts et donc possibililt\'e de   combiner leurs r\'esultats sans connaitre les donn\'ees utilis\'ees (et   donc notamment s'ils ont utilis\'e des donn\'ees diff\'erentes)
@@
 
-Je suis d\'esormais convaincu de la validit\'e du mod\`ele g\'en\'eratif i.i.d.
-pr\'esent\'e dans ma th\`ese pour le recalage: il est justifi\'e par le th\'eor\`eme d'\'echangeabilit\'e de de Finetti dans la mesure o\`u on suppose que les pixels/voxels sont m\'elang\'es au pr\'ealable, ce qui revient \`a dire que la repr\'esentation ``universelle'' est un ensemble de pixels (donc sans notion d'ordre).
+Je suis d\'esormais convaincu de la validit\'e du mod\`ele g\'en\'eratif "i.i.d" pr\'esent\'e dans ma th\`ese pour le recalage: il est justifi\'e par le th\'eor\`eme d'\'echangeabilit\'e de de Finetti dans la mesure o\`u on suppose que les pixels/voxels sont m\'elang\'es au pr\'ealable, ce qui revient \`a dire que la repr\'esentation "universelle" est un ensemble de pixels (donc sans notion d'ordre).
 
@@
 On peut voir l'\'elimination de param\`etres comme un probl\`eme d'approximation ou, si on pr\'ef\`ere, de {\em d\'ecision}, c-\`a-d un probl\`eme de choix d'une valeur unique dans un ensemble.
-Il s'agit ici de repr\'esenter un mod\`ele param\'etrique par une distribution unique afin d'\'evaluer {\em l'\'evidence du mod\`ele}, c-\`a-d la probabilit\'e qu'il conf\`ere aux donn\'ees.
+Il s'agit ici de repr\'esenter un mod\`ele param\'etrique par une distribution unique afin d'\'evaluer "l'\'evidence" du mod\`ele, c-\`a-d la probabilit\'e qu'il conf\`ere aux donn\'ees.
 
@@
 
-Une id\'ee naturelle est de choisir une sorte de centre de la famille, c'est-\`a-dire un \'el\'ement situ\'e \`a faible ``distance'' de tout autre \'el\'ement.
-La ``distance'' en question d\'efinit la fonction de perte et peut \^etre prise, par exemple, au hasard, comme la divergence KL:
+Une id\'ee naturelle est de choisir une sorte de centre de la famille, c'est-\`a-dire un \'el\'ement situ\'e \`a faible "distance" de tout autre \'el\'ement.
+La "distance" en question d\'efinit la fonction de perte et peut \^etre prise, par exemple, au hasard, comme la divergence KL:
 $$L(\theta, q) = D(p_\theta\|q)$$
@@
 Sa justification classique (cf.
-Gr\"unwald ou Rissanen) repose sur la notion de {\em regret}, c-\`a-d la diff\'erence de longueur de code entre la distribution caract\'eristique candidate et la distribution qui compresse le plus l'observation:
+Gr\"unwald ou Rissanen) repose sur la notion de "regret", c-\`a-d la diff\'erence de longueur de code entre la distribution caract\'eristique candidate et la distribution qui compresse le plus l'observation:
 $$
@@
 
-Le maximum de vraisemblance apparait naturellement dans cette formulation en tant que ``meilleur mod\`ele'' mais c'est un mod\`ele na\"if car $\ell_\star(y)$ ne d\'efinit pas une distribution pr\'edictive valide \`a cause du couplage entre l'observation et l'estim\'ee MV.
+Le maximum de vraisemblance apparait naturellement dans cette formulation en tant que "meilleur mod\`ele" mais c'est un mod\`ele "na\"if" car $\ell_\star(y)$ ne d\'efinit pas une distribution pr\'edictive valide \`a cause du couplage entre l'observation et l'estim\'ee MV.
 
 Une diff\'erence essentielle avec l'approche pr\'ec\'edente est que la fonction de perte d\'epend de l'observation.
-Tout se passe en fait comme si l'observation rempla\c{c}ait theta comme ``\'etat inconnu du monde''.
+Tout se passe en fait comme si l'observation rempla\c{c}ait theta comme "\'etat inconnu du monde".
 Cela a du sens dans la mesure o\`u la distribution caract\'eristique doit \^etre choisie ind\'ependamment de toute observation.
@@
 
-D\`es lors, il est naturel de chercher \`a minimiser le {\em pire regret}: 
+D\`es lors, il est naturel de chercher \`a minimiser le "pire" regret: 
 $$
@@
 
-Au-del\`a de la nature {\em Bayes robuste} de cette formulation, on voit qu'elle consiste plus fondamentalement \`a remplacer $D(p_\theta\|q)$ dans la 1\`ere formulation par $D(p\|q) - D(p\|p_\theta)$ avec l'id\'ee que $(p,\theta)$ repr\'esente l'\'etat du monde.
+Au-del\`a de la nature "Bayes robuste" de cette formulation, on voit qu'elle consiste plus fondamentalement \`a remplacer $D(p_\theta\|q)$ dans la 1\`ere formulation par $D(p\|q) - D(p\|p_\theta)$ avec l'id\'ee que $(p,\theta)$ repr\'esente l'\'etat du monde.
 
@@
 plut\^ot que via la notion empirique d'\'evidence.
-Dans cette perspective, la distribution r\'ealisant le maximum de S sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure {\em optimiste} qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:
+Dans cette perspective, la distribution r\'ealisant le maximum de S sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure "optimiste" qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:
 $$S_\star = \max_\theta S(p, p_\theta)$$
 
-Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la ``vraie'' distribution des donn\'ees, et que S* est donc inconnu. Mais Akaike justifie l'approximation: 
+Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la "vraie" distribution des donn\'ees, et que S* est donc inconnu. Mais Akaike justifie l'approximation: 
 $$S_\star \approx log \ell_\star(y) - k$$ 
@@
 Mais on voit aussi qu'il ne se g\'en\'eralise pas trivialement.
-Le biais d\'epend de la ``vraie'' distribution des donn\'ees; et s'il est grand, on pourrait se faire berner par un mod\`ele incorrect.
+Le biais d\'epend de la "vraie" distribution des donn\'ees; et s'il est grand, on pourrait se faire berner par un mod\`ele incorrect.
 On peut l'estimer empiriquement, ce biais, si \c{c}a rentre dans le budget calculatoire, ou essayer de le majorer...
@@
 Cela \'etant, le r\'esultat montre que le biais sur la log-vraisemblance {\em moyenne} est en $O(1/n)$, o\`u $n$ est le nombre de points.
-Ainsi, lorsqu'on compare des mod\`eles sur un ``grand'' \'echantillon, on peut raisonnablement supposer que le biais est n\'egligeable.
+Ainsi, lorsqu'on compare des mod\`eles sur un "grand" \'echantillon, on peut raisonnablement supposer que le biais est n\'egligeable.
 
@@
 
-J'avais not\'e en 2007 que le paradoxe ``disparait'' si on s'int\'eresse \`a des transformations libres.
+J'avais not\'e en 2007 que le paradoxe "disparait" si on s'int\'eresse \`a des transformations libres.
 Les param\`etres recherch\'es sont alors des appariements et sont d\'ependants de l'\'echantillonnage.
@@
 
-Contrairement \`a l'intuition qui a parfois fauss\'e mon jugement, l'ind\'ependance a posteriori des appariements n'est pas le signe d'un ``exc\`es de confiance'' de l'algorithme.
+Contrairement \`a l'intuition qui a parfois fauss\'e mon jugement, l'ind\'ependance a posteriori des appariements n'est pas le signe d'un "exc\`es de confiance" de l'algorithme.
 Bien au contraire!
 L'ind\'ependance a posteriori, en plus du fait que les lois marginales a posteriori sont ind\'ependantes du contexte (c-\`a-d, n'utilisent pas les pixels voisins du point cible orrespondant) illustre bien l'information limit\'ee fournie par la repr\'esentation.
-La distribution a posteriori pourrait dfficilement \^etre plus {\em entropique}!
+La distribution a posteriori pourrait dfficilement \^etre plus "entropique"!
 
@@
 l'in\'egalit\'e fondamentale du traitement des donn\'ees n'est pas bafou\'ee, au contraire elle est confirm\'ee dans toute sa gloire.
-Mais, alors, si l'incertitude sur des transformations libres est ``grande'', comment peut-elle \^etre nulle pour des transformations param\'etriques?
+Mais, alors, si l'incertitude sur des transformations libres est "grande", comment peut-elle \^etre nulle pour des transformations param\'etriques?
 
 Le probl\`eme n'a rien \`a voir avec notre mod\`ele particulier de repr\'esentation des images.
-Le simple fait d'imaginer qu'on puisse d\'efinir une vraisemblance sur des appariements denses conduit au ``paradoxe''.
+Le simple fait d'imaginer qu'on puisse d\'efinir une vraisemblance sur des appariements denses conduit au "paradoxe".
 
@@
 $$
-ce qui g\'en\'eralise notre {\em curse of dimensionality}: plus la grille sous-jacente \`a $Q$ est grande, plus l'ordre de grandeur de la vraisemblance sur $X$ est grande, et donc plus grands sont les rapports de vraisemblances entre transformations test\'ees.
+ce qui g\'en\'eralise notre "curse of dimensionality": plus la grille sous-jacente \`a $Q$ est grande, plus l'ordre de grandeur de la vraisemblance sur $X$ est grande, et donc plus grands sont les rapports de vraisemblances entre transformations test\'ees.
 
@@
 
-Ceci semble paradoxal mais c'est un effet ``m\'ecanique'' de la dimension.
+Ceci semble paradoxal mais c'est un effet "m\'ecanique" de la dimension.
 Consid\'erons le cas similaire de la r\'egression aux moindres carr\'es, pour laquelle il est bien connu que la variance d'estimation est donn\'ee par:
@@
 La matrice $X^\top X$ \'etant proportionnelle \`a la fr\'equence d'\'echantillonnage, V est une fonction d\'ecroissante de celle-ci, sauf \`a consid\'erer que la variance du bruit augmente avec la fr\'equence d'\'echantillonnage, ce qui a du sens physique si le bruit est moyenn\'e dans la p\'eriode d'\'echantillonnage.
-Ca n'en a pas pour nous car le ``bruit'' est ici l'incertitude marginale sur chaque appariement et n'a donc aucune raison ``physique'' de varier en fonction de l'\'echantillonnage.
+Ca n'en a pas pour nous car le "bruit" est ici l'incertitude marginale sur chaque appariement et n'a donc aucune raison "physique" de varier en fonction de l'\'echantillonnage.
 
@@
 
-Ensuite, notre mod\`ele suppose connues les position des points dans les images, que ce soit les pixels ``natifs'' ou les points interpol\'es.
+Ensuite, notre mod\`ele suppose connues les position des points dans les images, que ce soit les pixels "natifs" ou les points interpol\'es.
 Quand j'attribue \`a un pixel la position de son centre, n'est-ce pas arbitraire?
-La vraie position n'est-elle pas ``quelque part dans le pixel''?
+La vraie position n'est-elle pas "quelque part dans le pixel"?
 
@@
 Est-ce qu'on r\'ecup\`ere exactement la fonction originale?
-Est-ce que les {\em points saillants} (points de contour, coins...) peuvent \^etre localis\'es avec certitude?
-Evidemment pas
+Est-ce que les points "saillants" (points de contour, coins...) peuvent \^etre localis\'es avec certitude?
+Evidemment pas.
 
 D'une part, cela signifie qu'il est impossible \`a un op\'erateur humain d'\'etablir une v\'erit\'e terrain exacte \`a partir d'amers rep\'er\'es \`a l'oeil.
-D'autre part, l'algorithme de recalage op\`ere sur des images qui pr\'esentent la r\'ealit\'e d'une mani\`ere ``floue'' non prise en compte par notre mod\`ele statistique.
+D'autre part, l'algorithme de recalage op\`ere sur des images qui pr\'esentent la r\'ealit\'e d'une mani\`ere "floue" non prise en compte par notre mod\`ele statistique.
 
-Comment pourrait-il l'\^etre?
-La mani\`ere la plus intuitive, \`a mon sens, est de faire d\'ependre les appariements de la transformation de fa\c con non-d\'eterministe en introduisant une incertitude de localisation r\'esiduelle.
-C'est d\'ej\`a ce que Maes faisait implicitement en 1997 avec l'id\'ee de l'interpolation PV.
-Cela revient ici \`a modifier $P(Q|X)$ de fa\c{c}on simple.
-Pour calculer la vraisemblance $L(X)$, on peut utiliser un algorithme~EM dans lequel $Q$ est consid\'er\'e comme donn\'ee manquante.
-On voit alors que le calcul de la vraisemblance fait apparaitre des poids analogues \`a l'interpolation PV mais {\em modul\'es it\'erativement par la distribution conjointe d'intensitit\'e}.
+Comment pourrait-il l'\^etre? En introduisant une incertitude sur la localisation des points? Question ouverte.
 

### 2018-11-09 papers/composite_inference/draft.tex
Add revised composite inference paper project

\documentclass[english]{scrartcl}

\usepackage{fullpage}
\usepackage{amstext,amssymb,amsmath,amsthm}
\usepackage{graphicx,subfigure}
\usepackage{multirow}
\usepackage{url}
\usepackage{hyperref}
\usepackage{color}

\graphicspath{{pics/}}

%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{proposition}[theorem]{Proposition}

\title{Composite Bayesian inference}
\date{May 3, 2016}

\author{Alexis Roche\thanks{alexis.roche@\{centraliens.net,gmail.com,epfl.ch\}}}
%\\
%CHUV / Siemens Healthcare / EPFL \\ 
%CH-1015 Lausanne, Switzerland}

%\newcommand{\fix}{\marginpar{FIX}}
%\newcommand{\new}{\marginpar{NEW}}
%\newcommand{\matphi}{\boldsymbol{\Phi}} 
%\def\x{{\mathbf{x}}}
%\def\z{{\mathbf{z}}}
%\def\u{{\mathbf{u}}}
%\def\p{{\bar{\mathbf{p}}}}
%\def\q{{\bar{\mathbf{q}}}}

\begin{document}

\maketitle

\begin{abstract}
This paper revisits the concept of composite likelihood from the perspective of probabilistic inference, and proposes a generalization called {\em super composite likelihood} for sharper inference in multiclass problems. It is argued that, beside providing a new interpretation and a general justification of na\"ive Bayes procedures, super composite likelihood yields a much wider class of discriminative models suitable for unsupervised and weakly supervised learning.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Conventional Bayesian inference and other likelihood-based paradigms rest upon the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable. Because this is challenging when the data is inherently complex, common practice to implement feasible inference algorithms is to use deliberately misspecified data-generating models \cite{White-82,Walker-13} such as in {\em na\"ive Bayes} \cite{Ng-01} or minimum description length \cite{Grunwald-07}, or to resort to highly supervised discriminative modeling approaches \cite{Ng-01}, not to mention ad-hoc methodology.

Composite likelihood (see \cite{Varin-11} and the references therein) is a middle-way approach that extends the familiar notion of likelihood without requiring a full data-generating model. The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole. This may also be viewed as a {\em divide \& conquer} method of approximating the actual likelihood. While maximum composite likelihood does not inherit the general property of maximum likelihood to yield asymptotically minimum-variance estimators, it may offer an excellent trade-off between computational and statistical efficiency.

In this note, composite likelihood is interpreted as a probabilistic opinion pool \cite{Genest-86,Garg-04} of ``agents'' using different pieces of information, or clues, extracted from the data. Each agent acts as a local Bayesian statistician expressing an opinion in the form of a posterior distribution on the unknown parameters of interest, or hypotheses, given a specific clue. Composite likelihood can therefore be associated with a probability distribution on hypotheses, hence extending Bayesian analysis to problems where the proper likelihood function is intractable.

I further justify a generalization of composite likelihood called {\em super composite likelihood} whereby clues can be weighted differently depending on hypotheses to better deal with multiclass inference problems. This more general concept also encompasses another likelihood approximation strategy known as {\em PDF~projection} \cite{Baggenstoss-03,Minka-04,Baggenstoss-15}.

This paper is intended to serve as supporting material for a companion paper (in preparation), where the super composite likelihood framework is applied in computer vision to develop a probabilistic theory of image registration \cite{Viola-97}.

\section{Composite likelihood as opinion pooling}
\label{sec:pool}

Let $Y$ an observable multivariate random variable with sampling distribution $p(y|\theta)$ conditional on some unobserved parameter of interest, $\theta\in\Theta$, where $\Theta$ is a known set. Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $\theta$:
$$
L(\theta) = p(y|\theta)
.
$$

For a high-dimensional $Y$, this expression may be intractable if a plausible generative model is lacking, or involves nuisance parameters that are cumbersome to integrate out. A natural workaround known as {\em data reduction} is then to extract some lower-dimensional feature $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
$$
\ell(\theta) = p(z|\theta)
.
$$

Substituting $L(\theta)$ with $\ell(\theta)$ boils down to restricting the sample space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information. While it is valid for such an agent observing~$z$ only to consider $\ell(\theta)$ as the likelihood function of the problem, the drawback is that $\ell(\theta)$ might be poorly informative about~$\theta$ due to the information loss incurred by data reduction. To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to aggregate the likelihood functions $\ell_i(\theta) = p(z_i|\theta)$ that they elicit.

This leads to a classical problem of combining probabilistic opinions from possibly redundant agents \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. Genest {\em et al} \cite{Genest-86b} showed, in particular, that the only pooling operator that does not explicitly depend on~$\theta$ and preserves {\em external Bayesianity} is the generalized logarithmic opinion pool:
\begin{equation}
\label{eq:log_pool}
p_\star(\theta) = \frac{1}{Z} \pi(\theta) \prod_{i=1}^n \ell_i (\theta)^{w_i},
\qquad
{\rm with}\quad 
Z = \int \pi(\theta) \prod_{i=1}^n \ell_i (\theta)^{w_i} d\theta,
\end{equation} 
where $\pi(\theta)$ is some reference distribution or prior, and $w_i$ are arbitrary positive\footnote{Negative weights can be chosen only if the parameter set $\Theta$ is finite \cite{Genest-86b}.} weights that sum up to one,
$$
\sum_{i=1}^n w_i = 1
.
$$

External Bayesianity essentially means that it should not matter whether a prior on~$\theta$ is incorporated before or after pooling opinions, provided that all agents agree on the same prior. Importantly, the log-linear pool does not assume mutual feature independence, as would the same factorized form as (\ref{eq:log_pool}) with all unitary weights, $w_1=\ldots=w_n= 1$. Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature. This results in increasing their weights in the log-linear pool.}.

Strikingly, (\ref{eq:log_pool}) reduces to an analogous of Bayes rule: $p_\star(\theta)\propto \pi(\theta) L_c(\theta,\mathbf{w})$, where $\mathbf{w}=(w_1,w_2,\ldots,w_n)^\top$ denotes the vector of weights, and the quantity:
\begin{equation}
\label{eq:comp_lik}
L_c(\theta,\mathbf{w}) \equiv \prod_{i=1}^n \ell_i (\theta)^{w_i}
\end{equation} 
plays exactly the same role as a traditional likelihood function. $L_c(\theta,\mathbf{w})$ shares a convenient factorized form with the likelihood derived under mutual feature independence, sometimes called {\em na\"ive Bayes} likelihood in the literature \cite{Ng-01}. The key difference is that the single-feature likelihoods are scaled by positive weights~$w_i$ smaller than one, hence producing a flatter posterior distribution. In comparison with the proper likelihood (not assuming feature independence), the clear computational advantage is that we only need to evaluate the marginal feature distributions, rather than the joint distribution of all features.

We recognize in~(\ref{eq:comp_lik}) a general expression known as a {\em marginal composite likelihood} \cite{Varin-11}, although it is derived here under the restriction that the weights sum up to one (as already motivated in \cite{Wang-14} by a different argument using the maximum entropy principle). 
% Accounting here for Xi'An comment on xianblog.wordpress.com
% The sum of the powers is constrained to be equal to one, even though
% I do not understand why the dimensions of the projections play no
% role in this constraint. Simplicity is advanced as an argument,
% which sounds rather weak…
This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $\theta$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.

Previous attempts at Bayesian composite likelihood include tuning a constant weight so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or performing a close-in-spirit curvature adjustment \cite{Ribatet-12}. Such approaches reconcile the frequentist and Bayesian notions of uncertainty to some extent, but are not externally Bayesian since they do not warrant unit sum weights. When we refer to composite likelihood in the sequel, we assume unit sum weights.

%It is an open problem to determine weights that make it appropriate to plug composite likelihood into Bayes rule. The recommendation for unit sum weights is particularly simple compared to, {\em e.g.}, tuning a constant weight so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or performing a close-in-spirit curvature adjustment \cite{Ribatet-12}. Such approaches reconcile the frequentist and Bayesian notions of uncertainty to some extent, but are not externally Bayesian since they do not warrant unit sum weights. When we refer to composite likelihood in the sequel, we assume unit sum weights.

\section{Composite likelihood as message approximation}
\label{sec:message}

Composite likelihood may also be understood as a means to approximate the ``true'' likelihood function\footnote{By {\em true likelihood}, we mean the likelihood corresponding to a specified yet intractable generative model. In practice, such a model is obviously not required.} or, in the language of graphical models, the {\em message} that the data sends to the latent variable~$\theta$. Several ``clues'' are sent to~$\theta$ via different data features, and then integrated assuming {\em non-coalescent} emitting sources, meaning that statistical dependences between clues are treated as unknown, although not ignored. The whole idea is depicted by the factor graph\footnote{See \cite{Bishop-06} for a didactic introduction to factor graphs.} in Figure~\ref{fig:fgraph2}, where the latent variable is connected to multiple factors involving single clues (rather than a single factor involving multiple clues), thereby enabling efficient computations.

%A simplified message is computed by extracting arbitrary features from the data and treating them as {\em non-coalescent} clues in the sense that their statistical dependences are not modeled (although not ignored).

\begin{figure}[!ht]
\begin{center}
\subfigure[Generative model]{\includegraphics[width=.35\textwidth]{fgraph1.pdf}\label{fig:fgraph1}}
\subfigure[Composite likelihood model]{\includegraphics[width=.6\textwidth]{fgraph2.pdf}\label{fig:fgraph2}}
\caption{Factor graph representations of a generative model (a) and its approximation using composite likelihood (b). In (a), the factor connecting the data~$y$ and the latent variable~$\theta$ is the true likelihood, $L(y,\theta)=p(y|\theta)$. In (b), factors connecting~$y$ and the clues $z_1,z_2,\ldots$ represent feature extractions, $\alpha_i(y,z_i)=\delta[z_i-f_i(y)]$, while factors connecting the clues and~$\theta$ are scaled single-feature likelihoods, $\beta_i(z_i,\theta)=p(z_i|\theta)^{w_i}$. In both graphs, the factor shown in black is the prior $\pi(\theta)$.}
\label{fig:fgraph}
\end{center}
\end{figure}

This approximation scheme happens to be optimal in an information-theoretic sense. As noted in \cite{Garg-04}, the log-linear pool minimizes the average Kullback-Leibler (KL) divergence to the probabilistic opinions:
\begin{equation}
\label{eq:avg_kl}
p_\star = \arg\min_p \sum_{i=1}^n w_i D(p\|p_i),
\end{equation}
where $p_i(\theta) \propto \pi(\theta)\ell_i(\theta)$ is the posterior distribution of agent~$i$. Note that, even though the weights are arbitrary in~(\ref{eq:avg_kl}), the optimal solution normalizes them to unit sum. Composite likelihood thus arises as the best possible consensus among agents according to a natural criterion, in addition to satisfying the external Bayesianity axiom.

If one interprets the average divergence from agent opinions~(\ref{eq:avg_kl}) as a proxy for the divergence from ``the truth'', $p(\theta|y)$, composite likelihood can be seen as a variational approximation to the true likelihood. This corresponds to the intuitive notion that a consensus among sufficiently many experts should yield a reasonable guess. An essential difference with usual approximate inference methods \cite{Bishop-06,Minka-05} is that the likelihood function does not need to be computable owing to the use of a proxy. However, as we shall see in Section~\ref{sec:weights}, knowledge of the feature sampling distributions can be further exploited to optimize the likelihood approximation with respect to the composite likelihood weights.

\section{Super composite likelihood}
\label{sec:super}

Due to the distribution of weights between clues, a drawback of composite likelihood is that it is prone to {\em information overload} in the sense that it tends to ``flatten out'' when too many clues are included as relevant clues then get downweighted. If one decides not to merge clues for computational reasons (this would require handling their joint distribution), one could hope to mitigate information overload by assigning strong weights only to those clues that are believed to be ``most informative''.

However, when chosen for computational simplicity, clues may not only convey limited information at individual level: their informativeness may also be very much hypothesis-dependent. Consider, for instance, diagnosing a disease from a routine medical checkup. Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovascular disease -- and conversely for, say, blood pressure.

This motivates a more general setting where clues can be weighted differently depending on hypotheses. To avoid unnecessary technicalities, we will assume from now on a finite set of hypotheses, $\Theta=\{\theta_0,\theta_1,\ldots,\theta_m\}$, where one particular hypothesis, $\theta_0$, is given the special status of reference, or ``null'' hypothesis.

We start by introducing auxiliary binary variables~$t_j$, for $j=1,\ldots,m$, defined by truncation of~$\theta$:
\begin{equation}
\label{eq:binary}
t_j = 
\left\{
\begin{array}{lll}
1 & {\rm if} & \theta = \theta_j \\
0 & {\rm if} & \theta = \theta_0
\end{array}
\right.
.
\end{equation}
One can think of each~$t_j$ as an indicator light that flashes green or red whenever~$\theta$ is in one of the particular two states $\theta_j$ or $\theta_0$, and does not respond otherwise. The collection of all $t_j$'s may be thought of as a population code \cite{Knill-04} for~$\theta$.

The key idea is as follows: instead of approximating the message sent from the data to~$\theta$ in one piece as in Section~\ref{sec:message}, we may approximate each of the simpler messages sent to the $t_j$'s. To that end, we construct a factor graph using the population code, depicted in Figure~\ref{fig:fgraph3}, which is equivalent to the graph in Figure~\ref{fig:fgraph1} for the purpose of computing the posterior distribution~$p(\theta|y)$. This graph involves multiple factors representing the ``truncated'' likelihood functions, 
$$
L_j(t_j)=p(y|t_j)=
\left\{
\begin{array}{lll}
p(y|\theta_j) & {\rm if} & t_j=1 \\
p(y|\theta_0) & {\rm if} & t_j=0
\end{array}
\right.
,
\qquad j=1,\ldots,m,
$$
as opposed to a single factor for the full likelihood $L(\theta)$. In addition, there are factors $\gamma_j$ to synthesize the different messages sent to the binary variables $t_1,\ldots,t_m$ into one sent to~$\theta$:
\begin{equation}
\label{eq:factor_gamma}
\gamma_j(t_j,\theta)
= 
\left\{
\begin{array}{lll}
\delta_{1t_{j}} & {\rm if} & \theta=\theta_j \\
\delta_{0t_{j}} & {\rm if} & \theta\not=\theta_j
\end{array}
\right.
,
\end{equation}
where $\delta$ denotes the Kronecker delta.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.6\textwidth]{fgraph3.pdf}
  \end{center}
\caption{Alternative graph using population coding for the computation of $p(\theta|y)$. See text for details.}
\label{fig:fgraph3}
\end{figure}

The unnormalized joint distribution~$p_2(y,\theta)$ represented by the graph in Figure~\ref{fig:fgraph3} reads:
\begin{eqnarray*}
p_2(y,\theta) 
& = & 
\pi(\theta) \sum_{t_1=0}^{1}\ldots \sum_{t_m=0}^{1} 
\prod_{j=1}^m L_j(t_j) \gamma_j(t_j,\theta) \\
& = & 
\pi(\theta)  
\prod_{j=1}^m \left\{ 
\mathbf{1}_{\{\theta_j\}}(\theta) p(y|\theta_j) + [1-\mathbf{1}_{\{\theta_j\}}(\theta)]p(y|\theta_0)
\right\}\\
& = & 
\pi(\theta) p(y|\theta) p(y|\theta_0)^{m-1}
,
\end{eqnarray*}
clearly yielding the same posterior $p_2(\theta|y)=p(\theta|y)$ as the graph in Figure~\ref{fig:fgraph1}, although generally not the same generative model: $p_2(y|\theta)\propto p(y|\theta)p(y|\theta_0)^{m-1}$ hence $p_2(y|\theta)\not=p(y|\theta)$ unless there are two hypotheses only ($m=1$), or the null distribution~$p(y|\theta_0)$ is uniform. While the inconsistency between generative models is irrelevant to inference, the alternative graph enables a more flexible likelihood approximation scheme, where each factor $L_j(t_j)$ can be substituted with a specific composite likelihood $L_{cj}(t_j, \mathbf{w}_j)$ using own pre-determined weights $\mathbf{w}_j=(w_{1j},w_{2j},\ldots,w_{nj})^\top$ depending on~$j$:
$$
L_{cj}(t_j, \mathbf{w}_j) 
= \prod_{i=1}^n p(z_i|t_j)^{w_{ij}}
= \prod_{i=1}^n \beta_{ij}(z_i, t_j),
$$
with:
\begin{equation}
\label{eq:factor_beta}
\beta_{ij}(z_i, t_j)
= 
\left\{
\begin{array}{lll}
p(z_i|\theta_j)^{w_{ij}} = \ell_i (\theta_j)^{w_{ij}} & {\rm if} & t_j=1 \\
p(z_i|\theta_0)^{w_{ij}} = \ell_i (\theta_0)^{w_{ij}} & {\rm if} & t_j=0
\end{array}
\right.
,
\end{equation}
and:
$$
\forall j \in \{1,\ldots,m\},
\qquad
\sum_{i=1}^n w_{ij} = 1
.
$$

%Let $\mathbf{W}$ denote an $n\times m$ matrix of pre-determined unit sum weights $w_{ij}$ such that each column, denoted $\mathbf{w}_j$, sums up to one. The possibly sparse $\mathbf{W}$ encodes prior knowledge about the relative relevance of the selected clues to the different hypotheses under investigation.

This corresponds to replacing each factor $L_j$ in Figure~\ref{fig:fgraph3} with a subgraph of same structure as the subgraph highlighted in red in Figure~\ref{fig:fgraph2}, resulting in the further modified factor graph shown in Figure~\ref{fig:fgraph4}. Intuitively, hypothesis-dependent weights make it possible to emphasize the clues that are relevant to each particular hypothesis comparison $\theta_j$ {\em vs.}~$\theta_0$, leading to potentially better approximations to the true odds $p(\theta_j|y)/p(\theta_0|y)$ than using constant weights.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.7\textwidth]{fgraph4.pdf}
  \end{center}
\caption{Super composite likelihood factor graph. See respectively (\ref{eq:factor_beta}) and (\ref{eq:factor_gamma}) for the expression of the factors $\beta_{ij}$ and  $\gamma_j$.}
\label{fig:fgraph4}
\end{figure}

We shall point out that the subgraph connecting ${\bf z}=(z_1,z_2,\ldots,z_n)$ and ${\bf t}=(t_1,t_2,\ldots,t_m)$ in Figure~\ref{fig:fgraph4} (shown in red) is equivalent to an undirected bipartite graph, a property shared with restricted Boltzmann machines (RBM) \cite{Fischer-14,Hinton-06,Larochelle-08}. This means that the variables in one layer are independent conditionally on the other layer:
\begin{equation}
\label{eq:bipartite}
p_3({\bf t}|{\bf z}) = \prod_{j=1}^m p_3(t_j|{\bf z}),
\qquad
p_3({\bf z}|{\bf t}) = \prod_{i=1}^n p_3(z_i|{\bf t}),
\end{equation}
with, in this case, 
$$
p_3(t_j|{\bf z}) \propto L_{cj}(t_j, \mathbf{w}_j),
\qquad
p_3(z_i|{\bf t}) = p_3(z_i|\theta) \propto p(z_i|\theta)^{W_i(\theta)},
$$
where $W_i(\theta)$ is defined by $W_i(\theta_j)=w_{ij}$ if $j\in\{1,m\}$ and $W_i(\theta_0)=\sum_{j=1}^m w_{ij}$.

% and may hence be considered as a kind of restricted Boltzmann machine (RBM) \cite{Fischer-14,Hinton-06,Larochelle-08}: specifically, one with real-valued (or even vector-valued) inputs and truncated binary outputs. The parameters of this RBM are the feature sampling distribution parameters as well as the weights~${\bf W}$. Assuming that the former are known, there is a very natural method to learn~${\bf W}$, as we will see in Section~\ref{sec:weights}, which differs from traditional RBM~learning methods, namely contrastive and discriminative learning.

An essential difference with RBM, however, is that the generative distribution $p_3({\bf z}|{\bf t})$ is not intended to be used for model training. Instead, our core assumption is that the marginal feature distributions are learned ``before'' constructing the graph, a step that does not rely on assuming feature independence conditionally on~${\bf t}$ or, equivalently, $\theta$. Therefore, despite satisfying the bipartite condition (\ref{eq:bipartite}), the joint distribution $p_3({\bf t},{\bf z})$ represented in Figure~\ref{fig:fgraph2} is asymmetrical in use, and serves the only purpose of defining a posterior distribution~$p_3(\theta|y)$ via~$p_3({\bf t}|{\bf z})$.

Marginalizing out~${\bf t}$, we find:
\begin{eqnarray*}
p_3(\theta|y) 
& \propto & 
\pi(\theta) \sum_{{\bf t}\in \{0,1\}^m} 
\prod_{j=1}^m p_3({\bf t}|{\bf z}) \gamma_j(t_j,\theta) \\
& = & 
\pi(\theta) \sum_{t_1=0}^{1}\ldots \sum_{t_m=0}^{1} 
\prod_{j=1}^m L_{cj}(t_j, \mathbf{w}_j) \gamma_j(t_j,\theta) \\
& = & 
\pi(\theta) 
\prod_{i=1}^n
\prod_{j=1}^m
\ell_i \left(
\mathbf{1}_{\{\theta_j\}}(\theta) \theta + [1-\mathbf{1}_{\{\theta_j\}}(\theta)]\theta_0
\right)^{w_{ij}} \\
& = & 
\pi(\theta) 
\prod_{i=1}^n \ell_i(\theta_0)^{\sum_{j=1}^m w_{ij}}
\prod_{i=1}^n \left[
\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
\right]^{w_{ij}}
,
\end{eqnarray*}
therefore:
\begin{equation}
\label{eq:new_pool}
p_3(\theta|y) 
=
K \frac{\pi(\theta)}{\pi(\theta_0)}
\prod_{i=1}^n \left[
\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
\right]^{w_i(\theta)},
\end{equation}
where $K$ is a normalizing factor (dependent on~$y$) and the functions $w_i(\theta)$ are defined by $w_i(\theta_j)=w_{ij}$ for $j=1,\ldots,m$ and $w_i(\theta_0)=0$ conventionally.

As an opinion pooling rule, (\ref{eq:new_pool}) is more general than the log-linear pool~(\ref{eq:log_pool}) as it explicitly depends on~$\theta$ through the changing weights. Moreover, since the weights sum up to one for each~$\theta\not=\theta_0$, we have the equivalent expression:
$$
p_3(\theta|y) 
=
K
\prod_{i=1}^n \left[
\frac{\pi(\theta)\ell_i(\theta)}{\pi(\theta_0)\ell_i(\theta_0)}
\right]^{w_i(\theta)},
$$
showing that the pool does not change depending on whether the prior is incorporated before or after combining experts. In other words, (\ref{eq:new_pool}) is, again, externally Bayesian.

As in Section~\ref{sec:pool}, we recognize a Bayes rule-type expression: $p_3(\theta|y)\propto\pi(\theta){\cal L}_c(\theta,\mathbf{W})$, with:
\begin{equation}
\label{eq:super}
{\cal L}_c(\theta,\mathbf{W})
\equiv
\prod_{i=1}^n \left[
\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
\right]^{w_i(\theta)}
,
\end{equation}
where~$\mathbf{W}$ denotes the $n\times m$ weight matrix with general element $w_{ij}$. We call~(\ref{eq:super}) a super composite likelihood (SCL). Note that the SCL evaluated at a particular hypothesis~$\theta_j$ boils down to a composite likelihood ratio,
\begin{equation}
\label{eq:super_ratio}
{\cal L}_c(\theta_j,\mathbf{W}) = 
\frac{L_c(\theta_j,\mathbf{w}_j)}{L_c(\theta_0,,\mathbf{w}_j)}
.
\end{equation}

To see that SCL~(\ref{eq:super}) is indeed a generalization of composite likelihood, assume that $\mathbf{w}_j=\mathbf{w}$ is the same for all hypotheses. This implies that (\ref{eq:super_ratio}) simplifies to ${\cal L}_c(\theta,\mathbf{W})=L_c(\theta,\mathbf{w})/L_c(\theta_0,\mathbf{w})$, the denominator of which is independent from~$\theta$ and can therefore be safely ignored for inference about~$\theta$. In this special case, SCL is equivalent to standard composite likelihood regardless of the chosen reference~$\theta_0$.

Nevertheless, $\theta_0$ plays a crucial normalization role whenever the columns of $\mathbf{W}$ are different to enable complex, possibly sparse, weighting patterns. Consider, for instance, the case where each hypothesis gets evidence from a single clue, so that $\mathbf{W}$ has a single~1 in each column and all other elements~0. The SCL~(\ref{eq:super}) then reduces to a likelihood ratio involving a single, hypothesis-specific clue $z_{\iota(j)}$ determined by some mapping $\iota:\{1,\ldots,m\}\to \{1,\ldots,n\}$:
\begin{equation}
\label{eq:pdf_proj}
{\cal L}_c(\theta_j,\mathbf{W})
= \frac{\ell_{\iota(j)}(\theta_j)}{\ell_{\iota(j)}(\theta_0)}
= \frac{p(z_{\iota(j)}|\theta_j)}{p(z_{\iota(j)}|\theta_0)}
.
\end{equation}

This expression was already used in \cite{Baggenstoss-03,Minka-04,Baggenstoss-15} motivated by a different argument than here, namely the {\em PDF projection theorem}, which states the full data-generating model maximizing relative entropy (with respect to a chosen reference distribution) under knowledge of the feature sampling distributions. As discussed in \cite{Baggenstoss-03}, (\ref{eq:pdf_proj}) closely approximates the true likelihood if each $z_{\iota(j)}$ is a near-sufficient statistic for~$\theta_j$ {\em vs.}~$\theta_0$, a condition that may be difficult to meet if the choice of clues is driven by computational efficiency. SCL provides an alternative interpretation of PDF~projection, while also extending it to multiple clues with unknown statistical dependences.

\section{Weight optimization}
\label{sec:weights}

The general advantage of~SCL over standard composite likelihood is to define a broader class of pseudo-likelihood functions, hence with the potential to better approximate the true likelihood for a suitable choice of the weight matrix~$\mathbf{W}$. On the other hand, standard composite likelihood satisfies good asymptotic properties for any choice of unit sum positive weights, as shown in Appendix~\ref{sec:asymptotic}, but such properties do not extend in full generality to~SCL: a poor likelihood approximation is to be expected if~$\mathbf{W}$ is chosen at random, hence the importance of fine-tuning super composite weights.

To this end, a conventional machine learning strategy may be used. Assume (for now) that a number of examples of inputs and responses $(y^k, \theta^k)$, for $k=1,\ldots,N$, are sampled independently. Given a tentative weight matrix~$\mathbf{W}$, examples elicit different SCL functions ${\cal L}_{ck}(\theta, \mathbf{W})$ as $y^k$ varies across examples. By analogy with classical maximum likelihood model selection, we may consider rating the model ability to predict examples depending on~$\mathbf{W}$ via the sample average logarithm of the~SCL:
\begin{eqnarray*}
\hat{\cal U}(\mathbf{W})
& = & 
\frac{1}{N} \sum_{k=1}^N \log {\cal L}_{ck}(\theta^k, \mathbf{W})\\
& = &  
\frac{1}{N} \sum_{k=1}^N \sum_{i=1}^n w_i(\theta^k) \log \frac{p(z_i^k|\theta^k)}{p(z_i^k|\theta_0)}
,
\end{eqnarray*}
which essentially averages feature-based log-likelihood ratios relative to the chosen reference hypothesis~$\theta_0$.

This utility measure turns out to be particularly appealing to tune SCL weights, for two reasons. First, if examples are drawn from the true yet incompletely known joint distribution $p(y,\theta)=\pi(\theta)p(y|\theta)$, then, under the condition of existence, $\hat{\cal U}(\mathbf{W})$ converges in the limit of many examples to:
\begin{eqnarray}
{\cal U}(\mathbf{W})
& = & 
E\left[
\log {\cal L}_c(\theta, \mathbf{W})
\right] \nonumber \\
& = & 
\sum_{j=1}^m \pi(\theta_j) \sum_{i=1}^n w_{ij} 
\underbrace{\int p(z_i|\theta_j) \log \frac{p(z_i|\theta_j)}{p(z_i|\theta_0)} dz_i}_{u_{ij}},
\label{eq:discrim}
\end{eqnarray}
where $u_{ij}$ denotes the KL~divergence of~$p(z_i|\theta_0)$ from~$p(z_i|\theta_j)$, and can be interpreted as the utility of clue~$i$ regarding hypothesis~$j$. Since the $u_{ij}$'s are fully determined by the known feature-generating distributions~$p(z_i|\theta)$, (\ref{eq:discrim}) can be evaluated without further knowledge about~$p(y,\theta)$. Consequently, we do not need to constitute an actual training dataset!

Second, let~${\cal U}_\star$ denote the expected utility associated with the true distribution~$p(y,\theta)$, considered as a SCL built from a single clue, the full data~$y$. As is easy to check, ${\cal U}_\star$ is the conditional Kullback-Leibler divergence of~$p(y|\theta_0)$ from~$p(y|\theta)$:
$$
{\cal U}_\star
= \sum_{j=1}^m \pi(\theta_j) \int p(y|\theta_j) \log \frac{p(y|\theta_j)}{p(y|\theta_0)} dy
= D[p(y|\theta)\|p(y|\theta_0)]
.
$$

Applying in~(\ref{eq:discrim}) the data reduction inequality recalled in Appendix~\ref{sec:reduction_inequality}, we obtain the intuitive result that~${\cal U}(\mathbf{W})$ is upper bounded by the expected utility achievable without data reduction: 
$$
{\cal U}(\mathbf{W})\leq {\cal U}_\star
.
$$
This means that the full data-generating model maximizes the expected log-SCL over the whole space of SCL functions ({\em i.e.}, SCL functions built from arbitrary clues in arbitrary number and using arbitrary weights). Therefore, (\ref{eq:discrim}) qualifies as a measure of goodness of fit to the true likelihood.

Maximizing~(\ref{eq:discrim}) with respect to~$\mathbf{W}$ amounts to optimizing the fit over a subspace of SCL functions spanned by fixed user-specified features. Some constraints may be imposed on~$\mathbf{W}$, such as assuming identical columns as in the case of standard composite likelihood, or forcing some elements to zero if some feature likelihood functions are only known on subsets of~$\Theta$. We here focus on the case where no constraints are imposed beside, of course, that the columns of~$\mathbf{W}$ lie in the $n$-dimensional simplex. Clearly, the optimal weights are then determined independently for each hypothesis (and thus independently from the prior) as any positive weights satisfying, for all~$j\in\{1,\ldots,n\}$:
\begin{equation}
\label{eq:optimal_weights}
w_{ij} = 0 
\quad {\rm iff}\quad  
i \not\in \arg\max_{i'\in\{1,\ldots,n\}} u_{i'j},
\qquad {\rm and} \quad
\sum_{i=1}^n w_{ij}=1
,
\end{equation}
yielding a sparsity-enforcing rule which assigns non-zero weights only to the clues that achieve maximal KL~utility for a particular hypothesis. If there is a unique KL-maximal, or ``most exhaustive'' clue for each hypothesis, the resulting~SCL function essentially boils down to the PDF~projection method \cite{Baggenstoss-03,Minka-04,Baggenstoss-15}. More generally, if several clues~$z_i$ maximize $u_{ij}$ for some hypothesis~$\theta_j$, then optimal weights are not unique. In such case, it is preferable to assign equal weights to the winning clues in order to ensure that utility is not only large in average, but also stable across examples ({\em i.e.}, low in variance).

Also note that the weight optimality rule~(\ref{eq:optimal_weights}) implies that, for any $\theta_\star \in \Theta$,
$$
E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
\leq
E\left[ \log \frac{L_c(\theta_\star, \mathbf{w}_\star)}{L_c(\theta_0, \mathbf{w}_\star)} \right],
$$
where $\mathbf{w}_\star$ is the weight vector associated with~$\theta_\star$, $E$ stands for the expectation with respect to $p(y|\theta_\star)$, and~$\mathbf{w}$ is any weight vector. Moreover, the following inequality is shown in Appendix~\ref{sec:asymptotic}:
$$
E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
\leq
E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
,
$$
for any hypothesis~$\theta$ and weighting~$\mathbf{w}$ and, in particular, for the weight vector associated with~$\theta$ via~(\ref{eq:optimal_weights}). Therefore, provided that the weight matrix~$\mathbf{W}$ verifies~(\ref{eq:optimal_weights}), we have that:
$$
E \left[\log {\cal L}_c(\theta,\mathbf{W}) \right]
\leq 
E \left[\log {\cal L}_c(\theta_\star,\mathbf{W}) \right]
,
$$
meaning that the SCL is asymptotically consistent in the sense that the expectation of its logarithm is maximized by the ``true'' parameter value. This is a most awaited frequentist property which, as already pointed out, holds with any choice of weights for standard composite likelihood but not, in general, for~SCL.

%Regularized version.
%$$
%{\cal U}_\gamma(\mathbf{W})
%= \sum_{j=1}^m \pi(\theta_j) \sum_{i=1}^n (w_{ij}u_{ij}-\frac{\gamma}{2}w_{ij}^2)
%$$
%$\gamma>0$ is a constant. Maximizing this leads to independent quadratic programming problems for each~$j$. Can be solved using standard numerical methods \cite{Nocedal-99}. As $\gamma\to 0$, the optimal weights are uniform among clues satisfying (\ref{eq:optimal_weights}).

\section{Further extensions}

\subsection{Conditional super composite likelihood}
\label{sec:conditional}

The SCL derivation rests upon the definition of feature-based likelihood as $\ell_i(\theta)=p(z_i|\theta)$. As a straightforward  extension, $\ell_i(\theta)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
\begin{equation}
\label{eq:cond_feat_lik}
\ell_i(\theta) = p(z_i|\theta,z^c_i).
\end{equation}

Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $\theta$, but can provide relevant information when considered jointly with $z_i$, as in the case of regression covariates, for instance. Standard composite likelihood (\ref{eq:comp_lik}) then amounts to {\em conditional composite likelihood} \cite{Varin-11}, a more general form of composite likelihood also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74}, which was a major breakthrough in computer vision.

Likewise, the above derivation remains valid when ``independent'' features are used, and we can thus define a conditional version of SCL by plugging likelihood functions of the form~(\ref{eq:cond_feat_lik}) into (\ref{eq:super}).

\subsection{Nuisance parameters}
\label{sec:nuisance_params}

Most often, plausible feature-generating models involve unknown quantities of no direct interest. Such quantities may be estimated offline in a supervised learning phase if a suitable training dataset is available. However, if such ``pre-training'' is not feasible in practice, the feature-based likelihoods depend on an unknown parameter~$\psi$, in addition to the parameter of interest~$\theta$:  $\ell_i(\theta,\psi)=p(z_i|\theta,\psi)$, or $\ell_i(\theta,\psi)=p(z_i|z^c_i,\theta,\psi)$ if ``independent'' features are used. We then face two difficulties: 
\begin{itemize}
\item {\em Parameter integration.} How to make the inference on~$\theta$ independent from~$\psi$ assuming known SCL~weights?
\item {\em Weight optimization.} How to optimize the SCL~weights under unknown~$\psi$?
\end{itemize}

We address these two points in the sequel.

%We may then seek a means to eliminate~$\psi$.

\subsubsection{Parameter integration}
\label{sec:nuisance_integration}

When weighting clues independently from the hypotheses, a joint composite likelihood on both parameters may be derived:
$$
L_c(\theta,\psi,\mathbf{w}) = \prod_i \ell_i(\theta,\psi)^{w_i},
$$
and further integrated with respect to some prior on the nuisance parameter to yield a function of~$\theta$ only, which we may call the {\em composite evidence}:
\begin{equation}
\label{eq:comp_evidence}
\bar{L}_c(\theta, \mathbf{w}) = \int \pi(\psi) L_c(\theta,\psi,\mathbf{w}) d\psi
.
\end{equation}
This corresponds to using the factor graph represented in Figure~\ref{fig:fgraph5} to approximate the message from the data to~$\theta$, where the factors $\beta_i$ are now defined by $\beta_i(z_i,\theta,\psi)=\ell_i(\theta,\psi)^{w_i}$. To compute the associated posterior $p_4(\theta|y)$, $\psi$ is treated as an auxiliary variable to be integrated out, leading to $p_4(\theta|y)\propto \pi(\theta)\bar{L}_c(\theta, \mathbf{w})$.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.6\textwidth]{fgraph5.pdf}
  \end{center}
\caption{Extension of the composite likelihood factor graph in Figure~\ref{fig:fgraph2} to account for nuisance parameters.}
\label{fig:fgraph5}
\end{figure}

More generally, when weighting clues depending on hypotheses, we may use essentially the same idea as in Section~\ref{sec:super}, {\em i.e.}, replicate subgraphs such as the one depicted in red and magenta in Figure~\ref{fig:fgraph5} in order to approximate the different messages sent from the data to each truncated variable~$t_j$, as defined in~(\ref{eq:binary}). Note that, by replicating the same graph structure across variables~$t_j$, the variable~$\psi$ is also replicated, so that the resulting factor graph in Figure~\ref{fig:fgraph6} does not represent a joint distribution of the form $p(y,\theta,\psi)$, but rather one of the form~$p(y,\theta,\psi_1,\ldots,\psi_m)$. This makes the approximation more flexible by taking advantage of the fact that a marginal distribution on~$\psi$ is not required.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.7\textwidth]{fgraph6.pdf}
  \end{center}
\caption{Extension of the super composite likelihood factor graph in Figure~\ref{fig:fgraph4} to account for nuisance parameters. Here, the factors $\beta_{ij}(z_i,t_j,\psi_j)$ are defined by $\beta_{ij}(z_i,1,\psi_j)=\ell_i(\theta_j,\psi_j)^{w_{ij}}$ and $\beta_{ij}(z_i,0,\psi_j)=\ell_i(\theta_0,\psi_j)^{w_{ij}}$.}
\label{fig:fgraph6}
\end{figure}

The posterior distribution encoded by this graph (integrated with respect to the replicates $\psi_1,\ldots,\psi_m$) is easily found to be:
$$
p_5(\theta|y)
\propto
\pi(\theta)
\left\{
\begin{array}{lll}
\displaystyle
\frac{\bar{L}_c(\theta_j,\mathbf{w}_j)}{\bar{L}_c(\theta_0,\mathbf{w}_j)} & {\rm if} & \theta=\theta_j \quad {\rm with} \quad j\in\{1,\ldots,m\}\\
1 & {\rm if} & \theta=\theta_0
\end{array}
\right.
,
$$
where $\bar{L}_c(.,\mathbf{w}_j)$ is the above-defined composite evidence function~(\ref{eq:comp_evidence}) associated with the weight vector~$\mathbf{w}_j$. This justifies defining the {\em super composite evidence} as:
\begin{equation}
\label{eq:super_comp_evidence}
\bar{\cal L}_c(\theta_j,\mathbf{W}) = 
\left\{
\begin{array}{lll}
\displaystyle
\frac{\bar{L}_c(\theta_j,\mathbf{w}_j)}{\bar{L}_c(\theta_0,\mathbf{w}_j)} & {\rm if} & j\in\{1,\ldots,m\}\\
1 & {\rm if} & j=0
\end{array}
\right.
.
\end{equation}

Obvious from~(\ref{eq:super_comp_evidence}) is that the super composite evidence conserves all odds $\theta_j$ {\em vs.}~$\theta_0$ integrated within their respective associated subgraphs, very much like its nuisance parameter-free version~(\ref{eq:super_ratio}). The conservation of odds implies  that evaluating the SCL at a particular~$\theta$ is akin to computing a scaled Bayes factor, which may sometimes be efficiently approximated via a maximum likelihood ratio statistic.

%The philosophical question, maybe, is whether it's better to approximate the integrated likelihood, or integrate the approximated likelihood.

\subsubsection{Weight optimization}
\label{sec:nuisance_weights}

Consistently with the idea of nuisance parameter replication (see Figure~\ref{fig:fgraph5}), we may employ a nested strategy to optimize the SCL~weights in presence of nuisance parameters: for each $j\in\{1,\ldots,m\}$, apply the weight optimization of Section~\ref{sec:weights} to the  subgraph model connecting all clues~$z_i$ with the particular pair $(t_j,\psi_j)$. This leads to a set of linear programming problems: for each~$j$, find the weight vector~${\bf w}_j$ on the $n$-dimensional simplex that maximizes:
$$
\bar{\cal U}_j({\bf w}_j) 
= 
E\left[\log \frac{L_c(\theta_j,\psi_j,{\bf w}_j)}{L_c(\theta_0,\psi_j,{\bf w}_j)}\right],
$$
where the expectation is taken with respect to $(y,\psi_j)\sim p(y|\theta_j,\psi)\pi(\psi)$, a computation that only requires knowledge of the feature sampling distributions, $p(z_i|\theta,\psi)$ for $i=1,\ldots,n$, in addition to the prior $\pi(\psi)$.

In this way, every ${\bf w}_j$ is optimal for the posterior of~$(t_j,\psi_j)$ or, more precisely, yields the best approximation to the true joint likelihood $L(\theta, \psi)$ for~$\theta$ restricted to the binary set~$\{\theta_0,\theta_j\}$. The ensuing marginal likelihood approximations are therefore as tight as possible for all hypothesis comparisons with~$\theta_0$, as is the goal of SCL~weight optimization.

% En clair, c'est l'idee du college d'experts travaillant sur une hypothese binaire et integrant les parametres de nuisance sans tenir compte de ce que font les autres.

\section{Discussion}
\label{sec:discussion}

Composite likelihood is a relatively recent concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method. In this paper, we have shown (to our best knowledge, for the first time) a deep connection between composite likelihood and  probabilistic opinion pooling, thereby establishing composite likelihood as a class of {\em discriminative models}\footnote{We here define a discriminative model as a model of which the parameters describe the conditional distribution of an unobserved variable of interest given an observable variable, as opposed to a {\em generative} model, which involves parameters encoding for the conditional distribution of an observable given an unobserved variable.} for statistical inference and machine learning. This connection is possible under the mild restriction that composite likelihood weights are chosen to sum up to one.

\subsection{From na\"ive Bayes to super composite likelihood}

In the probabilistic opinion pooling perspective, composite likelihood is essentially a reinterpretation and a generalization of the {\em na\"ive Bayes} paradigm that relaxes the associated ``na\"ive'' mutual feature independence assumption. In particular, when all features are given equal weight, the composite likelihood is the na\"ive Bayes likelihood raised to the power~$1/n$, where~$n$ is the number of selected features, yielding for instance the same maximum a posteriori (MAP) estimator if a flat prior is used. However, beside providing a simple justification to the wide use of na\"ive Bayes MAP algorithms,  composite likelihood using unit sum weights also entails a conservative rescaling of credibility sets derived from na\"ive Bayes, which gets more drastic as the number of features increases. This comes as a consequence of relaxing feature independence.

We further argued that this rescaling may, to some extent, be unduly conservative in multiclass problems if features are weighted uniformly over the space of unobserved labels, leading us to propose the more general concept of super composite likelihood (SCL). SCL essentially approximates likelihood ratios relative to a fixed reference hypothesis using locally weighted composite likelihood functions. Owing to weight adaptability, SCL describes a more general class of discriminative models than standard composite likelihood.

The idea can also be understood in terms of approximating the factor graph in Figure~\ref{fig:fgraph3}, which encodes the true but intractable posterior distribution, by another factor graph of the type depicted in Figure~\ref{fig:fgraph4}. This substitution results in breaking into pieces both the observed and unobserved variable spaces, and assembling a series of concise {\em messages} passing from the former to the latter. Note that this is a pretty unusual case of approximate inference in factor graphs where the factors to be approximated are unknown (or need not be known).

\subsection{Super composite likelihood training}

Any SCL~model is fully determined by the marginal generative distributions of some pre-specified features, which may rely on a moderate number of parameters if low-dimensional features are chosen. There are two approaches to deal with these parameters.

One is to estimate them beforehand by supervised learning, if an adequate training dataset is available. This could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimizes parameters for generation of observable features. An important difference, however, is that contrastive pre-training is unsupervised. On the other hand, SCL pre-training relies on weaker assumptions as it does not assume conditional feature independence unlike RBMs. Once pre-training is complete, the SCL~weights may be tuned by maximum~SCL, as shown in Section~\ref{sec:weights}: this second training stage is, again, supervised in essence, but does not require additional training examples since it is fully determined by the feature distributions learned in the pre-training step.

%{\color{red} Moreover, while RBM parameters are typically refined in a supervised discriminative learning step, disjoint set of parameters for SCL. Dunno how to say that.}

%In such context, SCL competes with classical discriminative models (logistic regression, Gaussian processes \cite{Rasmussen-06}, maximum entropy models \cite{BergerA-96}, etc.), and may compare more or less favorably in practice depending on the amount of training data. For relatively small training datasets, we may hope for more accurate inference using~SCL than using traditional discriminative models, extrapolating from the results of \cite{Ng-01} regarding the comparison between logistic regression and na\"ive Bayes classifiers.
%{\color{red}Ici, il manque la comparaison avec les RBMs qui ne sont pas (forcement) des modeles discriminatifs.}

\begin{figure}[!ht]
\begin{center}
\subfigure[Generative model]{\includegraphics[width=.25\textwidth]{dgraph1.pdf}\label{fig:dgraph1}}
\hspace*{.05\textwidth}
\subfigure[Classical discriminative model]{\includegraphics[width=.25\textwidth]{dgraph2.pdf}\label{fig:dgraph2}}
\hspace*{.05\textwidth}
\subfigure[Composite likelihood model]{\includegraphics[width=.25\textwidth]{dgraph3.pdf}\label{fig:dgraph3}}
\caption{Belief networks representing, respectively: (a) a generative model, (b) a classical discriminative model, and (c) a composite likelihood model. Note the marginal independence between the data~$y$ and the nuisance parameter~$\psi$ in (b).}
\label{fig:dgraph}
\end{center}
\end{figure}

%%, hence alleviating the need for heavily supervised model training

The other approach to deal with feature distribution parameters is to consider them as nuisance parameters, as proposed in Section~\ref{sec:nuisance_params}, thereby avoiding pre-training. The method then becomes completely unsupervised. The weights can be tuned by maximizing the log-SCL integrated over the prior on nuisance parameters (see Section~\ref{sec:nuisance_weights}), which requires no training dataset whatsoever, and the nuisance parameters can be eliminated by substituting the~SCL function with a {\em super composite evidence} (see Section~\ref{sec:nuisance_integration}). In this version, SCL~essentially generalizes Bayesian integration.

The potential of SCL for unsupervised learning is perhaps suprising considering that it is a discriminative model. This stems from the fact that, unlike classical discriminative models, SCL~exploits direct information conveyed by the data about both the parameters of interest and the nuisance parameters, reflecting the generative nature of the underlying feature distribution models. This essential difference is illustrated in Figure~\ref{fig:dgraph}: the factor graph model underlying composite likelihood in its parametric version (see Figure~\ref{fig:fgraph5}) is not reducible to a belief network of the type represented in Figure~\ref{fig:dgraph2}, in which the data and the nuisance parameter are marginally independent. In contrast, the data is informative about the nuisance parameter in a composite likelihood model, as shown in Figure~\ref{fig:dgraph3}.

\section{Conclusion}
\label{sec:conclusion}

In summary, (super) composite likelihood has the potential to yield weakly supervised or unsupervised Bayesian-like inference procedures depending on the particular task at hand. This property reflects the encoding of statistical relationships between the data and {\em all} unknown parameters. Composite likelihood thus appears as a trade-off between generative models, which are optimal for unsupervised learning but possibly intractable, and traditional discriminative models (logistic regression, Gaussian processes \cite{Rasmussen-06}, maximum entropy models \cite{BergerA-96}, etc.), which are inherently supervised. Composite likelihood models are discriminative models assembled from atomic generative models and, from this point of view, may be considered as {\em semi-generative} models.

%composite likelihood may be considered as a {\em semi-generative} model: a discriminative model assembled from partial generative models.

%As a note of caution, we shall stress that the pre-determined weights assigned to the different associations between observed and unobserved values represent prior knowledge regarding the informativeness of clues. A poor choice of weights will inevitably result in a poor approximation to the ``true'' Bayesian posterior -- the posterior that would be obtained from a realistic generative model if it was tractable. In future work, we will investigate feature selection strategies to mitigate this problem.

% Improve the discussion on following aspects:
% ** Why is it compatible with unsupervised learning? Give more insight.
% ** Stress the contribution: class-specific weighting.
% * Pivotality argument.
% * Bayes is a special case of composite Bayes.

%{\color{red} {\em Mardi 12 avril 2016.} Je laisse reposer la version actuelle 3-4 semaines avant mise \`a jour sur arxiv. La plupart des points qui restaient obscurs en f\'evrier ont, me semble-t-il, \'et\'e lev\'es: lien avec les neural networks/RBM, strat\'egies d'apprentissage, optimisation des poids, int\'er\^et du super composite (multi-class problems). Sur l'optimisation des poids, il reste une petite zone d'ombre: faut-il r\'egulariser d'apprentissage de mani\`ere \`a \'eviter une pond\'eration \'eparse? L'argument pour serait d'acc\'el\'erer la convergence stochastique de la SCL vers l'utilit\'e th\'eorique. On en parle d\'ej\`a un peu dans la version actuelle en disant que si 2 attributs sont aussi informatifs l'un que l'autre, alors le mieux est de leur donner le m\^eme poids. Plus g\'en\'eralement, on pourrait aborder cet aspect via: 1) des contraintes de r\'epartition ``d\'emocratique'' des poids pour \'eviter une situation ``winner-take-all'' (eg, avec une p\'enalisation quadratique des poids); ou alors, 2) des contraintes de variations impos\'ees aux colomnes de la matrice de poids, ce qui donnerait une esp\`ece de compromis entre CL classique et SCL... mais je ne suis pas certain de l'int\'er\^et.}

\appendix

\section{Data reduction inequality}
\label{sec:reduction_inequality}

A fundamental property of the Kullback-Leibler divergence is that it decreases under feature extraction, {\em i.e.}, application of a deterministic transformation. This comes as a consequence of the logarithmic sum inequality \cite{Cover-06} (and also follows as a straightforward corollary of the PDF~projection theorem \cite{Minka-04,Baggenstoss-15}).

The proof is elementary and can be sketched as follows. Let $Z=f(Y)$ some feature extracted from a variable~$Y\sim p(y)$. Given an arbitrary distribution~$\pi(y)$, let  $\tilde{p}(z)$ and $\tilde{\pi}(z)$ the distributions induced on~$Z$ by $p(y)$ and $\pi(y)$, respectively. For any potential value~$z$ of~$Z$, consider the level set: $\Gamma(z)=\{y, f(y)=z\}$. Under conditions of existence, we can partition the integral involved in the KL~divergence $D(p\|\pi)$ using these level sets: 
$$
\int p(y) \log \frac{p(y)}{\pi(y)} dy
=
\int \left( \int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy \right) dz\\
.
$$

Applying the logarithmic sum inequality \cite{Cover-06} to each integral over~$\Gamma(z)$, we then readily get:
$$
\int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy
\geq 
\tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)}
,
$$
owing to the fact that $\displaystyle \int_{\Gamma(z)} p(y) dy = \tilde{p}(z)$ by definition. Therefore,
$$
\int \tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)} dz
\leq 
\int p(y) \log \frac{p(y)}{\pi(y)} dy
,
$$
or, in a more compact way, $D(\tilde{p}\|\tilde{\pi})\leq D(p\|\pi)$. The case of equality occurs if, and only if, $p(y)/\pi(y)=\tilde{p}(z)/\tilde{\pi}(z)$, in other words, if~$z$ is a sufficient statistic for the ``alternative'' hypothesis~$H_1:y\sim p(y)$ versus the ``null'' hypothesis~$H_0:y\sim\pi(y)$.

Since it is a well-known fact that $D(\tilde{p}\|\tilde{\pi})\geq 0$ as any KL~divergence, we conclude that:
\begin{equation}
\label{eq:reduction_inequality}
0 
\leq D(\tilde{p}\|\tilde{\pi}) 
\leq D(p\|\pi)
.
\end{equation}

\section{Basic asymptotic properties of composite likelihood}
\label{sec:asymptotic}

It follows from the double inequality~(\ref{eq:reduction_inequality}) that, for any extracted feature~$z_i$ and for any hypothesis~$\theta$,
$$
0 \leq
E\left[
\log \frac{p(z_i|\theta_\star)}{p(z_i|\theta)}
\right]
\leq
E\left[
\log \frac{p(y|\theta_\star)}{p(y|\theta)}
\right],
$$
where the expectation is taken with respect to the true distribution $p(y|\theta_\star)$. Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any standard composite likelihood function (assuming unit sum positive weights):
\begin{equation}
\label{eq:variation_bound}
0 \leq
E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta, \mathbf{w})} \right]
\leq 
E\left[ \log \frac{L(\theta_\star)}{L(\theta)} \right]
.
\end{equation}

This implies two asymptotic properties of standard composite likelihood: 
\begin{itemize}
\item {\em Consistency.} The expected log-composite likelihood is maximized by~$\theta_\star$, the true value of~$\theta$.
\item {\em Conservativeness.} The expected log-composite likelihood function is ``flatter'' than the true expected log-likelihood. In other words, composite likelihood ratios $L_c(\theta, \mathbf{w})/L_c(\theta_\star, \mathbf{w})$ relative to~$\theta_\star$ tend to be larger than the corresponding true likelihood ratios. 
\end{itemize}

These properties do not necessarily extend to the more general case of SCL. We may state a weaker consistency property by considering the upper envelope of expected log-composite likelihood ratios:
$$
M(\theta) = \max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
$$
where ${\cal W}$ is the $n$-dimensional simplex. Clearly, the expected logarithm of the SCL is upper bounded by~$M(\theta)$:
$$
E[\log {\cal L}_c(\theta, \mathbf{W})] \leq M(\theta)
.
$$

Moreover, $M(\theta)$ is maximized by $\theta_\star$ since (\ref{eq:variation_bound}) implies that, for any $(\theta,\theta_0)$,
$$
E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
\leq
E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
$$
which remains true when taking the maximum over the weights in both sides: 
$$
M(\theta) = 
\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
\leq
\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
= M(\theta_\star)
.
$$

Therefore, maximizing SCL corresponds to maximizing a lower bound on~$M(\theta)$, which can be considered as an objective function since it is maximized by~$\theta_\star$ (like the expected log-likelihood). Lower bound maximization guarantees a certain objective value, however not the maximum, hence asymptotic consistency may not hold.

\bibliographystyle{abbrv}
\bibliography{cvis,stat,alexis}

%%\input{draft.biblio}

\end{document}

### 2018-11-13 papers/composite_inference/draft.tex
Modifs on composite inference paper

@@
 
-\graphicspath{{pics/}}
+\graphicspath{{.}{pics/}}
 
@@
 \title{Composite Bayesian inference}
-\date{May 3, 2016}
-
-\author{Alexis Roche\thanks{alexis.roche@\{centraliens.net,gmail.com,epfl.ch\}}}
-%\\
-%CHUV / Siemens Healthcare / EPFL \\ 
-%CH-1015 Lausanne, Switzerland}
+\date{}
+\author{Alexis Roche\thanks{\url{alexis.roche@centraliens.net}}}
 
@@
 \begin{abstract}
-This paper revisits the concept of composite likelihood from the perspective of probabilistic inference, and proposes a generalization called {\em super composite likelihood} for sharper inference in multiclass problems.
-It is argued that, beside providing a new interpretation and a general justification of na\"ive Bayes procedures, super composite likelihood yields a much wider class of discriminative models suitable for unsupervised and weakly supervised learning.
+This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes entropy relative to a given prior and subject to multiple mean log-likelihood constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
+We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, and can be powerful in shallow learning problems, {\em e.g.}, in the context of transfer learning.
 \end{abstract}
@@
 
-Conventional Bayesian inference and other likelihood-based paradigms rest upon the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
-Because this is challenging when the data is inherently complex, common practice to implement feasible inference algorithms is to use deliberately misspecified data-generating models \cite{White-82,Walker-13} such as in {\em na\"ive Bayes} \cite{Ng-01} or minimum description length \cite{Grunwald-07}, or to resort to highly supervised discriminative modeling approaches \cite{Ng-01}, not to mention ad-hoc methodology.
+Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
+Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}, the minimum description length principle \cite{Grunwald-07} and, last but not least, discriminative models, which currently dominate the field of artificial intelligence and typically require supervised learning on large datasets -- these include a variety of deep learning \cite{Lecun-15,Goodfellow-16} and shallow learning techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00} and Gaussian processes \cite{Rasmussen-06}.
 
-Composite likelihood (see \cite{Varin-11} and the references therein) is a middle-way approach that extends the familiar notion of likelihood without requiring a full data-generating model.
+Composite likelihood (see \cite{Varin-11} and the references therein) is a semi-generative approach that extends the familiar notion of likelihood without requiring a full data-generating model.
 The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole.
-This may also be viewed as a {\em divide \& conquer} method of approximating the actual likelihood.
+This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood.
 While maximum composite likelihood does not inherit the general property of maximum likelihood to yield asymptotically minimum-variance estimators, it may offer an excellent trade-off between computational and statistical efficiency.
 
-In this note, composite likelihood is interpreted as a probabilistic opinion pool \cite{Genest-86,Garg-04} of ``agents'' using different pieces of information, or clues, extracted from the data.
-Each agent acts as a local Bayesian statistician expressing an opinion in the form of a posterior distribution on the unknown parameters of interest, or hypotheses, given a specific clue.
-Composite likelihood can therefore be associated with a probability distribution on hypotheses, hence extending Bayesian analysis to problems where the proper likelihood function is intractable.
+In this note, composite likelihood is interpreted as a probabilistic opinion pool of ``agents'' making use of different pieces of information, or clues, extracted from the input data.
+Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a specific clue in the form of a likelihood of the hidden variables.
+The opinion pool then compuptes a probability distribution on the hidden variables analogous to a Bayesian posterior, thereby extending Bayesian inference to problems where the complete likelihood is undefined.
 
-I further justify a generalization of composite likelihood called {\em super composite likelihood} whereby clues can be weighted differently depending on hypotheses to better deal with multiclass inference problems.
-This more general concept also encompasses another likelihood approximation strategy known as {\em PDF~projection} \cite{Baggenstoss-03,Minka-04,Baggenstoss-15}.
-
-This paper is intended to serve as supporting material for a companion paper (in preparation), where the super composite likelihood framework is applied in computer vision to develop a probabilistic theory of image registration \cite{Viola-97}.
+We further show that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum entropy \cite{Grunwald-04}.
+This perspective entails a method to tune the weights of the different agents from training data as in a classical discriminative learning scenario.
 
@@
 
-Let $Y$ an observable multivariate random variable with sampling distribution $p(y|\theta)$ conditional on some unobserved parameter of interest, $\theta\in\Theta$, where $\Theta$ is a known set.
-Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $\theta$:
+Let $Y$ an observable multivariate random variable with sampling distribution $p(y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is a known set.
+Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $x$:
 $$
-L(\theta) = p(y|\theta)
+L(x) = p(y|x)
 .
@@
 
-For a high-dimensional $Y$, this expression may be intractable if a plausible generative model is lacking, or involves nuisance parameters that are cumbersome to integrate out.
-A natural workaround known as {\em data reduction} is then to extract some lower-dimensional feature $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
+This requires a plausible generative model which, for complex data, may be out of reach or involve too many nuisance parameters.
+A natural workaround known as {\em data reduction} is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the more convenient likelihood function:
 $$
-\ell(\theta) = p(z|\theta)
+\ell(x) = p(z|x)
 .
@@
 
-Substituting $L(\theta)$ with $\ell(\theta)$ boils down to restricting the sample space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information.
-While it is valid for such an agent observing~$z$ only to consider $\ell(\theta)$ as the likelihood function of the problem, the drawback is that $\ell(\theta)$ might be poorly informative about~$\theta$ due to the information loss incurred by data reduction.
-To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to aggregate the likelihood functions $\ell_i(\theta) = p(z_i|\theta)$ that they elicit.
+Substituting $L(x)$ with $\ell(x)$ boils down to restricting the sample space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information.
+While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction.
+To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.
 
-This leads to a classical problem of combining probabilistic opinions from possibly redundant agents \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-Genest {\em et al} \cite{Genest-86b} showed, in particular, that the only pooling operator that does not explicitly depend on~$\theta$ and preserves {\em external Bayesianity} is the generalized logarithmic opinion pool:
+If we regard the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
+We will show in Section~\ref{sec:maxent} that one which is particularly well-suited to our case is the {\em log-linear opinion pool}:
 \begin{equation}
 \label{eq:log_pool}
-p_\star(\theta) = \frac{1}{Z} \pi(\theta) \prod_{i=1}^n \ell_i (\theta)^{w_i},
-\qquad
-{\rm with}\quad 
-Z = \int \pi(\theta) \prod_{i=1}^n \ell_i (\theta)^{w_i} d\theta,
+p_\lambda(x|y) = \frac{1}{Z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
 \end{equation} 
-where $\pi(\theta)$ is some reference distribution or prior, and $w_i$ are arbitrary positive\footnote{Negative weights can be chosen only if the parameter set $\Theta$ is finite \cite{Genest-86b}.} weights that sum up to one,
+where $\pi(x)$ is some reference distribution or prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of real-valued weights so that the normalizing factor:
 $$
-\sum_{i=1}^n w_i = 1
-.
+Z_\lambda(y) = \int \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i} dx
 $$
+is finite\footnote{It is possible to have the weights depend on~$y$, however there is a strong rational for constant weights as we shall see in the sequel.}.
 
-External Bayesianity essentially means that it should not matter whether a prior on~$\theta$ is incorporated before or after pooling opinions, provided that all agents agree on the same prior.
-Importantly, the log-linear pool does not assume mutual feature independence, as would the same factorized form as (\ref{eq:log_pool}) with all unitary weights, $w_1=\ldots=w_n= 1$.
-Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature.
+The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding the form: $p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x)$, where the quantity:
+\begin{equation}
+\label{eq:comp_lik}
+L^c_\lambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
+\end{equation} 
+plays the same role as a traditional likelihood function.
+We recognize in~(\ref{eq:comp_lik}) a general expression known as {\em marginal composite likelihood} \cite{Varin-11}.
+It shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, sometimes called {\em na\"ive Bayes} likelihood in the literature \cite{Ng-01}, which corresponds to the special case of all unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
+The general computational advantage of composite likelihood over the true likelihood is that it only requires to evaluate the marginal feature distributions rather than the joint distribution of all features.
+
+\section{The problem of tuning composite likelihood weights}
+
+There exist few general guidelines to choose the composite likelihood weights~$\lambda$ in~(\ref{eq:comp_lik}).
+Common practice is to impose nonnegative weights \cite{Varin-11} although this is not strictly required from the opinion aggregation viewpoint (the normalizing factor $Z_\lambda(y)$ in~(\ref{eq:log_pool}) only needs be finite for any~$y$ for the log-linear pool to yield a proper conditional distribution).
+
+In cases where the features can be considered to be {\em exchangeable}, it is natural to pick uniform weights.
+The composite likelihood function is then a scaled version of the {\em na\"ive Bayes} likelihood, the common weight value being irrelevant to the location of the composite likelihood maximizer.
+It can be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to reconcile the frequentist and Bayesian notions of uncertainty.
+
+Another common recommandation for the composite likelihood weights is to sum up to one.
+
+Going back to the log-linear pool underlying composite likelihood, Genest {\em et al} \cite{Genest-86b} showed that the only pooling operator that does not explicitly depend on~$x$ and preserves a property known as {\em external Bayesianity} is the log-linear opinion pool~(\ref{eq:log_pool}) with the additional constraint that the weights  sum up to one.
+External Bayesianity essentially means that it should not matter whether a prior on~$X$ is incorporated before or after combining opinions, provided that all agents agree on the same prior.
+Garg {\em et al} \cite{Garg-04} showed that the log-linear pool with unit weight sum also minimizes the average Kullback-Leibler (KL) divergence to the agent opinions.
+The MaxEnt argument was already given by \cite{Wang-14}.
+
+%Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature.
 This results in increasing their weights in the log-linear pool.}.
 
-Strikingly, (\ref{eq:log_pool}) reduces to an analogous of Bayes rule: $p_\star(\theta)\propto \pi(\theta) L_c(\theta,\mathbf{w})$, where $\mathbf{w}=(w_1,w_2,\ldots,w_n)^\top$ denotes the vector of weights, and the quantity:
-\begin{equation}
-\label{eq:comp_lik}
-L_c(\theta,\mathbf{w}) \equiv \prod_{i=1}^n \ell_i (\theta)^{w_i}
-\end{equation} 
-plays exactly the same role as a traditional likelihood function.
-$L_c(\theta,\mathbf{w})$ shares a convenient factorized form with the likelihood derived under mutual feature independence, sometimes called {\em na\"ive Bayes} likelihood in the literature \cite{Ng-01}.
-The key difference is that the single-feature likelihoods are scaled by positive weights~$w_i$ smaller than one, hence producing a flatter posterior distribution.
-In comparison with the proper likelihood (not assuming feature independence), the clear computational advantage is that we only need to evaluate the marginal feature distributions, rather than the joint distribution of all features.
+The log-linear pool has, in particular, the 0/1 forcing property, meaning that if any agent assigns zero likelihood to a value~$x$, then $x$ has zero probability in the pool.
 
-We recognize in~(\ref{eq:comp_lik}) a general expression known as a {\em marginal composite likelihood} \cite{Varin-11}, although it is derived here under the restriction that the weights sum up to one (as already motivated in \cite{Wang-14} by a different argument using the maximum entropy principle).
+Unit sum weights = crap.
+
+\section{Maximum entropy derivation of composite likelihood}
+\label{sec:maxent}
+
+The coordinator discards the priors used by the agents.
+
+The MaxEnt argument was already given by \cite{Wang-14}.
+
 % Accounting here for Xi'An comment on xianblog.wordpress.com
@@
 % which sounds rather weak…
-This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $\theta$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.
-
-Previous attempts at Bayesian composite likelihood include tuning a constant weight so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or performing a close-in-spirit curvature adjustment \cite{Ribatet-12}.
-Such approaches reconcile the frequentist and Bayesian notions of uncertainty to some extent, but are not externally Bayesian since they do not warrant unit sum weights.
-When we refer to composite likelihood in the sequel, we assume unit sum weights.
-
-%It is an open problem to determine weights that make it appropriate to plug composite likelihood into Bayes rule.
-The recommendation for unit sum weights is particularly simple compared to, {\em e.g.}, tuning a constant weight so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or performing a close-in-spirit curvature adjustment \cite{Ribatet-12}.
-Such approaches reconcile the frequentist and Bayesian notions of uncertainty to some extent, but are not externally Bayesian since they do not warrant unit sum weights.
-When we refer to composite likelihood in the sequel, we assume unit sum weights.
+This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $x$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.
 
@@
 Composite likelihood may also be understood as a means to approximate the ``true'' likelihood function\footnote{By {\em true likelihood}, we mean the likelihood corresponding to a specified yet intractable generative model.
-In practice, such a model is obviously not required.} or, in the language of graphical models, the {\em message} that the data sends to the latent variable~$\theta$.
-Several ``clues'' are sent to~$\theta$ via different data features, and then integrated assuming {\em non-coalescent} emitting sources, meaning that statistical dependences between clues are treated as unknown, although not ignored.
+In practice, such a model is obviously not required.} or, in the language of graphical models, the {\em message} that the data sends to the latent variable~$x$.
+Several ``clues'' are sent to~$x$ via different data features, and then integrated assuming {\em non-coalescent} emitting sources, meaning that statistical dependences between clues are treated as unknown, although not ignored.
 The whole idea is depicted by the factor graph\footnote{See \cite{Bishop-06} for a didactic introduction to factor graphs.} in Figure~\ref{fig:fgraph2}, where the latent variable is connected to multiple factors involving single clues (rather than a single factor involving multiple clues), thereby enabling efficient computations.
@@
 \caption{Factor graph representations of a generative model (a) and its approximation using composite likelihood (b).
-In (a), the factor connecting the data~$y$ and the latent variable~$\theta$ is the true likelihood, $L(y,\theta)=p(y|\theta)$.
-In (b), factors connecting~$y$ and the clues $z_1,z_2,\ldots$ represent feature extractions, $\alpha_i(y,z_i)=\delta[z_i-f_i(y)]$, while factors connecting the clues and~$\theta$ are scaled single-feature likelihoods, $\beta_i(z_i,\theta)=p(z_i|\theta)^{w_i}$.
-In both graphs, the factor shown in black is the prior $\pi(\theta)$.}
+In (a), the factor connecting the data~$y$ and the latent variable~$x$ is the true likelihood, $p(y|x)$.
+In (b), factors connecting~$y$ and the clues $z_1,z_2,\ldots$ represent feature extractions, $\alpha_i(y,z_i)=\delta[z_i-f_i(y)]$, while factors connecting the clues and~$x$ are scaled single-feature likelihoods, $\beta_i(z_i,x)=p(z_i|x)^{w_i}$.
+In both graphs, the factor shown in black is the prior $\pi(x)$.}
 \label{fig:fgraph}
@@
 
-If one interprets the average divergence from agent opinions~(\ref{eq:avg_kl}) as a proxy for the divergence from ``the truth'', $p(\theta|y)$, composite likelihood can be seen as a variational approximation to the true likelihood.
+If one interprets the average divergence from agent opinions~(\ref{eq:avg_kl}) as a proxy for the divergence from ``the truth'', $p(x|y)$, composite likelihood can be seen as a variational approximation to the true likelihood.
 This corresponds to the intuitive notion that a consensus among sufficiently many experts should yield a reasonable guess.
@@
 
-\section{Super composite likelihood}
-\label{sec:super}
+\section{Conditional composite likelihood}
+\label{sec:conditional}
 
-Due to the distribution of weights between clues, a drawback of composite likelihood is that it is prone to {\em information overload} in the sense that it tends to ``flatten out'' when too many clues are included as relevant clues then get downweighted.
-If one decides not to merge clues for computational reasons (this would require handling their joint distribution), one could hope to mitigate information overload by assigning strong weights only to those clues that are believed to be ``most informative''.
-
-However, when chosen for computational simplicity, clues may not only convey limited information at individual level: their informativeness may also be very much hypothesis-dependent.
-Consider, for instance, diagnosing a disease from a routine medical checkup.
-Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovascular disease -- and conversely for, say, blood pressure.
-
-This motivates a more general setting where clues can be weighted differently depending on hypotheses.
-To avoid unnecessary technicalities, we will assume from now on a finite set of hypotheses, $\Theta=\{\theta_0,\theta_1,\ldots,\theta_m\}$, where one particular hypothesis, $\theta_0$, is given the special status of reference, or ``null'' hypothesis.
-
-We start by introducing auxiliary binary variables~$t_j$, for $j=1,\ldots,m$, defined by truncation of~$\theta$:
+The SCL derivation rests upon the definition of feature-based likelihood as $\ell_i(x)=p(z_i|x)$.
+As a straightforward  extension, $\ell_i(x)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
 \begin{equation}
-\label{eq:binary}
-t_j = 
-\left\{
-\begin{array}{lll}
-1 & {\rm if} & \theta = \theta_j \\
-0 & {\rm if} & \theta = \theta_0
-\end{array}
-\right.
-.
-\end{equation}
-One can think of each~$t_j$ as an indicator light that flashes green or red whenever~$\theta$ is in one of the particular two states $\theta_j$ or $\theta_0$, and does not respond otherwise.
-The collection of all $t_j$'s may be thought of as a population code \cite{Knill-04} for~$\theta$.
-
-The key idea is as follows: instead of approximating the message sent from the data to~$\theta$ in one piece as in Section~\ref{sec:message}, we may approximate each of the simpler messages sent to the $t_j$'s.
-To that end, we construct a factor graph using the population code, depicted in Figure~\ref{fig:fgraph3}, which is equivalent to the graph in Figure~\ref{fig:fgraph1} for the purpose of computing the posterior distribution~$p(\theta|y)$.
-This graph involves multiple factors representing the ``truncated'' likelihood functions,
-$$
-L_j(t_j)=p(y|t_j)=
-\left\{
-\begin{array}{lll}
-p(y|\theta_j) & {\rm if} & t_j=1 \\
-p(y|\theta_0) & {\rm if} & t_j=0
-\end{array}
-\right.
-,
-\qquad j=1,\ldots,m,
-$$
-as opposed to a single factor for the full likelihood $L(\theta)$.
-In addition, there are factors $\gamma_j$ to synthesize the different messages sent to the binary variables $t_1,\ldots,t_m$ into one sent to~$\theta$:
-\begin{equation}
-\label{eq:factor_gamma}
-\gamma_j(t_j,\theta)
-= 
-\left\{
-\begin{array}{lll}
-\delta_{1t_{j}} & {\rm if} & \theta=\theta_j \\
-\delta_{0t_{j}} & {\rm if} & \theta\not=\theta_j
-\end{array}
-\right.
-,
-\end{equation}
-where $\delta$ denotes the Kronecker delta.
-
-\begin{figure}[!ht]
-  \begin{center}
-    \includegraphics[width=.6\textwidth]{fgraph3.pdf}
-  \end{center}
-\caption{Alternative graph using population coding for the computation of $p(\theta|y)$. See text for details.}
-\label{fig:fgraph3}
-\end{figure}
-
-The unnormalized joint distribution~$p_2(y,\theta)$ represented by the graph in Figure~\ref{fig:fgraph3} reads:
-\begin{eqnarray*}
-p_2(y,\theta) 
-& = & 
-\pi(\theta) \sum_{t_1=0}^{1}\ldots \sum_{t_m=0}^{1} 
-\prod_{j=1}^m L_j(t_j) \gamma_j(t_j,\theta) \\
-& = & 
-\pi(\theta)  
-\prod_{j=1}^m \left\{ 
-\mathbf{1}_{\{\theta_j\}}(\theta) p(y|\theta_j) + [1-\mathbf{1}_{\{\theta_j\}}(\theta)]p(y|\theta_0)
-\right\}\\
-& = & 
-\pi(\theta) p(y|\theta) p(y|\theta_0)^{m-1}
-,
-\end{eqnarray*}
-clearly yielding the same posterior $p_2(\theta|y)=p(\theta|y)$ as the graph in Figure~\ref{fig:fgraph1}, although generally not the same generative model: $p_2(y|\theta)\propto p(y|\theta)p(y|\theta_0)^{m-1}$ hence $p_2(y|\theta)\not=p(y|\theta)$ unless there are two hypotheses only ($m=1$), or the null distribution~$p(y|\theta_0)$ is uniform.
-While the inconsistency between generative models is irrelevant to inference, the alternative graph enables a more flexible likelihood approximation scheme, where each factor $L_j(t_j)$ can be substituted with a specific composite likelihood $L_{cj}(t_j, \mathbf{w}_j)$ using own pre-determined weights $\mathbf{w}_j=(w_{1j},w_{2j},\ldots,w_{nj})^\top$ depending on~$j$:
-$$
-L_{cj}(t_j, \mathbf{w}_j) 
-= \prod_{i=1}^n p(z_i|t_j)^{w_{ij}}
-= \prod_{i=1}^n \beta_{ij}(z_i, t_j),
-$$
-with:
-\begin{equation}
-\label{eq:factor_beta}
-\beta_{ij}(z_i, t_j)
-= 
-\left\{
-\begin{array}{lll}
-p(z_i|\theta_j)^{w_{ij}} = \ell_i (\theta_j)^{w_{ij}} & {\rm if} & t_j=1 \\
-p(z_i|\theta_0)^{w_{ij}} = \ell_i (\theta_0)^{w_{ij}} & {\rm if} & t_j=0
-\end{array}
-\right.
-,
-\end{equation}
-and:
-$$
-\forall j \in \{1,\ldots,m\},
-\qquad
-\sum_{i=1}^n w_{ij} = 1
-.
-$$
-
-%Let $\mathbf{W}$ denote an $n\times m$ matrix of pre-determined unit sum weights $w_{ij}$ such that each column, denoted $\mathbf{w}_j$, sums up to one.
-The possibly sparse $\mathbf{W}$ encodes prior knowledge about the relative relevance of the selected clues to the different hypotheses under investigation.
-
-This corresponds to replacing each factor $L_j$ in Figure~\ref{fig:fgraph3} with a subgraph of same structure as the subgraph highlighted in red in Figure~\ref{fig:fgraph2}, resulting in the further modified factor graph shown in Figure~\ref{fig:fgraph4}.
-Intuitively, hypothesis-dependent weights make it possible to emphasize the clues that are relevant to each particular hypothesis comparison $\theta_j$ {\em vs.}~$\theta_0$, leading to potentially better approximations to the true odds $p(\theta_j|y)/p(\theta_0|y)$ than using constant weights.
-
-\begin{figure}[!ht]
-  \begin{center}
-    \includegraphics[width=.7\textwidth]{fgraph4.pdf}
-  \end{center}
-\caption{Super composite likelihood factor graph. See respectively (\ref{eq:factor_beta}) and (\ref{eq:factor_gamma}) for the expression of the factors $\beta_{ij}$ and  $\gamma_j$.}
-\label{fig:fgraph4}
-\end{figure}
-
-We shall point out that the subgraph connecting ${\bf z}=(z_1,z_2,\ldots,z_n)$ and ${\bf t}=(t_1,t_2,\ldots,t_m)$ in Figure~\ref{fig:fgraph4} (shown in red) is equivalent to an undirected bipartite graph, a property shared with restricted Boltzmann machines (RBM) \cite{Fischer-14,Hinton-06,Larochelle-08}.
-This means that the variables in one layer are independent conditionally on the other layer:
-\begin{equation}
-\label{eq:bipartite}
-p_3({\bf t}|{\bf z}) = \prod_{j=1}^m p_3(t_j|{\bf z}),
-\qquad
-p_3({\bf z}|{\bf t}) = \prod_{i=1}^n p_3(z_i|{\bf t}),
-\end{equation}
-with, in this case, 
-$$
-p_3(t_j|{\bf z}) \propto L_{cj}(t_j, \mathbf{w}_j),
-\qquad
-p_3(z_i|{\bf t}) = p_3(z_i|\theta) \propto p(z_i|\theta)^{W_i(\theta)},
-$$
-where $W_i(\theta)$ is defined by $W_i(\theta_j)=w_{ij}$ if $j\in\{1,m\}$ and $W_i(\theta_0)=\sum_{j=1}^m w_{ij}$.
-
-% and may hence be considered as a kind of restricted Boltzmann machine (RBM) \cite{Fischer-14,Hinton-06,Larochelle-08}: specifically, one with real-valued (or even vector-valued) inputs and truncated binary outputs.
-The parameters of this RBM are the feature sampling distribution parameters as well as the weights~${\bf W}$.
-Assuming that the former are known, there is a very natural method to learn~${\bf W}$, as we will see in Section~\ref{sec:weights}, which differs from traditional RBM~learning methods, namely contrastive and discriminative learning.
-
-An essential difference with RBM, however, is that the generative distribution $p_3({\bf z}|{\bf t})$ is not intended to be used for model training.
-Instead, our core assumption is that the marginal feature distributions are learned ``before'' constructing the graph, a step that does not rely on assuming feature independence conditionally on~${\bf t}$ or, equivalently, $\theta$.
-Therefore, despite satisfying the bipartite condition (\ref{eq:bipartite}), the joint distribution $p_3({\bf t},{\bf z})$ represented in Figure~\ref{fig:fgraph2} is asymmetrical in use, and serves the only purpose of defining a posterior distribution~$p_3(\theta|y)$ via~$p_3({\bf t}|{\bf z})$.
-
-Marginalizing out~${\bf t}$, we find:
-\begin{eqnarray*}
-p_3(\theta|y) 
-& \propto & 
-\pi(\theta) \sum_{{\bf t}\in \{0,1\}^m} 
-\prod_{j=1}^m p_3({\bf t}|{\bf z}) \gamma_j(t_j,\theta) \\
-& = & 
-\pi(\theta) \sum_{t_1=0}^{1}\ldots \sum_{t_m=0}^{1} 
-\prod_{j=1}^m L_{cj}(t_j, \mathbf{w}_j) \gamma_j(t_j,\theta) \\
-& = & 
-\pi(\theta) 
-\prod_{i=1}^n
-\prod_{j=1}^m
-\ell_i \left(
-\mathbf{1}_{\{\theta_j\}}(\theta) \theta + [1-\mathbf{1}_{\{\theta_j\}}(\theta)]\theta_0
-\right)^{w_{ij}} \\
-& = & 
-\pi(\theta) 
-\prod_{i=1}^n \ell_i(\theta_0)^{\sum_{j=1}^m w_{ij}}
-\prod_{i=1}^n \left[
-\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
-\right]^{w_{ij}}
-,
-\end{eqnarray*}
-therefore:
-\begin{equation}
-\label{eq:new_pool}
-p_3(\theta|y) 
-=
-K \frac{\pi(\theta)}{\pi(\theta_0)}
-\prod_{i=1}^n \left[
-\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
-\right]^{w_i(\theta)},
-\end{equation}
-where $K$ is a normalizing factor (dependent on~$y$) and the functions $w_i(\theta)$ are defined by $w_i(\theta_j)=w_{ij}$ for $j=1,\ldots,m$ and $w_i(\theta_0)=0$ conventionally.
-
-As an opinion pooling rule, (\ref{eq:new_pool}) is more general than the log-linear pool~(\ref{eq:log_pool}) as it explicitly depends on~$\theta$ through the changing weights.
-Moreover, since the weights sum up to one for each~$\theta\not=\theta_0$, we have the equivalent expression:
-$$
-p_3(\theta|y) 
-=
-K
-\prod_{i=1}^n \left[
-\frac{\pi(\theta)\ell_i(\theta)}{\pi(\theta_0)\ell_i(\theta_0)}
-\right]^{w_i(\theta)},
-$$
-showing that the pool does not change depending on whether the prior is incorporated before or after combining experts. In other words, (\ref{eq:new_pool}) is, again, externally Bayesian.
-
-As in Section~\ref{sec:pool}, we recognize a Bayes rule-type expression: $p_3(\theta|y)\propto\pi(\theta){\cal L}_c(\theta,\mathbf{W})$, with:
-\begin{equation}
-\label{eq:super}
-{\cal L}_c(\theta,\mathbf{W})
-\equiv
-\prod_{i=1}^n \left[
-\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
-\right]^{w_i(\theta)}
-,
-\end{equation}
-where~$\mathbf{W}$ denotes the $n\times m$ weight matrix with general element $w_{ij}$.
-We call~(\ref{eq:super}) a super composite likelihood (SCL).
-Note that the SCL evaluated at a particular hypothesis~$\theta_j$ boils down to a composite likelihood ratio,
-\begin{equation}
-\label{eq:super_ratio}
-{\cal L}_c(\theta_j,\mathbf{W}) = 
-\frac{L_c(\theta_j,\mathbf{w}_j)}{L_c(\theta_0,,\mathbf{w}_j)}
-.
+\label{eq:cond_feat_lik}
+\ell_i(x) = p(z_i|x,z^c_i).
 \end{equation}
 
-To see that SCL~(\ref{eq:super}) is indeed a generalization of composite likelihood, assume that $\mathbf{w}_j=\mathbf{w}$ is the same for all hypotheses.
-This implies that (\ref{eq:super_ratio}) simplifies to ${\cal L}_c(\theta,\mathbf{W})=L_c(\theta,\mathbf{w})/L_c(\theta_0,\mathbf{w})$, the denominator of which is independent from~$\theta$ and can therefore be safely ignored for inference about~$\theta$.
-In this special case, SCL is equivalent to standard composite likelihood regardless of the chosen reference~$\theta_0$.
-
-Nevertheless, $\theta_0$ plays a crucial normalization role whenever the columns of $\mathbf{W}$ are different to enable complex, possibly sparse, weighting patterns.
-Consider, for instance, the case where each hypothesis gets evidence from a single clue, so that $\mathbf{W}$ has a single~1 in each column and all other elements~0.
-The SCL~(\ref{eq:super}) then reduces to a likelihood ratio involving a single, hypothesis-specific clue $z_{\iota(j)}$ determined by some mapping $\iota:\{1,\ldots,m\}\to \{1,\ldots,n\}$:
-\begin{equation}
-\label{eq:pdf_proj}
-{\cal L}_c(\theta_j,\mathbf{W})
-= \frac{\ell_{\iota(j)}(\theta_j)}{\ell_{\iota(j)}(\theta_0)}
-= \frac{p(z_{\iota(j)}|\theta_j)}{p(z_{\iota(j)}|\theta_0)}
-.
-\end{equation}
-
-This expression was already used in \cite{Baggenstoss-03,Minka-04,Baggenstoss-15} motivated by a different argument than here, namely the {\em PDF projection theorem}, which states the full data-generating model maximizing relative entropy (with respect to a chosen reference distribution) under knowledge of the feature sampling distributions.
-As discussed in \cite{Baggenstoss-03}, (\ref{eq:pdf_proj}) closely approximates the true likelihood if each $z_{\iota(j)}$ is a near-sufficient statistic for~$\theta_j$ {\em vs.}~$\theta_0$, a condition that may be difficult to meet if the choice of clues is driven by computational efficiency.
-SCL provides an alternative interpretation of PDF~projection, while also extending it to multiple clues with unknown statistical dependences.
-
-\section{Weight optimization}
-\label{sec:weights}
-
-The general advantage of~SCL over standard composite likelihood is to define a broader class of pseudo-likelihood functions, hence with the potential to better approximate the true likelihood for a suitable choice of the weight matrix~$\mathbf{W}$.
-On the other hand, standard composite likelihood satisfies good asymptotic properties for any choice of unit sum positive weights, as shown in Appendix~\ref{sec:asymptotic}, but such properties do not extend in full generality to~SCL: a poor likelihood approximation is to be expected if~$\mathbf{W}$ is chosen at random, hence the importance of fine-tuning super composite weights.
-
-To this end, a conventional machine learning strategy may be used.
-Assume (for now) that a number of examples of inputs and responses $(y^k, \theta^k)$, for $k=1,\ldots,N$, are sampled independently.
-Given a tentative weight matrix~$\mathbf{W}$, examples elicit different SCL functions ${\cal L}_{ck}(\theta, \mathbf{W})$ as $y^k$ varies across examples.
-By analogy with classical maximum likelihood model selection, we may consider rating the model ability to predict examples depending on~$\mathbf{W}$ via the sample average logarithm of the~SCL:
-\begin{eqnarray*}
-\hat{\cal U}(\mathbf{W})
-& = & 
-\frac{1}{N} \sum_{k=1}^N \log {\cal L}_{ck}(\theta^k, \mathbf{W})\\
-& = &  
-\frac{1}{N} \sum_{k=1}^N \sum_{i=1}^n w_i(\theta^k) \log \frac{p(z_i^k|\theta^k)}{p(z_i^k|\theta_0)}
-,
-\end{eqnarray*}
-which essentially averages feature-based log-likelihood ratios relative to the chosen reference hypothesis~$\theta_0$.
-
-This utility measure turns out to be particularly appealing to tune SCL weights, for two reasons.
-First, if examples are drawn from the true yet incompletely known joint distribution $p(y,\theta)=\pi(\theta)p(y|\theta)$, then, under the condition of existence, $\hat{\cal U}(\mathbf{W})$ converges in the limit of many examples to:
-\begin{eqnarray}
-{\cal U}(\mathbf{W})
-& = & 
-E\left[
-\log {\cal L}_c(\theta, \mathbf{W})
-\right] \nonumber \\
-& = & 
-\sum_{j=1}^m \pi(\theta_j) \sum_{i=1}^n w_{ij} 
-\underbrace{\int p(z_i|\theta_j) \log \frac{p(z_i|\theta_j)}{p(z_i|\theta_0)} dz_i}_{u_{ij}},
-\label{eq:discrim}
-\end{eqnarray}
-where $u_{ij}$ denotes the KL~divergence of~$p(z_i|\theta_0)$ from~$p(z_i|\theta_j)$, and can be interpreted as the utility of clue~$i$ regarding hypothesis~$j$.
-Since the $u_{ij}$'s are fully determined by the known feature-generating distributions~$p(z_i|\theta)$, (\ref{eq:discrim}) can be evaluated without further knowledge about~$p(y,\theta)$.
-Consequently, we do not need to constitute an actual training dataset!
-
-Second, let~${\cal U}_\star$ denote the expected utility associated with the true distribution~$p(y,\theta)$, considered as a SCL built from a single clue, the full data~$y$.
-As is easy to check, ${\cal U}_\star$ is the conditional Kullback-Leibler divergence of~$p(y|\theta_0)$ from~$p(y|\theta)$:
-$$
-{\cal U}_\star
-= \sum_{j=1}^m \pi(\theta_j) \int p(y|\theta_j) \log \frac{p(y|\theta_j)}{p(y|\theta_0)} dy
-= D[p(y|\theta)\|p(y|\theta_0)]
-.
-$$
-
-Applying in~(\ref{eq:discrim}) the data reduction inequality recalled in Appendix~\ref{sec:reduction_inequality}, we obtain the intuitive result that~${\cal U}(\mathbf{W})$ is upper bounded by the expected utility achievable without data reduction:
-$$
-{\cal U}(\mathbf{W})\leq {\cal U}_\star
-.
-$$
-This means that the full data-generating model maximizes the expected log-SCL over the whole space of SCL functions ({\em i.e.}, SCL functions built from arbitrary clues in arbitrary number and using arbitrary weights).
-Therefore, (\ref{eq:discrim}) qualifies as a measure of goodness of fit to the true likelihood.
-
-Maximizing~(\ref{eq:discrim}) with respect to~$\mathbf{W}$ amounts to optimizing the fit over a subspace of SCL functions spanned by fixed user-specified features.
-Some constraints may be imposed on~$\mathbf{W}$, such as assuming identical columns as in the case of standard composite likelihood, or forcing some elements to zero if some feature likelihood functions are only known on subsets of~$\Theta$.
-We here focus on the case where no constraints are imposed beside, of course, that the columns of~$\mathbf{W}$ lie in the $n$-dimensional simplex.
-Clearly, the optimal weights are then determined independently for each hypothesis (and thus independently from the prior) as any positive weights satisfying, for all~$j\in\{1,\ldots,n\}$:
-\begin{equation}
-\label{eq:optimal_weights}
-w_{ij} = 0 
-\quad {\rm iff}\quad  
-i \not\in \arg\max_{i'\in\{1,\ldots,n\}} u_{i'j},
-\qquad {\rm and} \quad
-\sum_{i=1}^n w_{ij}=1
-,
-\end{equation}
-yielding a sparsity-enforcing rule which assigns non-zero weights only to the clues that achieve maximal KL~utility for a particular hypothesis.
-If there is a unique KL-maximal, or ``most exhaustive'' clue for each hypothesis, the resulting~SCL function essentially boils down to the PDF~projection method \cite{Baggenstoss-03,Minka-04,Baggenstoss-15}.
-More generally, if several clues~$z_i$ maximize $u_{ij}$ for some hypothesis~$\theta_j$, then optimal weights are not unique.
-In such case, it is preferable to assign equal weights to the winning clues in order to ensure that utility is not only large in average, but also stable across examples ({\em i.e.}, low in variance).
-
-Also note that the weight optimality rule~(\ref{eq:optimal_weights}) implies that, for any $\theta_\star \in \Theta$,
-$$
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-\leq
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w}_\star)}{L_c(\theta_0, \mathbf{w}_\star)} \right],
-$$
-where $\mathbf{w}_\star$ is the weight vector associated with~$\theta_\star$, $E$ stands for the expectation with respect to $p(y|\theta_\star)$, and~$\mathbf{w}$ is any weight vector.
-Moreover, the following inequality is shown in Appendix~\ref{sec:asymptotic}:
-$$
-E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-\leq
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-,
-$$
-for any hypothesis~$\theta$ and weighting~$\mathbf{w}$ and, in particular, for the weight vector associated with~$\theta$ via~(\ref{eq:optimal_weights}).
-Therefore, provided that the weight matrix~$\mathbf{W}$ verifies~(\ref{eq:optimal_weights}), we have that:
-$$
-E \left[\log {\cal L}_c(\theta,\mathbf{W}) \right]
-\leq 
-E \left[\log {\cal L}_c(\theta_\star,\mathbf{W}) \right]
-,
-$$
-meaning that the SCL is asymptotically consistent in the sense that the expectation of its logarithm is maximized by the ``true'' parameter value.
-This is a most awaited frequentist property which, as already pointed out, holds with any choice of weights for standard composite likelihood but not, in general, for~SCL.
-
-%Regularized version.
-%$$
-%{\cal U}_\gamma(\mathbf{W})
-%= \sum_{j=1}^m \pi(\theta_j) \sum_{i=1}^n (w_{ij}u_{ij}-\frac{\gamma}{2}w_{ij}^2)
-%$$
-%$\gamma>0$ is a constant.
-Maximizing this leads to independent quadratic programming problems for each~$j$.
-Can be solved using standard numerical methods \cite{Nocedal-99}.
-As $\gamma\to 0$, the optimal weights are uniform among clues satisfying (\ref{eq:optimal_weights}).
-
-\section{Further extensions}
-
-\subsection{Conditional super composite likelihood}
-\label{sec:conditional}
-
-The SCL derivation rests upon the definition of feature-based likelihood as $\ell_i(\theta)=p(z_i|\theta)$.
-As a straightforward  extension, $\ell_i(\theta)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
-\begin{equation}
-\label{eq:cond_feat_lik}
-\ell_i(\theta) = p(z_i|\theta,z^c_i).
-\end{equation}
-
-Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $\theta$, but can provide relevant information when considered jointly with $z_i$, as in the case of regression covariates, for instance.
+Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $x$, but can provide relevant information when considered jointly with $x$, as in the case of regression covariates, for instance.
 Standard composite likelihood (\ref{eq:comp_lik}) then amounts to {\em conditional composite likelihood} \cite{Varin-11}, a more general form of composite likelihood also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74}, which was a major breakthrough in computer vision.
@@
 Likewise, the above derivation remains valid when ``independent'' features are used, and we can thus define a conditional version of SCL by plugging likelihood functions of the form~(\ref{eq:cond_feat_lik}) into (\ref{eq:super}).
-
-\subsection{Nuisance parameters}
-\label{sec:nuisance_params}
-
-Most often, plausible feature-generating models involve unknown quantities of no direct interest.
-Such quantities may be estimated offline in a supervised learning phase if a suitable training dataset is available.
-However, if such ``pre-training'' is not feasible in practice, the feature-based likelihoods depend on an unknown parameter~$\psi$, in addition to the parameter of interest~$\theta$:  $\ell_i(\theta,\psi)=p(z_i|\theta,\psi)$, or $\ell_i(\theta,\psi)=p(z_i|z^c_i,\theta,\psi)$ if ``independent'' features are used.
-We then face two difficulties:
-\begin{itemize}
-\item {\em Parameter integration.} How to make the inference on~$\theta$ independent from~$\psi$ assuming known SCL~weights?
-\item {\em Weight optimization.} How to optimize the SCL~weights under unknown~$\psi$?
-\end{itemize}
-
-We address these two points in the sequel.
-
-%We may then seek a means to eliminate~$\psi$.
-
-\subsubsection{Parameter integration}
-\label{sec:nuisance_integration}
-
-When weighting clues independently from the hypotheses, a joint composite likelihood on both parameters may be derived:
-$$
-L_c(\theta,\psi,\mathbf{w}) = \prod_i \ell_i(\theta,\psi)^{w_i},
-$$
-and further integrated with respect to some prior on the nuisance parameter to yield a function of~$\theta$ only, which we may call the {\em composite evidence}:
-\begin{equation}
-\label{eq:comp_evidence}
-\bar{L}_c(\theta, \mathbf{w}) = \int \pi(\psi) L_c(\theta,\psi,\mathbf{w}) d\psi
-.
-\end{equation}
-This corresponds to using the factor graph represented in Figure~\ref{fig:fgraph5} to approximate the message from the data to~$\theta$, where the factors $\beta_i$ are now defined by $\beta_i(z_i,\theta,\psi)=\ell_i(\theta,\psi)^{w_i}$.
-To compute the associated posterior $p_4(\theta|y)$, $\psi$ is treated as an auxiliary variable to be integrated out, leading to $p_4(\theta|y)\propto \pi(\theta)\bar{L}_c(\theta, \mathbf{w})$.
-
-\begin{figure}[!ht]
-  \begin{center}
-    \includegraphics[width=.6\textwidth]{fgraph5.pdf}
-  \end{center}
-\caption{Extension of the composite likelihood factor graph in Figure~\ref{fig:fgraph2} to account for nuisance parameters.}
-\label{fig:fgraph5}
-\end{figure}
-
-More generally, when weighting clues depending on hypotheses, we may use essentially the same idea as in Section~\ref{sec:super}, {\em i.e.}, replicate subgraphs such as the one depicted in red and magenta in Figure~\ref{fig:fgraph5} in order to approximate the different messages sent from the data to each truncated variable~$t_j$, as defined in~(\ref{eq:binary}).
-Note that, by replicating the same graph structure across variables~$t_j$, the variable~$\psi$ is also replicated, so that the resulting factor graph in Figure~\ref{fig:fgraph6} does not represent a joint distribution of the form $p(y,\theta,\psi)$, but rather one of the form~$p(y,\theta,\psi_1,\ldots,\psi_m)$.
-This makes the approximation more flexible by taking advantage of the fact that a marginal distribution on~$\psi$ is not required.
-
-\begin{figure}[!ht]
-  \begin{center}
-    \includegraphics[width=.7\textwidth]{fgraph6.pdf}
-  \end{center}
-\caption{Extension of the super composite likelihood factor graph in Figure~\ref{fig:fgraph4} to account for nuisance parameters.
-Here, the factors $\beta_{ij}(z_i,t_j,\psi_j)$ are defined by $\beta_{ij}(z_i,1,\psi_j)=\ell_i(\theta_j,\psi_j)^{w_{ij}}$ and $\beta_{ij}(z_i,0,\psi_j)=\ell_i(\theta_0,\psi_j)^{w_{ij}}$.}
-\label{fig:fgraph6}
-\end{figure}
-
-The posterior distribution encoded by this graph (integrated with respect to the replicates $\psi_1,\ldots,\psi_m$) is easily found to be:
-$$
-p_5(\theta|y)
-\propto
-\pi(\theta)
-\left\{
-\begin{array}{lll}
-\displaystyle
-\frac{\bar{L}_c(\theta_j,\mathbf{w}_j)}{\bar{L}_c(\theta_0,\mathbf{w}_j)} & {\rm if} & \theta=\theta_j \quad {\rm with} \quad j\in\{1,\ldots,m\}\\
-1 & {\rm if} & \theta=\theta_0
-\end{array}
-\right.
-,
-$$
-where $\bar{L}_c(.,\mathbf{w}_j)$ is the above-defined composite evidence function~(\ref{eq:comp_evidence}) associated with the weight vector~$\mathbf{w}_j$.
-This justifies defining the {\em super composite evidence} as:
-\begin{equation}
-\label{eq:super_comp_evidence}
-\bar{\cal L}_c(\theta_j,\mathbf{W}) = 
-\left\{
-\begin{array}{lll}
-\displaystyle
-\frac{\bar{L}_c(\theta_j,\mathbf{w}_j)}{\bar{L}_c(\theta_0,\mathbf{w}_j)} & {\rm if} & j\in\{1,\ldots,m\}\\
-1 & {\rm if} & j=0
-\end{array}
-\right.
-.
-\end{equation}
-
-Obvious from~(\ref{eq:super_comp_evidence}) is that the super composite evidence conserves all odds $\theta_j$ {\em vs.}~$\theta_0$ integrated within their respective associated subgraphs, very much like its nuisance parameter-free version~(\ref{eq:super_ratio}).
-The conservation of odds implies  that evaluating the SCL at a particular~$\theta$ is akin to computing a scaled Bayes factor, which may sometimes be efficiently approximated via a maximum likelihood ratio statistic.
-
-%The philosophical question, maybe, is whether it's better to approximate the integrated likelihood, or integrate the approximated likelihood.
-
-\subsubsection{Weight optimization}
-\label{sec:nuisance_weights}
-
-Consistently with the idea of nuisance parameter replication (see Figure~\ref{fig:fgraph5}), we may employ a nested strategy to optimize the SCL~weights in presence of nuisance parameters: for each $j\in\{1,\ldots,m\}$, apply the weight optimization of Section~\ref{sec:weights} to the  subgraph model connecting all clues~$z_i$ with the particular pair $(t_j,\psi_j)$.
-This leads to a set of linear programming problems: for each~$j$, find the weight vector~${\bf w}_j$ on the $n$-dimensional simplex that maximizes:
-$$
-\bar{\cal U}_j({\bf w}_j) 
-= 
-E\left[\log \frac{L_c(\theta_j,\psi_j,{\bf w}_j)}{L_c(\theta_0,\psi_j,{\bf w}_j)}\right],
-$$
-where the expectation is taken with respect to $(y,\psi_j)\sim p(y|\theta_j,\psi)\pi(\psi)$, a computation that only requires knowledge of the feature sampling distributions, $p(z_i|\theta,\psi)$ for $i=1,\ldots,n$, in addition to the prior $\pi(\psi)$.
-
-In this way, every ${\bf w}_j$ is optimal for the posterior of~$(t_j,\psi_j)$ or, more precisely, yields the best approximation to the true joint likelihood $L(\theta, \psi)$ for~$\theta$ restricted to the binary set~$\{\theta_0,\theta_j\}$.
-The ensuing marginal likelihood approximations are therefore as tight as possible for all hypothesis comparisons with~$\theta_0$, as is the goal of SCL~weight optimization.
-
-% En clair, c'est l'idee du college d'experts travaillant sur une hypothese binaire et integrant les parametres de nuisance sans tenir compte de ce que font les autres.
 

### 2018-11-14 papers/composite_inference/draft.tex
Updates composite inference

@@
 \begin{abstract}
-This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes entropy relative to a given prior and subject to multiple mean log-likelihood constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
-We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, and can be powerful in shallow learning problems, {\em e.g.}, in the context of transfer learning.
+This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes conditional entropy relative to a given prior and subject to multiple mean log-likelihood constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
+We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, which can be powerful in shallow learning and transfer learning problems.
 \end{abstract}
@@
 Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
-Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}, the minimum description length principle \cite{Grunwald-07} and, last but not least, discriminative models, which currently dominate the field of artificial intelligence and typically require supervised learning on large datasets -- these include a variety of deep learning \cite{Lecun-15,Goodfellow-16} and shallow learning techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00} and Gaussian processes \cite{Rasmussen-06}.
+Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavily supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16}.
 
-Composite likelihood (see \cite{Varin-11} and the references therein) is a semi-generative approach that extends the familiar notion of likelihood without requiring a full data-generating model.
+A strong limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter elimination because they treat the data and the model parameters as marginally independent, meaning that the data conveys no information about the parameters if the variable of interest is not observed.
+This is illustrated in Figure~\ref{fig:graph_comparison} by the respective oriented graph representations of generative and discriminative models.
+Due to the relative inefficiency of parameter estimation, discriminative models also tend to be outperformed by generative models in supervised learning tasks from small training sets \cite{Ng-01}.
+Pure discriminative models are thus of little use outside the context of big labeled data.
+
+% p(x,y|theta) = p(x|y,theta)p(y|theta)
+
+\begin{figure}[!ht]
+\begin{center}
+\subfigure[Generative model]{\includegraphics[width=.25\textwidth]{generative.pdf}\label{fig:generative}}
+\hspace*{.2\textwidth}
+\subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
+\caption{Belief networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively represent the variable of interest, the data, and the model parameters.
+Note the marginal independence of~$Y$ and~$\theta$ in the discriminative model.}
+\label{fig:graph_comparison}
+\end{center}
+\end{figure}
+
+Composite likelihood (CL, see \cite{Varin-11} and the references therein) is a semi-generative approach to statistical inference that extends the familiar notion of likelihood without requiring a full generative model.
 The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole.
 This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood.
-While maximum composite likelihood does not inherit the general property of maximum likelihood to yield asymptotically minimum-variance estimators, it may offer an excellent trade-off between computational and statistical efficiency.
+While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild regularity conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
-In this note, composite likelihood is interpreted as a probabilistic opinion pool of ``agents'' making use of different pieces of information, or clues, extracted from the input data.
-Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a specific clue in the form of a likelihood of the hidden variables.
-The opinion pool then compuptes a probability distribution on the hidden variables analogous to a Bayesian posterior, thereby extending Bayesian inference to problems where the complete likelihood is undefined.
+In this note, CL is interpreted as a probabilistic opinion pool of ``agents'' making use of several pieces of information, or clues, extracted from the input data.
+Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a single clue in the form of a likelihood function of the hidden variables.
+The agent opinions are then aggregated into a unique probability distribution analogous to a Bayesian posterior.
 
-We further show that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum entropy \cite{Grunwald-04}.
-This perspective entails a method to tune the weights of the different agents from training data as in a classical discriminative learning scenario.
+We further argue that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum relative conditional entropy.
+This perspective entails a method to tune the weights of the different agents from training data as in a typical discriminative learning scenario.
 
@@
 
-This requires a plausible generative model which, for complex data, may be out of reach or involve too many nuisance parameters.
-A natural workaround known as {\em data reduction} is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the more convenient likelihood function:
+This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated.
+A natural workaround known as {\em data reduction} is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
 $$
@@
 
-If we regard the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-We will show in Section~\ref{sec:maxent} that one which is particularly well-suited to our case is the {\em log-linear opinion pool}:
+If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
+We will show in Section~\ref{sec:maxent} that one that is particularly well-suited to our case is the {\em log-linear opinion pool}:
 \begin{equation}
@@
 \end{equation} 
-where $\pi(x)$ is some reference distribution or prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of real-valued weights so that the normalizing factor:
+where $\pi(x)$ is some reference distribution or prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights so that the normalizing factor:
 $$
@@
 $$
-is finite\footnote{It is possible to have the weights depend on~$y$, however there is a strong rational for constant weights as we shall see in the sequel.}.
+is finite, which holds whenever $\lambda\succeq 0$, {\em i.e.} all weights are positive, assuming that the agent opinions are always upper bounded.
+Negative weights should be ruled out as the existence condition $Z_\lambda(y)<\infty$ would then depend on the particular opinions expressed by the agents.
+Also note that, while it is possible to have the weights depend on~$y$, there is a strong rational for constant weights as we shall see in the sequel.
+
+%Strictly positive weights guarantee the so-called 0/1 forcing property, that is, if an hypothesis~$x$ has zero likelihood according to at least one agent, then its consensus probability vanishes too.
+
+% Not clear yet as to what a negative weight could mean!
 
@@
 plays the same role as a traditional likelihood function.
-We recognize in~(\ref{eq:comp_lik}) a general expression known as {\em marginal composite likelihood} \cite{Varin-11}.
-It shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, sometimes called {\em na\"ive Bayes} likelihood in the literature \cite{Ng-01}, which corresponds to the special case of all unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
-The general computational advantage of composite likelihood over the true likelihood is that it only requires to evaluate the marginal feature distributions rather than the joint distribution of all features.
+This expression happens to be known in statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
+See Appendix~\ref{sec:conditional} for the slightly more general form called {\em conditional composite likelihood}, which can be derived in the same way.
 
-\section{The problem of tuning composite likelihood weights}
+From~(\ref{eq:comp_lik}), we see that CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
+The clear computational advantage of CL over the true likelihood is that it only requires to evaluate the marginal feature distributions rather than the joint distribution of all features.
 
-There exist few general guidelines to choose the composite likelihood weights~$\lambda$ in~(\ref{eq:comp_lik}).
-Common practice is to impose nonnegative weights \cite{Varin-11} although this is not strictly required from the opinion aggregation viewpoint (the normalizing factor $Z_\lambda(y)$ in~(\ref{eq:log_pool}) only needs be finite for any~$y$ for the log-linear pool to yield a proper conditional distribution).
+\section{Classical tuning of composite likelihood weights}
 
-In cases where the features can be considered to be {\em exchangeable}, it is natural to pick uniform weights.
-The composite likelihood function is then a scaled version of the {\em na\"ive Bayes} likelihood, the common weight value being irrelevant to the location of the composite likelihood maximizer.
-It can be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to reconcile the frequentist and Bayesian notions of uncertainty.
+When the features can be considered {\em exchangeable}, it is natural to choose uniform CL weights.
+The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
+The weight may be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match the frequentist and composite Bayesian notions of uncertainty.
 
-Another common recommandation for the composite likelihood weights is to sum up to one.
+Ignoring such a goal, another frequent recommandation for CL weights is to sum up to one.
+This can be motivated in various ways.
+It turns out that the only pooling operator that does not explicitly depend on~$x$ and preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}.
+External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.
+Another appealing property of log-linear pooling with unit sum weights is to minimize the average Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
+A maximum relative entropy property is also given in \cite{Wang-14}, Theorem~3.
 
-Going back to the log-linear pool underlying composite likelihood, Genest {\em et al} \cite{Genest-86b} showed that the only pooling operator that does not explicitly depend on~$x$ and preserves a property known as {\em external Bayesianity} is the log-linear opinion pool~(\ref{eq:log_pool}) with the additional constraint that the weights  sum up to one.
-External Bayesianity essentially means that it should not matter whether a prior on~$X$ is incorporated before or after combining opinions, provided that all agents agree on the same prior.
-Garg {\em et al} \cite{Garg-04} showed that the log-linear pool with unit weight sum also minimizes the average Kullback-Leibler (KL) divergence to the agent opinions.
-The MaxEnt argument was already given by \cite{Wang-14}.
+Unit sum weights, however, correspond to the extreme situation where the features are assumed to be maximally redundant, but is clearly ineffective for independent or weakly correlated features, which require weights close to one for the CL to closely approximate the true likelihood.
+In this case, the CL may be much flatter than it should, leading to severly overestimated credibility sets.
+
+{\color{red} Why is it important to fine-tune the weights?}
+We therefore need a method to tune weights that can automatically adjust to the level of statistical dependence beween features.
 
@@
 This results in increasing their weights in the log-linear pool.}.
-
-The log-linear pool has, in particular, the 0/1 forcing property, meaning that if any agent assigns zero likelihood to a value~$x$, then $x$ has zero probability in the pool.
-
-Unit sum weights = crap.
 
@@
 
-The MaxEnt argument was already given by \cite{Wang-14}.
+A cheap MaxEnt argument was already given by \cite{Wang-14}.
+$I$-projection onto a set of predictive distributions compatible with mean log-likelihood contraints.
+If I know the generative distribution, I know the expected log-likelihood.
+
+Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. but is there some intuition behind that?
+
+BTW, we can work with inequality constraints to force $\lambda\geq 0$:
+$$
+E[\log \ell_i(x)] \geq \bar{\ell}_i
+$$
+
+What does the constraint mean? It means that a model is admissible if it compresses the data at least as well as all my agents.
+
+Game theoretic interpretation \cite{Grunwald-04}.
 
@@
 % The sum of the powers is constrained to be equal to one, even though
-% I do not understand why the dimensions of the projections play no
+% I do not understand why the dimensions of the projections play nog
 % role in this constraint. Simplicity is advanced as an argument,
 % which sounds rather weak…
-This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $x$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.
-
-\section{Composite likelihood as message approximation}
-\label{sec:message}
-
-Composite likelihood may also be understood as a means to approximate the ``true'' likelihood function\footnote{By {\em true likelihood}, we mean the likelihood corresponding to a specified yet intractable generative model.
-In practice, such a model is obviously not required.} or, in the language of graphical models, the {\em message} that the data sends to the latent variable~$x$.
-Several ``clues'' are sent to~$x$ via different data features, and then integrated assuming {\em non-coalescent} emitting sources, meaning that statistical dependences between clues are treated as unknown, although not ignored.
-The whole idea is depicted by the factor graph\footnote{See \cite{Bishop-06} for a didactic introduction to factor graphs.} in Figure~\ref{fig:fgraph2}, where the latent variable is connected to multiple factors involving single clues (rather than a single factor involving multiple clues), thereby enabling efficient computations.
-
-%A simplified message is computed by extracting arbitrary features from the data and treating them as {\em non-coalescent} clues in the sense that their statistical dependences are not modeled (although not ignored).
-
-\begin{figure}[!ht]
-\begin{center}
-\subfigure[Generative model]{\includegraphics[width=.35\textwidth]{fgraph1.pdf}\label{fig:fgraph1}}
-\subfigure[Composite likelihood model]{\includegraphics[width=.6\textwidth]{fgraph2.pdf}\label{fig:fgraph2}}
-\caption{Factor graph representations of a generative model (a) and its approximation using composite likelihood (b).
-In (a), the factor connecting the data~$y$ and the latent variable~$x$ is the true likelihood, $p(y|x)$.
-In (b), factors connecting~$y$ and the clues $z_1,z_2,\ldots$ represent feature extractions, $\alpha_i(y,z_i)=\delta[z_i-f_i(y)]$, while factors connecting the clues and~$x$ are scaled single-feature likelihoods, $\beta_i(z_i,x)=p(z_i|x)^{w_i}$.
-In both graphs, the factor shown in black is the prior $\pi(x)$.}
-\label{fig:fgraph}
-\end{center}
-\end{figure}
-
-This approximation scheme happens to be optimal in an information-theoretic sense.
-As noted in \cite{Garg-04}, the log-linear pool minimizes the average Kullback-Leibler (KL) divergence to the probabilistic opinions:
-\begin{equation}
-\label{eq:avg_kl}
-p_\star = \arg\min_p \sum_{i=1}^n w_i D(p\|p_i),
-\end{equation}
-where $p_i(\theta) \propto \pi(\theta)\ell_i(\theta)$ is the posterior distribution of agent~$i$.
-Note that, even though the weights are arbitrary in~(\ref{eq:avg_kl}), the optimal solution normalizes them to unit sum.
-Composite likelihood thus arises as the best possible consensus among agents according to a natural criterion, in addition to satisfying the external Bayesianity axiom.
-
-If one interprets the average divergence from agent opinions~(\ref{eq:avg_kl}) as a proxy for the divergence from ``the truth'', $p(x|y)$, composite likelihood can be seen as a variational approximation to the true likelihood.
-This corresponds to the intuitive notion that a consensus among sufficiently many experts should yield a reasonable guess.
-An essential difference with usual approximate inference methods \cite{Bishop-06,Minka-05} is that the likelihood function does not need to be computable owing to the use of a proxy.
-However, as we shall see in Section~\ref{sec:weights}, knowledge of the feature sampling distributions can be further exploited to optimize the likelihood approximation with respect to the composite likelihood weights.
-
-\section{Conditional composite likelihood}
-\label{sec:conditional}
-
-The SCL derivation rests upon the definition of feature-based likelihood as $\ell_i(x)=p(z_i|x)$.
-As a straightforward  extension, $\ell_i(x)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
-\begin{equation}
-\label{eq:cond_feat_lik}
-\ell_i(x) = p(z_i|x,z^c_i).
-\end{equation}
-
-Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $x$, but can provide relevant information when considered jointly with $x$, as in the case of regression covariates, for instance.
-Standard composite likelihood (\ref{eq:comp_lik}) then amounts to {\em conditional composite likelihood} \cite{Varin-11}, a more general form of composite likelihood also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74}, which was a major breakthrough in computer vision.
-
-Likewise, the above derivation remains valid when ``independent'' features are used, and we can thus define a conditional version of SCL by plugging likelihood functions of the form~(\ref{eq:cond_feat_lik}) into (\ref{eq:super}).
+%This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $x$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.
 
@@
 
-Composite likelihood is a relatively recent concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
-In this paper, we have shown (to our best knowledge, for the first time) a deep connection between composite likelihood and  probabilistic opinion pooling, thereby establishing composite likelihood as a class of {\em discriminative models}\footnote{We here define a discriminative model as a model of which the parameters describe the conditional distribution of an unobserved variable of interest given an observable variable, as opposed to a {\em generative} model, which involves parameters encoding for the conditional distribution of an observable given an unobserved variable.} for statistical inference and machine learning.
-This connection is possible under the mild restriction that composite likelihood weights are chosen to sum up to one.
+{\color{red}Product of fucking experts \cite{Hinton-02}.
+Log-linear pool to build a generative model (assuming in fact independence between experts).
+Better seen as a ``log-mixture model''.
+Sounds like each expert is associated with a class, so it's the same as a RBM.
+Contrastive learning approximates ML parameter estimation.
+The experts are learning jointly.
+It's just a generative model.}
 
-\subsection{From na\"ive Bayes to super composite likelihood}
+{\color{red}Two-step training: generative phase then discriminative phase. Why is it cool?}
 
-In the probabilistic opinion pooling perspective, composite likelihood is essentially a reinterpretation and a generalization of the {\em na\"ive Bayes} paradigm that relaxes the associated ``na\"ive'' mutual feature independence assumption.
-In particular, when all features are given equal weight, the composite likelihood is the na\"ive Bayes likelihood raised to the power~$1/n$, where~$n$ is the number of selected features, yielding for instance the same maximum a posteriori (MAP) estimator if a flat prior is used.
-However, beside providing a simple justification to the wide use of na\"ive Bayes MAP algorithms,  composite likelihood using unit sum weights also entails a conservative rescaling of credibility sets derived from na\"ive Bayes, which gets more drastic as the number of features increases.
-This comes as a consequence of relaxing feature independence.
+{\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
 
-We further argued that this rescaling may, to some extent, be unduly conservative in multiclass problems if features are weighted uniformly over the space of unobserved labels, leading us to propose the more general concept of super composite likelihood (SCL).
-SCL essentially approximates likelihood ratios relative to a fixed reference hypothesis using locally weighted composite likelihood functions.
-Owing to weight adaptability, SCL describes a more general class of discriminative models than standard composite likelihood.
+{\color{red}Boosting.}
 
-The idea can also be understood in terms of approximating the factor graph in Figure~\ref{fig:fgraph3}, which encodes the true but intractable posterior distribution, by another factor graph of the type depicted in Figure~\ref{fig:fgraph4}.
-This substitution results in breaking into pieces both the observed and unobserved variable spaces, and assembling a series of concise {\em messages} passing from the former to the latter.
-Note that this is a pretty unusual case of approximate inference in factor graphs where the factors to be approximated are unknown (or need not be known).
+CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
+In this paper, we have shown (to our best knowledge, for the first time) a deep connection between CL and  probabilistic opinion pooling, thereby establishing CL as a class of discriminative models.
 
-\subsection{Super composite likelihood training}
-
-Any SCL~model is fully determined by the marginal generative distributions of some pre-specified features, which may rely on a moderate number of parameters if low-dimensional features are chosen.
+Any CL~model is fully determined by the marginal generative distributions of some pre-specified features, which may rely on a moderate number of parameters if low-dimensional features are chosen.
 There are two approaches to deal with these parameters.
@@
 
-\begin{figure}[!ht]
-\begin{center}
-\subfigure[Generative model]{\includegraphics[width=.25\textwidth]{dgraph1.pdf}\label{fig:dgraph1}}
-\hspace*{.05\textwidth}
-\subfigure[Classical discriminative model]{\includegraphics[width=.25\textwidth]{dgraph2.pdf}\label{fig:dgraph2}}
-\hspace*{.05\textwidth}
-\subfigure[Composite likelihood model]{\includegraphics[width=.25\textwidth]{dgraph3.pdf}\label{fig:dgraph3}}
-\caption{Belief networks representing, respectively: (a) a generative model, (b) a classical discriminative model, and (c) a composite likelihood model.
-Note the marginal independence between the data~$y$ and the nuisance parameter~$\psi$ in (b).}
-\label{fig:dgraph}
-\end{center}
-\end{figure}
-
 %%, hence alleviating the need for heavily supervised model training
@@
 This stems from the fact that, unlike classical discriminative models, SCL~exploits direct information conveyed by the data about both the parameters of interest and the nuisance parameters, reflecting the generative nature of the underlying feature distribution models.
-This essential difference is illustrated in Figure~\ref{fig:dgraph}: the factor graph model underlying composite likelihood in its parametric version (see Figure~\ref{fig:fgraph5}) is not reducible to a belief network of the type represented in Figure~\ref{fig:dgraph2}, in which the data and the nuisance parameter are marginally independent.
-In contrast, the data is informative about the nuisance parameter in a composite likelihood model, as shown in Figure~\ref{fig:dgraph3}.
+This essential difference is illustrated in Figure~\ref{fig:dgraph}: the factor graph model underlying CL in its parametric version (see Figure~\ref{fig:fgraph5}) is not reducible to a belief network of the type represented in Figure~\ref{fig:dgraph2}, in which the data and the nuisance parameter are marginally independent.
+In contrast, the data is informative about the nuisance parameter in a CL model, as shown in Figure~\ref{fig:dgraph3}.
 
-\section{Conclusion}
-\label{sec:conclusion}
+In summary, CL has the potential to yield weakly supervised or unsupervised Bayesian-like inference procedures depending on the particular task at hand.
+This property reflects the encoding of statistical relationships between the data and {\em all} unknown parameters.
+CL thus appears as a trade-off between generative models, which are optimal for unsupervised learning but possibly intractable, and traditional discriminative models (logistic regression, Gaussian processes \cite{Rasmussen-06}, maximum entropy models \cite{BergerA-96}, etc.), which are inherently supervised.
+CL models are discriminative models assembled from atomic generative models and, from this point of view, may be considered as {\em semi-generative} models.
 
-In summary, (super) composite likelihood has the potential to yield weakly supervised or unsupervised Bayesian-like inference procedures depending on the particular task at hand.
-This property reflects the encoding of statistical relationships between the data and {\em all} unknown parameters.
-Composite likelihood thus appears as a trade-off between generative models, which are optimal for unsupervised learning but possibly intractable, and traditional discriminative models (logistic regression, Gaussian processes \cite{Rasmussen-06}, maximum entropy models \cite{BergerA-96}, etc.), which are inherently supervised.
-Composite likelihood models are discriminative models assembled from atomic generative models and, from this point of view, may be considered as {\em semi-generative} models.
-
-%composite likelihood may be considered as a {\em semi-generative} model: a discriminative model assembled from partial generative models.
+%CL may be considered as a {\em semi-generative} model: a discriminative model assembled from partial generative models.
 
@@
 
-\section{Data reduction inequality}
-\label{sec:reduction_inequality}
+\section{Conditional composite likelihood}
+\label{sec:conditional}
 
-A fundamental property of the Kullback-Leibler divergence is that it decreases under feature extraction, {\em i.e.}, application of a deterministic transformation.
-This comes as a consequence of the logarithmic sum inequality \cite{Cover-06} (and also follows as a straightforward corollary of the PDF~projection theorem \cite{Minka-04,Baggenstoss-15}).
+As a straightforward  extension of marginal CL, each feature-based likelihood $\ell_i(x)=p(z_i|x)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
+\begin{equation}
+\label{eq:cond_feat_lik}
+\ell_i(x) = p(z_i|x,z^c_i).
+\end{equation}
 
-The proof is elementary and can be sketched as follows.
-Let $Z=f(Y)$ some feature extracted from a variable~$Y\sim p(y)$.
-Given an arbitrary distribution~$\pi(y)$, let  $\tilde{p}(z)$ and $\tilde{\pi}(z)$ the distributions induced on~$Z$ by $p(y)$ and $\pi(y)$, respectively.
-For any potential value~$z$ of~$Z$, consider the level set: $\Gamma(z)=\{y, f(y)=z\}$.
-Under conditions of existence, we can partition the integral involved in the KL~divergence $D(p\|\pi)$ using these level sets:
+Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $x$, but can provide relevant information when considered jointly with $x$, as in the case of regression covariates, for instance.
+Equation~(\ref{eq:comp_lik}) then amounts to conditional CL \cite{Varin-11}, a more general form of CL also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74} developed for image segmentation.
+
+\section{Minimally discriminative model}
+
+Let $\pi(x)$ some reference distribution that represents full
+uncertainty about $X$. We wish to select the joint distribution
+$p(x,y)$ that minimizes:
 $$
-\int p(y) \log \frac{p(y)}{\pi(y)} dy
+I(p) = \int p(x,y) \log \frac{p(x|y)}{\pi(x)} dy,
+$$ subject to feature mean-value constraints of the form:
+$$
+\int p(x,y) f(x,y) dx dy = \mu.
+$$
+
+There are two special cases of this problem in the literature.  On the
+one hand, the {\em maximum entropy classifier} \cite{BergerA-96}
+incorporates the constraint that $p(y)$ be known, however we will see
+that this is not necessary. The {\em minimally informative likelihood}
+method \cite{Yuan-99b,Yuan-99}, on the other hand, imposes that
+$p(x)=\pi(x)$, hence assuming the form $p(x,y)=\pi(x)p(y|x)$. We won't
+make such assumptions here and will consider the case where $\mu$ is
+completely unknown.
+
+The above problem is seen to be equivalent to minimizing the auxiliary
+objective function:
+$$
+I(p,m) 
+= \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy,
+$$ over $p$ and $m$, subject to the same constraints on $p$. We note
+that $I(p,m)=I(p)+D(p_y\|m)$, showing that the auxiliary function
+essentially adds a penalty term to the actual objective in order to
+force $p(y)$ close to $m(y)$.
+
+Minimizing $I(p,m)$ along both $p$ and $m$ is basically a minimum KL
+divergence problem between two convex distribution spaces, so there
+must be a unique solution {\bf -- to be checked in
+  \cite{Cover-91}}. It is similar to the rate-distortion problem in
+information theory, which may be solved by alternate minimization,
+yielding a Blahut-Arimoto algorithm.
+\begin{itemize}
+\item {\em Let's call it A-step}. Optimize $p(x,y)$ at fixed $m$
+  s.t. constraint:
+$$
+\exists \lambda, \qquad
+p_{\lambda,m}(x,y) = \frac{1}{Z(\lambda, m)} \pi(x) m(y) e^{-\lambda^\top t(x,y)}
+$$
+The actual $\lambda$ is found my maximizing the dual function
+$\psi(\lambda,m)$, see below.
+\item {\em Let's call it B-step}. Optimize $m(y)$ at fixed $p(x,y)$:
+$$
+m(y) = \int p(x,y) dx
+$$
+\end{itemize}
+
+Upon convergence, the algorithm outputs both joint and marginal
+distributions $p_{\star}(x,y)$ and $m_{\star}(y)$ that both depend on
+$\mu$. By construction, we have that $p_{\star}(y) = m_{\star}(y)$,
+therefore $p_\star(x|y)=p_{\star}(x,y)/m_{\star}(y)$.
+
+\section{Comparison with maximum entropy classifier}
+
+Lagrangian...
+$$
+{\cal L}(p,m,\lambda)
 = 
-\int \left( \int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy \right) dz\\
+\int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy
++
+\lambda^\top \left( 
+\int p(x,y) f(x,y) dydy - \mu 
+\right)
+$$ where $\lambda$ represents a vector-valued Lagrange multiplier. At
+fixed $m$, the solution has the form:
+$$
+p_{\lambda,m}(x,y) = \frac{1}{Z(\lambda,m)}
+\pi(x) m(y) e^{-\lambda^\top f(x,y)} 
+$$
+
+Note that this implies that the conditional distribution is
+independent of $m$ once $\lambda$ is determined,
+$$
+p_\lambda(x|y) = \frac{1}{z(\lambda,y)} \pi(x) e^{-\lambda^\top f(x,y)} 
+$$
+
+Also, the marginal is a modulation of $m(y)$:
+$$
+p_{\lambda,m}(y) = \frac{z(\lambda,y)}{Z(\lambda,m)} m(y)
+$$
+
+Dual function at fixed $m$:
+$$
+\psi(\lambda,m) 
+\equiv \min_p {\cal L}(p,m,\lambda)
+= 
+- \log Z(\lambda,m) - \lambda^\top \mu
 .
@@
 
-Applying the logarithmic sum inequality \cite{Cover-06} to each integral over~$\Gamma(z)$, we then readily get:
+An alternative expression is:
 $$
-\int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy
-\geq 
-\tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)}
-,
+\psi(\lambda, m)
+= 
+\int h(x,y) 
+\log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(y)} dxdy,
+$$ where $h(x,y)$ is any distribution statisfying the
+constraints. This shows that maximizing the dual function at fixed $m$
+is essentially the same as minimizing the KL divergence
+$D(h\|p_{\lambda,m})$ over $\lambda$, in other words fitting $h(x,y)$
+by some distribution of the form $p_{\lambda,m}$. The fact that the
+result does not depend on the particular $h(x,y)$ that is chosen, as
+long as it satifies the constraint, is a general property of
+exponential families.
+
+We also have that the dual function associated with $I(p)$ reads:
+\begin{eqnarray*}
+\psi(\lambda) 
+ & = & \min_m \psi(\lambda, m)\\
+ & = & \int h(x,y) \log \frac{p_{\lambda}(x|y)}{\pi(x)} dxdy\\
+ & = & -\int h(y) \log z(\lambda,y) dy - \lambda^\top \mu
+\end{eqnarray*}
+
+But, wait, that's exactly what we get in the maximum entropy
+classifier! So, at the end of the day, we simply got an alternative
+method to learn $\lambda$ in the maximum entropy classifier, i.e. the
+Blahut-Arimoto algorithm as opposed to a brute-force maximization of
+$\psi(\lambda)$. Both methods will converge to the same $\lambda$...
+
+What it essentially means is that the optimal $\lambda$ is insensitive
+to $m(y)$ as long as $m(y)$ is compliant in the sense that:
 $$
-owing to the fact that $\displaystyle \int_{\Gamma(z)} p(y) dy = \tilde{p}(z)$ by definition. Therefore,
+m(y) = \int h(x,y) dx,
+$$ for some distribution $h(x,y)$ statisfying the constraint. This is
+true for the optimal $m_\star(y)$ output by the BA algorithm, but
+also, for instance, for the empirical distribution of observations in
+a training dataset used to estimate $\mu$, as proposed in
+\cite{BergerA-96}. The only special property of $m_\star(y)$ is to
+yield the full Bayesian model $p_\star(x,y)=p_\star(x|y)m_\star(y)$
+that minimizes the discrimination information. We may not care too
+much about that in practice since we will only use $p_\star(x|y)$.
+
+\section{Comparison with minimally informative likelihood}
+
+Yuan \cite{Yuan-99} considered the situation where we add the
+constraint that $p(x)=\pi(x)$ to the minimum discrimination inference
+problem. Unknown is then the conditional distribution
+$p(y|x)$. Lagrangian...
+\begin{eqnarray*}
+{\cal L}(p,m,\lambda)
+ & = & 
+\int \pi(x)p(y|x) \log \frac{p(y|x)}{m(y)} dxdy
++
+\lambda^\top \left( 
+\int \pi(x)p(y|x) f(x,y) dydy - \mu 
+\right) \\
+ & = & 
+\int \pi(x)
+\left( 
+\int
+p(y|x) \log \frac{p(y|x)}{m(y)} dy
++
+\lambda^\top 
+\int p(y|x) f(x,y) dy
+- \mu 
+\right)
+dx 
+\end{eqnarray*}
+
+The derivative is given by,
 $$
-\int \tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)} dz
-\leq 
-\int p(y) \log \frac{p(y)}{\pi(y)} dy
-,
-$$
-or, in a more compact way, $D(\tilde{p}\|\tilde{\pi})\leq D(p\|\pi)$.
-The case of equality occurs if, and only if, $p(y)/\pi(y)=\tilde{p}(z)/\tilde{\pi}(z)$, in other words, if~$z$ is a sufficient statistic for the ``alternative'' hypothesis~$H_1:y\sim p(y)$ versus the ``null'' hypothesis~$H_0:y\sim\pi(y)$.
-
-Since it is a well-known fact that $D(\tilde{p}\|\tilde{\pi})\geq 0$ as any KL~divergence, we conclude that:
-\begin{equation}
-\label{eq:reduction_inequality}
-0 
-\leq D(\tilde{p}\|\tilde{\pi}) 
-\leq D(p\|\pi)
-.
-\end{equation}
-
-\section{Basic asymptotic properties of composite likelihood}
-\label{sec:asymptotic}
-
-It follows from the double inequality~(\ref{eq:reduction_inequality}) that, for any extracted feature~$z_i$ and for any hypothesis~$\theta$,
-$$
-0 \leq
-E\left[
-\log \frac{p(z_i|\theta_\star)}{p(z_i|\theta)}
-\right]
-\leq
-E\left[
-\log \frac{p(y|\theta_\star)}{p(y|\theta)}
+\frac{\partial\cal L}{\partial p(y|x)}
+= 
+\pi(x)\left[ 
+1 + \log \frac{p(y|x)}{m(y)} 
++ \lambda^\top f(x,y)
 \right],
 $$
-where the expectation is taken with respect to the true distribution $p(y|\theta_\star)$.
-Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any standard composite likelihood function (assuming unit sum positive weights):
-\begin{equation}
-\label{eq:variation_bound}
-0 \leq
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta, \mathbf{w})} \right]
-\leq 
-E\left[ \log \frac{L(\theta_\star)}{L(\theta)} \right]
-.
-\end{equation}
-
-This implies two asymptotic properties of standard composite likelihood: 
-\begin{itemize}
-\item {\em Consistency.} The expected log-composite likelihood is maximized by~$\theta_\star$, the true value of~$\theta$.
-\item {\em Conservativeness.} The expected log-composite likelihood function is ``flatter'' than the true expected log-likelihood.
-In other words, composite likelihood ratios $L_c(\theta, \mathbf{w})/L_c(\theta_\star, \mathbf{w})$ relative to~$\theta_\star$ tend to be larger than the corresponding true likelihood ratios.
-\end{itemize}
-
-These properties do not necessarily extend to the more general case of SCL. We may state a weaker consistency property by considering the upper envelope of expected log-composite likelihood ratios:
+hence the optimal model at fixed $m$ has the form:
 $$
-M(\theta) = \max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
-$$
-where ${\cal W}$ is the $n$-dimensional simplex. Clearly, the expected logarithm of the SCL is upper bounded by~$M(\theta)$:
-$$
-E[\log {\cal L}_c(\theta, \mathbf{W})] \leq M(\theta)
-.
+p_{\lambda,m}(y|x) = \frac{1}{Z(\lambda,m,x)} m(y) e^{-\lambda^\top f(x,y)} 
 $$
 
-Moreover, $M(\theta)$ is maximized by $\theta_\star$ since (\ref{eq:variation_bound}) implies that, for any $(\theta,\theta_0)$,
+Note that the induced posterior distribution $p_{\lambda,m}(x|y)$ has
+a different form from above unless the normalizing factor
+$Z(\lambda,m,x)$ turns out independent from $x$. If this is not the
+case, we no longer have the property that $p_{\lambda,m}(x|y)$ is
+independent from $m$ given $\lambda$.
+
+Dual function...
 $$
-E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-\leq
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
-$$
-which remains true when taking the maximum over the weights in both sides: 
-$$
-M(\theta) = 
-\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-\leq
-\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-= M(\theta_\star)
-.
+\psi(\lambda,m) 
+=
+- \int \pi(x) \log Z(\lambda, m, x) dx
+- \lambda^\top \mu
 $$
 
-Therefore, maximizing SCL corresponds to maximizing a lower bound on~$M(\theta)$, which can be considered as an objective function since it is maximized by~$\theta_\star$ (like the expected log-likelihood).
-Lower bound maximization guarantees a certain objective value, however not the maximum, hence asymptotic consistency may not hold.
+Equivalently, for any distribution under the form $\pi(x)h(y|x)$ that
+satisfies the moment constraint, we have:
+$$
+\psi(\lambda,m) 
+=
+\int \pi(x) h(y|x) \log \frac{p_{\lambda,m}(y|x)}{m(y)} dxdy
+$$
+
+Now, the real question is why should we constrain $p(x)$, which boils
+down to a Bayesian prior in this context, to be the same as our
+reference $\pi(x)$? It only makes sense if we want our inference to
+stick to a generative modeling paradigm... but haven't we already
+given up on that? Therefore, unless we find a good reason not to, we
+won't impose the $p(x)=\pi(x)$ constraint, thereby allowing for a
+discrepancy between the reference and the prior.
+
+\section{Discriminative vs. semi-discriminative}
+
+Let's go back to the equivalence we found between our approach and the
+maximum entropy classifier (MCE) \cite{BergerA-96}. We have said that
+the former is essentially a re-formulation of MCE.
+
+But the re-formulation also conveys a generalization of MCE if,
+instead of letting $m$ being an arbitrary distribution, we restrict
+its search space to some set of acceptable reference
+distributions. Would such a strategy be useful?
+
+Recall that the method selects the joint distribution that minimizes
+discriminative information, as defined by $I(p)$:
+$$
+I(p) 
+= \int p(x,y)\log\frac{p(x|y)}{\pi(x)} dxdy
+= E_Y[D(p_{x|y}\|\pi)]
+$$
+
+The latter characterization reminds us that discriminative information
+is defined in an average sense. The corresponding posterior $p(x|y)$
+may not be conservative for some $y$, in particular those that are
+unlikely under $p(y)$. It would be a problem if that's the case for
+the particular data at hand. For that to happen rarely, the ideal
+$p(y)$ would be the ``true'' marginal distribution of $Y$.
+
+However, for vague mean value constraints, nothing may prevent the
+optimal $m_\star(y)$ from departing significantly from that ideal
+distribution. In fact, we can show that $m_\star(y)$ only has mass at
+$y$ values that maximize $z(\lambda_\star,y)$, and is therefore likely
+sparse. This is because $m_\star$ minimizes $\psi(\lambda_\star,m)$,
+which is equivalent to maximizing $Z(\lambda_\star,m)$ and we have:
+$$
+Z(\lambda, m) = \int m(y) z(\lambda, y) dy.
+$$
+
+In other words, the solution to our problem may be singular! It does
+not hurt since, as discussed above, one may alternatively fix $m(y)$
+to some pre-defined distribution... but this, in practice, restricts
+the method to supervised learning situations.
+
+We may not face this issue with the MIL method, which further imposes
+the prior on $X$, so all the information from the constraints goes
+into specifying a possibly reasonable generative model.
 

### 2018-11-15 papers/composite_inference/draft.tex
Night updates on CBI

@@
 Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
-Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavily supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16}.
+Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16}.
 
@@
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective oriented graph representations of generative and discriminative models.
-Due to the relative inefficiency of parameter estimation, discriminative models also tend to be outperformed by generative models in supervised learning tasks from small training sets \cite{Ng-01}.
+For the same reason, supervised learning in discriminative models is less efficient than in comparable generative models, explaining why the latter tend to outperform the former in small training sets \cite{Ng-01}.
 Pure discriminative models are thus of little use outside the context of big labeled data.
@@
 \subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
-\caption{Belief networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively represent the variable of interest, the data, and the model parameters.
+\caption{Belief networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the variable of interest, the data, and the model parameters.
 Note the marginal independence of~$Y$ and~$\theta$ in the discriminative model.}
@@
 The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole.
-This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood.
-While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild regularity conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
+This may be viewed as a {\em divide and conquer} method to approximate the true but intractable likelihood.
+While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
-In this note, CL is interpreted as a probabilistic opinion pool of ``agents'' making use of several pieces of information, or clues, extracted from the input data.
+In this note, CL is interpreted as a probabilistic opinion pool of ``agents'' making use of different pieces of information, or clues, extracted from the input data.
 Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a single clue in the form of a likelihood function of the hidden variables.
@@
 \label{eq:log_pool}
-p_\lambda(x|y) = \frac{1}{Z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
+p_\lambda(x|y) = \frac{1}{z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
 \end{equation} 
@@
 $$
-Z_\lambda(y) = \int \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i} dx
+z_\lambda(y) = \int \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i} dx
 $$
 is finite, which holds whenever $\lambda\succeq 0$, {\em i.e.} all weights are positive, assuming that the agent opinions are always upper bounded.
-Negative weights should be ruled out as the existence condition $Z_\lambda(y)<\infty$ would then depend on the particular opinions expressed by the agents.
+Negative weights should be ruled out as the existence condition $z_\lambda(y)<\infty$ would then depend on the particular opinions expressed by the agents.
 Also note that, while it is possible to have the weights depend on~$y$, there is a strong rational for constant weights as we shall see in the sequel.
@@
 
-\section{Classical tuning of composite likelihood weights}
+\section{Tuning composite likelihood weights}
 
-When the features can be considered {\em exchangeable}, it is natural to choose uniform CL weights.
-The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
+When the features can be considered exchangeable, it is natural to choose uniform CL weights.
+The CL is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
 The weight may be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match the frequentist and composite Bayesian notions of uncertainty.
@@
 
-{\color{red}Product of fucking experts \cite{Hinton-02}.
-Log-linear pool to build a generative model (assuming in fact independence between experts).
-Better seen as a ``log-mixture model''.
-Sounds like each expert is associated with a class, so it's the same as a RBM.
-Contrastive learning approximates ML parameter estimation.
-The experts are learning jointly.
-It's just a generative model.}
+CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
+In this note, we have shown a deep connection between CL and  probabilistic inference, thereby establishing CL as a class of discriminative models.
+The fact that it is built from a set of marginal feature-specific generative distributions makes it a semi-generative, semi-discriminative approach.
 
-{\color{red}Two-step training: generative phase then discriminative phase. Why is it cool?}
+Two-step training.
+Why potentially more efficient than pure discriminative training: because generative training is always more efficient than discriminative training if models are comparable.
+So the key point is that ``we have less parameters in the discriminant phase".
 
-{\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
-
-{\color{red}Boosting.}
-
-CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
-In this paper, we have shown (to our best knowledge, for the first time) a deep connection between CL and  probabilistic opinion pooling, thereby establishing CL as a class of discriminative models.
-
-Any CL~model is fully determined by the marginal generative distributions of some pre-specified features, which may rely on a moderate number of parameters if low-dimensional features are chosen.
-There are two approaches to deal with these parameters.
-
-One is to estimate them beforehand by supervised learning, if an adequate training dataset is available.
+First training phase.
+Supervised or unsupervised.
 This could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimizes parameters for generation of observable features.
-An important difference, however, is that contrastive pre-training is unsupervised.
-On the other hand, SCL pre-training relies on weaker assumptions as it does not assume conditional feature independence unlike RBMs.
-Once pre-training is complete, the SCL~weights may be tuned by maximum~SCL, as shown in Section~\ref{sec:weights}: this second training stage is, again, supervised in essence, but does not require additional training examples since it is fully determined by the feature distributions learned in the pre-training step.
+The main difference is that the two learning phases of CL optimize disjoint subsets of parameters unlike RBMs.
+CL pre-training relies on weaker assumptions as it does not assume conditional feature independence unlike RBMs.
+Once pre-training is complete, the CL~weights may be tuned: this second training stage is, again, supervised in essence, but does not require additional training examples since it is fully determined by the feature distributions learned in the pre-training step.
 
@@
 
-The other approach to deal with feature distribution parameters is to consider them as nuisance parameters, as proposed in Section~\ref{sec:nuisance_params}, thereby avoiding pre-training.
-The method then becomes completely unsupervised.
-The weights can be tuned by maximizing the log-SCL integrated over the prior on nuisance parameters (see Section~\ref{sec:nuisance_weights}), which requires no training dataset whatsoever, and the nuisance parameters can be eliminated by substituting the~SCL function with a {\em super composite evidence} (see Section~\ref{sec:nuisance_integration}).
-In this version, SCL~essentially generalizes Bayesian integration.
-
-The potential of SCL for unsupervised learning is perhaps suprising considering that it is a discriminative model.
-This stems from the fact that, unlike classical discriminative models, SCL~exploits direct information conveyed by the data about both the parameters of interest and the nuisance parameters, reflecting the generative nature of the underlying feature distribution models.
-This essential difference is illustrated in Figure~\ref{fig:dgraph}: the factor graph model underlying CL in its parametric version (see Figure~\ref{fig:fgraph5}) is not reducible to a belief network of the type represented in Figure~\ref{fig:dgraph2}, in which the data and the nuisance parameter are marginally independent.
-In contrast, the data is informative about the nuisance parameter in a CL model, as shown in Figure~\ref{fig:dgraph3}.
-
-In summary, CL has the potential to yield weakly supervised or unsupervised Bayesian-like inference procedures depending on the particular task at hand.
+%In summary, CL has the potential to yield weakly supervised or unsupervised Bayesian-like inference procedures depending on the particular task at hand.
 This property reflects the encoding of statistical relationships between the data and {\em all} unknown parameters.
@@
 
-%{\color{red} {\em Mardi 12 avril 2016.} Je laisse reposer la version actuelle 3-4 semaines avant mise \`a jour sur arxiv.
-La plupart des points qui restaient obscurs en f\'evrier ont, me semble-t-il, \'et\'e lev\'es: lien avec les neural networks/RBM, strat\'egies d'apprentissage, optimisation des poids, int\'er\^et du super composite (multi-class problems).
-Sur l'optimisation des poids, il reste une petite zone d'ombre: faut-il r\'egulariser d'apprentissage de mani\`ere \`a \'eviter une pond\'eration \'eparse?
-L'argument pour serait d'acc\'el\'erer la convergence stochastique de la SCL vers l'utilit\'e th\'eorique.
-On en parle d\'ej\`a un peu dans la version actuelle en disant que si 2 attributs sont aussi informatifs l'un que l'autre, alors le mieux est de leur donner le m\^eme poids.
-Plus g\'en\'eralement, on pourrait aborder cet aspect via: 1) des contraintes de r\'epartition ``d\'emocratique'' des poids pour \'eviter une situation ``winner-take-all'' (eg, avec une p\'enalisation quadratique des poids); ou alors, 2) des contraintes de variations impos\'ees aux colomnes de la matrice de poids, ce qui donnerait une esp\`ece de compromis entre CL classique et SCL...
-mais je ne suis pas certain de l'int\'er\^et.}
+{\color{red}Product of fucking experts \cite{Hinton-02}.
+Log-linear pool to build a generative model (assuming in fact independence between experts).
+Better seen as a ``log-mixture model''.
+Sounds like each expert is associated with a class, so it's the same as a RBM.
+Contrastive learning approximates ML parameter estimation.
+The experts are learning jointly.
+It's just a generative model.}
+
+{\color{red}Two-step training: generative phase then discriminative phase. Why is it cool?}
+
+{\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
+
+{\color{red}Boosting.}
 
@@
 
-As a straightforward  extension of marginal CL, each feature-based likelihood $\ell_i(x)=p(z_i|x)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
+As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
 \begin{equation}

### 2018-11-15 papers/composite_inference/draft.tex
Modifs 15/11 on CBI

@@
 \begin{abstract}
-This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes conditional entropy relative to a given prior and subject to multiple mean log-likelihood constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
+This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes conditional entropy relative to a given prior and subject to multiple mean log-likelihood inequality constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
 We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, which can be powerful in shallow learning and transfer learning problems.
@@
 Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
-Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16}.
+Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
 
-A strong limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter elimination because they treat the data and the model parameters as marginally independent, meaning that the data conveys no information about the parameters if the variable of interest is not observed.
-This is illustrated in Figure~\ref{fig:graph_comparison} by the respective oriented graph representations of generative and discriminative models.
-For the same reason, supervised learning in discriminative models is less efficient than in comparable generative models, explaining why the latter tend to outperform the former in small training sets \cite{Ng-01}.
-Pure discriminative models are thus of little use outside the context of big labeled data.
+A strong limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as {\em marginally independent}, meaning that the data conveys no information about the parameters unless the variable of interest is observed.
+This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models.
+For the same reason, supervised learning in discriminative models is less precise than in comparable generative models, hence less effective in small training sets \cite{Ng-01}.
+Overall, pure discriminative models are of little use outside the context of big labeled data.
 
@@
 \subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
-\caption{Belief networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the variable of interest, the data, and the model parameters.
+\caption{Bayesian networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the variable of interest, the data, and the model parameters.
 Note the marginal independence of~$Y$ and~$\theta$ in the discriminative model.}
@@
 Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a single clue in the form of a likelihood function of the hidden variables.
-The agent opinions are then aggregated into a unique probability distribution analogous to a Bayesian posterior.
+The agent opinions are then aggregated into a probability distribution on the hidden variables analogous to a Bayesian posterior.
 
 We further argue that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum relative conditional entropy.
-This perspective entails a method to tune the weights of the different agents from training data as in a typical discriminative learning scenario.
+This perspective entails a method to optimize the weights of the different agents from training data as in a typical discriminative learning scenario.
+
+%Something for true statisticians.
+Maybe we could clarify that we use the term ``Bayesian'' in the sense of ``empirical Bayesian'', implying that the parameters $\theta$ are to be estimated somehow contrary to a ``full Bayesian'' approach where they would be integrated out at inference time.
 
@@
 
-The coordinator discards the priors used by the agents.
+The coordinator discards the priors used by the agents (if non-uniform).
 
-A cheap MaxEnt argument was already given by \cite{Wang-14}.
-$I$-projection onto a set of predictive distributions compatible with mean log-likelihood contraints.
+A cheap MaxEnt argument was already given by \cite{Wang-14}  (standard, non-conditional MaxEnt at fixed $y$, hence just an existence result but no feasible weight tuning).
+
+What we do here is maximum {\em conditional} entropy.
+It's slightly more complex than a simple $I$-projection.
+The search space is a set of predictive distributions compatible with mean log-likelihood contraints.
 If I know the generative distribution, I know the expected log-likelihood.
-
-Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. but is there some intuition behind that?
 
@@
 $$
-E[\log \ell_i(x)] \geq \bar{\ell}_i
+E[\log p(z_i|x)] \geq c_i \equiv \int p(z_i|x) \log p(z_i|x) dx
 $$
 
-What does the constraint mean? It means that a model is admissible if it compresses the data at least as well as all my agents.
+What does the constraint mean?
+It means that a model is admissible if it compresses the data at least as well as all my agents.
+So we are not assuming that the generative models are {\em true}.
+We only assume that they are useful to compress the (reduced) data.
+Obvious connection to MDL.
+
+Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
 
@@
 CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
-In this note, we have shown a deep connection between CL and  probabilistic inference, thereby establishing CL as a class of discriminative models.
-The fact that it is built from a set of marginal feature-specific generative distributions makes it a semi-generative, semi-discriminative approach.
+We have shown a deep connection between CL and probabilistic inference, thereby establishing CL as a class of discriminative models.
+Because CL is built from a set of marginal feature-specific generative distributions, it is in essence a two-step semi-generative, semi-discriminative learning approach.
+In the first, ``generative'' phase, the feature distributions are learned; in the second, ``discriminative'' phase, the feature weights are learned.
+This strategy can be thought of as a form of non-adaptive boosting.
 
-Two-step training.
+%The first phase corresponds to the training of ``weak learners''. The second phase amounts to a form of boosting.
+
+A purely discriminative learning could be used instead but...
 Why potentially more efficient than pure discriminative training: because generative training is always more efficient than discriminative training if models are comparable.
 So the key point is that ``we have less parameters in the discriminant phase".
+Good in small datasets.
+But also in asymptotic regime if the features are weakly or highly correlated (???).
 
-First training phase.
-Supervised or unsupervised.
-This could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimizes parameters for generation of observable features.
+The first training phase is easier if supervised but could be unsupervised too (using mixture models).
+
+This first learning phase could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimizes parameters for generation of observable features.
 The main difference is that the two learning phases of CL optimize disjoint subsets of parameters unlike RBMs.
@@
 Once pre-training is complete, the CL~weights may be tuned: this second training stage is, again, supervised in essence, but does not require additional training examples since it is fully determined by the feature distributions learned in the pre-training step.
+
+Needs features. Not a representation learning method, but could be coupled why not.
 
@@
 {\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
-
-{\color{red}Boosting.}
 

### 2018-11-15 notes/composite_inference.tex
Some thoughts about discriminative models

--- 2018-10-29 notes/composite_inference.tex
@@
 \max_{\alpha, \beta} \left[
-\sum_k \log \pi_\alpha(y_k)/\pi_\alpha(z_{x_k}) + \sum_k \log p_\beta(z_{x_k}|x_k)
+\sum_k \log \frac{\pi_\alpha(y_k)}{\pi_\alpha(z_{x_k})} + \sum_k \log p_\beta(z_{x_k}|x_k)
 \right]
@@
 
-En supposant que les param\`etres $\alpha$ de $\pi$ et $\beta$ des autres lois g\'en\'eratives sont disjoints et que pi est la loi de la classe~0, cela revient \`a maximiser chaque terme ind\'ependamment, donc notamment:
+En supposant que les param\`etres $\alpha$ de $\pi$ et $\beta$ des autres lois g\'en\'eratives sont disjoints et que~$\pi$ est la loi de la classe~0, cela revient \`a maximiser chaque terme ind\'ependamment, donc notamment:
 $$
@@
 
-Si on suppose en outre que $\pi(z1)$ et $\pi(z2)$ correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution ``nulle'' de $z_1$ (l'attribut ``exhaustif'' pour la classe~1) sur les donn\'ees issues des classes~0 et~2...
+Si on suppose en outre que $\pi(z_1)$ et $\pi(z_2)$ correspondent \`a des sous-ensembles disjoints de alpha, on voit que la m\'ethode classique consiste \`a ajuster la distribution ``nulle'' de $z_1$ (l'attribut ``exhaustif'' pour la classe~1) sur les donn\'ees issues des classes~0 et~2...
 
@@
 \quad \Rightarrow 
-\arg\min R(q) = K \exp[\int \pi(\theta) \log p_\theta(y) dy ]
+\arg\min R(q) = K \exp\left[\int \pi(\theta) \log p_\theta(y) dy \right]
 $$
@@
 On voit alors que quelque soit $p$, la distribution de Shtarkov (1987) dite NML minimise {\em toujours} $L(p, q)$: 
-$$\arg\min_q L(p, q) = \frac{\ell_\star(y)}{Z_{nml}}$$ 
+$$\arg\min_q L(p, q) = \frac{\ell_\star(y)}{Z_{\rm NML}}$$ 
 avec en plus la propri\'et\'e que le minimum atteint est ind\'ependant de p: 
-$$\min_q L(p, q) = \log Z_{nml}$$
+$$\min_q L(p, q) = \log Z_{\rm NML}$$
 
@@
 plut\^ot que via la notion empirique d'\'evidence.
-Dans cette perspective, la distribution r\'ealisant le maximum de S sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure {\em optimiste} qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:
+Dans cette perspective, la distribution r\'ealisant le maximum de~$S$ sur la famille apparait comme distribution caract\'eristique naturelle et le maximum en question donne une mesure {\em optimiste} qu'il est naturel de consid\'erer comme log-vraisemblance du mod\`ele:
 $$S_\star = \max_\theta S(p, p_\theta)$$
 
-Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la ``vraie'' distribution des donn\'ees, et que S* est donc inconnu. Mais Akaike justifie l'approximation: 
-$$S_\star \approx log \ell_\star(y) - k$$ 
+Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la ``vraie'' distribution des donn\'ees, et que $S_\star$ est donc inconnu. Mais Akaike justifie l'approximation: 
+$$S_\star \approx \log \ell_\star(y) - k$$ 
 o\`u k est la dimension de $\theta$. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement... 
-
 J'ai l'impression qu'il suffit d'appliquer le fameux th\'eor\`eme de Wilks: 
 $$2 \log \frac{\ell_\star(y)}{p(y)} \sim \chi_2(k)$$ 
-valable asymptotiquement SI $p(y)$ APPARTIENT A LA FAMILLE et sous les conditions usuelles d'\'echantillonnage simple.
+valable asymptotiquement {\em si $p(y)$ appartient \`a la famille} et sous les conditions usuelles d'\'echantillonnage simple.
 Wilks nous dit alors que la log-vraisemblance profile sur-estime le score du mod\`ele par la moiti\'e d'un $\chi_2$ \`a $k$ degr\'es de libert\'e.
 En prenant l'esp\'erance de l'expression ci-dessus, les choses deviennent tr\`es claires:
-$$S_\star = E[log \ell_\star(y)] - \frac{k}{2}$$
+$$S_\star = E[\log \ell_\star(y)] - \frac{k}{2}$$
 
 Ainsi, la quantit\'e: 
-$$log \ell_\star(y) - k/2$$ 
+$$\log \ell_\star(y) - k/2$$ 
 fournit un estimateur sans biais de $S_\star$. C'est presque la correction d'Akaike... \`a un facteur 2 pr\`es. Akaike se serait-il plant\'e?
@@
 Il semble que cette incoh\'erence vienne du fait qu'Akaike cherche en fait \`a estimer: 
-$$S_{\star\star} = \int p(y) \log ML(y) dy$$ 
+$$S_{\star\star} = \int p(y) \log \ell_\star(y) dy$$ 
 et non $S_\star = \max_\theta S(p, p_\theta)$, ce qui me semble plus naturel mais c'est moi.
 
-Quoiqu'il en soit, tout se passe comme si on d\'efinissait l'\'evidence du mod\`ele comme: $q(y) = f(k) \ell_\star(y)$ avec $f(k) = \exp(-k/2)$ ou $\exp(-k)$ mais c'est justifi\'e sous l'hypoth\`ese {\em absurde dans le contexte de la s\'election de mod\`ele} que $p(y)$ est dans la famille.
+Quoiqu'il en soit, tout se passe comme si on d\'efinissait l'\'evidence du mod\`ele comme: $q(y) = \ell_\star(y)/Z_{\rm AIC}$ avec $Z_{\rm AIC} = \exp(k/2)$ ou $\exp(k)$ mais c'est justifi\'e sous l'hypoth\`ese {\em absurde dans le contexte de la s\'election de mod\`ele} que $p(y)$ est dans la famille.
 Si tous les mod\`eles test\'es sont corrects, il suffit de prendre le plus parcimonieux!
@@
 $$
-G = E[\frac{\partial^2 p_\theta}{\partial\theta^2}\big / p_\theta]
+G = E\left[\frac{\partial^2 p_\theta}{\partial\theta^2}\big / p_\theta\right]
 $$
@@
 En conclusion, l'int\'er\^et majeur de l'approche d'Akaike est, \`a mon sens, de justifier la vraisemblance profile comme mesure d'\'evidence \`a la limite des grands \'echantillons.
-Sous r\'eserve que n soit constant \`a travers les mod\`eles (ce qui n'est pas le cas dans le recalage).
+Sous r\'eserve que $n$ ne varie pas en fonction des mod\`eles (ce qui se produit en recalage).
 
@@
 
-===========================================================
+\section{Synth\`ese}
+
+Le terme "inférence composite" n'est pas bien choisi car la notion de vraisemblance composite est d\'efinie par ailleurs et est compl\`etement orthogonale \`a ce que nous faisons ici: elle correspond \`a l'utilisation de repr\'esentations multiples en parall\`ele pour l'inf\'erence.
+La "vraisemblance restreinte", elle aussi, est une notion orthogonale correspondant \`a l'utilisation de repr\'esentations emboit\'ees.
+
+Le terme {\em vraisemblance r\'eduite} semble plus ad\'equat.
+
+La notion centrale est celle de {\em quantit\'e exhaustive}, \`a savoir une fonction des observations et des variables cach\'ees telle que la distribution g\'en\'erative des observations admette la forme: $p(y|x)=f(z(x,y),x)g(y)$.
+Notons qu'il existe toujours une telle quantit\'e puisque $z(x,y)=y$ marche toujours.
+
+On montre alors que, \`a $x$ fixe, $z_x$ s'interpr\`ete comme une statistique exhaustive pour un certain probl\`eme d'inf\'erence binaire.
+D'o\`u l'interpr\'etation de la vraisemblance en termes d'aggr\'egation d'opinions.
+Ce n'est qu'une interpr\'etation possible d'un type de fonction de vraisemblances, concept par ailleurs on ne peut plus standard.
+
+Je trouve l'expression "repr\'esentation adaptative" plus jolie que "quantit\'e exhaustive" mais elle sous-entend que chaque expert serait libre du choix de sa repr\'esentation.
+Or une repr\'esentation adaptative arbitraire est susceptible de fournir des classifieurs idiots comme celui que j'avais imagin\'e en 2015.
+La construction inverse par combinaison des experts (c-\`a-d la m\'ethode de Baggenstoss) n'a de sens que si on suppose que la repr\'esentation adaptive est une quantit\'e exhaustive (vis-\`a-vis de la repr\'esentation globale choisie).
+
+Les choses deviennent non-standard quand on envisage l'\'elimination des param\`etres (de nuisance).
+L'intepr\'etation par aggr\'egation d'opinions justifie une approche distribu\'ee dans laquelle l'\'elimination des param\`etres est d\'el\'egu\'ee aux ``experts''.
+
+Notons que l'\'elimination distribu\'ee reste cependant compatible avec le paradigme classique car, lorsque la quantit\'e exhaustive n'est autre que l'observation elle-m\^eme, chaque expert travaille avec la m\^eme fonction de vraisemblance (en diff\'erents points).
+Ce que je propose est donc une g\'en\'eralisation de l'\'elimination traditionnelle de param\`etres.
+
+On peut encore g\'en\'eraliser cette approche \`a la vraisemblance composite en traitant ind\'ependamment chaque composante comme une vraisemblance r\'eductible.
+C'est une piste de recherche.
+Hors sujet ici.
+
+\section{Inf\'erence composite revisit\'ee}
+
+J'ai fait fausse route avec la vraisemblance super-composite en recalage.
+La vraisemblance du recalage n'est ni super-composite, ni m\^eme composite.
+C'est une vraisemblance classique.
+Faut-il pour autant jeter ma m\'ethode d'inf\'erence composite \`a la poubelle?
+
+Je doute fort aujourd'hui de l'int\'er\^et de la vraisemblance super-composite.
+J'avais montr\'e dans mon topo que l'apprentissage de poids adaptatifs induit un effet winner-take-all qui consiste \`a prendre un seul attribut par hypoth\`ese.
+On retombe alors sur une simple vraisemblance r\'eduite...
+
+C'est d\^u \`a la contrainte de poids de somme unitaire, directement h\'erit\'ee de notre interpr\'etation de la vraisemblance composite comme une aggr\'egation d'opinions.
+Faut-il remettre en cause cette vision des choses?
+
+Il y a quelque chose qui ne colle pas dans cette affaire.
+L'intuition du Bayes na\"if, aka vraisemblance composite classique (avec poids non-adaptatifs), est que chaque attribut apporte une contribution utile \`a l'inf\'erence, m\^eme s'il existe une certaine redondance entre attributs.
+
+Or, si on essaie d'apprendre les poids plut\^ot que de les choisir {\em a priori}, il faudrait alors prendre des poids adaptatifs selon ma th\'eorie...
+pour le r\'esultat qu'on conna\^it.
+Conclusion: la vraisemblance composite n'a aucun sens.
+
+C'est bien s\^ur absurde.
+Si, par exemple, les attributs sont vraiment ind\'ependants, cette m\'ethode fournit une pond\'eration totalement sous-optimale.
+Car elle suppose implicitement une sorte de "redondance maximale" entre les attributs qui n'est pas la r\'ealit\'e.
+
+Revenons \`a l'aggr\'egation de vraisemblances. Si j'ai plusieurs $p(z_i|x)$, il est naturel de former une {\em somme} g\'eom\'etrique (log-linear pooling):
+$$
+p_\lambda(x|y) = \frac{\pi(x)}{Z(\lambda, y)}\prod_i p(z_i|x)^{\lambda_i},
+\qquad 
+Z(\lambda, y) = \int \pi(x) \prod_i p_i(z_i|x)^{\lambda_i} dx
+$$
+
+Pourquoi g\'eom\'etrique plut\^ot qu'arithm\'etique ou autre r\`egle d'aggr\'egation d'opinions?
+
+On peut justifier cette forme par diff\'erents principes: Bay\'esien externe (Genest, 1986), axiomes de bon aloi (Tarentola et Valette, 1982), principes info-variationnels (Garg et al, 2004; Allard et al, 2012)...
+Toutes ces approches stipulent des poids unitaires ou une somme de poids unitaire qui, alli\'ee \`a un principe de raison insuffisante, justifie des poids uniformes, $\lambda_i\equiv 1/n$.
+En gros, elles justifient la m\'ethode Bayes na\"ive.
+
+Mais la meilleure justification, \`a mon sens, c'est le principe du maximum d'entropie conditionnelle (Berger et al, 1996) sous des contraintes du type:
+$$
+E[\log p(z_i|x)] = \bar{\ell}_i,
+$$
+o\`u l'esp\'erance est prise sur $x$ et $y$.
+
+Pourquoi est-ce la meilleure approche?
+Tout simplement parce que le principe du maximum d'entropie est la solution d'un probl\`eme de pr\'ediction optimale (cf Gr\"unwlad, 2005) qui correspond exactement \`a ce qu'on veut faire ici: pr\'edire $x$ au mieux en fonction de $y$ en utilisant les vraisemblances restreintes $p(z_i|x)$.
+Au fond, nous avons ici un probl\`eme d'apprentissage discriminant et il n'y a pas de magie: on peut initialiser les param\`etres de fa\c con raisonnable, mais on fera forc\'ement mieux en laissant parler les donn\'ees.
+
+Maxent nous dit que si on est capable d'estimer la log-vraisemblance moyenne pour chaque attribut, ce qui suppose une premi\`ere phase d'apprentissage "g\'en\'eratif", alors la forme ci-dessus est optimale en un certain sens pour la pr\'ediction de~$x$.
+Un deuxi\`eme apprentissage permet de d\'eterminer les poids des "experts" en tenant compte des d\'ependances entre attributs, en maximisant la vraisemblance (conditionnelle):
+$$
+L(\lambda) 
+= \sum_k \log \frac{p_\lambda(x_k|y_k)}{\pi(x_k)}
+= \sum_k \left[
+-\log Z(\lambda, y_k) + \sum_i \lambda_i \log p_i(z_{ik}|x_k)
+\right],
+$$
+ce qui revient \`a assimiler les moments $\bar{\ell}_i$ \`a leurs moyennes empiriques dans la fonction duale du MaxEnt.
+C'est ni plus ni moins qu'un apprentissage supervis\'e.
+On peut voir ce 2e apprentissage comme une forme de {\em boosting} puisque il s'agit de combiner des inf\'erences simples.
+
+Plusieurs remarques s'imposent:
+\begin{enumerate}
+    \item L'aggr\'egation d'opinions n'est autre qu'un classifieur MaxEnt d'un genre un peu particulier o\`u les contraintes de moments incorporent des mod\`eles g\'en\'eratifs plut\^ot que des quantit\'es arbitraires.
+    \item  Une fois r\'ealis\'e le premier apprentissage, le deuxi\`eme apprentissage est d\'etermin\'e par les distributions g\'en\'eratives apprises et par la distribution marginale empirique $h(y)$, cf.~ma note {\tt oct14.pdf}.
+    \item On peut voir le MaxEnt conditionnel comme la recherche d'une distribution jointe $p(x,y)$.
+Il n'y a pas unicit\'e de la solution.
+Toute distribution de la forme $h(y)p_\star(x|y)$, o\`u $h(y)$ est la marginale d'une distribution $h(x,y)$ compatible avec les contraintes, est optimale.
+    \item Le choix d'un {\em a priori} sur $x$ est n\'ecessaire pour d\'efinir les contraintes.
+Il est naturel de choisir $\pi(x)$ (qui est uniforme dans la pr\'esentation de Berger), ce qui revient \`a poser que la distribution empirique des donn\'ees ajust\'ee pour avoir $h(x)=\pi(x)$ est dans l'espace de recherche.
+    \item Toutefois, la m\'ethode n'impose pas que les distributions admissibles aient $\pi(x)$ comme loi marginale.
+En d'autres termes, tout mod\`ele ayant un fort pouvoir pr\'edictif est susceptible de nous int\'eresser qu'il utilise un {\em a priori} farfelu ou pas.
+C'est pragmatique \`a d\'efaut d'\^etre bay\'esien.
+    \item Cela \'etant, on peut voir la r\`egle d'aggr\'egation comme une g\'en\'eralisation de la loi de Bayes dans laquelle $\pi(x)$ joue le r\^ole d'{\em a priori}.
+On pourrait dire un {\em a priori} composite.
+    \item Une variante du MaxEnt, le principe MIL (Yuan et Clarke, 1999), permet d'imposer la contrainte $\pi(x)=p(x)$.
+Elle conduit \`a estimer en alternance les distributions marginale et conditionnelle $p(y)$ et $p(y|x)$ via l'algorithme de Blahut-Arimoto.
+Cette approche est adapt\'ee \`a des contraintes de moments arbitraires mais aurait peu de sens coupl\'ee avec le pr\'e-apprentissage "g\'en\'eratif".
+Le MIL est une s\'election de mod\`ele pour une inf\'erence authentiquement bay\'esienne et non composite.
+\end{enumerate}
+
+Au final, la diff\'erence entre l'inf\'erence composite et un bon vieux classifieur MaxEnt (qui comprend notamment la r\'egression logistique comme cas particulier) tient au type de contraintes.
+Quel avantage potentiel a l'inf\'erence composite?
+C'est LA question.
+
+Une bonne propri\'et\'e est d'\^etre asymptotiquement exacte dans le cas d'attributs mutuellement ind\'ependants (on a alors $\lambda\equiv 1$).
+Cela reste vrai en pr\'esence d'attributs r\'epliqu\'es:  les m\^emes attributs se partagent alors des poids de somme unitaire, c'est l'hypoth\`ese implicite de l'aggr\'egation bay\'esienne externe.
+
+Donc, intuitivement, l'inf\'erence composite est un bon mod\`ele pour des attributs peu ou tr\`es corr\'el\'es, le cas d\'efavorable \'etant a priori celui d'attributs mod\'er\'ement corr\'el\'es.
+
+Ce raisonnement porte sur la forme du mod\`ele pr\'edictif mais se pose aussi la question de l'estimation des param\`etres.
+Le pr\'e-apprentissage "g\'en\'eratif" (qui fait penser \`a Hinton 2006) permet de r\'eduire le nombre de param\`etres \`a apprendre dans la phase d'apprentissage discriminant.
+C'est int\'eressant si l'apprentissage "g\'en\'eratif" est statistiquement efficace.
+
+Pour comparer ce qui est comparable, l'alternative serait d'ajuster le mod\`ele pr\'edictif complet $p_{\lambda,\theta}(x|y)$, o\`u $\theta$ repr\'esente les param\`etres des lois g\'en\'eratives, par apprentissage discriminant.
+Malgr\'e son caract\`ere contre-intuitif compte tenu de la nature de $\theta$, cette approche est en principe indiqu\'ee dans la limite des grands \'echantillons (Ng et Jordan, 2001).
+
+Nous ne somme pas dans ce contexte.
+Le fait de travailleur avec une repr\'esentation donn\'ee $y=z_1,z_2,\ldots$, par exemple dans le cadre d'un apprentissage par transfert, revient \`a supposer que nous ne disposons pas d'un ensemble d'apprentissage gigantesque.
+
+Mais, m\^eme si c'\'etait le cas, est-ce que l'apprentissage purement discriminant ferait forc\'ement mieux?
+Dans le cas o\`u le mod\`ele est exact, la r\'eponse est non.
+Dans le cas d'attributs "mod\'er\'ement corr\'el\'es", c'est oui en principe.
+Beaucoup mieux?
+Je ne sais pas.
+
+Quoiqu'il en soit, l'inf\'erence composite d\'ecoule d'une approche {\em divide and conquer} qui a le m\'erite de simplifier l'apprentissage sur le plan calculatoire, donc d'apprendre plus vite, de r\'eduire les risques de convergence locale, etc.
+
+En d\'efinitive, cela fait pas mal d'arguments en faveur de l'inf\'erence composite.
+C'est du boosting de classifieurs probabilistes, une voie interm\'ediaire entre le Bayes na\"if et la r\'egression logistique, de l'apprentissage semi-g\'en\'eratif, c'est original, intuitif, relativement l\'eger sur le plan calculatoire...
+
+Je finis par retomber sur mes pattes.
+La morale de l'histoire, c'est qu'il peut y avoir un b\'en\'efice \`a  d\'ecomposer l'apprentissage en deux \'etapes: une 1ere phase d'apprentissage g\'en\'eratif pour obtenir les mod\`eles $p(z_i|x)$, une 2e phase d'apprentissage discriminant pour obtenir la pond\'eration.
+Ca semble id\'eal pour des probl\`emes d'apprentissage {\em small data}.
+
+On peut imaginer des variantes avec des a priori (conjugu\'es de pr\'ef\'erence) sur les coefficients pour am\'eliorer l'efficacit\'e statistique de l'apprentissage.
+Imposer, par exemple, qu'ils ne s'\'ecartent pas trop de $1$ ou $1/n$.
+
+Dans le r\'egime {\em big data}, \`a repr\'esentation fixe, on a l'option MIL quitte \`a balancer des attributs \`a la pelle dans l'entonnoir. Question ouverte.
+
+\section{Pivotalit\'e et remise en cause du principe de maximum de vraisemblance}
+
+C'est ici que je brille de mille feux. Le bouquet final.
+
+\section{Maximum entropie}
+
+\section{Are discriminative models Bayesian?}
+
+My introduction on composite Bayesian inference seems to imply that discriminative models are not Bayesian. Well, they can be represented by a Bayesian network, so they are Bayesian, right?
+
+So what is the matter with them?
+Well, they seem to contradict the likelihood principle because we don't use a likelihood function of the variable of interest.
+Instead, the discriminative model gives us directly what we are intested in, the posterior distribution, as it is a joint distribution of the form:
+$$
+p(x,y) = \pi(y) p(x|y)
+$$
+
+Least we forget, the goal is to make an inference about $x$ from $y$. But both the ``likelihood'' and the ``prior'' are things we won't even try to compute: 
+$$
+p(y|x) = \frac{p(x,y)}{p(x)}, 
+\qquad
+p(x) = \int \pi(y) p(x|y) dy
+$$
+
+It would be like doing a double inversion...
+pretty stupid, isn't it?
+Nonetheless, everything happens {\em as if} we were doing just that -- it's mathematically equivalent.
+So can we say that discriminative models violate the likelihood principle?
+No, we can't.
+
+Can we at least claim that $p(x,y)$ is not a realistic generative model?
+That would sound very picky if $\pi(y)$ is some large sample-based empirical distribution (we don't compute it in practice, but let's imagine...) and $p(x|y)$ is a reasonable predictive distribution.
+
+Again, what is the matter?
+The {\em real} difference between generative and discriminative models is that the former have all their parameters describing $p(y|x)$ while the latter have all their parameters describing $p(x|y)$.
+It's a question of parameters at the end of the day.
+
+Therefore, it's a question of learning.
+We can argue that generative models can ``learn faster'' than discriminative ones and it's because the learning process involves ``understanding the data''.
+When you model data by an empirical distribution, you clearly don't understand a shit about the data.
+You can make fairly good predictions but you're like a monkey.
+
+\section{Blablabla...}
 

### 2018-11-15 papers/composite_inference/draft.tex
Some thoughts about discriminative models

@@
 Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
-Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
+Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
 

### 2018-11-16 notes/composite_inference.tex
Modifs 16/11/18

@@
 The {\em real} difference between generative and discriminative models is that the former have all their parameters describing $p(y|x)$ while the latter have all their parameters describing $p(x|y)$.
-It's a question of parameters at the end of the day.
+Is it a question of parameters at the end of the day?
 
-Therefore, it's a question of learning.
-We can argue that generative models can ``learn faster'' than discriminative ones and it's because the learning process involves ``understanding the data''.
-When you model data by an empirical distribution, you clearly don't understand a shit about the data.
-You can make fairly good predictions but you're like a monkey.
+No, it's a question of what the models describe.
+A generative model describes the process of acquiring data, it has to reflect some physical understanding.
+A discriminative model describes a convenient approximating family for the recovery of a state, there is no physics involved.
+As a by-product, the generative model might well produce a good predictive distribution for the data, but it does that without understanding anything about the data.
+
+Just like an histogram is a pretty dumb but good predictor if we have a shit load of data to compute it.
+Instead, you can have a parametric model that reflects some understanding of what the distribution is.
+It takes much less data.
+
+La vraie question, c'est de savoir quel est le lien de cause \`a effet entre $x$ et $y$.
+Si $x$ est la ``cause'' de $y$, alors le mod\`ele g\'en\'eratif est naturel.
+Si c'est l'inverse, alors le mod\`ele discriminant est une \'evidence.
+
+Par exemple, supposons que $y$ soit une vitesse moyenne et $x$ la distance parcourue en un temps donn\'e.
+On a $x=t y$ et c'est m\^eme d\'eterministe, donc aucune question \`a se poser: ce mod\`ele purement discriminant dans ce contexte donne la r\'eponse.
+
+On peut se dire que la vitesse est la cause de la distance parcourue, mais la vitesse se d\'eduit tout aussi bien de la distance: $y=x/t$.
+Donc on pourrait aussi s'en sortir avec un mod\`ele g\'en\'eratif.
+
+Si maintenant notre mesure $y$ est bruit\'ee, il n'y a plus d'ambigu\"it\'e: on \'ecrit $y=x/t+\epsilon$, c'est-\`a-dire qu'on travaille avec un mod\`ele g\'en\'eratif.
+La vitesse mesur\'ee devient la ``cons\'equence'' de la distance \`a cause du bruit de mesure.
+
+On voit bien que la notion de causalit\'e est assez foireuse.
+
+On voit aussi que la d\'efinition officielle (Wikipedia et cie) de l'inf\'erence statistique est tr\`es restrictive: tirer des conclusions sur une population \`a partir d'un \'echantillon.
+C'est minable.
+Mais la d\'efinition permet de se mettre dans le cadre rassurant de l'\'echantillonnage.
+L'\'echantillon $y$ est bien ``g\'en\'er\'e'' \`a partir de la population.
+
+Je pr\'ef\`ere la d\'efinition de Floyd Bullard: ``Statistical inference means drawing conclusions based on data''.
+
+De mani\`ere g\'en\'erale, il y a l'id\'ee que les donn\'ees, ce qu'on mesure, peut \^etre pr\'edit par l\'etat du monde.
+Oui, il y a bien un lien de causali\'e implicite et la mesure arrive toujours en bout de cha\^ine.
+
+Dans l'exemple pr\'ec\'edent, l'\'etat du monde c'est $(v,x)$ la vraie vitesse et la vraie distance. On a $y=v+\epsilon$ mais par bonheur il existe une relation entre $x$ et $v$, 
+$p(x,v)=\delta(x-tv)p(v)$.
+Autrement dit le param\`etre de nuisance $v$ est facilement \'elimin\'e.
+Plus fondamentalement, les donn\'ees apportent une information sur l'\'etat du monde qui permet de mettre \`a jour sa connaissance.
 

### 2018-11-16 papers/composite_inference/draft.tex
Modifs 16/11/18

--- 2018-11-15 papers/composite_inference/draft.tex
@@
 
-Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable.
-Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; minimum description length \cite{Grunwald-07}; and, last but not least, discriminative models\footnote{A {\em discriminative} model describes the conditional distribution of an unobserved variable of interest given an observable variable, in contrast with a {\em generative} model, which describes the conditional distribution of an observable variable given an unobserved variable.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
+Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
+Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).
 
@@
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models.
-For the same reason, supervised learning in discriminative models is less precise than in comparable generative models, hence less effective in small training sets \cite{Ng-01}.
+For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence it is less effective in small training sets \cite{Ng-01}.
 Overall, pure discriminative models are of little use outside the context of big labeled data.
@@
 Game theoretic interpretation \cite{Grunwald-04}.
+
+The max entropy training turns out to be unsupervised since the moment
+constraints are determined by the 1st learning phase. We have to
+interpret that!!!
 

### 2018-11-22 papers/composite_inference/draft.tex
Modifs 22/11/18

@@
 We only assume that they are useful to compress the (reduced) data.
-Obvious connection to MDL.
+Connection to MDL.
 
-Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
+Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion ofR feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
 
@@
 
-The max entropy training turns out to be unsupervised since the moment
-constraints are determined by the 1st learning phase. We have to
-interpret that!!!
+The dual function amounts to the famous cross-entropy.
 
@@
 
+Logistic regression / naive Bayes example.
+Under homoscedasticity (assuming that each feature has class-independent variance), CBI is equivalent to logistic regression -- because the predictive distribution family is the same.
+This is a case where BCI brings nothing.
+But consider heteroscedasticity, then BCI yields a quadratic classifier.
+Compared to a fully discriminative model, the number of parameters to learned in the discriminative phase is reduced by half.
+
 The first training phase is easier if supervised but could be unsupervised too (using mixture models).
+How do we then deal with label switching issues?
+Can we safely assume conditional feature independence {\em in the generative training phase}?
+I believe so provided that the marginal distribution parameters are disjoint.
+It's obvious in the supervised learning scenario.
 
-This first learning phase could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimizes parameters for generation of observable features.
-The main difference is that the two learning phases of CL optimize disjoint subsets of parameters unlike RBMs.
-CL pre-training relies on weaker assumptions as it does not assume conditional feature independence unlike RBMs.
-Once pre-training is complete, the CL~weights may be tuned: this second training stage is, again, supervised in essence, but does not require additional training examples since it is fully determined by the feature distributions learned in the pre-training step.
+If unsupervised, the first learning phase could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimize parameters for generation of observable features.
+BCI is comparable with an RBM with a single output unit (which is just a generative model assuming conditionial feature independence).
+The key difference is that the RBM is a full generative model while BCI only deals with marginal models, hence relying on weaker assumptions.
+This won't change anything in the pre-training phase but RBMs have to deal with more parameters in the generative learning phase.
+In fully supervised context, RBM pre-training is pointless since all parameters learned in the first phase will be overwritten.
+In BCI, pre-training is crucial even in supervised context because the parameters learned in the discriminative phase (the feature weighs) describe a sub-manifold of the predictive distribution family.
 

### 2018-11-30 notes/composite_inference.tex
Update notes on conditional maxent
Modifs 22/11/18
Modifs 30/11/18

@@
 = \int p(y) \log \frac{p_\theta(y)}{q(y)} dy 
-= D(p\|q) - D(p\|p_theta) 
+= D(p\|q) - D(p\|p_\theta) 
 $$
@@
 
-Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la ``vraie'' distribution des donn\'ees, et que $S_\star$ est donc inconnu. Mais Akaike justifie l'approximation: 
+Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas $p$, la ``vraie'' distribution des donn\'ees, et que $S_\star$ est donc inconnu. Mais Akaike justifie l'approximation: 
 $$S_\star \approx \log \ell_\star(y) - k$$ 
-o\`u k est la dimension de $\theta$. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement... 
+o\`u $k$ est la dimension de $\theta$. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement... 
 J'ai l'impression qu'il suffit d'appliquer le fameux th\'eor\`eme de Wilks: 
@@
 pour le r\'esultat qu'on conna\^it.
-Conclusion: la vraisemblance composite n'a aucun sens.
-
 C'est bien s\^ur absurde.
@@
 \begin{enumerate}
-    \item L'aggr\'egation d'opinions n'est autre qu'un classifieur MaxEnt d'un genre un peu particulier o\`u les contraintes de moments incorporent des mod\`eles g\'en\'eratifs plut\^ot que des quantit\'es arbitraires.
-    \item  Une fois r\'ealis\'e le premier apprentissage, le deuxi\`eme apprentissage est d\'etermin\'e par les distributions g\'en\'eratives apprises et par la distribution marginale empirique $h(y)$, cf.~ma note {\tt oct14.pdf}.
-    \item On peut voir le MaxEnt conditionnel comme la recherche d'une distribution jointe $p(x,y)$.
-Il n'y a pas unicit\'e de la solution.
-Toute distribution de la forme $h(y)p_\star(x|y)$, o\`u $h(y)$ est la marginale d'une distribution $h(x,y)$ compatible avec les contraintes, est optimale.
+    \item L'aggr\'egation d'opinions n'est autre qu'un classifieur MaxEnt d'un genre un peu particulier o\`u les contraintes de moments incorporent des mod\`eles g\'en\'eratifs appris au pr\'ealable plut\^ot que des quantit\'es arbitraires.
+    \item  Une fois r\'ealis\'e le premier apprentissage, le deuxi\`eme apprentissage est d\'etermin\'e par les distributions g\'en\'eratives apprises et par la distribution marginale empirique $h(y)$, cf.~ma note {\tt oct14.pdf} qui contient par ailleurs des erreurs.
+    \item On peut voir le MaxEnt conditionnel comme un MaxEnt relatif standard sur une distribution jointe de marginale connue $h(y)$, donc de la forme $p(x,y)=h(y)p(x|y)$, et par rapport \`a une loi de r\'ef\'erence de la forme $\pi(x,y)=\pi(x)h(y)$.
+Il est donc capital d'avoir une estim\'ee de la marginale $h(y)$.
     \item Le choix d'un {\em a priori} sur $x$ est n\'ecessaire pour d\'efinir les contraintes.
 Il est naturel de choisir $\pi(x)$ (qui est uniforme dans la pr\'esentation de Berger), ce qui revient \`a poser que la distribution empirique des donn\'ees ajust\'ee pour avoir $h(x)=\pi(x)$ est dans l'espace de recherche.
+    \item L'espace de recherche est d\'efini par des contraintes portant sur les log-vraisemblances moyennes  qui sont {\em moins fortes} que supposer connues les distributions des attributs.
+Elles expriment l'id\'ee tr\`es MDL-esque que ce qui est important avec un mod\`ele, ce n'est pas qu'il soit exact, mais le taux de compression qu'il permet d'atteindre.
+    \item Doit-on utiliser des contraintes d'\'egalit\'e ou d'in\'egalit\'e? Les premi\`eres sont plus naturelles mais permettent des poids n\'egatifs... ce qui est malvenu pour des ``opinions''. 
     \item Toutefois, la m\'ethode n'impose pas que les distributions admissibles aient $\pi(x)$ comme loi marginale.
@@
 On pourrait dire un {\em a priori} composite.
-    \item Une variante du MaxEnt, le principe MIL (Yuan et Clarke, 1999), permet d'imposer la contrainte $\pi(x)=p(x)$.
+    \item Une g\'en\'eralisation du MaxEnt conditionnel, le MIL (Yuan et Clarke, 1999) permet de rel\^acher la contrainte que $p(y)$ est connue et (\'eventuellement) d'imposer la contrainte $\pi(x)=p(x)$.
 Elle conduit \`a estimer en alternance les distributions marginale et conditionnelle $p(y)$ et $p(y|x)$ via l'algorithme de Blahut-Arimoto.
-Cette approche est adapt\'ee \`a des contraintes de moments arbitraires mais aurait peu de sens coupl\'ee avec le pr\'e-apprentissage "g\'en\'eratif".
-Le MIL est une s\'election de mod\`ele pour une inf\'erence authentiquement bay\'esienne et non composite.
+\c Ca semble indiqu\'e si on n'a pas confiance dans la distribution empirique $h(y)$ mais, dans ce cas, pourquoi avoir confiance dans les moments?
 \end{enumerate}
@@
 
+\section{Compression}
+
+Let $X\sim p_\star(x)$ and consider a model $q(x)$.
+Assume we know the compression rate of $q$, but don't know $p_\star$.
+Does it make sense to consider admissible any distribution $p(x)$ for which the compression rate is at least as large, hence satisfying the following inequality:
+$$
+L(p,q) \geq L(p_\star,q),
+$$
+with:
+$$
+L(p,q) = \int p(x)\log q(x) dx 
+$$
+
+This is weird because if we somehow know $L(p_\star,q)$, we know for sure that it can't be larger than what it is, right?
+So why not using an equality constraint?
+Unless what we think we know happens to be a lower bound.
+
+$$
+\max_q\min_{p\in P} L(p,q) 
+=
+\min_{p\in P} \max_q L(p,q)
+= \min_{p\in P} L(p,p)
+= -\max_{p\in P} H(p)
+$$
+
 \section{Pivotalit\'e et remise en cause du principe de maximum de vraisemblance}
@@
 
-\section{Maximum entropie}
+\section{Maximum entropie conditionnelle}
+
+$$
+D(p) = \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
+$$
+
+Lagrangian:
+$$
+{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
+$$
+
+Optimal $p$ at fixed $\lambda$:
+$$
+\frac{\partial{\cal L}}{\partial p}
+=
+h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
+= h(y) c(y)
+\quad
+\Rightarrow
+\quad
+p_{\lambda}(x|y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
+$$
+with:
+$$
+z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
+$$
+
+Dual:
+$$
+\psi(\lambda) = \lambda^\top F - \int h(y) \log z(\lambda, y) dy 
+$$
+
+$$
+\nabla z
+=
+\int \pi(x) f(x,y)  e^{\lambda^\top f(x,y)} dx
+$$
+
+$$
+\nabla \nabla^\top z 
+=
+\int \pi(x) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dx
+$$
+
+$$
+\nabla\psi
+= F - \int h(y) \frac{\nabla z}{z} dy
+$$
+
+$$
+\nabla\nabla^\top\psi
+= - \int h(y) \left(
+\frac{\nabla \nabla^\top z}{z} 
+-
+\frac{\nabla z \nabla z^\top}{z^2}
+\right)
+dy
+$$
+
+Note the equivalent expression of the dual function as:
+$$
+\psi(\lambda)
+=
+\int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
+= D(h\|\pi) - D(h\|p_\lambda)
+$$
+
+\section{Maxent via the generalized KL}
+
+It's a bit different as the unit integral constraint is made implicit.
+
+Assume $\int\pi(x)dx=1$ and $\int h(y) dy = 1$. 
+$$
+D(p) 
+= \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
+- \int h(y)p(x|y)dx dy
++ 1
+$$
+
+Goal is to minimize $D(p)$ s.t. mean-value constraints on $p$.
+
+Lagrangian:
+$$
+{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
+$$
+
+Optimal $p$ at fixed $\lambda$:
+$$
+\frac{\partial{\cal L}}{\partial p}
+=
+h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
+= 0
+\quad
+\Rightarrow
+\quad
+p_{\lambda}(x|y) = \pi(x) e^{\lambda^\top f(x,y)}
+$$
+
+Dual:
+$$
+\psi(\lambda) = \lambda^\top F - \int h(y)z(\lambda, y) dy + 1
+$$
+with:
+$$
+z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
+$$
+
+$$
+\nabla\psi(\lambda)
+= F - \int \pi(x) h(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
+$$
+
+$$
+\nabla\nabla^\top\psi(\lambda)
+= - \int \pi(x) h(y) f(x,y)f(x,y)^\top e^{\lambda^\top f(x,y)} dx dy
+$$
 

### 2018-11-30 papers/composite_inference/draft.tex
Modifs 30/11/18

@@
 Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).
+Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence and typically require massive supervised learning -- these include classical techniques (maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
 
@@
 
-\section{Maximum entropy derivation of composite likelihood}
+\section{Maxent composite likelihood}
 \label{sec:maxent}
 
-The coordinator discards the priors used by the agents (if non-uniform).
+We can derive composite likelihood from the conditional maximum entropy principle \cite{BergerA-96}.
+It comes with a method of tuning weights, hence a particular composite likelihood function that we call maxent composite likelihood (MCL).
 
@@
 
-BTW, we can work with inequality constraints to force $\lambda\geq 0$:
+We work with inequality constraints to force $\lambda\geq 0$:
 $$
-E[\log p(z_i|x)] \geq c_i \equiv \int p(z_i|x) \log p(z_i|x) dx
+E[\log p(z_i|x)] \geq c_i \equiv \int \pi(x)p(z_i|x) \log p(z_i|x) dx dy
 $$
@@
 
-Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion ofR feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
+Why not using the feature-based posteriors or Bayes factors rather than the likelihoods?
+It's a possibility (perhaps a connection here with Gr\"unwald's luckiness).
+A justification for not doing it is that the coordinator should discard the priors used by the agents.
+The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.
+
+Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
+
+The derivation assumed the marginal distribution of the data $h(y)$ to be known. In practice, it is estimated by the empirical distribution just like the moments are estimated.
 
@@
 
-The dual function amounts to the famous cross-entropy.
+The dual function boils down to the famous cross-entropy.
+
+Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
+We expect to have $\lambda=1$ but it's not necessarily the case.
+It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
+This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
+So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
+
+Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
+The answer is no.
+The weights generally depend on the prior.
+Hence the maxent composite likelihood is prior-dependent.
+This is an important conceptual difference with the classical notion of likelihood.
+
+\section{Composite EM algorithm for unsupervised learning}
+\label{sec:gem}
+
+In a nutshell: alternate unsupervised learning of partial model parameters ($M$-step) and re-estimation of label posteriors via maxent ($E$-step).
+
+$M$-step: given a posterior estimate $q(x|y)$, do just as if the features were independent and solve: 
+$$
+\max_\theta
+\int \sum_i h(y)q(x|y) \log p_\theta(z_i,x) dx dy
+$$
+
+Note we can estimate cross-feature parameters such as class proportions.
+
+Surrogate $E$-step: given generative parameters, recompute the moment constraints and update $q(x|y)$ according to the conditional maxent principle:
+$$
+\min_q
+\int h(y)q(x|y) \log \frac{q(x|y)}{\pi(x)} dx dy
+$$
+s.t.
+$$
+\int h(y)q(x|y)\log p_\theta(z_i|x) dx dy \geq \int \pi(x)p_\theta(z_i|x)\log p_\theta(z_i|x) dx dy
+$$
+
+However, there is an important issue: how is $h(y)$ defined in both the $M$- and $E$-steps? In supervised context, $h(x,y)$ can be defined beforehand as the prior-corrected empirical distribution:
+$$
+h(x,y) = \pi(x) \frac{h_0(x,y)}{h_0(x)},
+$$
+where $h_0(x,y)$ is the raw empirical distribution.
+In unsupervised context, the best we can do is to replace $h_0(x,y)$ with $h_0(y)q(x|y)$, implying that the empirical distribution of $y$ is reweighted according to:
+$$
+h(y) = \sum_i w_i\delta(y-y_i),
+\qquad
+w_i = \int \pi(x)\frac{q(x|y_i)}{\sum_j q(x|y_j)} dx,
+$$
+which needs to be done each time $q(x|y)$ is updated, that is, after each $E$-step.
+
+The other option is to keep $h(y)$ constant throughout the iterations but adapt $\pi(x)$ as the marginal distribution of labels learned in the $M$-step. In this case, it's no longer a ``prior''.
+
+Are we minimizing a unique functional as in the variational EM algorithm?
+Possibly not.
+We may see this generalization as a two-player game.
+One player tries to predict several data features independently.
+The other player tries to predict labels.
+Together, they find patterns (clusters) in unlabeled data.
+Does the more general version converge?
+Does the game have a value?
+And so on.
+
+In the case of a single feature, is this the well-known EM?
+The $M$-step is clearly the same, what about the $E$-step?
+It is the same if $h(y)p_\theta(x|y)=p_\theta(y|x)\pi(x)$.
+However, this condition won't hold if $h(y)$ is an empirical distribution...
+In this special case, the ``E player'' could use the data distribution model produced by the ``M player'' instead of the empirical one but, more generally, there is no data distribution model so he has to stick to the empirical one.
+
+The maxent is a surrogate for the $E$-step. It reflects the lack of a full model $p_\theta(x,y)$.
 
@@
 \label{sec:discussion}
+
+Deep learning is discriminative...
 

### 2018-12-01 papers/composite_inference/draft.tex
re-birth of super composite likelihood

@@
 This is an important conceptual difference with the classical notion of likelihood.
+
+\section{Super composite likelihood}
+\label{sec:super}
+
+When chosen for computational simplicity, clues may not only convey limited information at individual level: their informativeness may also be very much hypothesis-dependent.
+Consider, for instance, diagnosing a disease from a routine medical checkup.
+Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovascular disease -- and conversely for, say, blood pressure.
+
+This motivates a more general setting where clues can be weighted differently depending on hypotheses.
+
+What happens if we use mean-values constraints conditional on $x$ instead of jointly averaged over $x$ and $y$? This boils down to picking functions of the form:
+$$
+\chi(x-a)\ell_i(x,y)
+$$
+
+The answer is that the maxent solution has the form:
+$$
+p_\lambda(x|y) = \pi(x) \prod_i p(z_i|x)^{\lambda_i(x)}
+$$
+
+In other words, the weights become dependent on the variable of interest. This is the super composite likelihood idea.
 

### 2019-01-11 cv/biblio.tex
Update composite inference

\newpage

%\section*{Publication list}
%\vspace*{-1cm}

%\subsection*{Book chapters}
%\subsection*{Journal papers}
%\subsection*{Peer-reviewed conference papers}
%\subsection*{Abstracts}
%\subsection*{Preprints}
%\subsection*{Theses}

\begin{thebibliography}{100}

\subsection*{Book chapters}

\bibitem{inbook:15}
\underline{A.~Roche}.
\newblock {\em {Image, Video \& 3D Data Registration: Medical, Satellite \&
  Video Processing Applications with Quality Metrics}}, chapter {Medical Image
  Registration Measures}.
\newblock Wiley, August 2015.

\bibitem{book:10}
J.-B. Poline, B.~Thirion, \underline{A.~Roche}, and S.~M\'eriaux.
\newblock {\em Foundational Issues in Human Brain Mapping}, chapter
  {Intersubject variability in fMRI data: causes, consequences, and related
  analysis strategies}, pages 173--192.
\newblock MIT press, 2010.

\bibitem{book:05}
X.~Pennec, N.~Ayache, \underline{A.~Roche}, and P.~Cachier.
\newblock {\em Multi-Sensor Image Fusion and Its Applications}, chapter
  Non-rigid {MR/US} registration for tracking brain deformations, pages
  107--143.
\newblock CRC Press, Boca Raton, USA, first edition, 2006.

\bibitem{book:02}
S.~Warfield, A.~Guimond, \underline{A.~Roche}, A.~Bharatha, A.~Tei, F.~Talos, J.~Rexilius,
  J.~Ruiz-Alzola, C.-F. Westin, S.~Haker, S.~Angenent, A.~Tannenbaum,
  F.~Jolesz, and R.~Kikinis.
\newblock {\em Brain Mapping: The Methods}, chapter Advanced Nonrigid
  Registration Algorithms for Image Fusion, pages 661--690.
\newblock Academic Press, San Diego, USA, second edition, 2002.

\subsection*{Journal papers}

\bibitem{flash:16}
\underline{A.~Roche}, B.~Mar\'echal, T.~Kober, G.~Krueger, P.~Hagmann, P.~Maeder, and
  R.~Meuli.
\newblock {Assessing brain volumes using MorphoBox prototype}.
\newblock {\em MAGNETOM Flash, the Siemens magazine of MRI: RSNA Edition},
  2017.

\bibitem{ijasp:13}
\underline{A.~Roche}.
\newblock {Approximate inference via variational sampling}.
\newblock {\em International Journal of Applied Statistics and Probability},
  1(3):110--120, 2013.

\bibitem{tmi:11}
\underline{A.~Roche}.
\newblock {A four-dimensional registration algorithm with application to joint
  correction of motion and slice timing in fMRI}.
\newblock {\em IEEE Transactions on Medical Imaging}, 30(8):1546--1554, 2011.

\bibitem{media:11}
\underline{A.~Roche}, D.~Ribes, M.~Bach~Cuadra, and G.~Krueger.
\newblock {On the Convergence of EM-Like Algorithms for Image Segmentation
  using Markov Random Fields}.
\newblock {\em Medical Image Analysis}, 15(6):830--839, 2011.

\bibitem{ni:07}
\underline{A.~Roche}, S.~M\'eriaux, M.~Keller, and B.~Thirion.
\newblock {Mixed-effects statistics for group analysis in fMRI: A nonparametric
  maximum likelihood approach}.
\newblock {\em Neuroimage}, 38:501--510, 2007.

\bibitem{tmi:01b}
\underline{A.~Roche}, X.~Pennec, G.~Malandain, and N.~Ayache.
\newblock {Rigid Registration of 3D Ultrasound with MR Images: a New Approach
  Combining Intensity and Gradient Information}.
\newblock {\em IEEE Transactions on Medical Imaging}, 20(10):1038--1049, 2001.

\bibitem{ijist:00}
\underline{A.~Roche}, G.~Malandain, and N.~Ayache.
\newblock {Unifying Maximum Likelihood Approaches in Medical Image
  Registration}.
\newblock {\em International Journal of Imaging Systems and Technology: Special
  issue on 3D imaging}, 11:71--80, 2000.

\bibitem{pv:16}
M.J. Fartaria, A.~Sorega, G.~Krueger, T.~Kober, C.~Granziera, M.~Bach~Cuadra,
  and \underline{A.~Roche}.
\newblock {Partial Volume-Aware Assessment of Multiple Sclerosis Lesions}.
\newblock 2018.
\newblock {\em NeuroImage:Clinical}.
\newblock Accepted.

\bibitem{plos:16}
G.~Bonnier, T.~Kober, M.~Schluep, R.~Du~Pasquier, G.~Krueger, R.~Meuli,
  C.~Granziera, and \underline{A.~Roche}.
\newblock {A New Approach for Deep Gray Matter Analysis Using Partial-Volume
  Estimation}.
\newblock {\em PlosOne}, 11(2):e0148631, February 2016.

\bibitem{plos:14}
K.~O'Brien, G.~Krueger, P.~Hagmann, P.~Maeder, T.~Kober, J.~Marques,
  F.~Lazeyras, and \underline{A.~Roche}.
\newblock {Robust T1-weighted structural brain imaging and morphometry at 7T
  using MP2RAGE}.
\newblock {\em PlosOne}, 9(6):e99676, June 2014.

\bibitem{neuroradiology:17}
J.~Boto, G.~Gkinis, \underline{A.~Roche}, T.~Kober, B.~Mar{\'e}chal, N.~Ortiz, K.-O.~L{\"o}vblad, F.~Lazeyras, and M.I.~Vargas.
\newblock {Evaluating anorexia-related brain atrophy using MP2RAGE-based morphometry}.
\newblock {\em European Radiology}, 27(12):5064--5072, 2017.

\bibitem{investigative_radiology:17}
M.J.~Fartaria, K.~O'Brien, A.~Sorega, G.~Bonnier, \underline{A.~Roche}, P.~Falkovskiy, G.~Krueger, T.~Kober, M.~Bach Cuadra, and C.~Granziera.
\newblock {An ultra-high field study of cerebellar pathology in early relapsing-remitting multiple sclerosis using MP2RAGE}.
\newblock {\em Investigative radiology}, 52(5):265--273, 2017.

\bibitem{ni:16}
P.~Falkovskiy, D.~Brenner, T.~Feiweier, S.~Kannengiesser, B.~Mar\'echal,
  T.~Kober, \underline{A.~Roche}, K.~Thostenson, R.~Meuli, D.~Reyes, T.~Stoecker, M.A.
  Bernstein, J.-P. Thiran, and G.~Krueger.
\newblock {Comparison of accelerated T1-weighted whole-brain structural-imaging
  protocols}.
\newblock {\em Neuroimage}, 124:157--167, 2016.

\bibitem{neurorad:16}
S.~Haller, P.~Falkovskiy, R.~Meuli, J.-P. Thiran, G.~Krueger, K.-O. Lovblad,
  \underline{A.~Roche}, T.~Kober, and B.~Mar\'echal.
\newblock {Basic MR sequence parameters systematically bias automated brain
  volume estimation}.
\newblock {\em Neuroradiology}, 2016.

\bibitem{jmri:15}
M.J. Fartaria, G.~Bonnier, \underline{A.~Roche}, T.~Kober, R.~Meuli, D.~Rotzinger,
  R.~Frackowiak, M.~Schluep, R.~Du~Pasquier, J.-P. Thiran, G.~Krueger,
  M.~Bach~Cuadra, and C.~Granziera.
\newblock Automated detection of white matter and cortical lesions in early
  stages of multiple sclerosis.
\newblock {\em Journal of Magnetic Resonance Imaging}, 43(6):1445--1454, 2015.

\bibitem{biomed:15}
G.~Bonnier, \underline{A.~Roche}, D.~Romascano, S.~Simioni, D.E. Meskaldji, D.~Rotzinger,
  Y.-C. Lin, G.~Menegaz, M.~Schluep, T.J. Du~Pasquier, R.~Sumpf, J.~Frahm,
  J.-P. Thiran, G.~Krueger, and C.~Granziera.
\newblock {Multicontrast MRI quantification of focal inflammation and
  degeneration in multiple sclerosis}.
\newblock {\em BioMed research international}, 2015(Article ID 569123), 2015.

\bibitem{nic:15}
C.~Granziera, A.~Daducci, A.~Donati, G.~Bonnier, D.~Romascano, \underline{A.~Roche},
  M.~Bach~Cuadra, D.~Schmitter, S.~Kl{\"o}ppel, R.~Meuli, A.~von Gunten, and
  G.~Krueger.
\newblock {A multi-contrast MRI study of microstructural brain damage in
  patients with mild cognitive impairment}.
\newblock {\em NeuroImage: Clinical}, 8:631--639, 2015.

\bibitem{nic:14}
D.~Schmitter, \underline{A.~Roche}, B.~Mar\'echal, D.~Ribes, A.~Abdulkadir, M.~Bach~Cuadra,
  A.~Daducci, C.~Granziera, S.~Kl\"oppel, P.~Maeder, R.~Meuli, and G.~Krueger.
\newblock {An evaluation of volume-based morphometry for prediction of mild
  cognitive impairment and Alzheimer's disease}.
\newblock {\em NeuroImage: Clinical}, 7:7--17, 2015.

\bibitem{hbm:15}
D.~Romascano, D.E. Meskaldji, G.~Bonnier, S.~Simioni, D.~Rotzinger, Y.-C. Lin,
  G.~Menegaz, \underline{A.~Roche}, M.~Schluep, R.~Du~Pasquier, J.~Richiardi, D.~Van
  De~Ville, A.~Daducci, T.~Sumpf, J.~Fraham, J.-P. Thiran, G.~Krueger, and
  C.~Granziera.
\newblock {Multicontrast connectometry: A new tool to assess cerebellum
  alterations in early relapsing-remitting multiple sclerosis}.
\newblock {\em Human Brain Mapping}, 36(4):1609--1619, 2015.

\bibitem{actn:14}
G.~Bonnier, \underline{A.~Roche}, D.~Romascano, S.~Simioni, D.~Meskaldji, D.~Rotzinger,
  Y.-C. Lin, G.~Menegaz, M.~Schluep, R.~Du~Pasquier, T.~Sumpf, J.~Frahm, J.-P.
  Thiran, G.~Krueger, and C.~Granziera.
\newblock {Advanced MRI unravels the nature of tissue alterations in early
  multiple sclerosis}.
\newblock {\em Annals of clinical and translational neurology}, 1(6):423--432,
  2014.

\bibitem{hbm:14}
C.~Granziera, A.~Daducci, D.~Romascano, \underline{A.~Roche}, G.~Helms, G.~Krueger, and
  N.~Hadjikhani.
\newblock {Structural abnormalities in the thalamus of migraineurs with aura: A
  multiparametric study at 3 T}.
\newblock {\em Human Brain Mapping}, 35(4):1461--1468, 2013.

\bibitem{cerebellum:13}
C.~Granziera, D.~Romascano, A.~Daducci, \underline{A.~Roche}, M.~Vincent, G.~Krueger, and
  N.~Hadjikhani.
\newblock {Migraineurs without aura show microstructural abnormalities in the
  cerebellum and frontal lobe}.
\newblock {\em The Cerebellum}, 12(6):812--818, 2013.

\bibitem{neurology:12}
C.~Granziera, A.~Daducci, D.~Meskaldji, \underline{A.~Roche}, P.~Maeder, P.~Michel,
  N.~Hadjikhani, G.~Sorensen, R.~Frackowiak, J.-P. Thiran, R.~Meuli, and
  G.~Krueger.
\newblock {A new early and automated MRI-based predictor of motor improvement}.
\newblock {\em Neurology}, 79(1):39--46, 2012.

\bibitem{sinica:08}
M.~Keller, \underline{A.~Roche}, A.~Tucholka, and B.~Thirion.
\newblock {Dealing with Spatial Normalization Errors in fMRI Group Inference
  using Hierarchical Modeling}.
\newblock {\em Statistica Sinica}, 18(4):1357--1374, 2008.

\bibitem{media:08}
C.~Poupon, \underline{A.~Roche}, J.~Dubois, J.-F. Mangin, and F.~Poupon.
\newblock {Real-Time MR Diffusion Tensor and Q-ball Imaging using Kalman
  Filtering}.
\newblock {\em Medical Image Analysis}, 12(5):527--534, 2008.
\newblock Elsevier Medical Image Analysis -- MICCAI'07 best paper.

\bibitem{tmi:07}
E.~Duchesnay, A.~Cachia, \underline{A.~Roche}, D.~Rivi\`ere, Y.~Cointepas,
  D.~Papadopoulos-Orfanos, M.~Zilbovicius, J.-L. Martinot, and J.-F. Mangin.
\newblock Classification from cortical folding patterns.
\newblock {\em IEEE Transactions on Medical Imaging}, 26(4):553--565, 2007.

\bibitem{ni:07b}
B.~Thirion, P.~Pinel, S.~M\'eriaux, \underline{A.~Roche}, S.~Dehaene, and J.-B. Poline.
\newblock {Analysis of a Large fMRI cohort: Statistical and Methodological
  Issues for Group Analyses}.
\newblock {\em Neuroimage}, 35(1):105--120, 2007.

\bibitem{tmi:07b}
B.~Thirion, P.~Pinel, A.~Tucholka, \underline{A.~Roche}, P.~Ciuciu, J.-F. Mangin, and J.-B.
  Poline.
\newblock {Structural Analysis of fMRI Data Revisited: Improving the
  Sensitivity and Reliability of fMRI Group Studies}.
\newblock {\em IEEE Transactions on Medical Imaging}, 26(9):1256--1269, 2007.

\bibitem{pnas:06}
G.~Dehaene-Lambertz, L.~Hertz-Pannier, J.~Dubois, S.~M\'eriaux, \underline{A.~Roche}, and
  M.~Sigman.
\newblock Functional organization of perisylvian activation during presentation
  of sentences in preverbal infants.
\newblock {\em Proceedings of the National Academy of Sciences},
  103:14240--14245, 2006.

\bibitem{hbm:06}
S.~M\'eriaux, \underline{A.~Roche}, G.~Dehaene-Lambertz, B.~Thirion, and J.-B. Poline.
\newblock {Combined permutation test and mixed-effect model for group average
  analysis in fMRI}.
\newblock {\em Human Brain Mapping}, 27(5):402--410, 2006.

\bibitem{hbm:06b}
B.~Thirion, G.~Flandin, P.~Pinel, \underline{A.~Roche}, and J.-B. Ciuciu, P.~Poline.
\newblock {Dealing with the shortcomings of spatial normalization:
  Multi-subject parcellation of fMRI datasets}.
\newblock {\em Human Brain Mapping}, 27(8):678--693, August 2006.

\bibitem{pls:05}
J.-B. Poline, A.~Ciuciu, \underline{A.~Roche}, and B.~Thirion.
\newblock Quelle confiance accorder aux images du cerveau en action?
\newblock {\em Pour la Science}, 338:138--142, 2005.

\bibitem{tmi:02}
L.~Freire, \underline{A.~Roche}, and J.-F. Mangin.
\newblock {What is the best similarity measure for motion correction in fMRI
  time series?}
\newblock {\em IEEE Transactions on Medical Imaging}, 21(5):470--484, 2002.

\bibitem{tmi:01}
A.~Guimond, \underline{A.~Roche}, N.~Ayache, and J.~Meunier.
\newblock {Multimodal Brain Warping Using the Demons Algorithm and Adaptative
  Intensity Corrections}.
\newblock {\em IEEE Transactions on Medical Imaging}, 20(1):58--69, 2001.

\bibitem{ivc:00}
S.~Ourselin, \underline{A.~Roche}, G.~Subsol, X.~Pennec, and N.~Ayache.
\newblock {Reconstructing a 3D Structure from Serial Histological Sections}.
\newblock {\em Image and Vision Computing}, 19(1--2):25--31, 2000.

\subsection*{Peer-reviewed conference papers}

\bibitem{miccai:14}
\underline{A.~Roche} and F.~Forbes.
\newblock {Partial volume estimation in brain MRI revisited}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'14)}, volume 8673 of {\em Lecture Notes in Computer Science}, pages
  771--778, Boston, USA, September 2014. Springer Verlag.

\bibitem{miccai:12}
\underline{A.~Roche}.
\newblock {Closed-Form Relaxation for MRF-MAP Tissue Classification Using
  Discrete Laplace Equations}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'12)}, volume 7511 of {\em Lecture Notes in Computer Science}, pages
  355--362, Nice, France, October 2012. Springer Verlag.

\bibitem{isbi:04}
\underline{A.~Roche}, P.-J. Layahe, and J.-B. Poline.
\newblock {Incremental Activation Detection in fMRI Time Series using Kalman
  Filtering}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'04)},
  pages 376--379, Washington, DC, USA, April 2004.

\bibitem{miccai:04}
\underline{A.~Roche}, P.~Pinel, S.~Dehaene, and J.-B. Poline.
\newblock {Solving Incrementally the Fitting and Detection Problems in fMRI
  Time Series}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'04)}, volume 3217 of {\em Lecture Notes in Computer Science}, pages
  719--726, Saint-Malo, France, September 2004. Springer Verlag.

\bibitem{eccv:00}
\underline{A.~Roche}, A.~Guimond, N.~Ayache, and J.~Meunier.
\newblock {Multimodal Elastic Matching of Brain Images}.
\newblock In {\em European Conference on Computer Vision (ECCV'00)}, volume
  1843 of {\em Lecture Notes in Computer Science}, pages 511--527, Dublin,
  Ireland, June 2000. Springer Verlag.

\bibitem{miccai:00}
\underline{A.~Roche}, X.~Pennec, M.~Rudolph, D.~P. Auer, G.~Malandain, S.~Ourselin, L.~M.
  Auer, and N.~Ayache.
\newblock {Generalized Correlation Ratio for Rigid Registration of 3D
  Ultrasound with MR Images}.
\newblock In S.~Delp, A.M. DiGioia, and B.~Jaramaz, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'00)}, volume 1935 of
  {\em Lecture Notes in Computer Science}, pages 567--577, Pittsburgh, USA,
  October 2000. Springer Verlag.

\bibitem{miccai:99}
\underline{A.~Roche}, G.~Malandain, N.~Ayache, and S.~Prima.
\newblock {Towards a Better Comprehension of Similarity Measures used in
  Medical Image Registration}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'99)}, volume 1679 of {\em Lecture Notes in Computer Science}, pages
  555--566, Cambridge, UK, September 1999. Springer Verlag.

\bibitem{miccai:98}
\underline{A.~Roche}, G.~Malandain, X.~Pennec, and N.~Ayache.
\newblock {The Correlation Ratio as a New Similarity Measure for Multimodal
  Image Registration}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'98)}, volume 1496 of {\em Lecture Notes in Computer Science}, pages
  1115--1124, Cambridge, USA, October 1998. Springer Verlag.

\bibitem{miccai:09}
M.~Keller, M.~Lavielle, M.~Perrot, and \underline{A.~Roche}.
\newblock {Anatomically Informed Bayesian Model Selection for fMRI Group Data
  Analysis}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'09)}, volume 5762 of {\em Lecture Notes in Computer Science}, pages
  450--457, London, UK, September 2009. Springer Verlag.

\bibitem{isbi:08}
M.~Keller and \underline{A.~Roche}.
\newblock {Increased Sensitivity in fMRI Group Analyses using Mixed-Effect
  Modeling}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'08)},
  pages 548--551, Paris, France, May 2008.

\bibitem{isbi:08b}
P.~Ciuciu, T.~Vincent, A.-L. Foulque, and \underline{A.~Roche}.
\newblock {Improved fMRI group studies based on spatially varying
  non-parametric BOLD signal modeling}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'08)},
  pages 1263--1266, Paris, France, May 2008.

\bibitem{isbi:07}
P.~Ciuciu, P.~Abry, C.~Rabrait, H.~Wendt, and \underline{A.~Roche}.
\newblock {Leader-Based Multifractal Analysis for Evi fmri Time Series: Ongoing
  vs Task-Related Brain Activity}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'07)},
  pages 404--407, Washington, DC, USA, May 2007.

\bibitem{miccai:01b}
S.~Granger, X.~Pennec, and \underline{A.~Roche}.
\newblock {Rigid Point-Surface Registration using an EM Variant of ICP for
  Computed Guided Oral Implantology}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'01)}, volume 2208 of
  {\em Lecture Notes in Computer Science}, pages 752--761, Utrecht,
  Netherlands, October 2001. Springer Verlag.

\bibitem{miccai:08}
M.~Bhattacharjee, A.~Pitiot, \underline{A.~Roche}, D.~Dormont, and E.~Bardinet.
\newblock {Anatomy-Preserving Nonlinear Registration of Deep Brain ROIs using
  Confidence-Based Block-Matching}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'08)}, Lecture Notes in Computer Science, New York, USA, September
  2008. Springer Verlag.

\bibitem{miccai:07}
C.~Poupon, F.~Poupon, \underline{A.~Roche}, Y.~Cointepas, J.~Dubois, and J.-F. Mangin.
\newblock {Real-Time MR Diffusion Tensor and Q-Ball Imaging Using Kalman
  Filtering}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'07)}, volume 4791 of {\em Lecture Notes in Computer Science}, pages
  27--35, Brisbane, Australia, November 2007. Springer Verlag.

\bibitem{ipmi:07}
B.~Thirion, A.~Tucholka, M.~Keller, P.~Pinel, \underline{A.~Roche}, J.-F. Mangin, and J.-B.
  Poline.
\newblock {High Level Group Analysis of fMRI Data Based on Dirichlet Process
  Mixture Models}.
\newblock In {\em Information Processing in Medical Imaging (IPMI'07)}, Lecture
  Notes in Computer Science, pages 482--494, Kerkrade, Netherlands, 2007.
  Springer Verlag.

\bibitem{isbi:06}
S.~M\'eriaux, \underline{A.~Roche}, B.~Thirion, and G.~Dehaene-Lambertz.
\newblock Robust statistics for nonparametric group analysis in {fMRI}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'06)},
  pages 936--939, Washington, DC, USA, April 2006.

\bibitem{mmbia:06}
B.~Thirion, \underline{A.~Roche}, P.~Ciuciu, and J.-B. Poline.
\newblock {Improving sensitivity and reliability of fMRI group studies through
  high level combination of individual subjects results}.
\newblock In {\em Mathematical Methods in Biomedical Image Analysis
  (MMBIA'06)}, page~62, New York, USA, June 2006.

\bibitem{isbi:04b}
P.~Ciuciu, J.~Idier, \underline{A.~Roche}, and C.~Pallier.
\newblock Outlier detection for robust region-based estimation of the
  hemodynamic response function in event-related f{MRI}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'04)},
  pages 392--395, Washington, DC, USA, April 2004.

\bibitem{isbi:04c}
E.~Duchesnay, \underline{A.~Roche}, D.~Rivi\`ere, D.~Papadopoulos-Orfanos, Y.~Cointepas,
  and J.-F. Mangin.
\newblock Population classification based on structural morphometry of cortical
  sulci.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'04)},
  pages 1276--1279, Washington, DC, USA, 2004.

\bibitem{miccai:02}
E.~Bardinet, P.~Cathier, \underline{A.~Roche}, and D.~Dormont.
\newblock {A Posteriori Validation of Pre-Operative Planning in Functional
  Neurosurgery By Quantification of Brain Pneumocephalus}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'02)}, volume 2488(1) of
  {\em Lecture Notes in Computer Science}, pages 323--330, Tokyo, Japan,
  November 2002. Springer Verlag.

\bibitem{miccai:01}
E.~Bardinet, A.C.F. Colchester, \underline{A.~Roche}, Y.~Zhu, Y.~He, S.~Ourselin,
  B.~Nailon, S.A. Hojjat, J.~Ironside, S.~Al-Sarraj, N.~Ayache, and J.~Wardlaw.
\newblock {Registration of Reconstructed Post Mortem Optical Data with MR Scans
  of the Same Patient}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'01)}, volume 2208 of
  {\em Lecture Notes in Computer Science}, pages 957--965, Utrecht,
  Netherlands, October 2001. Springer Verlag.

\bibitem{miccai:01d}
H.~Delingette, E.~Bardinet, D.~Rey, J.-D. Lemarechal, J.~Montagnat,
  S.~Ourselin, \underline{A.~Roche}, D.~Dormont, J.~Yelnik, and N.~Ayache.
\newblock {YAV++: a Software Platform for Medical Image Processing and
  Visualization}.
\newblock In {\em Workshop on Interactive Medical Image Visualization and
  Analysis, satellite symposia of MICCAI'01}, Utrecht, Netherlands, October
  2001.

\bibitem{miccai:01c}
S.~Ourselin, E.~Bardinet, D.~Dormont, G.~Malandain, \underline{A.~Roche}, and N.~Ayache.
\newblock {Fusion of Histological Sections and MR Images: Towards the
  Construction of an Atlas of the Human Basal Ganglia}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'01)}, volume 2208 of
  {\em Lecture Notes in Computer Science}, pages 743--751, Utrecht,
  Netherlands, October 2001. Springer Verlag.

\bibitem{miar:01}
X.~Pennec, N.~Ayache, \underline{A.~Roche}, and P.~Cachier.
\newblock {Non-rigid MR/US registration for tracking brain deformations}.
\newblock In {\em International Workshop on Medical Imaging and Augmented
  Reality (MIAR'01)}, pages 79--86, Shatin, Hong Kong, June 2001. IEEE Computer
  Society Press.

\bibitem{miccai:00c}
A.~Colchester, S.~Ourselin, Y.~Zhu, E.~Bardinet, Y.~He, \underline{A.~Roche}, S.~Al-Sarraj,
  B.~Nailon, J.~Ironside, and N.~Ayache.
\newblock {3-D Reconstruction of Macroscopic Optical Brain Slice Images}.
\newblock In S.~Delp, A.M. DiGioia, and B.~Jaramaz, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'00)}, volume 1935 of
  {\em Lecture Notes in Computer Science}, pages 95--105, Pittsburgh, USA,
  October 2000. Springer Verlag.

\bibitem{paris:00}
O.~Migneco, G.~Malandain, \underline{A.~Roche}, I.~Dygai, F.~Bussiere, N.~Ayache, and
  J.~Darcourt.
\newblock Absolute calibration of hmpao spect using xenon 133.
\newblock In {\em Congr\`es Europ\'een de M\'edecine Nucl\'eaire}, Paris,
  France, September 2000.

\bibitem{miccai:00b}
S.~Ourselin, \underline{A.~Roche}, S.~Prima, and N.~Ayache.
\newblock {Block Matching: a General Framework to Improve Robustness of Rigid
  Registration of Medical Images}.
\newblock In S.~Delp, A.M. DiGioia, and B.~Jaramaz, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'00)}, volume 1935 of
  {\em Lecture Notes in Computer Science}, pages 557--566, Pittsburgh, USA,
  October 2000. Springer Verlag.

\bibitem{wbir:99}
S.~Ourselin, \underline{A.~Roche}, G.~Subsol, and X.~Pennec.
\newblock {Automatic Alignment of Histological Sections}.
\newblock In F.~Pernus, S.~Kovacic, H.S. Stiehl, and M.A. Viergever, editors,
  {\em International Workshop on Biomedical Image Registration (WBIR'99)},
  pages 1--13, Bled, Slovenia, August 1999.

\subsection*{Abstracts}

\bibitem{adpd:17}
\underline{A.~Roche}, D.~Damian, F.~Pedron, B.~Mar\'echal, P.~Hagmann, P.~Maeder, R.~Meuli,
  T.~Kober, and J.-F. D\'emonet.
\newblock {Automated prediction of typical and mixed Alzheimer's disease using
  combined routine brain volumetry and cognitive assessment}.
\newblock In {\em Submitted to AD-PD conference}, Vienna, 2017.

\bibitem{ismrm:14}
\underline{A.~Roche}, D.~Schmitter, B.~Mar\'echal, D.~Ribes, A.~Abdulkadir, M.~Bach~Cuadra,
  A.~Daducci, C.~Granziera, S.~Kl\"oppel, P.~Maeder, R.~Meuli, and G.~Krueger.
\newblock {Volume-based vs. voxel-based brain morphometry in Alzheimer's
  disease prediction}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'14)}, Milan, Italy, 2014.
\newblock The first two authors contributed equally.

\bibitem{ohbm:05}
\underline{A.~Roche}.
\newblock {When is it a good time to smooth fMRI data?}
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'05) CD-Rom Neuroimage 26(1)}, Toronto, Canada, June 2005.

\bibitem{ohbm:04}
\underline{A.~Roche} and J.-B. Poline.
\newblock {Kalman Filtering for Real-Time fMRI Analysis}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{ohbm:03}
\underline{A.~Roche}, F.~Kherif, G.~Flandin, and J.-B. Poline.
\newblock Should {fMRI} data be analyzed using a single {BOLD} response model
  across regions?
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'03) CD-Rom Neuroimage 19(2)}, New York, USA, June 2003.

\bibitem{ismrm:16}
P.~Falkovskiy, B.~Mar\'echal, T.~Kober, P.~Maeder, R.~Meuli, J.-P. Thiran, and
  \underline{A.~Roche}.
\newblock {Impact of image acquisition systems on Alzheimer's disease-related
  atrophy detection}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:16b}
P.~Falkovskiy, B.~Mar\'echal, S.~Yan, Z.~Jin, T.~Qian, K.~O'Brien, R.~Meuli,
  J.-P. Thiran, G.~Krueger, T.~Kober, and \underline{A.~Roche}.
\newblock {Quantitative comparison of MP2RAGE skull-stripping strategies}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ohbm:16}
P.~Falkovskiy, B.~Mar\'echal, P.~Maeder, T.~Kober, J.-P. Thiran, R.~Meuli, and
  \underline{A.~Roche}.
\newblock {Correcting Effects of Magnetic Resonance Field Strength on Brain
  Volumetry}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'16)}, Geneva, Switzerland, 2016.

\bibitem{ismrm:15}
G.~Bonnier, J.-P Thiran, G.~Krueger, C.~Granziera, and \underline{A.~Roche}.
\newblock {Concentration maps improve detection of gray matter alteration in
  cerebellum and deep gray matter structures}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'15)}, Toronto, Canada, 2015.

\bibitem{ismrm:13}
K.~O'Brien, G.~Krueger, F.~Lazeyras, R.~Grueter, and \underline{A.~Roche}.
\newblock {A simple method to denoise MP2RAGE}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'13)}, Salt Lake City, USA, 2013.

\bibitem{ismrm:11}
D.~Ribes, B.~Mortamet, M.~Bach~Cuadra, C.R. Jack, R.~Meuli, G.~Krueger, and
  \underline{A.~Roche}.
\newblock {Comparison of tissue classification models for automatic brain MR
  segmentation}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'11)}, Montreal, Canada, 2011.

\bibitem{ohbm:09}
L.~Favre, A.-L. Fouque, T.~Vincent, A.~Tucholka, M.~Keller, G.~Operto,
  B.~Thyreau, C.~Clouchoux, L.~Risser, A.~Moreno, D.~Geffroy, Y.~Cointepas,
  O.~Coulon, P.~Ciuciu, B.~Thirion, and \underline{A.~Roche}.
\newblock {A Comprehensive fMRI Processing Toolbox for BrainVISA}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'09)}, San Francisco, USA, 2009.

\bibitem{ismrm:16c}
S.~Haller, P.~Falkovskiy, R.~Meuli, J.-P. Thiran, G.~Krueger, K.-O. Lovblad,
  \underline{A.~Roche}, T.~Kober, and B.~Mar\'echal.
\newblock {Basic MR sequence parameters systematically bias automated brain
  volume estimation}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:16d}
M.J. Fartaria, G.~Bonnier, T.~Kober, \underline{A.~Roche}, D.~Rotzinger, M.~Schluep,
  R.~Du~Pasquier, J.-P. Thiran, G.~Krueger, R.~Meuli, M.~Bach~Cuadra, and
  C.~Granziera.
\newblock {Longitudinal automated detection of white matter and cortical
  lesions in relapsing-remitting multiple sclerosis}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:16e}
M.J. Fartaria, G.~Bonnier, T.~Kober, K.~O'Brien, \underline{A.~Roche}, B.~Mar\'echal,
  D.~Rotzinger, R.~Meuli, J.-P. Thiran, G.~Krueger, M.~Bach~Cuadra, and
  C.~Granziera.
\newblock {Focal cerebellar pathology in early relapsing-remitting multiple
  sclerosis patients: a MP2RAGE study at 3T and 7T MRI }.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:15b}
M.J. Fartaria, G.~Bonnier, \underline{A.~Roche}, T.~Kober, R.~Meuli, D.~Rotinger,
  M.~Schluep, R.~Du~Pasquier, J.-P. Thiran, G.~Krueger, C.~Granziera, and
  M.~Bach~Cuadra.
\newblock {Exploration of advanced MR imaging contrasts for automated detection
  of white matter and cortical lesions in early-stages of multiple sclerosis}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'15)}, Toronto, Canada, 2015.

\bibitem{ismrm:15c}
Y.~Wang, B.~Mar\'echal, D.~Neumann, \underline{A.~Roche}, J.D. West, B.C. McDonald, M.A.
  Keiski, D.J. Smith, A.J. Saykin, and G.~Krueger.
\newblock {Detecting atrophy in chronic moderate and severe traumatic brain
  injury using an automated volume-based morphometry toolbox}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'15)}, Toronto, Canada, 2015.

\bibitem{ismrm:14b}
D.~Romascano, D.E. Meskaldji, G.~Bonnier, S.~Simioni, D.~Rotzinger, Y.-C. Lin,
  G.~Menegaz, \underline{A.~Roche}, M.~Schluep, and R.~Du~Pasquier.
\newblock {Cerebellar Connectomics Provide New Biomarkers in Early Multiple
  Sclerosis}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'14)}, Milan, Italy, 2014.

\bibitem{ismrm:14c}
G.~Bonnier, \underline{A.~Roche}, D.~Romascano, S.~Simioni, D.E. Meskaldji, D.~Rotzinger,
  Y.-C. Lin, G.~Menegaz, M.~Schluep, and R.~Du~Pasquier.
\newblock {Multiple Sclerosis Lesion Fingerprint Using Multicontrast MRI}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'14)}, Milan, Italy, 2014.

\bibitem{ismrm:13b}
M.~Bach~Cuadra, S.~Gelin, \underline{A.~Roche}, O.~Esteban, J.~Kober, T.~Marques,
  C.~Granziera, and G.~Krueger.
\newblock {Classical segmentation methods on novel MR imaging: a study of brain
  tissue segmentation of MP2RAGE vs MPRAGE}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'13)}, Salt Lake City, USA, 2013.

\bibitem{ismrm:13c}
G.~Bonnier, T.~Stumpf, D.~Romascano, \underline{A.~Roche}, M.~Schluep, R.~Dupasquier,
  C.~Granziera, and G.~Krueger.
\newblock {Ultrafast T2 mapping of multiple sclerosis pathology in early
  disease}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'13)}, Salt Lake City, USA, 2013.

\bibitem{ismrm:12}
T.~Kober, \underline{A.~Roche}, D.~Ribes, R.~Meuli, and G.~Krueger.
\newblock {Atlas-free brain tissue segmentation using a single T1-weighted MRI
  acquisition}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'12)}, Melbourne, Australia, 2012.

\bibitem{ismrm:12b}
B.~Mar\'echal, T.~Kober, T.~Hilbert, D.~Ribes, N.~Chevrey, \underline{A.~Roche}, J.-P.
  Thiran, R.~Meuli, and G.~Krueger.
\newblock {Automated quality control in MR-based brain morphometry}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'12)}, Melbourne, Australia, 2012.

\bibitem{ohbm:09b}
M.~Brett, J.~Taylor, C.~Burns, K.J. Millman, F.~Perez, \underline{A.~Roche}, B.~Thirion,
  and M.~D'Esposito.
\newblock {NIPY: an open library and development framework for FMRI data
  analysis}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'09) Neuroimage 47:S196}, San Francisco, USA, 2009.

\bibitem{ismrm:08}
F.~Poupon, \underline{A.~Roche}, J.-F. Mangin, and C.~Poupon.
\newblock Real-time {MR} diffusion tensor and {Q-ball} imaging using {Kalman}
  filtering.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'08)}, Toronto, Canada, 2008.

\bibitem{ohbm:06}
S.~M\'eriaux, \underline{A.~Roche}, G.~Dehaene-Lambertz, and J.-B. Poline.
\newblock When do mixed-effect models fail to improve detection sensitivity in
  f{MRI} group activation maps?
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'06) CD-Rom Neuroimage 31(1)}, Florence, Italy, 2006.

\bibitem{ohbm:05b}
S.~M\'eriaux, \underline{A.~Roche}, J.~Catry, O.~Chauvin, and J.-B. Poline.
\newblock Robust statistics for f{MRI} group analyses.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'05) CD-Rom Neuroimage 26(1)}, Toronto, Canada, June 2005.

\bibitem{ohbm:04c}
E.~Duchesnay, \underline{A.~Roche}, D.~Rivi\`ere, D.~Papadopoulos-Orfanos, Y.~Cointepas,
  and J.-F. Mangin.
\newblock Guessing the sex from the shapes of cortical folds.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{ohbm:04d}
G.~Flandin, X.~Pennec, \underline{A.~Roche}, W.~Penny, N.~Ayache, and J.-B. Poline.
\newblock Multi-subject anatomo-functional classification for activation
  studies.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{ohbm:04b}
S.~M\'eriaux, F.~Kherif, \underline{A.~Roche}, M.~Brett, L.~Garnero, and J.-B. Poline.
\newblock How frequently do we sample inhomogeneous group of subjects in f{MRI}
  studies?
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{cancero:99}
P.~Y. Bondiau, G.~Malandain, K.~Benezery, \underline{A.~Roche}, R.~Ferrand, D.~Ponvert,
  J.~L. Habrand, and J.~N. Bruneton.
\newblock {Fusions Scanner / IRM entri\`erement automatique: perspectives
  d'utilisation pour la radioth\'erapie}.
\newblock In {\em {Proc. du XIXe Forum de Canc\'erologie}}, Paris, June 1999.

\bibitem{rsna:99}
P.~Y. Bondiau, G.~Malandain, K.~Benezery, \underline{A.~Roche}, R.~Ferrand, D.~Ponvert,
  J.~L. Habrand, and J.~N. Bruneton.
\newblock {Automatic CT Scan and MRI Matching: Use for Brain Tumor
  Radiotherapy}.
\newblock In {\em {RSNA'99}}, Chicago, USA, November 1999. Published in
  Supplement to Radiology, 213:236.

\bibitem{dijon:99}
S.~Ourselin, C.~Sattonnet, \underline{A.~Roche}, and G.~Subsol.
\newblock {Automatic alignment of histological sections for 3D reconstruction
  and analysis}.
\newblock In {\em 6\`eme congr\`es annuel de l'Association Fran\c caise de
  Cytom\'etrie}, Dijon, France, October 1999. Published in the Analytical
  Cellular Pathology Journal, 19(3):123.

\subsection*{Preprints}

\bibitem{rr:15}
\underline{A.~Roche}.
\newblock {Composite Bayesian inference}.
\newblock Technical report, 2015.
\newblock Available at arxiv.org (arXiv:1512.07678).

\bibitem{rr:12}
\underline{A.~Roche}.
\newblock {EM algorithm and variants: an informal tutorial}.
\newblock Technical report, 2012.
\newblock Available at arxiv.org (arXiv:1105.1476).

\bibitem{rr:00}
\underline{A.~Roche}, X.~Pennec, G.~Malandain, N.~Ayache, and S.~Ourselin.
\newblock {Generalized Correlation Ratio for Rigid Registration of 3D
  Ultrasound with MR Images}.
\newblock Technical Report 3980, INRIA, July 2000.

\bibitem{rr:99}
\underline{A.~Roche}, G.~Malandain, and N.~Ayache.
\newblock {Unifying Maximum Likelihood Approaches in Medical Image
  Registration}.
\newblock Technical Report 3741, INRIA, July 1999.

\bibitem{rr:98}
\underline{A.~Roche}, G.~Malandain, N.~Ayache, and X.~Pennec.
\newblock {Multimodal Image Registration by Maximization of the Correlation
  Ratio}.
\newblock Technical Report 3378, INRIA, August 1998.

\bibitem{rr:01}
S.~Granger, X.~Pennec, and \underline{A.~Roche}.
\newblock {Rigid Point-Surface Registration using Oriented Points and an EM
  Variant of ICP for Computer Guided Oral Implantology}.
\newblock Technical Report 4169, INRIA, April 2001.

\bibitem{rr:99b}
A.~Guimond, \underline{A.~Roche}, N.~Ayache, and J.~Meunier.
\newblock {Multimodal Brain Warping Using the Demons Algorithm and Adaptative
  Intensity Corrections}.
\newblock Technical Report 3796, INRIA, November 1999.

\bibitem{rr:98b}
S.~Ourselin, \underline{A.~Roche}, G.~Subsol, X.~Pennec, and C.~Sattonet.
\newblock {Automatic Alignment of Histological Sections for 3D Reconstruction
  and Analysis}.
\newblock Technical Report 3595, INRIA, December 1998.

\subsection*{Patents}

\bibitem{patentapp:12}
\underline{A.~Roche}, T.~Kober, and G.~Krueger.
\newblock Automated atlas-free brain tissue classification using specific
  magnetic resonance imaging sequences, 2013.
\newblock Owner: Siemens Schweiz AG, Switzerland. US patent application
  13/855972.

\bibitem{patent:05}
\underline{A.~Roche}, M.~Brady, J.~Declerck, and O.~Corolleur.
\newblock Similarity measures.
\newblock United States patent US2005152617, July 2005.
\newblock Owner: Mirada Solutions Ltd, UK. Also published as EP1513104
  (Europe), JP2005169077 (Japan).

\bibitem{patent:03}
\underline{A.~Roche}, N.~Ayache, G.~Malandain, and X.~Pennec.
\newblock Electronic device for automatic registration of images.
\newblock United States patent US6539127, March 2003.
\newblock Owner: INRIA, France. Also published as EP0977148 (Europe), FR2781906
  (France).

\bibitem{patent:15}
K.~O'Brien and \underline{A.~Roche}.
\newblock {A simple method to denoise MP2RAGE}.
\newblock United States patent US2013P01055, 2013.
\newblock Owner: Siemens Schweiz AG, Switzerland. Also published as EP 2765437.

\bibitem{patent:12}
C.~Poupon, F.~Poupon, \underline{A.~Roche}, and J.-F. Mangin.
\newblock Real-time magnetic resonance diffusion imaging.
\newblock United States patent US8330462, 2007.
\newblock Owner: CEA, France. Also published as EP2203758A1 (Europe).

\subsection*{Theses}

\bibitem{these:01}
\underline{A.~Roche}.
\newblock {\em Recalage d'images m\'edicales par inf\'erence statistique}.
\newblock PhD thesis, Universit\'e de Nice-Sophia Antipolis, February 2001.

\bibitem{dea:96}
\underline{A.~Roche}.
\newblock {Mod\'elisation de la perception du tempo par oscillateurs
  adaptatifs}.
\newblock Master's thesis, {Universit\'e Pierre et Marie Curie, Paris VI},
  1996.

\end{thebibliography}

### 2019-01-11 cv/cv.tex
Update composite inference

\documentclass[english,10pt,a4paper]{scrartcl}

\usepackage{times}
\usepackage{fullpage}
\usepackage{hyperref} 
\usepackage{url}
\usepackage{tabularx}
\usepackage{multicol}
\usepackage{eurosym}
\usepackage{color}

\title{Alexis Roche}

\author{}
\date{}

\makeatletter
\newcommand{\matphi}{\boldsymbol{\Phi}}
\newcommand{\req}[1]{Equation~(\ref{#1})}
\newcommand*{\rom}[1]{\expandafter\@slowromancap\romannumeral #1@}

\newcommand{\heading}[1]{\vspace*{.75cm}\noindent{\Large{#1}}\vspace*{-.25cm}\newline\noindent\rule{\textwidth}{0.8pt}\vspace*{.25cm}}

\newcommand{\bighead}[3]{\noindent\begin{minipage}{.85\textwidth}{\bf #1}\newline {\bf #2}\end{minipage}\begin{minipage}{.15\textwidth}\begin{flushright}{\bf #3}\end{flushright}\end{minipage}\newline\noindent\rule{.85\textwidth}{0.4pt}}

\newcommand{\head}[2]{{\noindent {\bf #1} $\mid$ {\bf #2}}}

\makeatother

\begin{document}

%\maketitle

\begin{flushright} 
{\huge {\bf Alexis Roche}}\\
\begin{small}
+351 910 327 756 $\mid$ \url{alexis.roche@gmail.com}\\
\url{https://www.linkedin.com/in/alexis-roche-3815592}\\
French National $\mid$ DOB 12/8/1972 $\mid$ Married, 2 Children\\
\end{small}
\end{flushright}

\heading{Professional Profile}

\noindent Senior-level engineer and applied researcher with a specialist expertise in computer vision, medical image analysis and machine learning, operating in the fields of applied research, health industry and domestic appliance industry. Graduated from the French ``Grande \'Ecole" CentraleSupelec, and holder of a PhD. Strong experience in developing research and commercial software.

\noindent {\em Keywords:} image processing, statistics, machine learning, structural and functional brain magnetic resonance imaging, Python.

\heading{Key Skills}

\noindent {\bf DATA SCIENCE:} In-depth knowledge of statistics, computer vision, medical image analysis, traditional machine learning and deep learning.

\ \\
\noindent {\bf NEUROIMAGING:} Extensive experience in developing algorithms for structural and functional brain magnetic resonance imaging (MRI) in the form of open-source and commercial software for neuroscience, clinical research \& clinical practice.

\ \\
\noindent {\bf SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.

\ \\
\noindent {\bf INTERACTION:} Proven ability to work in multidisciplinary environments at the interface of industry and academic research, as well as to develop successful collaborations with physicians and fundamental life science researchers.

\ \\
\noindent {\bf SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 international patents.

\heading{Career Summary}

\bighead{SENIOR COMPUTER VISION SCIENTIST}{CoVii, Ar\c celik/Beko group, Porto, Portugal}{2017--present}
\begin{itemize}
\item Developing object recognition and tracking algorithms for smart domestic appliances (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017).
\end{itemize}

\ \\
\bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
\begin{itemize}
\item Led algorithmic development of the {\em MorphoBox} Siemens prototype for brain morphometry using anatomical MRI to help radiological reading for patients with suspected neurodegeneration.
\item Contributed a number of clinical validation studies of brain morphometry in collaboration with neurologists and radiologists, via statistical analysis and data mining techniques, in Alzheimer's disease and other dementia, mild cognitive impairment, multiple sclerosis, migraine, anorexia, \ldots 
\item Co-supervised three PhD students and two MSc students.
\item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
\end{itemize}

\ \\
\bighead{PERMANENT RESEARCHER -- NEUROIMAGING INSTITUTE}{French Atomic Commission (CEA), Paris, France}{2002--2011}
\begin{itemize}
\item Developed advanced algorithms for image processing and statistical analysis of functional, anatomical and diffusion-weighted MRI data (population analysis, spatio-temporal image registration, real-time imaging).
\item French ANR-funded project management: PI (Karametria project, statistical analysis of brain structures, budget: 620 K\euro, 2009--2011), team leader (NIBB project, studying language in infants via functional neuroimaging, 2006--2009).
\item Active contributor to the NiPy software library (Neuroimaging in Python, \url{www.nipy.org}) from 2006.
\item Main supervisor of two PhD students and four MSc students.
\item Worked as Academic Guest at Computer Vision Laboratory, Swiss Federal Institute of Technology Zurich (ETHZ), Switzerland, 2009--2011, on brain image registration. 
\end{itemize}

\ \\
\bighead{RESEARCH ASSISTANT -- WOLFSON MEDICAL VISION LABORATORY}{University of Oxford, UK}{2001--2002}
\begin{itemize}
\item Developed advanced medical image registration algorithms.
\item Consulting work for Mirada Solutions Ltd (now Siemens Molecular Imaging) on image-based deformation tracking (brain, liver).
\end{itemize}

\ \\
\bighead{CONTINGENT SCIENTIST -- FRENCH NATIONAL SERVICE}{General Directorate for Armament (DGA), Vernon, France}{1996--1997}
\begin{itemize}
\item Developed plane trajectory simulator using stochastic process models.
\end{itemize}

\heading{Education}

\head{PhD, Engineering Science (1997--2001)}{University of Nice-Sophia Antipolis, France}
\begin{itemize}
\item Funded by French National Research Institute (INRIA), Sophia Antipolis unit.
\item Developed generic algorithm for multimodal image registration, with applications in radiotherapy, image-guided surgery and neuroscience.
\end{itemize}

\ \\
\head{MSc, Cognitive Science (1995--1996)}{University Pierre \& Marie Curie, Paris \rom{6}, France}
\begin{itemize}
\item Internship at Experimental Psychology Laboratory, National Center for Scientific Research (CNRS), Paris, France, on modeling human perception of tempo using artificial neural networks.
\end{itemize}

\ \\
\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris (now CentraleSupelec), France}
\begin{itemize}
\item Third year specialization in Applied Mathematics.
\end{itemize}

%%\item Other training: Neuronal dynamics open course, EPFLx, Switzerland (2014).

\heading{Additional information}

\noindent 
\begin{minipage}{.15\textwidth}
{\bf Languages:}
\end{minipage}
\begin{minipage}{.85\textwidth}
Native French, fluent English, basic Italian \& Portuguese. %German.
\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
{\bf Programming:}
\ \\ 
\ \\
\ \\ 
\ \\
\ \\ 
\end{minipage}
\begin{minipage}{.85\textwidth}
{\tiny$\bullet$} C/C++ \\
{\tiny$\bullet$} Python (numpy, scipy, pylab, sklearn, TensorFlow, keras, nipy, cython)\\ 
{\tiny$\bullet$} Matlab\\
{\tiny$\bullet$} R. \\
{\tiny$\bullet$} Version control software (git, svn).\\
\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
{\bf Publications:}
\ \\ 
\ \\
\  \\
\end{minipage}
\begin{minipage}{.85\textwidth}
{\tiny$\bullet$} Over 100 scholarly publications. \\
{\tiny$\bullet$} Google scholar statistics: 9146 citations, h-index: 31, i10-index: 57 (December 2018).\\
{\tiny$\bullet$} International patents: 4 published + 1 pending application \\
\end{minipage}

%\ \\
%\noindent 
%\begin{minipage}{.15\textwidth}
%{\bf Reviewing:}
%\ \\ 
%\ \\
%\end{minipage}
%\begin{minipage}{.85\textwidth}
%IEEE Transactions on Medical Imaging, Medical Image Analysis, NeuroImage, Frontiers in
% Neuroscience, IEEE Transactions on Image Processing, Signal Image and Video Processing, MICCAI conference, \ldots
%\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
{\bf Teaching:}
\ \\ 
\ \\
\  \\
\end{minipage}
\begin{minipage}{.85\textwidth}
{\tiny$\bullet$} Brain morphometry post-graduate lecture (2h), Lausanne: UNIL (2014/16), EPFL (2015/16).\\
{\tiny$\bullet$} Statistical analysis in neuroimaging, INSERM/CNRS continuing education, France (10h/year from 2003 to 2015).\\
\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
{\bf References:}
\end{minipage}
\begin{minipage}{.85\textwidth}
Available on request.
\end{minipage}

%\input{biblio.tex}

\end{document}

### 2019-01-11 notes/composite_inference.tex
Update composite inference

@@
 
+\section{Composite EM algorithm for unsupervised learning}
+\label{sec:gem}
+
+In a nutshell: alternate unsupervised learning of partial model parameters ($M$-step) and re-estimation of label posteriors via maxent ($E$-step).
+
+\begin{itemize}
+\item $CM$-step: given a posterior estimate $q(x|y)$, do just as if the features were independent and solve: 
+$$
+\max_{\theta,\pi}
+\int \sum_i h(y)q(x|y) \log p_\theta(z_i|x)\pi(x) dx dy
+$$
+Assuming that the feature conditional distributions have non-overlapping parameters, this boils down to independent ML-type parameter estimations.
+In addition, the empirical proportions $\pi(x)$ of the unobserved labels are estimated via:
+$$
+\pi(x) = \int h(y)q(x|y) dy
+$$
+\item $CE$-step: given the generative parameters $\theta$ and the reference $\pi$, evaluate the average log-likelihoods:
+$$
+\bar{\ell}_i
+= \int \pi(x)p_\theta(z_i|x)\log p_\theta(z_i|x) dx dy
+$$
+and update $q(x|y)$ according to the conditional maxent principle:
+$$
+\min_q
+\int h(y)q(x|y) \log \frac{q(x|y)}{\pi(x)} dx dy
+$$
+subject to:
+$$
+\int h(y)q(x|y)\log p_\theta(z_i|x) dx dy = \bar{\ell}_i
+$$
+\end{itemize}
+
+A question arises: how is $h(y)$ defined in both the $M$- and $E$-steps?
+One obvious possibility is to take it as the empirical distribution of the data, meaning that the empirical distribution of the labels plays the role of the ``prior'', as implicitely assumed above, but is modified on each iteration.
+
+Another option is to impose the prior $\pi(x)$ and hold it fixed throughout iterations by re-defining the joint empirical distribution before each $M$-step depending on the current posterior estimate via:
+$$
+h(x,y)=\pi(x)\frac{h_0(y)q(x|y)}{\int h_0(y)q(x|y) dy} ,
+$$
+yielding the following correction to the empirical data distribution:
+$$
+h(y) = \sum_i w_i\delta(y-y_i),
+\qquad
+w_i = \int \pi(x)\frac{q(x|y_i)}{\sum_j q(x|y_j)} dx
+.
+$$
+
+Open question: are both methods equivalent in the case of the standard EM?
+I speculate that the answer is yes.
+But they are not in the more general case of multiple features, so we have to make a choice.
+
+The maxent problem involved in the $CE$-step is {\em exactly} the same as in the supervised case whether or not the data points are reweighted.
+We may see the $CE$-step as a surrogate for the actactual $E$-step, which can't be implemented because we lack a full generative model.
+
+Are we minimizing a unique functional as in the variational EM algorithm?
+Possibly not.
+We may see this generalization as a two-player game.
+One player tries to predict several data features independently.
+The other player tries to predict labels.
+Together, they find patterns (clusters) in unlabeled data.
+Does the more general version converge?
+Does the game have a value?
+And so on.
+
+In the case of a single feature, is this the well-known EM?
+The $CM$-step is clearly the same as the $M$-step, what about the $E$-step?
+It is the same if $h(y)p_\theta(x|y)=p_\theta(y|x)\pi(x)$.
+However, this condition won't hold if $h(y)$ is an empirical distribution...
+In this special case, the ``E player'' could use the data distribution model produced by the ``M player'' instead of the empirical one but, more generally, there is no data distribution model so he has to stick to the empirical one.
+
+The key question to be answered is: how is this composite EM trick different from a plain mixture model EM assuming independent features?
+Basically, what does the $CE$-step bring that the $E$-step doesn't?
+
+% Accounting here for Xi'An comment on xianblog.wordpress.com
+% The sum of the powers is constrained to be equal to one, even though
+% I do not understand why the dimensions of the projections play nog
+% role in this constraint. Simplicity is advanced as an argument,
+% which sounds rather weak…
+%This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $x$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.
+
 \section{Blablabla...}

### 2019-01-11 papers/composite_inference/draft.tex
Update composite inference

--- 2018-11-22 papers/composite_inference/draft.tex
@@
 \begin{abstract}
-This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes conditional entropy relative to a given prior and subject to multiple mean log-likelihood inequality constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
-We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, which can be powerful in shallow learning and transfer learning problems.
+This note revisits the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
+Our view makes it natural to use a machine learning approach to calibrate composite likelihood, that is, tune the associated weights.
+For instance, the weights can be set to minimize cross-entropy on a labeled training dataset, as is a typical supervised learning strategy, yielding an easy-to-solve convex problem.
+We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to some artificial intelligence tasks.
 \end{abstract}
+
+% which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes conditional entropy relative to a given prior and subject to multiple mean log-likelihood inequality constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
+{\color{red} Confusing: we simply maximize cross-entropy, which turns out to have a conditional maxent interpretation.} We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, which can be powerful in shallow learning and transfer learning problems.
+{\color{red} Rephrase this: extends Bayes and keeps introspection ability, which is crucial in some AI problems.
 
@@
 Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).
+Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate artificial intelligence (AI) and typically require supervised learning on large training sets, including ``shallow'' learning techniques (maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
 
-A strong limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as {\em marginally independent}, meaning that the data conveys no information about the parameters unless the variable of interest is observed.
+A common criticism of discriminative models is that they are ``black boxes'' in that they can provide a valuable answer to an inference problem, but no explanation to support it.
+This is problematic in application fields such as medical diagnosis, where AI is intended as a tool to assist humans in a decision making process, but decision is not delegated to AI.
+Automated disease predictions should therefore be corroborated by empirical evidence -- in other words, AI systems should be able to justify their opinion just like human experts.
+Unlike discriminative models, simple methods based on range analysis of univariate biomarkers can provide useful clues to clinicians despite being less powerful in terms of prediction performance.
+
+%Decision making is not about pooling opinions, it is about pooling empirical observations and find which hypothesis they point to.
+
+%Need to emphasize why we need generative models.
+Below are some valid reasons: unsupervised learning, potentially better performance on small training datasets...
+But the \#1 reason is that discriminative models are black boxes: they provide an answer but no explanation to support it.
+Truth is, they perform better in supervised big data context in the sense of accuracy, cross-entropy, and so on, no question about that -- but they are mere inference recipes.
+Is that what AI should be?
+Sometimes, yes.
+But there are application fields where this clearly cannot work, for instance medical diagnosis.
+
+%Imagine a doctor telling a patient: ``Sorry, sir, you have Alzheimer's disease because this fancy AI software said so''.
+Unless the said software is 100\% reliable, which is practically impossible, this would obviously be unacceptable.
+Acceptable would be to corroborate the diagnosis with some key quantitative empirical observations: low memory test scores, hippocampal atrophy, etc.
+In other words, inference has to come with some insight.
+An AI system is but one expert relying on hypotheses which may be invalid, as any expert, and is therefore prone to error.
+A typical error source is to train a classifier on a subset of classes from the real world.
+We cannot avoid inference errors but we can safeguard against them by being able to interrogate the system: why do you think what you think?
+To achieve that, the system should be capable of {\em introspection.
+
+%Vapnik: ``one should solve the classification problem directly and never solve a more general problem as an intermediate step''.
+True if the goal is classification only.
+Wrong if introspection is needed because the problem is then more general in nature.
+
+%Human inference does not work in a pre-determined universe of ``causes''.
+We confront theories with reality to evaluate how likely they are, while knowing that none of the theories considered so far may be ``right''.
+In fact, we are not looking for the truth, but for a conving enough theory.
+
+%Composite likelihood seen as a pool of experts comes with a potential introspection mechanism: first sort experts by decreasing influence on the predictive distribution, see how likely the different classes are to the most influential experts.
+
+Another limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as marginally independent, meaning that the data conveys no information about the parameters unless the variable of interest is observed.
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models.
 For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence it is less effective in small training sets \cite{Ng-01}.
-Overall, pure discriminative models are of little use outside the context of big labeled data.
+
+%Overall, pure discriminative models are of little use outside the context of big labeled data.
 
@@
 \caption{Bayesian networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the variable of interest, the data, and the model parameters.
-Note the marginal independence of~$Y$ and~$\theta$ in the discriminative model.}
+Note the marginal independence of data and parameters in the discriminative model.}
 \label{fig:graph_comparison}
@@
 The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole.
-This may be viewed as a {\em divide and conquer} method to approximate the true but intractable likelihood.
+This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood.
 While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
-In this note, CL is interpreted as a probabilistic opinion pool of ``agents'' making use of different pieces of information, or clues, extracted from the input data.
-Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a single clue in the form of a likelihood function of the hidden variables.
-The agent opinions are then aggregated into a probability distribution on the hidden variables analogous to a Bayesian posterior.
+In this note, CL is interpreted as a probabilistic log-linear opinion pool of ``agents'' making use of different pieces of information extracted from the input data.
+Each agent acts as an isolated statistician using a generative model of a single feature to express an opinion in the form of a likelihood function of the hidden variables.
+The agent opinions are then aggregated into a predictive probability distribution analogous to a Bayesian posterior.
 
-We further argue that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum relative conditional entropy.
+The relevance of each feature in the opinion pool is determined by an arbitrary weight used to scale the associated likelihood.
+For classification performance, the weights can be tuned in a typical supervised learning scenario so as to concurrently maximize discriminative likelihood (or, equivalently, minimize cross-entropy), thereby making the method hybrid between generative and discriminative modeling approaches.
+We shall note that this particular training strategy amounts to building a maximum entropy classifier \cite{BergerA-96} that uses the feature-based log-likelihoods as basis functions.
+
+{\color{red} Still, note that the likelihood functions themselves are trained beforehand so this is slightly different from a standard maxent classifier aka logistic regression.
+Double training phase.}
+
+{\color{red} Interpretability of composite likelihood.
+Obvious from the construction.
+Can achieve prediction performance comparable with fully discriminative models while maintaining the interpretability of generative models.}
+
+%We further argue that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum relative conditional entropy.
 This perspective entails a method to optimize the weights of the different agents from training data as in a typical discriminative learning scenario.
@@
 
+From a machine learning perspective, it is natural to tune the weights so as to minimize some prediction error, for example conditional cross-entropy.
+This brings the question: why then do it with ``pre-learned'' likelihood functions rather than with basis functions spanning the same predictive distribution family?
+Because of potential overfitting.
+The whole point of this note is that it may be better to run two learning phases than a single fully blown discriminative learning due to overfitting.
+It is a kind of tradeoff between pure generative and pure discriminative learning.
+
+The ``hybrid'' strategy necessarily achieves a larger cross-entropy than the pure discriminative one on the training dataset, but might generalize better. Should show examples of this.
+
+The rest of the discussion, i.e. the link with maximum entropy, is less important beside showing that the composite likelihood form is somehow optimal given some knowledge.
+
 %Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature.
@@
 
-BTW, we can work with inequality constraints to force $\lambda\geq 0$:
+We could work with inequality constraints to force $\lambda\geq 0$:
 $$
-E[\log p(z_i|x)] \geq c_i \equiv \int p(z_i|x) \log p(z_i|x) dx
+E[\log p(z_i|x)] \geq \bar{\ell}_i \equiv \int h(x)p(z_i|x) \log p(z_i|x) dx dy
 $$
+where $h(x)$ is the marginal distribution of labels.
 
@@
 
-The dual function amounts to the famous cross-entropy.
+The dual function boils down to the familiar notions of cross-entropy, log-likelihood, KL divergence...
+$$
+\psi(\lambda)
+= \int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
+$$
 
-% Accounting here for Xi'An comment on xianblog.wordpress.com
-% The sum of the powers is constrained to be equal to one, even though
-% I do not understand why the dimensions of the projections play nog
-% role in this constraint. Simplicity is advanced as an argument,
-% which sounds rather weak…
-%This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $x$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.
+If we introduce $p_\lambda(x,y)=p_\lambda(x|y)h(y)$, it is even clearer that this is a most classical distribution fitting problem:
+$$
+\psi(\lambda)
+= \int h(x,y) \log \frac{p_\lambda(x,y)}{\pi(x)h(y)} dx dy
+$$
+(we can obviously drop the reference distribution here).
+So the Maxent justifies the composite likelihood concept as a somewhat optimal parametric form but the remainder of the story is most classical.
+
+Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
+We expect to have $\lambda=1$ but it's not necessarily the case.
+It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
+This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
+So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
+
+Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
+The answer is no.
+The weights generally depend on the prior.
+Hence the maxent composite likelihood is prior-dependent.
+This is an important conceptual difference with the classical notion of likelihood.
+
+\subsection{Data weighting}
+
+The distribution $h(x)$ is the marginal of the labels, which is natural to approximate via their empirical distribution.
+We may then have a discrepancy between $h(x)$ and $\pi(x)$, which makes little sense if it is believed that $h(x)$ is a ``good encoder''.
+
+To correct for this, we may adjust the empirical distribution of labels and data so as to match the target prior, $\pi(x)$, leading to weight the data points as a preprocessing step:
+$$
+h(y) = \sum_i w_i \delta(y-y_i),
+\qquad
+w_i = \frac{\pi(x_i)}{h(x_i)}
+$$
+
+Are the two methods equivalent?
+No.
+The feature-based parameter estimates are obviously the same in both cases, but the search space for the predictive distribution is different, hence we can't get the same answer.
+Which one of the two methods is best?
+The one that allows you to choose your own prior makes more sense to me because the distribution of classes in your training dataset may be completely different from the one that you will deal with in practice.
+For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
+
+\subsection{Super composite likelihood}
+\label{sec:super}
+
+What happens if we use mean-values constraints conditional on $x$ instead of jointly averaged over $x$ and $y$?
+This boils down to picking functions of the form $\chi(x-a)\ell_i(x,y)$ and equate their expectation
+
+The answer is that the maxent solution has the form:
+$$
+p_\lambda(x|y) = \pi(x) \prod_i p(z_i|x)^{\lambda_i(x)}
+$$
+
+In other words, the weights become dependent on the variable of interest. This is the super composite likelihood idea in a nutshell.
+
+In 2016, I wrote:
+\begin{quote}
+When chosen for computational simplicity, clues may not only convey limited information at individual level: their informativeness may also be very much hypothesis-dependent.
+Consider, for instance, diagnosing a disease from a routine medical checkup.
+Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovascular disease -- and conversely for, say, blood pressure.
+This motivates a more general setting where clues can be weighted differently depending on hypotheses.
+\end{quote}
+
+However, I was thinking about the unit sum weight version...
+The diagnosis example could be one in which clues are independent, in which case $\lambda_i(x)\equiv 1$ would be optimal, making any super composite trick irrelevant.
+Also see that with hypothesis-dependent weights, we may lose the estimation consistency.
+At the end of the day, the super composite idea looks like a pretty pointless sophistication.
 

### 2019-01-15 papers/composite_inference/draft.tex
Modifs draft CL

@@
 This note revisits the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
-Our view makes it natural to use a machine learning approach to calibrate composite likelihood, that is, tune the associated weights.
-For instance, the weights can be set to minimize cross-entropy on a labeled training dataset, as is a typical supervised learning strategy, yielding an easy-to-solve convex problem.
+This view makes it natural to use a machine learning approach to calibrate the weights associated with composite likelihood.
+For instance, they can be tuned in a typical supervised learning phase so as to minimize cross-entropy on a labeled training dataset, yielding a convex problem.
 We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to some artificial intelligence tasks.
@@
 
-Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate artificial intelligence (AI) and typically require supervised learning on large training sets, including ``shallow'' learning techniques (maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.
+Frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
+Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate artificial intelligence (AI) and typically require supervised learning on large datasets, including ``shallow'' learning techniques (maximum entropy classifiers aka logistic regression \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).
 
-A common criticism of discriminative models is that they are ``black boxes'' in that they can provide a valuable answer to an inference problem, but no explanation to support it.
-This is problematic in application fields such as medical diagnosis, where AI is intended as a tool to assist humans in a decision making process, but decision is not delegated to AI.
-Automated disease predictions should therefore be corroborated by empirical evidence -- in other words, AI systems should be able to justify their opinion just like human experts.
-Unlike discriminative models, simple methods based on range analysis of univariate biomarkers can provide useful clues to clinicians despite being less powerful in terms of prediction performance.
+A common criticism of discriminative models is that they are ``black boxes'' in that they can make valid inferences but provide little or no information to support them \cite{Molnar-18}.
+This is problematic in application fields such as medical diagnosis, where correct inferences are not straightforward to humans and AI is intended as a tool to assist a decision making process.
+Automated disease predictions should therefore be corroborated by individualized findings to be of any clinical value -- in other words, AI systems should be able to justify their opinion just like clinicians.
+Unlike discriminative models, simple methods based on range analysis of univariate biomarkers can provide meaningful clues, and are thus more useful in practice although less powerful in terms of prediction performance.
 
@@
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models.
-For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence it is less effective in small training sets \cite{Ng-01}.
+For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence is less effective in small training sets \cite{Ng-01}.
 
@@
 
-Composite likelihood (CL, see \cite{Varin-11} and the references therein) is a semi-generative approach to statistical inference that extends the familiar notion of likelihood without requiring a full generative model.
+This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a possible compromise between generative and discriminative models.
+Rather than using a purely discriminative model or attempting at a full generative model, we may combine multiple low-dimensional generative models for different pieces of information extracted from the input data.
+Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the variable of interest.
+The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
+This simple idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
+Importantly, predictions made in such a way could be readily interpretable owing to the use of low-dimensional generative models.
+
+It turns out that, when the aggregation method is chosen as a log-linear pool, the consensus distribution is (up to a multiplicative prior) a quantity known as composite likelihood (CL); see \cite{Varin-11} and the references therein.
+CL was developed in computational statistics as an extension of the familiar notion of likelihood to situations where the likelihood is intractable.
+While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is proven to be consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
+
+%the data distribution is not modeled as a whole but separate feature-based generative models are available
+
+%CL is a semi-generative approach to statistical inference that extends the familiar notion of likelihood without requiring a full generative model.
 The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole.
@@
 
-In this note, CL is interpreted as a probabilistic log-linear opinion pool of ``agents'' making use of different pieces of information extracted from the input data.
-Each agent acts as an isolated statistician using a generative model of a single feature to express an opinion in the form of a likelihood function of the hidden variables.
-The agent opinions are then aggregated into a predictive probability distribution analogous to a Bayesian posterior.
+%An open question regarding CL is how to choose the weights associated with the respective features.
 
-The relevance of each feature in the opinion pool is determined by an arbitrary weight used to scale the associated likelihood.
+With the opinion pooling interpretation in mind, the feature weights in CL may be optimized for prediction performance in a typical discriminative learning scenario.
+As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.
+These likelihood functions themselves may need to be pre-trained, therefore the proposed model typically comes with a double training scheme: a generative training phase (to learn feature-based likelihood parameters) followed by a discriminative training phase (to learn feature weights in aggregation).
+
+Technical details are given in the remainder.
+
+%The relevance of each feature in the CL is determined by an arbitrary weight used to scale the associated likelihood.
 For classification performance, the weights can be tuned in a typical supervised learning scenario so as to concurrently maximize discriminative likelihood (or, equivalently, minimize cross-entropy), thereby making the method hybrid between generative and discriminative modeling approaches.
@@
 
-{\color{red} Still, note that the likelihood functions themselves are trained beforehand so this is slightly different from a standard maxent classifier aka logistic regression.
+%{\color{red} Still, note that the likelihood functions themselves are trained beforehand so this is slightly different from a standard maxent classifier aka logistic regression.
 Double training phase.}
 
-{\color{red} Interpretability of composite likelihood.
+%{\color{red} Classical composite likelihood not studied with nuisance parameters. In our scenario, nuisance parameters are learned beforehand.}
+
+%{\color{red} Interpretability of composite likelihood.
 Obvious from the construction.
@@
 \section{Composite likelihood as opinion pooling}
-\label{sec:pool}
+\label{sec:log_pool}
 
@@
 This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated.
-A natural workaround known as {\em data reduction} is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
+A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
 $$
@@
 If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-We will show in Section~\ref{sec:maxent} that one that is particularly well-suited to our case is the {\em log-linear opinion pool}:
+One that is appealing as it tends to produce single peaked distributions is the {\em generalized logarithmic opinion pool}:
 \begin{equation}
@@
 \end{equation} 
-where $\pi(x)$ is some reference distribution or prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights so that the normalizing factor:
-$$
-z_\lambda(y) = \int \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i} dx
-$$
-is finite, which holds whenever $\lambda\succeq 0$, {\em i.e.} all weights are positive, assuming that the agent opinions are always upper bounded.
-Negative weights should be ruled out as the existence condition $z_\lambda(y)<\infty$ would then depend on the particular opinions expressed by the agents.
-Also note that, while it is possible to have the weights depend on~$y$, there is a strong rational for constant weights as we shall see in the sequel.
+where $\pi(x)$ is some positive-value function which can be interpreted as a global prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights chosen so that the normalizing factor $z_\lambda(y)$ is finite.
+This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are always upper bounded.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$.
+Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and will therefore not be considered here.
 
@@
 
-The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding the form: $p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x)$, where the quantity:
+\section{Composite Bayes rule}
+\label{sec:bayes_rule}
+
+The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding  the form: 
+$$
+p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x),
+$$
+where the function:
 \begin{equation}
@@
 plays the same role as a traditional likelihood function.
-This expression happens to be known in statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
-See Appendix~\ref{sec:conditional} for the slightly more general form called {\em conditional composite likelihood}, which can be derived in the same way.
+This expression happens to be known as a {\em marginal composite likelihood} \cite{Varin-11}.
+The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{sec:conditional}.
+The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature  than the joint distribution of all features.
 
-From~(\ref{eq:comp_lik}), we see that CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
-The clear computational advantage of CL over the true likelihood is that it only requires to evaluate the marginal feature distributions rather than the joint distribution of all features.
+CL in equation~(\ref{eq:comp_lik}) shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
+While our derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method.
+However, unitary CL weights do not guarantee optimal prediction performance.
 
-\section{Tuning composite likelihood weights}
+\section{Composite weight tuning}
 
-When the features can be considered exchangeable, it is natural to choose uniform CL weights.
-The CL is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
-The weight may be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match the frequentist and composite Bayesian notions of uncertainty.
+\subsection{A priori tuning}
 
-Ignoring such a goal, another frequent recommandation for CL weights is to sum up to one.
-This can be motivated in various ways.
-It turns out that the only pooling operator that does not explicitly depend on~$x$ and preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}.
-External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.
-Another appealing property of log-linear pooling with unit sum weights is to minimize the average Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
-A maximum relative entropy property is also given in \cite{Wang-14}, Theorem~3.
+%When the features can be considered exchangeable, it is natural to choose constant CL weights.
 
-Unit sum weights, however, correspond to the extreme situation where the features are assumed to be maximally redundant, but is clearly ineffective for independent or weakly correlated features, which require weights close to one for the CL to closely approximate the true likelihood.
-In this case, the CL may be much flatter than it should, leading to severly overestimated credibility sets.
+In the absence of knowledge, a default rule is to choose constant CL weights.
+The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
+For the purpose of computing reasonable credibility sets, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
 
-{\color{red} Why is it important to fine-tune the weights?}
-We therefore need a method to tune weights that can automatically adjust to the level of statistical dependence beween features.
+Ignoring such a goal, another sensible prior recommendation for log-linear pooling weights is to sum up to one.
+This is motivated in \cite{Genest-86b} by the fact that the only pooling operator that preserves external Bayesianity\footnote{External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.} is the log-linear opinion pool with unit sum weights.
+A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the average Kullback-Leibler (KL) divergence to the agent opinions.
+See also \cite{Wang-14}, Theorem~3, for another information theoretic argument.
+
+Unit sum weights, however, implicitly assume strong redundancy between features and may therefore be ineffective for moderately correlated features.
+An ideal method to tune weights is one that can capture and adjust for existing statistical dependences in the data.
+This calls for a training approach, which, as we will see, is unlikely to pick constant or unit sum weights.
+
+%automatically adjust to the level of statistical dependence beween features.
+
+%are over-conservative for , which require weights close to one for the CL to closely approximate the true likelihood.
+Using weights smaller than one, the CL will be flatter than it should and lead to severly overestimated credibility sets.
+
+%correspond to the extreme situation where the features are assumed to be maximally redundant, but is
+
+%{\color{red} Why is it important to fine-tune the weights?}
+
+\subsection{Training}
+
+Brute force approach: optimize fit to Bayesian posterior.
 
@@
 
-\section{Maxent composite likelihood}
-\label{sec:maxent}
-
-We can derive composite likelihood from the conditional maximum entropy principle \cite{BergerA-96}.
-It comes with a method of tuning weights, hence a particular composite likelihood function that we call maxent composite likelihood (MCL).
-
-A cheap MaxEnt argument was already given by \cite{Wang-14}  (standard, non-conditional MaxEnt at fixed $y$, hence just an existence result but no feasible weight tuning).
-
-What we do here is maximum {\em conditional} entropy.
-It's slightly more complex than a simple $I$-projection.
-The search space is a set of predictive distributions compatible with mean log-likelihood contraints.
-If I know the generative distribution, I know the expected log-likelihood.
-
-We could work with inequality constraints to force $\lambda\geq 0$:
-$$
-E[\log p(z_i|x)] \geq \bar{\ell}_i \equiv \int h(x)p(z_i|x) \log p(z_i|x) dx dy
-$$
-where $h(x)$ is the marginal distribution of labels.
-
-What does the constraint mean?
-It means that a model is admissible if it compresses the data at least as well as all my agents.
-So we are not assuming that the generative models are {\em true}.
-We only assume that they are useful to compress the (reduced) data.
-Connection to MDL.
-
-Why not using the feature-based posteriors or Bayes factors rather than the likelihoods?
-It's a possibility (perhaps a connection here with Gr\"unwald's luckiness).
-A justification for not doing it is that the coordinator should discard the priors used by the agents.
-The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.
-
-Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
-
-The derivation assumed the marginal distribution of the data $h(y)$ to be known. In practice, it is estimated by the empirical distribution just like the moments are estimated.
-
-Game theoretic interpretation \cite{Grunwald-04}.
-
-The dual function boils down to the familiar notions of cross-entropy, log-likelihood, KL divergence...
-$$
-\psi(\lambda)
-= \int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
-$$
-
-If we introduce $p_\lambda(x,y)=p_\lambda(x|y)h(y)$, it is even clearer that this is a most classical distribution fitting problem:
-$$
-\psi(\lambda)
-= \int h(x,y) \log \frac{p_\lambda(x,y)}{\pi(x)h(y)} dx dy
-$$
-(we can obviously drop the reference distribution here).
-So the Maxent justifies the composite likelihood concept as a somewhat optimal parametric form but the remainder of the story is most classical.
-
-Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
-We expect to have $\lambda=1$ but it's not necessarily the case.
-It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
-This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
-So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
-
-Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
-The answer is no.
-The weights generally depend on the prior.
-Hence the maxent composite likelihood is prior-dependent.
-This is an important conceptual difference with the classical notion of likelihood.
-
-\subsection{Data weighting}
-
-The distribution $h(x)$ is the marginal of the labels, which is natural to approximate via their empirical distribution.
-We may then have a discrepancy between $h(x)$ and $\pi(x)$, which makes little sense if it is believed that $h(x)$ is a ``good encoder''.
-
-To correct for this, we may adjust the empirical distribution of labels and data so as to match the target prior, $\pi(x)$, leading to weight the data points as a preprocessing step:
-$$
-h(y) = \sum_i w_i \delta(y-y_i),
-\qquad
-w_i = \frac{\pi(x_i)}{h(x_i)}
-$$
-
-Are the two methods equivalent?
-No.
-The feature-based parameter estimates are obviously the same in both cases, but the search space for the predictive distribution is different, hence we can't get the same answer.
-Which one of the two methods is best?
-The one that allows you to choose your own prior makes more sense to me because the distribution of classes in your training dataset may be completely different from the one that you will deal with in practice.
-For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
+\section{Extensions}
 
@@
 Equation~(\ref{eq:comp_lik}) then amounts to conditional CL \cite{Varin-11}, a more general form of CL also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74} developed for image segmentation.
+
+\section{Maxent}
+
+\subsection{Training}
+\label{sec:maxent}
+
+{\color{red}This stuff should move to the appendix.
+It is not critical to understand the Maxent connection of trained composite likelihood, althought it gives some insight.
+Essentially, agents report compressing rates using feature-based models and we either trust them (equality constraints) or assume that they are underestimated (inequality constraints).
+Why underestimated?
+Well, it's just more general.
+Overestimated leads nowhere, we have to assume something.}
+
+We can derive composite likelihood from the conditional maximum entropy principle \cite{BergerA-96}.
+It comes with a method of tuning weights, hence a particular composite likelihood function that we call maxent composite likelihood (MCL).
+
+A cheap MaxEnt argument was already given by \cite{Wang-14}  (standard, non-conditional MaxEnt at fixed $y$, hence just an existence result but no feasible weight tuning).
+
+What we do here is maximum {\em conditional} entropy.
+It's slightly more complex than a simple $I$-projection.
+The search space is a set of predictive distributions compatible with mean log-likelihood contraints.
+If I know the generative distribution, I know the expected log-likelihood.
+
+We could work with inequality constraints to force $\lambda\geq 0$:
+$$
+E[\log p(z_i|x)] \geq \bar{\ell}_i \equiv \int h(x)p(z_i|x) \log p(z_i|x) dx dy
+$$
+where $h(x)$ is the marginal distribution of labels.
+
+What does the inequality constraint mean?
+It means that a model is admissible if it jointly allows for all agents to achieve the same compression rate {\em or better} than achieved individually.
+Somewhow, we forget that there should exist a true distribution that is consistent with the agent constraints both jointly and individually.
+We forget that not because it is not true but because we deem it irrelevant.
+Connection to MDL?
+We are reminded of these words by Rissanen:
+\begin{quote}
+No statistical model is ``true" or ``false", ``right" or ``wrong"; the models just have varying performance, which can be assessed.  
+\end{quote}
+
+Importantly, no agent will be proven wrong since individual performance is at least as good as announced.
+But if it is strictly better, it means that the agent is simply discarded (getting a zero weight rather than a difficult-to-intepret negative weight).
+
+We could also consider inequality constraints as a relaxation: explicitly making it easier for Nature to screw us up, but for a good cause.
+Not to make optimization easier, to make it easier to intepret.
+Why?
+Because we know that a posteriori from the fact that we get $\lambda_i\geq 0$.
+But a priori?
+
+%Okay, but then it's not ``nature'' playing, is it?
+The game is between a ``generator'' G and a ``predictor'' P.
+G tries to make it as difficult as possible to predict outcomes using an admissible model.
+We could argue that true models are at the boundaries of the admissible space, but G is not restricted to true models as long as it maintains the agents performance level.
+
+%In fact, P is the agent coordinator and G could be seen as some sort of test controller. Inequality constraints add some further robustness to the model.
+
+Why not using the feature-based posteriors or Bayes factors rather than the likelihoods?
+It's a possibility (perhaps a connection here with Gr\"unwald's luckiness).
+A justification for not doing it is that the coordinator should discard the priors used by the agents.
+The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.
+
+Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
+
+The derivation assumed the marginal distribution of the data $h(y)$ to be known. In practice, it is estimated by the empirical distribution just like the moments are estimated.
+
+Game theoretic interpretation \cite{Grunwald-04}.
+
+The dual function boils down to the familiar notions of cross-entropy, log-likelihood, KL divergence...
+$$
+\psi(\lambda)
+= \int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
+$$
+
+If we introduce $p_\lambda(x,y)=p_\lambda(x|y)h(y)$, it is even clearer that this is a most classical distribution fitting problem:
+$$
+\psi(\lambda)
+= \int h(x,y) \log \frac{p_\lambda(x,y)}{\pi(x)h(y)} dx dy
+$$
+(we can obviously drop the reference distribution here).
+So the Maxent justifies the composite likelihood concept as a somewhat optimal parametric form but the remainder of the story is most classical.
+
+Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
+We expect to have $\lambda=1$ but it's not necessarily the case.
+It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
+This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
+So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
+
+Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
+The answer is no.
+The weights generally depend on the prior.
+Hence the maxent composite likelihood is prior-dependent.
+This is an important conceptual difference with the classical notion of likelihood.
+
+\subsection{Data weighting}
+
+The distribution $h(x)$ is the marginal of the labels, which is natural to approximate via their empirical distribution.
+We may then have a discrepancy between $h(x)$ and $\pi(x)$, which makes little sense if it is believed that $h(x)$ is a ``good encoder''.
+
+To correct for this, we may adjust the empirical distribution of labels and data so as to match the target prior, $\pi(x)$, leading to weight the data points as a preprocessing step:
+$$
+h(y) = \sum_i w_i \delta(y-y_i),
+\qquad
+w_i = \frac{\pi(x_i)}{h(x_i)}
+$$
+
+Are the two methods equivalent?
+No.
+The feature-based parameter estimates are obviously the same in both cases, but the search space for the predictive distribution is different, hence we can't get the same answer.
+Which one of the two methods is best?
+The one that allows you to choose your own prior makes more sense to me because the distribution of classes in your training dataset may be completely different from the one that you will deal with in practice.
+For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
 

### 2019-01-16 papers/composite_inference/draft.tex
Modifs 16 janvier 2018

@@
 Frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate artificial intelligence (AI) and typically require supervised learning on large datasets, including ``shallow'' learning techniques (maximum entropy classifiers aka logistic regression \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).
+Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets, including ``shallow'' learning techniques (random forests \cite{Ho-95}, maximum entropy classifiers/logistic regression \cite{BergerA-96}, Gaussian processes \cite{Rasmussen-06}, support vector machines \cite{Vapnik-00}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06}).
 
-A common criticism of discriminative models is that they are ``black boxes'' in that they can make valid inferences but provide little or no information to support them \cite{Molnar-18}.
-This is problematic in application fields such as medical diagnosis, where correct inferences are not straightforward to humans and AI is intended as a tool to assist a decision making process.
-Automated disease predictions should therefore be corroborated by individualized findings to be of any clinical value -- in other words, AI systems should be able to justify their opinion just like clinicians.
-Unlike discriminative models, simple methods based on range analysis of univariate biomarkers can provide meaningful clues, and are thus more useful in practice although less powerful in terms of prediction performance.
+%Interpretability needed in life-impacting applications of AI. 
+%Whenever possible, proxy features should be avoided as they make models gameable. 
+%Local Interpretability for a Single Prediction: Why did the model make a certain prediction for an instance?
+%What about sensitivity analysis: study the prediction as a function of a single feature...
+
+Discriminative models are often considered as black boxes because their predictions may be valid  but difficult to interpret.
+This is problematic in life-impacting applications, where AI is confined to a supportive role in a human-driven decision making process \cite{Molnar-18}.
+For instance, in medical diagnosis, automated disease predictions should be corroborated by individualized findings to bring clinical value -- in other words, AI systems should be able to justify their opinions just like clinicians.
+Simple univariate generative models, although expectedly less powerful than discriminative models in terms of prediction performance, can provide decisive clues for diagnosis, {\em e.g.}, by pinpointing biomarkers that deviate from healthy reference ranges.
+
+% simple methods based on range analysis of univariate biomarkers can provide meaningful clues, and are thus more useful in practice although less powerful in terms of prediction performance.
+
+%With generative models you can compare data with expected outcomes to get insight.
+
+%the information they provide to support them is limited to the boundaries between classes, which are difficult to interpret for high-dimensional data representations.
 
@@
 It turns out that, when the aggregation method is chosen as a log-linear pool, the consensus distribution is (up to a multiplicative prior) a quantity known as composite likelihood (CL); see \cite{Varin-11} and the references therein.
-CL was developed in computational statistics as an extension of the familiar notion of likelihood to situations where the likelihood is intractable.
+CL was developed in computational statistics as an extension of the familiar notion of likelihood to situations where a genuine likelihood is intractable.
 While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is proven to be consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
@@
 \label{eq:log_pool}
-p_\lambda(x|y) = \frac{1}{z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
+p_\lambda(x|y) = \frac{1}{m_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
 \end{equation} 
-where $\pi(x)$ is some positive-value function which can be interpreted as a global prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights chosen so that the normalizing factor $z_\lambda(y)$ is finite.
+where $\pi(x)$ is some positive-value function which can be interpreted as a global prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights chosen so that the normalizing factor $m_\lambda(y)$ is finite.
 This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are always upper bounded.
@@
 
-Brute force approach: optimize fit to Bayesian posterior.
+Brute-force approach: optimize fit for optimal prediction.
+Discriminative model defined by (\ref{eq:log_pool}).
+Training objective, weight likelihood (which, in this context, is negated cross-entropy) better than accuracy:
+$$
+\max_{\lambda\succeq 0} {\cal L}(\lambda),
+\qquad {\rm with} \quad
+{\cal L}(\lambda)\equiv\sum_k \log p_\lambda(x_k|y_k)
+$$
 
-From a machine learning perspective, it is natural to tune the weights so as to minimize some prediction error, for example conditional cross-entropy.
+We recognize the dual function $\psi(\lambda)$ of a conditional maxent problem, see Appendix.
+$$
+{\cal L}(\lambda)
+= 
+\sum_k \log \pi(x_k)
++ n \psi(\lambda)
+$$
+
+Maxent defined by constraints given by mean log-likelihood values.
+Weaker than assuming feature distributed according to agent models.
+Can be seen in terms of detection threshold: all we care about is that our feature-based log-likelihoods reach guaranteed levels (on the average) for the true class.
+Hence inequality constraints are more natural than equality ones.
+Model is admissible if agent compression rates are achieved or surpassed.
+
+We know more about the model, so this is relaxation.
+Doesn't make optimization easier but, I speculate, the solution is more stable than if we were imposing feature-based marginals.
+Both search spaces are determined by the empirically estimated likelihood parameters, intuitively I would say that the bigger space is less sensitive to estimation errors.
+We can show an example where weights are much more stable by choosing inequality vs equality constraints.
+
+%From a machine learning perspective, it is natural to tune the weights so as to minimize some prediction error, for example conditional cross-entropy.
 This brings the question: why then do it with ``pre-learned'' likelihood functions rather than with basis functions spanning the same predictive distribution family?
@@
 
-The ``hybrid'' strategy necessarily achieves a larger cross-entropy than the pure discriminative one on the training dataset, but might generalize better. Should show examples of this.
+%The ``hybrid'' strategy necessarily achieves a larger cross-entropy than the pure discriminative one on the training dataset, but might generalize better. Should show examples of this.
 
-The rest of the discussion, i.e. the link with maximum entropy, is less important beside showing that the composite likelihood form is somehow optimal given some knowledge.
+%The rest of the discussion, i.e. the link with maximum entropy, is less important beside showing that the composite likelihood form is somehow optimal given some knowledge.
 
-%Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature.
-This results in increasing their weights in the log-linear pool.}.
+Intuition behind weights: Lagrange multipliers to enforce a constraint...
 
-\section{Extensions}
+Alternative: equality constraints (negative weights), less interpretable.
 
-\subsection{Super composite likelihood}
+%\section{Extensions}
+
+\section{Super composite likelihood}
 \label{sec:super}
 
-What happens if we use mean-values constraints conditional on $x$ instead of jointly averaged over $x$ and $y$?
-This boils down to picking functions of the form $\chi(x-a)\ell_i(x,y)$ and equate their expectation
+Extension: use mean-values constraints conditional on $x$ instead of jointly averaged over $x$ and $y$?
+This boils down to picking functions of the form $\chi(x-a)\rho_i(x,y)$ and equate their expectations.
+For this to work, we need the constraints to involve log-likelihood ratios vs a reference distribution chosen as one of the class distributions.
+Log-likelihood ratios make even more sense than log-likelihood in the constraints...
 
-The answer is that the maxent solution has the form:
+The maxent solution has the form:
 $$
-p_\lambda(x|y) = \pi(x) \prod_i p(z_i|x)^{\lambda_i(x)}
+p_\lambda(x|y) = \pi(x) \prod_i \left[\frac{p(z_i|x)}{\pi(z_i)}\right]^{\lambda_i(x)}
 $$
@@
 
-However, I was thinking about the unit sum weight version...
+I was thinking about the unit sum weight version...
 The diagnosis example could be one in which clues are independent, in which case $\lambda_i(x)\equiv 1$ would be optimal, making any super composite trick irrelevant.
 Also see that with hypothesis-dependent weights, we may lose the estimation consistency.
-At the end of the day, the super composite idea looks like a pretty pointless sophistication.
+At the end of the day, the interest of the super composite idea is questionable.
 
@@
 
-Deep learning is discriminative...
+{\color{red} Semi-generative method. Good for shallow learning or transfer learning. Still need representation learning methods.}
+
+Number of discriminative parameters in logistic regression: $K(F+1)$.
+In standard composite inference: $F$.
+In super-composite inference: $FK$.
+Number of generative parameters in composite inference using univariate Gaussian models: $2K$ with heteroscedasticity, $K+1$ with homoscedasticity.
+
+Intuition behind weights: notion of global feature relevance.
+But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$ for a particular instance?
+No.
+But we can assess the influence of each scaled likelihood in order to rank features according to their influence on the decision.
 
@@
 
-Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?
-
 The derivation assumed the marginal distribution of the data $h(y)$ to be known. In practice, it is estimated by the empirical distribution just like the moments are estimated.
@@
 For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
+
+\section{Maximum entropie conditionnelle}
+
+$$
+D(p) = \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
+$$
+
+Lagrangian:
+$$
+{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
+$$
+
+Optimal $p$ at fixed $\lambda$:
+$$
+\frac{\partial{\cal L}}{\partial p}
+=
+h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
+= h(y) c(y)
+\quad
+\Rightarrow
+\quad
+p_{\lambda}(x|y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
+$$
+with:
+$$
+z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
+$$
+
+Dual:
+$$
+\psi(\lambda) = \lambda^\top F - \int h(y) \log z(\lambda, y) dy 
+$$
+
+$$
+\nabla z
+=
+\int \pi(x) f(x,y)  e^{\lambda^\top f(x,y)} dx
+$$
+
+$$
+\nabla \nabla^\top z 
+=
+\int \pi(x) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dx
+$$
+
+$$
+\nabla\psi
+= F - \int h(y) \frac{\nabla z}{z} dy
+$$
+
+$$
+\nabla\nabla^\top\psi
+= - \int h(y) \left(
+\frac{\nabla \nabla^\top z}{z} 
+-
+\frac{\nabla z \nabla z^\top}{z^2}
+\right)
+dy
+$$
+
+Note the equivalent expression of the dual function as:
+$$
+\psi(\lambda)
+=
+\int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
+= D(h\|\pi) - D(h\|p_\lambda)
+$$
+
+\section{Maxent via the generalized KL}
+
+It's a bit different as the unit integral constraint is made implicit.
+
+Assume $\int\pi(x)dx=1$ and $\int h(y) dy = 1$. 
+$$
+D(p) 
+= \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
+- \int h(y)p(x|y)dx dy
++ 1
+$$
+
+Goal is to minimize $D(p)$ s.t. mean-value constraints on $p$.
+
+Lagrangian:
+$$
+{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
+$$
+
+Optimal $p$ at fixed $\lambda$:
+$$
+\frac{\partial{\cal L}}{\partial p}
+=
+h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
+= 0
+\quad
+\Rightarrow
+\quad
+p_{\lambda}(x|y) = \pi(x) e^{\lambda^\top f(x,y)}
+$$
+
+Dual:
+$$
+\psi(\lambda) = \lambda^\top F - \int h(y)z(\lambda, y) dy + 1
+$$
+with:
+$$
+z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
+$$
+
+$$
+\nabla\psi(\lambda)
+= F - \int \pi(x) h(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
+$$
+
+$$
+\nabla\nabla^\top\psi(\lambda)
+= - \int \pi(x) h(y) f(x,y)f(x,y)^\top e^{\lambda^\top f(x,y)} dx dy
+$$
 
@@
 
+\section{Data reduction inequality}
+\label{sec:reduction_inequality}
+
+A fundamental property of the Kullback-Leibler divergence is that it decreases under feature extraction, {\em i.e.}, application of a deterministic transformation.
+This comes as a consequence of the logarithmic sum inequality \cite{Cover-06} (and also follows as a straightforward corollary of the PDF~projection theorem \cite{Minka-04,Baggenstoss-15}).
+
+The proof is elementary and can be sketched as follows.
+Let $Z=f(Y)$ some feature extracted from a variable~$Y\sim p(y)$.
+Given an arbitrary distribution~$\pi(y)$, let  $\tilde{p}(z)$ and $\tilde{\pi}(z)$ the distributions induced on~$Z$ by $p(y)$ and $\pi(y)$, respectively.
+For any potential value~$z$ of~$Z$, consider the level set: $\Gamma(z)=\{y, f(y)=z\}$.
+Under conditions of existence, we can partition the integral involved in the KL~divergence $D(p\|\pi)$ using these level sets:
+$$
+\int p(y) \log \frac{p(y)}{\pi(y)} dy
+=
+\int \left( \int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy \right) dz\\
+.
+$$
+
+Applying the logarithmic sum inequality \cite{Cover-06} to each integral over~$\Gamma(z)$, we then readily get:
+$$
+\int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy
+\geq 
+\tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)}
+,
+$$
+owing to the fact that $\displaystyle \int_{\Gamma(z)} p(y) dy = \tilde{p}(z)$ by definition. Therefore,
+$$
+\int \tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)} dz
+\leq 
+\int p(y) \log \frac{p(y)}{\pi(y)} dy
+,
+$$
+or, in a more compact way, $D(\tilde{p}\|\tilde{\pi})\leq D(p\|\pi)$.
+The case of equality occurs if, and only if, $p(y)/\pi(y)=\tilde{p}(z)/\tilde{\pi}(z)$, in other words, if~$z$ is a sufficient statistic for the ``alternative'' hypothesis~$H_1:y\sim p(y)$ versus the ``null'' hypothesis~$H_0:y\sim\pi(y)$.
+
+Since it is a well-known fact that $D(\tilde{p}\|\tilde{\pi})\geq 0$ as any KL~divergence, we conclude that:
+\begin{equation}
+\label{eq:reduction_inequality}
+0 
+\leq D(\tilde{p}\|\tilde{\pi}) 
+\leq D(p\|\pi)
+.
+\end{equation}
+
+\section{Basic asymptotic properties of composite likelihood}
+\label{sec:asymptotic}
+
+It follows from the double inequality~(\ref{eq:reduction_inequality}) that, for any extracted feature~$z_i$ and for any hypothesis~$\theta$,
+$$
+0 \leq
+E\left[
+\log \frac{p(z_i|\theta_\star)}{p(z_i|\theta)}
+\right]
+\leq
+E\left[
+\log \frac{p(y|\theta_\star)}{p(y|\theta)}
+\right],
+$$
+where the expectation is taken with respect to the true distribution $p(y|\theta_\star)$.
+Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any standard composite likelihood function (assuming unit sum positive weights):
+\begin{equation}
+\label{eq:variation_bound}
+0 \leq
+E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta, \mathbf{w})} \right]
+\leq 
+E\left[ \log \frac{L(\theta_\star)}{L(\theta)} \right]
+.
+\end{equation}
+
+This implies two asymptotic properties of standard composite likelihood: 
+\begin{itemize}
+\item {\em Consistency.} The expected log-composite likelihood is maximized by~$\theta_\star$, the true value of~$\theta$.
+\item {\em Conservativeness.} The expected log-composite likelihood function is ``flatter'' than the true expected log-likelihood.
+In other words, composite likelihood ratios $L_c(\theta, \mathbf{w})/L_c(\theta_\star, \mathbf{w})$ relative to~$\theta_\star$ tend to be larger than the corresponding true likelihood ratios.
+\end{itemize}
+
+These properties do not necessarily extend to the more general case of SCL. We may state a weaker consistency property by considering the upper envelope of expected log-composite likelihood ratios:
+$$
+M(\theta) = \max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
+$$
+where ${\cal W}$ is the $n$-dimensional simplex. Clearly, the expected logarithm of the SCL is upper bounded by~$M(\theta)$:
+$$
+E[\log {\cal L}_c(\theta, \mathbf{W})] \leq M(\theta)
+.
+$$
+
+Moreover, $M(\theta)$ is maximized by $\theta_\star$ since (\ref{eq:variation_bound}) implies that, for any $(\theta,\theta_0)$,
+$$
+E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
+\leq
+E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
+$$
+which remains true when taking the maximum over the weights in both sides: 
+$$
+M(\theta) = 
+\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
+\leq
+\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
+= M(\theta_\star)
+.
+$$
+
+Therefore, maximizing SCL corresponds to maximizing a lower bound on~$M(\theta)$, which can be considered as an objective function since it is maximized by~$\theta_\star$ (like the expected log-likelihood).
+Lower bound maximization guarantees a certain objective value, however not the maximum, hence asymptotic consistency may not hold.
+
 \bibliographystyle{abbrv}

### 2019-02-04 papers/composite_inference/draft.tex
Update intro CL

@@
 %\def\x{{\mathbf{x}}}
-%\def\z{{\mathbf{z}}}
 %\def\u{{\mathbf{u}}}
@@
 \begin{abstract}
-This note revisits the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
+We revisit the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
 This view makes it natural to use a machine learning approach to calibrate the weights associated with composite likelihood.
-For instance, they can be tuned in a typical supervised learning phase so as to minimize cross-entropy on a labeled training dataset, yielding a convex problem.
-We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to some artificial intelligence tasks.
+For instance, they can be tuned in a typical supervised learning stage so as to minimize cross-entropy on a labeled training dataset, yielding a convex problem.
+We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
 \end{abstract}
@@
 
-Frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets, including ``shallow'' learning techniques (random forests \cite{Ho-95}, maximum entropy classifiers/logistic regression \cite{BergerA-96}, Gaussian processes \cite{Rasmussen-06}, support vector machines \cite{Vapnik-00}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06}).
+Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
+Because this requirement may be very challenging for multi-dimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include ``shallow'' learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
 
@@
 
-Discriminative models are often considered as black boxes because their predictions may be valid  but difficult to interpret.
-This is problematic in life-impacting applications, where AI is confined to a supportive role in a human-driven decision making process \cite{Molnar-18}.
-For instance, in medical diagnosis, automated disease predictions should be corroborated by individualized findings to bring clinical value -- in other words, AI systems should be able to justify their opinions just like clinicians.
-Simple univariate generative models, although expectedly less powerful than discriminative models in terms of prediction performance, can provide decisive clues for diagnosis, {\em e.g.}, by pinpointing biomarkers that deviate from healthy reference ranges.
+Discriminative models, however, are mere inference recipes and lack the built-in ability to ``support'' their predictions.
+Consider as a straightforward example the problem of deciding whether a pan is a sauce pan or a frying pan based on shape measurements.
+A simple threshold on the pan depth could work well in most practical cases, but only makes sense if known the fact that sauce pans tend to be deeper than frying pans, which is generative information left implicit in the decision rule.
+For multi-dimensional features, it quickly becomes impossible to relate adjusted boundaries between classes to generative information {\em a posteriori}.
+
+%Such common knowledge is rarely available in real-world apps.
+
+% merely trace boundaries between classes in the data representation space and therefore lack built-in information to interpret their predictions.
+
+There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
+For instance, in medical diagnosis, automated disease predictions should always be corroborated by individual findings to bring real clinical value -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like clinicians.
+Clinicians have long used methods based on simple generative models, such as range analysis of univariate biomarkers, that can provide decisive clues for diagnosis despite being outperformed by discriminative models for blind disease prediction.
+
+%, {\em e.g.}, by pinpointing biomarkers that deviate from healthy reference ranges.
 
@@
 
-Another limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as marginally independent, meaning that the data conveys no information about the parameters unless the variable of interest is observed.
-This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models.
-For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence is less effective in small training sets \cite{Ng-01}.
+Another limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as {\em marginally independent} variables, meaning that the data conveys no information about the parameters unless the target variable is observed.
+This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graphs representing generative and discriminative models.
+For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence requires larger training sets to achieve optimal performance \cite{Ng-01}.
 
@@
 \subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
-\caption{Bayesian networks representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the variable of interest, the data, and the model parameters.
+\caption{Directed graphs representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the target variable, the data, and the model parameters.
 Note the marginal independence of data and parameters in the discriminative model.}
@@
 
-This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a possible compromise between generative and discriminative models.
-Rather than using a purely discriminative model or attempting at a full generative model, we may combine multiple low-dimensional generative models for different pieces of information extracted from the input data.
-Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the variable of interest.
+This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a means to incorporate generative information into a discriminative model.
+Rather than attempting at a full generative model, one may combine multiple low-dimensional generative models for different low-dimensional pieces of information extracted from the input data.
+Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
 The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
-This simple idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
-Importantly, predictions made in such a way could be readily interpretable owing to the use of low-dimensional generative models.
-
-It turns out that, when the aggregation method is chosen as a log-linear pool, the consensus distribution is (up to a multiplicative prior) a quantity known as composite likelihood (CL); see \cite{Varin-11} and the references therein.
-CL was developed in computational statistics as an extension of the familiar notion of likelihood to situations where a genuine likelihood is intractable.
-While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is proven to be consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
+This idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
+Importantly, predictions made in such a way are interpretable owing to their underlying generative nature.
+
+%use of low-dimensional generative models.
+
+When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be (up to a multiplicative prior) a quantity known as composite likelihood (CL) in computational statistics.
+As described in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
+While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is asymptotically consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
@@
 As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.
-These likelihood functions themselves may need to be pre-trained, therefore the proposed model typically comes with a double training scheme: a generative training phase (to learn feature-based likelihood parameters) followed by a discriminative training phase (to learn feature weights in aggregation).
+These likelihood functions themselves may need to be pre-trained, therefore the composite inference approach typically invovles a double training scheme: a generative training stage (to learn feature-based likelihood parameters) followed by a discriminative training stage (to learn feature weights in aggregation).
 
@@
 
-Let $Y$ an observable multivariate random variable with sampling distribution $p(y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is a known set.
+Let $Y$ an observable multivariate random variable with sampling distribution $p(y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is assumed to be a finite set for simplicity.
 Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $x$:
@@
 
-Substituting $L(x)$ with $\ell(x)$ boils down to restricting the sample space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information.
+Substituting $L(x)$ with $\ell(x)$ boils down to restricting the feature space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information.
 While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction.
@@
 \label{eq:log_pool}
-p_\lambda(x|y) = \frac{1}{m_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
+p_\lambda(x|y) \propto \nu(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
 \end{equation} 
-where $\pi(x)$ is some positive-value function which can be interpreted as a global prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights chosen so that the normalizing factor $m_\lambda(y)$ is finite.
-This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are always upper bounded.
+where $\nu(x)$ is some arbitrary reference distribution (which may differ from a Bayesian prior as we shall see below), and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
 Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$.
-Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and will therefore not be considered here.
+Also note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
+
+%This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$.
+Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
 
@@
 $$
-p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x),
-$$
-where the function:
+p_\lambda(x|y)\propto \nu(x) L^c_\lambda(x),
+$$
+where the distribution $\nu(x)$ plays the role of a prior, and the function:
 \begin{equation}
@@
 \end{equation} 
-plays the same role as a traditional likelihood function.
+plays that of a traditional likelihood function.
 This expression happens to be known as a {\em marginal composite likelihood} \cite{Varin-11}.
@@
 CL in equation~(\ref{eq:comp_lik}) shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
-While our derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method.
+While the above derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method.
 However, unitary CL weights do not guarantee optimal prediction performance.
 
+\section{Prediction interpretation}
+
+A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$y$.
+Given two target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ simply adds up single feature contributions:
+$$
+\log \frac{p_\lambda(x_1|y)}{p_\lambda(x_0|y)}
+= 
+\underbrace{\log \frac{\nu(x_1)}{\nu(x_0)}}_{\text{offset}}
++ \sum_i \underbrace{\lambda_i \log \frac{p(z_i|x_1)}{p(z_i|x_0)}}_{\text{feature contribution}},
+$$
+therefore is straightforward to assess which features contribute most (either positively or negatively), see possible disagreements between features, and corroborate the overall prediction with feature-level testable information.
+
+Each feature contributes proportionally to its associated log-likelihood ratio test statistic for the comparison $x_1$ vs~$x_0$.
+Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction, and may therefore account for mutual feature dependence if tuned appropriately.
+Contrary to statistical hypothesis testing, likelihood ratio test statistics are not converted into $p$-values but are directly combined into predictive probabilities.
+In this perspective, composite weights may be seen as an alternative to multiple testing correction.
+
+% For instance, we can get insight such as: ``this patient is more likely to be AD than healthy because his hippocampus is severly atrophied even though his memory tests are quite normal.''
+
+%So, if a feature yields a significant test but has low weight, it may have little influence on the comparison.
+
+%How about overall feature influence?
+Well, if we want to really interpret the prediction, we shall go through each target comparison that is deemed important for the decision.
+For instance, the system says AD more likely than normal, but is AD more likely than MCI?
+To answer this, we need to apply the procedure to AD vs MCI, and maybe other features will pop up.
+
+In order to understand predictive distribution, the user(s) may go through successive pairwise comparisons of target values in a dynamic way.
+They may start comparing the most probable target, {\em i.e.} the automated prediction, to some educated guess or reference class (for instance, normality in medical diagnosis).
+Next, depending on the feature contributions highlighted by the comparison interpretation, they may form another hypothesis to compare to the automated prediction.
+This process, which can be thought of as a kind of human machine dialog, may be repeated until the composite prediction is felt to be understood by the user(s).
+
+%See if the automated prediction makes sense.
+If not, stop, otherwise consider what other class the interpretation could suggest and ask for comparison.
+Repeat the process until all other possibilities are subjectively ruled out or the user decides to reject the prediction.
+Think about this as a human AI dialog.
+
+%This is a recipe for {\em local} interpretation (why did the system make a particular prediction), but what about {\em global} interpretation (how does the system make predictions).
+The thing is, we may not care about global interpretation.
+The only thing we need to know is that the system uses a bunch of features that all potentially contribute to the predictions.
+We could investigate which features tend to contribute most notably and which ones do not {\em on the average} but this would only be interesting design purposes, i.e.
+if we are prepared to use another representation.
+
+%We may represent feature influence by a $K\times K$ matrix with general element $\lambda_i r_i(x_k,x_\ell)$.
+
+%This recipe addresses {\em local} interpretation (why did the system make a particular prediction) as opposed to {\em global} interpretation (how does the system make predictions in general).
+
 \section{Composite weight tuning}
-
-\subsection{A priori tuning}
-
-%When the features can be considered exchangeable, it is natural to choose constant CL weights.
+\label{sec:weight_tuning}
+
+\subsection{Agnostic calibration}
 
@@
 The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
-For the purpose of computing reasonable credibility sets, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
+For the purpose of computing credibility sets, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
 
 Ignoring such a goal, another sensible prior recommendation for log-linear pooling weights is to sum up to one.
-This is motivated in \cite{Genest-86b} by the fact that the only pooling operator that preserves external Bayesianity\footnote{External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.} is the log-linear opinion pool with unit sum weights.
+This is motivated in \cite{Genest-86b} by the fact that the only pooling operator that preserves external Bayesianity\footnote{External Bayesianity means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.} is the log-linear opinion pool with unit sum weights.
 A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the average Kullback-Leibler (KL) divergence to the agent opinions.
@@
 
-Unit sum weights, however, implicitly assume strong redundancy between features and may therefore be ineffective for moderately correlated features.
-An ideal method to tune weights is one that can capture and adjust for existing statistical dependences in the data.
+Unit sum weights, however, implicitly assume strong redundancy between features and may therefore be ineffective in practice since features are generally chosen to be complementary.
+An ideal method to tune weights is one that can capture and adjust for existing statistical dependences between features.
 This calls for a training approach, which, as we will see, is unlikely to pick constant or unit sum weights.
-
-%automatically adjust to the level of statistical dependence beween features.
-
-%are over-conservative for , which require weights close to one for the CL to closely approximate the true likelihood.
-Using weights smaller than one, the CL will be flatter than it should and lead to severly overestimated credibility sets.
-
-%correspond to the extreme situation where the features are assumed to be maximally redundant, but is
-
-%{\color{red} Why is it important to fine-tune the weights?}
 
@@
 Intuition behind weights: Lagrange multipliers to enforce a constraint...
+From empirical observation, there is a non-linear correlation between CL weight and feature-level discrimination power (as measure e.g.
+by maximum KL divergence).
+Non-linear means that features with average discrimination power may get low weight as they are superseded by the most discriminative features.
+But some features with low discrimination power may also get large weight, which does not mean that they are very influential in practice (see above).
+This is illustrated in Figure~\ref{fig:disc_weight_plot} with the UCI breast cancer dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}} used to predict whether breast lumps are benign or malignant based on a set of features extracted from images of fine-needle aspirates.
+
+\begin{figure}[!ht]
+  \begin{center}
+    \includegraphics[width=.6\textwidth]{disc_weight_plot.pdf}
+  \end{center}
+\caption{Distribution of optimal composite weights vs individual feature discrimination power in the breast cancer UCI dataset.
+Discrimination is measured by the maximum KL divergence between class-conditional generative distributions.
+The most discriminative feature, `worst concave points', has a moderate weight of $0.34$ while `worst texture' achieves a weight of $0.78$ but is among the least discriminative features.}
+\label{fig:disc_weight_plot}
+\end{figure}
 
 Alternative: equality constraints (negative weights), less interpretable.
+
+Data weights: 
+$$
+w_k = \frac{\pi(x_k)}{n(x_k)}
+$$
+
+Need to weight the data to adjust to prior as a general measure when minimizing cross-entropy, which should yield a conditional model that approximates the ``true'' conditional distribution, $q(x|y)\approx p(x|y)$.
+So the model will absorb the ``true'' marginal of $x$, which should thus be replaced by the prior if we want the model to mimic ideal Bayesian inference.
+In our case, it means that the composite likelihood approximates the true likelihood.
+
+We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\nu(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$.
+
+When including offsets, using homoscedastic generative variance estimation and equality constraints, composite likelihood is equivalent to binomial logistic regression.
+However, equality constraints are not easily interpretable.
+
+%In some sense, composite weights replace the multiple testing correction required in the frequentist inference paradigm for proper control of false positives.
+Here, we don't control false positives, we control prediction accuracy.
 
@@
 
-Extension: use mean-values constraints conditional on $x$ instead of jointly averaged over $x$ and $y$?
-This boils down to picking functions of the form $\chi(x-a)\rho_i(x,y)$ and equate their expectations.
-For this to work, we need the constraints to involve log-likelihood ratios vs a reference distribution chosen as one of the class distributions.
-Log-likelihood ratios make even more sense than log-likelihood in the constraints...
-
-The maxent solution has the form:
-$$
-p_\lambda(x|y) = \pi(x) \prod_i \left[\frac{p(z_i|x)}{\pi(z_i)}\right]^{\lambda_i(x)}
-$$
-
-In other words, the weights become dependent on the variable of interest. This is the super composite likelihood idea in a nutshell.
-
-In 2016, I wrote:
-\begin{quote}
-When chosen for computational simplicity, clues may not only convey limited information at individual level: their informativeness may also be very much hypothesis-dependent.
+When chosen for computational simplicity, features may not only convey limited information at individual level, their informativeness may also be very much class-dependent.
 Consider, for instance, diagnosing a disease from a routine medical checkup.
-Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovascular disease -- and conversely for, say, blood pressure.
-This motivates a more general setting where clues can be weighted differently depending on hypotheses.
-\end{quote}
-
-I was thinking about the unit sum weight version...
+Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
+This motivates a more general pooling than~(\ref{eq:log_pool}) setting where features can be weighted depending on the target variable:
+$$
+p_\lambda(x|y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)}
+$$
+
+$p_0$ is an arbitrary reference (implicit if weights are class-independent as it is then effectless).
+Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
+healthy population).
+In this case, the super composite likelihood is equivalent to the standard one in the two-class case.
+
+This is the super composite likelihood idea in a nutshell.
+
+Again, we can train this model and see that it is the same as solving a maxent problem.
+Difference with above is that the constraints are not averaged over $x$ (basis functions are of the form $\chi(x-a)\rho_i(x,y)$) and that they involve arbitrary log-likelihood ratios vs some reference distribution.
+Make sense to have a reference as it gives a customizable baseline for detection.
+
+If some classes are merged or split, then the other classes keep the same weights (assuming $h(y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
+So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
+
+Super composite likelihood is essentially about combining binary versions of composite likelihood. It is analogous to the ``one-versus-rest'' strategy in multiple logistic regression.
+
+%Link with PDF projection \cite{Baggenstoss-03}.
+
+Advantages: richer model (so better predictions), probabilty ratios stable to class addition.
+Drawbacks: require a reference, no more consistency.
+Still interpretable, that's why it is worthwile mentioning.
+
+%I was thinking about the unit sum weight version...
 The diagnosis example could be one in which clues are independent, in which case $\lambda_i(x)\equiv 1$ would be optimal, making any super composite trick irrelevant.
@@
 
+{\color{red} Why prediction performance is still important? Because it validates (or not) interpretation. Interpretation is the goal but prediction performance is a crucial intermediate step.}
+
+{\color{red} We cannot rank features according to their weights.
+A feature might incidentally get a large weight because its likelihood is flat.
+Feature relevance also depends on conditional likelihood distribution.
+Consider mutually independent features: each one recieves a unit weight, but some may be independent from target.}
+
 {\color{red} Semi-generative method. Good for shallow learning or transfer learning. Still need representation learning methods.}
 
-Number of discriminative parameters in logistic regression: $K(F+1)$.
-In standard composite inference: $F$.
-In super-composite inference: $FK$.
+{\color{red} It is easy to build a good classifier.
+You just need to pick a parametric family that is big enough to produce a good fit yet small enough to avoid overfitting.
+Less easy is to interpret what the classifier does.
+Deep learning is cool but it is relevant to AI only  if the representations it learns are generic enough to be used in other problems, i.e.
+it should ultimately be validated via transfer learning.}
+
+{\color{red} Composite Bayes is invariant by affine transformation of the data (unlike multinomial LR, which is problematic when damping is applied).}
+
+Some experimental observations.
+In the iris dataset (historical data used by Fisher to demonstrate linear discriminant analysis), composite Bayes finds that petal length is the most relevant feature followed by petal width, while sepal length and width get zero weight.
+In the wine dataset, the flavanoids is the most relevant feature for standard composite Bayes, but it's color intensity closely followed by flavanoids for the super composite version.
+In the breast cancer, both approaches are equivalent since it is a binary problem, weights are very sparse: only 13/30 are non-zero with worst area the most relevant.
+In the digits prolem, the most relevant features tend to be in the image center; composite Bayes can achieve similar performance to multinomial logistic regression with about 10 times less parameters.
+
+Number of independent discriminative parameters in logistic regression: $(K-1)(F+1)$.
+In standard composite inference: $K-1+F$.
+Difference $=(K-2)F$.
+In super-composite inference: $(K-1)(F+1)$ (if reference is a class).
 Number of generative parameters in composite inference using univariate Gaussian models: $2K$ with heteroscedasticity, $K+1$ with homoscedasticity.
@@
 
-\section{Maximum entropie conditionnelle}
-
-$$
-D(p) = \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
+\section{Optimal predictive model}
+
+Let $\pi(x)$ some prior distribution on~$X$. Optimal prediction aims to select the joint distribution $p(x,y)$ that minimizes the KL divergence $D(p(x,y)\|\pi(x)p(y))$:
+$$
+D(p) = -\int p(x,y) \log \frac{p(x|y)}{\pi(x)} dx dy,
+$$ subject to feature mean-value constraints of the form:
+$$
+\int p(x,y) f(x,y) dx dy = \mu.
+$$
+
+Note that with some constraints of this type, we can enforce $p(x)=\pi(x)$. In this case, the best predictive model minimizes the mutual information between the data and the target.
+
+One important special case of this problem arises when imposing the additional constraint that $p(y)=h(y)$ is known.
+This is the {\em maximum conditional entropy} method \cite{BergerA-96} which we review hereafter.
+We will also treat the case where the marginal constraint is not assumed, which corresponds to the {\em minimally informative likelihood} method \cite{Yuan-99b,Yuan-99}.
+
+%We could say that MIL enforces the prior constraint explicitly while Maxent does it implicitly by pre-weighting the data.
+So it's a soft constraint in the Maxent case (which ends up being loosely satisfied if the model fits the empirical distribution well enough), and it's a hard constraint in the MIL case.
+On the other hand, the MIL formulation does not exploit the  empirical marginal at all (it only exploits the moments).
+Which one do we prefer?
+
+\subsection{Maximum conditional entropy}
+
+$$
+D(p) = -\int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
 $$
@@
 $$
-{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
+{\cal L}(p,\lambda) = D(p) + \lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - \mu)
 $$
@@
 =
-h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
-= h(y) c(y)
-\quad
+- h(y) \log \frac{p(x|y)}{\pi(x)} + h(y) \lambda^\top f(x,y) + (\kappa(y)-1)h(y)
+$$
+$$
 \Rightarrow
@@
 $$
-\psi(\lambda) = \lambda^\top F - \int h(y) \log z(\lambda, y) dy 
+\psi(\lambda) = \lambda^\top \mu - \int h(y) \log z(\lambda, y) dy 
 $$
@@
 \nabla\psi
-= F - \int h(y) \frac{\nabla z}{z} dy
+= \mu - \int h(y) \frac{\nabla z}{z} dy
 $$
@@
 \frac{\nabla \nabla^\top z}{z} 
--
-\frac{\nabla z \nabla z^\top}{z^2}
+- \frac{\nabla z \nabla z^\top}{z^2}
 \right)
@@
 \int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
-= D(h\|\pi) - D(h\|p_\lambda)
-$$
-
-\section{Maxent via the generalized KL}
-
-It's a bit different as the unit integral constraint is made implicit.
-
-Assume $\int\pi(x)dx=1$ and $\int h(y) dy = 1$. 
-$$
-D(p) 
-= \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
-- \int h(y)p(x|y)dx dy
-+ 1
-$$
-
-Goal is to minimize $D(p)$ s.t. mean-value constraints on $p$.
-
-Lagrangian:
-$$
-{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
-$$
-
-Optimal $p$ at fixed $\lambda$:
-$$
-\frac{\partial{\cal L}}{\partial p}
-=
-h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
-= 0
-\quad
-\Rightarrow
-\quad
-p_{\lambda}(x|y) = \pi(x) e^{\lambda^\top f(x,y)}
-$$
-
-Dual:
-$$
-\psi(\lambda) = \lambda^\top F - \int h(y)z(\lambda, y) dy + 1
+= D(h\|\pi(x)h(y)) - D(h\|p_\lambda)
+$$
+
+\paragraph{Convexity}
+
+The convexity of any function of the form $A(\lambda)=\log z(\lambda)$ with $z(\lambda)=\int\pi(x)e^{\lambda^\top f(x)}dx$ can be proved using H\"older inequality:
+$$
+\int f_1 f_2 \leq 
+\left(\int f_1^{\frac{1}{\lambda}} \right)^\lambda  
+\left(\int f_2^{\frac{1}{1-\lambda}} \right)^{1-\lambda}  
+$$
+for $f\succeq 0$, $g\succeq 0$, $0\leq\lambda\leq 1$.
+Just do $f_1=\exp(\lambda\theta^\top f)$ and similarly for $f_2$ with $1-\lambda$.
+This proves that $A_y(\lambda) = \log z(\lambda,y)$ is convex for any~$y$  and the fact remains true by averaging $A_y(\lambda)$ over several $y$, therefore $-\psi(\lambda)$ is convex.
+
+There is another, more insightful way to prove that $A(\lambda)$ is convex. Consider the variable $F=f(X)$ with $X\sim p_\lambda(x)$. It is easy to verify the following facts:
+$$
+\nabla A = \frac{\nabla z}{z} = E(F),
+\qquad
+\nabla\nabla^\top A 
+= \frac{\nabla\nabla^\top z}{z} 
+- \frac{\nabla z \nabla z^\top}{z^2}
+= {\rm Var}(F),
+$$
+so the Hessian is a variance matrix and is therefore positive, which proves convexity.
+How about definite positiveness?
+To stand a chance, we need the number of features to be less than the number of labels otherwise the basis vectors are linearly dependent.
+
+Let $m_F(y)$ and $V_F(y)$ respectively denote the mean and variance of $f(x,y)$ conditional on~$y$. Denoting by $\langle \rangle$ the empirical mean wrt $y$:
+$$
+\nabla \psi = \mu - \langle m_F \rangle
+$$
+
+$$
+\nabla \nabla^\top \psi = - \langle V_F \rangle
+$$
+
+In general, we have more features than classes for good prediction performance so each matrix $V_F(y)$ is singular and tends to be smaller (in terms of spectral radius) as the number of features increases.
+The Hessian is a sample average of these matrices and will thus follow the same trend.
+
+In practice, it looks like the problem is always nearly singular so we may need regularization!
+
+\paragraph{Damping}
+
+We can regularize the Maxent problem via a $L_2$ penalty, yielding:
+$$
+\psi_r(\lambda) = \psi(\lambda) - \alpha \|\lambda\|^2
+$$
+
+This implements a callback force towards the prior $\pi(x)$. Can be interpreted as some way of saying that our trust in the empirical information is not infinite.
+
+\paragraph{Utility}
+
+Let us consider a reference target value $x_0$. By definition,
+$$
+p_\lambda(x_0|y) = \frac{\pi(x_0)}{z(\lambda,y)} e^{\lambda^\top f(x_0,y)}
+$$
+
+Therefore,
+$$
+\log z(\lambda,y) = \lambda^\top f(x_0,y) - \log \frac{p_\lambda(x_0|y)}{\pi(x_0)}
+$$
+
+We can use this fact to rewrite the dual function:
+$$
+\psi(\lambda) = \lambda^\top (\mu - \mu_0) + \int h(y) \log \frac{p_\lambda(x_0|y)}{\pi(x_0)} dy
 $$
@@
 $$
-z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
-$$
-
-$$
-\nabla\psi(\lambda)
-= F - \int \pi(x) h(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
-$$
-
-$$
-\nabla\nabla^\top\psi(\lambda)
-= - \int \pi(x) h(y) f(x,y)f(x,y)^\top e^{\lambda^\top f(x,y)} dx dy
-$$
-
-\section{Minimally discriminative model}
-
-Let $\pi(x)$ some reference distribution that represents full
-uncertainty about $X$. We wish to select the joint distribution
-$p(x,y)$ that minimizes:
-$$
-I(p) = \int p(x,y) \log \frac{p(x|y)}{\pi(x)} dy,
-$$ subject to feature mean-value constraints of the form:
-$$
-\int p(x,y) f(x,y) dx dy = \mu.
-$$
-
-There are two special cases of this problem in the literature.  On the
-one hand, the {\em maximum entropy classifier} \cite{BergerA-96}
-incorporates the constraint that $p(y)$ be known, however we will see
-that this is not necessary. The {\em minimally informative likelihood}
-method \cite{Yuan-99b,Yuan-99}, on the other hand, imposes that
-$p(x)=\pi(x)$, hence assuming the form $p(x,y)=\pi(x)p(y|x)$. We won't
-make such assumptions here and will consider the case where $\mu$ is
-completely unknown.
-
-The above problem is seen to be equivalent to minimizing the auxiliary
-objective function:
-$$
-I(p,m) 
-= \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy,
-$$ over $p$ and $m$, subject to the same constraints on $p$. We note
-that $I(p,m)=I(p)+D(p_y\|m)$, showing that the auxiliary function
-essentially adds a penalty term to the actual objective in order to
-force $p(y)$ close to $m(y)$.
-
-Minimizing $I(p,m)$ along both $p$ and $m$ is basically a minimum KL
-divergence problem between two convex distribution spaces, so there
-must be a unique solution {\bf -- to be checked in
-  \cite{Cover-91}}. It is similar to the rate-distortion problem in
-information theory, which may be solved by alternate minimization,
-yielding a Blahut-Arimoto algorithm.
-\begin{itemize}
-\item {\em Let's call it A-step}. Optimize $p(x,y)$ at fixed $m$
-  s.t. constraint:
-$$
-\exists \lambda, \qquad
-p_{\lambda,m}(x,y) = \frac{1}{Z(\lambda, m)} \pi(x) m(y) e^{-\lambda^\top t(x,y)}
-$$
-The actual $\lambda$ is found my maximizing the dual function
-$\psi(\lambda,m)$, see below.
-\item {\em Let's call it B-step}. Optimize $m(y)$ at fixed $p(x,y)$:
-$$
-m(y) = \int p(x,y) dx
-$$
-\end{itemize}
-
-Upon convergence, the algorithm outputs both joint and marginal
-distributions $p_{\star}(x,y)$ and $m_{\star}(y)$ that both depend on
-$\mu$. By construction, we have that $p_{\star}(y) = m_{\star}(y)$,
-therefore $p_\star(x|y)=p_{\star}(x,y)/m_{\star}(y)$.
-
-\section{Comparison with maximum entropy classifier}
-
-Lagrangian...
+\mu_0 = \int h(y) f(x_0, y) dy
+$$
+
+The first term $\lambda^\top (\mu-\mu_0)$ is essentially the same thing as the utility considered in my previous version of composite likelihood training (which was maximized on the simplex).
+Utility is label-dependent as it involves the moments.
+If the basis consists of feature-based log-likelihood functions, it measures the discrepancy vs the reference and pushes towards large coefficient values.
+Cross-entropy appears here as a regularized version of utility.
+The regularization term $E[\log p_\lambda(x_0|y)$ replaces the hard constraint that $\lambda$ lie in the simplex: it forces the reference class to be probable for any input, so it will tend to discard predictions that are peaked on a single class if it's not $x_0$.
+This term acts as a callback force.
+
+\subsection{Minimally informative likelihood}
+
+In this case, the constraint $p(y)=h(y)$ no longer applies. Consider:
+$$
+D(p,m) = \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dx dy
+$$
+and note that the prediction objective is: $D(p) = \min_m D(p,m)$. This is the Blahut-Arimoto trick, which enables us to consider an alternate optimization strategy wrt $p$ and $m$, respectively.
+
+At fixed $p$, this is a maximum (joint) relative entropy problem. Lagrangian...
 $$
@@
 =
-\int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy
+- \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy
 +
@@
 \int p(x,y) f(x,y) dydy - \mu 
+\right) \\
+$$
+
+The derivative is given by:
+$$
+\frac{\partial\cal L}{\partial p(x,y)}
+= 
+- \log \frac{p(x,y)}{\pi(x)m(y)} - 1
++ \lambda^\top f(x,y)
+$$
+hence the optimal model at fixed $m$ has the form:
+$$
+p_{\lambda,m}(x,y) = \frac{1}{z(\lambda,m)} \pi(x) m(y) e^{\lambda^\top f(x,y)} 
+$$
+with:
+$$
+z(\lambda,m) = \int \pi(x) m(y) e^{\lambda^\top f(x,y)} dx dy
+$$
+
+Note that the induced posterior distribution $p_{\lambda,m}(x|y)\propto \pi(x)p_{\lambda,m}(y|x)$ has the same form as in the conditional maxent case (if using the very same moment constraints).
+
+Dual function...
+$$
+\psi(\lambda,m) 
+= \lambda^\top \mu - \log z(\lambda, m)
+$$
+
+$$
+\nabla z = \int \pi(x) m(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
+$$
+
+$$
+\nabla\nabla^\top z = \int \pi(x)m(y) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dxdy
+$$
+
+$$
+\nabla \psi = \mu - \frac{\nabla z}{z} 
+$$
+
+$$
+\nabla\nabla^\top \psi = 
+- \left(
+\frac{\nabla \nabla^\top z}{z} 
+- \frac{\nabla z \nabla z^\top}{z^2}
 \right) 
-$$ where $\lambda$ represents a vector-valued Lagrange multiplier. At
-fixed $m$, the solution has the form:
-$$
-p_{\lambda,m}(x,y) = \frac{1}{Z(\lambda,m)}
-\pi(x) m(y) e^{-\lambda^\top f(x,y)} 
-$$
-
-Note that this implies that the conditional distribution is
-independent of $m$ once $\lambda$ is determined,
-$$
-p_\lambda(x|y) = \frac{1}{z(\lambda,y)} \pi(x) e^{-\lambda^\top f(x,y)} 
-$$
-
-Also, the marginal is a modulation of $m(y)$:
-$$
-p_{\lambda,m}(y) = \frac{z(\lambda,y)}{Z(\lambda,m)} m(y)
-$$
-
-Dual function at fixed $m$:
-$$
-\psi(\lambda,m) 
-\equiv \min_p {\cal L}(p,m,\lambda)
-= 
-- \log Z(\lambda,m) - \lambda^\top \mu
-.
-$$
-
-An alternative expression is:
-$$
-\psi(\lambda, m)
-= 
-\int h(x,y) 
-\log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(y)} dxdy,
-$$ where $h(x,y)$ is any distribution statisfying the
-constraints. This shows that maximizing the dual function at fixed $m$
-is essentially the same as minimizing the KL divergence
-$D(h\|p_{\lambda,m})$ over $\lambda$, in other words fitting $h(x,y)$
-by some distribution of the form $p_{\lambda,m}$. The fact that the
-result does not depend on the particular $h(x,y)$ that is chosen, as
-long as it satifies the constraint, is a general property of
-exponential families.
-
-We also have that the dual function associated with $I(p)$ reads:
-\begin{eqnarray*}
-\psi(\lambda) 
- & = & \min_m \psi(\lambda, m)\\
- & = & \int h(x,y) \log \frac{p_{\lambda}(x|y)}{\pi(x)} dxdy\\
- & = & -\int h(y) \log z(\lambda,y) dy - \lambda^\top \mu
-\end{eqnarray*}
-
-But, wait, that's exactly what we get in the maximum entropy
-classifier! So, at the end of the day, we simply got an alternative
-method to learn $\lambda$ in the maximum entropy classifier, i.e. the
-Blahut-Arimoto algorithm as opposed to a brute-force maximization of
-$\psi(\lambda)$. Both methods will converge to the same $\lambda$...
-
-What it essentially means is that the optimal $\lambda$ is insensitive
-to $m(y)$ as long as $m(y)$ is compliant in the sense that:
-$$
-m(y) = \int h(x,y) dx,
-$$ for some distribution $h(x,y)$ statisfying the constraint. This is
-true for the optimal $m_\star(y)$ output by the BA algorithm, but
-also, for instance, for the empirical distribution of observations in
-a training dataset used to estimate $\mu$, as proposed in
-\cite{BergerA-96}. The only special property of $m_\star(y)$ is to
-yield the full Bayesian model $p_\star(x,y)=p_\star(x|y)m_\star(y)$
-that minimizes the discrimination information. We may not care too
-much about that in practice since we will only use $p_\star(x|y)$.
-
-\section{Comparison with minimally informative likelihood}
-
-Yuan \cite{Yuan-99} considered the situation where we add the
-constraint that $p(x)=\pi(x)$ to the minimum discrimination inference
-problem. Unknown is then the conditional distribution
-$p(y|x)$. Lagrangian...
-\begin{eqnarray*}
-{\cal L}(p,m,\lambda)
- & = & 
-\int \pi(x)p(y|x) \log \frac{p(y|x)}{m(y)} dxdy
-+
-\lambda^\top \left( 
-\int \pi(x)p(y|x) f(x,y) dydy - \mu 
-\right) \\
- & = & 
-\int \pi(x)
-\left( 
-\int
-p(y|x) \log \frac{p(y|x)}{m(y)} dy
-+
-\lambda^\top 
-\int p(y|x) f(x,y) dy
-- \mu 
-\right)
-dx 
-\end{eqnarray*}
-
-The derivative is given by,
-$$
-\frac{\partial\cal L}{\partial p(y|x)}
-= 
-\pi(x)\left[ 
-1 + \log \frac{p(y|x)}{m(y)} 
-+ \lambda^\top f(x,y)
-\right],
-$$
-hence the optimal model at fixed $m$ has the form:
-$$
-p_{\lambda,m}(y|x) = \frac{1}{Z(\lambda,m,x)} m(y) e^{-\lambda^\top f(x,y)} 
-$$
-
-Note that the induced posterior distribution $p_{\lambda,m}(x|y)$ has
-a different form from above unless the normalizing factor
-$Z(\lambda,m,x)$ turns out independent from $x$. If this is not the
-case, we no longer have the property that $p_{\lambda,m}(x|y)$ is
-independent from $m$ given $\lambda$.
-
-Dual function...
+$$
+
+Equivalently, for any distribution $h(x,y)$ that satisfies the moment constraint, we have:
 $$
@@
 =
-- \int \pi(x) \log Z(\lambda, m, x) dx
-- \lambda^\top \mu
-$$
-
-Equivalently, for any distribution under the form $\pi(x)h(y|x)$ that
-satisfies the moment constraint, we have:
-$$
-\psi(\lambda,m) 
-=
-\int \pi(x) h(y|x) \log \frac{p_{\lambda,m}(y|x)}{m(y)} dxdy
-$$
-
-Now, the real question is why should we constrain $p(x)$, which boils
-down to a Bayesian prior in this context, to be the same as our
-reference $\pi(x)$? It only makes sense if we want our inference to
-stick to a generative modeling paradigm... but haven't we already
-given up on that? Therefore, unless we find a good reason not to, we
-won't impose the $p(x)=\pi(x)$ constraint, thereby allowing for a
-discrepancy between the reference and the prior.
-
-\section{Discriminative vs. semi-discriminative}
-
-Let's go back to the equivalence we found between our approach and the
-maximum entropy classifier (MCE) \cite{BergerA-96}. We have said that
-the former is essentially a re-formulation of MCE.
-
-But the re-formulation also conveys a generalization of MCE if,
-instead of letting $m$ being an arbitrary distribution, we restrict
-its search space to some set of acceptable reference
-distributions. Would such a strategy be useful?
-
-Recall that the method selects the joint distribution that minimizes
-discriminative information, as defined by $I(p)$:
-$$
-I(p) 
-= \int p(x,y)\log\frac{p(x|y)}{\pi(x)} dxdy
-= E_Y[D(p_{x|y}\|\pi)]
-$$
-
-The latter characterization reminds us that discriminative information
-is defined in an average sense. The corresponding posterior $p(x|y)$
-may not be conservative for some $y$, in particular those that are
-unlikely under $p(y)$. It would be a problem if that's the case for
-the particular data at hand. For that to happen rarely, the ideal
-$p(y)$ would be the ``true'' marginal distribution of $Y$.
-
-However, for vague mean value constraints, nothing may prevent the
-optimal $m_\star(y)$ from departing significantly from that ideal
-distribution. In fact, we can show that $m_\star(y)$ only has mass at
-$y$ values that maximize $z(\lambda_\star,y)$, and is therefore likely
-sparse. This is because $m_\star$ minimizes $\psi(\lambda_\star,m)$,
-which is equivalent to maximizing $Z(\lambda_\star,m)$ and we have:
-$$
-Z(\lambda, m) = \int m(y) z(\lambda, y) dy.
-$$
-
-In other words, the solution to our problem may be singular! It does
-not hurt since, as discussed above, one may alternatively fix $m(y)$
-to some pre-defined distribution... but this, in practice, restricts
-the method to supervised learning situations.
-
-We may not face this issue with the MIL method, which further imposes
-the prior on $X$, so all the information from the constraints goes
-into specifying a possibly reasonable generative model.
+\int h(x,y) \log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(y)} dxdy
+= D(h\|\pi_xm_y) - D(h\|p_{\lambda,m})
+$$
+showing that this is a generative learning problem!
+
+The BA algorithm alternates between estimating $\lambda$ at fixed $m$ via the maximization of $\psi(\lambda,m)$ and estimating $m$ at fixed $\lambda$ by marginalization:
+$$
+m(y) = \int p_{\lambda,m}(x,y) dx
+$$
+
+The MIL seems intractable if we don't impose constraints on $m(y)$.
+It becomes feasible by restricting $m(y)$ to have a support limited to the training data, similarly to the conditional maxent implementation:
+$$
+m(y) = \sum_k w_k \delta(y-y_k),
+$$
+hence $m$ is determined by a finite set of weights $w_k$. In practice, the optimal weights look very sparse.
+
+{\color{red} Big fat speculation: wouldn't MIL make more sense when passing generative mean-value constraints, i.e.
+empirical mean and variance per class?
+This could avoid the double learning stage!
+We don't do it with Maxent because we want feature-based generative models for interpretation purposes (if we did, it would be logistic regression).
+But using MIL, we get a full generative model for free...
+which is a game changer.
+It's no longer composite!
+The trouble is, this model is kinda degenerate (due to its ``empirical'' style, which is needed for computational reasons): the likelihood of any new~$y$ is almost surely zero.
+So not very good to do hypothesis testing.
+We have an interpolation problem, Houston.
+And that interpolation problem is the very reason why we trashed discriminative models from the start!
+So maybe not the way to go.}
 
@@
 
+\section{Log-likelihood ratio}
+
+$$
+\log r = \log \frac{N(x;\mu_1,\sigma_1)}{N(x;\mu_0,\sigma_0)}
+$$
+
+$$
+\log N(x;\mu,\sigma) = -\frac{1}{2}\left[
+\log(2\pi\sigma^2) + \frac{(x-\mu)^2}{\sigma^2}
+\right]
+$$
+
+$$
+\log r = -\frac{1}{2}
+\left[
+\log \frac{\sigma_1^2}{\sigma_0^2}
++ \frac{(x-\mu_1)^2}{\sigma_1^2} - \frac{(x-\mu_0)^2}{\sigma_0^2}
+\right]
+$$
+
+It follows:
+\begin{eqnarray*}
+\log r 
+ & = & 
+-\frac{1}{2}\left[
+\log \frac{\sigma_1^2}{\sigma_0^2}
++ \frac{\sigma_0^2(x-\mu_1)^2 - \sigma_1^2(x-\mu_0)^2}{\sigma_0^2 \sigma_1^2}
+\right] \\
+ & = & 
+\frac{1}{2}\left[
+\log \rho^2
++ \frac{(\rho^2-1)x^2 - 2(\rho^2\mu_0-\mu_1)x+ (\rho^2\mu_0 ^2-\mu_1^2)}{\sigma_0^2}
+\right]
+\end{eqnarray*}
+with $\rho=\sigma_1/\sigma_0$. In the case of same variance ($\rho=1$), this simplies to an affine function of $x$ with offset given by the center of $(\mu_0,\mu_1)$:
+$$
+\log r = \frac{\mu_1-\mu_0}{\sigma^2} \left(
+x - \frac{\mu_0 + \mu_1}{2}
+\right) 
+$$
+
+In any case,
+$$
+E[\log r | H_1] = 
+\frac{1}{2}
+\left[
+- \log \frac{\sigma_1^2}{\sigma_0^2}
++ \frac{\sigma_1^2 + (\mu_1-\mu_0)^2}{\sigma_0^2}
+- 1
+\right]
+$$
+
+Let $\delta = (\mu_1-\mu_0)/\sigma_0$. We have:
+$$
+E[\log r | H_1] = 
+\frac{1}{2}
+\left[
+\delta^2
+- \log \rho^2
++ \rho^2  - 1
+\right]
+$$
+
+\section{Feature relevance}
+
+Let $x_0$ a reference class and consider the log-probability ratio:
+$$
+\alpha = \log \frac{p_\lambda(x|y)}{p_\lambda(x_0|y)}
+$$
+
+The larger $\alpha$ for a given $x_0$, the happier we are.
+Let us average $\alpha$ wrt $x$ and $y$ as well as $x_0\sim \pi(x_0)$, which in this context may be interpreted as the outcome of a random classifier.
+Ignoring offsets, we have:
+$$
+E[\alpha]
+= \sum_i \lambda_i \int \pi(x_0)\pi(x) p(z_i|x) \log\frac{p(z_i|x)}{p(z_i|x_0)} dz_i dx dx_0
+= \sum_i \lambda_i \bar{d}_i
+$$
+with:
+$$
+\bar{d}_i =
+\int \int \pi(x_0) \pi(x) D(p_i^x\|p_i^{x_0}) dx dx_0
+$$
+
+If we have offsets, then there is an offset in this expresion, meaning that part of the log-probability ratio is accounted for by the offsets... this is not important.
+
+Hence, the expected log-likelihood ratio is a weighted sum of positive feature-specific contributions $r_i=\lambda_i \bar{d}_i$.
+The larger the contribution, the more influential the feature on the average.
+The analysis can be refined by skipping the averaging over $x_0$, in which case feature influence becomes dependent on the comparison class $x_0$...
+so it's all more complicated and potentially confusing.
+It would make sense to do that in the super-composite framework, where the reference class is fixed.
+
+The bottom line is that global feature relevance is determined by the CL weight(s) but also by the KL divergence matrix.
+
+Note that, by the information processing inequality, we have $D(p_i^x\|p_i^{x_0}) \leq d_{\rm lim}(x,x_0)\equiv D[p(y|x)\|p(y|x_0)]$, hence $\bar{d}_i\leq \bar{d}_{\rm lim}$ and:
+$$
+r_i \leq \lambda_i \bar{d}_{\rm lim}
+$$
+so if we normalize relevance according to unit sum weights, no single feature can be more relevant than the full data.
+
+How about {\em local} relevance? We simply replace $d_{k\ell i}$ by the absolute empirical log-likelihood ratio $|\log p_k(z_i)/p_\ell(z_i)|$. That's it.
+
 \bibliographystyle{abbrv}

### 2019-02-06 papers/composite_inference/draft.tex

@@
 
-%\newcommand{\fix}{\marginpar{FIX}}
-%\newcommand{\new}{\marginpar{NEW}}
+\newcommand{\E}{\mathbb{E}}
+\newcommand{\lda}{{\boldsymbol{\lambda}}}
 %\newcommand{\matphi}{\boldsymbol{\Phi}} 
@@
 
-% which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes conditional entropy relative to a given prior and subject to multiple mean log-likelihood inequality constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule.
-{\color{red} Confusing: we simply maximize cross-entropy, which turns out to have a conditional maxent interpretation.} We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, which can be powerful in shallow learning and transfer learning problems.
-{\color{red} Rephrase this: extends Bayes and keeps introspection ability, which is crucial in some AI problems.
-
 \section{Introduction}
@@
 Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this requirement may be very challenging for multi-dimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include ``shallow'' learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
-
-%Interpretability needed in life-impacting applications of AI. 
-%Whenever possible, proxy features should be avoided as they make models gameable. 
-%Local Interpretability for a Single Prediction: Why did the model make a certain prediction for an instance?
-%What about sensitivity analysis: study the prediction as a function of a single feature...
-
-Discriminative models, however, are mere inference recipes and lack the built-in ability to ``support'' their predictions.
-Consider as a straightforward example the problem of deciding whether a pan is a sauce pan or a frying pan based on shape measurements.
-A simple threshold on the pan depth could work well in most practical cases, but only makes sense if known the fact that sauce pans tend to be deeper than frying pans, which is generative information left implicit in the decision rule.
-For multi-dimensional features, it quickly becomes impossible to relate adjusted boundaries between classes to generative information {\em a posteriori}.
-
-%Such common knowledge is rarely available in real-world apps.
-
-% merely trace boundaries between classes in the data representation space and therefore lack built-in information to interpret their predictions.
+Because this double requirement may be very challenging for multi-dimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in the minimum description length paradigm \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include ``shallow'' learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
+
+Discriminative models, however, lack the ability to establish causality links between data instances and predictions, and therefore cannnot ``support'' their own predictions in a built-in manner.
+Consider as a straightforward example the problem of deciding whether a pan is a sauce pan or a frying pan based on its depth.
+A discriminative model can learn an optimal threshold that correctly classifies most pans in practice, but it cannot gain an internal representation that sauce pans tend to be deeper than frying pans, which is {\em why} the thresholding works.
+To understand a discriminative model, we need to ``reverse engineer'' it using external knowledge, which quickly becomes an impossible task as the data dimension increases.
 
 There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
-For instance, in medical diagnosis, automated disease predictions should always be corroborated by individual findings to bring real clinical value -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like clinicians.
-Clinicians have long used methods based on simple generative models, such as range analysis of univariate biomarkers, that can provide decisive clues for diagnosis despite being outperformed by discriminative models for blind disease prediction.
+For instance, in medical diagnosis, automated disease predictions should always be corroborated by individualized findings to bring communicable clinical value -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
+Clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
 
@@
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graphs representing generative and discriminative models.
-For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence requires larger training sets to achieve optimal performance \cite{Ng-01}.
+For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence requires a larger training set to achieve optimal performance \cite{Ng-01}.
 
@@
 This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a means to incorporate generative information into a discriminative model.
-Rather than attempting at a full generative model, one may combine multiple low-dimensional generative models for different low-dimensional pieces of information extracted from the input data.
-Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
+Rather than attempting at a full generative model, one may combine multiple low-dimensional generative models for different pieces of information extracted from the input data.
+Each ``small'' model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
 The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
@@
 
-%use of low-dimensional generative models.
-
-When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be (up to a multiplicative prior) a quantity known as composite likelihood (CL) in computational statistics.
-As described in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
-While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is asymptotically consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
-
-%the data distribution is not modeled as a whole but separate feature-based generative models are available
-
-%CL is a semi-generative approach to statistical inference that extends the familiar notion of likelihood without requiring a full generative model.
-The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole.
-This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood.
-While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.
-
-%An open question regarding CL is how to choose the weights associated with the respective features.
+When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known as composite likelihood (CL) in computational statistics.
+As detailed in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
+While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} and generally offers an excellent trade-off between computational and statistical efficiency.
 
@@
 Technical details are given in the remainder.
-
-%The relevance of each feature in the CL is determined by an arbitrary weight used to scale the associated likelihood.
-For classification performance, the weights can be tuned in a typical supervised learning scenario so as to concurrently maximize discriminative likelihood (or, equivalently, minimize cross-entropy), thereby making the method hybrid between generative and discriminative modeling approaches.
-We shall note that this particular training strategy amounts to building a maximum entropy classifier \cite{BergerA-96} that uses the feature-based log-likelihoods as basis functions.
-
-%{\color{red} Still, note that the likelihood functions themselves are trained beforehand so this is slightly different from a standard maxent classifier aka logistic regression.
-Double training phase.}
-
-%{\color{red} Classical composite likelihood not studied with nuisance parameters. In our scenario, nuisance parameters are learned beforehand.}
-
-%{\color{red} Interpretability of composite likelihood.
-Obvious from the construction.
-Can achieve prediction performance comparable with fully discriminative models while maintaining the interpretability of generative models.}
-
-%We further argue that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum relative conditional entropy.
-This perspective entails a method to optimize the weights of the different agents from training data as in a typical discriminative learning scenario.
-
-%Something for true statisticians.
-Maybe we could clarify that we use the term ``Bayesian'' in the sense of ``empirical Bayesian'', implying that the parameters $\theta$ are to be estimated somehow contrary to a ``full Bayesian'' approach where they would be integrated out at inference time.
 
@@
 \label{eq:log_pool}
-p_\lambda(x|y) \propto \nu(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
+p_\lda(x|y) \propto \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
 \end{equation} 
-where $\nu(x)$ is some arbitrary reference distribution (which may differ from a Bayesian prior as we shall see below), and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$.
+where $\pi(x)$ is a prior distribution, and $\lda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lda(x_1|y)\leq p_\lda(x_2|y)$.
 Also note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
@@
 $$
-p_\lambda(x|y)\propto \nu(x) L^c_\lambda(x),
-$$
-where the distribution $\nu(x)$ plays the role of a prior, and the function:
+p_\lda(x|y)\propto \pi(x) L^c_\lda(x),
+$$
+where the distribution $\pi(x)$ plays the role of a prior, and the function:
 \begin{equation}
 \label{eq:comp_lik}
-L^c_\lambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
+L^c_\lda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
 \end{equation} 
-plays that of a traditional likelihood function.
-This expression happens to be known as a {\em marginal composite likelihood} \cite{Varin-11}.
+plays that of a likelihood function.
+The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
 The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{sec:conditional}.
@@
 
-CL in equation~(\ref{eq:comp_lik}) shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$.
+CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lda\equiv 1$.
 While the above derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method.
@@
 A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$y$.
-Given two target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ simply adds up single feature contributions:
-$$
-\log \frac{p_\lambda(x_1|y)}{p_\lambda(x_0|y)}
+Given two target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ simply adds up single-feature contributions:
+$$
+\log \frac{p_\lda(x_1|y)}{p_\lda(x_0|y)}
 = 
-\underbrace{\log \frac{\nu(x_1)}{\nu(x_0)}}_{\text{offset}}
+\underbrace{\log \frac{\pi(x_1)}{\pi(x_0)}}_{\text{offset}}
 + \sum_i \underbrace{\lambda_i \log \frac{p(z_i|x_1)}{p(z_i|x_0)}}_{\text{feature contribution}},
 $$
-therefore is straightforward to assess which features contribute most (either positively or negatively), see possible disagreements between features, and corroborate the overall prediction with feature-level testable information.
+therefore it is straightforward to assess which features contribute most (either positively or negatively), see possible disagreements between features, and corroborate the overall prediction with feature-level testable information.
 
 Each feature contributes proportionally to its associated log-likelihood ratio test statistic for the comparison $x_1$ vs~$x_0$.
-Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction, and may therefore account for mutual feature dependence if tuned appropriately.
-Contrary to statistical hypothesis testing, likelihood ratio test statistics are not converted into $p$-values but are directly combined into predictive probabilities.
-In this perspective, composite weights may be seen as an alternative to multiple testing correction.
+Contrary to statistical hypothesis testing, likelihood ratio test statistics are not converted into $p$-values but are directly combined into a predictive probability ratio.
+Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\lda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately.
+Keeping the analogy with hypothesis testing, composite weights may be thought of as the counterpart of multiple testing correction \cite{Benjamini-10}.
 
@@
 
-In order to understand predictive distribution, the user(s) may go through successive pairwise comparisons of target values in a dynamic way.
+To interpret the predictive distribution, the user(s) may go through successive pairwise comparisons of target values in a dynamic way.
 They may start comparing the most probable target, {\em i.e.} the automated prediction, to some educated guess or reference class (for instance, normality in medical diagnosis).
-Next, depending on the feature contributions highlighted by the comparison interpretation, they may form another hypothesis to compare to the automated prediction.
-This process, which can be thought of as a kind of human machine dialog, may be repeated until the composite prediction is felt to be understood by the user(s).
+Next, depending on the feature contributions highlighted by the comparison, they may submit another hypothesis for comparison to the automated prediction.
+This process, which can be thought of as a human machine dialog, may be repeated until the composite prediction is subjectively understood by the user(s).
 
@@
 
+Let us now turn to the problem of tuning the weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in the composite predictive distribution~(\ref{eq:log_pool}).
+
 \subsection{Agnostic calibration}
@@
 The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
-For the purpose of computing credibility sets, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
-
-Ignoring such a goal, another sensible prior recommendation for log-linear pooling weights is to sum up to one.
-This is motivated in \cite{Genest-86b} by the fact that the only pooling operator that preserves external Bayesianity\footnote{External Bayesianity means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.} is the log-linear opinion pool with unit sum weights.
-A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the average Kullback-Leibler (KL) divergence to the agent opinions.
-See also \cite{Wang-14}, Theorem~3, for another information theoretic argument.
-
-Unit sum weights, however, implicitly assume strong redundancy between features and may therefore be ineffective in practice since features are generally chosen to be complementary.
-An ideal method to tune weights is one that can capture and adjust for existing statistical dependences between features.
+To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
+
+Another simpler recommendation for log-linear pooling weights is to sum up to one.
+This may be motivated by the fact that the only pooling operator that preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}.
+External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling opinions, provided that all agents agree on the same prior.
+A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the sum of Kullback-Leibler (KL) divergences to the agent opinions (see also \cite{Wang-14}, Theorem~3, for a related optimality argument).
+
+Unit sum weights, however, implicitly assume strong redundancy between features and thus tend to produce over-conservative predictions when features are weakly correlated, as is generally the case in practice since features are supposed to be chosen to be complementary.
+An ideal method to tune weights is one that can capture and adjust for the pattern of statistical dependences between features.
 This calls for a training approach, which, as we will see, is unlikely to pick constant or unit sum weights.
@@
 
-Brute-force approach: optimize fit for optimal prediction.
-Discriminative model defined by (\ref{eq:log_pool}).
-Training objective, weight likelihood (which, in this context, is negated cross-entropy) better than accuracy:
-$$
-\max_{\lambda\succeq 0} {\cal L}(\lambda),
-\qquad {\rm with} \quad
-{\cal L}(\lambda)\equiv\sum_k \log p_\lambda(x_k|y_k)
-$$
-
-We recognize the dual function $\psi(\lambda)$ of a conditional maxent problem, see Appendix.
-$$
-{\cal L}(\lambda)
-= 
-\sum_k \log \pi(x_k)
-+ n \psi(\lambda)
+Given a labeled training dataset ${\cal D}=\{(x_k, y_k), k=1,\ldots,N\}$, the most direct approach to optimize the CL weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'')  under the composite predictive distribution:
+$$
+\max_{\lda\succeq 0} {\cal L}(\lda)
+\equiv\sum_{k=1}^N \log p_\lda(x_k|y_k).
+$$
+
+From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\lda(x|y)$ spanned by the weights~$\lda$ is an exponential family with natural parameter~$\lda$:
+$$
+p_\lda(x|y) = \pi(x) \exp[\lda^\top \mathbf{f}(x,y) - A(\lda,y)],
+$$
+with:
+$$
+f_i(x,y) \equiv \log p(z_i|x),
+\qquad
+A(\lda, y) \equiv \log \int \pi(x) e^{\lda^\top \mathbf{f}(x,y)} dx.
+$$
+
+It follows that ${\cal L}(\lambda)$ is a concave function as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}.
+It also implies that maximizing ${\cal L}(\lambda)$ is dual to a maximum conditional entropy, or minimum conditional KL divergence problem \cite{BergerA-96}:
+$$
+\min_{p} \int p(x,y) \log \frac{p(x|y)}{\pi(x)} dx dy,
+$$
+subject to the constraints $p(y) = h(y)$ and, $\forall i=1,\ldots,n$, $\E_p[\ell_i]=\E_h[\ell_i]$, 
+where $h(x,y)$ is the joint empirical distribution of targets and data features:
+$$
+h(x,y) = \frac{1}{N} \sum_{k=1}^N \delta(x-x_k)\delta(y-y_k).
 $$
@@
 We can show an example where weights are much more stable by choosing inequality vs equality constraints.
+
+%Brute-force approach: optimize fit for optimal prediction.
+Discriminative model defined by (\ref{eq:log_pool}).
+Training objective, weight likelihood (which, in this context, is negated cross-entropy) better than accuracy:
 
@@
 
-We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\nu(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$.
+{\color{red} To offset or not to offset?}
+
+We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$.
+This ``deforms'' the prior to improve the training score.
+But why is it important to enforce this constraint?
+Again, it all makes things just less interpretable.
 

### 2019-02-07 papers/composite_inference/draft.tex
Modifs 7/2/19

@@
 A discriminative model can learn an optimal threshold that correctly classifies most pans in practice, but it cannot gain an internal representation that sauce pans tend to be deeper than frying pans, which is {\em why} the thresholding works.
-To understand a discriminative model, we need to ``reverse engineer'' it using external knowledge, which quickly becomes an impossible task as the data dimension increases.
+To interpret a discriminative model means to ``reverse engineer'' it using external knowledge, which quickly becomes an impossible task as the data dimension increases.
 
 There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
-For instance, in medical diagnosis, automated disease predictions should always be corroborated by individualized findings to bring communicable clinical value -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
+For instance, in medical diagnosis, automated disease predictions should always be corroborated by individualized findings to be communicable and thus be taken into consideration by clinicians -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
 Clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
@@
 The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
-The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{sec:conditional}.
+The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{appendix:conditional}.
 The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.
@@
 A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$y$.
-Given two target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ simply adds up single-feature contributions:
+Given two putative target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ is computed by simply adding up single-feature contributions:
 $$
@@
 $$
-therefore it is straightforward to assess which features contribute most (either positively or negatively), see possible disagreements between features, and corroborate the overall prediction with feature-level testable information.
-
-Each feature contributes proportionally to its associated log-likelihood ratio test statistic for the comparison $x_1$ vs~$x_0$.
-Contrary to statistical hypothesis testing, likelihood ratio test statistics are not converted into $p$-values but are directly combined into a predictive probability ratio.
+therefore it is straightforward to assess which features contribute most to a particular hypothesis comparison (either positively or negatively), see possible disagreements between features, and corroborate predictive probabilities with feature-level testable information.
+
+Each feature contributes to an hypothesis comparison proportionally to its associated log-likelihood ratio test statistic.
+Contrary to statistical hypothesis testing, likelihood ratios are not converted into $p$-values but are directly combined into a predictive probability ratio.
 Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\lda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately.
@@
 
-% For instance, we can get insight such as: ``this patient is more likely to be AD than healthy because his hippocampus is severly atrophied even though his memory tests are quite normal.''
-
-%So, if a feature yields a significant test but has low weight, it may have little influence on the comparison.
-
-%How about overall feature influence?
-Well, if we want to really interpret the prediction, we shall go through each target comparison that is deemed important for the decision.
-For instance, the system says AD more likely than normal, but is AD more likely than MCI?
-To answer this, we need to apply the procedure to AD vs MCI, and maybe other features will pop up.
-
-To interpret the predictive distribution, the user(s) may go through successive pairwise comparisons of target values in a dynamic way.
-They may start comparing the most probable target, {\em i.e.} the automated prediction, to some educated guess or reference class (for instance, normality in medical diagnosis).
+To interpret the predictive distribution globally, the user(s) may go through successive pairwise hypothesis comparisons in a dynamic way, starting by comparing the automated prediction, {\em i.e.} the most probable target, to some educated guess or reference class (for instance, normality in medical diagnosis).
 Next, depending on the feature contributions highlighted by the comparison, they may submit another hypothesis for comparison to the automated prediction.
-This process, which can be thought of as a human machine dialog, may be repeated until the composite prediction is subjectively understood by the user(s).
-
-%See if the automated prediction makes sense.
-If not, stop, otherwise consider what other class the interpretation could suggest and ask for comparison.
-Repeat the process until all other possibilities are subjectively ruled out or the user decides to reject the prediction.
-Think about this as a human AI dialog.
-
-%This is a recipe for {\em local} interpretation (why did the system make a particular prediction), but what about {\em global} interpretation (how does the system make predictions).
-The thing is, we may not care about global interpretation.
-The only thing we need to know is that the system uses a bunch of features that all potentially contribute to the predictions.
-We could investigate which features tend to contribute most notably and which ones do not {\em on the average} but this would only be interesting design purposes, i.e.
-if we are prepared to use another representation.
-
-%We may represent feature influence by a $K\times K$ matrix with general element $\lambda_i r_i(x_k,x_\ell)$.
-
-%This recipe addresses {\em local} interpretation (why did the system make a particular prediction) as opposed to {\em global} interpretation (how does the system make predictions in general).
+This process may be repeated until the composite prediction is subjectively understood by the user(s), yielding a form of human machine dialog.
 
@@
 
-Let us now turn to the problem of tuning the weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in the composite predictive distribution~(\ref{eq:log_pool}).
+%So far, we have assumed that the composite weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) are known.
+
+Let us now turn to the problem of tuning the composite weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration.
+It is assumed at this stage that the feature generative models are known, which may require a pre-training stage that should not cause technical difficulties as it may use standard parameter estimation techniques.
 
@@
 
-In the absence of knowledge, a default rule is to choose constant CL weights.
+In the absence of knowledge, a default rule is to choose constant weights.
 The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
-To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
+To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic MCLE variance matrix \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
 
@@
 This may be motivated by the fact that the only pooling operator that preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}.
-External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling opinions, provided that all agents agree on the same prior.
+External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling agent opinions, provided that all agents agree on the same prior.
 A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the sum of Kullback-Leibler (KL) divergences to the agent opinions (see also \cite{Wang-14}, Theorem~3, for a related optimality argument).
@@
 
-Given a labeled training dataset ${\cal D}=\{(x_k, y_k), k=1,\ldots,N\}$, the most direct approach to optimize the CL weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'')  under the composite predictive distribution:
-$$
-\max_{\lda\succeq 0} {\cal L}(\lda)
+Given a labeled training dataset ${\cal D}=\{(x_k, y_k), k=1,\ldots,N\}$, the most direct approach to tune the composite weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'') under the composite predictive distribution:
+\begin{equation}
+\label{eq:train_likelihood}
+\max_{\lda\succeq 0} U(\lda)
 \equiv\sum_{k=1}^N \log p_\lda(x_k|y_k).
-$$
-
-From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\lda(x|y)$ spanned by the weights~$\lda$ is an exponential family with natural parameter~$\lda$:
-$$
-p_\lda(x|y) = \pi(x) \exp[\lda^\top \mathbf{f}(x,y) - A(\lda,y)],
+\end{equation}
+
+From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\lda(x|y)$ spanned by the weights~$\lda$ is the exponential family with natural parameter~$\lda$ and basis functions given by the feature-based log-likelihoods:
+$$
+p_\lda(x|y) = \pi(x) \exp[\lda^\top \mathbf{L}(x,y) - A(\lda,y)],
 $$
@@
 $$
-f_i(x,y) \equiv \log p(z_i|x),
+L_i(x,y) \equiv \log p(z_i|x),
 \qquad
-A(\lda, y) \equiv \log \int \pi(x) e^{\lda^\top \mathbf{f}(x,y)} dx.
-$$
-
-It follows that ${\cal L}(\lambda)$ is a concave function as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}.
-It also implies that maximizing ${\cal L}(\lambda)$ is dual to a maximum conditional entropy, or minimum conditional KL divergence problem \cite{BergerA-96}:
-$$
-\min_{p} \int p(x,y) \log \frac{p(x|y)}{\pi(x)} dx dy,
-$$
-subject to the constraints $p(y) = h(y)$ and, $\forall i=1,\ldots,n$, $\E_p[\ell_i]=\E_h[\ell_i]$, 
-where $h(x,y)$ is the joint empirical distribution of targets and data features:
-$$
-h(x,y) = \frac{1}{N} \sum_{k=1}^N \delta(x-x_k)\delta(y-y_k).
-$$
-
-Maxent defined by constraints given by mean log-likelihood values.
-Weaker than assuming feature distributed according to agent models.
-Can be seen in terms of detection threshold: all we care about is that our feature-based log-likelihoods reach guaranteed levels (on the average) for the true class.
-Hence inequality constraints are more natural than equality ones.
-Model is admissible if agent compression rates are achieved or surpassed.
-
-We know more about the model, so this is relaxation.
-Doesn't make optimization easier but, I speculate, the solution is more stable than if we were imposing feature-based marginals.
-Both search spaces are determined by the empirically estimated likelihood parameters, intuitively I would say that the bigger space is less sensitive to estimation errors.
-We can show an example where weights are much more stable by choosing inequality vs equality constraints.
-
-%Brute-force approach: optimize fit for optimal prediction.
-Discriminative model defined by (\ref{eq:log_pool}).
-Training objective, weight likelihood (which, in this context, is negated cross-entropy) better than accuracy:
-
-%From a machine learning perspective, it is natural to tune the weights so as to minimize some prediction error, for example conditional cross-entropy.
-This brings the question: why then do it with ``pre-learned'' likelihood functions rather than with basis functions spanning the same predictive distribution family?
-Because of potential overfitting.
-The whole point of this note is that it may be better to run two learning phases than a single fully blown discriminative learning due to overfitting.
-It is a kind of tradeoff between pure generative and pure discriminative learning.
-
-%The ``hybrid'' strategy necessarily achieves a larger cross-entropy than the pure discriminative one on the training dataset, but might generalize better. Should show examples of this.
-
-%The rest of the discussion, i.e. the link with maximum entropy, is less important beside showing that the composite likelihood form is somehow optimal given some knowledge.
-
-Intuition behind weights: Lagrange multipliers to enforce a constraint...
-From empirical observation, there is a non-linear correlation between CL weight and feature-level discrimination power (as measure e.g.
-by maximum KL divergence).
-Non-linear means that features with average discrimination power may get low weight as they are superseded by the most discriminative features.
-But some features with low discrimination power may also get large weight, which does not mean that they are very influential in practice (see above).
-This is illustrated in Figure~\ref{fig:disc_weight_plot} with the UCI breast cancer dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}} used to predict whether breast lumps are benign or malignant based on a set of features extracted from images of fine-needle aspirates.
+A(\lda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\lda^\top \mathbf{F}(x,y)}.
+$$
+
+It follows that the utility function $U(\lda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:maxent} for details.
+It also implies that maximizing $U(\lda)$ is dual to a minimum KL~divergence problem \cite{Csiszar-84}:
+$$
+\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,y) \log \frac{p(x,y)}{\pi(x)h(y)},
+$$
+subject to the constraints $p(y)=h(y)$ and $\E_p[\mathbf{L}] \succeq \E_h[\mathbf{L}]$, where 
+${\cal Y}=\{y_1,\ldots,y_N\}$ is the set of training data values, ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$, $h(x,y)$ is the joint empirical distribution of targets and data, and $h(y)$ is the corresponding marginal distribution.
+
+Therefore, the optimal composite model may be interpreted as the $I$-projection of the independence distribution $\pi(x)h(y)$ onto a set of admissible joint distributions defined by two types of contraints.
+The marginal constraint $p(y)=h(y)$, on the one hand,  ensures that only the conditional distribution $p(x|y)$ is ``updated'', making the $I$-projection a special case of maximum conditional entropy \cite{BergerA-96}.
+The mean log-likelihood value constraints $\E_p[\mathbf{L}] \succeq \E_h[\mathbf{L}]$, on the other hand, encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
+This is weaker a constraint than assuming that the features are distributed according to the generative models $p(z_i|x)$: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.
+
+A particularly insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is under broad conditions a minimax strategy for prediction according to the log loss.
+Hence, any $I$-projection is in some sense a best possible predictive model under particular constraints.
+Being more specifically a maximum conditional entropy method, trained composite inference belongs to the same family as multinomial logistic regression, only differing by the choice of constraints, which, in the latter case, are not related to feature generative models.
+
+In the $I$-projection framework, the composite weights are the Lagrange multipliers associated with the mean log-likelihood inequality constraints.
+Weights corresponding to inactive constraints vanish, hence the training has a natural tendency to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that the training differs significantly from univariate feature selection as it takes feature redundancy into account.
 
@@
   \end{center}
-\caption{Distribution of optimal composite weights vs individual feature discrimination power in the breast cancer UCI dataset.
-Discrimination is measured by the maximum KL divergence between class-conditional generative distributions.
-The most discriminative feature, `worst concave points', has a moderate weight of $0.34$ while `worst texture' achieves a weight of $0.78$ but is among the least discriminative features.}
+\caption{Distribution of optimal composite weights vs individual feature discrimination power in a binary classification task (breast cancer UCI dataset).
+Discrimination is measured by the maximum KL divergence between class-conditional generative distributions.}
 \label{fig:disc_weight_plot}
@@
 
-Alternative: equality constraints (negative weights), less interpretable.
-
-Data weights: 
-$$
-w_k = \frac{\pi(x_k)}{n(x_k)}
-$$
-
-Need to weight the data to adjust to prior as a general measure when minimizing cross-entropy, which should yield a conditional model that approximates the ``true'' conditional distribution, $q(x|y)\approx p(x|y)$.
+As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (obtaining balanced classes if the prior is uniform).
+This does not affect the above maths provided that the data weights are reflected in the empirical distribution~$h(x,y)$.
+
+Other variants of the training scheme that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family.
+For instance, class-dependent offsets may be learned concurrently with the composite weights, which amounts to learning the ``prior'' in the training phase.
+It is also possible to replace the mean log-likelihood inequality constraints with equality constraints: the maximization of $U(\lda)$ is then unconstrained, therefore some trained composite weights may be negative.
+Finally, the mean log-likelihood constraints may be expressed for each class rather than being averaged wrt the prior: this leads to an extension that we call ``super composite likelihood'', in which weights become class-dependent (see Appendix~\ref{appendix:super}).
+
+%used to predict whether breast lumps are benign or malignant based on a set of features extracted from fine-needle aspirate images.
+
+%The most discriminative features tend to get large weight, however some quite discriminative features may get low weight if strongly correlated with a more discriminative feature
+
+%Non-linear means that features with average discrimination power may get low weight as they are superseded by the most discriminative features.
+But some features with low discrimination power may also get large weight, which does not mean that they are very influential in practice (see above).
+This is illustrated in Figure~\ref{fig:disc_weight_plot} with the UCI breast cancer dataset.
+
+%{\color{red} Equality constraints (negative weights) are less interpretable.}
+
+%{\color{red} Offsets.}
+
+%{\color{red} Data weights.}
+
+%Need to weight the data to adjust to prior as a general measure when minimizing cross-entropy, which should yield a conditional model that approximates the ``true'' conditional distribution, $q(x|y)\approx p(x|y)$.
 So the model will absorb the ``true'' marginal of $x$, which should thus be replaced by the prior if we want the model to mimic ideal Bayesian inference.
@@
 
-{\color{red} To offset or not to offset?}
-
-We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$.
+%We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$.
 This ``deforms'' the prior to improve the training score.
@@
 
-When including offsets, using homoscedastic generative variance estimation and equality constraints, composite likelihood is equivalent to binomial logistic regression.
+%When including offsets, using homoscedastic generative variance estimation and equality constraints, composite likelihood is equivalent to binomial logistic regression.
 However, equality constraints are not easily interpretable.
@@
 
-%\section{Extensions}
-
-\section{Super composite likelihood}
-\label{sec:super}
-
-When chosen for computational simplicity, features may not only convey limited information at individual level, their informativeness may also be very much class-dependent.
-Consider, for instance, diagnosing a disease from a routine medical checkup.
-Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
-This motivates a more general pooling than~(\ref{eq:log_pool}) setting where features can be weighted depending on the target variable:
-$$
-p_\lambda(x|y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)}
-$$
-
-$p_0$ is an arbitrary reference (implicit if weights are class-independent as it is then effectless).
-Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
-healthy population).
-In this case, the super composite likelihood is equivalent to the standard one in the two-class case.
-
-This is the super composite likelihood idea in a nutshell.
-
-Again, we can train this model and see that it is the same as solving a maxent problem.
-Difference with above is that the constraints are not averaged over $x$ (basis functions are of the form $\chi(x-a)\rho_i(x,y)$) and that they involve arbitrary log-likelihood ratios vs some reference distribution.
-Make sense to have a reference as it gives a customizable baseline for detection.
-
-If some classes are merged or split, then the other classes keep the same weights (assuming $h(y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
-So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
-
-Super composite likelihood is essentially about combining binary versions of composite likelihood. It is analogous to the ``one-versus-rest'' strategy in multiple logistic regression.
-
-%Link with PDF projection \cite{Baggenstoss-03}.
-
-Advantages: richer model (so better predictions), probabilty ratios stable to class addition.
-Drawbacks: require a reference, no more consistency.
-Still interpretable, that's why it is worthwile mentioning.
-
-%I was thinking about the unit sum weight version...
-The diagnosis example could be one in which clues are independent, in which case $\lambda_i(x)\equiv 1$ would be optimal, making any super composite trick irrelevant.
-Also see that with hypothesis-dependent weights, we may lose the estimation consistency.
-At the end of the day, the interest of the super composite idea is questionable.
+%Why not using the feature-based posteriors or Bayes factors rather than the likelihoods?
+It's a possibility (perhaps a connection here with Gr\"unwald's luckiness).
+A justification for not doing it is that the coordinator should discard the priors used by the agents.
+The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.
+
+%Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
+We expect to have $\lambda=1$ but it's not necessarily the case.
+It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
+This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
+So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
+
+%Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
+The answer is no.
+The weights generally depend on the prior.
+Hence the maxent composite likelihood is prior-dependent.
+This is an important conceptual difference with the classical notion of likelihood.
 
@@
 {\color{red} Composite Bayes is invariant by affine transformation of the data (unlike multinomial LR, which is problematic when damping is applied).}
+
+{\color{red} Is damping important? (it guarantees training cost is stricly convex).}
+
+{\color{red} Can we use the same dataset to learn both the feature generative distributions and the composite weights? In principle, nothing speaks about that.}
 
@@
 
+Future work: can we collapse generative and discriminative training into a single step? Maybe using the concept of {\em minimally informative likelihood} \cite{Yuan-99b}.
+
 \appendix
@@
 \section{Conditional composite likelihood}
-\label{sec:conditional}
+\label{appendix:conditional}
 
@@
 
-\section{Maxent}
-
-\subsection{Training}
-\label{sec:maxent}
-
-{\color{red}This stuff should move to the appendix.
-It is not critical to understand the Maxent connection of trained composite likelihood, althought it gives some insight.
-Essentially, agents report compressing rates using feature-based models and we either trust them (equality constraints) or assume that they are underestimated (inequality constraints).
-Why underestimated?
-Well, it's just more general.
-Overestimated leads nowhere, we have to assume something.}
-
-We can derive composite likelihood from the conditional maximum entropy principle \cite{BergerA-96}.
-It comes with a method of tuning weights, hence a particular composite likelihood function that we call maxent composite likelihood (MCL).
-
-A cheap MaxEnt argument was already given by \cite{Wang-14}  (standard, non-conditional MaxEnt at fixed $y$, hence just an existence result but no feasible weight tuning).
-
-What we do here is maximum {\em conditional} entropy.
-It's slightly more complex than a simple $I$-projection.
-The search space is a set of predictive distributions compatible with mean log-likelihood contraints.
-If I know the generative distribution, I know the expected log-likelihood.
-
-We could work with inequality constraints to force $\lambda\geq 0$:
-$$
-E[\log p(z_i|x)] \geq \bar{\ell}_i \equiv \int h(x)p(z_i|x) \log p(z_i|x) dx dy
-$$
-where $h(x)$ is the marginal distribution of labels.
-
-What does the inequality constraint mean?
-It means that a model is admissible if it jointly allows for all agents to achieve the same compression rate {\em or better} than achieved individually.
-Somewhow, we forget that there should exist a true distribution that is consistent with the agent constraints both jointly and individually.
-We forget that not because it is not true but because we deem it irrelevant.
-Connection to MDL?
-We are reminded of these words by Rissanen:
-\begin{quote}
-No statistical model is ``true" or ``false", ``right" or ``wrong"; the models just have varying performance, which can be assessed.  
-\end{quote}
-
-Importantly, no agent will be proven wrong since individual performance is at least as good as announced.
-But if it is strictly better, it means that the agent is simply discarded (getting a zero weight rather than a difficult-to-intepret negative weight).
-
-We could also consider inequality constraints as a relaxation: explicitly making it easier for Nature to screw us up, but for a good cause.
-Not to make optimization easier, to make it easier to intepret.
-Why?
-Because we know that a posteriori from the fact that we get $\lambda_i\geq 0$.
-But a priori?
-
-%Okay, but then it's not ``nature'' playing, is it?
-The game is between a ``generator'' G and a ``predictor'' P.
-G tries to make it as difficult as possible to predict outcomes using an admissible model.
-We could argue that true models are at the boundaries of the admissible space, but G is not restricted to true models as long as it maintains the agents performance level.
-
-%In fact, P is the agent coordinator and G could be seen as some sort of test controller. Inequality constraints add some further robustness to the model.
-
-Why not using the feature-based posteriors or Bayes factors rather than the likelihoods?
-It's a possibility (perhaps a connection here with Gr\"unwald's luckiness).
-A justification for not doing it is that the coordinator should discard the priors used by the agents.
-The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.
-
-The derivation assumed the marginal distribution of the data $h(y)$ to be known. In practice, it is estimated by the empirical distribution just like the moments are estimated.
-
-Game theoretic interpretation \cite{Grunwald-04}.
-
-The dual function boils down to the familiar notions of cross-entropy, log-likelihood, KL divergence...
-$$
-\psi(\lambda)
-= \int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
-$$
-
-If we introduce $p_\lambda(x,y)=p_\lambda(x|y)h(y)$, it is even clearer that this is a most classical distribution fitting problem:
-$$
-\psi(\lambda)
-= \int h(x,y) \log \frac{p_\lambda(x,y)}{\pi(x)h(y)} dx dy
-$$
-(we can obviously drop the reference distribution here).
-So the Maxent justifies the composite likelihood concept as a somewhat optimal parametric form but the remainder of the story is most classical.
-
-Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
-We expect to have $\lambda=1$ but it's not necessarily the case.
-It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
-This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
-So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
-
-Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
-The answer is no.
-The weights generally depend on the prior.
-Hence the maxent composite likelihood is prior-dependent.
-This is an important conceptual difference with the classical notion of likelihood.
-
-\subsection{Data weighting}
-
-The distribution $h(x)$ is the marginal of the labels, which is natural to approximate via their empirical distribution.
-We may then have a discrepancy between $h(x)$ and $\pi(x)$, which makes little sense if it is believed that $h(x)$ is a ``good encoder''.
-
-To correct for this, we may adjust the empirical distribution of labels and data so as to match the target prior, $\pi(x)$, leading to weight the data points as a preprocessing step:
-$$
-h(y) = \sum_i w_i \delta(y-y_i),
-\qquad
-w_i = \frac{\pi(x_i)}{h(x_i)}
-$$
-
-Are the two methods equivalent?
-No.
-The feature-based parameter estimates are obviously the same in both cases, but the search space for the predictive distribution is different, hence we can't get the same answer.
-Which one of the two methods is best?
-The one that allows you to choose your own prior makes more sense to me because the distribution of classes in your training dataset may be completely different from the one that you will deal with in practice.
-For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
-
 \section{Optimal predictive model}
+\label{appendix:maxent}
+
+{\color{red} Need cleaning up. Note offsets imply we minimize mutual information. But let's not discuss MIL here.}
 
@@
 This is the {\em maximum conditional entropy} method \cite{BergerA-96} which we review hereafter.
-We will also treat the case where the marginal constraint is not assumed, which corresponds to the {\em minimally informative likelihood} method \cite{Yuan-99b,Yuan-99}.
+
+%We will also treat the case where the marginal constraint is not assumed, which corresponds to the {\em minimally informative likelihood} method \cite{Yuan-99b,Yuan-99}.
 
@@
 
+\subsection{Data weighting}
+
+The distribution $h(x)$ is the marginal of the labels, which is natural to approximate via their empirical distribution.
+We may then have a discrepancy between $h(x)$ and $\pi(x)$, which makes little sense if it is believed that $h(x)$ is a ``good encoder''.
+
+To correct for this, we may adjust the empirical distribution of labels and data so as to match the target prior, $\pi(x)$, leading to weight the data points as a preprocessing step:
+$$
+h(y) = \sum_i w_i \delta(y-y_i),
+\qquad
+w_i = \frac{\pi(x_i)}{h(x_i)}
+$$
+
+Are the two methods equivalent?
+No.
+The feature-based parameter estimates are obviously the same in both cases, but the search space for the predictive distribution is different, hence we can't get the same answer.
+Which one of the two methods is best?
+The one that allows you to choose your own prior makes more sense to me because the distribution of classes in your training dataset may be completely different from the one that you will deal with in practice.
+For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
+
+\section{Super composite likelihood}
+\label{appendix:super}
+
+When chosen for computational simplicity, features may not only convey limited information at individual level, their informativeness may also be very much class-dependent.
+Consider, for instance, diagnosing a disease from a routine medical checkup.
+Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
+This motivates a more general pooling than~(\ref{eq:log_pool}) setting where features can be weighted depending on the target variable:
+$$
+p_\lambda(x|y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)}
+$$
+
+$p_0$ is an arbitrary reference (implicit if weights are class-independent as it is then effectless).
+Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
+healthy population).
+In this case, the super composite likelihood is equivalent to the standard one in the two-class case.
+
+This is the super composite likelihood idea in a nutshell.
+
+Again, we can train this model and see that it is the same as solving a maxent problem.
+Difference with above is that the constraints are not averaged over $x$ (basis functions are of the form $\chi(x-a)\rho_i(x,y)$) and that they involve arbitrary log-likelihood ratios vs some reference distribution.
+Make sense to have a reference as it gives a customizable baseline for detection.
+
+If some classes are merged or split, then the other classes keep the same weights (assuming $h(y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
+So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
+
+Super composite likelihood is essentially about combining binary versions of composite likelihood. It is analogous to the ``one-versus-rest'' strategy in multiple logistic regression.
+
+%Link with PDF projection \cite{Baggenstoss-03}.
+
+Advantages: richer model (so better predictions), probabilty ratios stable to class addition.
+Drawbacks: require a reference, no more consistency.
+Still interpretable, that's why it is worthwile mentioning.
+
+%I was thinking about the unit sum weight version...
+The diagnosis example could be one in which clues are independent, in which case $\lambda_i(x)\equiv 1$ would be optimal, making any super composite trick irrelevant.
+Also see that with hypothesis-dependent weights, we may lose the estimation consistency.
+At the end of the day, the interest of the super composite idea is questionable.
+
 \section{Data reduction inequality}
-\label{sec:reduction_inequality}
+\label{appendix:reduction_inequality}
 
@@
 \section{Basic asymptotic properties of composite likelihood}
-\label{sec:asymptotic}
+\label{appendix:asymptotic}
 

### 2019-02-08 papers/composite_inference/draft.tex
Init Madrugada

@@
 
+\def\x{{\mathbf{x}}}
+\def\y{{\mathbf{y}}}
+\def\y{{\mathbf{z}}}
+\newcommand{\blambda}{{\boldsymbol{\lambda}}}
+\newcommand{\bell}{{\boldsymbol{\ell}}}
 \newcommand{\E}{\mathbb{E}}
-\newcommand{\lda}{{\boldsymbol{\lambda}}}
 %\newcommand{\matphi}{\boldsymbol{\Phi}} 
-%\def\x{{\mathbf{x}}}
-%\def\u{{\mathbf{u}}}
 %\def\p{{\bar{\mathbf{p}}}}
@@
 
-% p(x,y|theta) = p(x|y,theta)p(y|theta)
-
 \begin{figure}[!ht]
@@
 
-Let $Y$ an observable multivariate random variable with sampling distribution $p(y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is assumed to be a finite set for simplicity.
-Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $x$:
+Let $\mathbf{Y}$ an observable multivariate random variable with sampling distribution $p(\y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is assumed to be a finite set for simplicity.
+Given an experimental outcome $\y$, the likelihood is the sampling distribution evaluated at $\y$, seen as a function of $x$:
 $$
-L(x) = p(y|x)
+L(x) = p(\y|x)
 .
@@
 This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated.
-A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
+A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(\y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
 $$
@@
 While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction.
-To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.
+To make the trick statistically more efficient, we may extract several features, $z_i=f_i(\y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.
 
@@
 \label{eq:log_pool}
-p_\lda(x|y) \propto \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
+p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
 \end{equation} 
-where $\pi(x)$ is a prior distribution, and $\lda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lda(x_1|y)\leq p_\lda(x_2|y)$.
+where $\pi(x)$ is a prior distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
 Also note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
@@
 %This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|\y)\leq p_\lambda(x_2|\y)$.
 Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
@@
 $$
-p_\lda(x|y)\propto \pi(x) L^c_\lda(x),
+p_\blambda(x|\y)\propto \pi(x) L^c_\blambda(x),
 $$
@@
 \label{eq:comp_lik}
-L^c_\lda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
+L^c_\blambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
 \end{equation} 
@@
 
-CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lda\equiv 1$.
+CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
 While the above derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method.
@@
 $$
-\log \frac{p_\lda(x_1|y)}{p_\lda(x_0|y)}
+\log \frac{p_\blambda(x_1|\y)}{p_\blambda(x_0|\y)}
 = 
@@
 Contrary to statistical hypothesis testing, likelihood ratios are not converted into $p$-values but are directly combined into a predictive probability ratio.
-Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\lda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately.
+Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\blambda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately.
 Keeping the analogy with hypothesis testing, composite weights may be thought of as the counterpart of multiple testing correction \cite{Benjamini-10}.
@@
 
-%So far, we have assumed that the composite weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) are known.
+%So far, we have assumed that the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) are known.
 
-Let us now turn to the problem of tuning the composite weights~$\lda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration.
+Let us now turn to the problem of tuning the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration.
 It is assumed at this stage that the feature generative models are known, which may require a pre-training stage that should not cause technical difficulties as it may use standard parameter estimation techniques.
@@
 
-Given a labeled training dataset ${\cal D}=\{(x_k, y_k), k=1,\ldots,N\}$, the most direct approach to tune the composite weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'') under the composite predictive distribution:
+Given a labeled training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct approach to tune the composite weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'') under the composite predictive distribution:
 \begin{equation}
 \label{eq:train_likelihood}
-\max_{\lda\succeq 0} U(\lda)
-\equiv\sum_{k=1}^N \log p_\lda(x_k|y_k).
+\max_{\blambda\succeq 0} U(\blambda)
+\equiv\sum_{k=1}^N \log p_\blambda(x_k|\y_k).
 \end{equation}
 
-From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\lda(x|y)$ spanned by the weights~$\lda$ is the exponential family with natural parameter~$\lda$ and basis functions given by the feature-based log-likelihoods:
+From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\blambda(x|\y)$ spanned by the weights~$\blambda$ is the exponential family with natural parameter~$\blambda$ and basis functions given by the feature-based log-likelihoods:
 $$
-p_\lda(x|y) = \pi(x) \exp[\lda^\top \mathbf{L}(x,y) - A(\lda,y)],
+p_\blambda(x|\y) = \pi(x) \exp[\blambda^\top \bell(x,\y) - A(\blambda,y)],
 $$
@@
 $$
-L_i(x,y) \equiv \log p(z_i|x),
+\ell_i(x,\y) \equiv \log p(z_i|x),
 \qquad
-A(\lda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\lda^\top \mathbf{F}(x,y)}.
+A(\blambda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
 $$
 
-It follows that the utility function $U(\lda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:maxent} for details.
-It also implies that maximizing $U(\lda)$ is dual to a minimum KL~divergence problem \cite{Csiszar-84}:
+It follows that the utility function $U(\blambda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:maxent} for details.
+It also implies that maximizing $U(\blambda)$ is dual to a minimum KL~divergence problem \cite{Csiszar-84}:
 $$
-\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,y) \log \frac{p(x,y)}{\pi(x)h(y)},
+\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{\pi(x)h(\y)},
 $$
-subject to the constraints $p(y)=h(y)$ and $\E_p[\mathbf{L}] \succeq \E_h[\mathbf{L}]$, where 
-${\cal Y}=\{y_1,\ldots,y_N\}$ is the set of training data values, ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$, $h(x,y)$ is the joint empirical distribution of targets and data, and $h(y)$ is the corresponding marginal distribution.
+subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where 
+${\cal Y}=\{\y_1,\ldots,\y_N\}$ is the set of training data values, ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$, $h(x,\y)$ is the joint empirical distribution of targets and data, and $h(\y)$ is the corresponding marginal distribution.
 
-Therefore, the optimal composite model may be interpreted as the $I$-projection of the independence distribution $\pi(x)h(y)$ onto a set of admissible joint distributions defined by two types of contraints.
-The marginal constraint $p(y)=h(y)$, on the one hand,  ensures that only the conditional distribution $p(x|y)$ is ``updated'', making the $I$-projection a special case of maximum conditional entropy \cite{BergerA-96}.
-The mean log-likelihood value constraints $\E_p[\mathbf{L}] \succeq \E_h[\mathbf{L}]$, on the other hand, encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
+Therefore, the optimal composite model may be interpreted as the $I$-projection of the independence distribution $\pi(x)h(\y)$ onto a set of admissible joint distributions defined by two types of contraints.
+The marginal constraint $p(\y)=h(\y)$, on the one hand,  ensures that only the conditional distribution $p(x|\y)$ is ``updated'', making the $I$-projection a special case of maximum conditional entropy \cite{BergerA-96}.
+The mean log-likelihood value constraints $\E_p[\bell] \succeq \E_h[\bell]$, on the other hand, encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
 This is weaker a constraint than assuming that the features are distributed according to the generative models $p(z_i|x)$: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.
@@
 As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (obtaining balanced classes if the prior is uniform).
-This does not affect the above maths provided that the data weights are reflected in the empirical distribution~$h(x,y)$.
+This does not affect the above maths provided that the data weights are reflected in the empirical distribution~$h(x,\y)$.
 
@@
 For instance, class-dependent offsets may be learned concurrently with the composite weights, which amounts to learning the ``prior'' in the training phase.
-It is also possible to replace the mean log-likelihood inequality constraints with equality constraints: the maximization of $U(\lda)$ is then unconstrained, therefore some trained composite weights may be negative.
+It is also possible to replace the mean log-likelihood inequality constraints with equality constraints: the maximization of $U(\blambda)$ is then unconstrained, therefore some trained composite weights may be negative.
 Finally, the mean log-likelihood constraints may be expressed for each class rather than being averaged wrt the prior: this leads to an extension that we call ``super composite likelihood'', in which weights become class-dependent (see Appendix~\ref{appendix:super}).
@@
 
-%We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(y)p_\lambda(x|y)dy = \pi(x)$.
+%We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(\y)p_\lambda(x|y)dy = \pi(x)$.
 This ``deforms'' the prior to improve the training score.
@@
 
-%Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$.
+%Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(\y|x)^\lambda$.
 We expect to have $\lambda=1$ but it's not necessarily the case.
-It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active).
+It is if $h(\y)=\int\pi(x)p(\y|x)dx$ because then $h(\y)p(x|y)=\pi(x)p(\y|x)$ so the constraint is verified (and active).
 This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
@@
 
-As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
+As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(\y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(\y)$, yielding the more general form:
 \begin{equation}
@@
 
-Let $\pi(x)$ some prior distribution on~$X$. Optimal prediction aims to select the joint distribution $p(x,y)$ that minimizes the KL divergence $D(p(x,y)\|\pi(x)p(y))$:
+Let $\pi(x)$ some prior distribution on~$X$. Optimal prediction aims to select the joint distribution $p(x,\y)$ that minimizes the KL divergence $D(p(x,\y)\|\pi(x)p(\y))$:
 $$
-D(p) = -\int p(x,y) \log \frac{p(x|y)}{\pi(x)} dx dy,
+D(p) = -\int p(x,\y) \log \frac{p(x|\y)}{\pi(x)} dx d\y,
 $$ subject to feature mean-value constraints of the form:
 $$
-\int p(x,y) f(x,y) dx dy = \mu.
+\int p(x,\y) f(x,\y) dx d\y = \mu.
 $$
@@
 
-One important special case of this problem arises when imposing the additional constraint that $p(y)=h(y)$ is known.
+One important special case of this problem arises when imposing the additional constraint that $p(\y)=h(\y)$ is known.
 This is the {\em maximum conditional entropy} method \cite{BergerA-96} which we review hereafter.
@@
 $$
-D(p) = -\int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
+D(p) = -\int h(\y)p(x|\y) \log \frac{p(x|\y)}{\pi(x)} dxdy
 $$
@@
 $$
-{\cal L}(p,\lambda) = D(p) + \lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - \mu)
+{\cal L}(p,\lambda) = D(p) + \lambda^\top (\int h(\y)p(x|\y)f(x,y)dxdy - \mu)
 $$
@@
 =
-- h(y) \log \frac{p(x|y)}{\pi(x)} + h(y) \lambda^\top f(x,y) + (\kappa(y)-1)h(y)
+- h(\y) \log \frac{p(x|\y)}{\pi(x)} + h(\y) \lambda^\top f(x,y) + (\kappa(\y)-1)h(\y)
 $$
@@
 \quad
-p_{\lambda}(x|y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
+p_{\lambda}(x|\y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
 $$
@@
 $$
-\psi(\lambda) = \lambda^\top \mu - \int h(y) \log z(\lambda, y) dy 
+\psi(\lambda) = \lambda^\top \mu - \int h(\y) \log z(\lambda, y) dy 
 $$
@@
 \nabla\psi
-= \mu - \int h(y) \frac{\nabla z}{z} dy
+= \mu - \int h(\y) \frac{\nabla z}{z} dy
 $$
@@
 \nabla\nabla^\top\psi
-= - \int h(y) \left(
+= - \int h(\y) \left(
 \frac{\nabla \nabla^\top z}{z} 
@@
 =
-\int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
-= D(h\|\pi(x)h(y)) - D(h\|p_\lambda)
+\int h(x,y) \log \frac{p_\lambda(x|\y)}{\pi(x)} dx dy
+= D(h\|\pi(x)h(\y)) - D(h\|p_\lambda)
 $$
@@
 
-Let $m_F(y)$ and $V_F(y)$ respectively denote the mean and variance of $f(x,y)$ conditional on~$y$. Denoting by $\langle \rangle$ the empirical mean wrt $y$:
+Let $m_F(\y)$ and $V_F(\y)$ respectively denote the mean and variance of $f(x,y)$ conditional on~$y$. Denoting by $\langle \rangle$ the empirical mean wrt $y$:
 $$
@@
 
-In general, we have more features than classes for good prediction performance so each matrix $V_F(y)$ is singular and tends to be smaller (in terms of spectral radius) as the number of features increases.
+In general, we have more features than classes for good prediction performance so each matrix $V_F(\y)$ is singular and tends to be smaller (in terms of spectral radius) as the number of features increases.
 The Hessian is a sample average of these matrices and will thus follow the same trend.
@@
 $$
-p_\lambda(x_0|y) = \frac{\pi(x_0)}{z(\lambda,y)} e^{\lambda^\top f(x_0,y)}
+p_\lambda(x_0|\y) = \frac{\pi(x_0)}{z(\lambda,y)} e^{\lambda^\top f(x_0,y)}
 $$
@@
 $$
-\log z(\lambda,y) = \lambda^\top f(x_0,y) - \log \frac{p_\lambda(x_0|y)}{\pi(x_0)}
+\log z(\lambda,y) = \lambda^\top f(x_0,y) - \log \frac{p_\lambda(x_0|\y)}{\pi(x_0)}
 $$
@@
 $$
-\psi(\lambda) = \lambda^\top (\mu - \mu_0) + \int h(y) \log \frac{p_\lambda(x_0|y)}{\pi(x_0)} dy
+\psi(\lambda) = \lambda^\top (\mu - \mu_0) + \int h(\y) \log \frac{p_\lambda(x_0|\y)}{\pi(x_0)} dy
 $$
@@
 $$
-\mu_0 = \int h(y) f(x_0, y) dy
+\mu_0 = \int h(\y) f(x_0, y) dy
 $$
@@
 Cross-entropy appears here as a regularized version of utility.
-The regularization term $E[\log p_\lambda(x_0|y)$ replaces the hard constraint that $\lambda$ lie in the simplex: it forces the reference class to be probable for any input, so it will tend to discard predictions that are peaked on a single class if it's not $x_0$.
+The regularization term $E[\log p_\lambda(x_0|\y)$ replaces the hard constraint that $\lambda$ lie in the simplex: it forces the reference class to be probable for any input, so it will tend to discard predictions that are peaked on a single class if it's not $x_0$.
 This term acts as a callback force.
@@
 
-In this case, the constraint $p(y)=h(y)$ no longer applies. Consider:
+In this case, the constraint $p(\y)=h(\y)$ no longer applies. Consider:
 $$
-D(p,m) = \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dx dy
+D(p,m) = \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(\y)} dx dy
 $$
@@
 =
-- \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(y)} dxdy
+- \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(\y)} dxdy
 +
@@
 = 
-- \log \frac{p(x,y)}{\pi(x)m(y)} - 1
+- \log \frac{p(x,y)}{\pi(x)m(\y)} - 1
 + \lambda^\top f(x,y)
@@
 $$
-p_{\lambda,m}(x,y) = \frac{1}{z(\lambda,m)} \pi(x) m(y) e^{\lambda^\top f(x,y)} 
+p_{\lambda,m}(x,y) = \frac{1}{z(\lambda,m)} \pi(x) m(\y) e^{\lambda^\top f(x,y)} 
 $$
@@
 $$
-z(\lambda,m) = \int \pi(x) m(y) e^{\lambda^\top f(x,y)} dx dy
+z(\lambda,m) = \int \pi(x) m(\y) e^{\lambda^\top f(x,y)} dx dy
 $$
 
-Note that the induced posterior distribution $p_{\lambda,m}(x|y)\propto \pi(x)p_{\lambda,m}(y|x)$ has the same form as in the conditional maxent case (if using the very same moment constraints).
+Note that the induced posterior distribution $p_{\lambda,m}(x|\y)\propto \pi(x)p_{\lambda,m}(\y|x)$ has the same form as in the conditional maxent case (if using the very same moment constraints).
 
@@
 $$
-\nabla z = \int \pi(x) m(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
+\nabla z = \int \pi(x) m(\y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
 $$
@@
 $$
-\nabla\nabla^\top z = \int \pi(x)m(y) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dxdy
+\nabla\nabla^\top z = \int \pi(x)m(\y) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dxdy
 $$
@@
 =
-\int h(x,y) \log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(y)} dxdy
+\int h(x,y) \log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(\y)} dxdy
 = D(h\|\pi_xm_y) - D(h\|p_{\lambda,m})
@@
 $$
-m(y) = \int p_{\lambda,m}(x,y) dx
+m(\y) = \int p_{\lambda,m}(x,y) dx
 $$
 
-The MIL seems intractable if we don't impose constraints on $m(y)$.
-It becomes feasible by restricting $m(y)$ to have a support limited to the training data, similarly to the conditional maxent implementation:
+The MIL seems intractable if we don't impose constraints on $m(\y)$.
+It becomes feasible by restricting $m(\y)$ to have a support limited to the training data, similarly to the conditional maxent implementation:
 $$
-m(y) = \sum_k w_k \delta(y-y_k),
+m(\y) = \sum_k w_k \delta(\y-\y_k),
 $$
@@
 $$
-h(y) = \sum_i w_i \delta(y-y_i),
+h(\y) = \sum_i w_i \delta(\y-\y_i),
 \qquad
@@
 $$
-p_\lambda(x|y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)}
+p_\lambda(x|\y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)}
 $$
@@
 
-If some classes are merged or split, then the other classes keep the same weights (assuming $h(y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
+If some classes are merged or split, then the other classes keep the same weights (assuming $h(\y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
 So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
@@
 $$
-\alpha = \log \frac{p_\lambda(x|y)}{p_\lambda(x_0|y)}
+\alpha = \log \frac{p_\lambda(x|\y)}{p_\lambda(x_0|\y)}
 $$

### 2019-02-12 papers/composite_inference/draft.tex
Changes Friday night 9/2/19

@@
 \def\y{{\mathbf{y}}}
-\def\y{{\mathbf{z}}}
 \newcommand{\blambda}{{\boldsymbol{\lambda}}}
@@
 
-A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$y$.
+A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$\y$.
 Given two putative target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ is computed by simply adding up single-feature contributions:
@@
 \section{Composite weight tuning}
-\label{sec:weight_tuning}
+\label{sec:tuning}
 
@@
 Let us now turn to the problem of tuning the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration.
-It is assumed at this stage that the feature generative models are known, which may require a pre-training stage that should not cause technical difficulties as it may use standard parameter estimation techniques.
+It is assumed at this stage that the feature generative models are known, which may require a pre-training stage using standard parameter estimation techniques.
 
-\subsection{Agnostic calibration}
+\subsection{Previous work}
 
 In the absence of knowledge, a default rule is to choose constant weights.
-The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
+The CL function is then a scaled version of na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
 To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic MCLE variance matrix \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
 
-Another simpler recommendation for log-linear pooling weights is to sum up to one.
+Another justifiable recommendation for log-linear pooling weights is to sum up to one.
 This may be motivated by the fact that the only pooling operator that preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}.
@@
 
-Unit sum weights, however, implicitly assume strong redundancy between features and thus tend to produce over-conservative predictions when features are weakly correlated, as is generally the case in practice since features are supposed to be chosen to be complementary.
+Unit sum weights, however, implicitly assume strong redundancy between features and thus tend to produce over-conservative predictions when features are weakly correlated, as is generally the case in practice since features are supposed to be chosen for complementarity.
 An ideal method to tune weights is one that can capture and adjust for the pattern of statistical dependences between features.
-This calls for a training approach, which, as we will see, is unlikely to pick constant or unit sum weights.
+This calls for a training approach, which, as we will see, is very unlikely to pick constant or unit sum weights.
 
-\subsection{Training}
+\subsection{Maximum likelihood training}
+\label{sec:training}
 
@@
 \label{eq:train_likelihood}
-\max_{\blambda\succeq 0} U(\blambda)
-\equiv\sum_{k=1}^N \log p_\blambda(x_k|\y_k).
+\max_{\blambda\succeq 0} U(\blambda),
+\qquad \text{with} \quad
+U(\blambda) \equiv\sum_{k=1}^N \log p_\blambda(x_k|\y_k).
 \end{equation}
@@
 $$
-p_\blambda(x|\y) = \pi(x) \exp[\blambda^\top \bell(x,\y) - A(\blambda,y)],
+p_\blambda(x|\y) = \pi(x) \exp[\blambda^\top \bell(x,\y) - a(\blambda,y)],
 $$
@@
 \qquad
-A(\blambda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
+a(\blambda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
 $$
 
-It follows that the utility function $U(\blambda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:maxent} for details.
-It also implies that maximizing $U(\blambda)$ is dual to a minimum KL~divergence problem \cite{Csiszar-84}:
-$$
-\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{\pi(x)h(\y)},
-$$
-subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where 
-${\cal Y}=\{\y_1,\ldots,\y_N\}$ is the set of training data values, ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$, $h(x,\y)$ is the joint empirical distribution of targets and data, and $h(\y)$ is the corresponding marginal distribution.
+It follows that the utility function $U(\blambda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:training} for details.
 
-Therefore, the optimal composite model may be interpreted as the $I$-projection of the independence distribution $\pi(x)h(\y)$ onto a set of admissible joint distributions defined by two types of contraints.
-The marginal constraint $p(\y)=h(\y)$, on the one hand,  ensures that only the conditional distribution $p(x|\y)$ is ``updated'', making the $I$-projection a special case of maximum conditional entropy \cite{BergerA-96}.
-The mean log-likelihood value constraints $\E_p[\bell] \succeq \E_h[\bell]$, on the other hand, encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
-This is weaker a constraint than assuming that the features are distributed according to the generative models $p(z_i|x)$: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.
-
-A particularly insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is under broad conditions a minimax strategy for prediction according to the log loss.
-Hence, any $I$-projection is in some sense a best possible predictive model under particular constraints.
-Being more specifically a maximum conditional entropy method, trained composite inference belongs to the same family as multinomial logistic regression, only differing by the choice of constraints, which, in the latter case, are not related to feature generative models.
-
-In the $I$-projection framework, the composite weights are the Lagrange multipliers associated with the mean log-likelihood inequality constraints.
-Weights corresponding to inactive constraints vanish, hence the training has a natural tendency to produce sparse weights.
-As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that the training differs significantly from univariate feature selection as it takes feature redundancy into account.
+Because some positive weight constraints may turn out inactive, the training has a natural tendency to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, clearly showing that maximum likelihood training differs from univariate feature selection as it takes feature redundancy into account.
 
@@
 
-As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (obtaining balanced classes if the prior is uniform).
-This does not affect the above maths provided that the data weights are reflected in the empirical distribution~$h(x,\y)$.
+As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (enforcing balanced classes if the prior is uniform).
+Also, some regularization may be useful for stability: we shall in practice maximize $U(\lambda)-\alpha \|\blambda\|^2$, where $\alpha>0$ is a small damping factor.
 
-Other variants of the training scheme that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family.
-For instance, class-dependent offsets may be learned concurrently with the composite weights, which amounts to learning the ``prior'' in the training phase.
-It is also possible to replace the mean log-likelihood inequality constraints with equality constraints: the maximization of $U(\blambda)$ is then unconstrained, therefore some trained composite weights may be negative.
-Finally, the mean log-likelihood constraints may be expressed for each class rather than being averaged wrt the prior: this leads to an extension that we call ``super composite likelihood'', in which weights become class-dependent (see Appendix~\ref{appendix:super}).
+Other training variants that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family.
+For instance, class-dependent offsets may be added, hence learning the ``prior'' together with the composite weights.
+It is also possible to perform unconstrained optimization to achieve higher training likelihood, meaning that some composite weights are then negative.
+
+\subsection{Maximum entropy interpretation}
+
+Exponential family properties also imply that maximum likelihood training~(\ref{eq:train_likelihood}) is dual to a maximum entropy problem or, more precisely, an $I$-projection \cite{Csiszar-84}:
+$$
+\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{\pi(x)h(\y)},
+$$
+subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where 
+${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$
+with ${\cal Y}=\{\y_1,\ldots,\y_N\}$ the set of training data values, $h(x,\y)$ is the joint empirical distribution of targets and data, and $h(\y)$ is the corresponding marginal data distribution.
+Note that the constraints implicitly approximate the data marginal and the mean log-likelihood values by empirical estimates, suggesting the need for regularization on small training sets.
+
+The marginal constraint $p(\y)=h(\y)$ makes the $I$-projection a special case of maximum entropy classifier \cite{BergerA-96}, ensuring that only the conditional distribution $p(x|\y)$ is optimized.
+The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
+This is weaker a constraint than assuming that the feature generative models $p(z_i|x)$ are true: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.
+
+%In the $I$-projection framework, the composite weights are the Lagrange multipliers associated with the mean log-likelihood inequality constraints.
+
+A particularly insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is under broad conditions a minimax strategy for prediction according to the log loss.
+Hence, any $I$-projection is in some sense a best possible predictive model under particular constraints (in the limit of validity of empirical approximations).
+Our training of composite weights belongs to the same family as multinomial logistic regression, another maximum entropy classifier, only differing by the choice of constraints, which, in the latter case, are not related to generative models.
+
+As an alternative to the strategy considered here, the mean log-likelihood constraints may be expressed for each class rather than being averaged w.r.t.~the prior, leading to a more general predictive model that we call ``super composite likelihood'', in which the weights become class-dependent (see Appendix~\ref{appendix:super}).
+This variant has the potential for higher prediction performance, but poses some conceptual difficulties, in particular that of having to define an arbitrary reference class.
 
@@
 
+{\color{red} It is easy to build a good classifier.
+You just need to pick a parametric family that is big enough to produce a good fit yet small enough to avoid overfitting.
+Less easy is to interpret what the classifier does.
+Deep learning is cool but it is relevant to AI only if the representations it learns are generic enough to be used in other problems, i.e.
+it should ultimately be validated via transfer learning.}
+
 {\color{red} Why prediction performance is still important? Because it validates (or not) interpretation. Interpretation is the goal but prediction performance is a crucial intermediate step.}
@@
 
-{\color{red} It is easy to build a good classifier.
-You just need to pick a parametric family that is big enough to produce a good fit yet small enough to avoid overfitting.
-Less easy is to interpret what the classifier does.
-Deep learning is cool but it is relevant to AI only if the representations it learns are generic enough to be used in other problems, i.e.
-it should ultimately be validated via transfer learning.}
+{\color{red} Alternative to logistic regression in deep learning? Would require a game-theoretic learning approach... future work.}
 
@@
 
-{\color{red} Can we use the same dataset to learn both the feature generative distributions and the composite weights? In principle, nothing speaks about that.}
+{\color{red} Can we use the same dataset to learn both the feature generative distributions and the composite weights? In principle, nothing speaks against this.}
 
@@
 Number of independent discriminative parameters in logistic regression: $(K-1)(F+1)$.
-In standard composite inference: $K-1+F$.
-Difference $=(K-2)F$.
+In standard composite inference: $F$ (or $K-1+F$ with offsets).
+Difference $=(K-2)(F+1)+1$ (or $=(K-2)F$).
 In super-composite inference: $(K-1)(F+1)$ (if reference is a class).
@@
 
+\section{Basic frequentist properties of composite likelihood}
+\label{appendix:frequentist}
+
+Assume that the data~$\y$ is distributed according to some target value~$x_\star$. For any extracted feature~$z_i$ and for any hypothetical target value~$x$,
+$$
+0 \leq
+E\left[
+\log \frac{p(z_i|x_\star)}{p(z_i|x)}
+\right]
+\leq
+E\left[
+\log \frac{p(\y|x_\star)}{p(\y|x)}
+\right],
+$$
+where the expectation is taken with respect to the true distribution $p(\y|x_\star)$.
+Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any composite likelihood function:
+\begin{equation}
+\label{eq:variation_bound}
+0 \leq
+E\left[ \log \frac{L_c(x_\star, \blambda)}{L_c(x,\blambda)} \right]
+\leq 
+\|\blambda\|_1 E\left[ \log \frac{L(x_\star)}{L(x)} \right]
+.
+\end{equation}
+
+This implies two asymptotic properties of standard composite likelihood: 
+\begin{itemize}
+\item {\em Consistency.} The expected composite log-likelihood is maximized by~$x_\star$.
+\item {\em Conservativeness.} For any~$x$, if the expected true log-likelihood of~$x$ is finite, then the corresponding expected composite log-likelihood ratio is finite too. 
+\end{itemize}
+
 \section{Conditional composite likelihood}
@@
 
-\section{Optimal predictive model}
-\label{appendix:maxent}
+\section{Composite likelihood training}
+\label{appendix:training}
 
-{\color{red} Need cleaning up. Note offsets imply we minimize mutual information. But let's not discuss MIL here.}
-
-Let $\pi(x)$ some prior distribution on~$X$. Optimal prediction aims to select the joint distribution $p(x,\y)$ that minimizes the KL divergence $D(p(x,\y)\|\pi(x)p(\y))$:
+Using the same notation as in Section~\ref{sec:training}, the likelihood function in~(\ref{eq:train_likelihood}) can be expanded as follows:
 $$
-D(p) = -\int p(x,\y) \log \frac{p(x|\y)}{\pi(x)} dx d\y,
-$$ subject to feature mean-value constraints of the form:
-$$
-\int p(x,\y) f(x,\y) dx d\y = \mu.
+U(\blambda) 
+= \sum_{k=1}^N \left[
+\log \pi(x_k) + \blambda^\top \bell(x_k, \y_k) - a(\blambda,\y_k)
+\right]
 $$
 
-Note that with some constraints of this type, we can enforce $p(x)=\pi(x)$. In this case, the best predictive model minimizes the mutual information between the data and the target.
+This reads: $U(\blambda) = \sum_k \log \pi(x_k) + N \psi(\blambda)$ with: 
+$$
+\psi(\blambda) \equiv \blambda^\top \bar{\bell} - \frac{1}{N} \sum_k a(\lambda,\y_k), 
+$$
+and:
+$$
+\bar{\bell} \equiv \frac{1}{N} \sum_k \bell(x_k,\y_k),
+$$
+therefore maximizing~$U(\blambda)$ is equivalent to maximizing~$\psi(\blambda)$.
 
-One important special case of this problem arises when imposing the additional constraint that $p(\y)=h(\y)$ is known.
-This is the {\em maximum conditional entropy} method \cite{BergerA-96} which we review hereafter.
-
-%We will also treat the case where the marginal constraint is not assumed, which corresponds to the {\em minimally informative likelihood} method \cite{Yuan-99b,Yuan-99}.
-
-%We could say that MIL enforces the prior constraint explicitly while Maxent does it implicitly by pre-weighting the data.
-So it's a soft constraint in the Maxent case (which ends up being loosely satisfied if the model fits the empirical distribution well enough), and it's a hard constraint in the MIL case.
-On the other hand, the MIL formulation does not exploit the  empirical marginal at all (it only exploits the moments).
-Which one do we prefer?
-
-\subsection{Maximum conditional entropy}
+Derivatives...
+$$
+\nabla\psi
+= \bar{\bell} - \frac{1}{N}\sum_k \nabla a(\blambda,\y_k)
+$$
 
 $$
-D(p) = -\int h(\y)p(x|\y) \log \frac{p(x|\y)}{\pi(x)} dxdy
+\nabla\nabla^\top\psi
+= - \frac{1}{N}\sum_k
+\nabla \nabla^\top a(\blambda, \y_k)
 $$
 
-Lagrangian:
+We have:
 $$
-{\cal L}(p,\lambda) = D(p) + \lambda^\top (\int h(\y)p(x|\y)f(x,y)dxdy - \mu)
+\nabla a = \E_{\blambda}(\bell|\y),
+\qquad
+\nabla \nabla^\top a = \text{Var}_{\blambda}(\bell|\y)
+$$
+proving that $a(\blambda,\y)$ is convex in $\blambda$ for any $\y$, and in turnes proves the concavity of $\psi$.
+
+To see this, let:
+$$
+z(\blambda,\y) 
+\equiv e^{a(\lambda,\y)}  
+= \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}
 $$
 
-Optimal $p$ at fixed $\lambda$:
+We have:
 $$
-\frac{\partial{\cal L}}{\partial p}
-=
-- h(\y) \log \frac{p(x|\y)}{\pi(x)} + h(\y) \lambda^\top f(x,y) + (\kappa(\y)-1)h(\y)
-$$
-$$
-\Rightarrow
-\quad
-p_{\lambda}(x|\y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
-$$
-with:
-$$
-z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
+\nabla a = \frac{\nabla z}{z}
 $$
 
-Dual:
 $$
-\psi(\lambda) = \lambda^\top \mu - \int h(\y) \log z(\lambda, y) dy 
+\nabla\nabla^\top a 
+= \left(
+\frac{\nabla \nabla^\top z}{z} 
+- \frac{\nabla z \nabla z^\top}{z^2}
+\right)
 $$
@@
 =
-\int \pi(x) f(x,y)  e^{\lambda^\top f(x,y)} dx
+\sum_{x\in{\cal X}} \pi(x) \bell(x,\y)  e^{\blambda^\top \bell(x,\y)}
 $$
@@
 =
-\int \pi(x) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dx
+\sum_{x\in{\cal X}} \pi(x) \bell(x,\y) \bell(x,\y)^\top e^{\blambda^\top \bell(x,\y)}
 $$
-
-$$
-\nabla\psi
-= \mu - \int h(\y) \frac{\nabla z}{z} dy
-$$
-
-$$
-\nabla\nabla^\top\psi
-= - \int h(\y) \left(
-\frac{\nabla \nabla^\top z}{z} 
-- \frac{\nabla z \nabla z^\top}{z^2}
-\right)
-dy
-$$
-
-Note the equivalent expression of the dual function as:
-$$
-\psi(\lambda)
-=
-\int h(x,y) \log \frac{p_\lambda(x|\y)}{\pi(x)} dx dy
-= D(h\|\pi(x)h(\y)) - D(h\|p_\lambda)
-$$
-
-\paragraph{Convexity}
-
-The convexity of any function of the form $A(\lambda)=\log z(\lambda)$ with $z(\lambda)=\int\pi(x)e^{\lambda^\top f(x)}dx$ can be proved using H\"older inequality:
-$$
-\int f_1 f_2 \leq 
-\left(\int f_1^{\frac{1}{\lambda}} \right)^\lambda  
-\left(\int f_2^{\frac{1}{1-\lambda}} \right)^{1-\lambda}  
-$$
-for $f\succeq 0$, $g\succeq 0$, $0\leq\lambda\leq 1$.
-Just do $f_1=\exp(\lambda\theta^\top f)$ and similarly for $f_2$ with $1-\lambda$.
-This proves that $A_y(\lambda) = \log z(\lambda,y)$ is convex for any~$y$  and the fact remains true by averaging $A_y(\lambda)$ over several $y$, therefore $-\psi(\lambda)$ is convex.
-
-There is another, more insightful way to prove that $A(\lambda)$ is convex. Consider the variable $F=f(X)$ with $X\sim p_\lambda(x)$. It is easy to verify the following facts:
-$$
-\nabla A = \frac{\nabla z}{z} = E(F),
-\qquad
-\nabla\nabla^\top A 
-= \frac{\nabla\nabla^\top z}{z} 
-- \frac{\nabla z \nabla z^\top}{z^2}
-= {\rm Var}(F),
-$$
-so the Hessian is a variance matrix and is therefore positive, which proves convexity.
-How about definite positiveness?
-To stand a chance, we need the number of features to be less than the number of labels otherwise the basis vectors are linearly dependent.
-
-Let $m_F(\y)$ and $V_F(\y)$ respectively denote the mean and variance of $f(x,y)$ conditional on~$y$. Denoting by $\langle \rangle$ the empirical mean wrt $y$:
-$$
-\nabla \psi = \mu - \langle m_F \rangle
-$$
-
-$$
-\nabla \nabla^\top \psi = - \langle V_F \rangle
-$$
-
-In general, we have more features than classes for good prediction performance so each matrix $V_F(\y)$ is singular and tends to be smaller (in terms of spectral radius) as the number of features increases.
-The Hessian is a sample average of these matrices and will thus follow the same trend.
-
-In practice, it looks like the problem is always nearly singular so we may need regularization!
-
-\paragraph{Damping}
-
-We can regularize the Maxent problem via a $L_2$ penalty, yielding:
-$$
-\psi_r(\lambda) = \psi(\lambda) - \alpha \|\lambda\|^2
-$$
-
-This implements a callback force towards the prior $\pi(x)$. Can be interpreted as some way of saying that our trust in the empirical information is not infinite.
-
-\paragraph{Utility}
-
-Let us consider a reference target value $x_0$. By definition,
-$$
-p_\lambda(x_0|\y) = \frac{\pi(x_0)}{z(\lambda,y)} e^{\lambda^\top f(x_0,y)}
-$$
-
-Therefore,
-$$
-\log z(\lambda,y) = \lambda^\top f(x_0,y) - \log \frac{p_\lambda(x_0|\y)}{\pi(x_0)}
-$$
-
-We can use this fact to rewrite the dual function:
-$$
-\psi(\lambda) = \lambda^\top (\mu - \mu_0) + \int h(\y) \log \frac{p_\lambda(x_0|\y)}{\pi(x_0)} dy
-$$
-with: 
-$$
-\mu_0 = \int h(\y) f(x_0, y) dy
-$$
-
-The first term $\lambda^\top (\mu-\mu_0)$ is essentially the same thing as the utility considered in my previous version of composite likelihood training (which was maximized on the simplex).
-Utility is label-dependent as it involves the moments.
-If the basis consists of feature-based log-likelihood functions, it measures the discrepancy vs the reference and pushes towards large coefficient values.
-Cross-entropy appears here as a regularized version of utility.
-The regularization term $E[\log p_\lambda(x_0|\y)$ replaces the hard constraint that $\lambda$ lie in the simplex: it forces the reference class to be probable for any input, so it will tend to discard predictions that are peaked on a single class if it's not $x_0$.
-This term acts as a callback force.
-
-\subsection{Minimally informative likelihood}
-
-In this case, the constraint $p(\y)=h(\y)$ no longer applies. Consider:
-$$
-D(p,m) = \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(\y)} dx dy
-$$
-and note that the prediction objective is: $D(p) = \min_m D(p,m)$. This is the Blahut-Arimoto trick, which enables us to consider an alternate optimization strategy wrt $p$ and $m$, respectively.
-
-At fixed $p$, this is a maximum (joint) relative entropy problem. Lagrangian...
-$$
-{\cal L}(p,m,\lambda)
-=
-- \int p(x,y) \log \frac{p(x,y)}{\pi(x)m(\y)} dxdy
-+
-\lambda^\top \left( 
-\int p(x,y) f(x,y) dydy - \mu 
-\right) \\
-$$
-
-The derivative is given by:
-$$
-\frac{\partial\cal L}{\partial p(x,y)}
-= 
-- \log \frac{p(x,y)}{\pi(x)m(\y)} - 1
-+ \lambda^\top f(x,y)
-$$
-hence the optimal model at fixed $m$ has the form:
-$$
-p_{\lambda,m}(x,y) = \frac{1}{z(\lambda,m)} \pi(x) m(\y) e^{\lambda^\top f(x,y)} 
-$$
-with:
-$$
-z(\lambda,m) = \int \pi(x) m(\y) e^{\lambda^\top f(x,y)} dx dy
-$$
-
-Note that the induced posterior distribution $p_{\lambda,m}(x|\y)\propto \pi(x)p_{\lambda,m}(\y|x)$ has the same form as in the conditional maxent case (if using the very same moment constraints).
-
-Dual function...
-$$
-\psi(\lambda,m) 
-= \lambda^\top \mu - \log z(\lambda, m)
-$$
-
-$$
-\nabla z = \int \pi(x) m(\y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
-$$
-
-$$
-\nabla\nabla^\top z = \int \pi(x)m(\y) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dxdy
-$$
-
-$$
-\nabla \psi = \mu - \frac{\nabla z}{z} 
-$$
-
-$$
-\nabla\nabla^\top \psi = 
-- \left(
-\frac{\nabla \nabla^\top z}{z} 
-- \frac{\nabla z \nabla z^\top}{z^2}
-\right) 
-$$
-
-Equivalently, for any distribution $h(x,y)$ that satisfies the moment constraint, we have:
-$$
-\psi(\lambda,m) 
-=
-\int h(x,y) \log \frac{p_{\lambda,m}(x,y)}{\pi(x)m(\y)} dxdy
-= D(h\|\pi_xm_y) - D(h\|p_{\lambda,m})
-$$
-showing that this is a generative learning problem!
-
-The BA algorithm alternates between estimating $\lambda$ at fixed $m$ via the maximization of $\psi(\lambda,m)$ and estimating $m$ at fixed $\lambda$ by marginalization:
-$$
-m(\y) = \int p_{\lambda,m}(x,y) dx
-$$
-
-The MIL seems intractable if we don't impose constraints on $m(\y)$.
-It becomes feasible by restricting $m(\y)$ to have a support limited to the training data, similarly to the conditional maxent implementation:
-$$
-m(\y) = \sum_k w_k \delta(\y-\y_k),
-$$
-hence $m$ is determined by a finite set of weights $w_k$. In practice, the optimal weights look very sparse.
-
-{\color{red} Big fat speculation: wouldn't MIL make more sense when passing generative mean-value constraints, i.e.
-empirical mean and variance per class?
-This could avoid the double learning stage!
-We don't do it with Maxent because we want feature-based generative models for interpretation purposes (if we did, it would be logistic regression).
-But using MIL, we get a full generative model for free...
-which is a game changer.
-It's no longer composite!
-The trouble is, this model is kinda degenerate (due to its ``empirical'' style, which is needed for computational reasons): the likelihood of any new~$y$ is almost surely zero.
-So not very good to do hypothesis testing.
-We have an interpolation problem, Houston.
-And that interpolation problem is the very reason why we trashed discriminative models from the start!
-So maybe not the way to go.}
-
-\subsection{Data weighting}
-
-The distribution $h(x)$ is the marginal of the labels, which is natural to approximate via their empirical distribution.
-We may then have a discrepancy between $h(x)$ and $\pi(x)$, which makes little sense if it is believed that $h(x)$ is a ``good encoder''.
-
-To correct for this, we may adjust the empirical distribution of labels and data so as to match the target prior, $\pi(x)$, leading to weight the data points as a preprocessing step:
-$$
-h(\y) = \sum_i w_i \delta(\y-\y_i),
-\qquad
-w_i = \frac{\pi(x_i)}{h(x_i)}
-$$
-
-Are the two methods equivalent?
-No.
-The feature-based parameter estimates are obviously the same in both cases, but the search space for the predictive distribution is different, hence we can't get the same answer.
-Which one of the two methods is best?
-The one that allows you to choose your own prior makes more sense to me because the distribution of classes in your training dataset may be completely different from the one that you will deal with in practice.
-For instance, it be weird to say that someone with all the symptoms of a very rare disease does not have it because it is very rare.
 
@@
 Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
-This motivates a more general pooling than~(\ref{eq:log_pool}) setting where features can be weighted depending on the target variable:
+
+This motivates a more general pooling than the common log-linear pool~(\ref{eq:log_pool}) where features can be weighted depending on the target variable:
 $$
-p_\lambda(x|\y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)}
+p_\lambda(x|\y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)},
 $$
-
-$p_0$ is an arbitrary reference (implicit if weights are class-independent as it is then effectless).
+where $p_0$ is an arbitrary reference feature distribution (which is implicit if weights are class-independent as it is then effectless).
+Make sense to have a reference as it gives a customizable baseline for detection.
 Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
@@
 
-This is the super composite likelihood idea in a nutshell.
+We can train this model by maximum likelihood, yielding a similar optimization problem as in Section~\ref{sec:training}.
 
-Again, we can train this model and see that it is the same as solving a maxent problem.
-Difference with above is that the constraints are not averaged over $x$ (basis functions are of the form $\chi(x-a)\rho_i(x,y)$) and that they involve arbitrary log-likelihood ratios vs some reference distribution.
-Make sense to have a reference as it gives a customizable baseline for detection.
-
-If some classes are merged or split, then the other classes keep the same weights (assuming $h(\y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
+%If some classes are merged or split, then the other classes keep the same weights (assuming $h(\y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
 So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
@@
 
-%Link with PDF projection \cite{Baggenstoss-03}.
+Link with PDF projection \cite{Baggenstoss-03}.
 
@@
 At the end of the day, the interest of the super composite idea is questionable.
-
-\section{Data reduction inequality}
-\label{appendix:reduction_inequality}
-
-A fundamental property of the Kullback-Leibler divergence is that it decreases under feature extraction, {\em i.e.}, application of a deterministic transformation.
-This comes as a consequence of the logarithmic sum inequality \cite{Cover-06} (and also follows as a straightforward corollary of the PDF~projection theorem \cite{Minka-04,Baggenstoss-15}).
-
-The proof is elementary and can be sketched as follows.
-Let $Z=f(Y)$ some feature extracted from a variable~$Y\sim p(y)$.
-Given an arbitrary distribution~$\pi(y)$, let  $\tilde{p}(z)$ and $\tilde{\pi}(z)$ the distributions induced on~$Z$ by $p(y)$ and $\pi(y)$, respectively.
-For any potential value~$z$ of~$Z$, consider the level set: $\Gamma(z)=\{y, f(y)=z\}$.
-Under conditions of existence, we can partition the integral involved in the KL~divergence $D(p\|\pi)$ using these level sets:
-$$
-\int p(y) \log \frac{p(y)}{\pi(y)} dy
-=
-\int \left( \int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy \right) dz\\
-.
-$$
-
-Applying the logarithmic sum inequality \cite{Cover-06} to each integral over~$\Gamma(z)$, we then readily get:
-$$
-\int_{\Gamma(z)} p(y) \log \frac{p(y)}{\pi(y)} dy
-\geq 
-\tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)}
-,
-$$
-owing to the fact that $\displaystyle \int_{\Gamma(z)} p(y) dy = \tilde{p}(z)$ by definition. Therefore,
-$$
-\int \tilde{p}(z) \log \frac{\tilde{p}(z)}{\tilde{\pi}(z)} dz
-\leq 
-\int p(y) \log \frac{p(y)}{\pi(y)} dy
-,
-$$
-or, in a more compact way, $D(\tilde{p}\|\tilde{\pi})\leq D(p\|\pi)$.
-The case of equality occurs if, and only if, $p(y)/\pi(y)=\tilde{p}(z)/\tilde{\pi}(z)$, in other words, if~$z$ is a sufficient statistic for the ``alternative'' hypothesis~$H_1:y\sim p(y)$ versus the ``null'' hypothesis~$H_0:y\sim\pi(y)$.
-
-Since it is a well-known fact that $D(\tilde{p}\|\tilde{\pi})\geq 0$ as any KL~divergence, we conclude that:
-\begin{equation}
-\label{eq:reduction_inequality}
-0 
-\leq D(\tilde{p}\|\tilde{\pi}) 
-\leq D(p\|\pi)
-.
-\end{equation}
-
-\section{Basic asymptotic properties of composite likelihood}
-\label{appendix:asymptotic}
-
-It follows from the double inequality~(\ref{eq:reduction_inequality}) that, for any extracted feature~$z_i$ and for any hypothesis~$\theta$,
-$$
-0 \leq
-E\left[
-\log \frac{p(z_i|\theta_\star)}{p(z_i|\theta)}
-\right]
-\leq
-E\left[
-\log \frac{p(y|\theta_\star)}{p(y|\theta)}
-\right],
-$$
-where the expectation is taken with respect to the true distribution $p(y|\theta_\star)$.
-Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any standard composite likelihood function (assuming unit sum positive weights):
-\begin{equation}
-\label{eq:variation_bound}
-0 \leq
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta, \mathbf{w})} \right]
-\leq 
-E\left[ \log \frac{L(\theta_\star)}{L(\theta)} \right]
-.
-\end{equation}
-
-This implies two asymptotic properties of standard composite likelihood: 
-\begin{itemize}
-\item {\em Consistency.} The expected log-composite likelihood is maximized by~$\theta_\star$, the true value of~$\theta$.
-\item {\em Conservativeness.} The expected log-composite likelihood function is ``flatter'' than the true expected log-likelihood.
-In other words, composite likelihood ratios $L_c(\theta, \mathbf{w})/L_c(\theta_\star, \mathbf{w})$ relative to~$\theta_\star$ tend to be larger than the corresponding true likelihood ratios.
-\end{itemize}
-
-These properties do not necessarily extend to the more general case of SCL. We may state a weaker consistency property by considering the upper envelope of expected log-composite likelihood ratios:
-$$
-M(\theta) = \max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
-$$
-where ${\cal W}$ is the $n$-dimensional simplex. Clearly, the expected logarithm of the SCL is upper bounded by~$M(\theta)$:
-$$
-E[\log {\cal L}_c(\theta, \mathbf{W})] \leq M(\theta)
-.
-$$
-
-Moreover, $M(\theta)$ is maximized by $\theta_\star$ since (\ref{eq:variation_bound}) implies that, for any $(\theta,\theta_0)$,
-$$
-E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-\leq
-E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right],
-$$
-which remains true when taking the maximum over the weights in both sides: 
-$$
-M(\theta) = 
-\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-\leq
-\max_{\mathbf{w}\in{\cal W}} E\left[ \log \frac{L_c(\theta_\star, \mathbf{w})}{L_c(\theta_0, \mathbf{w})} \right]
-= M(\theta_\star)
-.
-$$
-
-Therefore, maximizing SCL corresponds to maximizing a lower bound on~$M(\theta)$, which can be considered as an objective function since it is maximized by~$\theta_\star$ (like the expected log-likelihood).
-Lower bound maximization guarantees a certain objective value, however not the maximum, hence asymptotic consistency may not hold.
 
@@
 
-\section{Feature relevance}
-
-Let $x_0$ a reference class and consider the log-probability ratio:
-$$
-\alpha = \log \frac{p_\lambda(x|\y)}{p_\lambda(x_0|\y)}
-$$
-
-The larger $\alpha$ for a given $x_0$, the happier we are.
-Let us average $\alpha$ wrt $x$ and $y$ as well as $x_0\sim \pi(x_0)$, which in this context may be interpreted as the outcome of a random classifier.
-Ignoring offsets, we have:
-$$
-E[\alpha]
-= \sum_i \lambda_i \int \pi(x_0)\pi(x) p(z_i|x) \log\frac{p(z_i|x)}{p(z_i|x_0)} dz_i dx dx_0
-= \sum_i \lambda_i \bar{d}_i
-$$
-with:
-$$
-\bar{d}_i =
-\int \int \pi(x_0) \pi(x) D(p_i^x\|p_i^{x_0}) dx dx_0
-$$
-
-If we have offsets, then there is an offset in this expresion, meaning that part of the log-probability ratio is accounted for by the offsets... this is not important.
-
-Hence, the expected log-likelihood ratio is a weighted sum of positive feature-specific contributions $r_i=\lambda_i \bar{d}_i$.
-The larger the contribution, the more influential the feature on the average.
-The analysis can be refined by skipping the averaging over $x_0$, in which case feature influence becomes dependent on the comparison class $x_0$...
-so it's all more complicated and potentially confusing.
-It would make sense to do that in the super-composite framework, where the reference class is fixed.
-
-The bottom line is that global feature relevance is determined by the CL weight(s) but also by the KL divergence matrix.
-
-Note that, by the information processing inequality, we have $D(p_i^x\|p_i^{x_0}) \leq d_{\rm lim}(x,x_0)\equiv D[p(y|x)\|p(y|x_0)]$, hence $\bar{d}_i\leq \bar{d}_{\rm lim}$ and:
-$$
-r_i \leq \lambda_i \bar{d}_{\rm lim}
-$$
-so if we normalize relevance according to unit sum weights, no single feature can be more relevant than the full data.
-
-How about {\em local} relevance? We simply replace $d_{k\ell i}$ by the absolute empirical log-likelihood ratio $|\log p_k(z_i)/p_\ell(z_i)|$. That's it.
-
 \bibliographystyle{abbrv}

### 2019-02-14 papers/composite_inference/draft.tex
Modifs 14.2.19

@@
 
-Unit sum weights, however, implicitly assume strong redundancy between features and thus tend to produce over-conservative predictions when features are weakly correlated, as is generally the case in practice since features are supposed to be chosen for complementarity.
-An ideal method to tune weights is one that can capture and adjust for the pattern of statistical dependences between features.
-This calls for a training approach, which, as we will see, is very unlikely to pick constant or unit sum weights.
+Unit sum weights, however, implicitly assume strong redundancy between features and thus tend to produce over-conservative predictions when features are weakly correlated (see Appendix~\ref{appendix:frequentist}), as is generally the case in practice since features are usually chosen for complementarity.
+An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}.
+This may be tackled using a learning approach, which, as we will see, is unlikely to pick constant or unit sum weights.
 
@@
 
-Given a labeled training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct approach to tune the composite weights is to maximize their likelihood (or, equivalently, minimize their ``cross-entropy'') under the composite predictive distribution:
+Given a labeled training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct way to tune the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
 \begin{equation}
@@
 $$
-\ell_i(x,\y) \equiv \log p(z_i|x),
+\ell_i(x,\y) = \log p(z_i|x),
 \qquad
-a(\blambda, y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
+a(\blambda, \y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
 $$
 
-It follows that the utility function $U(\blambda)$ is concave as a general property of likelihood in exponential families, hence the training can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:training} for details.
+It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence the training of weights can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:training} for details.
 
-Because some positive weight constraints may turn out inactive, the training has a natural tendency to produce sparse weights.
-As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, clearly showing that maximum likelihood training differs from univariate feature selection as it takes feature redundancy into account.
+Because some positive weight constraints may turn out inactive, the optimization has a natural tendency to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood training differs from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
 
@@
 Other training variants that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family.
-For instance, class-dependent offsets may be added, hence learning the ``prior'' together with the composite weights.
+For instance, class-dependent offsets may be added so as to learn the ``prior'' together with the composite weights.
 It is also possible to perform unconstrained optimization to achieve higher training likelihood, meaning that some composite weights are then negative.
@@
 Exponential family properties also imply that maximum likelihood training~(\ref{eq:train_likelihood}) is dual to a maximum entropy problem or, more precisely, an $I$-projection \cite{Csiszar-84}:
-$$
+\begin{equation}
+\label{eq:maxent}
 \min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{\pi(x)h(\y)},
-$$
+\end{equation}
 subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where 
@@
 
-The marginal constraint $p(\y)=h(\y)$ makes the $I$-projection a special case of maximum entropy classifier \cite{BergerA-96}, ensuring that only the conditional distribution $p(x|\y)$ is optimized.
+The marginal constraint $p(\y)=h(\y)$ makes the $I$-projection equivalent to a maximum entropy classifier \cite{BergerA-96}, ensuring that only the conditional distribution $p(x|\y)$ is optimized.
 The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
@@
 Hence, any $I$-projection is in some sense a best possible predictive model under particular constraints (in the limit of validity of empirical approximations).
-Our training of composite weights belongs to the same family as multinomial logistic regression, another maximum entropy classifier, only differing by the choice of constraints, which, in the latter case, are not related to generative models.
+Our training of composite weights relates to multinomial logistic regression, another maximum entropy classifier, only differing by the choice of constraints, which, in the latter case, are not related to generative models.
 
-As an alternative to the strategy considered here, the mean log-likelihood constraints may be expressed for each class rather than being averaged w.r.t.~the prior, leading to a more general predictive model that we call ``super composite likelihood'', in which the weights become class-dependent (see Appendix~\ref{appendix:super}).
+As an alternative to the strategy considered here, the mean log-likelihood constraints may be expressed for each class rather than being averaged with respect to the prior, leading to a more general predictive model that we call ``super composite likelihood'', in which the weights become class-dependent (see Appendix~\ref{appendix:super}).
 This variant has the potential for higher prediction performance, but poses some conceptual difficulties, in particular that of having to define an arbitrary reference class.
@@
 
-{\color{red} It is easy to build a good classifier.
+It is easy to build a good classifier.
 You just need to pick a parametric family that is big enough to produce a good fit yet small enough to avoid overfitting.
@@
 Deep learning is cool but it is relevant to AI only if the representations it learns are generic enough to be used in other problems, i.e.
-it should ultimately be validated via transfer learning.}
+it should ultimately be validated via transfer learning.
 
@@
 Consider mutually independent features: each one recieves a unit weight, but some may be independent from target.}
+
+{\color{red} Can we use the same dataset to learn both the feature generative distributions and the composite weights? In principle, nothing speaks against this.}
 
@@
 
-{\color{red} Can we use the same dataset to learn both the feature generative distributions and the composite weights? In principle, nothing speaks against this.}
+{\color{red} Can we interpret maximum weight likelihood as a ``maximum relevance minimum redudancy'' feature selection approach?}
+
+%In other words, is $\bar{a}(\blambda)$ a measure of feature redundancy?
+At fixed $\y$, if one $\lambda_i\to\infty$, then $a(\blambda,\y)\ot\infty$ except if $\ell_i(x,\y)\preceq 0$, in which case it goes to $-\infty$.
+If using likelihood ratios, this only happens if the reference class is the most likely for $\y$.
+But then we can prove that it can't happen for all $\y$ (otherwise the associated moment would be negative), so basically $\bar{a}(\blambda)$ blows up when at least one weight gets large.
+And if several weights get large simultaneously, then it blows up even more so we might consider it a redundancy measure to some extent...
 
@@
 
-Assume that the data~$\y$ is distributed according to some target value~$x_\star$. For any extracted feature~$z_i$ and for any hypothetical target value~$x$,
+Let a multivariate observable variable~$\y \sim p(\y|x_\star)$ distributed according to some target value~$x_\star$.
+For any extracted feature~$z_i$ and any hypothetical target value~$x$, the following double inequality holds:
 $$
@@
 $$
-where the expectation is taken with respect to the true distribution $p(\y|x_\star)$.
-Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any composite likelihood function:
+where the expectation is taken with respect to the true distribution $p(\y|x_\star)$. This fact follows from two basic properties of KL~divergence: positivity and partition inequality.
+
+Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any composite likelihood function~(\ref{eq:comp_lik}) with positive weights~$\blambda\succeq 0$:
 \begin{equation}
@@
 \leq 
-\|\blambda\|_1 E\left[ \log \frac{L(x_\star)}{L(x)} \right]
-.
+s_\blambda E\left[ \log \frac{L(x_\star)}{L(x)} \right]
+,
 \end{equation}
+with $s_\blambda\equiv \|\blambda\|_1 =\sum_i \lambda_i$, and $L(x)$ the true likelihood function, {\em i.e.} $L(x)\equiv \log p(\y|x)$.
 
-This implies two asymptotic properties of standard composite likelihood: 
+This implies two general asymptotic properties of composite likelihood:
 \begin{itemize}
 \item {\em Consistency.} The expected composite log-likelihood is maximized by~$x_\star$.
-\item {\em Conservativeness.} For any~$x$, if the expected true log-likelihood of~$x$ is finite, then the corresponding expected composite log-likelihood ratio is finite too. 
+\item {\em Conservativeness for unit sum weights.} If the weights sum up to one or less ($s_\blambda\leq 1$), composite likelihood ratios of the true target~$x_\star$ vs.~other target values~$x$ tend to be underestimated, {\em i.e.} are smaller on the average than corresponding true likelihood ratios.
 \end{itemize}
@@
 
-This reads: $U(\blambda) = \sum_k \log \pi(x_k) + N \psi(\blambda)$ with: 
+Maximizing $U(\blambda)$ is therefore equivalent to maximizing: 
 $$
-\psi(\blambda) \equiv \blambda^\top \bar{\bell} - \frac{1}{N} \sum_k a(\lambda,\y_k), 
+\psi(\blambda) \equiv \blambda^\top \bar{\bell} - \bar{a}(\blambda), 
 $$
-and:
+with:
 $$
 \bar{\bell} \equiv \frac{1}{N} \sum_k \bell(x_k,\y_k),
+\qquad
+\bar{a}(\blambda) \equiv \frac{1}{N} \sum_k a(\blambda,\y_k).
 $$
-therefore maximizing~$U(\blambda)$ is equivalent to maximizing~$\psi(\blambda)$.
+
+Note that $\psi(\blambda)$ is nothing but the dual function associated with the maxent problem~(\ref{eq:maxent}).
 
@@
 $$
-\nabla\psi
-= \bar{\bell} - \frac{1}{N}\sum_k \nabla a(\blambda,\y_k)
+\nabla\psi(\blambda)
+= \bar{\bell} - \frac{1}{N} \sum_k \nabla a(\blambda,\y_k) 
 $$
@@
 $$
-\nabla\nabla^\top\psi
-= - \frac{1}{N}\sum_k
-\nabla \nabla^\top a(\blambda, \y_k)
+\nabla\nabla^\top\psi(\blambda)
+= - \frac{1}{N} \sum_k \nabla \nabla^\top a(\blambda,\y_k)  
 $$
@@
 $$
-\nabla a = \E_{\blambda}(\bell|\y),
+\nabla a(\blambda,\y) = \E_{\blambda}(\bell|\y),
 \qquad
-\nabla \nabla^\top a = \text{Var}_{\blambda}(\bell|\y)
+\nabla \nabla^\top a(\blambda,\y) = \text{Var}_{\blambda}(\bell|\y)
 $$
-proving that $a(\blambda,\y)$ is convex in $\blambda$ for any $\y$, and in turnes proves the concavity of $\psi$.
+proving that $a(\blambda,\y)$ is convex in $\blambda$ for any $\y$, and in turn proves the concavity of $\psi$.
 
@@
 Still interpretable, that's why it is worthwile mentioning.
-
-%I was thinking about the unit sum weight version...
-The diagnosis example could be one in which clues are independent, in which case $\lambda_i(x)\equiv 1$ would be optimal, making any super composite trick irrelevant.
-Also see that with hypothesis-dependent weights, we may lose the estimation consistency.
-At the end of the day, the interest of the super composite idea is questionable.
 

### 2019-04-01 papers/composite_inference/draft.tex
Changes last week of March 2019

@@
 \title{Composite Bayesian inference}
-\date{}
+%\date{}
 \author{Alexis Roche\thanks{\url{alexis.roche@centraliens.net}}}
@@
 \begin{abstract}
-We revisit the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
-This view makes it natural to use a machine learning approach to calibrate the weights associated with composite likelihood.
-For instance, they can be tuned in a typical supervised learning stage so as to minimize cross-entropy on a labeled training dataset, yielding a convex problem.
-We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
+We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
+The resulting class of predictive models, which we call ``composite Bayesian'', is a middle way between purely generative and purely discriminative models.
+This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning: in the latter case, they may be tuned so as to maximize training likelihood, yielding an easy-to-solve convex problem.
+We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
 \end{abstract}
@@
 Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this double requirement may be very challenging for multi-dimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in the minimum description length paradigm \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include ``shallow'' learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
+Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
 
-Discriminative models, however, lack the ability to establish causality links between data instances and predictions, and therefore cannnot ``support'' their own predictions in a built-in manner.
-Consider as a straightforward example the problem of deciding whether a pan is a sauce pan or a frying pan based on its depth.
-A discriminative model can learn an optimal threshold that correctly classifies most pans in practice, but it cannot gain an internal representation that sauce pans tend to be deeper than frying pans, which is {\em why} the thresholding works.
-To interpret a discriminative model means to ``reverse engineer'' it using external knowledge, which quickly becomes an impossible task as the data dimension increases.
+Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves.
+Consider as a straightforward example the problem of deciding whether an object is a sauce pan or a frying pan based on depth.
+A discriminative model can learn a threshold that correctly classifies most pans in practice, but is unable to see that, say, a blender does not match either category.
+The impossibility of such confirmatory analysis strongly limits interpretation if outlier inputs are to be expected.
 
 There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
-For instance, in medical diagnosis, automated disease predictions should always be corroborated by individualized findings to be communicable and thus be taken into consideration by clinicians -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
-Clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
-
-%, {\em e.g.}, by pinpointing biomarkers that deviate from healthy reference ranges.
-
-% simple methods based on range analysis of univariate biomarkers can provide meaningful clues, and are thus more useful in practice although less powerful in terms of prediction performance.
-
-%With generative models you can compare data with expected outcomes to get insight.
-
-%the information they provide to support them is limited to the boundaries between classes, which are difficult to interpret for high-dimensional data representations.
-
-%Decision making is not about pooling opinions, it is about pooling empirical observations and find which hypothesis they point to.
-
-%Need to emphasize why we need generative models.
-Below are some valid reasons: unsupervised learning, potentially better performance on small training datasets...
-But the \#1 reason is that discriminative models are black boxes: they provide an answer but no explanation to support it.
-Truth is, they perform better in supervised big data context in the sense of accuracy, cross-entropy, and so on, no question about that -- but they are mere inference recipes.
-Is that what AI should be?
-Sometimes, yes.
-But there are application fields where this clearly cannot work, for instance medical diagnosis.
+For instance, in medical diagnosis, automated disease predictions should be corroborated by individualized findings to be communicable to clinicians.
+In other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
+On the other hand, clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
 
@@
 
-Another limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as {\em marginally independent} variables, meaning that the data conveys no information about the parameters unless the target variable is observed.
+A related limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as {\em marginally independent} variables, meaning that the data conveys no information about the parameters unless the target variable is observed.
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graphs representing generative and discriminative models.
-For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence requires a larger training set to achieve optimal performance \cite{Ng-01}.
+For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence it requires more training data to achieve optimal performance \cite{Ng-01}.
 
@@
 
-This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a means to incorporate generative information into a discriminative model.
-Rather than attempting at a full generative model, one may combine multiple low-dimensional generative models for different pieces of information extracted from the input data.
-Each ``small'' model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
+This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both computationally efficient and interpretable, resulting in a class of hybrid models.
+The key idea is to combine multiple low-dimensional generative models corresponding to different pieces of information extracted from the input data.
+Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
 The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
-This idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
-Importantly, predictions made in such a way are interpretable owing to their underlying generative nature.
+This may be understood as a probabilistic version of boosting, or as a {\em divide and conquer} approximation to the intractable Bayesian posterior.
 
-When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known as composite likelihood (CL) in computational statistics.
-As detailed in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
-While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} and generally offers an excellent trade-off between computational and statistical efficiency.
+%Importantly, predictions made in such a way are interpretable owing to their underlying generative nature.
 
-With the opinion pooling interpretation in mind, the feature weights in CL may be optimized for prediction performance in a typical discriminative learning scenario.
+When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL).
+As discussed in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
+While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency.
+
+With the opinion pooling interpretation in mind, the weights assigned to the different features may be optimized for prediction performance in a typical discriminative learning scenario.
 As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.
-These likelihood functions themselves may need to be pre-trained, therefore the composite inference approach typically invovles a double training scheme: a generative training stage (to learn feature-based likelihood parameters) followed by a discriminative training stage (to learn feature weights in aggregation).
+The likelihood functions themselves may need to be pre-trained, therefore the composite inference approach typically involves a double training scheme: a generative training stage (to learn the feature-based likelihood parameters) followed by a discriminative training stage (to learn the feature weights in aggregation).
 
 Technical details are given in the remainder.
+We warn the reader that slightly abusive mathematical notations are used deliberately for the sake of clarity in places where we think that they are unambiguous.
 
@@
 
-This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated.
-A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(\y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
+This requires a plausible generative model which, for complex data, may be out of reach or involve too many nuisance parameters.
+A natural workaround known as data reduction is to extract some lower-dimensional representation $z(\y)\sim f(z|x)$ using a many-to-one mapping, and consider the potentially more convenient likelihood function:
 $$
-\ell(x) = p(z|x)
+\ell(x) = f(z|x)
 .
@@
 While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction.
-To make the trick statistically more efficient, we may extract several features, $z_i=f_i(\y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.
+To make the trick statistically more efficient, we may extract several features, $z_i(\y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = f(z_i|x)$ that they elicit.
 
-If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-One that is appealing as it tends to produce single peaked distributions is the {\em generalized logarithmic opinion pool}:
+If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
+One that is appealing as it tends to produce single peaked distributions is the (generalized) log-linear opinion pool:
 \begin{equation}
 \label{eq:log_pool}
-p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
+p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n f(z_i|x)^{\lambda_i},
 \end{equation} 
 where $\pi(x)$ is a prior distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
-Also note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
+Note that it is possible to have the weights depend on the data~$\y$, but this would make the method considerably more complex and less interpretable.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
 
@@
 
-The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding  the form: 
+The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule as it yields the form: 
 $$
@@
 The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
-The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{appendix:conditional}.
+The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}.
 The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.
 
-CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
-While the above derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method.
-However, unitary CL weights do not guarantee optimal prediction performance.
+CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
+While the CL derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method, however unitary CL weights do not guarantee optimal prediction performance.
 
-\section{Prediction interpretation}
+\section{Composite likelihood calibration}
+\label{sec:calibration}
 
-A compelling advantage of the composite approach is to make it easy to interpret a prediction for any instance~$\y$.
-Given two putative target values $x_0$ and $x_1$, the log-probability ratio of $x_1$ vs~$x_0$ is computed by simply adding up single-feature contributions:
-$$
-\log \frac{p_\blambda(x_1|\y)}{p_\blambda(x_0|\y)}
-= 
-\underbrace{\log \frac{\pi(x_1)}{\pi(x_0)}}_{\text{offset}}
-+ \sum_i \underbrace{\lambda_i \log \frac{p(z_i|x_1)}{p(z_i|x_0)}}_{\text{feature contribution}},
-$$
-therefore it is straightforward to assess which features contribute most to a particular hypothesis comparison (either positively or negatively), see possible disagreements between features, and corroborate predictive probabilities with feature-level testable information.
+CL involves two types of parameters:
+\begin{enumerate}
+    \item ``generative'' parameters, {\em i.e.}, the parameters of the feature generation models $f(z_i|x)$; 
+    \item ``discriminative'' parameters, {\em i.e.},  the weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ assigned to the feature-based likelihood functions in~(\ref{eq:log_pool}).
+\end{enumerate}
 
-Each feature contributes to an hypothesis comparison proportionally to its associated log-likelihood ratio test statistic.
-Contrary to statistical hypothesis testing, likelihood ratios are not converted into $p$-values but are directly combined into a predictive probability ratio.
-Non-unitary composite weights reinforce or inhibit features compared to na\"ive Bayes prediction ($\blambda\equiv 1$), and may therefore account for mutual feature dependence if tuned appropriately.
-Keeping the analogy with hypothesis testing, composite weights may be thought of as the counterpart of multiple testing correction \cite{Benjamini-10}.
+If training data is available, the generative parameters may be set in a first stage using standard estimation techniques.
+Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to composite likelihood one of the same techniques as for traditional likelihood \cite{Berger-99}.
+We focus in the sequel on tuning the composite weights.
 
-To interpret the predictive distribution globally, the user(s) may go through successive pairwise hypothesis comparisons in a dynamic way, starting by comparing the automated prediction, {\em i.e.} the most probable target, to some educated guess or reference class (for instance, normality in medical diagnosis).
-Next, depending on the feature contributions highlighted by the comparison, they may submit another hypothesis for comparison to the automated prediction.
-This process may be repeated until the composite prediction is subjectively understood by the user(s), yielding a form of human machine dialog.
+\subsection{Agnostic weights}
 
-\section{Composite weight tuning}
-\label{sec:tuning}
+In the absence of knowledge, uniform weights should be chosen under a rule of indifference.
+CL is then a scaled version of na\"ive Bayes likelihood, the unique weight value being irrelevant to the maximum CL estimator (MCLE).
+To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic MCLE variance matrix \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, as proposed in previous attempts at Bayesian inference from composite likelihood.
 
-%So far, we have assumed that the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) are known.
+% in attempts to match frequentist and Bayesian notions of uncertainty.
 
-Let us now turn to the problem of tuning the composite weights~$\blambda=(\lambda_1,\ldots,\lambda_n)$ in~(\ref{eq:log_pool}) as part of the method calibration.
-It is assumed at this stage that the feature generative models are known, which may require a pre-training stage using standard parameter estimation techniques.
+Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one.
+This is motivated by the opinion pooling perspective and a characterization theorem \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator (for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome) is the log-linear opinion pool with unit sum weights.
+The external Bayesian  property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
+The log-linear pool with unit sum weights also happens to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
 
-\subsection{Previous work}
+%While other divergence measures lead to other optimal pooling operators (for instance, the exclusive KL divergence leads to the linear pool), they all enforce unit sum weights, as is natural in the search for a best compromise.
 
-In the absence of knowledge, a default rule is to choose constant weights.
-The CL function is then a scaled version of na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE).
-To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic MCLE variance matrix \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.
+Unit sum weights, however, implicitly assume maximum redundancy between features.
+Consequently, they tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}), meaning reduced prediction confidence when features are weakly correlated, as is desirable in general.
+An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}.
+This suggests a learning approach if possible, which, as we shall see, is unlikely to pick uniform or unit sum weights.
 
-Another justifiable recommendation for log-linear pooling weights is to sum up to one.
-This may be motivated by the fact that the only pooling operator that preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}.
-External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling agent opinions, provided that all agents agree on the same prior.
-A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the sum of Kullback-Leibler (KL) divergences to the agent opinions (see also \cite{Wang-14}, Theorem~3, for a related optimality argument).
+\subsection{Learning the weights}
+\label{sec:learning}
 
-Unit sum weights, however, implicitly assume strong redundancy between features and thus tend to produce over-conservative predictions when features are weakly correlated (see Appendix~\ref{appendix:frequentist}), as is generally the case in practice since features are usually chosen for complementarity.
-An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}.
-This may be tackled using a learning approach, which, as we will see, is unlikely to pick constant or unit sum weights.
-
-\subsection{Maximum likelihood training}
-\label{sec:training}
-
-Given a labeled training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct way to tune the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
+Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
 \begin{equation}
@@
 $$
-\ell_i(x,\y) = \log p(z_i|x),
+\ell_i(x,\y) = \log f(z_i|x),
 \qquad
@@
 
-It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence the training of weights can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}, see Appendix~\ref{appendix:training} for details.
+It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence the learning can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}.
+See Appendix~\ref{app:training} for some implementation details.
 
-Because some positive weight constraints may turn out inactive, the optimization has a natural tendency to produce sparse weights.
-As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood training differs from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
+Because some positive weight constraints may turn out inactive, learning has a natural tendency to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
 
@@
 
-As is customary and is generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$ (enforcing balanced classes if the prior is uniform).
-Also, some regularization may be useful for stability: we shall in practice maximize $U(\lambda)-\alpha \|\blambda\|^2$, where $\alpha>0$ is a small damping factor.
+As is customary and generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$, enforcing balanced classes if the prior is uniform.
+Also, some regularization may be needed for stability: we shall in practice maximize $U(\lambda)-\alpha \|\blambda\|^2$, where $\alpha>0$ is a small damping factor.
 
-Other training variants that we mention here but do not particularly recommend for interpretability include adding basis functions to the exponential family.
+Other training variants that we mention here but do not particularly recommend for interpretability arise from adding basis functions to the exponential family.
 For instance, class-dependent offsets may be added so as to learn the ``prior'' together with the composite weights.
-It is also possible to perform unconstrained optimization to achieve higher training likelihood, meaning that some composite weights are then negative.
+It is also possible to perform unconstrained optimization to achieve higher training likelihood, but this means that some composite weights are then negative, in violation of the monotonicity property mentionned in Section~\ref{sec:log_pool}.
 
-\subsection{Maximum entropy interpretation}
+\paragraph{$I$-projection interpretation.}
 
-Exponential family properties also imply that maximum likelihood training~(\ref{eq:train_likelihood}) is dual to a maximum entropy problem or, more precisely, an $I$-projection \cite{Csiszar-84}:
+%D(p\|p_0) = \sum_{x\in{\cal X}}\sum_{\y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{p_0(x,\y)},
+Exponential family properties also imply that learning via likelihood maximization~(\ref{eq:train_likelihood}) is dual to an $I$-projection \cite{Csiszar-84}: 
 \begin{equation}
 \label{eq:maxent}
-\min_{p\in{\cal P}} D(p\|\pi \cdot h) = \sum_{x\in{\cal X}}\sum_{y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{\pi(x)h(\y)},
+\min_{p\in{\cal P}} D(p\|p_0),
+\qquad \text{with} \quad
+p_0(x,\y) \equiv \pi(x)h(\y),
 \end{equation}
-subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where 
-${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$
-with ${\cal Y}=\{\y_1,\ldots,\y_N\}$ the set of training data values, $h(x,\y)$ is the joint empirical distribution of targets and data, and $h(\y)$ is the corresponding marginal data distribution.
-Note that the constraints implicitly approximate the data marginal and the mean log-likelihood values by empirical estimates, suggesting the need for regularization on small training sets.
+subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where $h(x,\y)$ is the joint empirical distribution of targets and data examples, $h(\y)$ is the corresponding marginal data distribution, and ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$ with ${\cal Y}\equiv\{\y_1,\ldots,\y_N\}$.
+Note that the constraints implicitly approximate the data marginal and the mean log-likelihood values by their empirical estimates, confirming the need for regularization in practice.
 
-The marginal constraint $p(\y)=h(\y)$ makes the $I$-projection equivalent to a maximum entropy classifier \cite{BergerA-96}, ensuring that only the conditional distribution $p(x|\y)$ is optimized.
+The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized, hence the $I$-projection is equivalent to maximum conditional entropy \cite{BergerA-96}.
 The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
-This is weaker a constraint than assuming that the feature generative models $p(z_i|x)$ are true: it only assumes that the ``description length'' \cite{Grunwald-07} achieved by each agent is truthworthy.
+This is weaker a constraint than assuming that the feature generative models $f(z_i|x)$ are true: it only assumes that the {\em description length} \cite{Grunwald-07} achieved by each feature model is a truthworthy upper bound.
 
@@
 
-A particularly insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is under broad conditions a minimax strategy for prediction according to the log loss.
-Hence, any $I$-projection is in some sense a best possible predictive model under particular constraints (in the limit of validity of empirical approximations).
-Our training of composite weights relates to multinomial logistic regression, another maximum entropy classifier, only differing by the choice of constraints, which, in the latter case, are not related to generative models.
+An insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss.
+Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is inexact due to the limited training set size.
+Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by the underlying knowledge, which in this case may be interpreted in terms of maximum feature description length.
 
-As an alternative to the strategy considered here, the mean log-likelihood constraints may be expressed for each class rather than being averaged with respect to the prior, leading to a more general predictive model that we call ``super composite likelihood'', in which the weights become class-dependent (see Appendix~\ref{appendix:super}).
+\section{Super composite likelihood}
+\label{sec:super}
+
+So far, we have assumed that each agent is given a single weight in the opinion aggregation~(\ref{eq:log_pool}).
+Let us now consider a more general pooling operator where agents can be weighted depending on the target variable:
+\begin{equation}
+\label{eq:super_pool}
+p_\lambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
+\end{equation}
+where $x_0$ is some fixed target value, which we call the reference.
+As discussed in Appendix~\ref{app:super}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting functions~$w_i(x)$ sum up to one uniformly,
+$$
+\forall x\in{\cal X},\quad
+\sum_{i=1}^n w_i(x) = 1,
+$$
+hence providing a sensible default rule to tune the weights.
+Note that (\ref{eq:super_pool}) amounts to a log-linear pool if the weights are target-independent as the reference is then effectless.
+Using the same analogy with Bayes rule as in Section~\ref{sec:bayes_rule}, we see that the quantity:
+\begin{equation}
+\label{eq:super_comp_lik}
+L^{sc}_\blambda(x) \equiv 
+\prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)}
+\end{equation} 
+plays the role of a standard likelihood at prediction time.
+We call this form super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs.
+SCL further generalizes CL since $L^{sc}_\blambda(x)=L^c_\blambda(x)/L^c_\blambda(x_0)$ for target-independent weights $\lambda_i(x)\equiv \lambda_i$, hence it is in this case proportional to CL and yields the same predictive distribution.
+
+Conversely, when each target assigns a weight one to a single feature and zero to all the others, SCL turns out to be identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, the ``PDF projection theorem'', which in fact boils down to an $I$-projection \cite{Minka-04}.
+While our derivation stems from a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
+
+%Case of a single nonzero agent per feature.
+Then SCL is equivalent to the class-specific likelihood derived in \cite{Baggenstoss-03} based on a generative model selection argument  Our derivation is much different but interesting to see that we end up with a more general form.
+
+A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which happen if the feature generative distributions $f(z_i|x)$ are unknown for some target values~$x$.
+It is possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs.
+Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence on the predictive distribution.
+
+%Interest?
+If some feature likelihood values are missing, i.e.
+agents are not able to model the whole set of targets.
+It's a bit like censoring.
+Then it's ok if they can model at least two targets including the reference: $\ell_i(x)$ is unknown for the other values but we can give them zero weight so it doesn't matter.
+
+If training data is available, the SCL weights may be learned by maximum likelihood as in Section~\ref{sec:learning}, yielding a similar convex problem similar.
+Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen.
+The $I$-projection interpretation still holds, yet with class-specific mean-value constraints:
+$$
+\forall a\in{\cal X}, \forall i=1,\ldots,n,
+\quad
+\E_p\left[\delta_{xa}\log \frac{f(z_i|x)}{f(z_i|x_0)}\right]
+\geq \E_h\left[\delta_{xa}\log \frac{f(z_i|x)}{f(z_i|x_0)}\right],
+$$
+where $\delta$ denotes the Kronecker delta.
+%$$
+%\forall x\in{\cal X}, \forall i=1,\ldots,n,
+%\quad
+%p(x) \E_{f(z_i|x)}\left[\log \frac{f(z_i|x)}{f(z_i|x_0)}\right]
+%\geq h(x) \E_{h(z_i|x)}\left[\log \frac{f(z_i|x)}{f(z_i|x_0)}\right]
+%$$
+
+%Another shortcoming is that SCL is not guaranteed to be asymptotically consistent, see Appendix~\ref{app:frequentist}, so it may yield biased MCLE despite
+
+%Disadvantage: interpetation is relative to an arbitrary class $x_0$.
+Advantages: richer model (so better predictions), probabilty ratios stable to class addition.
+Drawbacks: require a reference, no more consistency.
+Still interpretable, that's why it is worthwile mentioning.
+
+%Argue that SCL more robust than CL to mispecification of feature models.
+Intuitively, if a feature model is poor, the empirical description length is large, hence can be achieved with a `small' Lagrange multiplier $\lambda_i(x)$.
+Therefore, unreliable likelihoods $\ell_i(x)$ tend to get low weights.
+
+% which does not contradict the axiomatic derivation in \cite{Genest-86b} as it restricted itself to operators that only depend on the probas evaluated at~$x$ unlike~(\ref{eq:super_pool}).
+
+%The real motivation: when agents operate on restricted target sets.
+Without loss of generality, assume binary sets.
+Trick: for each such agent, say we have $p(z_i|x_1)$ and $p(z_i|x_0)$.
+We define $p(z_i|x)=p(z_i|x_0)$ if $x\not= x_1$.
+It does not make any sense but the likelihood ratio is then one for any $x\not= x_1$ so won't hamper the system.
+
+%Due to the distribution of weights between clues, a drawback of composite likelihood is that it is prone to {\em information overload} in the sense that it tends to ``flatten out'' when too many clues are included as relevant clues then get downweighted.
+If one decides not to merge clues for computational reasons (this would require handling their joint distribution), one could hope to mitigate information overload by assigning strong weights only to those clues that are believed to be ``most informative''.
+% When chosen for computational simplicity, features may not only convey limited information at individual level, their informativeness may also be very much class-dependent.
+Consider, for instance, diagnosing a disease from a routine medical checkup.
+Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
+
+%Reference is implicit if weights are class-independent as it is then effectless.
+Make sense to have a reference as it gives a customizable baseline for detection.
+Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
+healthy population).
+In this case, the super composite likelihood is equivalent to the standard one in the two-class case.
+
+%If some classes are merged or split, then the other classes keep the same weights (assuming $h(\y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
+So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
+
+%Super composite likelihood is essentially about combining binary versions of composite likelihood. It is analogous to the ``one-versus-rest'' strategy in multiple logistic regression.
+
+%Still externally Bayesian if weights sum up to one in each class.
+Alternatively, we can train this model by maximum likelihood, yielding a similar optimization problem as in Section~\ref{sec:learning}.
+
+%Maxent interpretation.
+It is an alternative to the strategy considered above, the mean log-likelihood constraints may be expressed for each class rather than being averaged with respect to the prior, leading to a more general predictive model that we call ``super composite likelihood'', in which the weights become class-dependent (see Appendix~\ref{sec:super}).
 This variant has the potential for higher prediction performance, but poses some conceptual difficulties, in particular that of having to define an arbitrary reference class.
@@
 
+Beside greater flexibility than CL, a disadvantage of SCL it that interpretation becomes relative to an arbitrary reference~$x_0$.
+In some applications, there may exist a natural reference class to cling to in common practice (for instance, normality in medical diagnosis), but this is not always the case.
+
 It is easy to build a good classifier.
@@
 {\color{red} Can we interpret maximum weight likelihood as a ``maximum relevance minimum redudancy'' feature selection approach?}
+
+Is it new to see composite likelihood as opinion pooling?
+Yes it is.
+Some previous attempts were made to derive composite likelihood as some sort of consensus generative model \cite{Wang-14}.
+Also some previsous attemps at Bayesian inference from CL using asymptotic theory.
 
@@
 \section{Basic frequentist properties of composite likelihood}
-\label{appendix:frequentist}
+\label{app:frequentist}
 
 Let a multivariate observable variable~$\y \sim p(\y|x_\star)$ distributed according to some target value~$x_\star$.
-For any extracted feature~$z_i$ and any hypothetical target value~$x$, the following double inequality holds:
+For any hypothetical target value~$x$ and any extracted feature $z_i\sim f(z_i|x)$, the following double inequality holds:
 $$
@@
 E\left[
-\log \frac{p(z_i|x_\star)}{p(z_i|x)}
+\log \frac{f(z_i|x_\star)}{f(z_i|x)}
 \right]
@@
 
+What about SCL???
+
 \section{Conditional composite likelihood}
-\label{appendix:conditional}
+\label{app:conditional}
 
-As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(\y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(\y)$, yielding the more general form:
+As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i(\y)$ considered as a predictor of the ``dependent'' feature, $z_i(\y)$, yielding the more general form:
 \begin{equation}
 \label{eq:cond_feat_lik}
-\ell_i(x) = p(z_i|x,z^c_i).
+\ell_i(x) = f(z_i|x,z^c_i).
 \end{equation}
@@
 \section{Composite likelihood training}
-\label{appendix:training}
+\label{app:training}
 
-Using the same notation as in Section~\ref{sec:training}, the likelihood function in~(\ref{eq:train_likelihood}) can be expanded as follows:
+Using the same notation as in Section~\ref{sec:learning}, the likelihood function in~(\ref{eq:train_likelihood}) can be expanded as follows:
 $$
@@
 
-\section{Super composite likelihood}
-\label{appendix:super}
+\section{Super log-linear pool}
+\label{app:super}
 
-When chosen for computational simplicity, features may not only convey limited information at individual level, their informativeness may also be very much class-dependent.
-Consider, for instance, diagnosing a disease from a routine medical checkup.
-Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
+Opinions: $p_i(x)$.
 
-This motivates a more general pooling than the common log-linear pool~(\ref{eq:log_pool}) where features can be weighted depending on the target variable:
+Log-linear pool:
 $$
-p_\lambda(x|\y) \propto \pi(x) \prod_i \left[\frac{p(z_i|x)}{p_0(z_i)}\right]^{\lambda_i(x)},
+p(x)\propto \prod_i p_i(x)^{\lambda_i}
 $$
-where $p_0$ is an arbitrary reference feature distribution (which is implicit if weights are class-independent as it is then effectless).
-Make sense to have a reference as it gives a customizable baseline for detection.
-Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
-healthy population).
-In this case, the super composite likelihood is equivalent to the standard one in the two-class case.
 
-We can train this model by maximum likelihood, yielding a similar optimization problem as in Section~\ref{sec:training}.
+What about having $\lambda_i(x)$ depend on the target?
 
-%If some classes are merged or split, then the other classes keep the same weights (assuming $h(\y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
-So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
+Can we do: $p(x)\propto \prod_i p_i(x)^{\lambda_i(x)}$? Is it externally Bayesian?
 
-Super composite likelihood is essentially about combining binary versions of composite likelihood. It is analogous to the ``one-versus-rest'' strategy in multiple logistic regression.
+Let the opinions be multiplied by some function $f(x)$. The new opinions read: $q_i(x)=f(x)p_i(x)/z_i$ with $z_i=\int f p_i$. So the pool gives:
+$$
+q(x) \propto p(x) f(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)}
+$$
 
-Link with PDF projection \cite{Baggenstoss-03}.
+So, it's not EB because $z_i$ may differ from one expert to the other.
 
-Advantages: richer model (so better predictions), probabilty ratios stable to class addition.
-Drawbacks: require a reference, no more consistency.
-Still interpretable, that's why it is worthwile mentioning.
+It feels more reasonable to do:
+$$
+p(x) \propto \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)}
+$$
+
+Is it EB? We have:
+$$
+\frac{q_i(x)}{q_i(x_0)} 
+= \frac{f(x)p_i(x)}{f(x_0)p_i(x_0)}
+$$
+because the $z_i$'s cancel out. Therefore,
+$$
+q(x) \propto p(x) \left[\frac{f(x)}{f(x_0)}\right]^{\sum_i \lambda_i(x)}
+$$
+
+So it's EB if the unit sum weight constraints are satisfied.
+
+\section{Notes Genest}
+
+External Bayesian: pooling and updating commute.
+A pooling operator is compliant with the `likelihood principle' if the pool at $x$ depends on the opinions only through the probabilities $p_i(x)$.
+
+Characterizaton of externally Bayesian pooling operators (EBPOs).
+
+Any EBPO LP-compliant that does not depend on $x$ is a log-pool with unit sum weights:
+$$
+p(x) \propto \prod_i p_i(x)^{w_i}
+$$
+
+Any EBPO LP-compliant is a generalized log-pool with unit sum weights:
+$$
+p(x) \propto \pi(x) \prod_i p_i(x)^{w_i}
+$$
+
+Weights can be negative if domain of $x$ is finite.
+
+To characterize more general EBPOs (not LP-compliant), they introduce an equivalence class.
+Two sets of opinions are equivalent if any ratio of two experts is the same up to a constant factor.
+It means that $q\sim p$ if there exists $f(x)$ such that
+$q_i(x) \propto f(x) p_i(x)$, $\forall i$.
+
+Define an arbitrary member $p_{a1},p_{a2},\ldots$ in each equivalence class.
+
+Any EBPO is of the form:
+$$
+T(p_1,p_2,\ldots) \propto \pi_a(x) v_a(x) \frac{p_1(x)}{p_{a1}(x)},
+\qquad
+v_a \geq \max (p_{a1},p_{a2},\ldots)
+$$
 

### 2019-04-01 papers/composite_inference/draft.21mar19.tex
Modifs Apr 1 2019

--- 2019-02-14 papers/composite_inference/draft.tex
@@
 This view makes it natural to use a machine learning approach to calibrate the weights associated with composite likelihood.
-For instance, they can be tuned in a typical supervised learning stage so as to minimize cross-entropy on a labeled training dataset, yielding a convex problem.
+In a supervised learning scenario, for instance, they can be set so as to minimize training cross-entropy, yielding an easy-to-solve convex problem.
 We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
@@
 Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this double requirement may be very challenging for multi-dimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in the minimum description length paradigm \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include ``shallow'' learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
+Because this double requirement may be very challenging for multi-dimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in the minimum description length paradigm \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical ``shallow'' learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
 
-Discriminative models, however, lack the ability to establish causality links between data instances and predictions, and therefore cannnot ``support'' their own predictions in a built-in manner.
+% Retravailler cette histoire de causalite.
+Ce qui est dit dans ce pragaraphe est faux.
+N'import quel seuillgage fait apparaitre la profondeur de l'objet comme "cause" de la classification.
+Ce qu'un modele discriminant ne permet pas, c'est de dire si les donnees observees sont "typiques" de telle ou telle classe.
+
+Discriminative models, however, lack the ability to establish {\em causality} links between data instances and predictions, and therefore cannnot {\em explain} their own predictions.
 Consider as a straightforward example the problem of deciding whether a pan is a sauce pan or a frying pan based on its depth.
-A discriminative model can learn an optimal threshold that correctly classifies most pans in practice, but it cannot gain an internal representation that sauce pans tend to be deeper than frying pans, which is {\em why} the thresholding works.
-To interpret a discriminative model means to ``reverse engineer'' it using external knowledge, which quickly becomes an impossible task as the data dimension increases.
+A discriminative model can learn a threshold that correctly classifies most pans in practice, but it cannot gain an explicit internal representation that sauce pans tend to be deeper than frying pans, which is {\em why} the thresholding works.
+The only way to interpret a discriminative model is to ``reverse engineer'' it using external knowledge, which is an almost impossible task for high-dimensional models.
 
 There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
-For instance, in medical diagnosis, automated disease predictions should always be corroborated by individualized findings to be communicable and thus be taken into consideration by clinicians -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
-Clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
+For instance, in medical diagnosis, automated disease predictions should always be corroborated by individualized findings to be communicable to clinicians -- in other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
+On the other hand, clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
 
@@
 This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graphs representing generative and discriminative models.
-For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence requires a larger training set to achieve optimal performance \cite{Ng-01}.
+For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence it requires more training data to achieve optimal performance \cite{Ng-01}.
 
@@
 
-This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a means to incorporate generative information into a discriminative model.
-Rather than attempting at a full generative model, one may combine multiple low-dimensional generative models for different pieces of information extracted from the input data.
-Each ``small'' model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
+This note advocates probabilistic opinion pooling \cite{Genest-86} as a means to incorporate generative information into a discriminative model.
+Rather than attempting at a full generative model, one may combine multiple low-dimensional generative models corresponding to different pieces of information extracted from the input data.
+Each such model acts as an isolated ``agent'' using a single feature to express an opinion in the form of a likelihood function of the target variable.
 The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
+
 This idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
@@
 
-With the opinion pooling interpretation in mind, the feature weights in CL may be optimized for prediction performance in a typical discriminative learning scenario.
+With the opinion pooling interpretation in mind, the weights assigned to the different features may be optimized for prediction performance in a typical discriminative learning scenario.
 As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.

### 2019-04-08 papers/composite_inference/draft.8apr19.tex
Modifs apr 8

--- 2019-04-01 papers/composite_inference/draft.tex
@@
 
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{proposition}[theorem]{Proposition}
-
 \title{Composite Bayesian inference}
@@
 \newcommand{\blambda}{{\boldsymbol{\lambda}}}
+\newcommand{\Blambda}{{\boldsymbol{\Lambda}}}
 \newcommand{\bell}{{\boldsymbol{\ell}}}
 \newcommand{\E}{\mathbb{E}}
-%\newcommand{\matphi}{\boldsymbol{\Phi}} 
-%\def\p{{\bar{\mathbf{p}}}}
-%\def\q{{\bar{\mathbf{q}}}}
 
@@
 The resulting class of predictive models, which we call ``composite Bayesian'', is a middle way between purely generative and purely discriminative models.
-This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning: in the latter case, they may be tuned so as to maximize training likelihood, yielding an easy-to-solve convex problem.
+This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning; in the latter case, they may be tuned so as to minimize training cross-entropy, yielding an easy-to-solve convex problem.
 We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
@@
 Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
+Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the conditional distribution of the target variable given the data, while a {\em generative} model describes the joint distribution of the target and the data.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
 
-Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves.
+In a determined universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions.
+Consider as a straightforward example deciding whether an object is a sauce pan or a frying pan based on depth.
+A two-class discriminative model can learn a threshold that correctly classifies most pans in practice, but will confidently classify a blender as a sauce pan, while linear discriminant analysis\footnote{Despite the name, linear discriminant analysis is based on a generative model.}, for instance, could hint that a blender is a poor match to both pan categories.
+Inability to ``confirm'' or ``justify'' their predictions makes discriminative models difficult to interpret.
+
+%Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves.
 Consider as a straightforward example the problem of deciding whether an object is a sauce pan or a frying pan based on depth.
@@
 
-There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
-For instance, in medical diagnosis, automated disease predictions should be corroborated by individualized findings to be communicable to clinicians.
-In other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
-On the other hand, clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
+There is growing awareness that AI should be interpretable in life-impacting applications \cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision-making processes, if only because decisions need to be explained and justified to other humans, while determining the set of possible outcomes may also be part of the process.
+For instance, in medical diagnosis, automated disease predictions need to be corroborated by individualized findings to be taken into account.
+In other words, there is little clinical value in ``black boxes'' and certainly more in automated methods that can justify their outputs just like human experts.
+One such method that has long been used by clinicians is the familiar reference range analysis, which stems from simple generative models.
 
@@
 \subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
-\caption{Directed graphs representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the target variable, the data, and the model parameters.
+\caption{Directed graphs representing a generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the target variable, the data, and the model parameters.
 Note the marginal independence of data and parameters in the discriminative model.}
@@
 
-This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both computationally efficient and interpretable, resulting in a class of hybrid models.
+This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both accurate and interpretable.
 The key idea is to combine multiple low-dimensional generative models corresponding to different pieces of information extracted from the input data.
-Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
-The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
-This may be understood as a probabilistic version of boosting, or as a {\em divide and conquer} approximation to the intractable Bayesian posterior.
+Each such model acts as an isolated ``agent'' that uses a single feature to express a testable opinion in the form of a likelihood function of the target variable.
+The agent opinions are then aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
+This strategy may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
 
-%Importantly, predictions made in such a way are interpretable owing to their underlying generative nature.
+%Importantly, each agent opinion is justifiable as it is based on a generative model.
 
-When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL).
-As discussed in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
-While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency.
+When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL), see \cite{Varin-11} and the references therein.
+CL was developed as a surrogate of traditional likelihood for parameter estimation.
+While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} (see Appendix~\ref{app:frequentist} for a basic weak consistency proof), and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
-With the opinion pooling interpretation in mind, the weights assigned to the different features may be optimized for prediction performance in a typical discriminative learning scenario.
+With the opinion pooling interpretation in mind, the weights assigned to the different features in~CL may be optimized for prediction performance in a typical supervised learning scenario.
 As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.
@@
 
-Technical details are given in the remainder.
-We warn the reader that slightly abusive mathematical notations are used deliberately for the sake of clarity in places where we think that they are unambiguous.
+The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on the target values, leading to a more flexible predictive model that can be trained similarly to CL.
+
+%Technical details are given in the remainder.
+
+The reader is warned that slightly abusive mathematical notation is sometimes used in the remainder.
+This is done deliberately for the sake of clarity in places where we think that notation can be lightened without raising ambiguity.
 
@@
 If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-One that is appealing as it tends to produce single peaked distributions is the (generalized) log-linear opinion pool:
+One that is appealing as it tends to produce single peaked distributions is the log-linear opinion pool\footnote{When $\pi(x)$ is not uniform, it is also called {\em generalized} log-linear pool, however this terminology could be confusing here since we will present another generalization.}:
 \begin{equation}
@@
 \end{equation} 
-where $\pi(x)$ is a prior distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
-Note that it is possible to have the weights depend on the data~$\y$, but this would make the method considerably more complex and less interpretable.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
-
-%This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|\y)\leq p_\lambda(x_2|\y)$.
-Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
-
-%Strictly positive weights guarantee the so-called 0/1 forcing property, that is, if an hypothesis~$x$ has zero likelihood according to at least one agent, then its consensus probability vanishes too.
-
-% Not clear yet as to what a negative weight could mean!
+where $\pi(x)$ is a reference distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+Note that the weights could depend on the data~$\y$, but this would make the method considerably more complex and less interpretable.
+Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less probable than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
 
@@
 \end{equation} 
-plays that of a likelihood function.
-The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
+plays that of a likelihood function, and happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
 The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}.
-The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.
 
-CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
-While the CL derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method, however unitary CL weights do not guarantee optimal prediction performance.
+The computational advantage of CL over genuine likelihood stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.
+CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} method in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
+Since log-linear opinion pooling does not assume feature independence, it yields both an alternative interpretation and a generalization of na\"ive Bayes, whereby unitary feature weights may be used by default but do not guarantee optimal prediction performance.
 
@@
 If training data is available, the generative parameters may be set in a first stage using standard estimation techniques.
-Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to composite likelihood one of the same techniques as for traditional likelihood \cite{Berger-99}.
+Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to CL one of the same techniques as for traditional likelihood \cite{Berger-99}.
 We focus in the sequel on tuning the composite weights.
@@
 
-In the absence of knowledge, uniform weights should be chosen under a rule of indifference.
+In the absence of knowledge, uniform weights may be chosen under a rule of indifference.
 CL is then a scaled version of na\"ive Bayes likelihood, the unique weight value being irrelevant to the maximum CL estimator (MCLE).
@@
 
-% in attempts to match frequentist and Bayesian notions of uncertainty.
-
 Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one.
-This is motivated by the opinion pooling perspective and a characterization theorem \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator (for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome) is the log-linear opinion pool with unit sum weights.
-The external Bayesian  property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
-The log-linear pool with unit sum weights also happens to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
+This is motivated by a characterization theorem for opinion pooling \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator (for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome) is the log-linear opinion pool with unit sum weights.
+Externally Bayesian essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
+The log-linear pool with unit sum weights also turns out to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
 
@@
 
-Unit sum weights, however, implicitly assume maximum redundancy between features.
-Consequently, they tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}), meaning reduced prediction confidence when features are weakly correlated, as is desirable in general.
+Unit sum weights, however, implicitly assume maximum redundancy between features, hence tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}).
+This means low prediction confidence (large credibility sets) when features are weakly correlated, as is desirable in general.
+Conversely, unitary weights ($\blambda\equiv 1$) as in Na\"ive Bayes assume independent features and may thus yield over-confident predictions.
 An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}.
-This suggests a learning approach if possible, which, as we shall see, is unlikely to pick uniform or unit sum weights.
+This suggests a learning approach whenever feasible.
+
+%, which, as we shall see, is unlikely to pick uniform or unit sum weights.
 
@@
 
-Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
+Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, possibly the same as for generative pre-training, the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
 \begin{equation}
@@
 $$
-\ell_i(x,\y) = \log f(z_i|x),
+\ell_i(x,\y) \equiv \log f(z_i|x),
 \qquad
@@
 
-It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence the learning can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}.
+It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence this learning can be implemented using a standard convex optimization algorithm such as limited-memory BFGS \cite{Byrd-95}.
 See Appendix~\ref{app:training} for some implementation details.
 
-Because some positive weight constraints may turn out inactive, learning has a natural tendency to produce sparse weights.
-As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
+Because some positive weight constraints may turn out inactive, optimization has a natural tendency to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs radically from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
 
@@
   \end{center}
-\caption{Distribution of optimal composite weights vs individual feature discrimination power in a binary classification task (breast cancer UCI dataset).
+\caption{Distribution of optimal composite weights vs.~individual feature discrimination power in a binary classification task (breast cancer UCI dataset).
 Discrimination is measured by the maximum KL divergence between class-conditional generative distributions.}
@@
 
-As is customary and generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$, enforcing balanced classes if the prior is uniform.
+As is customary and generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$, thus enforcing balanced classes if the prior is uniform.
 Also, some regularization may be needed for stability: we shall in practice maximize $U(\lambda)-\alpha \|\blambda\|^2$, where $\alpha>0$ is a small damping factor.
@@
 
-%D(p\|p_0) = \sum_{x\in{\cal X}}\sum_{\y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{p_0(x,\y)},
 Exponential family properties also imply that learning via likelihood maximization~(\ref{eq:train_likelihood}) is dual to an $I$-projection \cite{Csiszar-84}: 
 \begin{equation}
-\label{eq:maxent}
+\label{eq:i_proj}
 \min_{p\in{\cal P}} D(p\|p_0),
@@
 \end{equation}
-subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where $h(x,\y)$ is the joint empirical distribution of targets and data examples, $h(\y)$ is the corresponding marginal data distribution, and ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$ with ${\cal Y}\equiv\{\y_1,\ldots,\y_N\}$.
-Note that the constraints implicitly approximate the data marginal and the mean log-likelihood values by their empirical estimates, confirming the need for regularization in practice.
+subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where $\E_h$ denotes the expectation with respect to the joint {\em empirical} distribution of targets and inputs, $h(\y)$ is the corresponding empirical marginal distribution of inputs, and ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$ with ${\cal Y}\equiv\{\y_1,\ldots,\y_N\}$.
+Note that the problem statement implicitly approximates the true (unknown) joint distribution of $(x,\y)$ by its empirical estimate, hence the need for regularization in practice.
 
-The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized, hence the $I$-projection is equivalent to maximum conditional entropy \cite{BergerA-96}.
+The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized and that the $I$-projection is equivalent to maximum conditional entropy \cite{BergerA-96}.
 The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
@@
 
-An insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss.
+An insightful analysis given in \cite{Grunwald-04} shows that $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss.
 Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is inexact due to the limited training set size.
-Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by the underlying knowledge, which in this case may be interpreted in terms of maximum feature description length.
+Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by its underlying inexact knowledge, which in this case may be interpreted in terms of  feature description length.
 
@@
 \label{eq:super_pool}
-p_\lambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
+p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
 \end{equation}
-where $x_0$ is some fixed target value, which we call the reference.
-As discussed in Appendix~\ref{app:super}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting functions~$w_i(x)$ sum up to one uniformly,
+where~$x_0$ is an arbitrary fixed target value, which we call the reference, and $\blambda: {\cal X}\to \mathbb{R}_+^n$ is a weighting function here.
+As discussed in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting function sums up to one uniformly,
 $$
 \forall x\in{\cal X},\quad
-\sum_{i=1}^n w_i(x) = 1,
+\sum_{i=1}^n \lambda_i(x) = 1,
 $$
@@
 We call this form super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs.
-SCL further generalizes CL since $L^{sc}_\blambda(x)=L^c_\blambda(x)/L^c_\blambda(x_0)$ for target-independent weights $\lambda_i(x)\equiv \lambda_i$, hence it is in this case proportional to CL and yields the same predictive distribution.
+SCL further generalizes CL since both forms are obviously proportional for target-independent weights, $\lambda_i(x)\equiv \lambda_i$.
 
-Conversely, when each target assigns a weight one to a single feature and zero to all the others, SCL turns out to be identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, the ``PDF projection theorem'', which in fact boils down to an $I$-projection \cite{Minka-04}.
+%$L^{sc}_\blambda(x)=L^c_\blambda(x)/L^c_\blambda(x_0)$ for target-independent weights $\lambda_i(x)\equiv \lambda_i$, hence it is in this case proportional to CL and yields the same predictive distribution.
+
+Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, for each target, there is a single feature with non-zero weight and this weight is one), then SCL turns out to be identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, the ``PDF projection theorem'', which in fact boils down to an $I$-projection \cite{Minka-04}.
 While our derivation stems from a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
-
-%Case of a single nonzero agent per feature.
-Then SCL is equivalent to the class-specific likelihood derived in \cite{Baggenstoss-03} based on a generative model selection argument  Our derivation is much different but interesting to see that we end up with a more general form.
 
@@
 It is possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs.
-Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence on the predictive distribution.
+Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence whatsoever on the predictive distribution.
 
@@
 Then it's ok if they can model at least two targets including the reference: $\ell_i(x)$ is unknown for the other values but we can give them zero weight so it doesn't matter.
-
-If training data is available, the SCL weights may be learned by maximum likelihood as in Section~\ref{sec:learning}, yielding a similar convex problem similar.
-Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen.
+If training data is available, the SCL weights may be learned by maximum likelihood like for CL (see Section~\ref{sec:learning}), yielding a similar convex problem.
+Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen (although SCL may not be asymptotically consistent unlike CL, see Appendix~\ref{app:frequentist}).
 The $I$-projection interpretation still holds, yet with class-specific mean-value constraints:
@@
 where $\delta$ denotes the Kronecker delta.
-%$$
-%\forall x\in{\cal X}, \forall i=1,\ldots,n,
-%\quad
-%p(x) \E_{f(z_i|x)}\left[\log \frac{f(z_i|x)}{f(z_i|x_0)}\right]
-%\geq h(x) \E_{h(z_i|x)}\left[\log \frac{f(z_i|x)}{f(z_i|x_0)}\right]
-%$$
+
+%Note that
+
+%In summary, SCL is a natural generalization of CL with greater flexibility in weight assignments, hence potentially superior prediction performance if trained on a large enough dataset.
+It, however, depends on an arbitrary reference class.
+
+%Beside greater flexibility than CL, a disadvantage of SCL it that interpretation becomes relative to an arbitrary reference~$x_0$.
+In some applications, there may exist a natural reference class to cling to in common practice (for instance, normality in medical diagnosis), but this is not always the case.
 
@@
 
-Beside greater flexibility than CL, a disadvantage of SCL it that interpretation becomes relative to an arbitrary reference~$x_0$.
-In some applications, there may exist a natural reference class to cling to in common practice (for instance, normality in medical diagnosis), but this is not always the case.
+The composite Bayesian framework presented here encompasses and unifies several seemingly unrelated methods available in computational statistics and machine learning: composite likelihood \cite{Varin-11}, na\"ive Bayes, and the class-specific method \cite{Baggenstoss-03}.
+It provides a rich class of trainable predictive models that work on pre-determined data features, and is therefore suitable for shallow learning tasks (including transfer learning), where it competes with popular methods such as logistic regression, support vector machines or random forests.
 
-It is easy to build a good classifier.
-You just need to pick a parametric family that is big enough to produce a good fit yet small enough to avoid overfitting.
-Less easy is to interpret what the classifier does.
-Deep learning is cool but it is relevant to AI only if the representations it learns are generic enough to be used in other problems, i.e.
-it should ultimately be validated via transfer learning.
+By construction, composite Bayesian models are discriminative but, contrary to the above mentionned methods, they are based on feature generation models.
+This makes it possible to test how consistent the data at hand is with each feature-induced ``opinion'', hence validating the overall model prediction.
+Moreover, since opinions are aggregated in an additive way, it is easy to determine which features contribute most to a particular comparison of target values, and thus to explain the rational for the prediction by highlighting a few decisive features.
+
+[EXAMPLE].
+
+We believe that there are many applications of statistical inference in which interpretation is the real goal and prediction performance is only an intermediate requisite.
+A prediction method is only relevant if it proves reliable; but, once proven reliable, it may need to justifiy its predictions.
+The {\em why} question may be more important than the {\em what} question in life-impacting applications of AI.
+
+% (we want to listen to reliable experts, but experts have to be able to argue their case).
+
+Previous Bayesian-like inference attempts at CL.
+\cite{Wang-14}: generative model selection approach to CL, \cite{Pauli-11,Ribatet-12} calibrate pseudo-posterior using asymptotic theory.
+Our approach is explicitly taylored to prediction.
+We don't tortille du cul.
+
+Gaussian log-likelihood:
+$$
+\log p(x)
+= - \frac{1}{2} \log (2\pi\sigma^2) - \frac{(x-m)^2}{2\sigma^2}
+= - \frac{1}{2} [\log (2\pi\sigma^2) + \chi_1^2]
+$$
+
+LL can be low for two reasons: either the data is implausible under the model (large chi-square), or the model is intrinsically random (high variance).
+
+It could be that a model is more likely than another just because it has smaller variance despite larger chi-square.
+
+When we interpret composite likelihood, the following questions need to be answered sequentially: 1) what are the main features driving the prediction?
+2) for these features, is the data consistent with the prediction?
+
+1) is easy from the additive nature of CL.
+2) requires evaluating chi-square stat for each feature (better said Mahalanobis distance)
+
+\begin{enumerate}
+\item Scope.
+
+How to build predictive model from a set of features.
+
+We assume known features, so this is good for shallow learning, including transfer learning.
+Still need representation learning methods.
+Needs features.
+Not a representation learning method, but could be coupled why not.
+
+\item Contribution.
+
+L'id\'ee forte c'est la modelisation semi-generative.
+Ni purement generatif, ni purement discriminant.
+On peut ainsi creer des systemes capables de justifier leurs inferences (introspection?).
+Un modele purement discriminant ne le permet pas.
+Un modele purement generatif le permet mais s'avere souvent trop couteux.
+On n'a pas besoin de simuler des donnees completes...
 
@@
 
-{\color{red} We cannot rank features according to their weights.
-A feature might incidentally get a large weight because its likelihood is flat.
-Feature relevance also depends on conditional likelihood distribution.
-Consider mutually independent features: each one recieves a unit weight, but some may be independent from target.}
+\item Cette approche est naturellement mise en oeuvre par aggregation d'opinions, ce qui nous mene tout droit a CL et a des generalisations (SCL).
+
+SCL has the potential to achieve higher prediction performance than~CL if trained on a large enough dataset.
+
+\item {\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
+
+Well, it's a bit unfair...
+CBI is a way of boosting naive Bayes prediction performance almost at the level of the best discriminative methods, while keeping interpretation alive.
+We have even more flexibility to do this with SCL.
+And SCL can also work with incomplete likelihoods.
+
+\item Previous work.
+
+Moins ambitieux qu'une approche selection de modele generative \cite{Wang-14,Baggenstoss-03,Yuan-99b}.
+En quoi est-ce different du produit d'experts \cite{Hinton-02}?
+Plus simple, direct et efficace que les approches bayesiano-asymptotiques a la con.
+
+\item Interpretation de CL: facile du fait que la difference de log-probabilites entre deux cibles est la somme de contributions specifiques aux attributs.
+On peut donc non seulement determiner quels attributs contribuent le plus a une difference mais aussi expliquer chaque contribution en comparant la valeur de l'attribut aux distributions respectives.
+Pour SCL, c'est un peu plus complique car il faut faire intervenir une troisieme distribution, celle de la reference, dans l'equation.
+GIVE EXAMPLE.
+
+\item En pratique, double apprentissage.
 
@@
 
-{\color{red} Semi-generative method. Good for shallow learning or transfer learning. Still need representation learning methods.}
+{\color{red}Two-step training: generative phase then discriminative phase. Why is it cool?}
 
-{\color{red} Alternative to logistic regression in deep learning? Would require a game-theoretic learning approach... future work.}
-
-{\color{red} Composite Bayes is invariant by affine transformation of the data (unlike multinomial LR, which is problematic when damping is applied).}
-
-{\color{red} Is damping important? (it guarantees training cost is stricly convex).}
-
-{\color{red} Can we interpret maximum weight likelihood as a ``maximum relevance minimum redudancy'' feature selection approach?}
-
-Is it new to see composite likelihood as opinion pooling?
-Yes it is.
-Some previous attempts were made to derive composite likelihood as some sort of consensus generative model \cite{Wang-14}.
-Also some previsous attemps at Bayesian inference from CL using asymptotic theory.
-
-%In other words, is $\bar{a}(\blambda)$ a measure of feature redundancy?
-At fixed $\y$, if one $\lambda_i\to\infty$, then $a(\blambda,\y)\ot\infty$ except if $\ell_i(x,\y)\preceq 0$, in which case it goes to $-\infty$.
-If using likelihood ratios, this only happens if the reference class is the most likely for $\y$.
-But then we can prove that it can't happen for all $\y$ (otherwise the associated moment would be negative), so basically $\bar{a}(\blambda)$ blows up when at least one weight gets large.
-And if several weights get large simultaneously, then it blows up even more so we might consider it a redundancy measure to some extent...
-
-Some experimental observations.
-In the iris dataset (historical data used by Fisher to demonstrate linear discriminant analysis), composite Bayes finds that petal length is the most relevant feature followed by petal width, while sepal length and width get zero weight.
-In the wine dataset, the flavanoids is the most relevant feature for standard composite Bayes, but it's color intensity closely followed by flavanoids for the super composite version.
-In the breast cancer, both approaches are equivalent since it is a binary problem, weights are very sparse: only 13/30 are non-zero with worst area the most relevant.
-In the digits prolem, the most relevant features tend to be in the image center; composite Bayes can achieve similar performance to multinomial logistic regression with about 10 times less parameters.
-
-Number of independent discriminative parameters in logistic regression: $(K-1)(F+1)$.
-In standard composite inference: $F$ (or $K-1+F$ with offsets).
-Difference $=(K-2)(F+1)+1$ (or $=(K-2)F$).
-In super-composite inference: $(K-1)(F+1)$ (if reference is a class).
-Number of generative parameters in composite inference using univariate Gaussian models: $2K$ with heteroscedasticity, $K+1$ with homoscedasticity.
-
-Intuition behind weights: notion of global feature relevance.
-But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$ for a particular instance?
-No.
-But we can assess the influence of each scaled likelihood in order to rank features according to their influence on the decision.
-
-CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
-We have shown a deep connection between CL and probabilistic inference, thereby establishing CL as a class of discriminative models.
-Because CL is built from a set of marginal feature-specific generative distributions, it is in essence a two-step semi-generative, semi-discriminative learning approach.
-In the first, ``generative'' phase, the feature distributions are learned; in the second, ``discriminative'' phase, the feature weights are learned.
-This strategy can be thought of as a form of non-adaptive boosting.
-
-%The first phase corresponds to the training of ``weak learners''. The second phase amounts to a form of boosting.
-
-A purely discriminative learning could be used instead but...
-Why potentially more efficient than pure discriminative training: because generative training is always more efficient than discriminative training if models are comparable.
-So the key point is that ``we have less parameters in the discriminant phase".
-Good in small datasets.
-But also in asymptotic regime if the features are weakly or highly correlated (???).
-
-Logistic regression / naive Bayes example.
-Under homoscedasticity (assuming that each feature has class-independent variance), CBI is equivalent to logistic regression -- because the predictive distribution family is the same.
-This is a case where BCI brings nothing.
-But consider heteroscedasticity, then BCI yields a quadratic classifier.
-Compared to a fully discriminative model, the number of parameters to learned in the discriminative phase is reduced by half.
+\item Unsupervised learning
 
@@
 
-Needs features. Not a representation learning method, but could be coupled why not.
+\item {\color{red} Alternative to logistic regression in deep learning? Would require a game-theoretic learning approach... future work.} But why should we do that? C'est un terrain Bernard min\'e...
+
+\end{enumerate}
+
+%Beside greater flexibility than CL, a disadvantage of SCL it that interpretation becomes relative to an arbitrary reference~$x_0$.
+In some applications, there may exist a natural reference class to cling to in common practice (for instance, normality in medical diagnosis), but this is not always the case.
+
+%It is easy to build a good classifier.
+You just need to pick a parametric family that is big enough to produce a good fit and small enough to avoid overfitting.
+Less easy is to interpret what the classifier does.
+Deep learning is cool but it is relevant to AI only if the representations it learns are generic enough to be used in other problems, i.e.
+it should ultimately be validated via transfer learning.
+
+%{\color{red} We cannot rank features according to their weights.
+A feature might incidentally get a large weight because its likelihood is flat.
+Feature relevance also depends on conditional likelihood distribution.
+Consider mutually independent features: each one recieves a unit weight, but some may be independent from target.}
+
+%{\color{red} Composite Bayes is invariant by affine transformation of the data (unlike multinomial LR, which is problematic when damping is applied).}
+
+%{\color{red} Is damping important? (it guarantees training cost is stricly convex).}
+
+%{\color{red} Can we interpret maximum weight likelihood as a ``maximum relevance minimum redudancy'' feature selection approach?}
+
+%Is it new to see composite likelihood as opinion pooling?
+Yes it is.
+Some previous attempts were made to derive composite likelihood as some sort of consensus generative model \cite{Wang-14}.
+Also some previsous attemps at Bayesian inference from CL using asymptotic theory.
+
+%In other words, is $\bar{a}(\blambda)$ a measure of feature redundancy?
+At fixed $\y$, if one $\lambda_i\to\infty$, then $a(\blambda,\y)\ot\infty$ except if $\ell_i(x,\y)\preceq 0$, in which case it goes to $-\infty$.
+If using likelihood ratios, this only happens if the reference class is the most likely for $\y$.
+But then we can prove that it can't happen for all $\y$ (otherwise the associated moment would be negative), so basically $\bar{a}(\blambda)$ blows up when at least one weight gets large.
+And if several weights get large simultaneously, then it blows up even more so we might consider it a redundancy measure to some extent...
+
+%Some experimental observations.
+In the iris dataset (historical data used by Fisher to demonstrate linear discriminant analysis), composite Bayes finds that petal length is the most relevant feature followed by petal width, while sepal length and width get zero weight.
+In the wine dataset, the flavanoids is the most relevant feature for standard composite Bayes, but it's color intensity closely followed by flavanoids for the super composite version.
+In the breast cancer, both approaches are equivalent since it is a binary problem, weights are very sparse: only 13/30 are non-zero with worst area the most relevant.
+In the digits prolem, the most relevant features tend to be in the image center; composite Bayes can achieve similar performance to multinomial logistic regression with about 10 times less parameters.
+
+%Number of independent discriminative parameters in logistic regression: $(K-1)(F+1)$.
+In standard composite inference: $F$ (or $K-1+F$ with offsets).
+Difference $=(K-2)(F+1)+1$ (or $=(K-2)F$).
+In super-composite inference: $(K-1)(F+1)$ (if reference is a class).
+Number of generative parameters in composite inference using univariate Gaussian models: $2K$ with heteroscedasticity, $K+1$ with homoscedasticity.
+
+%Intuition behind weights: notion of global feature relevance.
+But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$ for a particular instance?
+No.
+But we can assess the influence of each scaled likelihood in order to rank features according to their influence on the decision.
+
+%CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
+We have shown a deep connection between CL and probabilistic inference, thereby establishing CL as a class of discriminative models.
+Because CL is built from a set of marginal feature-specific generative distributions, it is in essence a two-step semi-generative, semi-discriminative learning approach.
+In the first, ``generative'' phase, the feature distributions are learned; in the second, ``discriminative'' phase, the feature weights are learned.
+This strategy can be thought of as a form of non-adaptive boosting.
+
+%The first phase corresponds to the training of ``weak learners''. The second phase amounts to a form of boosting.
+
+%A purely discriminative learning could be used instead but...
+Why potentially more efficient than pure discriminative training: because generative training is always more efficient than discriminative training if models are comparable.
+So the key point is that ``we have less parameters in the discriminant phase".
+Good in small datasets.
+But also in asymptotic regime if the features are weakly or highly correlated (???).
+
+%Logistic regression / naive Bayes example.
+Under homoscedasticity (assuming that each feature has class-independent variance), CBI is equivalent to logistic regression -- because the predictive distribution family is the same.
+This is a case where BCI brings nothing.
+But consider heteroscedasticity, then BCI yields a quadratic classifier.
+Compared to a fully discriminative model, the number of parameters to learned in the discriminative phase is reduced by half.
 
@@
 
-{\color{red}Product of fucking experts \cite{Hinton-02}.
+%{\color{red}Product of fucking experts \cite{Hinton-02}.
 Log-linear pool to build a generative model (assuming in fact independence between experts).
@@
 
-{\color{red}Two-step training: generative phase then discriminative phase. Why is it cool?}
-
-{\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
-
-Future work: can we collapse generative and discriminative training into a single step? Maybe using the concept of {\em minimally informative likelihood} \cite{Yuan-99b}.
+%Future work: can we collapse generative and discriminative training into a single step? Maybe using the concept of {\em minimally informative likelihood} \cite{Yuan-99b}.
 
@@
 $$
-where the expectation is taken with respect to the true distribution $p(\y|x_\star)$. This fact follows from two basic properties of KL~divergence: positivity and partition inequality.
+where the expectation is taken with respect to the true distribution $p(\y|x_\star)$ at fixed~$x_\star$.
+This fact follows from two basic properties of KL~divergence: positivity and partition inequality.
 
@@
 \end{equation}
-with $s_\blambda\equiv \|\blambda\|_1 =\sum_i \lambda_i$, and $L(x)$ the true likelihood function, {\em i.e.} $L(x)\equiv \log p(\y|x)$.
+with $s_\blambda\equiv \|\blambda\|_1 =\sum_i \lambda_i$, and $L(x)\equiv \log p(\y|x)$ is the true likelihood function.
 
@@
 \begin{itemize}
-\item {\em Consistency.} The expected composite log-likelihood is maximized by~$x_\star$.
-\item {\em Conservativeness for unit sum weights.} If the weights sum up to one or less ($s_\blambda\leq 1$), composite likelihood ratios of the true target~$x_\star$ vs.~other target values~$x$ tend to be underestimated, {\em i.e.} are smaller on the average than corresponding true likelihood ratios.
+\item {\em Weak consistency.} The expected composite log-likelihood is maximized by~$x_\star$.
+\item {\em Conservativeness.} If the weights sum up to one or less ($s_\blambda\leq 1$), composite likelihood ratios of the true target~$x_\star$ vs.~other target values~$x$ tend to  underestimate true likelihood ratios.
 \end{itemize}
 
-What about SCL???
+Note that the derivation assumes that the weights~$\blambda$ are independent from the target~$x$, therefore these properties do not necessarily extend to SCL, for which weights are target-dependent.
 
@@
 
-As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $z^c_i(\y)$ considered as a predictor of the ``dependent'' feature, $z_i(\y)$, yielding the more general form:
+As a straightforward  extension of marginal CL, each feature-based likelihood may be conditioned by an additional ``independent'' feature $\nu_i(\y)$ considered as a predictor of the ``dependent'' feature, $z_i(\y)$, yielding the more general form:
 \begin{equation}
 \label{eq:cond_feat_lik}
-\ell_i(x) = f(z_i|x,z^c_i).
+\ell_i(x) = f(z_i|x,\nu_i).
 \end{equation}
 
-Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $x$, but can provide relevant information when considered jointly with $x$, as in the case of regression covariates, for instance.
-Equation~(\ref{eq:comp_lik}) then amounts to conditional CL \cite{Varin-11}, a more general form of CL also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74} developed for image segmentation.
+Conditioning may be useful if it is believed that $\nu_i$ alone provides little or no information about~$x$, but is informative when considered jointly with~$z_i$, as in the case of regression covariates, for instance.
+Equation~(\ref{eq:comp_lik}) then amounts to conditional CL \cite{Varin-11}, a more general form of CL which also includes Besag's historical {\em pseudo-likelihood} \cite{Besag-74} developed for image segmentation.
 
@@
 
-Note that $\psi(\blambda)$ is nothing but the dual function associated with the maxent problem~(\ref{eq:maxent}).
+Note that $\psi(\blambda)$ is nothing but the dual function associated with the $I$-projection problem~(\ref{eq:i_proj}). The derivatives of~$\psi(\blambda)$ are found to be:
+\begin{eqnarray*}
+\nabla\psi(\blambda)
+ & = & \bar{\bell} - \frac{1}{N} \sum_k \nabla a(\blambda,\y_k), \\
+\nabla\nabla^\top\psi(\blambda)
+ & = & - \frac{1}{N} \sum_k \nabla \nabla^\top a(\blambda,\y_k),
+\end{eqnarray*}
+with:
+$$
+\nabla a(\blambda,\y) = \E_{\blambda}(\bell),
+\qquad
+\nabla \nabla^\top a(\blambda,\y) = {\rm Var}_{\blambda}(\bell),
+$$
+where $\E_{\blambda}$ and ${\rm Var}_{\blambda}$ respectively denote the expectation and variance with respect to $p_\blambda(x|\y)\propto \pi(x)\exp[\blambda^\top \bell(x,\y)]$ at fixed~$\y$.
+This shows that $a(\blambda,\y)$ is convex in $\blambda$ since the variance matrix is positive, which, in turn, proves the concavity of~$\psi$.
 
-Derivatives...
+\section{Heterogeneous log-linear pooling}
+\label{app:hetero}
+
+Consider a sequence of probability distributions $p_i(x)$, $i=1,\ldots,n$, representing multiple opinions on a variable $x\in{\cal X}$.
+The general goal of opinion aggregation is to define an operator $F(p_1,\ldots,p_n)$ that takes opinions as inputs and outputs a single distribution $p(x)$ representing a consensus.
+
+Log-linear pooling \cite{Genest-86} is one particular such operator:
 $$
-\nabla\psi(\blambda)
-= \bar{\bell} - \frac{1}{N} \sum_k \nabla a(\blambda,\y_k) 
+p(x)\propto \pi(x) \prod_{i=1}^n p_i(x)^{\lambda_i},
+$$
+where $\pi(x)$ is an arbitrary distribution and $\lambda_i\in\mathbb{R}$ are arbitrary weights.
+It is easily seen that, if the weights sum up to one, log-linear pooling is externally Bayesian in the sense that, for any nonvanishing positive valued function $q:{\cal X}\to \mathbb{R}^\star_+$, opinion pooling and opinion updating commute:
+$$
+F[N(q p_1), \ldots, N(q p_n)]
+=
+N[q F(p_1,\ldots, p_n)],
+$$
+where~$N$ denotes the normalization endomorphism of $\mathbb{R}_+^{\cal X}$:
+$$
+N(f)(x) = \frac{f(x)}{\sum_{x'\in{\cal X}} f(x')}
+.
 $$
 
+Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$, is of the above form with unit sum weights and is therefore a log-linear opinion pool.
+
+A question that arises naturally is whether it is possible to weight opinions depending on targets, so that $\lambda_i(x)$ becomes a function of~$x$.
+This would make it possible, for instance, to consider an opinion reliable when it is about a certain hypothesis~$x_1$, but unreliable when it comes to another hypothesis~$x_2$.
+As a first try, let us consider a straightforward extension of the log-linear pool:
 $$
-\nabla\nabla^\top\psi(\blambda)
-= - \frac{1}{N} \sum_k \nabla \nabla^\top a(\blambda,\y_k)  
+p(x)\propto \prod_i p_i(x)^{\lambda_i(x)},
+$$
+where $\lambda_i : {\cal X}\to\mathbb{R}$ is some weighting function. Unfortunately, this operator is generally not externally Bayesian because $F[N(q p_1), \ldots, N(q p_n)](x)
+\propto r(x) F(p_1,\ldots, p_n)(x)$ with:
+$$
+r(x) = q(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)},
+\qquad 
+z_i = \sum_{x\in{\cal X}} q(x) p_i(x),
+$$
+and $r(x)$ cannot be made proportional to $q(x)$, except with constant weights, because the normalizing factors~$z_i$ may differ from one agent to the other.
+
+Consider, however, the following alternative:
+\begin{equation}
+\label{eq:hetero_log_pool}
+p(x) \propto \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)},    
+\end{equation}
+where $x_0$ is some arbitrary reference value.
+This does the trick because, for any positive function $q(x)$, the ratio $q(x)p_i(x)/q(x_0)p_i(x_0)$ is left unchanged by normalization of $q p_i$ given that the normalizing factors are the same at both the numerator and the numerator, hence they cancel out, yielding:
+$$
+F[N(q p_1), \ldots, N(q p_n)](x)
+\propto
+\left[\frac{q(x)}{q(x_0)}\right]^{\sum_i \lambda_i(x)}
+F(p_1, \ldots, p_n)(x),
+$$
+which shows that~(\ref{eq:hetero_log_pool}) is externally Bayesian whenever the weighting function verifies:
+$$
+\forall x \in {\cal X},
+\qquad
+\sum_i \lambda_i(x) = 1.
 $$
 
-We have:
-$$
-\nabla a(\blambda,\y) = \E_{\blambda}(\bell|\y),
-\qquad
-\nabla \nabla^\top a(\blambda,\y) = \text{Var}_{\blambda}(\bell|\y)
-$$
-proving that $a(\blambda,\y)$ is convex in $\blambda$ for any $\y$, and in turn proves the concavity of $\psi$.
+The operator~(\ref{eq:hetero_log_pool}) clearly generalizes standard log-linear pooling and has the property that the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
+The ``trick'' to enable heterogeneous (target-dependent) weights is essentially to combine the odds relative to a fixed reference.
 
-To see this, let:
-$$
-z(\blambda,\y) 
-\equiv e^{a(\lambda,\y)}  
-= \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}
-$$
+%Let us stress that this result does not contradict the above mentionned characterization theorem \cite{Genest-86b}, which restricts itself to operators for which the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ while, in the heterogeneous log-linear pool, $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
+Note that \cite{Genest-86b} also provides the general characterization of externally Bayesian operators, of which the heterogeneous log-linear pool is just a special case slightly more general than the log-linear pool.
 
-We have:
-$$
-\nabla a = \frac{\nabla z}{z}
-$$
-
-$$
-\nabla\nabla^\top a 
-= \left(
-\frac{\nabla \nabla^\top z}{z} 
-- \frac{\nabla z \nabla z^\top}{z^2}
-\right)
-$$
-
-$$
-\nabla z 
-=
-\sum_{x\in{\cal X}} \pi(x) \bell(x,\y)  e^{\blambda^\top \bell(x,\y)}
-$$
-
-$$
-\nabla \nabla^\top z 
-=
-\sum_{x\in{\cal X}} \pi(x) \bell(x,\y) \bell(x,\y)^\top e^{\blambda^\top \bell(x,\y)}
-$$
-
-\section{Super log-linear pool}
-\label{app:super}
-
-Opinions: $p_i(x)$.
-
-Log-linear pool:
-$$
-p(x)\propto \prod_i p_i(x)^{\lambda_i}
-$$
-
-What about having $\lambda_i(x)$ depend on the target?
-
-Can we do: $p(x)\propto \prod_i p_i(x)^{\lambda_i(x)}$? Is it externally Bayesian?
-
-Let the opinions be multiplied by some function $f(x)$. The new opinions read: $q_i(x)=f(x)p_i(x)/z_i$ with $z_i=\int f p_i$. So the pool gives:
-$$
-q(x) \propto p(x) f(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)}
-$$
-
-So, it's not EB because $z_i$ may differ from one expert to the other.
-
-It feels more reasonable to do:
-$$
-p(x) \propto \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)}
-$$
-
-Is it EB? We have:
-$$
-\frac{q_i(x)}{q_i(x_0)} 
-= \frac{f(x)p_i(x)}{f(x_0)p_i(x_0)}
-$$
-because the $z_i$'s cancel out. Therefore,
-$$
-q(x) \propto p(x) \left[\frac{f(x)}{f(x_0)}\right]^{\sum_i \lambda_i(x)}
-$$
-
-So it's EB if the unit sum weight constraints are satisfied.
-
-\section{Notes Genest}
-
-External Bayesian: pooling and updating commute.
-A pooling operator is compliant with the `likelihood principle' if the pool at $x$ depends on the opinions only through the probabilities $p_i(x)$.
-
-Characterizaton of externally Bayesian pooling operators (EBPOs).
-
-Any EBPO LP-compliant that does not depend on $x$ is a log-pool with unit sum weights:
-$$
-p(x) \propto \prod_i p_i(x)^{w_i}
-$$
-
-Any EBPO LP-compliant is a generalized log-pool with unit sum weights:
-$$
-p(x) \propto \pi(x) \prod_i p_i(x)^{w_i}
-$$
-
-Weights can be negative if domain of $x$ is finite.
-
-To characterize more general EBPOs (not LP-compliant), they introduce an equivalence class.
-Two sets of opinions are equivalent if any ratio of two experts is the same up to a constant factor.
-It means that $q\sim p$ if there exists $f(x)$ such that
-$q_i(x) \propto f(x) p_i(x)$, $\forall i$.
-
-Define an arbitrary member $p_{a1},p_{a2},\ldots$ in each equivalence class.
-
-Any EBPO is of the form:
-$$
-T(p_1,p_2,\ldots) \propto \pi_a(x) v_a(x) \frac{p_1(x)}{p_{a1}(x)},
-\qquad
-v_a \geq \max (p_{a1},p_{a2},\ldots)
-$$
+%It may seem at first sight that this result contradicts the characterization theorem \cite{Genest-86b} since~(\ref{eq:hetero_log_pool}) is clearly more general than log-linear pooling.
+There is no contradiction because the said theorem restricts itself to operators for which the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ while, in the heterogeneous log-linear pool, $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
 
@@
 
+We note that $\log r$ is a second-order polynomial of~$x$:
+$$
+f(x) = a x^2 + b x + c,
+$$
+with $a=(\rho^2-1)/2\sigma_0^2$ and $b=(\mu_1-\rho^2\mu_0)/\sigma_0^2$.
+
+Assuming $X\sim N(\mu_1,\sigma_1)$, and given an observation $x$, we would like to compute the probability:
+\begin{eqnarray*}
+p 
+& = & P[f(X)>f(x)] \\
+& = & P\{[f(X)-f(x)]>0\}\\
+& = & P\{[a(X^2-x^2)+b(X-x)]>0\}\\
+& = & P\{(X-x)[a(X+x)+b]>0)\}\\
+& = & P[(X-x)(X-y)>0]\\
+& = & 1 - P[X\in(x,y)]\\
+& = & 1 - |G(x)-G(y)|
+\end{eqnarray*}
+where $y=-x-b/a$, and $G=G_{\mu_1,\sigma_1}$ is the Gaussian partition function. In this case, we have:
+$$
+\frac{b}{a} = 2 \frac{\mu_1-\rho^2\mu_0}{\rho^2-1}
+$$
+
+Special case if $\rho=0$, then $a=0$ and $b=\mu_1-\mu_0$, we can't introduce $y$. Going backward, we see that:
+$$
+p = P[(X-x)b>0]
+$$
+
+If $\mu_1>\mu_0$, the answer is $p=1-G(x)$. If $\mu_1<\mu_0$, it is $G(x)$. If $\mu_1=\mu_0$, it is just zero.
+
+So, the story is quite easy to tell.
+Say that we want to compare two hypotheses $x_0$ and $x_1$, one of which is assumed to hold.
+It is easy using CL to isolate the features that produce the most significant contribution to the log-CL difference of $x_1$ and $x_0$.
+Assume that $x_1$ is the one that yields the largest predictive probability, so we are interested in positive contributions (trying to ask: why does the system think $x_1$ is more probable?).
+
+For any such feature, we may ask if such a large contribution is expected under $x_1$ in order to confirm the prediction.
+We then compute the above probability $p$.
+If $p\approx 0$ or $p\approx 1$, the contribution is abnormal, respectively too large or too small.
+This raises a flag, indicating that $x$ may NOT be distributed according to $x_1$.
+And, since $x_1$ is more probable than $x_0$, neither hypothesis holds in fact.
+It could be a distribution intermediate between $x_0$ and $x_1$ (if $p\approx 1$), or farther away from $x_0$ than $x_1$ (if $p\approx 0$).
+
+Note that the $p$ is insensitive to the feature weight in CL...
+The weights do however play a role in identifying the features relevant to a particular comparison.
+If a feature is not relevant, it does not matter if
+
 \bibliographystyle{abbrv}

### 2019-04-10 papers/composite_inference/draft.tex
Modifs 9 apr 2019

@@
 We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
-The resulting class of predictive models, which we call ``composite Bayesian'', is a middle way between purely generative and purely discriminative models.
-This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning: in the latter case, they may be tuned so as to maximize training likelihood, yielding an easy-to-solve convex problem.
+The resulting class of predictive models, which we call composite Bayesian models, is a middle way between purely generative and purely discriminative models.
+This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning; in the latter case, they may be tuned so as to minimize training cross-entropy, yielding an easy-to-solve convex problem.
 We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
@@
 Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
+Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the conditional distribution of the target variable given the data, while a {\em generative} model describes the joint distribution of the target and the data.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
 
-Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves.
+In a closed universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions.
+Consider as a straightforward example deciding whether an object is a sauce pan or a frying pan based on its depth.
+A two-class discriminative model can learn a threshold that correctly classifies most pans in practice, but will confidently classify a blender as a sauce pan, while linear discriminant analysis\footnote{Despite the name, linear discriminant analysis is based on a generative model.}, for instance, could hint that a blender is a poor match to both pan categories.
+Inability to ``confirm'' or ``justify'' predictions makes discriminative models difficult to interpret.
+
+%Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves.
 Consider as a straightforward example the problem of deciding whether an object is a sauce pan or a frying pan based on depth.
@@
 
-There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes.
-For instance, in medical diagnosis, automated disease predictions should be corroborated by individualized findings to be communicable to clinicians.
-In other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts.
-On the other hand, clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.
+There is growing awareness that AI should be interpretable in life-impacting applications \cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision-making processes, if only because decisions need to be explained and justified to other humans, while determining the set of possible outcomes may also be part of the process.
+For instance, in medical diagnosis, automated disease predictions need to be corroborated by individualized findings to be taken into account.
+In other words, there is little clinical value in ``black boxes'', but certainly more in automated methods that can justify their outputs just like human experts.
+One such method that has long been used by clinicians is the familiar reference range analysis, which stems from simple generative models.
 
@@
 
-This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both computationally efficient and interpretable, resulting in a class of hybrid models.
+This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both accurate and interpretable.
 The key idea is to combine multiple low-dimensional generative models corresponding to different pieces of information extracted from the input data.
-Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable.
-The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
-This may be understood as a probabilistic version of boosting, or as a {\em divide and conquer} approximation to the intractable Bayesian posterior.
+Each such model acts as an isolated ``agent'' that uses a single feature to express a testable opinion in the form of a likelihood function of the target variable.
+The agent opinions are then aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
+This strategy may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
 
-%Importantly, predictions made in such a way are interpretable owing to their underlying generative nature.
+%Importantly, each agent opinion is justifiable as it is based on a generative model.
 
-When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL).
-As discussed in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation.
-While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency.
+When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL), see \cite{Varin-11} and the references therein.
+CL was developed as a surrogate of traditional likelihood for parameter estimation.
+While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} (see Appendix~\ref{app:frequentist} for a basic weak consistency proof), and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
-With the opinion pooling interpretation in mind, the weights assigned to the different features may be optimized for prediction performance in a typical discriminative learning scenario.
+Based on the opinion pooling interpretation, the weights assigned to the different features in~CL may be optimized for prediction performance in a typical supervised learning scenario.
 As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.
-The likelihood functions themselves may need to be pre-trained, therefore the composite inference approach typically involves a double training scheme: a generative training stage (to learn the feature-based likelihood parameters) followed by a discriminative training stage (to learn the feature weights in aggregation).
+The likelihood functions themselves may need to be pre-trained, therefore the composite Bayesian approach typically involves a two-stage training scheme: a generative training (to learn the feature-based likelihood parameters) followed by a discriminative training (to learn the feature weights in aggregation).
 
-Technical details are given in the remainder.
-We warn the reader that slightly abusive mathematical notations are used deliberately for the sake of clarity in places where we think that they are unambiguous.
+The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on the target values, leading to a more flexible predictive model that can be trained similarly to CL.
+
+%Technical details are given in the remainder.
+
+The reader is warned that slightly abusive mathematical notation is sometimes used in the remainder.
+This is done deliberately for the sake of clarity in places where we think that notation can be lightened without raising ambiguity.
 
@@
 If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-One that is appealing as it tends to produce single peaked distributions is the (generalized) log-linear opinion pool:
+A common choice, owing to its simplicity and tendency to produce single peaked distributions, is log-linear pooling\footnote{When $\pi(x)$ is not uniform, it is also called {\em generalized} log-linear pooling, however this terminology could be confusing here since we will present another generalization.}:
 \begin{equation}
@@
 \end{equation} 
-where $\pi(x)$ is a prior distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
-Note that it is possible to have the weights depend on the data~$\y$, but this would make the method considerably more complex and less interpretable.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
-
-%This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded.
-Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|\y)\leq p_\lambda(x_2|\y)$.
-Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.
-
-%Strictly positive weights guarantee the so-called 0/1 forcing property, that is, if an hypothesis~$x$ has zero likelihood according to at least one agent, then its consensus probability vanishes too.
-
-% Not clear yet as to what a negative weight could mean!
+where $\pi(x)$ is a reference distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+We shall asume for feasibility that the weights are intrinsic to the agents and are therefore independent from the data~$\y$.
+Also, negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less probable than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
 
@@
 
-The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule as it yields the form: 
+Log-linear pooling~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule as it yields the form: 
 $$
@@
 \end{equation} 
-plays that of a likelihood function.
-The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
+plays that of a likelihood function, and happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
 The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}.
-The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.
 
-CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
-While the CL derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method, however unitary CL weights do not guarantee optimal prediction performance.
+The clear computational advantage of CL over genuine likelihood is that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.
+CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} method in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
+Since log-linear opinion pooling does not assume feature independence, it yields both an alternative interpretation and a generalization of na\"ive Bayes, whereby unitary feature weights may be used by default but do not guarantee optimal prediction performance.
 
@@
 
-% in attempts to match frequentist and Bayesian notions of uncertainty.
-
 Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one.
-This is motivated by the opinion pooling perspective and a characterization theorem \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator (for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome) is the log-linear opinion pool with unit sum weights.
-The external Bayesian  property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
-The log-linear pool with unit sum weights also happens to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
+This is motivated by a characterization theorem for opinion pooling \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome, is the log-linear opinion pool with unit sum weights.
+Externally Bayesian essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
+Log-linear pooling with unit sum weights also turns out to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
 
@@
 
-Unit sum weights, however, implicitly assume maximum redundancy between features.
-Consequently, they tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}), meaning reduced prediction confidence when features are weakly correlated, as is desirable in general.
+Unit sum weights, however, implicitly assume maximum redundancy between features, hence tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}).
+This means low prediction confidence (over-estimated credibility sets) when features are weakly correlated, as is desirable in general.
+Conversely, unitary weights ($\blambda\equiv 1$) as in Na\"ive Bayes assume independent features and may thus yield over-confident predictions.
 An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}.
-This suggests a learning approach if possible, which, as we shall see, is unlikely to pick uniform or unit sum weights.
+This suggests a learning approach whenever feasible.
+
+%, which, as we shall see, is unlikely to pick uniform or unit sum weights.
 
@@
 
-Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
+Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$ (possibly the same as for generative pre-training), the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
 \begin{equation}
@@
 
-It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence the learning can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}.
+It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence this learning can be implemented using a standard convex optimization algorithm such as limited-memory BFGS \cite{Byrd-95}.
 See Appendix~\ref{app:training} for some implementation details.
 
-Because some positive weight constraints may turn out inactive, learning has a natural tendency to produce sparse weights.
-As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
+Because some positive weight constraints may turn out inactive, optimization has a natural tendency to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer diagnostic UCI dataset \cite{Wolberg-94}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs radically from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
+
+%\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}
 
@@
 \label{eq:super_pool}
-p_\lambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
+p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
 \end{equation}
-where $x_0$ is some fixed target value, which we call the reference.
-As discussed in Appendix~\ref{app:super}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting functions~$w_i(x)$ sum up to one uniformly,
+where~$x_0$ is an arbitrary fixed target value, which we call the reference, and $\blambda: {\cal X}\to \mathbb{R}_+^n$ is a weighting function here.
+Note that the reference weight values $\lambda_i(x_0)$ play no role and can therefore be conventionally assumed to be zero.
+As discussed in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting function sums up to one uniformly,
 $$
-\forall x\in{\cal X},\quad
-\sum_{i=1}^n w_i(x) = 1,
+\forall x\in{\cal X}\setminus \{x_0\},\quad
+\sum_{i=1}^n \lambda_i(x) = 1,
 $$
@@
 \end{equation} 
-plays the role of a standard likelihood at prediction time.
+acts as a standard likelihood at prediction time.
 We call this form super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs.
-SCL further generalizes CL since $L^{sc}_\blambda(x)=L^c_\blambda(x)/L^c_\blambda(x_0)$ for target-independent weights $\lambda_i(x)\equiv \lambda_i$, hence it is in this case proportional to CL and yields the same predictive distribution.
+SCL further generalizes CL since both forms are seen to be proportional for target-independent weights.
 
-Conversely, when each target assigns a weight one to a single feature and zero to all the others, SCL turns out to be identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, the ``PDF projection theorem'', which in fact boils down to an $I$-projection \cite{Minka-04}.
-While our derivation stems from a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
+Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, $\forall x\in{\cal X}\setminus \{x_0\}$, $\lambda_i(x)=1$ for a single feature~$i$, and zero otherwise), then SCL turns out identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument; namely, the {\em PDF projection theorem}, which in fact boils down to an $I$-projection \cite{Minka-04}.
+While our derivation follows a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
 
-%Case of a single nonzero agent per feature.
-Then SCL is equivalent to the class-specific likelihood derived in \cite{Baggenstoss-03} based on a generative model selection argument  Our derivation is much different but interesting to see that we end up with a more general form.
-
-A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which happen if the feature generative distributions $f(z_i|x)$ are unknown for some target values~$x$.
-It is possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs.
-Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence on the predictive distribution.
+A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which happen if the feature generative distributions $f(z_i|x)$ are unknown for some targets~$x$.
+SCL makes it possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs.
+Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence whatsoever on the predictive distribution.
 
@@
 
-If training data is available, the SCL weights may be learned by maximum likelihood as in Section~\ref{sec:learning}, yielding a similar convex problem similar.
-Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen.
+If training data is available, the SCL weights may be learned by maximum likelihood as described for CL in Section~\ref{sec:learning}, yielding a similar convex problem.
+Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen (although SCL may not be asymptotically consistent unlike CL, see Appendix~\ref{app:frequentist}).
 The $I$-projection interpretation still holds, yet with class-specific mean-value constraints:
@@
 
-Beside greater flexibility than CL, a disadvantage of SCL it that interpretation becomes relative to an arbitrary reference~$x_0$.
-In some applications, there may exist a natural reference class to cling to in common practice (for instance, normality in medical diagnosis), but this is not always the case.
+The composite Bayesian framework presented here unifies several seemingly unrelated concepts from computational statistics and machine learning: na\"ive Bayes, composite likelihood \cite{Varin-11}, class-specific method \cite{Baggenstoss-03}.
+It ultimately provides a rich class of trainable predictive models that work on pre-determined data features, and is therefore suitable for shallow learning or transfer learning tasks, where it may compete with popular discriminative models such as logistic regression, support vector machines or random forests.
 
-It is easy to build a good classifier.
-You just need to pick a parametric family that is big enough to produce a good fit yet small enough to avoid overfitting.
-Less easy is to interpret what the classifier does.
-Deep learning is cool but it is relevant to AI only if the representations it learns are generic enough to be used in other problems, i.e.
-it should ultimately be validated via transfer learning.
+Unlike previous attempts at Bayesian inference from incomplete statistical models relying on generative model selection \cite{Yuan-99b,Wang-14} or asymptotic theory \cite{Pauli-11,Ribatet-12}, our construction is based on a discriminative model selection approach.
+Composite Bayesian models are thus explicitly taylored to prediction and cannot simulate complete data but, contrary to conventional discriminative models, they encapsulate feature generation models.
 
-{\color{red} Why prediction performance is still important? Because it validates (or not) interpretation. Interpretation is the goal but prediction performance is a crucial intermediate step.}
+Training a composite Bayesian model involves the sequential estimation of feature generative parameters and feature relevance weights.
+This two-stage training is reminiscent of restricted Boltzmann machines (RBMs) \cite{Hinton-06,Fischer-14}, which are fully generative models, with the important difference that it operates on {\em disjoint} sets of parameters in the case of composite Bayesian models, while RBM training modifies the same parameters twice.
+Also, the generative pre-training of RBMs is typically unsupervised, and it is an open research issue whether composite Bayesian models are suitable for unsupervised learning.
 
-{\color{red} We cannot rank features according to their weights.
-A feature might incidentally get a large weight because its likelihood is flat.
-Feature relevance also depends on conditional likelihood distribution.
-Consider mutually independent features: each one recieves a unit weight, but some may be independent from target.}
+The semi-generative nature of composite Bayesian models enables to test, at prediction time, how consistent the data at hand is with each feature distribution, which in turn enables to validate the overall model prediction.
+Moreover, since feature log-likelihoods are combined additively, it is easy to determine which features contribute most to a particular comparison of hypotheses about the target, and summarize the rational for a prediction by highlighting a few decisive features.
 
-{\color{red} Can we use the same dataset to learn both the feature generative distributions and the composite weights? In principle, nothing speaks against this.}
+{\color{red}Example.}
 
-{\color{red} Semi-generative method. Good for shallow learning or transfer learning. Still need representation learning methods.}
-
-{\color{red} Alternative to logistic regression in deep learning? Would require a game-theoretic learning approach... future work.}
-
-{\color{red} Composite Bayes is invariant by affine transformation of the data (unlike multinomial LR, which is problematic when damping is applied).}
-
-{\color{red} Is damping important? (it guarantees training cost is stricly convex).}
-
-{\color{red} Can we interpret maximum weight likelihood as a ``maximum relevance minimum redudancy'' feature selection approach?}
-
-Is it new to see composite likelihood as opinion pooling?
-Yes it is.
-Some previous attempts were made to derive composite likelihood as some sort of consensus generative model \cite{Wang-14}.
-Also some previsous attemps at Bayesian inference from CL using asymptotic theory.
-
-%In other words, is $\bar{a}(\blambda)$ a measure of feature redundancy?
-At fixed $\y$, if one $\lambda_i\to\infty$, then $a(\blambda,\y)\ot\infty$ except if $\ell_i(x,\y)\preceq 0$, in which case it goes to $-\infty$.
-If using likelihood ratios, this only happens if the reference class is the most likely for $\y$.
-But then we can prove that it can't happen for all $\y$ (otherwise the associated moment would be negative), so basically $\bar{a}(\blambda)$ blows up when at least one weight gets large.
-And if several weights get large simultaneously, then it blows up even more so we might consider it a redundancy measure to some extent...
-
-Some experimental observations.
-In the iris dataset (historical data used by Fisher to demonstrate linear discriminant analysis), composite Bayes finds that petal length is the most relevant feature followed by petal width, while sepal length and width get zero weight.
-In the wine dataset, the flavanoids is the most relevant feature for standard composite Bayes, but it's color intensity closely followed by flavanoids for the super composite version.
-In the breast cancer, both approaches are equivalent since it is a binary problem, weights are very sparse: only 13/30 are non-zero with worst area the most relevant.
-In the digits prolem, the most relevant features tend to be in the image center; composite Bayes can achieve similar performance to multinomial logistic regression with about 10 times less parameters.
-
-Number of independent discriminative parameters in logistic regression: $(K-1)(F+1)$.
-In standard composite inference: $F$ (or $K-1+F$ with offsets).
-Difference $=(K-2)(F+1)+1$ (or $=(K-2)F$).
-In super-composite inference: $(K-1)(F+1)$ (if reference is a class).
-Number of generative parameters in composite inference using univariate Gaussian models: $2K$ with heteroscedasticity, $K+1$ with homoscedasticity.
-
-Intuition behind weights: notion of global feature relevance.
-But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$ for a particular instance?
-No.
-But we can assess the influence of each scaled likelihood in order to rank features according to their influence on the decision.
-
-CL is a concept from computational statistics that has mainly been developed so far in a frequentist perspective as a surrogate for the maximum likelihood method.
-We have shown a deep connection between CL and probabilistic inference, thereby establishing CL as a class of discriminative models.
-Because CL is built from a set of marginal feature-specific generative distributions, it is in essence a two-step semi-generative, semi-discriminative learning approach.
-In the first, ``generative'' phase, the feature distributions are learned; in the second, ``discriminative'' phase, the feature weights are learned.
-This strategy can be thought of as a form of non-adaptive boosting.
-
-%The first phase corresponds to the training of ``weak learners''. The second phase amounts to a form of boosting.
-
-A purely discriminative learning could be used instead but...
-Why potentially more efficient than pure discriminative training: because generative training is always more efficient than discriminative training if models are comparable.
-So the key point is that ``we have less parameters in the discriminant phase".
-Good in small datasets.
-But also in asymptotic regime if the features are weakly or highly correlated (???).
-
-Logistic regression / naive Bayes example.
-Under homoscedasticity (assuming that each feature has class-independent variance), CBI is equivalent to logistic regression -- because the predictive distribution family is the same.
-This is a case where BCI brings nothing.
-But consider heteroscedasticity, then BCI yields a quadratic classifier.
-Compared to a fully discriminative model, the number of parameters to learned in the discriminative phase is reduced by half.
-
-The first training phase is easier if supervised but could be unsupervised too (using mixture models).
-How do we then deal with label switching issues?
-Can we safely assume conditional feature independence {\em in the generative training phase}?
-I believe so provided that the marginal distribution parameters are disjoint.
-It's obvious in the supervised learning scenario.
-
-If unsupervised, the first learning phase could be compared with contrastive pre-training of RBMs \cite{Hinton-06,Fischer-14}, which also optimize parameters for generation of observable features.
-BCI is comparable with an RBM with a single output unit (which is just a generative model assuming conditionial feature independence).
-The key difference is that the RBM is a full generative model while BCI only deals with marginal models, hence relying on weaker assumptions.
-This won't change anything in the pre-training phase but RBMs have to deal with more parameters in the generative learning phase.
-In fully supervised context, RBM pre-training is pointless since all parameters learned in the first phase will be overwritten.
-In BCI, pre-training is crucial even in supervised context because the parameters learned in the discriminative phase (the feature weighs) describe a sub-manifold of the predictive distribution family.
-
-Needs features. Not a representation learning method, but could be coupled why not.
-
-%{\color{red} Moreover, while RBM parameters are typically refined in a supervised discriminative learning step, disjoint set of parameters for SCL. Dunno how to say that.}
-
-%In such context, SCL competes with classical discriminative models (logistic regression, Gaussian processes \cite{Rasmussen-06}, maximum entropy models \cite{BergerA-96}, etc.), and may compare more or less favorably in practice depending on the amount of training data.
-For relatively small training datasets, we may hope for more accurate inference using~SCL than using traditional discriminative models, extrapolating from the results of \cite{Ng-01} regarding the comparison between logistic regression and na\"ive Bayes classifiers.
-%{\color{red}Ici, il manque la comparaison avec les RBMs qui ne sont pas (forcement) des modeles discriminatifs.}
-
-%%, hence alleviating the need for heavily supervised model training
-
-%In summary, CL has the potential to yield weakly supervised or unsupervised Bayesian-like inference procedures depending on the particular task at hand.
-This property reflects the encoding of statistical relationships between the data and {\em all} unknown parameters.
-CL thus appears as a trade-off between generative models, which are optimal for unsupervised learning but possibly intractable, and traditional discriminative models (logistic regression, Gaussian processes \cite{Rasmussen-06}, maximum entropy models \cite{BergerA-96}, etc.), which are inherently supervised.
-CL models are discriminative models assembled from atomic generative models and, from this point of view, may be considered as {\em semi-generative} models.
-
-%CL may be considered as a {\em semi-generative} model: a discriminative model assembled from partial generative models.
-
-%As a note of caution, we shall stress that the pre-determined weights assigned to the different associations between observed and unobserved values represent prior knowledge regarding the informativeness of clues.
-A poor choice of weights will inevitably result in a poor approximation to the ``true'' Bayesian posterior -- the posterior that would be obtained from a realistic generative model if it was tractable.
-In future work, we will investigate feature selection strategies to mitigate this problem.
-
-% Improve the discussion on following aspects:
-% ** Why is it compatible with unsupervised learning? Give more insight.
-% ** Stress the contribution: class-specific weighting.
-% * Pivotality argument.
-% * Bayes is a special case of composite Bayes.
-
-{\color{red}Product of fucking experts \cite{Hinton-02}.
-Log-linear pool to build a generative model (assuming in fact independence between experts).
-Better seen as a ``log-mixture model''.
-Sounds like each expert is associated with a class, so it's the same as a RBM.
-Contrastive learning approximates ML parameter estimation.
-The experts are learning jointly.
-It's just a generative model.}
-
-{\color{red}Two-step training: generative phase then discriminative phase. Why is it cool?}
-
-{\color{red}CBI is just a way to reweight Naive Bayes. What's the big deal? Can we really expect massively superior performance? Are we just talking about realistic credibility sets?}
-
-Future work: can we collapse generative and discriminative training into a single step? Maybe using the concept of {\em minimally informative likelihood} \cite{Yuan-99b}.
+We believe that there are many applications of statistical inference in which interpretation is the real goal while prediction performance is only an intermediate requisite.
+A prediction method needs to prove reliable to be relevant; but, once proven reliable, it may need to be interpreted.
+Composite Bayesian inference may come into play when the {\em why} question is more important than the {\em what} question.
 
@@
 
-Note that $\psi(\blambda)$ is nothing but the dual function associated with the maxent problem~(\ref{eq:maxent}).
+Note that $\psi(\blambda)$ is nothing but the dual function associated with the $I$-projection problem~(\ref{eq:i_proj}). The derivatives of~$\psi(\blambda)$ are found to be:
+\begin{eqnarray*}
+\nabla\psi(\blambda)
+ & = & \bar{\bell} - \frac{1}{N} \sum_k \nabla a(\blambda,\y_k), \\
+\nabla\nabla^\top\psi(\blambda)
+ & = & - \frac{1}{N} \sum_k \nabla \nabla^\top a(\blambda,\y_k),
+\end{eqnarray*}
+with:
+$$
+\nabla a(\blambda,\y) = \E_{\blambda}(\bell),
+\qquad
+\nabla \nabla^\top a(\blambda,\y) = {\rm Var}_{\blambda}(\bell),
+$$
+where $\E_{\blambda}$ and ${\rm Var}_{\blambda}$ respectively denote the expectation and variance with respect to $p_\blambda(x|\y)\propto \pi(x)\exp[\blambda^\top \bell(x,\y)]$ at fixed~$\y$.
+This shows that $a(\blambda,\y)$ is convex in $\blambda$ since the variance matrix is positive, which, in turn, proves the concavity of~$\psi$.
 
-Derivatives...
+\section{Heterogeneous log-linear pooling}
+\label{app:hetero}
+
+Consider a sequence of probability distributions $p_i(x)$, $i=1,\ldots,n$, representing multiple opinions on a variable $x\in{\cal X}$.
+The general goal of opinion aggregation is to define an operator $F(p_1,\ldots,p_n)$ that takes opinions as inputs and outputs a single distribution $p(x)$ representing a consensus.
+
+Log-linear pooling \cite{Genest-86} is one particular such operator:
 $$
-\nabla\psi(\blambda)
-= \bar{\bell} - \frac{1}{N} \sum_k \nabla a(\blambda,\y_k) 
+p(x)\propto \pi(x) \prod_{i=1}^n p_i(x)^{\lambda_i},
+$$
+where $\pi(x)$ is an arbitrary distribution and $\lambda_i\in\mathbb{R}$ are arbitrary weights.
+It is easily seen that, if the weights sum up to one, log-linear pooling is externally Bayesian in the sense that, for any nonvanishing positive valued function $q:{\cal X}\to \mathbb{R}^\star_+$, opinion pooling and opinion updating commute:
+$$
+F[N(q p_1), \ldots, N(q p_n)]
+=
+N[q F(p_1,\ldots, p_n)],
+$$
+where~$N$ denotes the normalization endomorphism of $\mathbb{R}_+^{\cal X}$:
+$$
+N(f)(x) = \frac{f(x)}{\sum_{x'\in{\cal X}} f(x')}
+.
 $$
 
+Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$, is of the above form with unit sum weights and is therefore a log-linear opinion pool.
+
+A question that arises naturally is whether it is possible to weight opinions depending on targets, so that $\lambda_i(x)$ becomes a function of~$x$.
+This would make it possible, for instance, to consider an agent reliable when expressing an opinion about a certain hypothesis~$x_1$, but unreliable when it comes to another hypothesis~$x_2$.
+As a first try, let us consider a straightforward extension of log-linear pooling:
 $$
-\nabla\nabla^\top\psi(\blambda)
-= - \frac{1}{N} \sum_k \nabla \nabla^\top a(\blambda,\y_k)  
+p(x)\propto \prod_i p_i(x)^{\lambda_i(x)},
+$$
+where $\lambda_i : {\cal X}\to\mathbb{R}$ is some weighting function. Unfortunately, this operator is generally not externally Bayesian because $F[N(q p_1), \ldots, N(q p_n)](x)
+\propto r(x) F(p_1,\ldots, p_n)(x)$ with:
+$$
+r(x) = q(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)},
+\qquad 
+z_i = \sum_{x\in{\cal X}} q(x) p_i(x),
+$$
+and $r(x)$ cannot be made proportional to $q(x)$, except with constant weights, because the normalizing factors~$z_i$ may differ from one agent to the other.
+
+Consider, however, the following alternative:
+\begin{equation}
+\label{eq:hetero_log_pool}
+p(x) \propto \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)},    
+\end{equation}
+where $x_0$ is some arbitrary reference value.
+This does the trick because, for any positive function $q(x)$, the ratio $q(x)p_i(x)/q(x_0)p_i(x_0)$ is left unchanged by normalization of $q p_i$ since the normalizing factors are the same at both the numerator and the numerator, hence they cancel out, yielding:
+$$
+F[N(q p_1), \ldots, N(q p_n)](x)
+\propto
+\left[\frac{q(x)}{q(x_0)}\right]^{\sum_i \lambda_i(x)}
+F(p_1, \ldots, p_n)(x),
+$$
+which shows that~(\ref{eq:hetero_log_pool}) is externally Bayesian whenever the weighting function verifies:
+$$
+\forall x \in {\cal X}\setminus \{x_0\},
+\qquad
+\sum_i \lambda_i(x) = 1.
 $$
 
-We have:
-$$
-\nabla a(\blambda,\y) = \E_{\blambda}(\bell|\y),
-\qquad
-\nabla \nabla^\top a(\blambda,\y) = \text{Var}_{\blambda}(\bell|\y)
-$$
-proving that $a(\blambda,\y)$ is convex in $\blambda$ for any $\y$, and in turn proves the concavity of $\psi$.
+The operator~(\ref{eq:hetero_log_pool}) clearly generalizes standard log-linear pooling and has the property that the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
+The ``trick'' to enable heterogeneous (target-dependent) weights is essentially to combine the odds relative to a fixed reference.
 
-To see this, let:
-$$
-z(\blambda,\y) 
-\equiv e^{a(\lambda,\y)}  
-= \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}
-$$
+%Let us stress that this result does not contradict the above mentionned characterization theorem \cite{Genest-86b}, which restricts itself to operators for which the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ while, in the heterogeneous log-linear pool, $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
+Note that \cite{Genest-86b} also provides the general characterization of externally Bayesian operators, of which the heterogeneous log-linear pool is just a special case slightly more general than the log-linear pool.
 
-We have:
-$$
-\nabla a = \frac{\nabla z}{z}
-$$
-
-$$
-\nabla\nabla^\top a 
-= \left(
-\frac{\nabla \nabla^\top z}{z} 
-- \frac{\nabla z \nabla z^\top}{z^2}
-\right)
-$$
-
-$$
-\nabla z 
-=
-\sum_{x\in{\cal X}} \pi(x) \bell(x,\y)  e^{\blambda^\top \bell(x,\y)}
-$$
-
-$$
-\nabla \nabla^\top z 
-=
-\sum_{x\in{\cal X}} \pi(x) \bell(x,\y) \bell(x,\y)^\top e^{\blambda^\top \bell(x,\y)}
-$$
-
-\section{Super log-linear pool}
-\label{app:super}
-
-Opinions: $p_i(x)$.
-
-Log-linear pool:
-$$
-p(x)\propto \prod_i p_i(x)^{\lambda_i}
-$$
-
-What about having $\lambda_i(x)$ depend on the target?
-
-Can we do: $p(x)\propto \prod_i p_i(x)^{\lambda_i(x)}$? Is it externally Bayesian?
-
-Let the opinions be multiplied by some function $f(x)$. The new opinions read: $q_i(x)=f(x)p_i(x)/z_i$ with $z_i=\int f p_i$. So the pool gives:
-$$
-q(x) \propto p(x) f(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)}
-$$
-
-So, it's not EB because $z_i$ may differ from one expert to the other.
-
-It feels more reasonable to do:
-$$
-p(x) \propto \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)}
-$$
-
-Is it EB? We have:
-$$
-\frac{q_i(x)}{q_i(x_0)} 
-= \frac{f(x)p_i(x)}{f(x_0)p_i(x_0)}
-$$
-because the $z_i$'s cancel out. Therefore,
-$$
-q(x) \propto p(x) \left[\frac{f(x)}{f(x_0)}\right]^{\sum_i \lambda_i(x)}
-$$
-
-So it's EB if the unit sum weight constraints are satisfied.
-
-\section{Notes Genest}
-
-External Bayesian: pooling and updating commute.
-A pooling operator is compliant with the `likelihood principle' if the pool at $x$ depends on the opinions only through the probabilities $p_i(x)$.
-
-Characterizaton of externally Bayesian pooling operators (EBPOs).
-
-Any EBPO LP-compliant that does not depend on $x$ is a log-pool with unit sum weights:
-$$
-p(x) \propto \prod_i p_i(x)^{w_i}
-$$
-
-Any EBPO LP-compliant is a generalized log-pool with unit sum weights:
-$$
-p(x) \propto \pi(x) \prod_i p_i(x)^{w_i}
-$$
-
-Weights can be negative if domain of $x$ is finite.
-
-To characterize more general EBPOs (not LP-compliant), they introduce an equivalence class.
-Two sets of opinions are equivalent if any ratio of two experts is the same up to a constant factor.
-It means that $q\sim p$ if there exists $f(x)$ such that
-$q_i(x) \propto f(x) p_i(x)$, $\forall i$.
-
-Define an arbitrary member $p_{a1},p_{a2},\ldots$ in each equivalence class.
-
-Any EBPO is of the form:
-$$
-T(p_1,p_2,\ldots) \propto \pi_a(x) v_a(x) \frac{p_1(x)}{p_{a1}(x)},
-\qquad
-v_a \geq \max (p_{a1},p_{a2},\ldots)
-$$
+%It may seem at first sight that this result contradicts the characterization theorem \cite{Genest-86b} since~(\ref{eq:hetero_log_pool}) is clearly more general than log-linear pooling.
+There is no contradiction because the said theorem restricts itself to operators for which the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ while, in the heterogeneous log-linear pool, $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
 
@@
 
+We note that $\log r$ is a second-order polynomial of~$x$:
+$$
+f(x) = a x^2 + b x + c,
+$$
+with $a=(\rho^2-1)/2\sigma_0^2$ and $b=(\mu_1-\rho^2\mu_0)/\sigma_0^2$.
+
+Assuming $X\sim N(\mu_1,\sigma_1)$, and given an observation $x$, we would like to compute the probability:
+\begin{eqnarray*}
+p 
+& = & P[f(X)>f(x)] \\
+& = & P\{[f(X)-f(x)]>0\}\\
+& = & P\{[a(X^2-x^2)+b(X-x)]>0\}\\
+& = & P\{(X-x)[a(X+x)+b]>0)\}\\
+& = & P[(X-x)(X-y)>0]\\
+& = & 1 - P[X\in(x,y)]\\
+& = & 1 - |G(x)-G(y)|
+\end{eqnarray*}
+where $y=-x-b/a$, and $G=G_{\mu_1,\sigma_1}$ is the Gaussian partition function. In this case, we have:
+$$
+\frac{b}{a} = 2 \frac{\mu_1-\rho^2\mu_0}{\rho^2-1}
+$$
+
+Special case if $\rho=0$, then $a=0$ and $b=\mu_1-\mu_0$, we can't introduce $y$. Going backward, we see that:
+$$
+p = P[(X-x)b>0]
+$$
+
+If $\mu_1>\mu_0$, the answer is $p=1-G(x)$. If $\mu_1<\mu_0$, it is $G(x)$. If $\mu_1=\mu_0$, it is just zero.
+
+So, the story is quite easy to tell.
+Say that we want to compare two hypotheses $x_0$ and $x_1$, one of which is assumed to hold.
+It is easy using CL to isolate the features that produce the most significant contribution to the log-CL difference of $x_1$ and $x_0$.
+Assume that $x_1$ is the one that yields the largest predictive probability, so we are interested in positive contributions (trying to ask: why does the system think $x_1$ is more probable?).
+
+For any such feature, we may ask if such a large contribution is expected under $x_1$ in order to confirm the prediction.
+We then compute the above probability $p$.
+If $p\approx 0$ or $p\approx 1$, the contribution is abnormal, respectively too large or too small.
+This raises a flag, indicating that $x$ may NOT be distributed according to $x_1$.
+And, since $x_1$ is more probable than $x_0$, neither hypothesis holds in fact.
+It could be a distribution intermediate between $x_0$ and $x_1$ (if $p\approx 1$), or farther away from $x_0$ than $x_1$ (if $p\approx 0$).
+
+Note that the $p$ is insensitive to the feature weight in CL...
+The weights do however play a role in identifying the features relevant to a particular comparison.
+If a feature is not relevant, it does not matter if
+
+%\bibliographystyle{ieeetr}
 \bibliographystyle{abbrv}

### 2019-04-11 papers/composite_inference/draft.tex
Finalize CB paper

@@
 We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
-The resulting class of predictive models, which we call composite Bayesian models, is a middle way between purely generative and purely discriminative models.
-This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning; in the latter case, they may be tuned so as to minimize training cross-entropy, yielding an easy-to-solve convex problem.
+The resulting class of predictive models, which we call composite Bayesian models, is a middle way between generative and discriminative models.
+This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning; in the latter case, they may be tuned so as to minimize prediction cross-entropy, yielding an easy-to-solve convex problem.
 We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
@@
 Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable.
-Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the conditional distribution of the target variable given the data, while a {\em generative} model describes the joint distribution of the target and the data.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
+Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the conditional distribution of the target variable given the data, while a {\em generative} model describes the joint distribution of the target and the data.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical machine learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).
 
-In a closed universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions.
-Consider as a straightforward example deciding whether an object is a sauce pan or a frying pan based on its depth.
+In a well-defined universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions.
+Consider as a straightforward example deciding whether an object is a sauce pan or a frying pan based on depth.
 A two-class discriminative model can learn a threshold that correctly classifies most pans in practice, but will confidently classify a blender as a sauce pan, while linear discriminant analysis\footnote{Despite the name, linear discriminant analysis is based on a generative model.}, for instance, could hint that a blender is a poor match to both pan categories.
-Inability to ``confirm'' or ``justify'' predictions makes discriminative models difficult to interpret.
-
-%Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves.
-Consider as a straightforward example the problem of deciding whether an object is a sauce pan or a frying pan based on depth.
-A discriminative model can learn a threshold that correctly classifies most pans in practice, but is unable to see that, say, a blender does not match either category.
-The impossibility of such confirmatory analysis strongly limits interpretation if outlier inputs are to be expected.
+Intrinsic inability to ``confirm'' or ``justify'' predictions makes discriminative models difficult to interpret.
 
@@
 For instance, in medical diagnosis, automated disease predictions need to be corroborated by individualized findings to be taken into account.
-In other words, there is little clinical value in ``black boxes'', but certainly more in automated methods that can justify their outputs just like human experts.
+In other words, there is little clinical value in ``black boxes'', and certainly more in automated methods that can ``justify'' themselves just like human experts.
 One such method that has long been used by clinicians is the familiar reference range analysis, which stems from simple generative models.
@@
 The agent opinions are then aggregated into a unique predictive probability distribution analogous to a Bayesian posterior.
-This strategy may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
+This idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.
 
@@
 
-When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL), see \cite{Varin-11} and the references therein.
+When choosing the aggregation method as log-linear pooling, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL), see \cite{Varin-11} and the references therein.
 CL was developed as a surrogate of traditional likelihood for parameter estimation.
@@
 
-Based on the opinion pooling interpretation, the weights assigned to the different features in~CL may be optimized for prediction performance in a typical supervised learning scenario.
+Following the opinion pooling interpretation, the weights assigned to the different features in~CL may be optimized for prediction performance in a typical supervised learning scenario.
 As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions.
@@
 
-The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on the target values, leading to a more flexible predictive model that can be trained similarly to CL.
+The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on the target values, leading to a more flexible predictive model that can be trained similarly.
 
@@
 The reader is warned that slightly abusive mathematical notation is sometimes used in the remainder.
-This is done deliberately for the sake of clarity in places where we think that notation can be lightened without raising ambiguity.
+This is done deliberately for the sake of clarity in places where we think that notation can be lightened without ambiguity.
 
@@
 $$
-where the distribution $\pi(x)$ plays the role of a prior, and the function:
+where the distribution $\pi(x)$ plays the role of a prior, while the quantity:
 \begin{equation}
@@
 
-If training data is available, the generative parameters may be set in a first stage using standard estimation techniques.
+If training data is available, the generative parameters may be tuned in a first stage using standard estimation techniques.
 Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to CL one of the same techniques as for traditional likelihood \cite{Berger-99}.
@@
 This is motivated by a characterization theorem for opinion pooling \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome, is the log-linear opinion pool with unit sum weights.
-Externally Bayesian essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
+The external Bayesian property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
 Log-linear pooling with unit sum weights also turns out to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
-
-%While other divergence measures lead to other optimal pooling operators (for instance, the exclusive KL divergence leads to the linear pool), they all enforce unit sum weights, as is natural in the search for a best compromise.
 
@@
 
-%, which, as we shall see, is unlikely to pick uniform or unit sum weights.
-
 \subsection{Learning the weights}
@@
 
-Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$ (possibly the same as for generative pre-training), the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
+Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$ (possibly the same as for generative pre-training), the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution or, equivalently, minimize prediction ``cross-entropy'':
 \begin{equation}
@@
 As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer diagnostic UCI dataset \cite{Wolberg-94}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs radically from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
-
-%\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}
 
@@
 
-The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized and that the $I$-projection is equivalent to maximum conditional entropy \cite{BergerA-96}.
+The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized, making the $I$-projection equivalent to maximum conditional entropy \cite{BergerA-96}.
 The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
 This is weaker a constraint than assuming that the feature generative models $f(z_i|x)$ are true: it only assumes that the {\em description length} \cite{Grunwald-07} achieved by each feature model is a truthworthy upper bound.
-
-%In the $I$-projection framework, the composite weights are the Lagrange multipliers associated with the mean log-likelihood inequality constraints.
+The composite likelihood weights may be interpreted as the Lagrange multipliers associated with these description length constraints.
 
 An insightful analysis given in \cite{Grunwald-04} shows that $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss.
-Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is inexact due to the limited training set size.
+Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is {\em inexact} in practice due to the finite training set.
 Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by its underlying inexact knowledge, which in this case may be interpreted in terms of feature description length.
@@
 where~$x_0$ is an arbitrary fixed target value, which we call the reference, and $\blambda: {\cal X}\to \mathbb{R}_+^n$ is a weighting function here.
-Note that the reference weight values $\lambda_i(x_0)$ play no role and can therefore be conventionally assumed to be zero.
-As discussed in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting function sums up to one uniformly,
+Note that the reference weights $\lambda_i(x_0)$ play no role and can therefore be conventionally assumed to cancel.
+As shown in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make pooling externally Bayesian if the weighting function sums up to one uniformly,
 $$
@@
 hence providing a sensible default rule to tune the weights.
-Note that (\ref{eq:super_pool}) amounts to a log-linear pool if the weights are target-independent as the reference is then effectless.
+Note that (\ref{eq:super_pool}) amounts to log-linear pooling if the weights are target-independent as the reference is then effectless.
 Using the same analogy with Bayes rule as in Section~\ref{sec:bayes_rule}, we see that the quantity:
@@
 \end{equation} 
-acts as a standard likelihood at prediction time.
-We call this form super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs.
+acts as a likelihood function at prediction time.
+We call the form~(\ref{eq:super_comp_lik}) super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs.
 SCL further generalizes CL since both forms are seen to be proportional for target-independent weights.
 
-Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, $\forall x\in{\cal X}\setminus \{x_0\}$, $\lambda_i(x)=1$ for a single feature~$i$, and zero otherwise), then SCL turns out identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument; namely, the {\em PDF projection theorem}, which in fact boils down to an $I$-projection \cite{Minka-04}.
+Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, $\forall x\in{\cal X}\setminus \{x_0\}$, $\lambda_i(x)=1$ for a single feature~$i$, and zero otherwise), then SCL turns out identical to the class-specific likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument; namely, the {\em PDF projection theorem}, which in fact boils down to an $I$-projection \cite{Minka-04}.
 While our derivation follows a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
@@
 
-%Interest?
-If some feature likelihood values are missing, i.e.
-agents are not able to model the whole set of targets.
-It's a bit like censoring.
-Then it's ok if they can model at least two targets including the reference: $\ell_i(x)$ is unknown for the other values but we can give them zero weight so it doesn't matter.
-
 If training data is available, the SCL weights may be learned by maximum likelihood as described for CL in Section~\ref{sec:learning}, yielding a similar convex problem.
-Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen (although SCL may not be asymptotically consistent unlike CL, see Appendix~\ref{app:frequentist}).
+Since SCL enables to explore a wider class of predictive models than CL, it has the potential to achieve better prediction performance as long as overfitting does not happen during training (although SCL may not be asymptotically consistent unlike CL, see Appendix~\ref{app:frequentist}).
 The $I$-projection interpretation still holds, yet with class-specific mean-value constraints:
@@
 
-%Note that
-
-%In summary, SCL is a natural generalization of CL with greater flexibility in weight assignments, hence potentially superior prediction performance if trained on a large enough dataset.
-It, however, depends on an arbitrary reference class.
-
-%Beside greater flexibility than CL, a disadvantage of SCL it that interpretation becomes relative to an arbitrary reference~$x_0$.
-In some applications, there may exist a natural reference class to cling to in common practice (for instance, normality in medical diagnosis), but this is not always the case.
-
-%Another shortcoming is that SCL is not guaranteed to be asymptotically consistent, see Appendix~\ref{app:frequentist}, so it may yield biased MCLE despite
-
-%Disadvantage: interpetation is relative to an arbitrary class $x_0$.
-Advantages: richer model (so better predictions), probabilty ratios stable to class addition.
-Drawbacks: require a reference, no more consistency.
-Still interpretable, that's why it is worthwile mentioning.
-
-%Argue that SCL more robust than CL to mispecification of feature models.
-Intuitively, if a feature model is poor, the empirical description length is large, hence can be achieved with a `small' Lagrange multiplier $\lambda_i(x)$.
-Therefore, unreliable likelihoods $\ell_i(x)$ tend to get low weights.
-
-% which does not contradict the axiomatic derivation in \cite{Genest-86b} as it restricted itself to operators that only depend on the probas evaluated at~$x$ unlike~(\ref{eq:super_pool}).
-
-%The real motivation: when agents operate on restricted target sets.
-Without loss of generality, assume binary sets.
-Trick: for each such agent, say we have $p(z_i|x_1)$ and $p(z_i|x_0)$.
-We define $p(z_i|x)=p(z_i|x_0)$ if $x\not= x_1$.
-It does not make any sense but the likelihood ratio is then one for any $x\not= x_1$ so won't hamper the system.
-
-%Due to the distribution of weights between clues, a drawback of composite likelihood is that it is prone to {\em information overload} in the sense that it tends to ``flatten out'' when too many clues are included as relevant clues then get downweighted.
-If one decides not to merge clues for computational reasons (this would require handling their joint distribution), one could hope to mitigate information overload by assigning strong weights only to those clues that are believed to be ``most informative''.
-% When chosen for computational simplicity, features may not only convey limited information at individual level, their informativeness may also be very much class-dependent.
-Consider, for instance, diagnosing a disease from a routine medical checkup.
-Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovasc ular disease -- and conversely for, say, blood pressure.
-
-%Reference is implicit if weights are class-independent as it is then effectless.
-Make sense to have a reference as it gives a customizable baseline for detection.
-Can be the feature distribution averaged over all classes (not recommended I would say) or the distribution of a particular class if that makes sense (e.g.
-healthy population).
-In this case, the super composite likelihood is equivalent to the standard one in the two-class case.
-
-%If some classes are merged or split, then the other classes keep the same weights (assuming $h(\y)$ is kept constant, so we need to make sure the prior is adapted accordingly)!
-So if a patient was found to have 10 times more probability to be MCI than healthy (and healthy is the reference class), and we later split AD population into, say, pure AD and mixed AD, the MCI probably ratio fact will remain.
-
-%Super composite likelihood is essentially about combining binary versions of composite likelihood. It is analogous to the ``one-versus-rest'' strategy in multiple logistic regression.
-
-%Still externally Bayesian if weights sum up to one in each class.
-Alternatively, we can train this model by maximum likelihood, yielding a similar optimization problem as in Section~\ref{sec:learning}.
-
-%Maxent interpretation.
-It is an alternative to the strategy considered above, the mean log-likelihood constraints may be expressed for each class rather than being averaged with respect to the prior, leading to a more general predictive model that we call ``super composite likelihood'', in which the weights become class-dependent (see Appendix~\ref{sec:super}).
-This variant has the potential for higher prediction performance, but poses some conceptual difficulties, in particular that of having to define an arbitrary reference class.
-
-%used to predict whether breast lumps are benign or malignant based on a set of features extracted from fine-needle aspirate images.
-
-%The most discriminative features tend to get large weight, however some quite discriminative features may get low weight if strongly correlated with a more discriminative feature
-
-%Non-linear means that features with average discrimination power may get low weight as they are superseded by the most discriminative features.
-But some features with low discrimination power may also get large weight, which does not mean that they are very influential in practice (see above).
-This is illustrated in Figure~\ref{fig:disc_weight_plot} with the UCI breast cancer dataset.
-
-%{\color{red} Equality constraints (negative weights) are less interpretable.}
-
-%{\color{red} Offsets.}
-
-%{\color{red} Data weights.}
-
-%Need to weight the data to adjust to prior as a general measure when minimizing cross-entropy, which should yield a conditional model that approximates the ``true'' conditional distribution, $q(x|y)\approx p(x|y)$.
-So the model will absorb the ``true'' marginal of $x$, which should thus be replaced by the prior if we want the model to mimic ideal Bayesian inference.
-In our case, it means that the composite likelihood approximates the true likelihood.
-
-%We can further enforce exact prior matching by including offsets ($K-1$ free parameters), in which case the training phase learns $\pi(x)$ so that it matches the prior $\pi(x)$ in the sense $\int h(\y)p_\lambda(x|y)dy = \pi(x)$.
-This ``deforms'' the prior to improve the training score.
-But why is it important to enforce this constraint?
-Again, it all makes things just less interpretable.
-
-%When including offsets, using homoscedastic generative variance estimation and equality constraints, composite likelihood is equivalent to binomial logistic regression.
-However, equality constraints are not easily interpretable.
-
-%In some sense, composite weights replace the multiple testing correction required in the frequentist inference paradigm for proper control of false positives.
-Here, we don't control false positives, we control prediction accuracy.
-
-%Why not using the feature-based posteriors or Bayes factors rather than the likelihoods?
-It's a possibility (perhaps a connection here with Gr\"unwald's luckiness).
-A justification for not doing it is that the coordinator should discard the priors used by the agents.
-The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.
-
-%Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(\y|x)^\lambda$.
-We expect to have $\lambda=1$ but it's not necessarily the case.
-It is if $h(\y)=\int\pi(x)p(\y|x)dx$ because then $h(\y)p(x|y)=\pi(x)p(\y|x)$ so the constraint is verified (and active).
-This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold.
-So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.
-
-%Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general?
-The answer is no.
-The weights generally depend on the prior.
-Hence the maxent composite likelihood is prior-dependent.
-This is an important conceptual difference with the classical notion of likelihood.
-
 \section{Discussion}
@@
 
-The composite Bayesian framework presented here unifies several seemingly unrelated concepts from computational statistics and machine learning: na\"ive Bayes, composite likelihood \cite{Varin-11}, class-specific method \cite{Baggenstoss-03}.
-It ultimately provides a rich class of trainable predictive models that work on pre-determined data features, and is therefore suitable for shallow learning or transfer learning tasks, where it may compete with popular discriminative models such as logistic regression, support vector machines or random forests.
+The composite Bayesian framework unifies several seemingly unrelated concepts from computational statistics and machine learning: composite likelihood \cite{Varin-11}, PDF projection \cite{Baggenstoss-03}, na\"ive Bayes.
+It ultimately provides a rich class of trainable predictive models that work on pre-determined data features, and is therefore suitable for ``shallow'' learning tasks, where it may compete with classical discriminative models such as logistic regression, support vector machines or random forests.
+Potential use cases include transfer learning \cite{Goodfellow-16}, {\em i.e.}, when features are learned automatically in a related task via deep learning, for instance.
 
-Unlike previous attempts at Bayesian inference from incomplete statistical models relying on generative model selection \cite{Yuan-99b,Wang-14} or asymptotic theory \cite{Pauli-11,Ribatet-12}, our construction is based on a discriminative model selection approach.
+Unlike previous attempts at Bayesian inference from incomplete statistical models that relied on generative model selection \cite{Yuan-99b,Wang-14} or asymptotic theory \cite{Pauli-11,Ribatet-12}, our construction is based on a discriminative model selection approach.
 Composite Bayesian models are thus explicitly taylored to prediction and cannot simulate complete data but, contrary to conventional discriminative models, they encapsulate feature generation models.
@@
 This two-stage training is reminiscent of restricted Boltzmann machines (RBMs) \cite{Hinton-06,Fischer-14}, which are fully generative models, with the important difference that it operates on {\em disjoint} sets of parameters in the case of composite Bayesian models, while RBM training modifies the same parameters twice.
-Also, the generative pre-training of RBMs is typically unsupervised, and it is an open research issue whether composite Bayesian models are suitable for unsupervised learning.
+Also, the generative training stage of RBMs is typically unsupervised, and it is an open research issue whether composite Bayesian models are suitable for unsupervised learning.
 
-The semi-generative nature of composite Bayesian models enables to test, at prediction time, how consistent the data at hand is with each feature distribution, which in turn enables to validate the overall model prediction.
+\begin{figure}[!ht]
+\begin{center}
+\subfigure[Case study: clearly categorized class-3 wine]{
+\includegraphics[width=.9\textwidth]{interp_case1.pdf}}
+%\hspace*{.2\textwidth}
+\subfigure[Case study: atypical class-2 wine]{
+\includegraphics[width=.9\textwidth]{interp_case2.pdf}}
+\end{center}
+\caption{Graphical representation of feature influence on wine classification in a composite Bayesian model using heteroscedastic Gaussian feature generation models and uniform prior.
+Bars represent feature contributions to the difference of log-composite likelihoods between the most probable and second most probable class.
+Dots represent $\chi^2$ feature percentiles under the most probable class; green and red colors mark values respectively smaller and larger than 95\%.}
+\label{fig:interp}
+\end{figure}
+
+Interpretability is from our point of view the key advantage of composite Bayesian models.
+It results from their semi-generative nature, which enables to test at prediction time how consistent the input data is with each feature distribution, thereby providing the user with crucial clues to understand and validate the prediction.
 Moreover, since feature log-likelihoods are combined additively, it is easy to determine which features contribute most to a particular comparison of hypotheses about the target, and summarize the rational for a prediction by highlighting a few decisive features.
 
-{\color{red}Example.}
+This is illustrated in Figure~\ref{fig:interp} for a composite Bayesian expert fully trained (including weight optimization as in Section~\ref{sec:learning}) on a famous UCI dataset \cite{Aeberhard-92} to classify wines into 3 categories from 13~features.
+The top panel shows how the expert (correctly) classifies a particular wine in category~3 with near $100\%$ probability; it can be seen that the prediction is mainly based on the flavanoids concentration and the color intensity.
+Since both have typical values for category~3 wines, this is a clear-cut decision.
+In another case depicted in the bottom panel, the wine is (again, correctly) classified in category~2 with $98.9\%$ probability, however two of the main features driving the prediction (alcanality of ash and magnesium concentration) have peculiar values for category~2, while proline concentration quite strongly points to category~1.
+Despite high prediction confidence, this case may require further investigation.
 
@@
 
-Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any composite likelihood function~(\ref{eq:comp_lik}) with positive weights~$\blambda\succeq 0$:
+Using a weighted sum of such inequalities, we can bracket the expected variations of the logarithm of any composite likelihood function~(\ref{eq:comp_lik}) using positive weights~$\blambda\succeq 0$:
 \begin{equation}
@@
 \leq 
-s_\blambda E\left[ \log \frac{L(x_\star)}{L(x)} \right]
+\|\blambda\|_1 E\left[ \log \frac{L(x_\star)}{L(x)} \right]
 ,
 \end{equation}
-with $s_\blambda\equiv \|\blambda\|_1 =\sum_i \lambda_i$, and $L(x)\equiv \log p(\y|x)$ is the true likelihood function.
-
-This implies two general asymptotic properties of composite likelihood:
+where $\|\blambda\|_1 =\sum_i \lambda_i$ and $L(x)\equiv \log p(\y|x)$ is the true likelihood function. This trivially implies two general asymptotic properties of composite likelihood:
 \begin{itemize}
 \item {\em Weak consistency.} The expected composite log-likelihood is maximized by~$x_\star$.
-\item {\em Conservativeness.} If the weights sum up to one or less ($s_\blambda\leq 1$), composite likelihood ratios of the true target~$x_\star$ vs.~other target values~$x$ tend to  underestimate true likelihood ratios.
+\item {\em Conservativeness.} If the weights sum up to one or less ($\|\blambda\|_1\leq 1$), composite likelihood ratios of the true target~$x_\star$ vs.~other target values~$x$ tend to  underestimate true likelihood ratios.
 \end{itemize}
 
-Note that the derivation assumes that the weights~$\blambda$ are independent from the target~$x$, therefore these properties do not necessarily extend to SCL, for which weights are target-dependent.
+Note that the derivation assumes that the weights~$\blambda$ are independent from the target~$x$, therefore these properties do not necessarily extend to SCL.
 
@@
 
-The operator~(\ref{eq:hetero_log_pool}) clearly generalizes standard log-linear pooling and has the property that the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
-The ``trick'' to enable heterogeneous (target-dependent) weights is essentially to combine the odds relative to a fixed reference.
-
-%Let us stress that this result does not contradict the above mentionned characterization theorem \cite{Genest-86b}, which restricts itself to operators for which the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ while, in the heterogeneous log-linear pool, $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
-Note that \cite{Genest-86b} also provides the general characterization of externally Bayesian operators, of which the heterogeneous log-linear pool is just a special case slightly more general than the log-linear pool.
-
-%It may seem at first sight that this result contradicts the characterization theorem \cite{Genest-86b} since~(\ref{eq:hetero_log_pool}) is clearly more general than log-linear pooling.
-There is no contradiction because the said theorem restricts itself to operators for which the consensus evaluated at~$x$, $F(p)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$ while, in the heterogeneous log-linear pool, $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
-
-\section{Log-likelihood ratio}
-
-$$
-\log r = \log \frac{N(x;\mu_1,\sigma_1)}{N(x;\mu_0,\sigma_0)}
-$$
-
-$$
-\log N(x;\mu,\sigma) = -\frac{1}{2}\left[
-\log(2\pi\sigma^2) + \frac{(x-\mu)^2}{\sigma^2}
-\right]
-$$
-
-$$
-\log r = -\frac{1}{2}
-\left[
-\log \frac{\sigma_1^2}{\sigma_0^2}
-+ \frac{(x-\mu_1)^2}{\sigma_1^2} - \frac{(x-\mu_0)^2}{\sigma_0^2}
-\right]
-$$
-
-It follows:
-\begin{eqnarray*}
-\log r 
- & = & 
--\frac{1}{2}\left[
-\log \frac{\sigma_1^2}{\sigma_0^2}
-+ \frac{\sigma_0^2(x-\mu_1)^2 - \sigma_1^2(x-\mu_0)^2}{\sigma_0^2 \sigma_1^2}
-\right] \\
- & = & 
-\frac{1}{2}\left[
-\log \rho^2
-+ \frac{(\rho^2-1)x^2 - 2(\rho^2\mu_0-\mu_1)x+ (\rho^2\mu_0 ^2-\mu_1^2)}{\sigma_0^2}
-\right]
-\end{eqnarray*}
-with $\rho=\sigma_1/\sigma_0$. In the case of same variance ($\rho=1$), this simplies to an affine function of $x$ with offset given by the center of $(\mu_0,\mu_1)$:
-$$
-\log r = \frac{\mu_1-\mu_0}{\sigma^2} \left(
-x - \frac{\mu_0 + \mu_1}{2}
-\right) 
-$$
-
-In any case,
-$$
-E[\log r | H_1] = 
-\frac{1}{2}
-\left[
-- \log \frac{\sigma_1^2}{\sigma_0^2}
-+ \frac{\sigma_1^2 + (\mu_1-\mu_0)^2}{\sigma_0^2}
-- 1
-\right]
-$$
-
-Let $\delta = (\mu_1-\mu_0)/\sigma_0$. We have:
-$$
-E[\log r | H_1] = 
-\frac{1}{2}
-\left[
-\delta^2
-- \log \rho^2
-+ \rho^2  - 1
-\right]
-$$
-
-We note that $\log r$ is a second-order polynomial of~$x$:
-$$
-f(x) = a x^2 + b x + c,
-$$
-with $a=(\rho^2-1)/2\sigma_0^2$ and $b=(\mu_1-\rho^2\mu_0)/\sigma_0^2$.
-
-Assuming $X\sim N(\mu_1,\sigma_1)$, and given an observation $x$, we would like to compute the probability:
-\begin{eqnarray*}
-p 
-& = & P[f(X)>f(x)] \\
-& = & P\{[f(X)-f(x)]>0\}\\
-& = & P\{[a(X^2-x^2)+b(X-x)]>0\}\\
-& = & P\{(X-x)[a(X+x)+b]>0)\}\\
-& = & P[(X-x)(X-y)>0]\\
-& = & 1 - P[X\in(x,y)]\\
-& = & 1 - |G(x)-G(y)|
-\end{eqnarray*}
-where $y=-x-b/a$, and $G=G_{\mu_1,\sigma_1}$ is the Gaussian partition function. In this case, we have:
-$$
-\frac{b}{a} = 2 \frac{\mu_1-\rho^2\mu_0}{\rho^2-1}
-$$
-
-Special case if $\rho=0$, then $a=0$ and $b=\mu_1-\mu_0$, we can't introduce $y$. Going backward, we see that:
-$$
-p = P[(X-x)b>0]
-$$
-
-If $\mu_1>\mu_0$, the answer is $p=1-G(x)$. If $\mu_1<\mu_0$, it is $G(x)$. If $\mu_1=\mu_0$, it is just zero.
-
-So, the story is quite easy to tell.
-Say that we want to compare two hypotheses $x_0$ and $x_1$, one of which is assumed to hold.
-It is easy using CL to isolate the features that produce the most significant contribution to the log-CL difference of $x_1$ and $x_0$.
-Assume that $x_1$ is the one that yields the largest predictive probability, so we are interested in positive contributions (trying to ask: why does the system think $x_1$ is more probable?).
-
-For any such feature, we may ask if such a large contribution is expected under $x_1$ in order to confirm the prediction.
-We then compute the above probability $p$.
-If $p\approx 0$ or $p\approx 1$, the contribution is abnormal, respectively too large or too small.
-This raises a flag, indicating that $x$ may NOT be distributed according to $x_1$.
-And, since $x_1$ is more probable than $x_0$, neither hypothesis holds in fact.
-It could be a distribution intermediate between $x_0$ and $x_1$ (if $p\approx 1$), or farther away from $x_0$ than $x_1$ (if $p\approx 0$).
-
-Note that the $p$ is insensitive to the feature weight in CL...
-The weights do however play a role in identifying the features relevant to a particular comparison.
-If a feature is not relevant, it does not matter if
+The operator~(\ref{eq:hetero_log_pool}) generalizes standard log-linear pooling and has the property that the consensus $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
+The ``trick'' to enable heterogeneous (target-dependent) weights is essentially to consider the odds relative to a fixed reference.
 

### 2019-04-12 papers/composite_inference/draft.tex
Rewrite heterogeneous pooling story

@@
   \begin{center}
-    \includegraphics[width=.6\textwidth]{disc_weight_plot.pdf}
+    \includegraphics[width=.7\textwidth]{disc_weight_plot.pdf}
   \end{center}
@@
 The general goal of opinion aggregation is to define an operator $F(p_1,\ldots,p_n)$ that takes opinions as inputs and outputs a single distribution $p(x)$ representing a consensus.
-
 Log-linear pooling \cite{Genest-86} is one particular such operator:
@@
 $$
-p(x)\propto \prod_i p_i(x)^{\lambda_i(x)},
+p(x)\propto \pi(x) \prod_i p_i(x)^{\lambda_i(x)},
 $$
@@
 \label{eq:hetero_log_pool}
-p(x) \propto \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)},    
+p(x) \propto \pi(x) \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)},    
 \end{equation}
@@
 
+Caca.
+$$
+\log p(x) \equiv \mu(x) + \sum_i \lambda_i(x) \log p_i(x)
+$$
+
+If we look for a pool of the form:
+$$
+\log p(x) \equiv \mu(x) + \sum_i \lambda_i(x) \log p_i(x)
+$$
+
 %\bibliographystyle{ieeetr}

### 2019-04-14 papers/composite_inference/draft.tex
Modifs prob ratio invariance

@@
 $$
-L(x) = p(\y|x)
-.
+L(x) = p(\y|x).
 $$
@@
 $$
-\ell(x) = f(z|x)
-.
+\ell(x) = f(z|x).
 $$
@@
 If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
-A common choice, owing to its simplicity and tendency to produce single peaked distributions, is log-linear pooling\footnote{When $\pi(x)$ is not uniform, it is also called {\em generalized} log-linear pooling, however this terminology could be confusing here since we will present another generalization.}:
+One that is simple and satisfies a much desirable unbiasedness property is log-linear pooling\footnote{When $\pi(x)$ is not uniform, it is also called {\em generalized} log-linear pooling, however this terminology could be confusing here since we will present another generalization.}:
 \begin{equation}
@@
 \end{equation} 
-where $\pi(x)$ is a reference distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+where $\pi(x)$ is a reference distribution and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+Log-linear pooling is {\em unbiased} in the sense that eliminating some hypotheses does not alter the probability ratios between the remaining hypotheses (see Appendix~\ref{app:log_pool_standard}).
 We shall asume for feasibility that the weights are intrinsic to the agents and are therefore independent from the data~$\y$.
-Also, negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less probable than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
+Also, negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an hypothesis $x_1$ is less probable than an hypothesis $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
+As a convenient effect of using positive weights, log-linear pooling tends to produce single peaked distributions.
 
@@
 Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one.
-This is motivated by a characterization theorem for opinion pooling \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome, is the log-linear opinion pool with unit sum weights.
-The external Bayesian property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
+This is motivated by a characterization theorem \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome, is log-linear opinion pooling constrained to unit sum weights.
+The external Bayesian property is stronger than the above-mentionned unbiasedness property; it essentially means that the consensus should not change if any agent opinion is taken into account by the other agents as opposed to being stated independently.
 Log-linear pooling with unit sum weights also turns out to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
@@
 Note that the reference weights $\lambda_i(x_0)$ play no role and can therefore be conventionally assumed to cancel.
-As shown in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make pooling externally Bayesian if the weighting function sums up to one uniformly,
+As shown in Appendix~\ref{app:log_pool_hetero}, the main reason for introducing the normalization by a reference is to make pooling externally Bayesian if the weighting function sums up to one uniformly,
 $$
@@
 
-\section{Heterogeneous log-linear pooling}
-\label{app:hetero}
+\section{Log-linear pooling}
+\label{app:log_pool}
 
-Consider a sequence of probability distributions $p_i(x)$, $i=1,\ldots,n$, representing multiple opinions on a variable $x\in{\cal X}$.
+Consider a sequence of probability distributions $p_i(x)$, $i=1,\ldots,n$, representing multiple opinions on a variable of interest~$x\in{\cal X}$.
 The general goal of opinion aggregation is to define an operator $F(p_1,\ldots,p_n)$ that takes opinions as inputs and outputs a single distribution $p(x)$ representing a consensus.
+
+\subsection{Standard log-linear pooling}
+\label{app:log_pool_standard}
+
 Log-linear pooling \cite{Genest-86} is one particular such operator:
@@
 where $\pi(x)$ is an arbitrary distribution and $\lambda_i\in\mathbb{R}$ are arbitrary weights.
+
+A key property of log-linear pooling is that probability ratios are stable under hypothesis elimination: if the opinions are restricted to any subset ${\cal X}_s\subset {\cal X}$, the consensus~$p_s$ on ${\cal X}_s$ verifies $p_s(x_1)/p_s(x_2)=p(x_1)/p(x_2)$ for any pair $(x_1,x_2)\in{\cal X}_s\times {\cal X}_s$.
+In other words, eliminating some hypotheses does not bias inference towards any of the remaining hypotheses.
+
+Another way to look at this invariance property is to state that, for any boolean-valued function $q:{\cal X}\to \{0,1\}$, opinion pooling and opinion updating commute: 
+\begin{equation}
+\label{eq:external_bayesian}
+F[N(q p_1), \ldots, N(q p_n)]
+=
+N[q F(p_1,\ldots, p_n)],
+\end{equation}
+where~$N$ denotes the normalization endomorphism of $\mathbb{R}_+^{\cal X}$:
+$$
+N(f)(x) = \frac{f(x)}{\sum_{x'\in{\cal X}} f(x')}
+.
+$$
+
+Note that if (\ref{eq:external_bayesian}) holds for any nonvanishing positive valued function $q:{\cal X}\to\mathbb{R}^\star_+$,  the opinion pool is said to be {\em externally Bayesian}, which is thus a more general property than probability ratio invariance.
+It is easily seen that log-linear pooling is externally Bayesian if the weights~$\lambda_i$ sum up to one.
+Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which the consensus evaluated at~$x$, $F(p_1,\ldots,p_n)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$, is a log-linear opinion pool with unit sum weights.
+
+\subsection{Heterogeneous log-linear pooling}
+\label{app:log_pool_hetero}
+
+Log-linear pooling may not be ideal if agent have {\em selective} opinions in the sense that they respond more to certain outcomes  than to others.
+For instance, some agent may produce a sharply peaked distribution only when the target variable is in a certain state but a flat distribution otherwise, and some other agents may selectively respond to other states.
+In such situation, it is natural to try and weight agents depending on outcomes, so that $\lambda_i$ becomes a function of~$x$.
+
+As a first try, let us consider log-linear pooling in this scenario:
+$$
+p(x)\propto \pi(x) \prod_i p_i(x)^{\lambda_i(x)},
+$$
+where $\lambda_i : {\cal X}\to\mathbb{R}$ is now a weighting function. Unfortunately, this more general operator does not satisfy hypothesis reduction invariance. The reason
+
 It is easily seen that, if the weights sum up to one, log-linear pooling is externally Bayesian in the sense that, for any nonvanishing positive valued function $q:{\cal X}\to \mathbb{R}^\star_+$, opinion pooling and opinion updating commute:
@@
 
-Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$, is of the above form with unit sum weights and is therefore a log-linear opinion pool.
+Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which the consensus evaluated at~$x$, $F(p_1,\ldots,p_n)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$, is of the above form with unit sum weights and is therefore a log-linear opinion pool.
 
-A question that arises naturally is whether it is possible to weight opinions depending on targets, so that $\lambda_i(x)$ becomes a function of~$x$.
-This would make it possible, for instance, to consider an agent reliable when expressing an opinion about a certain hypothesis~$x_1$, but unreliable when it comes to another hypothesis~$x_2$.
-As a first try, let us consider a straightforward extension of log-linear pooling:
+Log-linear pooling may not be ideal if agent opinions are {\em selective}.
+Assume, for instance, that some agent produces a sharply peaked distribution only when the target variable is in a certain state but a flat distribution otherwise, and that some other agents selectively respond to other states.
+In such situation where opinions reflect some outcomes better than others, it would be natural to try and weight agents depending on outcomes, so that $\lambda_i$ becomes a function of~$x$.
+
+As a first try, let us consider log-linear pooling in this scenario:
 $$
@@
 $$
-where $\lambda_i : {\cal X}\to\mathbb{R}$ is some weighting function. Unfortunately, this operator is generally not externally Bayesian because $F[N(q p_1), \ldots, N(q p_n)](x)
-\propto r(x) F(p_1,\ldots, p_n)(x)$ with:
+where $\lambda_i : {\cal X}\to\mathbb{R}$ is now a weighting function. Unfortunately, this more general operator is not externally Bayesian:
 $$
-r(x) = q(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)},
-\qquad 
-z_i = \sum_{x\in{\cal X}} q(x) p_i(x),
+F[N(q p_1), \ldots, N(q p_n)](x)
+\propto
+\underbrace{q(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)}}_{r(x)}
+F(p_1,\ldots, p_n)(x),
 $$
-and $r(x)$ cannot be made proportional to $q(x)$, except with constant weights, because the normalizing factors~$z_i$ may differ from one agent to the other.
+with:
+$$
+z_i \equiv \sum_{x\in{\cal X}} q(x) p_i(x),
+$$
+and the only way to make $r(x)$ proportional to $q(x)$ for any $q, p_1,\ldots, p_n$ is to impose constant weights, because the normalizing factors~$z_i$ may differ from one agent to the other.
 
@@
 
+However, log-linear pooling is not appropriate in situations where some agent opinions are not expressed on the whole set~${\cal X}$ of possibilities.
+Assume that the opinion~$p_i$ of agent~$i$ is only defined on a subset of~${\cal X}$: it could trivially be extended to~${\cal X}$ by assigning zero probability to missing outcomes, however that would mean that the probabilities provided by the agent are either underestimated or overestimated, and therefore unreliable.
+
+In such situation, agent opinions somehow need normalization...
+
+In such situation, it should be possible to weight opinions depending on target values, so that $\lambda_i$ becomes a function of~$x$.
+
+have unreliable opinions about certain hypotheses. Assume, for instance, that an agent
+
+From this point on, only crap.
+
+Ponderer les agents en fonction de $x$, c'est comme considerer des mini-agents ayant chacun une mini-opinion. Mais alors comment représenter ces mini-opinions?
+
+Une option est de choisir une fonction constante \'egale \`a $1$ sauf pour la cible particuliere du mini-agent, ou ca vaut $p_i(x)/p_i(x_0)$.
+Et là on fait du log-linear pooling.
+Sauf que ce n'est pas sur les opinions 'originelles'.
+Au passage, les mini-opinions n'ont pas besoin d'être normalisés.
+
+Du coup, on se ramene au log-linear pooling. La question est de savoir quelles strategies de construction de mini-agents sont acceptables. Opinions -> mini-opinions -> pooling.
+
+Par exemple, si au lieu de prendre le rapport de proba, je prends juste $p_i(x)$ tout en laissant la ligne de base à 1, que se passe-t-il?
+On sent bien que c'est FOIREUX mais pourquoi?
+Ben, le rapport de proba n'est pas conserv\'e.
+Vu que la ligne de base est arbitraire, il faut d'une maniere ou d'une autre normaliser la valeur cible.
+
+Le mini-agent doit refleter un minimum l'agent dont il est issu. Conserver au moins un rapport de proba est une facon de garantir une certaine fidelite.
+
+Mais pourquoi conserver au moins un rapport?
+
 %\bibliographystyle{ieeetr}

### 2019-04-17 papers/composite_inference/draft.tex
ArXiv submission

--- 2019-04-12 papers/composite_inference/draft.tex
@@
 
+\newcommand{\isep}{\mathrel{{.}\,{.}}\nobreak}
+
 \begin{document}
@@
 \begin{abstract}
-We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents.
-The resulting class of predictive models, which we call composite Bayesian models, is a middle way between generative and discriminative models.
+We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents, thereby defining a class of predictive models which we call composite Bayesian.
 This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning; in the latter case, they may be tuned so as to minimize prediction cross-entropy, yielding an easy-to-solve convex problem.
-We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
+We argue that composite Bayesian inference is a middle way between generative and discriminative models that trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
 \end{abstract}
@@
 
-In a well-defined universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions.
+In a closed universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions.
 Consider as a straightforward example deciding whether an object is a sauce pan or a frying pan based on depth.
@@
 For instance, in medical diagnosis, automated disease predictions need to be corroborated by individualized findings to be taken into account.
-In other words, there is little clinical value in ``black boxes'', and certainly more in automated methods that can ``justify'' themselves just like human experts.
+In other words, there is little clinical value in ``black boxes'', and certainly more in automated methods that can support their predictions just like human experts.
 One such method that has long been used by clinicians is the familiar reference range analysis, which stems from simple generative models.
-
-%Imagine a doctor telling a patient: ``Sorry, sir, you have Alzheimer's disease because this fancy AI software said so''.
-Unless the said software is 100\% reliable, which is practically impossible, this would obviously be unacceptable.
-Acceptable would be to corroborate the diagnosis with some key quantitative empirical observations: low memory test scores, hippocampal atrophy, etc.
-In other words, inference has to come with some insight.
-An AI system is but one expert relying on hypotheses which may be invalid, as any expert, and is therefore prone to error.
-A typical error source is to train a classifier on a subset of classes from the real world.
-We cannot avoid inference errors but we can safeguard against them by being able to interrogate the system: why do you think what you think?
-To achieve that, the system should be capable of {\em introspection.
-
-%Vapnik: ``one should solve the classification problem directly and never solve a more general problem as an intermediate step''.
-True if the goal is classification only.
-Wrong if introspection is needed because the problem is then more general in nature.
-
-%Human inference does not work in a pre-determined universe of ``causes''.
-We confront theories with reality to evaluate how likely they are, while knowing that none of the theories considered so far may be ``right''.
-In fact, we are not looking for the truth, but for a conving enough theory.
-
-%Composite likelihood seen as a pool of experts comes with a potential introspection mechanism: first sort experts by decreasing influence on the predictive distribution, see how likely the different classes are to the most influential experts.
 
@@
 For the same basic reason, supervised learning in a discriminative model is statistically less efficient than in a generative model spanning the same family of posteriors, hence it requires more training data to achieve optimal performance \cite{Ng-01}.
-
-%Overall, pure discriminative models are of little use outside the context of big labeled data.
 
@@
 
-%Importantly, each agent opinion is justifiable as it is based on a generative model.
-
 When choosing the aggregation method as log-linear pooling, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL), see \cite{Varin-11} and the references therein.
-CL was developed as a surrogate of traditional likelihood for parameter estimation.
-While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} (see Appendix~\ref{app:frequentist} for a basic weak consistency proof), and may offer an excellent trade-off between computational and statistical efficiency in practice.
+CL was developed as a proxy for traditional likelihood in parameter estimation.
+While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotic minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} (see Appendix~\ref{app:frequentist} for a basic weak consistency proof), and may offer an excellent trade-off between computational and statistical efficiency in practice.
 
@@
 
-The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on the target values, leading to a more flexible predictive model that can be trained similarly.
-
-%Technical details are given in the remainder.
+The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on target values, leading to a more flexible predictive model that can be trained similarly to CL.
 
 The reader is warned that slightly abusive mathematical notation is sometimes used in the remainder.
-This is done deliberately for the sake of clarity in places where we think that notation can be lightened without ambiguity.
+This is done deliberately for the sake of clarity in places where we think that notation can be lightened without creating ambiguity.
 
@@
 
-Let $\mathbf{Y}$ an observable multivariate random variable with sampling distribution $p(\y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is assumed to be a finite set for simplicity.
-Given an experimental outcome $\y$, the likelihood is the sampling distribution evaluated at $\y$, seen as a function of $x$:
+Let $\mathbf{Y}$ an observable multivariate random variable with sampling distribution $p(\y|x)$ conditional on some unobserved variable of interest~$X$ taking on values in a set~${\cal X}$ assumed to be finite for simplicity.
+Given an experimental outcome~$\y$, the likelihood is the sampling distribution evaluated at~$\y$, seen as a function of~$x$:
 $$
-L(x) = p(\y|x)
-.
+L(x) = p(\y|x).
 $$
@@
 This requires a plausible generative model which, for complex data, may be out of reach or involve too many nuisance parameters.
-A natural workaround known as data reduction is to extract some lower-dimensional representation $z(\y)\sim f(z|x)$ using a many-to-one mapping, and consider the potentially more convenient likelihood function:
+A natural workaround known as data reduction is to extract some lower-dimensional representation using a many-to-one mapping $z(\y)\sim f(z|x)$, and consider the potentially more convenient likelihood function:
 $$
-\ell(x) = f(z|x)
-.
+\ell(x) = f(z|x).
 $$
 
-Substituting $L(x)$ with $\ell(x)$ boils down to restricting the feature space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information.
+Substituting $L(x)$ with $\ell(x)$ boils down to restricting the feature space, thereby  ``delegating'' inference to an ``agent'' provided with partial information.
 While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction.
@@
 
-If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
+If we see the likelihoods as Bayesian posterior distributions corresponding to uniform priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}.
 A common choice, owing to its simplicity and tendency to produce single peaked distributions, is log-linear pooling\footnote{When $\pi(x)$ is not uniform, it is also called {\em generalized} log-linear pooling, however this terminology could be confusing here since we will present another generalization.}:
@@
 \end{equation} 
-where $\pi(x)$ is a reference distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
-We shall asume for feasibility that the weights are intrinsic to the agents and are therefore independent from the data~$\y$.
-Also, negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less probable than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
+where $\pi(x)$ is a prior distribution and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights.
+An essential property of log-linear pooling, in our opinion, is to be {\em invariant by hypothesis elimination} in the sense that  probability ratios are not altered by restricting the state space~${\cal X}$ (see Appendix~\ref{app:log_pool_standard}).
+Negative weights should be ruled out to warrant a no less expected monotonicity property: if all agents agree that an hypothesis $x_1$ is less probable than an hypothesis $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.
+Also, we shall assume for feasibility that the weights are intrinsic to the agents and are therefore independent from the data~$\y$.
 
@@
 $$
-where the distribution $\pi(x)$ plays the role of a prior, while the quantity:
+where the distribution $\pi(x)$ plays the role of a Bayesian prior, while the quantity:
 \begin{equation}
@@
 \end{equation} 
-plays that of a likelihood function, and happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}.
-The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}.
+plays that of a likelihood function, and happens to be known in computational statistics as {\em marginal} CL \cite{Varin-11}.
+The slightly more general {\em conditional} CL form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}.
 
@@
 CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} method in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$.
-Since log-linear opinion pooling does not assume feature independence, it yields both an alternative interpretation and a generalization of na\"ive Bayes, whereby unitary feature weights may be used by default but do not guarantee optimal prediction performance.
+Since log-linear pooling does not assume feature independence, it yields both a generalization and an alternative interpretation of na\"ive Bayes, whereby unitary feature weights may be used by default but do not guarantee optimal prediction performance.
 
@@
 
-If training data is available, the generative parameters may be tuned in a first stage using standard estimation techniques.
+If training data is available, the generative parameters may be tuned in a first phase using standard estimation techniques.
 Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to CL one of the same techniques as for traditional likelihood \cite{Berger-99}.
@@
 Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one.
-This is motivated by a characterization theorem for opinion pooling \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome, is the log-linear opinion pool with unit sum weights.
-The external Bayesian property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently.
-Log-linear pooling with unit sum weights also turns out to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.
+This is motivated by a characterization theorem \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome, is log-linear pooling constrained to unit sum weights.
+The external Bayesian property is stronger than the above-mentioned invariance by hypothesis elimination; it essentially means that the consensus does not change if external opinions are taken into account before or after combining the agents.
+Log-linear pooling with unit sum weights also turns out to be an  optimal compromise in the sense of the inclusive Kullback-Leibler (KL) divergence \cite{Garg-04}.
 
@@
 
-Because some positive weight constraints may turn out inactive, optimization has a natural tendency to produce sparse weights.
-As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer diagnostic UCI dataset \cite{Wolberg-94}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs radically from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
+Because some positive weight constraints may turn out inactive, optimization naturally tends to produce sparse weights.
+As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer diagnostic UCI dataset \cite{Wolberg-94}, there is no clear relation between learned weights and feature-level discrimination power, showing that maximum likelihood learning differs radically from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.
 
@@
 For instance, class-dependent offsets may be added so as to learn the ``prior'' together with the composite weights.
-It is also possible to perform unconstrained optimization to achieve higher training likelihood, but this means that some composite weights are then negative, in violation of the monotonicity property mentionned in Section~\ref{sec:log_pool}.
+It is also possible to perform unconstrained optimization to achieve higher training likelihood, but this means that some composite weights are then negative, in violation of the monotonicity property mentioned in Section~\ref{sec:log_pool}.
 
@@
 The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized, making the $I$-projection equivalent to maximum conditional entropy \cite{BergerA-96}.
-The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
-This is weaker a constraint than assuming that the feature generative models $f(z_i|x)$ are true: it only assumes that the {\em description length} \cite{Grunwald-07} achieved by each feature model is a truthworthy upper bound.
-The composite likelihood weights may be interpreted as the Lagrange multipliers associated with these description length constraints.
+The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependencies between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature.
+This is weaker a constraint than assuming that the feature generative models $f(z_i|x)$ are true: it only assumes that the {\em description length} \cite{Grunwald-07} achieved by each feature model is a trustworthy upper bound.
+Learned composite likelihood weights may be interpreted as Lagrange multipliers associated with such description length constraints.
 
-An insightful analysis given in \cite{Grunwald-04} shows that $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss.
-Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is {\em inexact} in practice due to the finite training set.
-Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by its underlying inexact knowledge, which in this case may be interpreted in terms of feature description length.
+An insightful analysis given in \cite{Grunwald-04} expands on the fact that $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss.
+Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge extracted from the training data, which is represented by mean-value constraints and is {\em inexact} in practice due to the finite training set.
+Trained composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by its underlying inexact knowledge, which in this case may be interpreted in terms of feature description length.
 
@@
 
-So far, we have assumed that each agent is given a single weight in the opinion aggregation~(\ref{eq:log_pool}).
+So far, we have assumed that each agent is given a fixed weight in the opinion pool~(\ref{eq:log_pool}).
 Let us now consider a more general pooling operator where agents can be weighted depending on the target variable:
@@
 \end{equation}
-where~$x_0$ is an arbitrary fixed target value, which we call the reference, and $\blambda: {\cal X}\to \mathbb{R}_+^n$ is a weighting function here.
-Note that the reference weights $\lambda_i(x_0)$ play no role and can therefore be conventionally assumed to cancel.
-As shown in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make pooling externally Bayesian if the weighting function sums up to one uniformly,
+where~$x_0\in{\cal X}$ is an arbitrary fixed reference and $\blambda: {\cal X}\to \mathbb{R}_+^n$ is now a weighting function.
+Note that the weights associated with the reference, $\lambda_i(x_0)$, play no role and can therefore be conventionally assumed to vanish.
+As shown in Appendix~\ref{app:log_pool_selective}, the main reason for introducing a reference is to maintain pooling invariance by hypothesis elimination.
+Moreover, (\ref{eq:super_pool}) is externally Bayesian if the weighting function sums up to one uniformly,
 $$
@@
 $$
-hence providing a sensible default rule to tune the weights.
-Note that (\ref{eq:super_pool}) amounts to log-linear pooling if the weights are target-independent as the reference is then effectless.
-Using the same analogy with Bayes rule as in Section~\ref{sec:bayes_rule}, we see that the quantity:
+which provides a sensible default rule to tune the weights. Using the same analogy with Bayes rule as in Section~\ref{sec:bayes_rule}, we see that the quantity:
 \begin{equation}
@@
 We call the form~(\ref{eq:super_comp_lik}) super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs.
-SCL further generalizes CL since both forms are seen to be proportional for target-independent weights.
+SCL further generalizes CL since both forms are seen to be proportional in the case of binary classification ($|{\cal X}|=2$) or, generally, if the weights are constant across target values.
 
-Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, $\forall x\in{\cal X}\setminus \{x_0\}$, $\lambda_i(x)=1$ for a single feature~$i$, and zero otherwise), then SCL turns out identical to the class-specific likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument; namely, the {\em PDF projection theorem}, which in fact boils down to an $I$-projection \cite{Minka-04}.
-While our derivation follows a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
+Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, $\forall x\in{\cal X}\setminus \{x_0\}$, $\lambda_i(x)=1$ for a single feature~$i$, and zero otherwise), then SCL turns out identical to the class-specific likelihood proxy derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, {\em PDF projection}, which, in fact, is a type of $I$-projection \cite{Minka-04}.
+While our derivation follows from a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.
 
-A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which happen if the feature generative distributions $f(z_i|x)$ are unknown for some targets~$x$.
+A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which occur if the feature generative distributions $f(z_i|x)$ are unknown for some targets~$x$.
 SCL makes it possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs.
-Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence whatsoever on the predictive distribution.
+Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown shall receive uniformly zero weight so as to have no influence whatsoever on the predictive distribution.
 
@@
 $$
-\forall a\in{\cal X}, \forall i=1,\ldots,n,
+\forall a\in{\cal X}\setminus\{x_0\}, \forall i\in [1\isep n],
 \quad
@@
 \subfigure[Case study: clearly categorized class-3 wine]{
-\includegraphics[width=.9\textwidth]{interp_case1.pdf}}
+\includegraphics[width=\textwidth]{interp_case1.pdf}}
 %\hspace*{.2\textwidth}
 \subfigure[Case study: atypical class-2 wine]{
-\includegraphics[width=.9\textwidth]{interp_case2.pdf}}
+\includegraphics[width=\textwidth]{interp_case2.pdf}}
 \end{center}
@@
 This is illustrated in Figure~\ref{fig:interp} for a composite Bayesian expert fully trained (including weight optimization as in Section~\ref{sec:learning}) on a famous UCI dataset \cite{Aeberhard-92} to classify wines into 3 categories from 13~features.
-The top panel shows how the expert (correctly) classifies a particular wine in category~3 with near $100\%$ probability; it can be seen that the prediction is mainly based on the flavanoids concentration and the color intensity.
+The top panel shows how the expert (correctly) classifies a particular wine in category~3 with near $100\%$ probability; it can be seen that the prediction is mainly based on the flavanoid concentration and the color intensity.
 Since both have typical values for category~3 wines, this is a clear-cut decision.
-In another case depicted in the bottom panel, the wine is (again, correctly) classified in category~2 with $98.9\%$ probability, however two of the main features driving the prediction (alcanality of ash and magnesium concentration) have peculiar values for category~2, while proline concentration quite strongly points to category~1.
+In another case depicted in the bottom panel, the wine is (again, correctly) classified in category~2 with $98.9\%$ probability, however two of the main features driving the prediction (alcanality of ash and magnesium concentration) have peculiar values for category~2, while the proline concentration quite strongly points to category~1.
 Despite high prediction confidence, this case may require further investigation.
 
-We believe that there are many applications of statistical inference in which interpretation is the real goal while prediction performance is only an intermediate requisite.
-A prediction method needs to prove reliable to be relevant; but, once proven reliable, it may need to be interpreted.
+We believe that there are many applications of statistical inference in which interpretation is the real goal while prediction performance is only but an intermediate requisite.
+A prediction method needs to prove reliable to be of any use; but, once proven reliable, it may need to be interpreted.
 Composite Bayesian inference may come into play when the {\em why} question is more important than the {\em what} question.
+
+%Imagine a doctor telling a patient: ``Sorry, sir, you have Alzheimer's disease because this fancy AI software said so''.
+Unless the said software is 100\% reliable, which is practically impossible, this would obviously be unacceptable.
+Acceptable would be to corroborate the diagnosis with some key quantitative empirical observations: low memory test scores, hippocampal atrophy, etc.
+In other words, inference has to come with some insight.
+An AI system is but one expert relying on hypotheses which may be invalid, as any expert, and is therefore prone to error.
+A typical error source is to train a classifier on a subset of classes from the real world.
+We cannot avoid inference errors but we can safeguard against them by being able to interrogate the system: why do you think what you think?
+To achieve that, the system should be capable of {\em introspection.
+
+%Vapnik: ``one should solve the classification problem directly and never solve a more general problem as an intermediate step''.
+True if the goal is classification only.
+Wrong if introspection is needed because the problem is then more general in nature.
+
+%Human inference does not work in a pre-determined universe of ``causes''.
+We confront theories with reality to evaluate how likely they are, while knowing that none of the theories considered so far may be ``right''.
+In fact, we are not looking for the truth, but for a conving enough theory.
+
+%Composite likelihood seen as a pool of experts comes with a potential introspection mechanism: first sort experts by decreasing influence on the predictive distribution, see how likely the different classes are to the most influential experts.
 
@@
 Note that the derivation assumes that the weights~$\blambda$ are independent from the target~$x$, therefore these properties do not necessarily extend to SCL.
+
+\section{Log-linear pooling}
+\label{app:log_pool}
+
+Consider a sequence of probability distributions $p_i(x)$, $i=1,\ldots,n$, representing multiple opinions on a variable of interest~$x\in{\cal X}$.
+The general goal of opinion aggregation is to define an operator $F(p_1,\ldots,p_n)$ that takes opinions as inputs and outputs a single distribution $p(x)$ representing a consensus.
+
+\subsection{Standard log-linear pooling}
+\label{app:log_pool_standard}
+
+Log-linear pooling \cite{Genest-86} is one particular such operator:
+$$
+p(x)\propto \pi(x) \prod_{i=1}^n p_i(x)^{\lambda_i},
+$$
+where $\pi(x)$ is an arbitrary distribution and $\lambda_i\in\mathbb{R}$ are arbitrary weights.
+
+A key property of log-linear pooling is that probability ratios are stable by hypothesis elimination: if the opinions are restricted to any subset ${\cal X}_s\subset {\cal X}$ by simple renormalization, the consensus~$p_s$ on~${\cal X}_s$ is proportional to the consensus~$p$ on~${\cal X}$, {\em i.e.}, $p_s(x)\propto p(x)$ for $x\in{\cal X}_s$.
+In other words, eliminating some hypotheses does not bias inference towards any of the remaining hypotheses.
+
+Another way to look at this invariance property is to state that, for any boolean-valued function $q:{\cal X}\to \{0,1\}$, opinion pooling and opinion updating commute: 
+\begin{equation}
+\label{eq:external_bayesian}
+F[N(q p_1), \ldots, N(q p_n)]
+=
+N[q F(p_1,\ldots, p_n)],
+\end{equation}
+where~$N$ denotes the normalization endomorphism of $\mathbb{R}_+^{\cal X}$:
+$$
+N(f)(x) = \frac{f(x)}{\sum_{x'\in{\cal X}} f(x')}
+.
+$$
+
+Under the more general condition that~(\ref{eq:external_bayesian}) holds for any non-vanishing positive valued function $q:{\cal X}\to\mathbb{R}^\star_+$,  the opinion pool is said to be {\em externally Bayesian}.
+It is easily seen that log-linear pooling is externally Bayesian if the weights~$\lambda_i$ sum up to one.
+Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which the consensus evaluated at~$x$, $F(p_1,\ldots,p_n)(x)$, is a function of $x,p_1(x),\ldots,p_n(x)$, is a log-linear pool with unit sum weights.
+
+\subsection{Generalization: selective pooling}
+\label{app:log_pool_selective}
+
+Log-linear pooling may not be ideal if agent have {\em selective} opinions in the sense that they respond more to certain target outcomes than to others.
+For instance, some agent may produce a sharply peaked distribution when the target is in a certain state but a flat distribution otherwise, and some other agents may selectively respond to other states.
+In such situation, a stronger consensus might be reached by weighting agents depending on the target.
+
+If we extend log-linear pooling by letting $\lambda_i : {\cal X}\to\mathbb{R}$ become a function of~$x$, the resulting operator,
+$$
+p(x)\propto \pi(x) \prod_i p_i(x)^{\lambda_i(x)},
+$$
+is in general not invariant by hypothesis elimination.
+For a subset ${\cal X}_s\subset {\cal X}$, let $p_s$ denote the consensus from opinions restricted to ${\cal X}_s$.
+We have, for any $x\in{\cal X}_s$,
+$$
+p_s(x)
+\propto
+p(x)
+\prod_i z_i^{-\lambda_i(x)},
+\qquad \text{with} \quad
+z_i \equiv \sum_{x\in{\cal X}_s} p_i(x),
+$$
+hence the only way to have $p_s(x)$ proportional to $p(x)$ for any~${\cal X}_s$ is to impose constant weights, because the normalizing factors~$z_i$ may differ from one agent to the other.
+
+Consider, however, the following alternative:
+\begin{equation}
+\label{eq:hetero_log_pool}
+p(x) \propto \pi(x) \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)},    
+\end{equation}
+where~$x_0\in{\cal X}$ is some arbitrary reference value.
+Using a reference does the trick because, for any restriction~${\cal X}_s$, the odds $p_i(x)/p_i(x_0)$ are left unchanged by renormalization since the normalizing factors are the same at both the numerator and the numerator, so they cancel out.
+Moreover, it is easily seen that~(\ref{eq:hetero_log_pool}) is externally Bayesian whenever the weighting function verifies:
+$$
+\forall x \in {\cal X}\setminus \{x_0\},
+\qquad
+\sum_i \lambda_i(x) = 1.
+$$
+
+Clearly, (\ref{eq:hetero_log_pool}) includes log-linear pooling as a special case since the reference has no effect for a constant weighting function $\lambda_i(x)\equiv \lambda_i$.
+This ``selective'' generalization may be interpreted as a two-stage opinion aggregation process: first, binary classifiers operating on half overlapping sets $\{x_0,x\}$ are formed by standard log-linear pooling; second, a multi-class classifier is derived from the binary classifiers assuming preservation of odds relative to the ``pivotal'' class~$x_0$.
+
+%However, log-linear pooling is not appropriate in situations where some agent opinions are not expressed on the whole set~${\cal X}$ of possibilities.
+Assume that the opinion~$p_i$ of agent~$i$ is only defined on a subset of~${\cal X}$: it could trivially be extended to~${\cal X}$ by assigning zero probability to missing outcomes, however that would mean that the probabilities provided by the agent are either underestimated or overestimated, and therefore unreliable.
 
@@
 \qquad
-\nabla \nabla^\top a(\blambda,\y) = {\rm Var}_{\blambda}(\bell),
+\nabla \nabla^\top a(\blambda,\y) = \text{Var}_{\blambda}(\bell),
 $$
-where $\E_{\blambda}$ and ${\rm Var}_{\blambda}$ respectively denote the expectation and variance with respect to $p_\blambda(x|\y)\propto \pi(x)\exp[\blambda^\top \bell(x,\y)]$ at fixed~$\y$.
-This shows that $a(\blambda,\y)$ is convex in $\blambda$ since the variance matrix is positive, which, in turn, proves the concavity of~$\psi$.
+where $\E_{\blambda}$ and $\text{Var}_{\blambda}$ respectively denote the expectation and variance with respect to $p_\blambda(x|\y)\propto \pi(x)\exp[\blambda^\top \bell(x,\y)]$ at fixed~$\y$.
+This shows that $a(\blambda,\y)$ is convex in $\blambda$ since any variance matrix is positive, which, in turn, proves the concavity of~$\psi$.
 
-\section{Heterogeneous log-linear pooling}
-\label{app:hetero}
-
-Consider a sequence of probability distributions $p_i(x)$, $i=1,\ldots,n$, representing multiple opinions on a variable $x\in{\cal X}$.
-The general goal of opinion aggregation is to define an operator $F(p_1,\ldots,p_n)$ that takes opinions as inputs and outputs a single distribution $p(x)$ representing a consensus.
-Log-linear pooling \cite{Genest-86} is one particular such operator:
-$$
-p(x)\propto \pi(x) \prod_{i=1}^n p_i(x)^{\lambda_i},
-$$
-where $\pi(x)$ is an arbitrary distribution and $\lambda_i\in\mathbb{R}$ are arbitrary weights.
-It is easily seen that, if the weights sum up to one, log-linear pooling is externally Bayesian in the sense that, for any nonvanishing positive valued function $q:{\cal X}\to \mathbb{R}^\star_+$, opinion pooling and opinion updating commute:
-$$
-F[N(q p_1), \ldots, N(q p_n)]
-=
-N[q F(p_1,\ldots, p_n)],
-$$
-where~$N$ denotes the normalization endomorphism of $\mathbb{R}_+^{\cal X}$:
-$$
-N(f)(x) = \frac{f(x)}{\sum_{x'\in{\cal X}} f(x')}
-.
-$$
-
-Conversely, \cite{Genest-86b} proved that any externally Bayesian operator for which $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$, is of the above form with unit sum weights and is therefore a log-linear opinion pool.
-
-A question that arises naturally is whether it is possible to weight opinions depending on targets, so that $\lambda_i(x)$ becomes a function of~$x$.
-This would make it possible, for instance, to consider an agent reliable when expressing an opinion about a certain hypothesis~$x_1$, but unreliable when it comes to another hypothesis~$x_2$.
-As a first try, let us consider a straightforward extension of log-linear pooling:
-$$
-p(x)\propto \pi(x) \prod_i p_i(x)^{\lambda_i(x)},
-$$
-where $\lambda_i : {\cal X}\to\mathbb{R}$ is some weighting function. Unfortunately, this operator is generally not externally Bayesian because $F[N(q p_1), \ldots, N(q p_n)](x)
-\propto r(x) F(p_1,\ldots, p_n)(x)$ with:
-$$
-r(x) = q(x)^{\sum_i \lambda_i(x)} \prod_i z_i^{-\lambda_i(x)},
-\qquad 
-z_i = \sum_{x\in{\cal X}} q(x) p_i(x),
-$$
-and $r(x)$ cannot be made proportional to $q(x)$, except with constant weights, because the normalizing factors~$z_i$ may differ from one agent to the other.
-
-Consider, however, the following alternative:
-\begin{equation}
-\label{eq:hetero_log_pool}
-p(x) \propto \pi(x) \prod_i \left[\frac{p_i(x)}{p_i(x_0)}\right]^{\lambda_i(x)},    
-\end{equation}
-where $x_0$ is some arbitrary reference value.
-This does the trick because, for any positive function $q(x)$, the ratio $q(x)p_i(x)/q(x_0)p_i(x_0)$ is left unchanged by normalization of $q p_i$ since the normalizing factors are the same at both the numerator and the numerator, hence they cancel out, yielding:
-$$
-F[N(q p_1), \ldots, N(q p_n)](x)
-\propto
-\left[\frac{q(x)}{q(x_0)}\right]^{\sum_i \lambda_i(x)}
-F(p_1, \ldots, p_n)(x),
-$$
-which shows that~(\ref{eq:hetero_log_pool}) is externally Bayesian whenever the weighting function verifies:
-$$
-\forall x \in {\cal X}\setminus \{x_0\},
-\qquad
-\sum_i \lambda_i(x) = 1.
-$$
-
-The operator~(\ref{eq:hetero_log_pool}) generalizes standard log-linear pooling and has the property that the consensus $F(p)(x)$ is a function of $x,p_1(x),\ldots,p_n(x)$ but also $p_1(x_0),\ldots,p_n(x_0)$.
-The ``trick'' to enable heterogeneous (target-dependent) weights is essentially to consider the odds relative to a fixed reference.
-
-Caca.
-$$
-\log p(x) \equiv \mu(x) + \sum_i \lambda_i(x) \log p_i(x)
-$$
-
-If we look for a pool of the form:
-$$
-\log p(x) \equiv \mu(x) + \sum_i \lambda_i(x) \log p_i(x)
-$$
-
-%\bibliographystyle{ieeetr}
 \bibliographystyle{abbrv}

### 2020-01-13 cv/cv.tex
Nouveau texte pour chanson de Tom

@@
 
-\newcommand{\bighead}[3]{\noindent\begin{minipage}{.85\textwidth}{\bf #1}\newline {\bf #2}\end{minipage}\begin{minipage}{.15\textwidth}\begin{flushright}{\bf #3}\end{flushright}\end{minipage}\newline\noindent\rule{.85\textwidth}{0.4pt}}
+\newcommand{\bighead}[3]{\noindent\begin{minipage}{.85\textwidth}{\textbf{#1}}\newline {\textbf{#2}}\end{minipage}\begin{minipage}{.15\textwidth}\begin{flushright}{\textbf{#3}}\end{flushright}\end{minipage}\newline\noindent\rule{.85\textwidth}{0.4pt}}
 
-\newcommand{\head}[2]{{\noindent {\bf #1} $\mid$ {\bf #2}}}
+\newcommand{\head}[2]{{\noindent {\textbf{#1}} $\mid$ {\textbf{#2}}}}
 
@@
 \begin{flushright} 
-{\huge {\bf Alexis Roche}}\\
+{\huge {\textbf{Alexis Roche}}}\\
 \begin{small}
@@
 
-\noindent Senior-level engineer and applied researcher with a specialist expertise in computer vision, medical image analysis and machine learning, operating in the fields of applied research, health industry and domestic appliance industry.
-Graduated from the French ``Grande \'Ecole" CentraleSupelec, and holder of a PhD.
+\noindent Senior-level engineer and applied researcher with a specialist expertise in computer vision, medical image analysis and machine learning, operating in the fields of applied research, healthcare, electro-domestic and entertainment industry.
+Graduated from the French ``Grande \'Ecole" CentraleSupelec (formely, Centrale Paris), and holder of a PhD.
 Strong experience in developing research and commercial software.
-
-\noindent {\em Keywords:} image processing, statistics, machine learning, structural and functional brain magnetic resonance imaging, Python.
+{\em Keywords:} artificial intelligence, image processing, computer vision, machine learning, deep learning, statistics, brain imaging, Python.
 
@@
 
-\noindent {\bf DATA SCIENCE:} In-depth knowledge of statistics, computer vision, medical image analysis, traditional machine learning and deep learning.
+\noindent \textbf{DATA SCIENCE:} In-depth knowledge of statistics, computer vision, medical image analysis, traditional machine learning and deep learning.
 
 \ \\
-\noindent {\bf NEUROIMAGING:} Extensive experience in developing algorithms for structural and functional brain magnetic resonance imaging (MRI) in the form of open-source and commercial software for neuroscience, clinical research \& clinical practice.
+\noindent \textbf{NEUROIMAGING:} Extensive experience in developing algorithms for structural and functional brain magnetic resonance imaging (MRI) in the form of open-source and commercial software for neuroscience, clinical research \& clinical practice.
 
 \ \\
-\noindent {\bf SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.
+\noindent \textbf{SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.
 
 \ \\
-\noindent {\bf INTERACTION:} Proven ability to work in multidisciplinary environments at the interface of industry and academic research, as well as to develop successful collaborations with physicians and fundamental life science researchers.
+\noindent \textbf{INTERACTION:} Proven ability to work in multidisciplinary environments at the interface of industry and academic research, as well as to develop successful collaborations with physicians and fundamental life science researchers.
 
 \ \\
-\noindent {\bf SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 international patents.
+\noindent \textbf{SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 international patents.
 
@@
 
-\bighead{SENIOR COMPUTER VISION SCIENTIST}{CoVii, Ar\c celik/Beko group, Porto, Portugal}{2017--present}
+\bighead{SENIOR COMPUTER VISION ENGINEER}{Didimo, Porto, Portugal}{2019--present}
+\nopagebreak
 \begin{itemize}
-\item Developing object recognition and tracking algorithms for smart domestic appliances (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017).
+\item Developing computer vision algorithms for high-fidelity 3D avatars.
+\end{itemize}
+\ \\
+
+\bighead{SENIOR COMPUTER VISION SCIENTIST}{CoVii, Ar\c celik/Beko group, Porto, Portugal}{2017--2019}
+\nopagebreak
+\begin{itemize}
+\item Developed embedded object recognition and tracking algorithms for smart domestic appliances using traditional computer vision and deep learning (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017--18).
 \end{itemize}
@@
 \bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
+\nopagebreak
 \begin{itemize}
@@
 \bighead{PERMANENT RESEARCHER -- NEUROIMAGING INSTITUTE}{French Atomic Commission (CEA), Paris, France}{2002--2011}
+\nopagebreak
 \begin{itemize}
@@
 \bighead{RESEARCH ASSISTANT -- WOLFSON MEDICAL VISION LABORATORY}{University of Oxford, UK}{2001--2002}
+\nopagebreak
 \begin{itemize}
@@
 \bighead{CONTINGENT SCIENTIST -- FRENCH NATIONAL SERVICE}{General Directorate for Armament (DGA), Vernon, France}{1996--1997}
+\nopagebreak
 \begin{itemize}
@@
 \heading{Education}
-
+\nopagebreak
 \head{PhD, Engineering Science (1997--2001)}{University of Nice-Sophia Antipolis, France}
@@
 \head{MSc, Cognitive Science (1995--1996)}{University Pierre \& Marie Curie, Paris \rom{6}, France}
+\nopagebreak
 \begin{itemize}
@@
 \head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris (now CentraleSupelec), France}
+\nopagebreak
 \begin{itemize}
@@
 \heading{Additional information}
-
+\nopagebreak
 \noindent 
 \begin{minipage}{.15\textwidth}
-{\bf Languages:}
+\textbf{Languages:}
 \end{minipage}
@@
 \begin{minipage}{.15\textwidth}
-{\bf Programming:}
+\textbf{Programming:}
 \ \\ 
@@
 {\tiny$\bullet$} C/C++ \\
-{\tiny$\bullet$} Python (numpy, scipy, pylab, sklearn, TensorFlow, keras, nipy, cython)\\ 
+{\tiny$\bullet$} Python (numpy, scipy, pylab, skimage, opencv, sklearn, TensorFlow, keras, cython...)\\ 
 {\tiny$\bullet$} Matlab\\
@@
 \begin{minipage}{.15\textwidth}
-{\bf Publications:}
+\textbf{Publications:}
+\ \\ 
 \ \\
@@
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} Over 100 scholarly publications. \\
-{\tiny$\bullet$} Google scholar statistics: 9146 citations, h-index: 31, i10-index: 57 (December 2018).\\
-{\tiny$\bullet$} International patents: 4 published + 1 pending application \\
+{\tiny$\bullet$} 115 scholarly publications.\\
+{\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
+{\tiny$\bullet$} Google scholar statistics: 5754 citations, h-index: 32, i10-index: 56 (September 2019).\\
+{\tiny$\bullet$} 5 published international patents. \\
 \end{minipage}
@@
 %\begin{minipage}{.15\textwidth}
-%{\bf Reviewing:}
+%\textbf{ Reviewing:}
 %\ \\ 
@@
 \begin{minipage}{.15\textwidth}
-{\bf Teaching:}
+\textbf{Teaching:}
 \ \\ 
@@
 \begin{minipage}{.15\textwidth}
-{\bf References:}
+\textbf{References:}
 \end{minipage}

### 2020-03-31 notes/global_optimizer/main.tex
Grabbed global optimizer project from overleaf

\documentclass{article}

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,subfigure}
%\usepackage[utf8]{inputenc}

\title{Global optimizer}
\author{Alexis Roche}
%\date{}

\usepackage{natbib}
\usepackage{graphicx}

\begin{document}

\maketitle

\section{Introduction}

We propose to see online learning as an iterative fitting procedure in which each iteration performs an update of the form:
$$
q \leftarrow F(q^{1-\alpha} p^\alpha),
$$
where $F$ is a fitting operator, $q$ is the current objective approximation, $p$ is an empirical target distribution (possibly based on a random mini-batch), and $0\leq \alpha\leq 1$ is a learning rate.

%If $\alpha=1$, this is in fact classical optimization applied to~$h$ (assuming the same one across iterations). This shows that offline learning is just a special case of online learning.

This general iteration may be seen as a stochastic version of message passing for learning, as proposed in my amazing 2016 paper \cite{rr:16}, which treats mini-batches in a deterministic manner and enables factor-wise approximations, but requires heavy memory load. Conceptually, it's pretty much the same story but the stochastic version is much more feasible. In the message passing framework, the target is of the form $p=cf$, with $c=q^{1-\alpha}$ defining the ``context'' and $f=p^\alpha$ defining the ``factor'' to be approximated, hence the whole story is about fitting factors and the learning specifics are brushed under the carpet.

The iteration is doubly approximative: the fitting objective at each iteration is an approximation, and the fitting procedure itself is approximative. How damaging is it? If the objective is exact, then the only thing we need is a good fitting scheme. If the fitting works well, then the only constraint is the limited batch size, which forces us to approximate the objective iteratively.

We shall consider these as {\em separate problems}. Would it hurt to have an exact fitting scheme? It would not, for sure. We will focus here on minimizing (generalized) KL divergence as a fitting strategy:
$$
F_\star(p, q) = \arg\min_q D(p\|q),
$$
hence defining our {\em ideal} fitting operator. If the target is of the form $p=wf$, this may be interpreted as minimizing the local KL divergence $D(wf\|wg)$ along~$g$, so it can be seen as factor approximation in the sense of the local KL divergence. But we may also consider that we are minimizing an adaptive target {\em globally}.

Assume for now that we are dealing with a fixed target~$p$. Ideally, we would set $\alpha=1$ to fit~$p$ as quickly as possible. However, the fit can only be approximated in practice and may be poor if the target is far off the approximating family. A target of the form $q^{1-\alpha}p^\alpha$ is intuitively `easier' to fit if~$q$ lies in the approximating family.

%On the other hand, for a narrow context, VS will pretty much behave like a Taylor expansion. So, if we have to choose between the two options, we could say we are on the safe side using a narrow context, but the reality is that we may get a poor fit in both cases.

%We may assume that the best option is a context that is more or less centered at the target peak and just wide enough to capture relevant variations. If the target is close to the approximation space, then it does not matter much whether~$w$ is narrow or wide. But if it's not, then the question is which scale is relevant to describe it. The global KL fit would certainly help. Problem is: it's exactly what we are looking for. Chicken and egg problem.

The proposed scheme could be interpreted as gradual `non-gaussianity' or some sort of bridge sampling. Say we start with a very small~$\alpha$ and some prior guess~$q=\pi$. Since~$\alpha$ is small, our target $\pi^{1-\alpha}p^\alpha$ at first iteration is almost Gaussian. We get an update~$q$ that is just a little skewed towards~$p$:
$$
q \approx \pi^{1-\alpha} p^\alpha
$$

Now, if we update the target according to $p=q^{1-\alpha}p^\alpha$ and re-fit it, the new fit is a little bit more skewed towards~$p$. Roughly,
$$
q^{1-\alpha} p^\alpha 
\approx \pi^{(1-\alpha)^2} p^{(1-\alpha)\alpha + \alpha}
= \pi^{1-\beta} p^{\beta},
$$
with $\beta=\alpha+(1-\alpha)\alpha$ evaluating the larger weight of~$p$ relatively to the prior~$\pi$ in the new target, $\beta=\alpha+(1-\alpha)\alpha\approx 2\alpha$.

%The updated~$q$ is a natural choice for the context since it is, in principle, a closer fit to~$p$ than the prior.

So the idea is pretty simple: pick a fixed value~$\alpha>0$ and iterate the process of fitting $q^{1-\alpha}p^\alpha$. After several (many?) iterations, we hope that the target closely resembles~$p$. Is that naive?

What can go wrong? Well, if we use too large a value for~$\alpha$, this clearly won't work... So one first question is: how do we make sure that~$\alpha$ is small enough?

\section{Consistency}

Let~$p$ a target distribution and $Q$ an approximating family. A sufficient condition for the KL fit (in an exponential family of the type considered in the seminal paper) to have finite integral is obviously that~$p$ have a finite integral because $\int q_\star = \int p$. Moreover, if~$p$ has a finite integral, the same goes for $wp$ for any bounded function~$w$, therefore the contextual KL fit will always satisfy $\int q < \infty$. This implies that, if the target distribution is proper, the iterative KL minimization considered here yields a proper distribution.

\section{Fixed point}

Assume that~$Q$ is an exponential family $\exp(\theta^\top \phi)$. The fixed point~$q_*$ of our algorithm is such that:
$$
q_* = \arg\min_q D(q_*^{1-\alpha} p^\alpha \| q),
$$
therefore it satisfies the moment condition
$$
\int q_*^{1-\alpha}p^\alpha\phi 
= \int q_* \phi
$$

Consider the $\alpha$-divergence:
$$
D_\alpha(p\|q) = 
\frac{1}{\alpha(1-\alpha)}
\int \alpha p + (1-\alpha) q - p^\alpha q^{1-\alpha} 
$$

Consider minimizing it on~$Q$. Whenever $\alpha\not=0,1$, the first order optimality condition reads:
\begin{eqnarray*}
\alpha(1-\alpha) \frac{\partial D_\alpha}{\partial \theta} 
& = & \int (1-\alpha) q \phi - p^\alpha (1-\alpha) q^{1-\alpha} \phi \\
& = & (1-\alpha) \int (q - p^\alpha q^{1-\alpha}) \phi \\
& = & 0
\end{eqnarray*}

Incredible!!! I am realizing now that Minka \cite{Minka-05} noted this result before, but maybe not the implication: it means that the $\alpha$-divergence optimality condition is exactly the same as for the iterative scheme considered here using a {\em fixed target}, call it simple EP or whatever you like.  Since $D_\alpha$ is convex, this means that, whenever simple EP converges, it converges to the unique minimizer of $D_\alpha(p\|q)$ on~$Q$.  But how comes Minka or \cite{Dehaene-18} did not include this result in their convergence analysis of EP or AEP? Because they were using EP or AEP with variable targets in a data partitioning perspective.

The remaining issue is convergence. Basically, AEP boils down to iterated $M$-projections:
\begin{equation}
\label{eq:iter_m_proj}
q \leftarrow \arg\min_{q'} D(q^{1-\alpha}p^\alpha\|q').
\end{equation}

Iterated projection is a typical algorithm to find the intersection point between two sets, but AEP is not quite the same thing. There is only one projection set here and, at each iteration, the current projection is slightly pushed away before re-projecting.

We know that, if convergence occurs, then AEP minimizes $D_\alpha(p\|q)$, but is AEP guaranteed to converge? It would help to have a monotonicity property, {\em i.e.}, the update~(\ref{eq:iter_m_proj}) always decreases $D_\alpha(p\|q)$. Is it true? Minka \cite{Minka-05} does not seem to think so.

In practice, anyway we can only approximate the $M$-projection, so there might be no theoretical convergence guarantee. It may thus be wise to try and modify the iterative scheme so as to force convergence. A similar situation occurs in gradient descent schemes, which are not guaranteed to converge unless the step size parameter is tuned on each iteration so as to make the objective decrease.

Can we force convergence? Well, we have a fixed point iteration:
$$
\theta_{n+1} = f(\theta_n),
$$
with:
$$
f(\theta) = \arg\min_{\theta'} \tilde{D}(q_\theta^{1-\alpha}p^\alpha\|q_{\theta'}),
$$
where $\tilde{D}$ is some tractable KL divergence proxy. The function $f(\theta)$ is fully determined by~$p$ and~$\alpha$. The update may be generalized using a damping factor or ``learning rate'' $0<\epsilon_n\leq 1$:
$$
\theta_{n+1} = (1-\epsilon_n)\theta_n + \epsilon_n f(\theta_n).
$$

% We can always tune $\epsilon_n$ so that $\|\theta_{n+1}-\theta_n\|\leq \|\theta_n-\theta_{n-1}\|$, thereby ensuring that $\theta_n$ is a Cauchy sequence.

In practice, a small constant value for~$\epsilon$ might do. In a safer version of the algorithm, we may tune $\epsilon_n$ on each iteration so as to guarantee that some $\alpha$-divergence proxy decreases.

\section{AEP as relaxed KL~minimization}

The fact that fixed-target AEP is a convex minimization algorithm is certainly an interesting result that qualifies it as a relaxation method, which replaces the KL divergence objective with another $\alpha$-divergence for $0<\alpha<1$, corresponding to an {\em exclusive} and {\em non-zero forcing} divergence.

Why performing such a relaxation? Because the original problem is intractable. That is the only valid reason. By adopting this relaxation, we assume that we are able to minimize the KL divergence at least approximately when the target is `close enough' to the approximating family. So the existence of an approximate KL minimization scheme is at the heart of this strategy.

What are the candidates? 
\begin{enumerate}
\item Laplace method.
\item Approximate integration (remarking the equivalence between KL min and moment matching), so anything from the quadrature and Monte Carlo literature.
\item Variational sampling. (We could include in this category approaches such as Bayesian Monte Carlo, i.e. VS variants that make use of other costs than the empirical KL divergence, but such methods do not make much sense in the context of approximate KL minimization, do they?). 
\end{enumerate}

We could dismiss approximate integration from the very beginning based on a consistency argument: the fact that they are inexact even if the target lies in the approximating family, unlike both Laplace and VS. So, we have two remaining candidates.

What are the compared pros and cons of Laplace and VS? I can think about the following:
\begin{itemize}
\item[$-$] Laplace is potentially more efficient if the gradient and hessian have closed-form expressions.
\item[$+$] Laplace is local: it won't work well if the target is `rugged' or skewed, while VS is arguably more robust to such features. 
\item[$+$] Laplace does not handle singularities unlike VS (think about a... Laplace distribution, for which the Hessian does not exist at the mode).
\item[$+$] Laplace can output non-positive hessians, while VS can be constrained not to.
\end{itemize}

To put it in Christopher Bishop's terms: ``The most serious limitation of the Laplace framework, however, is that it is based purely on the aspects of the true distribution at a specific value of the variable, and so can fail to capture important global properties.'' In short, Laplace is potentially faster but less accurate than VS. At least, this is what I could have shown experimentally in my introducing paper, but I decided instead to consider VS as a moment matching method... how clever.

%We can't do it when it is far off, but can do it if is close enough -- how weird does that sound? I don't know. Arguably, it is easier to minimize the $\alpha$-divergence for small~$\alpha$ than for $\alpha\approx 1$. I am not even talking about $\alpha\leq 0$ or $\alpha > 1$, fuck it.

\section{Bridge sampling scheme}

Consider the series:
$$
\alpha_{n+1}=\alpha_n+(1-\alpha_n)\alpha_n=(2-\alpha_n)\alpha_n,
\qquad
0 < \alpha_0 < 1 .
$$

Clearly, $\alpha_n$ is increasing, upper bounded and therefore converges to a limit satisfying $\ell=(2-\ell)\ell$, hence $\ell=1$ cause $\ell=0$ is impossible if $\alpha_0>0$.

Let $\beta=(2-\alpha)\alpha$. We note that: $1-\beta=1-\alpha - \alpha(1-\alpha)=(1-\alpha)^2$.

\section{Laplace approximation}

Conjecture: if~$w$ approaches a Dirac distribution, the KL fit is the Laplace approximation, {\em i.e.} the fit derived from the Taylor expansion at the cavity center. Is that true? Sounds obvious. What else could it be?

If it's true, then we can claim that gradient descent and Newton method are just limiting cases of local KL minimization. If it's not, then KL minimization might raise some suspicion.

\section{Gradient descent}

One simple and classical way to perform fitting on each iteration is via a second-order Taylor expansion of the target logarithm:
$$
F = \log t = (1-\alpha)\log q + \alpha \log p.
$$

In the `quick Laplace' approach \cite{rr:16}, we perform the expansion at the cavity center~$a$, which verifies $\nabla \log q(a)=0$, hence:
$$
\nabla F(a) = \alpha g_p(a),
\qquad
\nabla\nabla^\top F(a) = (1-\alpha)H_q + \alpha H_p(a)
$$

Assuming that~$H$ is positive definite, the value that minimizes the local quadratic approximation is $b = a - H^{-1}g$, hence the update formulas:
$$
H \leftarrow (1-\alpha)H + \alpha H_p(a),
\qquad
a \leftarrow a - \alpha H ^{-1}g_p(a).
$$

If $\alpha=1$, this is the famous Newton algorithm. Using a learning rate $\alpha < 1$ may be useful even if~$p$ is a fixed function to stabilize the algorithm. For instance, if we start far away from the solution, the initial hessian may be off, so it makes sense to regularize it starting, e.g., with a positive scalar matrix for~$H$. Weirdly enough, this somewhat natural strategy does not seem to be much used in practice.

If the target is truly Gaussian, this will clearly delay convergence. However, if it is not even log-concave, this stabilization scheme may help avoiding convergence to saddle points. Instead of seeing~$\alpha$ as a learning rate, we may see it as a `prudence' parameter. Prudence is required for complex targets.

\section{Variational sampling}

What is VS? It is a numerical method to approximately solve the local KL problem. Canonical VS uses~$w$ as a sampling distribution. It's possible to distinguish the context from the sampling distributions, as I did in my original presentation \cite{ijasp:13}, but the price to pay is the potential numerical instability of the underlying importance weights. Besides, we can always fall back to the canonical version by pre-conditioning the target, therefore we shall restrict ourselves to canonical VS in the sequel.

Some basic results. Surrogate cost:
$$
\sum_k 
w_k \left[
p(x_k) \log \frac{p(x_k)}{q(x_k)} + q(x_k) - p(x_k)
\right],
$$
where $w_k$ are the weights associated with a quadrature rule for $\int f(x)dw$.

For random independent sampling from $w(x)$, we have $w_k\equiv 1/n$ and the error on the local target integral $z_w(p)= \int wp$, which represents the ideal approximation peak, is asymptotically Gaussian with zero mean and variance:
$$
v = \frac{1}{n} \int w(p-q_w)^2 ,
$$
where $q_w\equiv\arg\min_q D_w(p,q)$.

Intuitively, the tighter~$w$, the closer~$q_w$ approximates~$p$ on the support of~$w$, therefore the smaller the variance. However, the tighter~$w$, the more loosely~$q_w$ approximates~$q_\star$, the global KL fit, hence the larger the bias. So, we have a bias/variance tradeoff issue.

To limit the bias, we need a cavity that is wide and centered as much as possible around the actual peak. But wide means high variance. Moreover, how do we guarantee that we are centered on the actual peak? How do we go about that?

Canonical form. Minimize $D(wf\|wg)$ or, equivalently,
$$
C(g) = \int w(-f \log g + g),
$$
along functions of the form:
$$
g_\theta(x) = e^{\theta^\top \phi(x)},
$$
yielding:
$$
\nabla C (\theta) = \int w (g_\theta-f) \phi,
\qquad
\nabla \nabla^\top C(\theta) = \int w g_\theta \phi\phi^\top.
$$

In practice, we perform approximate minimization by sampling from~$w$ (so the evaluation points $x_k$ depend on~$w$, please keep in mind):
$$
\hat{C}(g) = \sum_k w_k \left[
-f(x_k) \log g(x_k) + g(x_k)
\right]
$$

Assume:
$$
w(x) = e^{\theta_0^\top \phi(x)},
\qquad
g(x) = e^{\theta^\top \phi(x)},
\qquad
q(x) = e^{(\theta_0+\theta)^\top \phi(x)}
$$

$$
\hat{C}(\theta) = \sum_k w_k \left[
-f(x_k) \theta^\top\phi(x_k) + e^{\theta^\top \phi(x_k)}
\right]
$$

$$
\nabla \hat{C}(\theta) 
= \sum_k w_k 
[g(x_k) - f(x_k)] \phi(x_k)
$$

$$
\nabla\nabla^\top \hat{C}(\theta) 
= \sum_k w_k 
g(x_k) \phi(x_k)\phi(x_k)^\top
$$

We may use a standard optimization algorithm such as L-BFGS to solve this. It could be quite costly in memory, though, cause we need to store $\theta_0$, $\theta$, as well as all the sampling points. But with big computers, we should be able to do this!

A sensible starting value for~$\theta$ could be derived assuming a constant fit: $g(x)\equiv c$, for which the optimal~$c$ is seen to be:
$$
c = \frac{\sum_k w_k f(x_k)}{\sum_k w_k}
$$

The optimizer outputs a value~$\theta$, that needs to be added to the current~$\theta_0$ to produce the new parameter representation of~$q$. If we want to run another iteration, we infer the new~$\theta_0$ by scaling~$q$ by a factor $1-\alpha$ and run the same process.

What I am sketching here is an iterative scheme encompassing VS. Instead of trying to minimize $D(p\|q)$ in a single shot, we iteratively minimize divergences of the form $D(\tilde{p}\|q)$ where $\tilde{p}$ itself is an adaptive approximation to~$p$. This is a way of solving the chicken and egg problem underlying~VS.

We could call this fractionated VS or bridge VS but, this algorithm turns out to be a special case of average or stochastic expectation propagation (AEP) \cite{Dehaene-18,Li-15}. To see this, write down the target distribution as a straightforward product of factors:
$$
p = \prod_{i=1}^n f_i,
\qquad\text{with}\quad
f_i = p^\frac{1}{n}.
$$

When factors represent independent opinions as in Minka's original work \cite{Minka-05}, it does make sense to use the fully blown EP algorithm, provided that we have enough memory for that, but when factors are hugely redundant, then it is intuitively more appropriate to use the AEP~variant as we clearly don't need an approximation for each factor.

At the end of the day, VS appears once again as a particular distribution fitting method, competing with Laplace approximation and Monte Carlo methods. It is easy to argue that VS has the potential to work around local convergence unlike Laplace, which gets easily trapped in saddle points for `rugged' distributions. It is also clear in practice that VS works much better than comparable importance sampling, and the same can be expected for many other MCMC methods since the sampling kernel is not really an issue here -- and we are not mentioning computation time.

So, basically, VS is a strong candidate but, still, my work on VS has attracted little attention, if not zero. There are possibly several reasons to this, including the valid one that the computational complexity of VS appears to be discouraging: $O(d^3)$ using Newton's method \cite{rr:16}. Of course, it is significantly reduced using limited-memory BFGS. I was fairly stupid not to point that out.

In the context where VS is used iteratively to perform $\alpha$-divergence minimization, each iteration approximately minimizes $D(q^{1-\alpha}p^\alpha\|q')$, so we have $w=q^{1-\alpha}$, $f=p^\alpha$ and $q'=q^{1-\alpha}g$. It was mentioned above that some damping may be used to make the iterative process converge by tuning the damping factor~$\epsilon_n$ so that the $\alpha$-divergence proxy decreases. Since the $\alpha$-divergence is not tractable, a proxy is needed. One that is consistent with VS in the sense that it yields the {\em same} first-order condition is:
$$
D_\alpha (p\|q) 
\equiv 
\int q^{1-\alpha} [(1-\alpha)q^\alpha - p^\alpha]
\approx 
(1-\alpha)\langle q^\alpha \rangle_w - \langle p^\alpha \rangle_w
$$

\section{Coordinate-wise KL minimization}

In very high-dimensional problems, it is unfeasible to run VS in a single shot on the whole approximating family. More specifically, VS cannot be run to directly fit $q^{1-\alpha}p^\alpha$ if only because of memory load.

A natural trick to work around this issue is to perform something similar to coordinate block optimization, that is, replace full dimensional factors by coordinate applications. Consider a partition $(x_0,x_1)$ of the coordinate space, and let us assume that we want to optimize the parameters associated with~$x_1$ while holding the others fixed.

Due to the fully factorial nature of the approximating family, the factor approximation verifies $q(x)=q_0(x_0)q_1(x_1)$ (this would not be true for a full Gaussian approximation). Note that this decomposition is not unique.
This separation property, however, only allows for dimension reduction if the tilted target, $q^{1-\alpha}p^\alpha$, is also separable, which only happens if~$p$ is separable and it is seldom the case.

To enforce target separation, we may make our contextual target definition a bit more general by including a dimension reduction mechanism. Let us pick some point~$a$ depending on~$q$ (for instance the mode) and consider an approximation of the form:
$$
t(x) = \gamma q_0(x_0) q_1(x_1)^{1-\alpha}p_1(x_1)^\alpha,
$$
where $p_1(x_1)=p(a_0, x_1)$ is the coordinate application of~$p$ along~$x_1$ at fixed $x_0=a_0$.

The intuition behind this is that, if~$q=q_0q_1$ is a good fit to~$p$, then $\tilde{p}\equiv q_0p_1$ should approximate~$p$ even better up to a constant factor. How to pick the constant factor? To make $\tilde{p}$ look as much as possible like~$p$, it is natural to require $\tilde{p}(a)=p(a)$, therefore:
$$
\tilde{p}(x) = \frac{q_0(x_0)}{q_0(a_0)} p_1(x_1),
$$
hence normalizing~$q_0$ in some way. In other words, for a given coordinate block, we `simplify' the target distribution as $\tilde{p}(x)=q_0(x_0)p_1(x_1)/q_0(a_0)$ and try to fit~$\tilde{p}$ by the same model as before, $q(x)=q_0(x_0)q_1(x_1)$, all this while holding~$q_0$ fixed. Because the simplified target is intuitively `closer' to the actual target than the current fit, it may be useful to further improve it.

To achieve this, we need to be careful because there may still be large discrepancies between~$q$ and~$\tilde{p}$ in the early iterations, so we will consider the tilted target:
\begin{eqnarray*}
t(x) 
 & = & q(x)^{1-\alpha} \tilde{p}(x)^\alpha \\
 & = & q_0(a_0)^{-\alpha} q_0(x_0) q_1(x_1)^{1-\alpha} p_1(x_1)^\alpha . 
\end{eqnarray*}

Therefore, the constant factor~$\gamma$ is defined as: 
$$
\gamma = q_0(a_0)^{-\alpha}
= \left[\frac{q_1(a_1)}{q(a)}\right]^{\alpha}.
$$

We are approximating a form $q_0(x_0)p_1(x_1)$ with a form $q_0(x_0)q_1(x_1)$. While it is `obvious' from the factorial structure that the KL minimization problem can be performed in the lower dimensional space~$x_1$, by simply dropping the term~$q_0(x_0)$. This is confirmed by the KL divergence decomposition itself. Assume~$q_0$ is fixed:
\begin{eqnarray*}
D(q_0 p_1 \| q_0 q_1) 
 & \equiv & 
 \int \left[
 q_0(x_0)p_1(x_1) 
 \log q_0(x_0)q_1(x_1) - q_0(x_0)q_1(x_1)
 \right]
 dx_0 dx_1 \\
& = & 
 c + K_0 D(p_1\|q_1),
\end{eqnarray*}
so, if we maximize $D(q_0p_1\|q_0q_1)$ at fixed~$q_0$, we maximize $D(p_1\|q_1)$. By the way, we shall easily prove that the maximization at fixed~$q_1$ maintains~$q_0$. Here, we exploit a separability property of the KL divergence: another strong argument for choosing this cool divergence measure.

Using coordinate block optimization, we also displace the fix point since we relax the $\alpha$-divergence minimization. The new fixed point is defined by:
$$
\int q_b^{1-\alpha}(x_b)p_b^\alpha(x_b) \phi(x_b) dx_b
= \int q_b(x_b) \phi(x_b) dx_b,
$$
for each block~$b$. In other words, for each block, $q_b$ minimizes the $\alpha$-divergence from the coordinate application $p_b$, itself implicitly defined by~$q_b$ since it depends on the center of~$q_b$.

What is the limit of this iteration if it exists? For any point~$a$, the coordinate applications $p_b(x_b)=p(a_{-b},x_b)$ are well defined, and there is a unique factorial distribution~$q_a=Q(a)$ that minimizes the reduced $\alpha$-divergence for each coordinate block. Consider the function~$C$ that maps any~$q$ to its center. The limit of the iteration must be a fixed point of $C\circ Q$, that is, we need to have $a = C(Q(a)$. We will admit for now that this function has a fixed point.

It is a parametric distribution $q$ such

\section{Constrained KL minimization}

To warrant that the Gaussian fit we are looking for is a proper distribution (has strictly negative second-order coefficients), we may assume that the parameters~$\theta$ associated with second-degree monomials are upper bounded by fixed values. These bounds may be interpreted in terms of maximum variance along each coordinate.

Roughly speaking, if we have a prior on~$x$ (which amounts to some kind of regularization), we expect the target to be (much) more sharply peaked than the prior, so we may just set the prior variance as a maximum variance parameter.

\section{Why?}

Why not trying to directly minimize the global KL divergence $D(p\|q)$? The short answer is: because this is intractable.

Expectation propagation and other message passing schemes \cite{Minka-01,Minka-05} provide ways to work around this problem. However, the core motivation for EP has remained unclear. Minka illustrated how EP provides explicit update rules in a `clutter problem' where direct KL minimization is intractable, but never gave a clear general motivation for EP.

Other authors stressed the computational advantage of distributed inference via data partitioning \cite{Vehtari-14}. This would tend to restrict EP to `big data' problems... However, direct KL minimization may be intractable in many `small data' problems.

The whole EP framework rests upon the `contextual approximation trick', which can be summarized by the formula: $p\leftarrow q^{1-\alpha}p^\alpha$. This can be interpreted as a relaxation: replace in the KL divergence the target $p$ with another target that is presumably much closer to the approximation space, hence it is intuitively `easier' to solve the relaxed KL minimization problem.

Why is it so? Minka's clutter example is one in which the relaxed problem enjoys a simple closed-form solution, while the original problem requires a heavy combinatorial computation. In many cases, there is not even a closed-form solution to either problem, so that an approximate KL minimization needs be performed. This is where some methods, including VS, come into play. In such situations, relaxing the problem is a way to reduce the approximation error. Nothing to do with data partitioning!

How comes the same idea has too very distinct motivations: memory load, on the one hand, and computational tractability, on the other hand? Is it a fortunate coincidence? Which one of the two is the stronger?

%The way I see it at the moment is that there are two distinct ideas underlying the EP~algorithm family.

Distributed inference means varying the target factor across iterations by resampling data (randomly or not). The most straightforward implementation of this idea is actually the AEP algorithm \cite{Dehaene-18}. The EP algorithm itself may be seen as a more sophisticated variant of AEP, one that is also much more memory demanding, that keeps track of all the available data in some sense. The question is: do we need this at all? I see no reason why, but one remark we shall make is that AEP requires an annealing scheme for the learning rate~$\alpha$ in order to converge, while~$\alpha$ can be kept constant in~EP. I will further elaborate on this later on.

If we keep the factor constant, which means that it is actually our target distribution, then AEP becomes a stochastic version of an adaptive $\alpha$-divergence minimization scheme already suggested by Minka \cite{Minka-05} that works by iterative $M$-projections (IMP). IMP is not concerned in any way with distributed inference, its only purpose is to implement $\alpha$-divergence minimization, which may be seen as a relaxation of KL divergence minimization. In this context, $\alpha$ is no longer a learning rate, it is a numerical parameter that controls the relaxation error: the larger, the better, however this is counterbalanced by the practical KL~approximation error: the larger~$\alpha$, the worse. In practice, $\alpha$ should be either fixed, or increased throughout the process to better approximate KL~minimization.

It thus seems that~$\alpha$ is trying to serve two distinct and somewhat contradictory objectives. The way I see to solve the dilemma is to introduce two parameters: a relaxation rate~$\alpha$ and a learning rate~$\rho$, and consider the more general scheme whereby the target~$p$ itself is adaptive:
$$
p \leftarrow q^{1-\rho} h^\rho,
\qquad
q \leftarrow \text{proj}(q^{1-\alpha}p^\alpha),
$$
where~$h$ is some empirical innovation (typically, the distribution associated with a mini-batch). For the target update to converge, we seem to need the learning rate to decrease to zero across iterations. We also note the following:
$$
q^{1-\alpha}p^\alpha = q^{1-\alpha\rho}h^{\alpha\rho},
$$
so, this is exactly the same as the above algorithm if we substitute~$\alpha$ by $\alpha\rho$. In other words, this algorithm cannot distinguish between learning and relaxation. If using a small learning rate~$\rho$, which should also be decreased across iterations for convergence, we can afford setting $\alpha=1$ so that $\alpha\rho$ may be interpreted as a pure learning rate.

The original AEP corresponds to a slightly different setting where we have multiple deterministic factors $h_1,\ldots,h_n$ as opposed to a single stochastic one. In this case, each iteration picks one of the $h_i$'s and it makes sense to set $\rho=1/n$.  Hence, AEP corresponds to a fixed $\alpha\rho=1/n$. The problem is that this may not converge since~$\rho$ is fixed!

I presume that \cite{Dehaene-18} makes no mention of this little issue because they consider AEP as a theoretical tool to study the asymptotic behavior of EP. However, \cite{Li-15} use AEP for practical matters under the name `stochastic EP', and they note that ``SEP converges to the same fixed points as AEP if the learning rates satisfy the Robbins-Monro condition'', thereby implicitly stating that AEP does not converge if the learning rate is kept constant.

Can we have $\alpha\rho$ constant with~$\rho$ decreasing to zero? Say, $\rho=1/N$ where~$N$ is the iteration number. It would imply that $\alpha = N/n \uparrow \infty$. I assume that IMP requires $\alpha \leq 1$ to work, so this is not good. Basically, if $\alpha\leq 1$, $\alpha\rho\leq \rho$ so it needs to converge to zero as well.

EP avoids this convergence issue using separate approximations for each fixed factor. Using fixed factors is not enough, as AEP proves, therefore one wonders if fixed factors have any advantage in realistic large-scale learning context. Since EP can maintain a fixed learning rate, it might converge faster than my simple AEP, but it's not clear what problem EP solves and the memory demand is a real burden. My simple AEP has only one fixed point, which is the KL~minimizer.

\section{Stochastic fixed point}

We know that, in the special case where~$h=p$ is fixed, the IMP fixed point for any fixed~$\alpha$ is the $\alpha$-divergence minimizer. If we decrease~$\alpha$ to zero according or not to RM, we expect the algorithm to converge to the exclusive KL divergence minimizer $\arg\min_q D(q\|p)$, because the 0-divergence is the exclusive KL divergence.

What happens in the stochastic IMP, that is, in the more general case where~$h$ is a random estimate of~$p$? Assume $\mathbb{E}(h)=p$. I speculate that, in the limit of small~$\alpha$,
$$
\mathbb{E}[\text{proj}(q^{1-\alpha}h^\alpha)]
\approx \text{proj}(q^{1-\alpha}p^\alpha).
$$

This is, at least, what intuition tells us, since the result is obviously true if~$h$ is deterministic, so it has to hold at first order. Therefore, under the condition $\alpha\approx 0$, we are back to the deterministic case: the stochastic iteration will minimize the exclusive KL divergence~$D(q\|p)$ on~$Q$. Not the inclusive one, as was our initial objective, but something related... not too bad an outcome. If~$p$ is unimodal, it does not matter much which divergence we choose. If it has multiple modes, then the algorithm will likely pick the strongest (not necessarily the highest). Anyway, it is not clear which $\alpha$-divergence is most appropriate to our problem.

\section{Stochastic approximation}

BTW, why is Li referring to Robbins-Monro? Consider a stochastic iteration of the form: $$
x_{n+1} = f(x_n, \epsilon_n, \omega_n),
$$
where $\epsilon_n$ is a deterministic learning rate and $\omega_n$ is random. Assuming $\omega_n\perp x_n$, this defines a Markov chain. Under the regime of small learning rates, we have by a order-1 Taylor expansion:
$$
f(x_n, \epsilon_n, \omega_n) 
\approx f(x_n, 0, \omega_n) + \epsilon_n \frac{\partial f}{\partial \epsilon} (x_n, 0,\omega_n)
$$

Let us now assume that $f(x, 0, \omega)=x$ holds whatever~$\omega$ and consider the derivative of~$f$ wrt~$\epsilon$ at $\epsilon=0$. The iteration is approximated by a Robbins-Monro algorithm:
$$
x_{n+1} = x_n + \epsilon_n \frac{\partial f}{\partial \epsilon}(x_n,0,\omega_n),
$$

Therefore, in the small learning rate regime, it should have more or less the same asymptotic behavior as Robbins-Monro under the same conditions. In particular, if~$\epsilon_n\sim O(n^{-\alpha})$ with $1/2<\alpha\leq 1$, $x_n$ should converge with probability one to the root of:
$$F(x)
\equiv 
= \frac{\partial \mathbb{E}(f)}{\partial \epsilon} (x,0),
$$
(by linearity), therefore:
$$
x_n \stackrel{p}{\to} x_\star,
\qquad 
\frac{\partial \mathbb{E}(f)}{\partial \epsilon} (x_\star,0)=0.
$$

In our case:
$$
f(q,\epsilon,h) = \text{proj}(q^{1-\epsilon}h^\epsilon)
$$

\section{Exclusive KL divergence minimization}

I remember being asked by Merlin what is the analytical expression of the Gaussian that minimizes exclusive KL divergence, replying that it is easy to find out from the equations, and then realizing that there is no closed-form solution.
$$
D(q\|p) = \int (q \log\frac{q}{p} - q + p)
$$

Say $q=\exp(\theta^\top \phi)$. 
$$
\nabla D(\theta) = 
\int (\log\frac{q_\theta}{p}) q_\theta\phi.
$$

Optimality condition:
$$
\int q_\theta \phi \phi^\top \theta
= \int q_\theta (\log p) \phi
$$

Forget it... What about the hessian?
$$
H = \int q_\theta(1 + \log\frac{q_\theta}{p}) \phi\phi^\top 
$$

\section{Exponentially weighted averages}

How do you compute a moving average in real time? Measure $y_n$ and smooth over time using the recursion:
$$
x_0 = 0,
\qquad
x_{n+1} = (1-\lambda) x_n + \lambda y_n.
$$

You can think of this as averaging approximately over the latest $1/\lambda$ samples. More precisely,
$$
x_n = \lambda \sum_{i=1}^n (1-\lambda)^i y_{n-i}.
$$

So the weight assigned to sample~$i$ is $w_i=\lambda (1-\lambda)^{n-i}$. The sum of weights is:
$$
\sum_{i=1}^n w_i = 1 - (1-\lambda)^n,
$$
so it is closer to one as $n$ increases. For small~$n$, we have bias. To correct $x_n$ for it, we just need to divide it by the sum of weights:
$$
x_n^c = \frac{x_n}{1 - (1-\lambda)^n}
$$

Just a thought: if you want to predict the future from the past, your prediction will always be late on reality, and this is simply because you need to average past data to make accurate predictions.

\section{Stabilization}

The bridge approximation algorithm described in this amazingly well organized document is an iterative process that seeks a fixed point of an $\alpha$-divergence proxy. How can we make sure it converges in practice?

There are different parameters that may significantly impact convergence:
\begin{enumerate}
\item The learning rate $\epsilon$. A small learning rate will help stabilize the algorithm, and that's what it is intended for. 
\item The divergence index $\alpha$. It looks like $\alpha$ needs to lie in the range $[0,1]$. The closer to zero, the smaller the changes on each iteration, hence the more stable the algorithm, although the slower the convergence of course.
\item The initial fit. The closer the initial fit to the target, the more likely convergence. As a default rule, the initial fit could be chosen as the prior distribution of the underlying Bayesian problem, hence an annealing procedure on the prior variance could help convergence: sequentially run bridge approximation for increasingly larger prior variances.
\item Block coordinate size acts on optimization dimensionality: the lower the dimension, the more accurate VS.  condition.
\end{enumerate}

However, among all these parameters,

\section{One-step VS}

In the bridge approximation scheme, where VS is used repeatedly to improve the fit by decreasing $D(q^{1-\alpha}p^\alpha\|q')$ on each iteration, it might not be needed to use a fully blown optimizer such as L-BFGS. Instead, one could do a mere gradient update to try and save computation time and memory load.

Assume we are using VS to try and fit $p=wf$ with $w g_\theta$. Consider first the case where $g_\theta = c$ is simply a constant. We can find the optimal constant explicitly via:
$$
c  = \langle f \rangle_w,
$$
assuming unit sum of weights, {\em i.e.}, $\langle 1 \rangle_w=1$.

From there, we can simply improve the fit by running a gradient update in a larger parametric space. Let us consider $\theta_c = (\log c, 0, 0, \ldots)$ the Gaussian parameter corresponding to the constant fit: 
$$
\theta = \theta_c + \rho D(\theta_c),
\qquad \text{with} \quad
D(\theta_c) = - \nabla \hat{C}(\theta_c)
= \langle f\phi \rangle_w - c \langle \phi \rangle_w.
$$

For a sufficiently small damping value~$\rho$, this will decrease the KL divergence, hence improving the fit $wg_\theta$ over~$w$. We note that the first component of $D(\theta_c)$ (the one associated with the constant term) is zero since $c\langle 1 \rangle_w - \langle f \rangle_w = 0$ by definition.

$$
D(\theta_c) = c [E_{wf}(\phi) - E_{w}(\phi)]
$$

\section{BS}

$$
F(q,h) = \lim_{\epsilon\to 0} 
\frac{\text{proj}(q^{1-\epsilon}h^\epsilon)-q}{\epsilon}
$$

The projection operator is defined implicitly by:
$$
f(p,q) = \int (q-p) \phi = 0
$$

Thus, by the implicit function theorem:
$$
\frac{\partial \text{proj}()}{\partial \epsilon}
= - \left(\frac{\partial F}{\partial \theta}\right)^{-1} \frac{\partial F}{\partial p}
$$

$
f^{\epsilon}= \exp [\epsilon\log f]
\to \log f f^{\epsilon}$

We have:
$$
\frac{\partial q^{1-\epsilon}h^\epsilon}{\partial \epsilon}
=
q (\log h/q) (h/q)^\epsilon
$$

The algorithm is of the form:
$$
q \leftarrow f(q, \epsilon)
$$
where, in our case,
$$
f(q,\epsilon) = \text{proj}(q^{1-\epsilon} h^\epsilon)
$$

$$
f(q,0) = q
$$

$$
f(q,\epsilon) \approx f(q,0) + \epsilon \frac{\partial f}{\partial \epsilon}
$$

$$
\int q_{n+1} \phi = \int q_n^{1-\epsilon} h_n^\epsilon \phi
$$

If $q_n$ converges to a fixed value $q_*$, the expectations need be equal so we need to have: 
$$
\int q_*\phi = \int q_*^{1-\epsilon}p^\epsilon \phi
$$

$$
x_* = E[f(x_*)] 
$$

We note that the projection operator is defined implicitly by:
$$
F(p,\theta) = \int (q_\theta-p)\phi = 0
$$

Thus, 
$$
\frac{\partial \theta}{\partial p}
= - \left(\frac{\partial F}{\partial \theta}\right)^{-1} \frac{\partial F}{\partial p}
$$

$$
\frac{\partial \theta}{\partial p} = \left(\int q_\theta \phi\phi^\top \right)^{-1} \phi
$$

My take on this is simply that we should use a single random factor in practice.

Robbins-Monro for KL-min. We want to solve:
$$
\theta_\star = \arg\min_\theta C(\theta),
\qquad
C(\theta) = -\int p \log q_\theta + \int q_\theta
$$

Optimality condition:
$$
M(\theta)\nabla D(\theta) = 0 = \int (q_\theta-p)\phi
$$

Robbins-Monro algorithm. Consider a random batch via the associated empirical distribution~$h$, and define:
$$
N(\theta) = \int (q_\theta-h) \phi
$$

Clearly, $E(N)=M$ so we can do:
$$
\theta_{n+1} = \theta_n - \epsilon_n N(\theta_n)
$$

Why is this bad? Because $\int h\phi$ is intractable.

\section{More BS}

MisoSoupe
Thorstentungstène
Irina
Edrot
LouicheSantouche
Conovaire
HamBaLaBa
Thierry Margarine
EricDARTELAshburner
Heartscore
Ze Gazbites
Gudrun Eikili

If the context is close enough to the target, VS may improve the fit. However, if the context is too `wide', the fit will likely be meaningless. On the other hand, for a `narrow' context, VS will pretty much behave like a Taylor expansion. So, if we have to choose between the two options, it's clear that we prefer the second one: a narrow $w$.

The choice of context is, however, external to VS and rests upon the context update rule. Given a current fit $q$, what is the best context for the next update? It would be natural to answer $w=q$, but it's not the only reasonable choice. We can think of gradient descent as picking $w(x)=\delta(x-x_q)$ with $x_q=\arg\max q$ as a context function.

But the target update is only part of a wider picture...

The problem I see with this, is that the algorithm may get stuck in a poor local approximation if the actual target is `rugged'.

\begin{figure}[h!]
\centering
\includegraphics[scale=1.7]{universe}
\caption{The Universe}
\label{fig:universe}
\end{figure}

\bibliographystyle{plain}
\bibliography{alexis,stat}
\end{document}

### 2020-03-31 notes/global_optimizer/online.tex
Grabbed global optimizer project from overleaf

\documentclass{article}

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,subfigure}

\def\x{\mathbf{x}}
\def\th{{\boldsymbol{\theta}}}
\def\n{{\boldsymbol{\nu}}}
\def\u{\mathbf{u}}
\def\m{\mathbf{m}}
\def\v{\mathbf{v}}
\def\f{\mathbf{f}}
\def\E{\mathbb{E}}
\def\g{\mathbf{g}}
\def\h{\mathbf{h}}
\def\A{\mathbf{A}}
\def\B{\mathbf{B}}

\title{Online variational sampling} 
\author{Alexis Roche} 
%\date{}

\begin{document}

\maketitle

\section{Basic algorithm}

Given a factor~$f(\x)$, we aim to minimize the local KL divergence $D(wf\|wg_\theta)$, where $w(\x)$ is a known distribution and $g_\th(\x)=\exp[\th^\top \phi(\x)]$ is a parametric distribution to be determined. Instead of a plain variational sampling implementation, consider the following iteration (which defines a Markov chain if I am not mistaken). Start with initial guesses~$\th_0$ and $\u_0$, then:
\begin{eqnarray*}
\x_n & \sim & w(\x),\\
\u_{n+1} & = & (1-\beta)\u_n + \beta [g_{\th_n} (\x_n) - f(\x_n)] \phi(\x_n),\\
\th_{n+1} & = & \th_n - \gamma_n \u_{n+1},
\end{eqnarray*}
where $0\leq \beta\leq 1$ and $\gamma_n$ satisfies the Robbins-Monro condition, therefore we can hope as a general property of stochastic approximation algorithms that $\th_n$ converges to a fixed value~$\th_\star$ such that $\E(\u_n) = 0$. Now, from the second update rule, this condition implies:
$$
\E \left\{[g_{\th_\star} (\x_n) - f(\x_n)] \phi(\x_n) \right\}= 0,
$$ 
but, since $\x_n$ is distributed like $w(\x)$ by construction, this means:
$$
\int w(\x) [g_\th (\x) - f(\x)] \phi(\x) d\x = 0.
$$

This is exactly the first-order KL minimality condition. So, IF convergence occurs, the limit~$\th_\star$ is necessarily the KL minimizer, which is unique since the KL divergence is convex:
$$
\th_\star = \arg\min_\theta D(wf\|wg_\th). 
$$

In fact, we see that~$\u_n$ is an online estimate of the KL divergence gradient using an exponentially weighted average, which may be thought as roughly computing the average of the latest $1/\beta$ values. The smaller~$\beta$, the more stable this estimate but the slower it reaches steady state.

\section{Alternative formulation}

Rather than saying that we fit the factor~$f(\x)$, we may say that we fit a target distribution $p(\x)$ with an exponential family distribution~$q_\th(\x)$. The above iteration may then be rewritten as follows:
\begin{eqnarray*}
\x_n & \sim & w(\x),\\
\u_{n+1} & = & (1-\beta)\u_n + \beta \frac{q_{\th_n} (\x_n) - p(\x_n)}{w(\x)} \phi(\x_n),\\
\th_{n+1} & = & \th_n - \gamma_n \u_{n+1}.
\end{eqnarray*}

The only difference is that~$\th$ now encodes for~$q_\th$ as opposed to~$g_\th$. In parameter space (assuming that~$w$ belongs to the parametric family), one is just found by applying an offset to the other. This form is  more general as we don't assume the target to be of the form $p(\x)=w(\x)f(\x)$. As a result, we get the ``infamous'' division by the kernel value.

\section{Adaptive sampling version}

The above algorithm is dramatically dependent on the kernel~$w(\x)$, not only by the result of divergence minimization, but also by the speed of convergence. We cannot guarantee in advance that the chosen kernel is suitable? To work around this problem, let us consider a variant of the above iteration in which the sampling kernel is updated on each iteration depending on the current fit $q_\th(\x)$. Assume that, given a current~$\th$, we decide to do pick:
$$
w(\x) = q_\th^{1-\alpha}(\x), 
$$
for some fixed $0<\alpha<1$. Why exactly we do this will be explained later. It amounts to sampling from a slightly more widespread distribution than the current fit. The other trick is to also re-define the target distribution on each iteration as being:
$$
t(\x) = w(\x) p(\x)^\alpha = q_\th^{1-\alpha}(\x) p(\x)^\alpha,
$$
which means that it is somewhere on the geometric path between the current fit and the actual target~$p(\x)$.

We may then apply the above update rules in the exact same way as above. By construction, we have:
$$
\frac{q_\th(\x)}{w(\x)} = q_\th(\x)^\alpha,
\qquad
\frac{p(\x)}{w(\x)} = p(\x)^\alpha.
$$

Therefore, the iteration reads:
\begin{eqnarray*}
\x_n & \sim & q_{\th_n}^{1-\alpha}(\x),\\
\u_{n+1} & = & (1-\beta)\u_n + \beta \big[q_{\th_n}(\x_n)^\alpha - p(\x_n)^\alpha \big] \phi(\x_n),\\
\th_{n+1} & = & \th_n - \gamma_n \u_{n+1}.
\end{eqnarray*}

We can check similarly to above that, IF $\th_n$ converges almost surely, it does so towards the unique and well-defined $\alpha$-divergence minimizer:
$$
\th_\star = \arg\min_{\th} D_\alpha(p \| q_\th).
$$

Hence, we have solved the ``chicken and egg'' problem of choosing a sampling kernel by replacing the local KL divergence objective with an $\alpha$-divergence. It took me time to realize that this was the right thing to do after being stuck for so many years in the asymptotic approach.

In large-scale problems, the target is computed from a random mini-batch, hence further adding noise to the procedure. Is that a problem? I don't think so.

We will argue that this iterative algorithm can replace stochastic gradient methods for large-scale machine learning. One fairly pleasant and maybe surprising property is that it does not require any gradient computation (hence, no back-propagation involved)! The other one is that its fixed point is `more global' and better-defined than for gradient methods (although not necessarily unique except for $\alpha=1$), which only converge to one of the many saddle points of the learning objective.

\section{KL divergence minimization}

Consider the problem:
$$
\min_\theta D(wf \| wg_\th),
\qquad
g_\th(\x) = e^{\th^\top \phi(\x)}
$$

$$
D(\th) \equiv  \int (-wf \log g_\th + w g_\th) 
$$

$$
\nabla D = \int w(g_\th - f)\phi
$$

$$
\nabla\nabla^\top D = \int wg_\th \phi\phi^\top
$$

Note that the Hessian is independent from the target factor.

\section{Basis change}

Consider minimizing the KL divergence via a gradient descent and consider a single gradient update. If we cannot afford to compute the Hessian (and to invert it), the gradient update relies on an arbitrary metric and may be rather ineffective. An important property of the KL divergence with exponential families is that its Hessian is independent from the target distribution and is associated with a dot product on the linear space spanned by~$\phi$ (which is a subspace of polynomials of order~2 for the factorial Gaussian family):
$$
H = \int q(\x) \phi(\x) \phi(\x)^\top d\x 
\quad \Rightarrow \quad
H_{ij} = \langle \phi_i , \phi_j \rangle .
$$

A natural idea is then to pick an orthonormal basis to use temporarily for a single gradient update as the divergence Hessian is scalar in such a basis.

Can we easily compute such as basis? The answer is yes. Assume for now and for simplicity that $\int q = 1$. In one dimension, it is easy to find an orthonormal basis:
$$
\phi_0(x) = 1,
\quad \phi_1(x) = \frac{(x-m)}{s},
\quad \phi_2(x) = \frac{1}{\sqrt{2}}\left[\frac{(x-m)^2}{s^2} - 1\right].
$$

It is straightforward to see that $\|P_0\|=1$, $\|P_1\|=1$, $\langle P_0, P_1 \rangle=0$ and $\langle P_0, P_2 \rangle=0$. The reason why $\langle P_1, P_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric. Finally, note that:
$$
\phi_2(x)^2 = \frac{1}{2} 
\left[\frac{(x-m)^4}{s^4} - 2 \frac{(x-m)^2}{s^2} + 1 \right],
$$
hence, given that the 4th-order centered moment of a Gaussian is~3 (which can be seen using the $\chi_1^2$ distribution), we have:
$$
\E_q [\phi_2^2(x)] = \|P_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
$$

In multiple dimensions, we do:
$$
\phi_0(\x) = 1, 
\quad
\phi_{1i}(\x) = \frac{(x_i-m_i)}{s_i},
\quad
\phi_{2i}(\x) = \frac{1}{\sqrt{2}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
$$
This defines an orthonormal basis because the components of~$\x$ are mutually independent under the factorial distribution~$q(\x)$, hence for different basis vectors: 
$$
\forall (i,j), \quad i\not= j, \quad
\langle \phi_{ki}, \phi_{\ell j} \rangle = \E_q[\phi_{ki}]\E_q[\phi_{ki}] = \langle \phi_{ki}, 1 \rangle \langle P_{\ell j}, 1 \rangle = 0.
$$
Moreover, when it comes to dot products involving the same component, the one-dimensional results apply.

In the general case where~$q$ does not integrate to one, we simply need to divide each basis vector by $Z=\int q_0$ to fulfill the unit norm conditions. Alternatively, we may leave the basis as is and later remember that the Hessian is $Z\text{Id}$.

So we have an orthonormal basis that is very easy to compute! Now, consider decreasing the KL divergence $D(p\|q)$ starting from some current guess~$q_0$. To this end, we may compute an orthonormal basis associated with~$q_0$ and perform one Newton iteration in that basis. More specifically, we search for a new fit under the form:
$$
q_\th(\x) = q_0(\x) e^{\th^\top \phi(\x)},
$$
where~$\phi$ is the orthonormal basis. Clearly, the initial parameter is $\th_0=0$. The Newton update yields:
$$
\th = \gamma \f,
\qquad
\f = -\frac{1}{Z} \u,
\quad
\u \equiv \nabla D(\th_0) = \int (q_0 - p)\phi,
$$
where $\gamma$ is a given step size ($\gamma=1$ for a genuine Newton iteration) and~$Z$ is there for the Hessian correction.

The gradient is generally intractable as it depends on~$p$, but we can approximate it via random sampling or quadrature. There are essentially two options. One is:
$$
\u \approx \int q_0\phi - {\cal I}(p\phi),
$$
where~${\cal I}$ denotes a numerical integration operator. The other option is:
$$
\u \approx {\cal I}[(q_0- p)\phi].
$$
This is where the variational sampling story may start. Option~1 amounts to a direct moment approximation. Option~2, which corresponds to our initial variational sampling idea, is perhaps less intuitive as it ignores the fact that we can compute the ``fitted moment'' exactly, but is arguably better because the error is smaller whenever~$q_0$ is close enough to~$p$. Hence, for a near Gaussian target, we expect less numerical instabilities. Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.

So, our gradient update takes on the simple form $\th = \gamma\tilde{\f}$, but this is computed in a moving coordinate system since the orthonormal basis changes on each iteration. Is it easy to compute the `natural' parameters of $q_\th=q_0\exp(\th^\top \phi)$ from the moving coordinates~$\th$? Yes, it is.

First, consider the right hand side and try to do the identification:
$$
e^{\th^\top \phi} = \kappa e^{-\frac{(\x-\mu)^2}{2\sigma^2}},
$$
hence, by taking the log:
$$
\log\kappa - \frac{1}{2} \sum_i \frac{(x_i-\mu_i)^2}{\sigma_i^2}
= 
\theta_0 + \sum_i \theta_{1i}\frac{x_i-m_i}{s_i}
+ \frac{1}{\sqrt{2}} \sum_i \theta_{2i}\big[ \frac{(x_i-m_i)^2}{s_i^2} -1 \big]. 
$$
We may differentiate this wrt $x_i$, yielding: 
$$
\frac{x_i-\mu_i}{\sigma_i^2} 
= \frac{\theta_{1i}}{s_i} + \sqrt{2} \theta_{2i} \frac{x_i-m_i}{s_i^2}
= \frac{\sqrt{2} \theta_{2i}}{s_i^2} \big[x_i - m_i + \frac{s_i\theta_{1i}}{\sqrt{2} \theta_{2i}} \big].
$$
By identification, we find:
$$
\sigma_i^2 = - \frac{s_i^2}{\sqrt{2}\theta_{2i}},
$$
and:
$$
\mu_i = m_i - \frac{s_i\theta_{1i}}{\sqrt{2}\theta_{2i}} = m_i + \frac{\sigma_i^2}{s_i}\theta_{1i}.
$$

To find $\kappa$, we note that the integral of $\th^\top \phi$ wrt~$q_0$ is equal to its dot product with~$\phi_0\equiv 1$, which is $\theta_0 \|\phi_0\|^2=\theta_0 Z$ since the basis $\phi/\sqrt{Z}$ is orthonormal. The constant $Z$ cancels out as it appears on both sides, and we find, using the K\"onig formula:
$$
\log \kappa = \theta_0 + \frac{1}{2}\sum_i \frac{s_i^2 + (\mu_i-m_i)^2}{\sigma_i^2}.
$$

We also need to get the parameter~$\th_0$ of~$q_0$, which act as an offset. For this, we just do $\kappa=K_0$, $\mu_i=m_i$, and $\sigma_i=s_i$, yielding:
$$
\theta_{2i}^0=-\frac{1}{\sqrt{2}},
\quad
\theta_{1i}^0 = 0,
\quad
\theta_0^0 = \log K_0 - \frac{d}{2}.
$$

Therefore, the updated fit is a function $\Psi(q_0,\th)$ of the current fit and parameter variation:
$$
v_{i1} =  \frac{v_{i0}}{1-\sqrt{2}\theta_{2i}},
\quad
m_{i1} = m_{i0} + \frac{v_{i1}}{\sqrt{v_{i0}}} \theta_{1i},
\quad
\log K_1 = \log K_0 + \theta_0 + \frac{1}{2} \left[\sum_i \frac{v_{i0} + (m_{i1}-m_{i0})^2}{v_{i1}} - d\right].
$$

\section{Stochastic gradient}

To approximate the gradient~$\u$, we may sample a point~$\x$ (or several ones) from some distribution $w(\x)$ and compute one of the following approximations.
\begin{enumerate}
\item Discrete KL proxy:
$$
\u \approx Z_w \frac{q(\x)-p(\x)}{w(\x)} \phi(\x)
$$
\item Likelihood proxy:
$$
\u \approx \int q\phi - Z_w\frac{p(\x)}{w(\x)} 
\phi(\x)
$$
\end{enumerate}

The discrete KL proxy provides a lower variance estimate of~$\u$ whenever the fit~$q$ is close enough to the target~$p$, and is even exact if $q=p$. The likelihood proxy is considered for historical reasons as it is the online version of direct Monte Carlo moment estimation.

Note that:
$$
\int q\phi_0 = Z,
\quad
\int q\phi_1 = 0,
\quad
\int q\phi_2 = 0.
$$

Therefore, the likelihood approximation to $\f=-\u/Z$ is:
$$
f_0 = \frac{Z_w}{Z}\frac{p(\x)}{w(\x)} - 1 
$$

$$
f_{i>0} = \frac{Z_w}{Z}\frac{p(\x)}{w(\x)}\phi_i(\x) 
$$

\section{Optimal sampling}

Using the discrete KL proxy, we estimate $\u=\int (q-p)\phi$ by sampling a point from an arbitrary distribution~$w(\x)$:
$$
\tilde{\u} = Z_w \frac{q(\x)-p(\x)}{w(\x)} \phi(\x),
$$
which clearly satisfies $\E(\tilde{\u})=\u$ whatever the sampling distribution and is therefore unbiased. However, its variance depends on~$w$, and we ought to pick~$w$ so as to make the variance as small as possible.

Since the mean is independent from~$w$, the variance depends on~$w$ only through the squared expectation matrix:
$$
\text{Var}(\tilde{\u}) 
= \E(\tilde{\u}\tilde{\u}^\top) - \E(\tilde{\u})\E(\tilde{\u})^\top
= \E(\tilde{\u}\tilde{\u}^\top) + {\rm cte},
$$
with:
$$
\E(\tilde{\u}\tilde{\u}^\top) = Z_w \int\frac{(q-p)^2}{w}\phi\phi^\top.
$$

This is a matrix, not a real-valued function, so it can't be minimized in a strict sense, reflecting the fact that no sampling distribution is optimal for all coordinates of a multidimensional integral. What matters most is that the function $(q-p)^2/w$ be a well-behaved distribution in the sense that  $\E_f(\phi\phi^\top)$ is finite. Since~$\phi$ represents second-order polynomials, this essentially means that it should vanish exponentially for $\|x\|\to\infty$. If this is the case, then all the elements of the squared matrix will be finite, therefore we may focus on one particular element, e.g., the constant term, for the sake of practicality:
$$
V(w) = Z_w \int\frac{(q-p)^2}{w}.
$$

The minimizer of~$V(w)$ is $w\propto |q-p|$ but this does not help much as there is no closed-form expression for~$p$. Note that, for the likelihood proxy, the problem is to estimate $\int p\phi$, for which the variance reads (up to an additive constant):
$$
V_{\ell}(w) = Z_w \int \frac{p^2}{w},
$$
in which case the optimal kernel is~$p$, which is neat and simple, although impractical... Intuition would tell us to choose a sampling distribution similar to the target, so the current fit~$q$ would be a candidate, but $V_\ell(q)$ is infinite whenever~$p$ has heavier distribution tails than a Gaussian. Intuition is thus misleading.

Back to the discrete KL proxy, the choice~$w=q$ is unsafe for the very same reason. Then, you say: ok, heavy tails do not happen if one uses a Gaussian prior. But then a sufficient condition for $p^2/q$ to have finite integral is that the variance of~$q$ be larger than half the prior variance (assuming a finite likelihood). This condition is clearly unrealistic, therefore the prior does not do the trick.

We remark that, if there happens to exist a context distribution $c(\x)$ such that $p=cf$, $q=cg$ {\em and} both~$f$ and~$g$ are upper bounded, then~$c$ is a valid candidate for the sampling distribution since $V(c)$ is finite:
$$
V(c)= Z_c \int c (f-g)^2 \leq Z_c^2 (f_{\max}^2 + g_{\max}^2) 
$$ 
The gradient estimator is then:
$$
\tilde{\u} = [g(\x)-f(\x)] \phi(\x).
$$

One way to meet this condition is to replace the target by $q^{1-\alpha}p^\alpha$ for $0<\alpha<1$ and set $c=q^{1-\alpha}$, $f=p^\alpha$ and $g=q^\alpha$. If the target is upper bounded (which is usually easy to verify), then it works cause $g=q^\alpha$ is upper bounded if~$q$ is. This means that we are changing the target at each iteration of the algorithm, but that's the idea of the star approximation.

The larger~$\alpha$, the larger the variance tends to be. It blows up when $\alpha\to 1$ as $Z(q^{1-\alpha})$ becomes infinite, meaning that this strategy then becomes ineffective.

%What about the extreme cases $\alpha=0$ and $\alpha=1$? When $\alpha\to 0$, $\tilde{\u}\to 0$, and the variance vanishes. It means that this strategy is not appropriate to

This contextual factor is also the very principle behind expectation propagation and more general message passing \cite{Minka-05}. Very interesting to note that it can be motivated by sampling purposes! Here, it is not brought in by the target factorial structure -- it is imposed by brute force.

\section{More coordinates}

We might wish to average increments in the above procedure for variance reduction purpose. The problem is that these are expressed in a moving coordinate system, so it does not really make sense to average them. To that end, we may need to convert increments back to a fixed coordinate system which is linearly related to the `moving theta', e.g.,
$$
\log q_\th(\x) = \theta_0 + \sum_i \theta_{1i} x_i + \sum_i \theta_{2i} x_i^2 .
$$

We denote by~$\n$ the moving coordinates to avoid confusion. A linear update in~$\n$ amounts to a linear update in~$\th$. If the increment in moving coordinates is $\Delta\n$, is it easy to find the equivalent $\Delta\th$? Yes it is. Let us equate $q_{\Delta\th}$ with $\exp(\Delta\n^\top \phi)$:
\begin{eqnarray*}
\Delta\theta_0 + \sum_i \Delta\theta_{1i} x_i + \sum_i \Delta\theta_{2i} x_i^2
& = & 
\Delta\nu_0 
+ \sum_i \Delta\nu_{1i} \frac{x_i-m_i}{s_i} 
+ \sum_i \frac{\Delta\nu_{2i}}{\sqrt{2}} \left[\frac{(x_i-m_i)^2}{s_i^2}-1\right] \\
& = & 
\Delta\nu_0 
+ \sum_i \Delta\nu_{1i} \frac{x_i-m_i}{s_i}
+ \sum_i \frac{\Delta\nu_{2i}}{\sqrt{2}} 
\left[\frac{x_i^2 - 2m_ix_i + m_i^2}{s_i^2}-1\right]
\end{eqnarray*}

We find by identification: 
\begin{eqnarray*}
\Delta\theta_{2i} & = & \frac{\Delta\nu_{2i}}{\sqrt{2}s_i^2},\\
\Delta\theta_{1i} & = & 
\frac{\Delta\nu_{1i}}{s_i} - \frac{\sqrt{2}m_i}{s_i^2}\Delta\nu_{2i},\\
\Delta\theta_0 & = & \Delta\nu_0 + \sum_i \left[
-\frac{m_i}{s_i}\Delta\nu_{1i} + \frac{m_i^2 - s_i^2}{\sqrt{2}s_i^2} \Delta\nu_{2i}
\right].
\end{eqnarray*}

These imply $\Delta\nu_{2i} = \sqrt{2}s_i^2 \Delta\theta_{2i}$, hence:
$$
\Delta\theta_{1i} = 
\frac{\Delta\nu_{1i}}{s_i} - 2m_i \Delta \theta_{2i},
$$
implying in turn $\Delta\nu_{1i} = s_i(2m_i \Delta\theta_{2i} + \Delta\theta_{1i})$, hence:
$$
\Delta\theta_0 = \Delta\nu_0 - \sum_i \left[
m_i \Delta\theta_{1i} + (m_i^2 + s_i^2) \Delta\theta_{2i}
\right]
$$

Now, assume that we have just performed an update: $\th_1 = \th_0 + \Delta$. How are the natural parameters $K$, $m$ and $s$ affected? This is a third coordinate system in this story, which we need for sampling and is non-linearly related to the other two! However, we can easily switch from~$\th$ to natural parameters by equating the two forms:
\begin{eqnarray*}
\theta_0 + \sum_i \theta_{1i} x_i + \sum_i \theta_{2i} x_i^2
& = & \log K - \frac{1}{2} \sum_i \frac{(x_i-m_i)^2}{s_i^2},\\
& = & \log K - \frac{1}{2} \sum_i \frac{x_i^2 - 2 m_ix_i + m_i^2}{s_i^2},\\
& = & \log K - \frac{1}{2} \sum_i \frac{m_i^2}{s_i^2} 
+ \sum_i \frac{m_i x_i}{s_i^2} 
- \frac{1}{2} \sum_i \frac{x_i^2}{s_i^2},
\end{eqnarray*}
yielding:
\begin{eqnarray*}
\theta_0 & = & \log K -\frac{1}{2}\sum_i \frac{m_i^2}{s_i^2},\\
\theta_1 & = & \frac{m}{s^2},\\
\theta_2 & = & -\frac{1}{2 s^2}.
\end{eqnarray*}

Therefore:
\begin{eqnarray*}
s^2 & = & -\frac{1}{2\theta_2},\\
m & = & s^2 \theta_1,\\
\log K & = & \theta_0 + \frac{1}{2}\sum_i \frac{m_i^2}{s_i^2}.
\end{eqnarray*}

\section{Improved algorithm}

Taking the basis change into account, we may perform the following iteration:
\begin{eqnarray*}
\x_n & \sim & q_n^{1-\alpha}(\x),\\
\f_n & = & \rho_n \big[p(\x_n)^\alpha - q_n(\x_n)^\alpha \big] \Phi(q_n, \x_n),\\
q_{n+1} & = & \Psi(q_n, \gamma \f_n),
\end{eqnarray*}
where $\Phi(q,\x)$ is the basis associated with~$q$ evaluated at~$\x$, and $\Psi(q,\th)$ is the fit updater depending on moving coordinates~$\th$. The factor~$\rho_n$ normalizes the basis function values to unit norm, and is defined by:
$$
\rho_n
= \frac{Z(q_n^{1-\alpha})}{Z(q_n)},
$$
and the reason why $Z(q_n^{1-\alpha})$ is in the numerator is that $\x_n$ is actually distributed like the normalized version of $q_n^{1-\alpha}$. The integral~$Z$ of an unnormalized distribution depends on~$K$ and~$v$ via: 
$$
Z = (2\pi)^\frac{d}{2} K  \prod_i v_i^\frac{1}{2}.
$$

For $q^{1-\alpha}$, we have $K_\alpha = K^{1-\alpha}$ and $v_\alpha=v/(1-\alpha)$, thus:
$$
\rho_n = K^{-\alpha} (1-\alpha)^{-\frac{d}{2}}.
$$
The dimension-dependent term $(1-\alpha)^{-d/2}$ scales exponentially with dimension, suggesting the need to adjust~$\alpha$ for the dimension in practice. The larger the dimension, the further away $q^{1-\alpha}p^\alpha$ from~$q$ for a constant~$\alpha$, so intuition confirms that~$\alpha$ should decrease with dimension.

Throughout the algorithm, we need to keep track of the natural parameters $K$, $m$ and $v$ of the fit~$q$. Besides the correction factor~$\rho_n$, the basis vectors are defined from those via:
$$
\phi_0(\x) = 1, 
\quad
\phi_{1i}(\x) = \frac{(x_i-m_i)}{s_i},
\quad
\phi_{2i}(\x) = \frac{1}{\sqrt{2}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right],
$$
which are computed on-the-fly (no need to store a `design matrix' as in offline variational sampling). After computing the force~$\f_n$ (the negated stochastic gradient normalized by the `genuine' Hessian), we update the moving coordinates as $\th=\gamma \f_n$ and update the fit accordingly:
$$
v_{i1} =  \frac{v_{i0}}{1-\sqrt{2}\theta_{2i}},
\quad
m_{i1} = m_{i0} + \frac{v_{i1}}{\sqrt{v_{i0}}} \theta_{1i},
\quad
\log K_1 = \log K_0 + \theta_0 + \frac{1}{2} \left[\sum_i \frac{v_{i0} + (m_{i1}-m_{i0})^2}{v_{i1}} - d\right].
$$

%%Are $\f_n$ values comparable (and thus can we average them) across iterations given that they rely on

%$(n+1)x_{n+1} = n x_n + y_n$

\section{Implementation details}

We note that $\rho_n=K_n^{-\alpha}(1-\alpha)^{-d/2}$ where $K_n=q_n(\mu_n)$ is the current fit's peak height, while the second term is a constant, and a cumbersome one at that as it takes on large values in high~D. For the sake of numerical stability, as will be seen below, we are better off re-writing the `force' as: 
$$
\f_n = \varepsilon_n \Phi(q_n,\x_n),
\qquad
\varepsilon_n \equiv
(1-\alpha)^{-\frac{d}{2}}\left\{
\left[\frac{p(\x_n)}{K_n}\right]^\alpha
-\left[\frac{q(\x_n)}{K_n}\right]^\alpha
\right\},
$$
showing that the peak height $K_n$ acts as a normalizing factor.

%In practice, we do not need to compute $(1-\alpha)^{-d/2}$ in~$\rho_n$ as it is a constant, and a cumbersome one at that as it takes on large values in high~D. For the sake of numerical stability, we are better off re-defining the `force' as:

%All it means is that the computed force is $(1-\alpha)^{d/2}$ times the effective one, hence it is smaller, and conversely the effective step size is smaller than~$\gamma$: $\gamma_\star = (1-\alpha)^{d/2} \gamma$. A constant~$\gamma$ thus results in an effective step size that decreases with the dimension.

%In practice, $(1-\alpha)^{-d/2}\approx 1.0513$ is constant across dimensions, hence we may drop this term as it boils down to a (slight) change in step size. The remaining term in~$\rho_n$ is~$K_n^{-\alpha}$, where $K_n$ is the current fit's peak height, $K_n=q_n(\mu_n)$, which acts as normalizing the distribution values for the `delta' computation:
%$$
%\varepsilon_n = \gamma\left\{
%\left[\frac{p(\x_n)}{K_n}\right]^\alpha
%-\left[\frac{q(\x_n)}{K_n}\right]^\alpha
%\right\}.
%$$

Next, we update the fit parameters via $\delta_n\equiv\gamma\varepsilon_n$ as follows:
$$
\frac{1}{\v_{n+1}} = \left\{
1 - \left[\frac{(\x_n-\m_n)^2}{\v_n} - 1\right] \delta_n
\right\} \frac{1}{\v_{n}},
$$
$$
\m_{n+1} = \m_{n} + \frac{\v_{n+1}}{\v_n}
(\x_n-\m_n)\delta_n,
$$
$$
\log K_{n+1} = \log K_n + \frac{1}{2}\left[\mathbf{1}^\top
\frac{\v_n + (\m_{n+1}-\m_n)^2}{\v_{n+1}} - d
\right]
+ \delta_n.
$$

Note that it is easy to enforce positive variances by thresholding $\delta_n=\gamma\varepsilon_n$ appropriately. In practice, however, this is not needed as $\delta_n$ should be small enough on each iteration. If it does happen, then it is an indication that the effective step size is too large.

\section{Ultimate algorithm}

The scaling factor $(1-\alpha)^{-d/2}$ is a constant, so it can be dropped. 
$$
\delta_n = \gamma' \varepsilon_n',
\qquad 
\varepsilon' = (p/K)^\alpha - (q/K)^\alpha,
\quad
\gamma = (1-\alpha)^{d/2} \gamma'
$$

Noise control (constant variance):
$$
\gamma' = \frac{k_\gamma}{\alpha\sqrt{d}}
$$

For $\alpha\approx 0$,  we fall back to the minimum exclusive KL limiting case (see below). For $\alpha\to 1$, $\delta_n$ is magically well-conditioned.

Default $\alpha$ setting (to make~$\gamma'$ independent from~$d$):
$$
\alpha = \frac{k_\alpha}{\sqrt{d}}
\quad
\Rightarrow
\quad \gamma' = \frac{k_\alpha}{k_\gamma}.
$$
This is a good starting point, but makes the algorithm quite similar to Carbonetto so only useful if one plans to further increase~$\alpha$.

Speculation: when increasing $\alpha$ in a progressive manner, there is a way to adjust $\gamma'$ so as to maintain stochastic convergence (wishing that the increase in noise variance would be compensated for by the smaller fitting error compared to the start). There could be a warming up phase in which $\alpha$ is increased to one in such a way, followed by an annealing phase in which $\gamma$ is decreased to zero for deterministic convergence.

What is the best way to adjust~$\gamma'$ in such context? We have to trade off between the intrinsically slower convergence for larger~$\alpha$ and the larger noise level for a constant step size. There are two extreme choices: keep the effective step size~$\gamma$ constant (try and maintain deterministic convergence speed), or keep the noise level constant (try and maintain stochastic convergence speed). Noise control (i.e. dividing the initial~$\gamma'$ by the ratio $\alpha/\alpha_{init}$) seems to make overall convergence hopelessly slow as the effective step size $\gamma=(1-\alpha)^{d/2}\gamma'$ is a fast decreasing function of~$\alpha$. What about keeping the effective step size constant? This means that $\gamma'$ is chosen according to $\gamma' \propto (1-\alpha)^{-\frac{d}{2}}$ and is not reasonable as the variance blows up!

The heuristic I lean to is to keep~$\gamma'$ constant. This seems to make convergence quite smooth.

Removing the `prime' symbols, this leads to the implementation-ready form:
\begin{eqnarray*}
\x_n & \sim & q_n^{1-\alpha}(\x),\\
\varepsilon_n & = & 
\left[\frac{p(\x_n)}{K_n}\right]^\alpha
-\left[\frac{q(\x_n)}{K_n}\right]^\alpha,\\
\f_n & = & \varepsilon_n \Phi(q_n, \x_n),\\
q_{n+1} & = & \Psi(q_n, \gamma \f_n),
\end{eqnarray*}

\section{Parameter tuning}

The algorithm has two parameters: the divergence index~$\alpha$ and the stepsize~$\gamma$, which should not be tuned independently, in particular because the correction factor $(1-\alpha)^{d/2}$ blows up in high~D, leading to huge increments unless~$\alpha$ is set to a lower value as the dimension decreases. The basic reason why this problem is happening is because the `force' is noisy, and its variance depends on both~$\alpha$ and the dimension.

One way to tackle the problem is to assume that~$q$ is the $\alpha$-divegence minimizer and thus has zero gradient. The values of $\f_n$ are then pure noise. We see that the variance is then:
$$
\text{Var}(\f_n) = (1-\alpha)^{-d} 
\int \tilde{q}_{1-\alpha} 
\left[(p/K)^\alpha - (q/K)^\alpha\right]^2\phi\phi^\top,
$$
where $\tilde{q}$ is the normalized version of~$q^{1-\alpha}$. For small~$\alpha$, the integral can be approximated at order~1, yielding:
$$
\text{Var}(\f_n) = (1-\alpha)^{-d} \alpha^2
\int \tilde{q} [\log(p/q)]^2 \phi\phi^\top . 
$$

The integral above depends on both~$p$ and~$q$, its $\alpha$-proximal distribution. For the sake of getting an order of magnitude, we may assume that $\log p/q=a$ is roughly constant, therefore: 
$$
\text{Var}(\f_n) \approx 
(1-\alpha)^{-d} \alpha^2
a^2 \int \tilde{q} \phi\phi^\top
= (1-\alpha)^{-d} \alpha^2
a^2 \mathbf{I}_d.
$$

In other words, the expected noise level roughly depends on both~$\alpha$ and the dimension~$d$ via: 
$$
\sqrt{\E[\|\f_n\|^2]} \propto (1-\alpha)^{-\frac{d}{2}} \alpha \sqrt{d} .
$$

We may thus pick~$\gamma$ depending on~$\alpha$ and~$d$ so as to control the noise level at a fixed value, yielding the rule:
$$
\gamma = k_\gamma \frac{(1-\alpha)^\frac{d}{2}}{\alpha\sqrt{d}},
$$
which is a decreasing function of~$\alpha$ that vanishes when $a\to 1$. As a function of dimension, $\gamma\sim \/\sqrt{d}$ for small~$\alpha$ and $\gamma\to 0$ for~$\alpha\to 1$ (keeping in mind that the derivation relies on a Taylor expansion around $\alpha=0$).

In practice, $k_\gamma=0.01$ seems to warrant convergence most often. The convergence speed is, however, highly dependent on~$\alpha$ because it critically depends on the step size~$\gamma$, itself determined by both~$\alpha$ and the dimension. To make sure that convergence speed scales well with dimension, we may force  $\gamma$ to keep a constant order of magnitude across dimensions. To this end, we may tune~$\alpha$ depending on the dimension so that~$\gamma$ is roughly constant. Since the solution is near zero, we may approximate~$\gamma$ by $\gamma\approx k_\gamma/\alpha\sqrt{d}$ to find it, hence:
$$
\alpha \approx \frac{k_\alpha}{\sqrt{d}}
\qquad \Rightarrow \quad
\gamma \approx \frac{k_\gamma}{k_\alpha}.
$$

A simple rule of thumb is therefore to choose~$\alpha$ inversely proportional to the square rooted dimension. The factor $k_\alpha$ represents the divergence index in dimension~$1$, and $k_\alpha=0.1$ seems to work well in practice, implying $\gamma\approx 0.1$ in any dimension. In this setting, the number of iterations required for convergence seems to scale proportionally to~$\sqrt{d}$: $1000\sqrt{d}$ seems to be fair.

%In this scenario, $\alpha_1=0.1$ seems to work fine with a step size $\gamma=0.1$ independent from the dimension. Larger values of~$\gamma$ sometimes enable faster convergence, but also make the algorithm less stable, so convergence may not happen at all.

For the iteration to converge almost surely, $\gamma$ needs be decreased according to the Robbins-Monro condition. When~$\gamma$ is tied to~$\alpha$ through the above relationship, this amounts to increasing~$\alpha$ throughout the algorithm, hence converging to the KL~optimal distribution provided convergence occurs... That's what we wanted, isn't it?

%How do we go about arbitrary~$\alpha$ values? We then have the above mentioned issue that the correction factor may be orders of magnitude larger, making the algorithm diverge if we use the same step size. One general way to address this is to start the algorithm with the auto-tuned~$\alpha$, which is close to zero, and increase~$\alpha$ as the fit gets better. Still, we need to be able to adapt the step size to increases in~$\alpha$.

\section{Variant}

Rather than sampling from~$q_n^{1-\alpha}$ and considering~$p^\alpha$ as the factor to be fitted, why not sampling from~$q_n$ with target factor $(p/q_n)^\alpha$? By doing so, we always have $\rho_n=1$, and the `epsilon' simplifies to:
$$
\varepsilon_n = 
\left[\frac{p(\x_n)}{q(\x_n)}\right]^\alpha -1 ,
$$
which no longer involves peak height. For the `likelihood' proxy, the `delta' is the same for the constant term and does not involve the subtraction by one for the other terms:
$$
\varepsilon_{ni} = 
\left[\frac{p(\x_n)}{q(\x_n)}\right]^\alpha - \delta_{0i}.
$$

Hence,   
\begin{eqnarray*}
\x_n & \sim & q_n(\x),\\
\f_n & = & \left\{\left[\frac{p(\x_n)}{q(\x_n)}\right]^\alpha -1 \right\} \Phi(q_n, \x_n),\\
q_{n+1} & = & \Psi(q_n, \gamma \f_n),
\end{eqnarray*}

It looks like this variant works similarly to the above under the same conditions (in particular, $\alpha$ needs be smaller in larger dimension), however it happens to diverge occasionally. Using the likelihood proxy, it is clearly less stable in practice than the previous version.

The underlying question is: what is the ideal sampling kernel to approximate the gradient:
$$
\u = \int q^{1-\alpha} (q^\alpha-p^\alpha) \phi
$$

The potential trouble with this version of~$\varepsilon_n$ is the division by~$q(\x_n)$. One can easily see that~$\varepsilon_n$ can be arbitrarily large -- that happens with low probability but still can and will happen in many trials. The previous version does not have this drawback, so we shall stick to it.

Generally speaking, the variance of the gradient estimator is related to:
$$
\int w(f-g)^2 \phi\phi^\top,
$$
where~$w$ is the sampling distribution, $f$ is the target factor and $g$ is the current factor fit. A sufficient condition for this integral to be finite is that both factors are bounded given that~$w$ is a proper gaussian distribution, hence $\int w \phi\phi^\top$ is finite in each element as a linear combination of moments up to order~4.

We have two options here. One is to set $w=q^{1-\alpha}$, $f=p^\alpha$, $g=q^\alpha$, in which the two factors are bounded, the other is to set $w=q$, $f=(p/q)^\alpha$ and $g=1$, in which the target factor is not necessarily bounded due to the division by~$q^\alpha$. It does not take much effort to realize which version is more risky. In fact, using~$q$ as a sampling distribution means that we simply have no convergence warranty, so it's not just risky, it's crap. This reflects the fact that, without assumptions regarding the target~$p$, we cannot reliably estimate the integral $\int (p-q)\phi$ by just looking at~$p$ in the neighborhood defined by~$q$.

The `tilting' $p\to q^{1-\alpha}p^\alpha$ is the trick to save the day but, for $\alpha\approx 1$, it becomes ineffective as the variance controlled approach tends to draw points very far off the peak, leading to very small increments and very slow convergence... We thus need to make further assumptions about the target to deal with~$\alpha$ close to one.

The VS variance when sampling from a distribution~$\pi$ is driven by the ratio $f={(p-q)^2}/{\pi}$. Specifically,
$$
V = \int f \phi \phi^\top
- gg^\top,
$$
where $g=\int (p-q)\phi$. We may see~$f$ as an unnormalized distribution. For the VS variance to be finite, this distribution needs have finite moments of order 0, 1 and 2 (in particular, it needs to be proper).

Assume now that~$\pi$ is some Gaussian prior distribution and that the likelihood $p/\pi$ is bounded. We may constrain the algorithm to approximating distributions~$q$ verifying the same condition, {\em i.e.}, $q/\pi$ is bounded. This amounts to a maximum variance constraint. If this condition holds for both~$p$ and~$q$, then $(p-q)^2 \leq \pi^2 [(\max p/\pi)^2 + (\max q/\pi)^2]$, therefore $f\leq \pi M$ verifies the same condition. This implies that we can safely sample from the prior~$\pi$, which is certainly better than sampling from the uniform distribution.

Therefore, as $\alpha\to 1$, we may implement a maximum variance constraint in the sampling kernel...

\section{Carbonetto}

This a paper from NIPS'09: \cite{Carbonetto-09}. The authors employ a stochastic approximation scheme that shares some similarity with ours. Their objective is the exclusive KL divergence $D(q_\theta\|p)$, which they attempt to minimize on an exponential family. We have:
$$
D(q_\theta\|p) \equiv \int (q_\theta \log \frac{q_\theta}{p} - q_\theta),
$$
hence,
$$
\nabla D = \int q_\theta (\log q_\theta - \log p) \phi.
$$

This is not quite the presentation of Carbonetto (they do not consider the generalized KL divergence and seem to struggle with other technicalities) but this is where their logic would lead. The potential advantage of the above expression is that the expectation is taken wrt the known fitting distribution, hence it is natural to estimate it by sampling from this same distribution. The idea is then to update $\theta$ via a quasi-Newton iteration (hence estimating the inverse Hessian by accumulating gradient estimates across iterations).

What is the difference between this scheme and ours? Essentially, the choice of divergence. At a given iteration, we consider $D(q_n^{1-\alpha}p^\alpha\|q_\theta)$, which is dependent on the current fit, they consider $D(q_\theta\|p)$, which is not. Our criterion is arguably easier to minimize, but it can also be argued that each iteration does not attempt to perform a full minimization, just to update the parameter in a sensible manner.

So... let's see if we could implement this. Rather than the damped BFGS hessian approximation, we may try and approximate it directly. We note that:
$$
\nabla \nabla^\top D = \int q_\theta \left[1 + \log \frac{q_\theta}{p} \right]\phi\phi^\top.
$$

At the optimum, we expect $q_\theta \approx p$ so the hessian is roughly:
$$
\nabla \nabla^\top D (\theta_\star) \approx \int q_\theta \phi\phi^\top.
$$

So, if we use the moving orthonormal basis, we will get a reasonable hessian approximation for free. The revisited Carbonetto algorithm becomes:
\begin{eqnarray*}
\x_n & \sim & q_n(\x),\\
\f_n & = & \big[\log p(\x_n) - \log q_n(\x_n) \big] \Phi(q_n, \x_n),\\
q_{n+1} & = & \Psi(q_n, \gamma \f_n),
\end{eqnarray*}

This looks quite similar to our algorithm, doesn't it? In fact, if we do a Taylor expansion at order~1 in ~$\alpha$ of the $\varepsilon_n$ above, what we get is:
$$
\varepsilon_n \approx \alpha \left[
\log p(\x_n) - \log q_n(\x_n)
\right],
$$
so our algorithm with step size~$\gamma$ boils down to the revisited Carbonetto algorithm with step size~$\alpha\gamma$ in the limit $\alpha\to 0$. Quite interesting, isn't it? Since the step size in Carbonetto has no reason to depend on~$\alpha$, it also suggests that our~$\gamma$ should vary in $1/\alpha$.

Does it mean my contribution is worth it zero? No!

First, we shall note a very important conceptual difference with Carbonetto's original work: they don't try to solve the same problem as us. What they try to do is to construct a sequential sampling scheme to perform standard Monte Carlo integration. They are not really trying to approximate the target distribution since it is assumed to lie in the exponential family. However, the integral of interest is not tractable and sampling from the target is not straighforward, which is why stochastic approximation comes into play. Their exponential family is not the Gaussian family as it would make their problem trivial.

Second, on the technical side of things, it is not straightforward to adapt their algorithm to our situation because of the expression they use to compute the gradient (which inherently assumes the target distribution to belong to the exponential family). The generalized KL divergence trick saves the day -- and that's the key idea behind VS. It may look trivial, but it's such a game changer that many people would have considered it before if it was so obvious. Also, we (implicitly) compute the exact Hessian in a very efficient way, while they stick to an LBGS incremental approximation, which is clearly heavier in computation and memory.

Finally, our work connects to expectation propagation and, even more closely, to its practical variants, average EP and stochastic EP.

\section{The Laplace side of things}

What happens in the star approximation algorithm if we replace the stochastic gradient update with a Laplace approximation? At a given iteration, we have a current fit~$q_0$ and we wish to approximate~$p^\alpha$ in the context~$q_0^{1-\alpha}$. A Taylor expansion around the current mode estimate~$\m_0$ may not be far off:
$$
\log p(\x) \approx a + \g^\top (\x-\m_0) + \frac{1}{2} (\x-\m_0)^\top\h \odot (\x-\m_0),
$$
with: 
$$
a \equiv \log p(\m_0),
\qquad
\g \equiv \nabla \log p,
\qquad
\h = \text{diag}(\nabla \nabla^\top) \log p.
$$

Note that this is not a fully blown second-order Taylor expansion since we approximate the hessian by its diagonal. An approximation to the tilted target~$t\equiv q^{1-\alpha}p^\alpha$ follows:
$$
\log t \approx 
(1-\alpha)\left[ \log K_0 + \frac{1}{2} (\x-\m_0)^\top\h_0 (\x-\m_0) \right]
+ \alpha \left[ \log p(\m_0) + \g^\top (\x-\m_0) + \frac{1}{2} (\x-\m_0)^\top \h (\x-\m_0) \right],
$$
with $\h_0\equiv -1/\v_0$, and where we omit the elementwise product symbol $\odot$ as it should be clear from the context.

This is of the form $\log t\approx a_1 + \g_1^\top(\x-\m_0) + (\x-\m_0)^\top\h_1(\x-\m_0)/2$ with:
$$
\h_1  = (1-\alpha) \h_0 + \alpha \h,
\qquad
\g_1 = \alpha \g,
\qquad
a_1 = (1-\alpha)\log K_0 + \alpha \log p(\m_0).
$$

So here we have our full fit update, but we still need to represent it in a fixed coordinate system. The hessian is~$\h_1$ and it is the negated inverse variance (precision) so $\v_1=-1/\h_1$. We find the mean by equating $\nabla \log t$ to zero, yielding:
$$
\m_1 = \m_0 - \g_1 / \h_1 = \m_0 + \alpha \v_1 \g .
$$

We get the constant factor $K_1$ by equating~$\log K_1$ to $\log t(\m_1)$:
$$
\log K_1 = a_1 + \g_1^\top (\m_1-\m_0) + \frac{1}{2}(\m_1-\m_0)^\top\h_1(\m_1-\m_0).
$$
Since $\m_1-\m_0 = -\g_1/\h_1$, this simplifies to:
$$
\log K_1 
= a_1 - \frac{1}{2}\g_1^\top (\g_1/\h_1)
= a_1 + \frac{\alpha^2}{2}\g^\top\v_1\g.
$$
Hence,
$$
\log K_1 = (1-\alpha)\log K_0 + \alpha \log p(\m_0) + \frac{\alpha^2}{2}\g^\top\v_1\g.
$$

To summarize, the `Laplace star' update rule is as follows:
\begin{eqnarray*}
\frac{1}{\v_{n+1}} & = & (1-\alpha)\frac{1}{\v_n} - \alpha\h \\
\m_{n+1} & = & \m_n + \alpha \v_{n+1} \g \\
\log K_{n+1} & = & (1-\alpha)\log K_n + \alpha \log p(\m_n) + \frac{\alpha^2}{2}\g^\top\v_{n+1}\g.
\end{eqnarray*}

The advantages of this algorithm over the stochastic ones considered above is that no step size is required, it's deterministic, it may be very fast (if the gradient and hessian have closed-form expressions). The drawback is that the fixed point condition is one of a saddle point of $\log p(\x)$, which is the usual crap.

\section{An algorithm for the inclusive KL divergence}

The inclusive KL divergence corresponds to the case $\alpha=1$ and is of special interest because the divergence minimization is then a convex problem and admits a unique solution under mild conditions. The problem is that our SA algorithm is pretty inefficient for $\alpha\to 1$ because it samples from too wide a distribution, hence generating most of the time $\x$ values for which the force is very small, yielding hopelessly slow convergence...

An idea to avoid this situation is to sample from a mixture distribution: $w = (1-\lambda) q + \lambda\pi$, where $\pi$ is some `prior' for which we assume that both the likelihood $p/\pi$ and the fitted likelihood $q/\pi$ are upper bounded. Since $w \geq \lambda \pi$, we find that $|p-q|/w\leq (1/\lambda)|p-q|/\pi$ is also bounded whenever $\lambda>0$. The reason for keeping the prior as part of the sampling distribution appears in its full glory: it is to guarantee bounded importance weights!

Assuming that $Z_\pi=1$,
\begin{eqnarray*}
\x_n & \sim & (1-\lambda) q_n(\x) + \lambda \pi(\x),\\
\f_n & = & \frac{p(\x_n) - q_n(\x_n)}{(1-\lambda)q_n(\x_n) + \lambda Z_n\pi(\x_n)} \Phi(q_n, \x_n),\\
q_{n+1} & = & \Psi(q_n, \gamma \f_n),
\end{eqnarray*}

\section{The learning problem}

In a typical learning scenario, the target has a factorial form:
$$
P(\x) = \prod_{i=1}^N f_i(\x),
$$
where each contribution $f_i$ arises from a single data point. The scale of such a distribution obviously depends on the training set size~$N$. However, for computational reasons, it is out of question to evaluate the full target at each iteration. We instead resort to the usual stochastic learning trick: pick a random batch of given size~$n$ and consider the `fake' target:
$$
\tilde{p}(\x) = \prod_{i=1}^n f_i(\x).
$$

This introduces a second source of randomness in the fitting algorithm. We have random data selection and random parameter generation, both of which are used for computational feasibility. We may treat $\tilde{p}(\x)$ exactly as before, as if it was a fixed target -- the only drawback that we incur is additional noise, so we might need to adjust the step size for that.

But there is another consequence of doing so, which is the implicit re-scaling of the target distribution. With a fixed batch size, the actual target considered by the algorithm is:
$$
p(\x) = P(\x)^\lambda,
\qquad \lambda \equiv \frac{n}{N}.
$$

An alternative approach would be to re-scale the `batch target' at each iteration by raising it to the power $N/n$, but that would probably cause numerical issues for very large~$N$... Intuitively, we'd better proceed as suggested using a `fake' target.

Is it a problem? Not really. We may simply fit~$p$ and then rescale the fit as a post-processing, i.e. do $q\leftarrow q^\lambda$, to get the right variance. Is it equivalent to fitting~$P$ directly? Good question. It's not, unless we use the exclusive KL divergence (the case $\alpha\to 0$).

Indeed, the exclusive KL divergence is equivariant by re-scaling, unlike other $\alpha$-divergences. It's not a convex criterion on the Gaussian family but this may also be considered a safe property: for a multimodal target, it will pick one mode rather than computing a meaningless mean. We may have a strong case for the exclusive KL divergence!

However, we may wonder about the numerical stability of directly injecting $\tilde{p}$ into the above algorithms. More generally, we may consider as a target the tilted distribution:
$p = q_n^{1-\rho} \tilde{p}^\rho$, where $\rho$ is a learning rate parameter, which is formally very similar to the previous~$\alpha$, but fulfills a different goal: leverage the noise induced by batch selection.

How does that play in the case of the Taylor-style fitting? If we have a current fit~$q$ and our current target is updated to $q^{1-\rho}\tilde{p}^\rho$, the question is how to update~$q$ so as to refine the fit... Loosely speaking, this is the same situation as before with $\rho\leftarrow\alpha$. We can forget about $\alpha$ as it is overshadowed by the learning rate in this context.

\section{Issues}

Star approx: how to choose the starting fit?

Star-Taylor: does vmax constraint work ok?

What about varying vmax across iterations (starting with a small value and increasing)?

What we were doing previously with the star Taylor version was to start from $\alpha=1$ (considered to be a learning rate) so as to avoid the initialization problem... but this may make the

\section{Discussion}

La raison pour laquelle j'ai \'echou\'e en partie \`a valoriser l'id\'ee du VS jusqu'\`a pr\'esent est que je l'ai mal situ\'ee par rapport \`a l'\'etat de l'art.

D'un c\^ot\'e, nous avons les m\'ethodes de Monte Carlo cens\'ees estimer exactement des int\'egrales pertinentes pour l'inf\'erence \`a la limite de grands \'echantillons (on peut consid\'erer dans la m\^eme cat\'egorie leur pendant d\'eterministe, les m\'ethodes de quadrature). De l'autre, nous avons les m\'ethodes d'inf\'erence variationnelle cens\'ees \^etre analytiques et donc rapides, pour lesquelles on accepte de substituer \`a l'erreur num\'erique des m\'ethodes de Monte Carlo ou de  quadrature une erreur syst\'ematique.

Le tableau est tel qu'il n'y a pas, ou tr\`es peu, de recouvrement entre ces deux mondes. Le VS se situe justement dans cette zone de recouvrement t\'enue. C'est une m\'ethode de Monte Carlo d'un genre bizarre car {\em pas n\'ecessairement asymptotiquement exacte} et une m\'ethode variationnelle d'un genre bizarre car {\em non-analytique}.

Il y a pourtant de la place pour ce genre de m\'ethodes. Les pures m\'ethodes de Monte Carlo sont peu utilis\'ees en IA car trop lentes, et les pures m\'ethodes variationnelles ne sont pas applicables en toute circonstance. C'\'etait le vide que tentait de combler le `Bayesian Monte Carlo' de Rasmussen et Ghahramani, par ailleurs peu convaincant sur le plan technique, mais peu d'auteurs se sont aventur\'es sur ce terrain. Dans le m\^eme genre d'id\'ee, il y avait aussi la m\'ethode cross-entropy, inf\'eod\'ee \`a l'importance sampling et dont on se demande si elle a d\'ej\`a produit le moindre r\'esultat tant le sch\'ema it\'eratif sur lequel elle se base est na\"if.

Mon erreur de conception du VS a \'et\'e de m'enfermer dans la contrainte d'exactitude asymptotique des m\'ethodes de Monte Carlo. Dans un monde o\`u il fallait ``choisir son erreur'', c'\'etait assez naturel, mais cette contrainte rend le VS inapplicable en grande dimension. Pour s'affranchir de cette limitation, il faut passer de la minimisation directe de la divergence de Kullback-Leibler \`a celle d'une $\alpha$-divergence pour $\alpha$ proche de z\'ero. On accepte ainsi de cumuler deux erreurs: une erreur num\'erique et une erreur syst\'ematique.

Le recours \`a la divergence~$\alpha$ est le m\^eme chemin que celui de Minka dans son papier de 2005 sur le {\em message passing}. La question qui reste en suspend pour moi, c'est pourquoi il n'\'evoque jamais le cas o\`u les messages n'ont pas d'expression analytique, ce qui arrive plus que le contraire en pratique! En fait, Minka semble avoir fait tout son possible pour \'eviter de marcher sur les plate-bandes du Monte Carlo...

L'id\'ee fondamentale de son {\em message passing}, qui g\'en\'eralise son {\em expectation propagation}, c'est l'approximation contextuelle des facteurs, mais personne ne comprend vraiment pourquoi \c ca marche, si ce n'est en constatant que \c ca donne lieu \`a des formules analytiques de mise \`a jour dans quelques cas d'int\'er\^et (comme la r\'egression probit, par exemple). Ce qui est fou, c'est que la justification sous-jacente, la divergence~$\alpha$, se trouve dans ses papiers, mais pas \`a la bonne place selon moi. J'\'evoque Minka car son papier m'a beaucoup influenc\'e en 2010--11 lorsque j'essayais de mettre en place mon histoire du VS. Il m'a pouss\'e \`a voir \`a l'\'epoque le VS comme une alternative radicale au `message passing', l'un \'etant asymptotiquement exact, l'autre rapide car exploitant la structure de la loi cible.

Ces deux pr\'ejug\'es sont faux. Le VS a peu d'int\'er\^et pratique si on le contraint \`a \^etre asymptotiquement exact. Quant au message passing, il repose sur l'approximation contextuelle qui ne requiert pas fondamentalement de structure factorielle! D\`es qu'on prend conscience de cela, on voit apparaitre une variante du `message passing' plus g\'en\'erale, dans laquelle le concept de message n'est d'ailleurs pas pertinent, th\'eoriquement plus limpide car elle minimise une divergence~$\alpha$ sous r\'eserve de convergence, beaucoup plus parcimonieuse en m\'emoire... et qui s'imbrique naturellement avec le VS, offrant une solution naturelle au probl\`eme de la poule et de l'\oe uf qui a plomb\'e mes premiers articles: comment choisir la distribution d'\'echantillonnage si on n'a pas une bonne estim\'ee de la solution?

J'ai tent\'e de r\'epondre \`a ce probl\`eme de diff\'erentes mani\`eres. La premi\`ere a \'et\'e de sugg\'erer une m\'ethode d'\'echantillonnage efficace et donc co\^uteuse, en l'occurrence annealed importance sampling, pour se mettre dans les conditions du VS de l'\'epoque. Pas si absurde car si on opte pour une m\'ethode de Monte Carlo sophistiqu\'ee, il y a mieux \`a faire que des moyennes empiriques, mais peu excitant car ce type de m\'ethode reste prohibitif en grande dimension. La seconde r\'eponse, en 2013, a \'et\'e de sugg\'erer l'approximation de Laplace comme noyau d'\'echantillonnage du VS, faisant apparaitre VS comme un l\'eger raffinement pas vraiment indispensable en pratique. La troisi\`eme r\'eponse, en 2016, a \'et\'e d'int\'egrer le VS \`a l'algorithme EP, ce qui nous rapprochait de la v\'erit\'e mais ne pr\'esentait pas d'avantage pratique marqu\'e par rapport \`a la m\'ethode de Laplace, encore elle, dans les probl\`emes convexes sur lesquels j'ai fait mes tests.

Tant qu'on reste confront\'e \`a des probl\`emes convexes, Laplace aka Newton aka Taylor aura toujours le dernier mot en raison de son efficacit\'e calculatoire imbattable. On peut toujours s'amuser \`a trouver des exemples de distributions unimodales dont l'approximation de Laplace est mauvaise (la loi de Laplace par exemple, LOL!), mais de tels cas pathologiques ne se rencontrent pas en pratique.

Le champ d'application du VS, ce sont les distributions multimodales. M\^eme s'il s'agit de faire une approximation gaussienne. La question \'etant de savoir si on pr\'ef\`ere faire une approximation de Laplace \`a partir d'un mode (ou plut\^ot d'un point selle) choisi au petit bonheur la chance, ou minimiser num\'eriquement une divergence~$\alpha$, probl\`eme convexe s'il en est et qui admet donc une solution unique. En somme: remplacer le paradigme de l'optimisation, mal pos\'e et vou\'e \`a l'\'echec pour des crit\`eres non-convexes, par l'approximation dudit crit\`ere au sens de la divergence~$\alpha$. Ou comment relaxer un probl\`eme non-convexe en probl\`eme convexe.

Comme je l'ai pressenti d\`es le d\'epart, le VS peut jouer un r\^ole dans ce changement de paradigme comme brique de base d'un solveur num\'erique analogue \`a la descente de gradient stochastique pour l'optimisation.

Ce point de vue m'a fait \'evoluer vers un algorithme de type approximation stochastique qui rappelle notamment Carbonetto et al, NIPS 2009. Quelles sont les diff\'erences notables entre ce papier et la star approximation, ma vision actuelle du VS?

\appendix

\section{Gaussian moments}

Let a univariate Gaussian variable of mean $m$ and variance $v$.
We have:
$$
\E(X^2) = m^2 + v
$$

We want to compute $\E(X^4)$. Note:
$$
X = \sqrt{v} \varepsilon + m,
$$
with $\varepsilon\sim N(0,1)$. We have: $\E(\varepsilon^2)=1$, and $\E(\varepsilon^4)={\rm Var}(\chi_1^2)+\E(\chi_1)^2 = 2+1 =3$. 
From the identity:
$$
(x + y)^4 
= x^4 + 4x^3y + 6x^2y^2 + 4xy^3 + y^4,
$$
we see that:
$$
X^4 = v^2 \varepsilon^4 + 6 v \varepsilon^2 m^2 + m^4 + \ldots,
$$
where we have omitted the monomials of odd degree in $\varepsilon$, which have zero mean. Therefore:
$$
\E(X^4) = 3 v^2 + 6 v m^2 + m^4
$$

\section{Wise man extension}

In the general algorithm for arbitrary~$\alpha$, it seems wiser to sample from $\pi^\alpha q^{1-\alpha}$ rather than $q^{1-\alpha}$, where $\pi$ is our `maximum variance' prior. By doing this, we avoid uniform sampling over~$\mathbb{R}^d$ as $\alpha\to 1$. Instead, we will be sampling from the prior, which sucks but not to infinite extents. Looks like this simply involves replacing $p$ with $p/\pi$ and $q$ with $q/\pi$ above... but it's not that simple.

Assume $q$ has parameters $K$, $m$, $v$ and $\pi$ has parameters $1$, $0$, $v_{\text{max}}$. What are the parameters of $q_\alpha = \pi^\alpha q^{1-\alpha}$?

In 1D:
$$
\log q_\alpha(x) = -\alpha\frac{x^2}{2v_{\text{max}}} + (1-\alpha)
\left[\log K - \frac{(x-m)^2}{2v} \right]
$$

We find:
$$
\frac{1}{v_\alpha}
=
\frac{\alpha}{v_{\text{max}}}
+ \frac{1-\alpha}{v}.
$$

$$
v_\alpha = \frac{v}{1-\alpha + \alpha \epsilon},
\qquad
\epsilon \equiv \frac{v}{v_{\text{max}}}
$$

$$
m_\alpha = (1-\alpha)\frac{v_\alpha m}{v}
= \frac{m}{1 + \epsilon'},
\qquad
\epsilon' \equiv \frac{\alpha}{1-\alpha} \epsilon.
$$

$$
m-m_\alpha = \frac{\epsilon' m}{1+\epsilon'}
= \frac{\alpha\epsilon}{1-\alpha+\alpha\epsilon}m
$$

$$
\log K_\alpha
= (1-\alpha)\log K + \frac{1}{2}
\left[
\frac{(m-m_\alpha)^2}{v} - \alpha \frac{m^2}{v_{\text{max}}}
\right]
$$

$$
\log K_\alpha
= (1-\alpha)\log K + 
\frac{(m-m_\alpha)^2 - \alpha \epsilon m^2}{2v} 
$$

$$
\log K_\alpha
= (1-\alpha)\log K + 
\alpha\epsilon
\left[\frac{\alpha\epsilon}{(1-\alpha + \alpha\epsilon)^2}-1
\right]
\frac{m^2}{2v} 
$$

\bibliography{stat}
\bibliographystyle{plain}

\end{document}

### 2020-03-31 notes/global_optimizer/powell_vs.tex
Grabbed global optimizer project from overleaf

--- 2018-09-13 notes/powell_vs.tex
@@
 \def\m{{\boldsymbol{\mu}}}
+\def\S{\mathbf{S}}
+\def\H{\mathbf{H}}
+\def\g{\mathbf{g}}
+\def\A{\mathbf{A}}
+\def\B{\mathbf{B}}
 \def\param{{\boldsymbol{\theta}}}
@@
 
+\section{Adding quadratic forms}
+\label{sec:quad_add}
+
+Consider a quadratic form parameterized by a tuple $(a,\m,\H)$ as follows:
+$$
+q(\x) = \alpha + \frac{1}{2} (\x-\m)^\top \H (\x-\m)
+$$
+
+Let now two quadratic forms $q_1$ and $q_2$ with respective parameters $(\alpha_1,\m_1,\H_1)$ and $(\alpha_2,\m_2,\H_2)$. What are the parameters $(\alpha,\m,\H)$ of $q_1+q_2$?
+
+\subsection{Basic formula}
+
+By definition,
+$$
+(q_1 + q_2)(\x)
+= 
+\alpha_1 + \frac{1}{2} (x-\m_1)^\top \H_1 (x-\m_1)
++ \alpha_2 + \frac{1}{2} (x-\m_2)^\top \H_2 (x-\m_2)
+$$
+
+Clearly, $\H$ is the Hessian of $q_1+q_2$. To find $\m$, we need to solve for $\nabla(q_1 + q_2)(\m)=0$. Finally, $\alpha=(q_1+q_2)(\m)$. We have:
+$$
+\nabla(q_1 + q_2)(\x)
+= 
+\H_1 (x-\m_1) + \H_2 (x-\m_2)
+$$
+and:
+$$
+\nabla\nabla^\top(q_1 + q_2)(\x)
+= 
+\H_1 + \H_2
+$$
+
+Therefore, we find that:
+$$
+\H=\H_1+\H_2,
+\quad
+\m = \H^{-1}(\H_1\m_1 + \H_2\m_2),
+\quad
+\alpha =
+\alpha_1 + \alpha_2 +
+\frac{1}{2} 
+\left[
+\|\m-\m_1\|_{\H_1}^2 + \|\m-\m_2\|_{\H_2}^2
+\right]
+.
+$$
+
+\subsection{Incremental form}
+
+We can derive the following alternative expressions of incremental nature:
+\begin{itemize}
+\item For the Hessian:
+$$
+\H = \H_1 + \Delta\H,
+\qquad 
+\Delta\H \equiv \H_2
+.
+$$
+\item For the center:
+$$
+\m = \m_1 + \Delta\m,
+\qquad 
+\Delta\m \equiv \H^{-1}\H_2(\m_2 - \m_1)
+.
+$$
+\item For the offset:
+$$
+\alpha = \alpha_1 + \Delta \alpha,
+\qquad
+\Delta \alpha \equiv \alpha_2 + \frac{1}{2}
+\left[
+\|\Delta\m\|_{\H_1}^2
++ \|\Delta_2\m\|_{\H_2}^2
+\right]
+,
+$$
+with $\Delta_2\m = \H^{-1}\H_1(\m_2-\m_1)$.
+\end{itemize}
+
+These show the final parameters as updates to the initial parameters of one of the two quadratic forms (it doesn't matter which one, of course).
+Essentially, the Hessian increment is the other quadratic Hessian, the center increment is based on the center difference, and the offset increment combines both the other offset and the center difference.
+
+Both the Hessian and center update rules are straightfoward from the above. To derive the update rule for the offset, we first note that $\m-\m_1=\Delta\m$ by definition, hence: 
+$$
+\|\m-\m_1\|_{\H_1}^2 = 
+\Delta\m^\top \H_1 \Delta\m,
+$$
+and the remainder follows by symmetry.
+
+\section{Laplace approximation}
+\label{sec:laplace_approx}
+
+Consider now another type of parameterization for a quadratic form (one that naturally follows from a Taylor expansion, for instance):
+$$
+q(\x) = \beta + \g^\top (\x-\a) + \frac{1}{2} (\x-\a)^\top \H (\x-\a)
+$$
+
+What are the $(\alpha,\m,\H)$ parameters of $q$?
+
+Obviously, the Hessian has to be the same in both forms since $\nabla\nabla^\top q = \H$.
+Using the gradient,
+$$
+\nabla q (\x) = \g + \H (\x-\a),
+$$
+we identify $\m$ via $\nabla q(\m)=0$, yielding $\H(\m-\a)=-\g$:
+$$
+\m = \a - \H^{-1}\g.
+$$
+
+Finally, we find $\alpha$ via $\alpha=q(\m)$, hence:
+$$
+\alpha = \beta - \frac{1}{2} \g^\top \H^{-1}\g
+= \beta + \frac{1}{2} \g^\top(\m-\a)
+.
+$$
+
+\section{Laplace-EP}
+
+Consider a target: 
+$$
+p(\x)=\prod_{i=1}^n f_i(\x)
+$$
+
+Assume we have some current Gaussian approximations $g_i$ of each factor and want to improve one of them. We form the cavity distribution, or context:
+$$
+c(\x) = \prod_{j\not=i} g_j(\x)
+$$
+in order to minimize $D(cf_i \| cg_i)$ over $g_i$, either exactly or approximately, yielding an unnormalized Gaussian with parameters $(\alpha_i,\m_i,\H_i)$.
+Let us denote the context parameters by $(\alpha_{-i},\m_{-i},\H_{-i})$ and the new global approximation parameters by $(\alpha,\m,\H)$.
+
+Following Section~\ref{sec:quad_add}, we have:
+\begin{enumerate}
+\item Hessian update:
+$$
+\H = \H_{-i} + \H_i,
+$$
+\item Center update:
+$$
+\m = \m_{-i} + (\H_{-i} + \H_i)^{-1}\H_i(\m_i-\m_{-i})
+$$
+\item Offset update:
+$$
+\alpha = \alpha_{-i} + \alpha_i + \ldots
+$$
+\end{enumerate}
+
+One may wonder what is the interest of updating the offset...
+If the ultimate goal of running EP is to estimate the mode and the associated uncertainty, the offset is useless.
+We would need it, for instance, to evaluate the evidence for comparison with other models (e.g., decide which CNN architecture is best for some specific task and training dataset).
+
+If we assume now that the factor approximation has a Taylor-like expression at the cavity center $\m_{-i}$, then according to Section~\ref{sec:laplace_approx} and denoting $\g_i$ the log-factor gradient, $\H_i (\m_i-\m_{-i}) = -\g_i$, we have:
+$$
+\m = \m_{-i} - (\H_{-i} + \H_i)^{-1}\g_i
+$$
+
+Doesn't it remind us of something?
+Yes, the second-order gradient descent.
+More precisely, the EP algorithm is comparable with a second-order version of {\em stochastic} gradient descent.
+Instead of taking a step propotional to the gradient using some (decaying) {\em learning rate} value, the gradient is scaled using anisotropic and adaptive curvature information.
+Updates tend to be large In the early iterations (assuming $\H\equiv 0$ initially) and get smaller as the Hessian stabilizes to a (hopefully) positive-definite matrix.
+
+As for the offset update,
+$$
+\alpha = \alpha_{-i} + \frac{1}{2} \beta_i \g_i^\top(\m_i-\m_{-i}) + \ldots
+$$
+but we said we don't care, didn't we?
+
+\section{Practical adaptations of EP}
+\label{sec:practical_ep}
+
+In large dimension, the EP algorithm is intractable for at least two reasons.
+First, the Hessian computation and inversion is too expansive.
+This turns out to be easy to work around using a fully factorized parametric form, which simply amounts to replacing the Hessian matrices with their diagonals in the computations.
+
+Another issue is parameter proliferation as EP keeps track of a complete set of parameters corresponding to each and every factor, hence requesting ridiculous memory load.
+Memory can be saved by assuming a unique approximation to all factors, yielding an EP variant known as the Averaged~EP (AEP) algorithm.
+
+On each iteration, we have a current unique factor approximation $g(\x)$ that fully determines the current {\em overall} approximation: $q(\x)=g(\x)^n$.
+Therefore, the cavity is $c(\x)=g(\x)^{n-1}$ independently from the factor under consideration, implying:
+$$
+\alpha_{-i} = \lambda \alpha,
+\quad
+\m_{-i} = \m,
+\quad
+\H_{-i} = \lambda \H,
+\quad 
+\lambda = \frac{n-1}{n}
+$$
+
+We then proceed to the factor fitting exactly as before, yielding a new overall approximation $q(\x)=c(\x)g_i(\x)$.
+However, instead of storing $g_i$, we collapse the new factor approximation to the previous ones via: $g(\x)=q(\x)^{1/n}$.
+This step makes the algorithm keep memory of all past factor approximations.
+In the standard EP, each new factor approximation simply substitutes the previous one.
+
+The problem I have with tis is: does AEP converge?
+From the center updating formula, it is clear that the condition for the algorithm to stop moving is that each factor gradient vanishes at the overall center estimate: this won't happen because the factors have  different centers (be them slightly or massively different)...
+This lack of convergence seems to arise due to the ``factor collapsing" step.
+
+We might need to introduce some decay to make convergence happen.
+One way to go is to say: OK, let's consider some scaling $f_i^\epsilon$, with $0<\epsilon\leq 1$ , as the factor under consideration, meaning that the context is made up of ``$n-\epsilon$ factors'', hence $\lambda=(n-\epsilon)/n$.
+The gradient and Hessian of the log-factor are respectively $\epsilon \g_i$ and $\epsilon \H_i$, hence the update rules generalize to:
+$$
+\H \leftarrow (1-\epsilon/n)\H + \epsilon \H_i,
+\qquad
+\m \leftarrow \m - \epsilon\H^{-1}\g_i
+.
+$$
+
+Clearly, when the decay parameter $\epsilon\to 0$, these updates are effectless hence the algorithm converges. But does it converge to a reasonable solution? That's the question. More on this later.
+
+\section{EP as Game Theory}
+
+It occured to me that EP can be viewed as a game, where each factor is associated with a player whose aim is to fit a parametric form to the factor.
+Each player's payoff depends on all other players as the fit needs to be more accurate in the subspace considered most important collectively.
+
+Player interference arises in a mere alternate minimization of the KL divergence.
+The key insight into EP is to make the ``where'' task much easier than in the alternate KL minimization approach as it essentially provides a relevance window function that is parametric, hence greatly simplifying each fitting problem.
+
+From this standpoint, the question of convergence of EP boils down to the existence of a Nash equilibrium...
+
+How about AEP?
+The difference with EP is that players do not see what the other players are doing.
+They only get aware of the overall fit, which is updated after each player's move but is not specific to one player.
+One way to account for this is to introduce an additional player that we may call the mediator.
+He's the only player the others speak to and he can speak with all of them.
+When a player makes a move, the mediator can reflect it.
+Problem is: the mediator's strategy depends on which factor he speaks to, so it doesn't seem to be standard game theory.
+
+\section{AEP as Stochastic Approximation}
+
+AEP is better understood in terms of an iterative algorithm minimizing factor-dependent costs $D_i(q,q')$ and performing updates $q\to F_i(q)$ where  $F_i(q)\equiv \arg\min_{q'}D(q^\lambda f_i\|q')$ with $q'=q^\lambda g$ or $F_i(q)=q^\lambda g$ with $g$ the Laplace approximation of $f_i$ around the center of $q$.
+
+So it's a weird kind of ``game'' where players always move together.
+It's not in the framework of Nash equilibrium theory.
+Convergence in this scenario relies on the existence of a multiple fix point $q=F_1(q)=\ldots=F_n(q)$.
+This is too much to ask, insn't it?
+If all factors were the same, all transition functions would be the same and that could work, but there is no way this will happen in practice.
+
+One important property of these transition functions is that they may all be seen as noisy versions of the same ``ideal'' transition function: ${\cal F}(q)=q_\star=\arg\min_{q'}D(p\|q')$ which is actually independent from $q$ and is the Holy Grail distribution this whole stuff is about trying to approximate.
+So, whenever we update~$q$, the result is slightly or even completely off, and it's only by averaging updates across iterations that we can achieve an accurate result.
+
+Therefore, the iteration will jump from one update to another frantically unless it has some memory. Take a look at the update formula in Section~\ref{sec:practical_ep}. Defining:
+$$
+F_i(\m, \H) = (\m-\H^{-1}\g_i, \lambda\H+\H_i),
+$$
+they correspond to a Robbins–Monro algorithm to track the fix point of $E[F_i(\m,\H)]=(\m,\H)$ where $\epsilon$ can be chosen according to the Blum theorem to guarantee convergence.
+Bascially, $\sum_n\epsilon_n=\infty$ and $\sum_n\epsilon_n^2<\infty$.
+
+Keras stochastic gradient implementation seems to use a time-based decay of the form:
+$$
+\epsilon_n = \frac{\epsilon_0}{1 + \alpha n},
+$$
+where $\epsilon_0$ is the initial learning rate and $\alpha$ is the decay value, therefore $\epsilon_n=O(1/n)$ and satisfies the Blum condition.
+
+Keras also shuffles the training data at the beginning of each epoch. This is not compliant with the native EP algorithm but is natural and easy to do with AEP. So shuffle or not shuffle?
+
+\section{Short biblio on stochastic gradient descent}
+
+\subsection{First-order SGD}
+
+The plain steepest descent looks like this:
+$$
+\x \leftarrow \x - \gamma_i \g_i
+,
+$$
+where $\gamma_i$ is an adaptive learning rate parameter and $\g_i$ is the stochastic gradient.
+
+The so-called RMS propagation implemented in Keras normalizes the gradient using some running average of its magnitude:
+$$
+a^2\leftarrow \rho a^2 + (1-\rho) \|\g_i\|^2,
+\qquad
+\x \leftarrow \x - \frac{\gamma_i}{a + \epsilon} \g_i,
+$$
+where $0\leq\rho\leq 1$ defaults to 0.9 and $\epsilon$, which defaults to $10^{-7}$, is  here to avoid divisions by zero.
+The learning rate is computed as $\gamma_i = \gamma/(1 + \alpha i)$ with $\gamma=10^{-3}$ (initial learning rate) and $\alpha=0$ (decay) by default.
+
+\subsection{Second-order SGD}
+
+Bottou (2010) defines 2nd order SGD as an iteration of the form:
+$$
+\x \leftarrow \x - \gamma_i \Gamma_i \g_i,
+$$
+where $\Gamma_i$ is a ``positive definite matrix approaching the inverse of the Hessian'' and $\gamma_i$ is some decay parameter which, like in 1st order SGD, is needed to achieve convergence under the single-point approximation of the gradient (the ``stochastic'' part).
+But he notes that ``this modification does not reduce the stochastic noise and therefore does not significantly improve the variance" (of the parameter estimate).
+
+Note sure if everyone agrees with that.
+It seems that some researchers still believe that 2nd order SGD is potentially faster to converge than its 1st order counterpart.
+One issue in the context of SGD is to come up with an effective online approximation to the Hessian.
+Then we might want to use the conjugate gradient (CG) method to solve the linear system associated with the SGD update, rather than inverting the Hessian itself.
+In practice, CG can be run for a few iterations to reach a good enough approximation -- this is known as the {\em Hessian-free} approach (HF).
+See James Martens (2010).
+
+So what kind of online Hessian estimates can be used?
+Martens simply use the Hessian of a large enough mini-batch based on purely empirical arguments.
+Argawal et al (2017) in a paper called {\em Second-Order Stochastic Optimization for Machine Learning in Linear Time} revisit the question and propose an algorithm called LiSSA (linear time second-order stochastic algorithm) that estimates the Hessian inverse incrementally.
+
+The trick is to exploit the famous formula:
+$$
+\A^{-1} 
+= \sum^{\infty}_{i=0} (I-\A)^i 
+$$
+
+In other words,
+$$
+\s_0 \equiv \g,
+\qquad
+\s_j \equiv \sum_{i=0}^j (I-\A)^{i}\g
+$$
+converges to $\A^{-1}\g$ as $j\to\infty$.
+The key observation is that $\s_j$ can be computed recursively via $\s_j=\g + (I-\A)\s_{j-1}$, giving rise to a linear solver algorithm that can be seen as an alternative to CG.
+Now, we would like to replace $\A$ with the Hessian but it's more effective to estimate the Hessian from a mini-batch.
+More specifically, the paper uses a different mini-batch at each step ``$j$'' of the recursion -- which makes it somewhat more sophisticated than the plain HF mini-batch approach.
+
+To summarize, the fully factorial EP standpoint provides two things: 1) a neat justification to the diagonal approximation to the Hessian; 2) an online Hessian estimate that implicitely depends on all datapoints and likely becomes stable across iterations (unlike the gradient estimate).
+
+\section{Robbins-Monro algorithm}
+
+A brief refresher. The Robbins-Monro algorithm is a stochastic approximation method that uses an iteration of the form:
+$$
+\x_{n+1} = \x_n - \epsilon_n f_n(\x_n)
+,
+$$
+in order to solve for $f(\x)=E[f_n(\x)]=0$, assuming that $f_n(\x)$ are random functions with same mean $f(\x)$.
+Equivalently, the iteration aims to find a fix point of $F(\x)=\x-f(\x)$ and also reads:
+$$
+\x_{n+1} = (1-\epsilon_n)\x_n + \epsilon_n F_n(\x_n),
+$$
+with $F_n(\x)\equiv \x - f_n(\x)$.
+
+Now, consider the computation of a sample mean:
+$$
+\x_n = \frac{1}{n}\sum_{i=1}^n \a_{i-1}
+.
+$$
+
+We note that $(n+1)\x_{n+1} = n\x_n + \a_n$ hence $\x_{n+1}=[n\x_n+a_n]/(n+1)$, so this is equivalent to the following iteration:
+$$
+\x_{n+1} = \x_n - \frac{1}{n+1} (\x_n-\a_n),
+$$
+which is of Robbins-Monro type with $f_n(\x)=\x-\a_n$, or $F_n(\x)=\a_n$, where $\a$ is the random variable, and $\epsilon_n=1/(n+1)$.
+The above general considerations confirm that this iteration is meant to track $\x=E(\a)$.
+
+But it also shows, by substituting formally $\a_n$ with $F_n(\x_n)$, that the Robins-Monro iteration with $\epsilon_n=1/(n+1)$ computes the average of $F_1(\x_1), F_2(\x_2), \ldots, F_{n-1}(\x_{n-1})$.
+
+If it turns out that $F$ is constant, RM is just a way to compute a sample mean.
+More generally, the Robbins-Monro may be seen as a procedure to estimate some quantity from repeated noisy measurements.
+
+\section{Some more stuff about learning}
+
+Why using placeholders?
+Apparently, people like placeholders because it makes it possible to implement a computational graph without using any data.
+So we can decorrelate the network architecture from the training.
+I can see a potential advantage in terms of memory management but it looks like Keras does not use it (to fit a model using Keras, the whole training dataset shall be loaded into memory beforehand).
+
+Dynamic learning phase? What is it? No clue.
+
+How does Keras initialize the network weights?
+
 \bibliographystyle{abbrv}

### 2020-03-31 notes/global_optimizer/sgd_ep.tex
Grabbed global optimizer project from overleaf

\documentclass{article}

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,subfigure}

\title{Point matching notes} 
\author{Alexis Roche} 
%\date{}

\begin{document}

\maketitle

\section{Newton method and expectation-propagation}

EP may be seen as a generalization of Newton's method.

\section{Gaussian updates}

Let $c(x)=K \exp[-\frac{1}{2} (x-\mu)^\top \Sigma^{-1} (x-\mu)]$ in canonical form and $g(x)$ such that $-\log g(x)=a + g^\top (x-\mu) + \frac{1}{2} (x-\mu)^\top H (x-\mu)$. Let us give the canonical expression of $c(x)g(x)$:
\begin{eqnarray*}
\log cg (x)
& = & 
\log K -\frac{1}{2} (x-\mu)^\top \Sigma^{-1} (x-\mu)
- a - g^\top (x-\mu) - \frac{1}{2} (x-\mu)^\top H (x-\mu) \\
& = & 
\log K -a - g^\top (x-\mu) - \frac{1}{2} (x-\mu)^\top (\Sigma^{-1} + H) (x-\mu) \\
& = & 
\log K -a - g^\top (x-\mu) - \frac{1}{2} (x-\mu)^\top \Sigma_n^{-1} (x-\mu),
\end{eqnarray*}
with the new covariance matrix:
$$
\Sigma_n = (\Sigma^{-1} + H)^{-1}
.
$$

To find the new mean, just cancel the gradient of the above expression:
$$
\nabla \log cg 
= 
- g - \Sigma_n^{-1} (x-\mu),
$$
yielding:
$$
\mu_n = \mu - \Sigma_n g
.
$$

Finally, let us evaluate the multiplicative constant. By definition,
$$
\log K_n = \log cg(\mu_n),
$$
hence, denoting $\delta=\mu_n-\mu=-\Sigma_n g$:
$$
\log K_n = \log K -a - g^\top \delta - \frac{1}{2} \delta^\top \Sigma_n^{-1} \delta
$$

$$
\log K_n = \log K -a + \frac{1}{2} g^\top \Sigma_n g.
$$

\end{document}

### 2020-05-06 cv/cv.tex
CV update

@@
 Strong experience in developing research and commercial software.
-{\em Keywords:} artificial intelligence, image processing, computer vision, machine learning, deep learning, statistics, brain imaging, Python.
+{\em Keywords:} artificial intelligence, image processing, computer vision, machine learning, deep learning, statistics, brain imaging, Python, C.
 
@@
 {\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 5754 citations, h-index: 32, i10-index: 56 (September 2019).\\
+{\tiny$\bullet$} Google scholar statistics: 6102 citations, h-index: 34, i10-index: 62 (April 2020).\\
 {\tiny$\bullet$} 5 published international patents. \\

### 2020-09-28 cv/cv.tex
Some CV modifs

@@
 
-\noindent Senior-level engineer and applied researcher with a specialist expertise in computer vision, medical image analysis and machine learning, operating in the fields of applied research, healthcare, electro-domestic and entertainment industry.
-Graduated from the French ``Grande \'Ecole" CentraleSupelec (formely, Centrale Paris), and holder of a PhD.
-Strong experience in developing research and commercial software.
-{\em Keywords:} artificial intelligence, image processing, computer vision, machine learning, deep learning, statistics, brain imaging, Python, C.
+\noindent Scientific team leader with a specialist expertise in computer vision, data science and machine learning, who has operated in the fields of academic research, healthcare, electro-domestic and entertainment industry.
+Graduated from \'Ecole Centrale Paris, and holder of a PhD in Engineering Science.
+Strong experience in the development of both open source and commercial software.
+{\em Keywords:} artificial intelligence, image processing, computer vision, medical image analysis, machine learning, deep learning, statistics, Python,~C.
 
@@
 \ \\
-\noindent \textbf{NEUROIMAGING:} Extensive experience in developing algorithms for structural and functional brain magnetic resonance imaging (MRI) in the form of open-source and commercial software for neuroscience, clinical research \& clinical practice.
-
-\ \\
 \noindent \textbf{SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.
@@
 \ \\
-\noindent \textbf{INTERACTION:} Proven ability to work in multidisciplinary environments at the interface of industry and academic research, as well as to develop successful collaborations with physicians and fundamental life science researchers.
+\noindent \textbf{SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 published international patents.
 
 \ \\
-\noindent \textbf{SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 international patents.
+\noindent \textbf{TECHNICAL LEADERSHIP:} Proven ability to coordinate a team and manage objectives in multidisciplinary environments.
+
+%\ \\
+%\noindent \textbf{NEUROIMAGING:} Extensive experience in developing algorithms for structural and functional brain magnetic resonance imaging (MRI) in the form of open-source and commercial software for neuroscience, clinical research \& clinical practice.
 
@@
 
-\bighead{SENIOR COMPUTER VISION ENGINEER}{Didimo, Porto, Portugal}{2019--present}
+\bighead{COMPUTER VISION \& ARTIFICIAL INTELLIGENCE TEAM LEADER}{Didimo, Porto, Portugal}{2019--present}
 \nopagebreak
 \begin{itemize}
-\item Developing computer vision algorithms for high-fidelity 3D avatars.
+\item Developing computer vision and machine learning algorithms for high-fidelity 3D avatars.
 \end{itemize}
@@
 {\tiny$\bullet$} C/C++ \\
-{\tiny$\bullet$} Python (numpy, scipy, pylab, skimage, opencv, sklearn, TensorFlow, keras, cython...)\\ 
+{\tiny$\bullet$} Python (numpy, scipy, pylab, skimage, opencv, sklearn, TensorFlow, Keras, ...)\\
+{\tiny$\bullet$} C/Python integration via Cython\\
 {\tiny$\bullet$} Matlab\\
-{\tiny$\bullet$} R. \\
-{\tiny$\bullet$} Version control software (git, svn).\\
+{\tiny$\bullet$} R \\
+{\tiny$\bullet$} Version control software: Git, SVN, Perforce. \\
 \end{minipage}
@@
 {\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 6102 citations, h-index: 34, i10-index: 62 (April 2020).\\
+{\tiny$\bullet$} Google scholar statistics: 6396 citations, h-index: 35, i10-index: 64 (September 2020).\\
 {\tiny$\bullet$} 5 published international patents. \\

### 2020-11-05 cv/cv.tex
Update CV

@@
 \url{https://www.linkedin.com/in/alexis-roche-3815592}\\
-French National $\mid$ DOB 12/8/1972 $\mid$ Married, 2 Children\\
+%French National $\mid$ DOB 12/8/1972 $\mid$ Married, 2 Children\\
 \end{small}
@@
 
-\bighead{COMPUTER VISION \& ARTIFICIAL INTELLIGENCE TEAM LEADER}{Didimo, Porto, Portugal}{2019--present}
+\bighead{HEAD OF COMPUTER VISION \& ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
 \nopagebreak
@@
 {\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 6396 citations, h-index: 35, i10-index: 64 (September 2020).\\
+{\tiny$\bullet$} Google scholar statistics: 6147 citations, h-index: 32, i10-index: 62 (October 2020).\\
 {\tiny$\bullet$} 5 published international patents. \\

### 2021-01-07 cv/cv.tex
update cv

@@
 \begin{itemize}
-\item Developing computer vision and machine learning algorithms for high-fidelity 3D avatars.
+\item Developing computer vision and machine learning algorithms for reconstruction of high-fidelity digital humans from photos.
 \end{itemize}
@@
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} C/C++ \\
-{\tiny$\bullet$} Python (numpy, scipy, pylab, skimage, opencv, sklearn, TensorFlow, Keras, ...)\\
-{\tiny$\bullet$} C/Python integration via Cython\\
-{\tiny$\bullet$} Matlab\\
-{\tiny$\bullet$} R \\
-{\tiny$\bullet$} Version control software: Git, SVN, Perforce. \\
+{\tiny$\bullet$} Programming languages: C, Python, Matlab, R. \\
+{\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, TensorFlow, keras.\\
+{\tiny$\bullet$} C/Python integration via cython.\\
+{\tiny$\bullet$} Version control software: git, svn, perforce. \\
 \end{minipage}
@@
 {\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 6147 citations, h-index: 32, i10-index: 62 (October 2020).\\
+{\tiny$\bullet$} Google scholar statistics: 6225 citations, h-index: 32, i10-index: 63 (December 2020).\\
 {\tiny$\bullet$} 5 published international patents. \\

### 2021-01-13 cv/cv.tex
Update cv

@@
 %French National $\mid$ DOB 12/8/1972 $\mid$ Married, 2 Children\\
+French National $\mid$ Married, 2 Children\\
 \end{small}
@@
 \item Main supervisor of two PhD students and four MSc students.
-\item Worked as Academic Guest at Computer Vision Laboratory, Swiss Federal Institute of Technology Zurich (ETHZ), Switzerland, 2009--2011, on brain image registration. 
+\item Academic Guest at Computer Vision Laboratory, Swiss Federal Institute of Technology Zurich (ETHZ), Switzerland, 2009--2011.
 \end{itemize}
@@
 \ \\
-\bighead{RESEARCH ASSISTANT -- WOLFSON MEDICAL VISION LABORATORY}{University of Oxford, UK}{2001--2002}
+\bighead{POST-DOCTORAL RESEARCHER -- WOLFSON MEDICAL VISION LABORATORY}{University of Oxford, UK}{2001--2002}
 \nopagebreak
@@
 \item Developed advanced medical image registration algorithms.
-\item Consulting work for Mirada Solutions Ltd (now Siemens Molecular Imaging) on image-based deformation tracking (brain, liver).
+\item Consulting for Mirada Solutions Ltd (now Siemens Molecular Imaging) on image-based deformation tracking (brain, liver).
 \end{itemize}
@@
 \nopagebreak
-\head{PhD, Engineering Science (1997--2001)}{University of Nice-Sophia Antipolis, France}
+\head{PhD, Engineering Science (1997--2001)}{French National Research Institute (INRIA) \& University of Nice-Sophia Antipolis, France}
 \begin{itemize}
-\item Funded by French National Research Institute (INRIA), Sophia Antipolis unit.
-\item Developed generic algorithm for multimodal image registration, with applications in radiotherapy, image-guided surgery and neuroscience.
+\item Development of multimodal image registration algorithms, with applications in radiotherapy, image-guided surgery and neuroscience. With highest honor.
 \end{itemize}
@@
 \ \\
-\ \\ 
 \end{minipage}
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} Programming languages: C, Python, Matlab, R. \\
+{\tiny$\bullet$} Languages: C, Python, Matlab, R. \\
 {\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, TensorFlow, keras.\\

### 2021-03-05 cv/cv.tex
update cv

@@
 
-\noindent Scientific team leader with a specialist expertise in computer vision, data science and machine learning, who has operated in the fields of academic research, healthcare, electro-domestic and entertainment industry.
-Graduated from \'Ecole Centrale Paris, and holder of a PhD in Engineering Science.
+\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD, with a specialist expertise in computer vision, data science and machine learning.
+Has operated in the fields of academic research, healthcare, electro-domestic and entertainment industry.
 Strong experience in the development of both open source and commercial software.
@@
 {\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 6225 citations, h-index: 32, i10-index: 63 (December 2020).\\
+{\tiny$\bullet$} Google scholar statistics: 6322 citations, h-index: 32, i10-index: 65 (March 2021).\\
 {\tiny$\bullet$} 5 published international patents. \\

### 2021-04-07 cv/cv.tex
CV update

@@
 {\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 6322 citations, h-index: 32, i10-index: 65 (March 2021).\\
+{\tiny$\bullet$} Google scholar statistics: 6376 citations, h-index: 32, i10-index: 65 (April 2021).\\
 {\tiny$\bullet$} 5 published international patents. \\

### 2021-07-16 cv/cv.tex
Update cv

@@
 \begin{minipage}{.15\textwidth}
-\textbf{Programming:}
+\textbf{Computer skills:}
 \ \\ 
@@
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} Languages: C, Python, Matlab, R. \\
-{\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, TensorFlow, keras.\\
+{\tiny$\bullet$} Programming languages: C, Python, Matlab, R. \\
+{\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, tensorflow, keras.\\
 {\tiny$\bullet$} C/Python integration via cython.\\
 {\tiny$\bullet$} Version control software: git, svn, perforce. \\
+{\tiny$\bullet$} Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs. \\
 \end{minipage}
@@
 \ \\
-\ \\
 \end{minipage}
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} 115 scholarly publications.\\
-{\tiny$\bullet$} Full list available at ORCID: \url{http://orcid.org/0000-0002-4821-6893} \\
-{\tiny$\bullet$} Google scholar statistics: 6376 citations, h-index: 32, i10-index: 65 (April 2021).\\
-{\tiny$\bullet$} 5 published international patents. \\
+{\tiny$\bullet$} 115 scholarly publications. Full list: \url{http://orcid.org/0000-0002-4821-6893}. \\
+{\tiny$\bullet$} Google scholar statistics: 6602 citations, h-index: 33, i10-index: 66 (July 2021).\\
+{\tiny$\bullet$} 5 published international patents. 2 ongoing applications.\\
 \end{minipage}
@@
 \begin{minipage}{.15\textwidth}
-\textbf{Teaching:}
+\textbf{Academic work:}
+\ \\ 
+\ \\
 \ \\
@@
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} Brain morphometry post-graduate lecture (2h), Lausanne: UNIL (2014/16), EPFL (2015/16).\\
-{\tiny$\bullet$} Statistical analysis in neuroimaging, INSERM/CNRS continuing education, France (10h/year from 2003 to 2015).\\
+  {\tiny$\bullet$} Post-graduate teaching in computer vision and medical image analysis: about 100~hours (Ecole Centrale Paris, EPFL Lausanne, University of Nice, University of Lausanne, INSERM/CNRS continuing education, \ldots).\ \\
+  {\tiny$\bullet$} Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\ \\
 \end{minipage}

### 2021-08-13 cv/cv.tex
Modif Nenuphars

@@
 
-\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD, with a specialist expertise in computer vision, data science and machine learning.
+\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD in engineering science, with a specialist expertise in computer vision, data science and machine learning.
 Has operated in the fields of academic research, healthcare, electro-domestic and entertainment industry.
 Strong experience in the development of both open source and commercial software.
-{\em Keywords:} artificial intelligence, image processing, computer vision, medical image analysis, machine learning, deep learning, statistics, Python,~C.
+{\em Keywords:} applied mathematics, artificial intelligence, image processing, computer vision, medical image analysis, machine learning, deep learning, statistics, Python,~C.
 

### 2021-09-02 cv/cv.tex
Add CV in PDF format
Update CV

@@
 \ \\
-\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris (now CentraleSupelec), France}
+\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris (now CentraleSupelec, University Paris-Saclay), France}
 \nopagebreak
@@
 \begin{minipage}{.15\textwidth}
-\textbf{Computer skills:}
+\textbf{IT skills:}
+\ \\ 
+\ \\
+\ \\ 
 \ \\
@@
 \begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} Programming languages: C, Python, Matlab, R. \\
-{\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, tensorflow, keras.\\
-{\tiny$\bullet$} C/Python integration via cython.\\
+{\tiny$\bullet$} Programming languages: Python, C, C++, Matlab, R. \\
+{\tiny$\bullet$} Long-term experience with scientific Python pacakges (numpy, scipy, pylab).\\
+{\tiny$\bullet$} CV/AI Python packages: skimage, PIL, opencv, sklearn, pytorch, tensorflow, keras.\\
+{\tiny$\bullet$} C/Python integration via Cython.\\
 {\tiny$\bullet$} Version control software: git, svn, perforce. \\
+{\tiny$\bullet$} Agile workflow (mainly Scrum) using JIRA \& Confluence. \\
 {\tiny$\bullet$} Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs. \\

### 2021-10-06 cv/cv.tex
CV updates

@@
 \ \\
-\noindent \textbf{SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.
+\noindent \textbf{SCIENTIFIC COMPUTING:} 15 years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.
 
@@
 \begin{itemize}
-\item Developing computer vision and machine learning algorithms for reconstruction of high-fidelity digital humans from photos.
+\item Leading development of computer vision and machine learning algorithms for generation of high-fidelity digital humans from photos (facial reconstruction, texture image enhancement, animation retargetting...).
 \end{itemize}
@@
 \begin{itemize}
-\item Developed embedded object recognition and tracking algorithms for smart domestic appliances using traditional computer vision and deep learning (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017--18).
+\item Developed embedded image classification, object recognition and tracking algorithms for smart domestic appliances using traditional computer vision and deep learning (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017--18).
 \end{itemize}
+\ \\
 
-\ \\
 \bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
@@
 \item Led algorithmic development of the {\em MorphoBox} Siemens prototype for brain morphometry using anatomical MRI to help radiological reading for patients with suspected neurodegeneration.
-\item Contributed a number of clinical validation studies of brain morphometry in collaboration with neurologists and radiologists, via statistical analysis and data mining techniques, in Alzheimer's disease and other dementia, mild cognitive impairment, multiple sclerosis, migraine, anorexia, \ldots
-\item Co-supervised three PhD students and two MSc students.
+\item Contributed a number of clinical validation studies of brain morphometry in collaboration with neurologists and radiologists, via statistical analysis and data mining techniques (in Alzheimer's disease and other dementia, mild cognitive impairment, multiple sclerosis, migraine, anorexia, \ldots
+\item Co-supervised three PhD students and two MSc students).
 \item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
@@
 {\tiny$\bullet$} 115 scholarly publications. Full list: \url{http://orcid.org/0000-0002-4821-6893}. \\
-{\tiny$\bullet$} Google scholar statistics: 6602 citations, h-index: 33, i10-index: 66 (July 2021).\\
+{\tiny$\bullet$} Google scholar statistics: 6698 citations, h-index: 33, i10-index: 67 (October 2021).\\
 {\tiny$\bullet$} 5 published international patents. 2 ongoing applications.\\

### 2022-01-31 papers/variana/notes.tex
Include custom optimizer in VS notes

--- 2020-03-31 notes/global_optimizer/main.tex
@@
 
-\title{Global optimizer}
+\def\x{\mathbf{x}}
+\def\th{{\boldsymbol{\theta}}}
+\def\n{{\boldsymbol{\nu}}}
+\def\u{\mathbf{u}}
+\def\m{\mathbf{m}}
+\def\v{\mathbf{v}}
+\def\f{\mathbf{f}}
+\def\E{\mathbb{E}}
+\def\g{\mathbf{g}}
+\def\h{\mathbf{h}}
+\def\A{\mathbf{A}}
+\def\B{\mathbf{B}}
+
+\title{Variana Reloaded}
 \author{Alexis Roche}
@@
 
-\section{Introduction}
+\section{Variational sampling recap}
+
+Consider:
+\begin{itemize}
+\item a reference distribution~$\pi(x)$ (from which we can sample)
+\item a factor~$f(x)$ so that $p\equiv \pi f$ is a distribution of interest
+\item an approximating factor~$g$ searched under the exponential form:
+  $$
+  g(x) = e^{\theta^\top \phi(x)},
+  $$ where $\phi$ is a known vector-valued function and $\theta$ needs
+  to be determined.
+\end{itemize}
+
+VS is a stochastic method to approximate the parametric factor~$g$ that minimizes the generalized Kullback-Leibler divergence:
+$$
+D(\pi f\|\pi g)
+= \int \pi (f \log \frac{f}{g} + g - f)
+$$
+
+It does so by sampling independent points $(x_1,\ldots,x_n)$ from~$\pi$ and minimizing a Monte Carlo approximation to the KL~divergence, leading to the surrogate criterion:
+$$
+\tilde{D}(\theta) = \frac{1}{n} \sum_i \ell[ f(x_i), g_\theta(x_i) ],
+$$
+where $\ell(f, g)$ is the loss function associated with the KL~divergence:
+$$
+\ell(f, g) \equiv f \log \frac{f}{g} - f + g
+$$
+
+We have proved in our initial paper from 2013 that the resulting estimator of the integral, or moment, $I(\phi) \equiv \int \pi f \phi$, is asymptotically unbiased with variance matrix:
+$$
+\Sigma_n = \frac{1}{n} \int \pi (f - g_\star)^2 \phi \phi^\top,
+$$
+where $g_\star$ is the genuine (intractable) KL~minimizer:
+$$
+g_\star \equiv \arg\min_g D(\pi f\|\pi g).
+$$
+
+This makes VS fundamentally different from importance sampling, for which the moment is directly estimated via:
+$$
+\hat{\mu} \equiv \frac{1}{n} \sum_i f(x_i)\phi(x_i),
+$$
+yielding an unbiased estimator with asymptotic variance:
+$$
+\Sigma_n^{IS} = \frac{1}{n} \left[
+\int \pi f^2 \phi \phi^\top
+- I(\phi) I(\phi)^\top
+\right]
+$$
+
+In particular, VS has the key consistency property unlike IS to be exact when the target factor lies in the approximating family.
+
+\section{Variance upper bound}
+
+Intuitively, the closer the target factor~$f$ to the approximating family, the smaller the VS~asymptotic variance.
+To show that formally, we would like to upper bound the mean square error $\int\pi(f-g)^2$ by some increasing function of the KL~divergence $D(\pi f\|\pi g)$.
+
+I have been bumping into this issue for years...
+only to realize now (January 2022) that such an upper bound does exist and is actually pretty simple to exhibit.
+We will need a straightforward intermediate result:
+
+\paragraph{Alexis theorem:} if $f:\Omega\in\mathbb{R}\to \mathbb{R}$ is a smooth convex function from some open interval reaching its minimum in $a\in\Omega$, assuming for simplicity that $f(a)=0$, and such that the second derivative $f''$ is a decreasing function, then we have that:
+$$
+\forall x \leq a,
+\qquad
+f(x) \geq \frac{1}{2} f''(a) (x - a)^2 .
+$$
+
+The proof is straightforward from the fact that $f''(x) \geq f''(a)$ since $f''$ is decreasing, then we just need to integrate this relationship twice and take into account that $f'(a)=f(a)=0$.
+
+Next, let us consider the loss function $\ell(f,g)=f\log(f/g)+g-f$ as a function of $g$. We easily see that:
+$$
+\frac{\partial\ell}{\partial g} = 1 - \frac{f}{g},
+\qquad
+\frac{\partial^2\ell}{\partial g^2} = \frac{f}{g^2},
+$$
+implying that the loss is not only convex in~$g$ but also staisfies the conditions of Alexis's theorem with $a=f$. Hence,
+$$
+\forall g\leq f,
+\qquad
+\ell(f, g) \geq \frac{1}{2f} (g - f)^2. 
+$$
+
+But we can also consider the loss as a function of~$f$ at fixed~$g$. We have:
+$$
+\frac{\partial\ell}{\partial f} = \log(f),
+\qquad
+\frac{\partial^2\ell}{\partial g^2} = \frac{1}{f},
+$$
+which means that Alexis's theorem can be applied again, this time with $a=g$, yielding:
+$$
+\forall f\leq g,
+\qquad
+\ell(f, g) \geq \frac{1}{2g} (f - g)^2. 
+$$
+
+Concatenating both results, we see that for all $(f, g)$:
+$$
+(f-g)^2 \leq 2 \max(f, g)\ell(f, g).
+$$
+
+Integrating this over~$\pi$, we get this long-waited inequality:
+$$
+\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g).
+$$
+
+This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes.
+But this holds under the mild condition that both $f$ and $g$ remain bounded.
+This practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.
+
+Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.
+
+The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi$. The above result implies straighforwardly:
+$$
+(f-g)\phi^2
+\leq 2 \max(f\phi, g\phi) \ell(f\phi, g\phi)
+= 2 \max(f\phi, g\phi) \phi\ell(f, g)
+= 2 \max(f\phi^2, g\phi^2) \ell(f, g)
+$$
+where we use the fact that $\ell(f\phi,g\phi)=\phi\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
+$$
+\int \pi (f-g)^2\phi^2 \leq 2\max(\|f\phi^2\|_\infty, \|g\phi^2\|_\infty) D(\pi f\|\pi g).
+$$
+
+This yields the same kind of variance upper bound for any moment estimator we have.
+
+\section{Custom optimizer}
+
+In the following, we consider the canonical problem of minimizing $D(p\|q)$ for some target~$p$, wrt some distribution log-quadratic distribution~$q$.
+I wrote the following section at Covii in Spring~2019.
+The goals is to have a scalable optimizer for VS, and the core idea is to re-parameterize the approximating family locally so as to implement~VS using a first-order gradient descent rather than a quasi-Newton method, which requires maninpulations on some Hessian approximation in one way or another and may thus be challenging in very large dimension.
+
+\subsection{Basis change}
+
+Consider minimizing the KL divergence via a gradient descent and consider a single gradient update.
+If we cannot afford to compute the Hessian (and to invert it), the gradient update relies on an arbitrary metric and may be rather ineffective.
+An important property of the KL divergence with exponential families is that its Hessian is independent from the target distribution and is associated with a dot product on the linear space spanned by~$\phi$ (which is a subspace of polynomials of order~2 for the factorial Gaussian family):
+$$
+H = \int q(\x) \phi(\x) \phi(\x)^\top d\x 
+\quad \Rightarrow \quad
+H_{ij} = \langle \phi_i , \phi_j \rangle .
+$$
+
+A natural idea is then to pick an orthonormal basis to use temporarily for a single gradient update as the divergence Hessian is scalar in such a basis.
+
+Can we easily compute such as basis? The answer is yes. Assume for now and for simplicity that $\int q = 1$. In one dimension, it is easy to find an orthonormal basis:
+$$
+\phi_0(x) = 1,
+\quad \phi_1(x) = \frac{(x-m)}{s},
+\quad \phi_2(x) = \frac{1}{\sqrt{2}}\left[\frac{(x-m)^2}{s^2} - 1\right].
+$$
+
+It is straightforward to see that $\|P_0\|=1$, $\|P_1\|=1$, $\langle P_0, P_1 \rangle=0$ and $\langle P_0, P_2 \rangle=0$.
+The reason why $\langle P_1, P_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric.
+Finally, note that:
+$$
+\phi_2(x)^2 = \frac{1}{2} 
+\left[\frac{(x-m)^4}{s^4} - 2 \frac{(x-m)^2}{s^2} + 1 \right],
+$$
+hence, given that the 4th-order centered moment of a Gaussian is~3 (which can be seen using the $\chi_1^2$ distribution), we have:
+$$
+\E_q [\phi_2^2(x)] = \|P_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
+$$
+
+In multiple dimensions, we do:
+$$
+\phi_0(\x) = 1, 
+\quad
+\phi_{1i}(\x) = \frac{(x_i-m_i)}{s_i},
+\quad
+\phi_{2i}(\x) = \frac{1}{\sqrt{2}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
+$$
+This defines an orthonormal basis because the components of~$\x$ are mutually independent under the factorial distribution~$q(\x)$, hence for different basis vectors: 
+$$
+\forall (i,j), \quad i\not= j, \quad
+\langle \phi_{ki}, \phi_{\ell j} \rangle = \E_q[\phi_{ki}]\E_q[\phi_{ki}] = \langle \phi_{ki}, 1 \rangle \langle P_{\ell j}, 1 \rangle = 0.
+$$
+Moreover, when it comes to dot products involving the same component, the one-dimensional results apply.
+
+In the general case where~$q$ does not integrate to one, we simply need to divide each basis vector by $Z=\int q_0$ to fulfill the unit norm conditions.
+Alternatively, we may leave the basis as is and later remember that the Hessian is $Z\text{Id}$.
+
+\subsection{Iteration}
+
+So we have an orthonormal basis that is very easy to compute!
+Now, consider decreasing the KL divergence $D(p\|q)$ starting from some current guess~$q_0$.
+To this end, we may compute an orthonormal basis associated with~$q_0$ and perform one Newton iteration in that basis.
+More specifically, we search for a new fit under the form:
+$$
+q_\th(\x) = q_0(\x) e^{\th^\top \phi(\x)},
+$$
+where~$\phi$ is the orthonormal basis. Clearly, the initial parameter is $\th_0=0$. The Newton update yields:
+$$
+\th = \gamma \f,
+\qquad
+\f = -\frac{1}{Z} \u,
+\quad
+\u \equiv \nabla D(\th_0) = \int (q_0 - p)\phi,
+$$
+where $\gamma$ is a given step size ($\gamma=1$ for a genuine Newton iteration) and~$Z$ is there for the Hessian correction.
+
+The gradient is generally intractable as it depends on~$p$, but we can approximate it via random sampling or quadrature. There are essentially two options. One is:
+$$
+\u \approx \int q_0\phi - {\cal I}(p\phi),
+$$
+where~${\cal I}$ denotes a numerical integration operator. The other option is:
+$$
+\u \approx {\cal I}[(q_0- p)\phi].
+$$
+This is where the variational sampling story may start.
+Option~1 amounts to a direct moment approximation.
+Option~2, which corresponds to our initial variational sampling idea, is perhaps less intuitive as it ignores the fact that we can compute the ``fitted moment'' exactly, but is arguably better because the error is smaller whenever~$q_0$ is close enough to~$p$.
+Hence, for a near Gaussian target, we expect less numerical instabilities.
+Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.
+
+So, our gradient update takes on the simple form $\th = \gamma\tilde{\f}$, but this is computed in a moving coordinate system since the orthonormal basis changes on each iteration.
+Is it easy to compute the `natural' parameters of $q_\th=q_0\exp(\th^\top \phi)$ from the moving coordinates~$\th$?
+Yes, it is.
+
+First, consider the right hand side and try to do the identification:
+$$
+e^{\th^\top \phi} = \kappa e^{-\frac{(\x-\mu)^2}{2\sigma^2}},
+$$
+hence, by taking the log:
+$$
+\log\kappa - \frac{1}{2} \sum_i \frac{(x_i-\mu_i)^2}{\sigma_i^2}
+= 
+\theta_0 + \sum_i \theta_{1i}\frac{x_i-m_i}{s_i}
++ \frac{1}{\sqrt{2}} \sum_i \theta_{2i}\big[ \frac{(x_i-m_i)^2}{s_i^2} -1 \big]. 
+$$
+We may differentiate this wrt $x_i$, yielding: 
+$$
+\frac{x_i-\mu_i}{\sigma_i^2} 
+= \frac{\theta_{1i}}{s_i} + \sqrt{2} \theta_{2i} \frac{x_i-m_i}{s_i^2}
+= \frac{\sqrt{2} \theta_{2i}}{s_i^2} \big[x_i - m_i + \frac{s_i\theta_{1i}}{\sqrt{2} \theta_{2i}} \big].
+$$
+By identification, we find:
+$$
+\sigma_i^2 = - \frac{s_i^2}{\sqrt{2}\theta_{2i}},
+$$
+and:
+$$
+\mu_i = m_i - \frac{s_i\theta_{1i}}{\sqrt{2}\theta_{2i}} = m_i + \frac{\sigma_i^2}{s_i}\theta_{1i}.
+$$
+
+To find $\kappa$, we note that the integral of $\th^\top \phi$ wrt~$q_0$ is equal to its dot product with~$\phi_0\equiv 1$, which is $\theta_0 \|\phi_0\|^2=\theta_0 Z$ since the basis $\phi/\sqrt{Z}$ is orthonormal.
+The constant $Z$ cancels out as it appears on both sides, and we find, using the K\"onig formula:
+$$
+\log \kappa = \theta_0 + \frac{1}{2}\sum_i \frac{s_i^2 + (\mu_i-m_i)^2}{\sigma_i^2}.
+$$
+
+We also need to get the parameter~$\th_0$ of~$q_0$, which act as an offset. For this, we just do $\kappa=K_0$, $\mu_i=m_i$, and $\sigma_i=s_i$, yielding:
+$$
+\theta_{2i}^0=-\frac{1}{\sqrt{2}},
+\quad
+\theta_{1i}^0 = 0,
+\quad
+\theta_0^0 = \log K_0 - \frac{d}{2}.
+$$
+
+Therefore, the updated fit is a function $\Psi(q_0,\th)$ of the current fit and parameter variation:
+$$
+v_{i1} =  \frac{v_{i0}}{1-\sqrt{2}\theta_{2i}},
+\quad
+m_{i1} = m_{i0} + \frac{v_{i1}}{\sqrt{v_{i0}}} \theta_{1i},
+\quad
+\log K_1 = \log K_0 + \theta_0 + \frac{1}{2} \left[\sum_i \frac{v_{i0} + (m_{i1}-m_{i0})^2}{v_{i1}} - d\right].
+$$
+
+\section{Naive VS}
+
+What the variance inequalities essentially prescribe is that the target factor~$f$ be a fast decreasing bounded function (it should not have heavier tails than polynomial in the Gaussian case).
+In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function.
+Other compliant factors can be derived from this case, as we will see.
+However, importance weighting functions resulting from distribution ratios are ruled out.
+
+Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function.
+A natural approach would be to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated.
+This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior.
+Makes sense.
+This is the canonical approach exposed in my previous publications on VS.
+
+However, there are some issues with that.
+Firstly, the likelihood might not be fast-decreasing (this is the case in image registration), meaning that VS convergence is doubtful.
+It may not converge at all, or convergence may be super slow...
+But, even if $L$ is a nice friendly sharply peaked function, sampling from the prior is intuitively not the most efficient thing to do as most simulated points will have low likelihood.
+
+This is what we may call the ``grasp all, loose all'' effect: unless the factor is stricly Gaussian, which never happens in practice (and if it did, VS would not be useful), the VS~variance blows up as the context gets wider.
+In a narrow context, i.e.
+$\pi$ is sharp wrt $f$, the factor doesn't vary much within the sampled region, hence it can be very well approximated by a Gaussian (think of the fact that $\log f$ may be close to a quadratic function in this small neighborhood by Taylor's expansion).
+Such an approximation is bound to be less accurate as the context gets wider.
+
+\section{Variance reduction}
+
+So, how can we make VS more stable?
+
+Instead of trying to solve the problem in one shot, maybe we can do it iteratively. Assume we have a current factor approximation $g_n$. How can we improve it?
+
+Perhaps a natural idea, at least this is among the first that came to my mind when I started to work on VS, is to apply VS sampling from~$\pi g_n$ to the factor $f/g_n$.
+In my initial framework, it means approximating the target $p=\pi f$ by the distribution $q=\pi g$, using the current approximation $q_n$ as a sampling distribution and $p/q_n$ as the target importance weight function.
+
+But we know that this is a bad idea because $p/q_n=f/g_n$ may not be upper bounded, hence VS is not even guaranteed to converge.
+In fact, it will diverge if $p$ is a heavy-tailed distribution, which is impossible to know in advance.
+
+A more reasonable thing to do is to note that, since $g_n$ approximates $f$, the combined function $g_n^{1-\alpha}f^\alpha$, for some $0< \alpha\leq 1$, yields an even better approximation.
+If $g_n$ is far off, it would be a little less far off if we pick a small~$\alpha$, and much closer if we pick a value close to one.
+If $g_n$ is already a good fit, it shouldn't matter too much which $\alpha$ we pick, at least it should have less impact.
+
+This simple remark suggests that the objective $D(\pi f\|\pi g)$ could be proxied by another KL divergence, $D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)$, yielding the following iteration:
+$$
+g_{n+1} = \arg\min_g D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)
+$$
+
+We will soon realize that this is heavily related to the minimization of the $\alpha$-divergence $D_\alpha(\pi f\|\pi g)$, and also has a strong connection with the Expectation Propagation algorithm, more specifically its ``averaged'' variant (AEP).
+
+But, for the time being, let us see why this is interesting for VS.
+
+\section{Scaling}
+
+We may interpret the KL divergence $D(\pi f\|\pi g)$ as a divergence between the distributions $\pi f$ and $\pi g$, but also as a divergence between the factors $f$ and $g$ in the context defined by~$\pi$.
+This second interpretation opens a new perspective on~VS, and enables generalizations.
+For instance, we may consider the more general divergence:
+$$
+\delta_\varepsilon (f, g) \equiv
+D(\pi f^\varepsilon \|\pi g^\varepsilon),
+$$
+where $\varepsilon$ is a positive scaling factor.
+The intuition behind this generalization is that the scaled target factor $f^\varepsilon$ gets flatter as $\varepsilon$ approaches zero, hence the VS estimator should have lower variance considering that the approximating family contains constant functions.
+On the other hand, applying a scaling factor~$\varepsilon$ smaller than one amounts to ``shrinking'' the context with respect to the factor, hence biasing the divergence minimizer (the VS target) towards the particular context.
+This yields a typical bias/variance tradeoff.
+
+Could the Expectation Propagation (EP) solve this?
+
+$$
+p = \pi \prod_{k=1}^n f^\varepsilon,
+\quad
+\varepsilon = \frac{1}{k}.
+$$
+
+Say current approx is $g^\epsilon$ for all factors.
+An EP update is to fit the factor $f^\varepsilon$ in the cavity $c=\pi g^{1-\varepsilon}$ by minimizing $D(cf^\varepsilon\|cg^\varepsilon)$ over a new $g$.
+This can be approximated using VS, with $c$ as the context, and $f^\varepsilon$ as the target factor.
+This is arguably an easier problem than the original one.
+
+\section{Expectation Propagation}
 
@@
 
-\section{Consistency}
+\subsection{Consistency}
 
@@
 
-\section{Fixed point}
+\subsection{Fixed point}
 
@@
 
-\section{AEP as relaxed KL~minimization}
+\subsection{AEP as relaxed KL~minimization}
 
@@
 
-\begin{figure}[h!]
-\centering
-\includegraphics[scale=1.7]{universe}
-\caption{The Universe}
-\label{fig:universe}
-\end{figure}
+%\begin{figure}[h!]
+%\centering
+%\includegraphics[scale=1.7]{universe}
+%\caption{The Universe}
+%\label{fig:universe}
+%\end{figure}
 

### 2022-02-16 papers/variana/notes.tex
Notes on custom Variana optimizer

@@
 \def\x{\mathbf{x}}
+\def\y{\mathbf{y}}
+\def\e{\mathbf{e}}
 \def\th{{\boldsymbol{\theta}}}
@@
 
-In the following, we consider the canonical problem of minimizing $D(p\|q)$ for some target~$p$, wrt some distribution log-quadratic distribution~$q$.
-I wrote the following section at Covii in Spring~2019.
-The goals is to have a scalable optimizer for VS, and the core idea is to re-parameterize the approximating family locally so as to implement~VS using a first-order gradient descent rather than a quasi-Newton method, which requires maninpulations on some Hessian approximation in one way or another and may thus be challenging in very large dimension.
-
-\subsection{Basis change}
-
-Consider minimizing the KL divergence via a gradient descent and consider a single gradient update.
-If we cannot afford to compute the Hessian (and to invert it), the gradient update relies on an arbitrary metric and may be rather ineffective.
-An important property of the KL divergence with exponential families is that its Hessian is independent from the target distribution and is associated with a dot product on the linear space spanned by~$\phi$ (which is a subspace of polynomials of order~2 for the factorial Gaussian family):
-$$
-H = \int q(\x) \phi(\x) \phi(\x)^\top d\x 
-\quad \Rightarrow \quad
-H_{ij} = \langle \phi_i , \phi_j \rangle .
-$$
-
-A natural idea is then to pick an orthonormal basis to use temporarily for a single gradient update as the divergence Hessian is scalar in such a basis.
-
-Can we easily compute such as basis? The answer is yes. Assume for now and for simplicity that $\int q = 1$. In one dimension, it is easy to find an orthonormal basis:
-$$
-\phi_0(x) = 1,
-\quad \phi_1(x) = \frac{(x-m)}{s},
-\quad \phi_2(x) = \frac{1}{\sqrt{2}}\left[\frac{(x-m)^2}{s^2} - 1\right].
-$$
-
-It is straightforward to see that $\|P_0\|=1$, $\|P_1\|=1$, $\langle P_0, P_1 \rangle=0$ and $\langle P_0, P_2 \rangle=0$.
-The reason why $\langle P_1, P_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric.
+We shall note that the basis functions $\phi_i$ underlying the fitting family are somewhat arbitrary, hence comes a natural question: which basis is most convenient for optimization?
+
+Consider for now the case of the KL divergence $D(\theta)=D(\pi f\|\pi g_\theta)$ for a given set of $\phi$-functions. We have the relations:
+$$
+\nabla D = \int \pi (g_\theta-f) \phi,
+\qquad
+\nabla^2 D = \int \pi g_\theta \phi\phi^\top.
+$$
+
+From there, a key remark is that the Hessian at a particular $\theta$ is the identity matrix if $\phi$ defines an orthonormal basis for the dot product:
+$$
+\langle h_1, h_2 \rangle \equiv \int \pi g_{\theta} h_1 h_2,
+$$
+where $h_1$ and $h_2$ are real-valued functions...
+We have some Hilbert space structure here, but I'll skip these details for now as I am not a mathematician.
+What really matters is that the Hessian being the identity means that a Newton iteration is very easy to compute as it boils down to a first-order gradient iteration.
+Importantly, however, this property is achieved for a particular basis function that is {\em dependent on the current fit estimate}.
+
+So, let's say that we have a current fit $g_n$ (which may be defined, for instance, by its coordinates in the canonical, fixed basis) and we want to perform one Newton iteration to improve it.
+Given an orthonormal basis $\phi_n$ associated with $g_n$ (in the sense of the above-defined dot product), we may look for the update under the form $g_\theta = g_n \exp({\theta^\top \phi_n})$.
+The Newton iteration from $\theta=0$ then takes the form:
+$$
+\hat{\theta} = -\lambda \nabla D(0) = \lambda \int \pi(f - g_n)\phi_n,
+$$
+for some damping factor $\lambda > 0$ (the genuine Newton iteration corresponds to $\lambda=1$), yielding the update:
+$$
+g_{n+1} = g_n \exp (\hat{\theta}^\top \phi_n).
+$$
+
+We will further need to represent $g_{n+1}$ in the same fixed coordinate system as $g_n$, which will require a linear coordinate change from $\phi_n$ to the fixed basis.
+
+\subsection{Orthonormal basis}
+
+Let's start in one dimension for simplicity.
+Given a current fit $g$, let $q=\pi g$ (pig\'e?),  with integral, mean and scale parameters $(Z, m, s)$, the following three functions define an orthonormal basis:
+$$
+\phi_0(x) = \frac{1}{\sqrt{Z}},
+\quad \phi_1(x) = \frac{(x-m)}{\sqrt{Z}s},
+\quad \phi_2(x) = \frac{1}{\sqrt{2Z}}\left[\frac{(x-m)^2}{s^2} - 1\right].
+$$
+
+{\bf Proof.} It is easy to see that $\|\phi_0\|=1$, $\|\phi_1\|=1$, $\langle \phi_0, \phi_1 \rangle=0$ and $\langle \phi_0, \phi_2 \rangle=0$.
+The reason why $\langle \phi_1, \phi_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric.
 Finally, note that:
 $$
-\phi_2(x)^2 = \frac{1}{2} 
+\phi_2(x)^2 = \frac{1}{2Z} 
 \left[\frac{(x-m)^4}{s^4} - 2 \frac{(x-m)^2}{s^2} + 1 \right],
@@
 $$
-\E_q [\phi_2^2(x)] = \|P_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
+\int q \phi_2^2 = \|\phi_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
 $$
@@
 $$
-\phi_0(\x) = 1, 
+\phi_0(\x) = \frac{1}{\sqrt{Z}}, 
 \quad
-\phi_{1i}(\x) = \frac{(x_i-m_i)}{s_i},
+\phi_{1i}(\x) = \frac{(x_i-m_i)}{\sqrt{Z} s_i},
 \quad
-\phi_{2i}(\x) = \frac{1}{\sqrt{2}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
+\phi_{2i}(\x) = \frac{1}{\sqrt{2Z}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
 $$
@@
 \forall (i,j), \quad i\not= j, \quad
-\langle \phi_{ki}, \phi_{\ell j} \rangle = \E_q[\phi_{ki}]\E_q[\phi_{ki}] = \langle \phi_{ki}, 1 \rangle \langle P_{\ell j}, 1 \rangle = 0.
+\langle \phi_{ki}, \phi_{\ell j} \rangle
+= \left(\int q \phi_{ki}\right)\left(\int q \phi_{\ell j}\right)
+= \langle \phi_{ki}, 1 \rangle \langle \phi_{\ell j}, 1 \rangle = 0.
 $$
@@
 
-In the general case where~$q$ does not integrate to one, we simply need to divide each basis vector by $Z=\int q_0$ to fulfill the unit norm conditions.
-Alternatively, we may leave the basis as is and later remember that the Hessian is $Z\text{Id}$.
-
-\subsection{Iteration}
-
-So we have an orthonormal basis that is very easy to compute!
-Now, consider decreasing the KL divergence $D(p\|q)$ starting from some current guess~$q_0$.
-To this end, we may compute an orthonormal basis associated with~$q_0$ and perform one Newton iteration in that basis.
-More specifically, we search for a new fit under the form:
-$$
-q_\th(\x) = q_0(\x) e^{\th^\top \phi(\x)},
-$$
-where~$\phi$ is the orthonormal basis. Clearly, the initial parameter is $\th_0=0$. The Newton update yields:
-$$
-\th = \gamma \f,
-\qquad
-\f = -\frac{1}{Z} \u,
-\quad
-\u \equiv \nabla D(\th_0) = \int (q_0 - p)\phi,
-$$
-where $\gamma$ is a given step size ($\gamma=1$ for a genuine Newton iteration) and~$Z$ is there for the Hessian correction.
-
-The gradient is generally intractable as it depends on~$p$, but we can approximate it via random sampling or quadrature. There are essentially two options. One is:
-$$
-\u \approx \int q_0\phi - {\cal I}(p\phi),
-$$
-where~${\cal I}$ denotes a numerical integration operator. The other option is:
-$$
-\u \approx {\cal I}[(q_0- p)\phi].
+\subsection{Coordinate change}
+
+Given some ``moving'' coordinates in the orthonormal basis, we aim to compute the corresponding coordinates in the canonical basis $b_0=1$, $b_{11}=x_1$, $b_{12}=x_1^2$, $b_{21}=x_2$, $b_{22}x_2^2$, etc.
+This is a linear mapping.
+We easily get by identification, owing to the separability of Gaussian factors:
+$$
+\phi_0 =
+\frac{1}{\sqrt{Z}} b_0,
+\qquad
+\phi_{1i} =
+-\frac{m_i}{\sqrt{Z}s_i} b_0
++ \frac{1}{\sqrt{Z}s_i} b_{1i},
+\qquad
+\phi_{2i} =
+\frac{1}{\sqrt{2Z}} \left(\frac{m_i^2}{s_i^2} - 1\right) b_0
+- \frac{\sqrt{2}m_i}{\sqrt{Z}s_i^2} b_{1i}
++ \frac{1}{\sqrt{2Z}s^2} b_{2i}.
+$$
+
+This shows that it the transition matrix from moving to fixed coordinates is block-diagonal, up to a first row corresponding to the constant terms.
+In the 1D case, it is upper triangular.
+In general, it never has more than three nonzero elements per column.
+
+\subsection{Application to VS}
+
+So far in this section, we have considered the problem of minimizing the KL~divergence.
+We have seen that we can elegantly work around the problem of inverting the Hessian in a Newton iteration by using appropriate orthonormal bases.
+However, as we know, the KL gradient may be intractable for the target factor under consideration, meaning that we need to approximate it via random sampling or quadrature, leading respectively to stochastic gradient and numerical approximation methods.
+
+There are essentially two options.
+
+One option is to approximate the moment involved in the KL gradient:
+$$
+\nabla D \approx \int \pi g_n\phi - {\cal I}(\pi f \phi),
+$$
+where~${\cal I}$ denotes a numerical integration operator.
+
+The other option is to approximate the KL gradient altogether:
+$$
+\nabla D \approx {\cal I}[\pi(g_n - f)\phi].
 $$
 This is where the variational sampling story may start.
-Option~1 amounts to a direct moment approximation.
-Option~2, which corresponds to our initial variational sampling idea, is perhaps less intuitive as it ignores the fact that we can compute the ``fitted moment'' exactly, but is arguably better because the error is smaller whenever~$q_0$ is close enough to~$p$.
-Hence, for a near Gaussian target, we expect less numerical instabilities.
+Option~2, which corresponds to our initial variational sampling idea, is arguably better because it is exact whenever $g_n=f$, meaning that the Newton iteration will converge deterministically whenever $f$ is in the approximating family.
+Hence, for a near Gaussian target, we expect less numerical instabilities with Option~2.
 Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.
-
-So, our gradient update takes on the simple form $\th = \gamma\tilde{\f}$, but this is computed in a moving coordinate system since the orthonormal basis changes on each iteration.
-Is it easy to compute the `natural' parameters of $q_\th=q_0\exp(\th^\top \phi)$ from the moving coordinates~$\th$?
-Yes, it is.
-
-First, consider the right hand side and try to do the identification:
-$$
-e^{\th^\top \phi} = \kappa e^{-\frac{(\x-\mu)^2}{2\sigma^2}},
-$$
-hence, by taking the log:
-$$
-\log\kappa - \frac{1}{2} \sum_i \frac{(x_i-\mu_i)^2}{\sigma_i^2}
-= 
-\theta_0 + \sum_i \theta_{1i}\frac{x_i-m_i}{s_i}
-+ \frac{1}{\sqrt{2}} \sum_i \theta_{2i}\big[ \frac{(x_i-m_i)^2}{s_i^2} -1 \big]. 
-$$
-We may differentiate this wrt $x_i$, yielding: 
-$$
-\frac{x_i-\mu_i}{\sigma_i^2} 
-= \frac{\theta_{1i}}{s_i} + \sqrt{2} \theta_{2i} \frac{x_i-m_i}{s_i^2}
-= \frac{\sqrt{2} \theta_{2i}}{s_i^2} \big[x_i - m_i + \frac{s_i\theta_{1i}}{\sqrt{2} \theta_{2i}} \big].
-$$
-By identification, we find:
-$$
-\sigma_i^2 = - \frac{s_i^2}{\sqrt{2}\theta_{2i}},
-$$
-and:
-$$
-\mu_i = m_i - \frac{s_i\theta_{1i}}{\sqrt{2}\theta_{2i}} = m_i + \frac{\sigma_i^2}{s_i}\theta_{1i}.
-$$
-
-To find $\kappa$, we note that the integral of $\th^\top \phi$ wrt~$q_0$ is equal to its dot product with~$\phi_0\equiv 1$, which is $\theta_0 \|\phi_0\|^2=\theta_0 Z$ since the basis $\phi/\sqrt{Z}$ is orthonormal.
-The constant $Z$ cancels out as it appears on both sides, and we find, using the K\"onig formula:
-$$
-\log \kappa = \theta_0 + \frac{1}{2}\sum_i \frac{s_i^2 + (\mu_i-m_i)^2}{\sigma_i^2}.
-$$
-
-We also need to get the parameter~$\th_0$ of~$q_0$, which act as an offset. For this, we just do $\kappa=K_0$, $\mu_i=m_i$, and $\sigma_i=s_i$, yielding:
-$$
-\theta_{2i}^0=-\frac{1}{\sqrt{2}},
-\quad
-\theta_{1i}^0 = 0,
-\quad
-\theta_0^0 = \log K_0 - \frac{d}{2}.
-$$
-
-Therefore, the updated fit is a function $\Psi(q_0,\th)$ of the current fit and parameter variation:
-$$
-v_{i1} =  \frac{v_{i0}}{1-\sqrt{2}\theta_{2i}},
-\quad
-m_{i1} = m_{i0} + \frac{v_{i1}}{\sqrt{v_{i0}}} \theta_{1i},
-\quad
-\log K_1 = \log K_0 + \theta_0 + \frac{1}{2} \left[\sum_i \frac{v_{i0} + (m_{i1}-m_{i0})^2}{v_{i1}} - d\right].
-$$
 

### 2022-03-07 cv/cv.tex
Updated cv

--- 2021-08-13 cv/cv.tex
@@
 \usepackage{color}
+
+\newcommand{\tabitem}{~~\llap{\textbullet}~~}
 
@@
 
-\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD in engineering science, with a specialist expertise in computer vision, data science and machine learning.
-Has operated in the fields of academic research, healthcare, electro-domestic and entertainment industry.
-Strong experience in the development of both open source and commercial software.
-{\em Keywords:} applied mathematics, artificial intelligence, image processing, computer vision, medical image analysis, machine learning, deep learning, statistics, Python,~C.
+\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD in engineering science, with a specialist expertise in computer vision and machine learning.
+Has operated in academic research for 15 years before moving to R\&D in industry (healthcare, electro-domestic, entertainment).
+Strong experience in the development of both open source and commercial software in Python, C, C++.
+{\em Keywords:} applied mathematics, statistics, artificial intelligence, machine learning, computer vision, medical image analysis, clinical studies, computer graphics.
 
@@
 
-\noindent \textbf{DATA SCIENCE:} In-depth knowledge of statistics, computer vision, medical image analysis, traditional machine learning and deep learning.
+\noindent \textbf{STATISTICAL MODELING:} Strong experience developing model-based algorithms for statistical inference, with applications in machine learning, computer vision, computer graphics, \ldots
 
 \ \\
-\noindent \textbf{SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.
+\noindent \textbf{TECHNICAL LEADERSHIP:} Proven ability to coordinate a team and manage objectives in multidisciplinary environments.
+
+\ \\
+\noindent \textbf{SCIENTIFIC COMPUTING:} 15 years experience in Python programming within the NumPy/SciPy ecosystem and interfacing C~language with Python.
 
@@
 \noindent \textbf{SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 published international patents.
-
-\ \\
-\noindent \textbf{TECHNICAL LEADERSHIP:} Proven ability to coordinate a team and manage objectives in multidisciplinary environments.
 
@@
 \begin{itemize}
-\item Developing computer vision and machine learning algorithms for reconstruction of high-fidelity digital humans from photos.
+\item Leading development of algorithms for automated generation of high-fidelity digital humans from photos.
+{\em Keywords:} facial reconstruction, ``deep face'', animation retargetting, image enhancement.
 \end{itemize}
@@
 \begin{itemize}
-\item Developed embedded object recognition and tracking algorithms for smart domestic appliances using traditional computer vision and deep learning (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017--18).
+\item Developed embedded algorithms for smart domestic appliances (VUXHub and Artisan intelligent oven prototypes demonstrated at IFA Berlin, 2017--18).
+{\em Keywords:} image classification, object recognition \& tracking, deep learning.
 \end{itemize}
+\ \\
 
-\ \\
 \bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
@@
 \begin{itemize}
-\item Led algorithmic development of the {\em MorphoBox} Siemens prototype for brain morphometry using anatomical MRI to help radiological reading for patients with suspected neurodegeneration.
-\item Contributed a number of clinical validation studies of brain morphometry in collaboration with neurologists and radiologists, via statistical analysis and data mining techniques, in Alzheimer's disease and other dementia, mild cognitive impairment, multiple sclerosis, migraine, anorexia, \ldots
-\item Co-supervised three PhD students and two MSc students.
+\item Led algorithmic development of a brain morphometry tool (released as part of the {\em AI-Rad Companion} Siemens solutions) to help radiological reading for patients with suspected neurodegeneration.
+{\em Keywords:} magnetic resonance imaging (MRI), brain image segmentation, machine learning for disease classification.
 \item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
@@
 \begin{itemize}
-\item Developed advanced algorithms for image processing and statistical analysis of functional, anatomical and diffusion-weighted MRI data (population analysis, spatio-temporal image registration, real-time imaging).
-\item French ANR-funded project management: PI (Karametria project, statistical analysis of brain structures, budget: 620 K\euro, 2009--2011), team leader (NIBB project, studying language in infants via functional neuroimaging, 2006--2009).
+\item Developed advanced algorithms for image processing and statistical analysis of brain imaging data.
+{\em Keywords:} functional, anatomical and diffusion-weighted MRI, image-based population analysis, spatio-temporal image registration, real-time imaging.
+\item Funded project management: principal investigator ({\em Karametria} -- statistical analysis of brain structures, French National Research Agency project, budget: 620 K\euro, 2009--2011); team leader ({\em NIBB} -- studying language in infants via functional neuroimaging, French National Research Agency project, 2006--2009).
 \item Active contributor to the NiPy software library (Neuroimaging in Python, \url{www.nipy.org}) from 2006.
-\item Main supervisor of two PhD students and four MSc students.
 \item Academic Guest at Computer Vision Laboratory, Swiss Federal Institute of Technology Zurich (ETHZ), Switzerland, 2009--2011.
@@
 \ \\
+\bighead{PhD CANDIDATE -- EPIDAURE PROJECT}{French National Research Institute (INRIA), Sophia Antipolis, France}{1997--2001}
+\nopagebreak
+\begin{itemize}
+\item Development of multimodal image registration algorithms, with applications in radiotherapy, image-guided surgery and neuroscience. 
+\end{itemize}
+
+\ \\
 \bighead{CONTINGENT SCIENTIST -- FRENCH NATIONAL SERVICE}{General Directorate for Armament (DGA), Vernon, France}{1996--1997}
@@
 \nopagebreak
-\head{PhD, Engineering Science (1997--2001)}{French National Research Institute (INRIA) \& University of Nice-Sophia Antipolis, France}
+\head{PhD, Engineering Science (1997--2001)}{University of Nice-Sophia Antipolis, France}
 \begin{itemize}
-\item Development of multimodal image registration algorithms, with applications in radiotherapy, image-guided surgery and neuroscience. With highest honor.
+\item With highest honor.
 \end{itemize}
@@
 \ \\
-\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris (now CentraleSupelec), France}
+\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris, France}
 \nopagebreak
@@
 \noindent
-\begin{minipage}{.15\textwidth}
-\textbf{Languages:}
-\end{minipage}
-\begin{minipage}{.85\textwidth}
-Native French, fluent English, basic Italian \& Portuguese. %German.
-\end{minipage}
 
-\ \\
-\noindent 
-\begin{minipage}{.15\textwidth}
-\textbf{Computer skills:}
-\ \\ 
-\ \\
-\ \\ 
-\ \\
-\end{minipage}
-\begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} Programming languages: C, Python, Matlab, R. \\
-{\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, tensorflow, keras.\\
-{\tiny$\bullet$} C/Python integration via cython.\\
-{\tiny$\bullet$} Version control software: git, svn, perforce. \\
-{\tiny$\bullet$} Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs. \\
-\end{minipage}
-
-\ \\
-\noindent 
-\begin{minipage}{.15\textwidth}
-\textbf{Publications:}
-\ \\ 
-\ \\
-\ \\
-\end{minipage}
-\begin{minipage}{.85\textwidth}
-{\tiny$\bullet$} 115 scholarly publications. Full list: \url{http://orcid.org/0000-0002-4821-6893}. \\
-{\tiny$\bullet$} Google scholar statistics: 6602 citations, h-index: 33, i10-index: 66 (July 2021).\\
-{\tiny$\bullet$} 5 published international patents. 2 ongoing applications.\\
-\end{minipage}
-
-%\ \\
-%\noindent 
-%\begin{minipage}{.15\textwidth}
-%\textbf{ Reviewing:}
-%\ \\ 
-%\ \\
-%\end{minipage}
-%\begin{minipage}{.85\textwidth}
-%IEEE Transactions on Medical Imaging, Medical Image Analysis, NeuroImage, Frontiers in
-% Neuroscience, IEEE Transactions on Image Processing, Signal Image and Video Processing, MICCAI conference, \ldots
-%\end{minipage}
-
-\ \\
-\noindent 
-\begin{minipage}{.15\textwidth}
-\textbf{Academic work:}
-\ \\ 
-\ \\
-\ \\
-\ \\
-\ \\
-\end{minipage}
-\begin{minipage}{.85\textwidth}
-  {\tiny$\bullet$} Post-graduate teaching in computer vision and medical image analysis: about 100~hours (Ecole Centrale Paris, EPFL Lausanne, University of Nice, University of Lausanne, INSERM/CNRS continuing education, \ldots).\ \\
-  {\tiny$\bullet$} Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\ \\
-\end{minipage}
-
-\ \\
-\noindent 
-\begin{minipage}{.15\textwidth}
-\textbf{References:}
-\end{minipage}
-\begin{minipage}{.85\textwidth}
-Available on request.
-\end{minipage}
+\begin{tabular}{p{0.2\linewidth}p{0.8\linewidth}}
+  \textbf{Languages:} & Native French, fluent English, basic Italian \& Portuguese.\\
+  \hline
+  \textbf{IT skills:} &
+  \tabitem Programming languages: Python, C, C++, Matlab, R.\\
+  & \tabitem Long-term experience with scientific Python pacakges (numpy, scipy, pylab).\\
+  & \tabitem CV/AI Python packages: skimage, PIL, opencv, sklearn, pytorch, tensorflow, keras.\\
+  & \tabitem  C/Python integration via Cython.\\
+  & \tabitem Version control software: git, svn, perforce.\\
+  & \tabitem Agile workflow (mainly Scrum) using JIRA \& Confluence.\\
+  & \tabitem Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs.\\
+  \hline
+  \textbf{Publications:} &
+  \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}. \\
+  & \tabitem 115 scholarly publications.\\
+  & \tabitem Google scholar statistics: 6862 citations, h-index: 33, i10-index: 68 (March 2022).\\
+  & \tabitem 5 published international patents. 2 ongoing applications.\\
+  \hline
+  \textbf{Academic work:} & 
+  \tabitem Post-graduate teaching in computer vision and medical image analysis: about 100~hours (Ecole Centrale Paris, EPFL Lausanne, University of Nice, University of Lausanne, INSERM/CNRS continuing education, \ldots).\\
+  & \tabitem Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\\
+  & \tabitem Supervision: two PhD students and four MSc students as main supervisor, three PhD students and two MSc students as co-supervisor.\\
+  \hline
+  \textbf{References:} & Available on request.
+\end{tabular}
 

### 2022-03-07 papers/variana/notes.tex
Updated cv

@@
 
+\section{VS vs IS}
+
+The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$.
+This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor.
+The IS estimator is just a sample mean:
+$$
+\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
+\qquad x_i \stackrel{iid}{\sim} \pi(x),
+$$
+which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
+$$
+\mathbb{E}[f(X)] = \int\pi f,
+$$
+we also have that:
+$$
+{\rm Var}[f(X)]
+= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
+$$
+
+Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.
+
+As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$.
+Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.
+
+Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
+Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''.
+We can make this a little bit more precise using :
+$$
+\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
+$$
+
 \section{Custom optimizer}
@@
 $$
-This is where the variational sampling story may start.
+
 Option~2, which corresponds to our initial variational sampling idea, is arguably better because it is exact whenever $g_n=f$, meaning that the Newton iteration will converge deterministically whenever $f$ is in the approximating family.
@@
 
-\section{Naive VS}
-
-What the variance inequalities essentially prescribe is that the target factor~$f$ be a fast decreasing bounded function (it should not have heavier tails than polynomial in the Gaussian case).
+What we have sketched here is nothing but a stochastic gradient method to minimize the KL~divergence, which is quite different in spirit from the initial VS~idea in that resampling could take place at each iteration of the optimization process.
+While this ensures that the successive gradient estimates are mutually independent, hence the method is unbiased and exact (not only asymptotically).
+This could be costly if the target factor takes time to evaluate, though.
+
+\section{Probability updating}
+
+What the variance inequalities essentially prescribe is that the target factor~$f$ be a bounded function.
+If this condition holds, the product distribution $\pi f$ does not have heavier tails than $\pi$, therefore IS converges whenever $\pi \phi\phi^\top$ exists, which is guaranteed for instance if $\pi$ is Gaussian (proper) and  $\phi$ is polynomial, which is our standard setting.
+VS convergence follows from the same argument provided that $(f-g)$ is bounded, in other words that the best fitting~$g$ is also bounded.
+
+polynomial in the Gaussian case.
 In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function.
@@
 Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function.
-A natural approach would be to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated.
+It is natural to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated.
 This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior.

### 2022-11-03 cv/cv.tex
Moving to canonical coordinates explained in variana notes

@@
 French National $\mid$ Married, 2 Children\\
+Based in Porto, Portugal.
 \end{small}
@@
 
-\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD in engineering science, with a specialist expertise in computer vision and machine learning.
-Has operated in academic research for 15 years before moving to R\&D in industry (healthcare, electro-domestic, entertainment).
+\noindent Hands-on technical team leader, engineer at heart, PhD, with a specialist expertise in artificial intelligence and data analysis.
+Has operated in public research and private R\&D in various fields (healthcare, electro-domestic, entertainment).
 Strong experience in the development of both open source and commercial software in Python, C, C++.
-{\em Keywords:} applied mathematics, statistics, artificial intelligence, machine learning, computer vision, medical image analysis, clinical studies, computer graphics.
+{\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, medical image analysis, clinical studies.
 
@@
 
-\noindent \textbf{STATISTICAL MODELING:} Strong experience developing model-based algorithms for statistical inference, with applications in machine learning, computer vision, computer graphics, \ldots
+\noindent
 
-\ \\
 \noindent \textbf{TECHNICAL LEADERSHIP:} Proven ability to coordinate a team and manage objectives in multidisciplinary environments.
@@
 \ \\
-\noindent \textbf{SCIENTIFIC COMPUTING:} 15 years experience in Python programming within the NumPy/SciPy ecosystem and interfacing C~language with Python.
+\noindent \textbf{SCIENTIFIC PROGRAMMING:} 15 years experience programming in Python and~C language in the SciPy ecosystem.
 
 \ \\
-\noindent \textbf{SCIENTIFIC WRITING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 published international patents.
+\textbf{STATISTICS:} Strong experience developing inferential statistical models for machine learning, computer vision, computer graphics, and performing data analysis for cognitive research, clinical studies, software validation, \ldots
+
+\ \\
+\noindent \textbf{SCIENTIFIC REPORTING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 published international patents.
 
@@
 
-\bighead{HEAD OF COMPUTER VISION \& ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
+\bighead{HEAD OF DIDIMO AUTOMATION TEAM}{Didimo, Porto, Portugal}{2019--present}
 \nopagebreak
 \begin{itemize}
-\item Leading development of algorithms for automated generation of high-fidelity digital humans from photos.
-{\em Keywords:} facial reconstruction, ``deep face'', animation retargetting, image enhancement.
+\item Leading development of generation pipeline of high-fidelity avatars from photos.
+{\em Keywords:} facial reconstruction, ``deep face'', animation retargetting, image feature extraction, image restoration.
 \end{itemize}
@@
 \ \\
-\bighead{PERMANENT RESEARCHER -- NEUROIMAGING INSTITUTE}{French Atomic Commission (CEA), Paris, France}{2002--2011}
+\bighead{PERMANENT RESEARCHER -- NEUROSPIN (NEUROIMAGING INSTITUTE)}{French Atomic Commission (CEA), Paris, France}{2002--2011}
 \nopagebreak
@@
 \begin{tabular}{p{0.2\linewidth}p{0.8\linewidth}}
-  \textbf{Languages:} & Native French, fluent English, basic Italian \& Portuguese.\\
-  \hline
+  \textbf{Languages:} & Native French, fluent English, basic Portuguese \& Italian.\\
+  %\hline
+  \ \\
   \textbf{IT skills:} &
   \tabitem Programming languages: Python, C, C++, Matlab, R.\\
-  & \tabitem Long-term experience with scientific Python pacakges (numpy, scipy, pylab).\\
+  & \tabitem Long-term experience with scientific Python packages (numpy, scipy, pylab).\\
   & \tabitem CV/AI Python packages: skimage, PIL, opencv, sklearn, pytorch, tensorflow, keras.\\
@@
   & \tabitem Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs.\\
-  \hline
+  %\hline
+  \ \\
   \textbf{Publications:} &
   \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}. \\
-  & \tabitem 115 scholarly publications.\\
-  & \tabitem Google scholar statistics: 6862 citations, h-index: 33, i10-index: 68 (March 2022).\\
+  & \tabitem 116 scholarly publications.\\
+  & \tabitem Google scholar statistics: 7308 citations, h-index: 35, i10-index: 72 (October 2022).\\
   & \tabitem 5 published international patents. 2 ongoing applications.\\
-  \hline
+  %\hline
+  \ \\
   \textbf{Academic work:} & 
-  \tabitem Post-graduate teaching in computer vision and medical image analysis: about 100~hours (Ecole Centrale Paris, EPFL Lausanne, University of Nice, University of Lausanne, INSERM/CNRS continuing education, \ldots).\\
+  \tabitem Post-graduate teaching in computer vision and medical image analysis: about 100~hours (France: Ecole Centrale Paris, University of Nice, INSERM/CNRS continuing education; Switzerland: EPFL Lausanne, University of Lausanne).\\
   & \tabitem Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\\
-  & \tabitem Supervision: two PhD students and four MSc students as main supervisor, three PhD students and two MSc students as co-supervisor.\\
-  \hline
+  & \tabitem Supervision of two PhD students and four MSc students as main supervisor, three PhD students and two MSc students as co-supervisor.\\
+  %\hline
+  \ \\
   \textbf{References:} & Available on request.

### 2022-11-03 papers/variana/notes.tex
Updates summer 2022
Moving to canonical coordinates explained in variana notes

@@
 
+Note that this is only tight if $\max(f,g)$ is constant,
+
 This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes.
@@
 So, let's say that we have a current fit $g_n$ (which may be defined, for instance, by its coordinates in the canonical, fixed basis) and we want to perform one Newton iteration to improve it.
-Given an orthonormal basis $\phi_n$ associated with $g_n$ (in the sense of the above-defined dot product), we may look for the update under the form $g_\theta = g_n \exp({\theta^\top \phi_n})$.
-The Newton iteration from $\theta=0$ then takes the form:
-$$
-\hat{\theta} = -\lambda \nabla D(0) = \lambda \int \pi(f - g_n)\phi_n,
+Given an orthonormal basis $\phi_n$ associated with $g_n$ (in the sense of the above-defined dot product), we may look for the update under the form $g_\nu = g_n \exp({\nu^\top \phi_n})$.
+The Newton iteration from $\nu=0$ then takes the form:
+$$
+\hat{\nu} = -\lambda \nabla D(0) = \lambda \int \pi(f - g_n)\phi_n,
 $$
@@
 $$
-g_{n+1} = g_n \exp (\hat{\theta}^\top \phi_n).
-$$
-
-We will further need to represent $g_{n+1}$ in the same fixed coordinate system as $g_n$, which will require a linear coordinate change from $\phi_n$ to the fixed basis.
+g_{n+1} = g_n \exp (\hat{\nu}^\top \phi_n).
+$$
+
+We will further need to represent $g_{n+1}$ in the same fixed coordinate system as $g_n$, i.e.
+using a ``$\theta$ parameter'', which will require a linear coordinate change from $\phi_n$ to the fixed basis.
 
@@
 - \frac{\sqrt{2}m_i}{\sqrt{Z}s_i^2} b_{1i}
-+ \frac{1}{\sqrt{2Z}s^2} b_{2i}.
-$$
-
-This shows that it the transition matrix from moving to fixed coordinates is block-diagonal, up to a first row corresponding to the constant terms.
++ \frac{1}{\sqrt{2Z}s_i^2} b_{2i}.
+$$
+
+This shows that the transition matrix from moving to fixed coordinates is block-diagonal, up to a first row corresponding to the constant terms.
 In the 1D case, it is upper triangular.
@@
 
+$$
+T =
+\frac{1}{\sqrt{Z}}
+\left[
+\begin{array}{ccccccc}
+1 & -\frac{m_1}{s_1} & -\frac{m_2}{s_2} & \ldots &
+  \frac{1}{\sqrt{2}} \left(\frac{m_1^2}{s_1^2} - 1\right) &
+  \frac{1}{\sqrt{2}} \left(\frac{m_2^2}{s_2^2} - 1\right) &
+  \ldots \\
+0 & \frac{1}{s_1} & 0 & \ldots &
+-\frac{\sqrt{2}m_1}{s_1^2} & 0 & \ldots \\
+  0 & 0 & \frac{1}{s_2} & 0 &
+  0 & -\frac{\sqrt{2}m_2}{s_2^2} & \ldots \\
+\vdots & \vdots & \vdots & \vdots &
+\vdots & \vdots & \vdots\\
+0 & 0 & 0 & 0 &
+  \frac{1}{\sqrt{2}s_1^2} & 0 & \ldots\\
+0 & 0 & 0 & 0 &
+  0 & \frac{1}{\sqrt{2}s_2^2} & \ldots\\
+\end{array}
+\right].
+$$
+
 \subsection{Application to VS}
+
+In VS, we approximate the KL divergence using some integration rule with weights~$w_i$:
+$$
+\hat{D}(\nu)
+\equiv \sum_i w_i \left[-f_i
+  \log g_\nu(\x_i) + g_\nu(\x_i) \right],
+$$
+yielding the gradient (with simplified notation):
+$$
+\nabla\hat{D} =
+\sum_i w_i (g_i - f_i) \phi(\x_i),
+$$
+where $\phi$ is the above defined orthonormal basis, which depends on the current fit~$g$.
+
+Hence, the VS approximation to the Newton iteration on the KL~divergence is:
+$$
+\hat{\nu} = -\lambda \sum_i w_i (g_i - f_i) \phi(\x_i).
+$$
+
+This requires evaluating both the current fit and the associated orthonormal basis functions at each sample point.
+Note that, in the canonical basis, we don't have to re-evaluate the basis functions on each iteration, but we have to re-evaluate the KL Hessian if we are to implement the genuine Newton method.
+
+Once the moving coordinates are updated through $\hat{\nu}$, we convert them to canonical coordinates by left multiplying $\hat{\nu}$ by the above transition matrix.
+
+\subsection{Utter bulshit}
 

### 2022-11-16 papers/variana/notes.tex
Fall 2022 changes

@@
 
+Let $\theta=T\nu$.
+
+$$
+\sqrt{Z}\theta_0 =
+\nu_0
+- \sum_i \frac{m_i}{s_i} \nu_{1i}
++ \frac{1}{\sqrt{2}} \sum_i \left(\frac{m_i^2}{s_i^2} - 1\right) \nu_{2i}
+.
+$$
+
+$$
+\sqrt{Z}\theta_{1i}
+=
+\frac{\nu_{1i}}{s_i} - \sqrt{2} \frac{m_i\nu_{2i}}{s_i^2}.
+$$
+
+$$
+\sqrt{Z}\theta_{2i}
+=
+\frac{\nu_{2i}}{\sqrt{2}s_i^2}.
+$$
+
 \subsection{Application to VS}
@@
 This requires evaluating both the current fit and the associated orthonormal basis functions at each sample point.
+Evaluating the basis function requires computing the natural parameters $(Z, m, s)$.
 Note that, in the canonical basis, we don't have to re-evaluate the basis functions on each iteration, but we have to re-evaluate the KL Hessian if we are to implement the genuine Newton method.
 
-Once the moving coordinates are updated through $\hat{\nu}$, we convert them to canonical coordinates by left multiplying $\hat{\nu}$ by the above transition matrix.
+Once the moving coordinates are updated through $\hat{\nu}$, we convert them to canonical coordinates by left multiplying $\hat{\nu}$ by the above transition matrix, which gives us the increment $\delta\theta=T\hat{\nu}$ to add to the current $\theta$ estimate for the next iteration.
 

### 2023-01-19 papers/robayes/notes.tex
Add notes about robust Bayes

--- 2022-03-07 papers/variana/notes.tex
@@
 
-\title{Variana Reloaded}
+\title{Robust Bayes Reloaded}
 \author{Alexis Roche}
@@
 
-\section{Variational sampling recap}
+\section{Initial toughts}
 
-Consider:
-\begin{itemize}
-\item a reference distribution~$\pi(x)$ (from which we can sample)
-\item a factor~$f(x)$ so that $p\equiv \pi f$ is a distribution of interest
-\item an approximating factor~$g$ searched under the exponential form:
-  $$
-  g(x) = e^{\theta^\top \phi(x)},
-  $$ where $\phi$ is a known vector-valued function and $\theta$ needs
-  to be determined.
-\end{itemize}
+The Bayesian paradigm is semi-generative in essence in that it relies on a parametric distribution model that incorporates a priori on unknown parameters.
+This a priori is non-testable in essence, and can thus be chosen in an arbitrary manner, which may lead to really poor inference models for high-dimensional models...
+This is an undeniable {\em weakness} of classical Bayesian analysis.
 
-VS is a stochastic method to approximate the parametric factor~$g$ that minimizes the generalized Kullback-Leibler divergence:
-$$
-D(\pi f\|\pi g)
-= \int \pi (f \log \frac{f}{g} + g - f)
-$$
+Robust Bayes analysis can be used to work around this problem.
+We may see it as a minimax risk strategy (instead of Bayesian) to select a reasonably low-dimensional parametric model to fight the curse of dimensionality.
+It is, in essence, a dimension reduction scheme.
 
-It does so by sampling independent points $(x_1,\ldots,x_n)$ from~$\pi$ and minimizing a Monte Carlo approximation to the KL~divergence, leading to the surrogate criterion:
-$$
-\tilde{D}(\theta) = \frac{1}{n} \sum_i \ell[ f(x_i), g_\theta(x_i) ],
-$$
-where $\ell(f, g)$ is the loss function associated with the KL~divergence:
-$$
-\ell(f, g) \equiv f \log \frac{f}{g} - f + g
-$$
-
-We have proved in our initial paper from 2013 that the resulting estimator of the integral, or moment, $I(\phi) \equiv \int \pi f \phi$, is asymptotically unbiased with variance matrix:
-$$
-\Sigma_n = \frac{1}{n} \int \pi (f - g_\star)^2 \phi \phi^\top,
-$$
-where $g_\star$ is the genuine (intractable) KL~minimizer:
-$$
-g_\star \equiv \arg\min_g D(\pi f\|\pi g).
-$$
-
-This makes VS fundamentally different from importance sampling, for which the moment is directly estimated via:
-$$
-\hat{\mu} \equiv \frac{1}{n} \sum_i f(x_i)\phi(x_i),
-$$
-yielding an unbiased estimator with asymptotic variance:
-$$
-\Sigma_n^{IS} = \frac{1}{n} \left[
-\int \pi f^2 \phi \phi^\top
-- I(\phi) I(\phi)^\top
-\right]
-$$
-
-In particular, VS has the key consistency property unlike IS to be exact when the target factor lies in the approximating family.
-
-\section{Variance upper bound}
-
-Intuitively, the closer the target factor~$f$ to the approximating family, the smaller the VS~asymptotic variance.
-To show that formally, we would like to upper bound the mean square error $\int\pi(f-g)^2$ by some increasing function of the KL~divergence $D(\pi f\|\pi g)$.
-
-I have been bumping into this issue for years...
-only to realize now (January 2022) that such an upper bound does exist and is actually pretty simple to exhibit.
-We will need a straightforward intermediate result:
-
-\paragraph{Alexis theorem:} if $f:\Omega\in\mathbb{R}\to \mathbb{R}$ is a smooth convex function from some open interval reaching its minimum in $a\in\Omega$, assuming for simplicity that $f(a)=0$, and such that the second derivative $f''$ is a decreasing function, then we have that:
-$$
-\forall x \leq a,
-\qquad
-f(x) \geq \frac{1}{2} f''(a) (x - a)^2 .
-$$
-
-The proof is straightforward from the fact that $f''(x) \geq f''(a)$ since $f''$ is decreasing, then we just need to integrate this relationship twice and take into account that $f'(a)=f(a)=0$.
-
-Next, let us consider the loss function $\ell(f,g)=f\log(f/g)+g-f$ as a function of $g$. We easily see that:
-$$
-\frac{\partial\ell}{\partial g} = 1 - \frac{f}{g},
-\qquad
-\frac{\partial^2\ell}{\partial g^2} = \frac{f}{g^2},
-$$
-implying that the loss is not only convex in~$g$ but also staisfies the conditions of Alexis's theorem with $a=f$. Hence,
-$$
-\forall g\leq f,
-\qquad
-\ell(f, g) \geq \frac{1}{2f} (g - f)^2. 
-$$
-
-But we can also consider the loss as a function of~$f$ at fixed~$g$. We have:
-$$
-\frac{\partial\ell}{\partial f} = \log(f),
-\qquad
-\frac{\partial^2\ell}{\partial g^2} = \frac{1}{f},
-$$
-which means that Alexis's theorem can be applied again, this time with $a=g$, yielding:
-$$
-\forall f\leq g,
-\qquad
-\ell(f, g) \geq \frac{1}{2g} (f - g)^2. 
-$$
-
-Concatenating both results, we see that for all $(f, g)$:
-$$
-(f-g)^2 \leq 2 \max(f, g)\ell(f, g).
-$$
-
-Integrating this over~$\pi$, we get this long-waited inequality:
-$$
-\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g).
-$$
-
-This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes.
-But this holds under the mild condition that both $f$ and $g$ remain bounded.
-This practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.
-
-Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.
-
-The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi$. The above result implies straighforwardly:
-$$
-(f-g)\phi^2
-\leq 2 \max(f\phi, g\phi) \ell(f\phi, g\phi)
-= 2 \max(f\phi, g\phi) \phi\ell(f, g)
-= 2 \max(f\phi^2, g\phi^2) \ell(f, g)
-$$
-where we use the fact that $\ell(f\phi,g\phi)=\phi\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
-$$
-\int \pi (f-g)^2\phi^2 \leq 2\max(\|f\phi^2\|_\infty, \|g\phi^2\|_\infty) D(\pi f\|\pi g).
-$$
-
-This yields the same kind of variance upper bound for any moment estimator we have.
-
-\section{VS vs IS}
-
-The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$.
-This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor.
-The IS estimator is just a sample mean:
-$$
-\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
-\qquad x_i \stackrel{iid}{\sim} \pi(x),
-$$
-which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
-$$
-\mathbb{E}[f(X)] = \int\pi f,
-$$
-we also have that:
-$$
-{\rm Var}[f(X)]
-= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
-$$
-
-Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.
-
-As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$.
-Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.
-
-Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
-Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''.
-We can make this a little bit more precise using :
-$$
-\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
-$$
-
-\section{Custom optimizer}
-
-We shall note that the basis functions $\phi_i$ underlying the fitting family are somewhat arbitrary, hence comes a natural question: which basis is most convenient for optimization?
-
-Consider for now the case of the KL divergence $D(\theta)=D(\pi f\|\pi g_\theta)$ for a given set of $\phi$-functions. We have the relations:
-$$
-\nabla D = \int \pi (g_\theta-f) \phi,
-\qquad
-\nabla^2 D = \int \pi g_\theta \phi\phi^\top.
-$$
-
-From there, a key remark is that the Hessian at a particular $\theta$ is the identity matrix if $\phi$ defines an orthonormal basis for the dot product:
-$$
-\langle h_1, h_2 \rangle \equiv \int \pi g_{\theta} h_1 h_2,
-$$
-where $h_1$ and $h_2$ are real-valued functions...
-We have some Hilbert space structure here, but I'll skip these details for now as I am not a mathematician.
-What really matters is that the Hessian being the identity means that a Newton iteration is very easy to compute as it boils down to a first-order gradient iteration.
-Importantly, however, this property is achieved for a particular basis function that is {\em dependent on the current fit estimate}.
-
-So, let's say that we have a current fit $g_n$ (which may be defined, for instance, by its coordinates in the canonical, fixed basis) and we want to perform one Newton iteration to improve it.
-Given an orthonormal basis $\phi_n$ associated with $g_n$ (in the sense of the above-defined dot product), we may look for the update under the form $g_\theta = g_n \exp({\theta^\top \phi_n})$.
-The Newton iteration from $\theta=0$ then takes the form:
-$$
-\hat{\theta} = -\lambda \nabla D(0) = \lambda \int \pi(f - g_n)\phi_n,
-$$
-for some damping factor $\lambda > 0$ (the genuine Newton iteration corresponds to $\lambda=1$), yielding the update:
-$$
-g_{n+1} = g_n \exp (\hat{\theta}^\top \phi_n).
-$$
-
-We will further need to represent $g_{n+1}$ in the same fixed coordinate system as $g_n$, which will require a linear coordinate change from $\phi_n$ to the fixed basis.
-
-\subsection{Orthonormal basis}
-
-Let's start in one dimension for simplicity.
-Given a current fit $g$, let $q=\pi g$ (pig\'e?),  with integral, mean and scale parameters $(Z, m, s)$, the following three functions define an orthonormal basis:
-$$
-\phi_0(x) = \frac{1}{\sqrt{Z}},
-\quad \phi_1(x) = \frac{(x-m)}{\sqrt{Z}s},
-\quad \phi_2(x) = \frac{1}{\sqrt{2Z}}\left[\frac{(x-m)^2}{s^2} - 1\right].
-$$
-
-{\bf Proof.} It is easy to see that $\|\phi_0\|=1$, $\|\phi_1\|=1$, $\langle \phi_0, \phi_1 \rangle=0$ and $\langle \phi_0, \phi_2 \rangle=0$.
-The reason why $\langle \phi_1, \phi_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric.
-Finally, note that:
-$$
-\phi_2(x)^2 = \frac{1}{2Z} 
-\left[\frac{(x-m)^4}{s^4} - 2 \frac{(x-m)^2}{s^2} + 1 \right],
-$$
-hence, given that the 4th-order centered moment of a Gaussian is~3 (which can be seen using the $\chi_1^2$ distribution), we have:
-$$
-\int q \phi_2^2 = \|\phi_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
-$$
-
-In multiple dimensions, we do:
-$$
-\phi_0(\x) = \frac{1}{\sqrt{Z}}, 
-\quad
-\phi_{1i}(\x) = \frac{(x_i-m_i)}{\sqrt{Z} s_i},
-\quad
-\phi_{2i}(\x) = \frac{1}{\sqrt{2Z}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
-$$
-This defines an orthonormal basis because the components of~$\x$ are mutually independent under the factorial distribution~$q(\x)$, hence for different basis vectors: 
-$$
-\forall (i,j), \quad i\not= j, \quad
-\langle \phi_{ki}, \phi_{\ell j} \rangle
-= \left(\int q \phi_{ki}\right)\left(\int q \phi_{\ell j}\right)
-= \langle \phi_{ki}, 1 \rangle \langle \phi_{\ell j}, 1 \rangle = 0.
-$$
-Moreover, when it comes to dot products involving the same component, the one-dimensional results apply.
-
-\subsection{Coordinate change}
-
-Given some ``moving'' coordinates in the orthonormal basis, we aim to compute the corresponding coordinates in the canonical basis $b_0=1$, $b_{11}=x_1$, $b_{12}=x_1^2$, $b_{21}=x_2$, $b_{22}x_2^2$, etc.
-This is a linear mapping.
-We easily get by identification, owing to the separability of Gaussian factors:
-$$
-\phi_0 =
-\frac{1}{\sqrt{Z}} b_0,
-\qquad
-\phi_{1i} =
--\frac{m_i}{\sqrt{Z}s_i} b_0
-+ \frac{1}{\sqrt{Z}s_i} b_{1i},
-\qquad
-\phi_{2i} =
-\frac{1}{\sqrt{2Z}} \left(\frac{m_i^2}{s_i^2} - 1\right) b_0
-- \frac{\sqrt{2}m_i}{\sqrt{Z}s_i^2} b_{1i}
-+ \frac{1}{\sqrt{2Z}s^2} b_{2i}.
-$$
-
-This shows that it the transition matrix from moving to fixed coordinates is block-diagonal, up to a first row corresponding to the constant terms.
-In the 1D case, it is upper triangular.
-In general, it never has more than three nonzero elements per column.
-
-\subsection{Application to VS}
-
-So far in this section, we have considered the problem of minimizing the KL~divergence.
-We have seen that we can elegantly work around the problem of inverting the Hessian in a Newton iteration by using appropriate orthonormal bases.
-However, as we know, the KL gradient may be intractable for the target factor under consideration, meaning that we need to approximate it via random sampling or quadrature, leading respectively to stochastic gradient and numerical approximation methods.
-
-There are essentially two options.
-
-One option is to approximate the moment involved in the KL gradient:
-$$
-\nabla D \approx \int \pi g_n\phi - {\cal I}(\pi f \phi),
-$$
-where~${\cal I}$ denotes a numerical integration operator.
-
-The other option is to approximate the KL gradient altogether:
-$$
-\nabla D \approx {\cal I}[\pi(g_n - f)\phi].
-$$
-
-Option~2, which corresponds to our initial variational sampling idea, is arguably better because it is exact whenever $g_n=f$, meaning that the Newton iteration will converge deterministically whenever $f$ is in the approximating family.
-Hence, for a near Gaussian target, we expect less numerical instabilities with Option~2.
-Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.
-
-What we have sketched here is nothing but a stochastic gradient method to minimize the KL~divergence, which is quite different in spirit from the initial VS~idea in that resampling could take place at each iteration of the optimization process.
-While this ensures that the successive gradient estimates are mutually independent, hence the method is unbiased and exact (not only asymptotically).
-This could be costly if the target factor takes time to evaluate, though.
-
-\section{Probability updating}
-
-What the variance inequalities essentially prescribe is that the target factor~$f$ be a bounded function.
-If this condition holds, the product distribution $\pi f$ does not have heavier tails than $\pi$, therefore IS converges whenever $\pi \phi\phi^\top$ exists, which is guaranteed for instance if $\pi$ is Gaussian (proper) and  $\phi$ is polynomial, which is our standard setting.
-VS convergence follows from the same argument provided that $(f-g)$ is bounded, in other words that the best fitting~$g$ is also bounded.
-
-polynomial in the Gaussian case.
-In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function.
-Other compliant factors can be derived from this case, as we will see.
-However, importance weighting functions resulting from distribution ratios are ruled out.
-
-Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function.
-It is natural to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated.
-This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior.
-Makes sense.
-This is the canonical approach exposed in my previous publications on VS.
-
-However, there are some issues with that.
-Firstly, the likelihood might not be fast-decreasing (this is the case in image registration), meaning that VS convergence is doubtful.
-It may not converge at all, or convergence may be super slow...
-But, even if $L$ is a nice friendly sharply peaked function, sampling from the prior is intuitively not the most efficient thing to do as most simulated points will have low likelihood.
-
-This is what we may call the ``grasp all, loose all'' effect: unless the factor is stricly Gaussian, which never happens in practice (and if it did, VS would not be useful), the VS~variance blows up as the context gets wider.
-In a narrow context, i.e.
-$\pi$ is sharp wrt $f$, the factor doesn't vary much within the sampled region, hence it can be very well approximated by a Gaussian (think of the fact that $\log f$ may be close to a quadratic function in this small neighborhood by Taylor's expansion).
-Such an approximation is bound to be less accurate as the context gets wider.
-
-\section{Variance reduction}
-
-So, how can we make VS more stable?
-
-Instead of trying to solve the problem in one shot, maybe we can do it iteratively. Assume we have a current factor approximation $g_n$. How can we improve it?
-
-Perhaps a natural idea, at least this is among the first that came to my mind when I started to work on VS, is to apply VS sampling from~$\pi g_n$ to the factor $f/g_n$.
-In my initial framework, it means approximating the target $p=\pi f$ by the distribution $q=\pi g$, using the current approximation $q_n$ as a sampling distribution and $p/q_n$ as the target importance weight function.
-
-But we know that this is a bad idea because $p/q_n=f/g_n$ may not be upper bounded, hence VS is not even guaranteed to converge.
-In fact, it will diverge if $p$ is a heavy-tailed distribution, which is impossible to know in advance.
-
-A more reasonable thing to do is to note that, since $g_n$ approximates $f$, the combined function $g_n^{1-\alpha}f^\alpha$, for some $0< \alpha\leq 1$, yields an even better approximation.
-If $g_n$ is far off, it would be a little less far off if we pick a small~$\alpha$, and much closer if we pick a value close to one.
-If $g_n$ is already a good fit, it shouldn't matter too much which $\alpha$ we pick, at least it should have less impact.
-
-This simple remark suggests that the objective $D(\pi f\|\pi g)$ could be proxied by another KL divergence, $D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)$, yielding the following iteration:
-$$
-g_{n+1} = \arg\min_g D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)
-$$
-
-We will soon realize that this is heavily related to the minimization of the $\alpha$-divergence $D_\alpha(\pi f\|\pi g)$, and also has a strong connection with the Expectation Propagation algorithm, more specifically its ``averaged'' variant (AEP).
-
-But, for the time being, let us see why this is interesting for VS.
-
-\section{Scaling}
-
-We may interpret the KL divergence $D(\pi f\|\pi g)$ as a divergence between the distributions $\pi f$ and $\pi g$, but also as a divergence between the factors $f$ and $g$ in the context defined by~$\pi$.
-This second interpretation opens a new perspective on~VS, and enables generalizations.
-For instance, we may consider the more general divergence:
-$$
-\delta_\varepsilon (f, g) \equiv
-D(\pi f^\varepsilon \|\pi g^\varepsilon),
-$$
-where $\varepsilon$ is a positive scaling factor.
-The intuition behind this generalization is that the scaled target factor $f^\varepsilon$ gets flatter as $\varepsilon$ approaches zero, hence the VS estimator should have lower variance considering that the approximating family contains constant functions.
-On the other hand, applying a scaling factor~$\varepsilon$ smaller than one amounts to ``shrinking'' the context with respect to the factor, hence biasing the divergence minimizer (the VS target) towards the particular context.
-This yields a typical bias/variance tradeoff.
-
-Could the Expectation Propagation (EP) solve this?
-
-$$
-p = \pi \prod_{k=1}^n f^\varepsilon,
-\quad
-\varepsilon = \frac{1}{k}.
-$$
-
-Say current approx is $g^\epsilon$ for all factors.
-An EP update is to fit the factor $f^\varepsilon$ in the cavity $c=\pi g^{1-\varepsilon}$ by minimizing $D(cf^\varepsilon\|cg^\varepsilon)$ over a new $g$.
-This can be approximated using VS, with $c$ as the context, and $f^\varepsilon$ as the target factor.
-This is arguably an easier problem than the original one.
-
-\section{Expectation Propagation}
-
-We propose to see online learning as an iterative fitting procedure in which each iteration performs an update of the form:
-$$
-q \leftarrow F(q^{1-\alpha} p^\alpha),
-$$
-where $F$ is a fitting operator, $q$ is the current objective approximation, $p$ is an empirical target distribution (possibly based on a random mini-batch), and $0\leq \alpha\leq 1$ is a learning rate.
-
-%If $\alpha=1$, this is in fact classical optimization applied to~$h$ (assuming the same one across iterations). This shows that offline learning is just a special case of online learning.
-
-This general iteration may be seen as a stochastic version of message passing for learning, as proposed in my amazing 2016 paper \cite{rr:16}, which treats mini-batches in a deterministic manner and enables factor-wise approximations, but requires heavy memory load.
-Conceptually, it's pretty much the same story but the stochastic version is much more feasible.
-In the message passing framework, the target is of the form $p=cf$, with $c=q^{1-\alpha}$ defining the ``context'' and $f=p^\alpha$ defining the ``factor'' to be approximated, hence the whole story is about fitting factors and the learning specifics are brushed under the carpet.
-
-The iteration is doubly approximative: the fitting objective at each iteration is an approximation, and the fitting procedure itself is approximative.
-How damaging is it?
-If the objective is exact, then the only thing we need is a good fitting scheme.
-If the fitting works well, then the only constraint is the limited batch size, which forces us to approximate the objective iteratively.
-
-We shall consider these as {\em separate problems}.
-Would it hurt to have an exact fitting scheme?
-It would not, for sure.
-We will focus here on minimizing (generalized) KL divergence as a fitting strategy:
-$$
-F_\star(p, q) = \arg\min_q D(p\|q),
-$$
-hence defining our {\em ideal} fitting operator.
-If the target is of the form $p=wf$, this may be interpreted as minimizing the local KL divergence $D(wf\|wg)$ along~$g$, so it can be seen as factor approximation in the sense of the local KL divergence.
-But we may also consider that we are minimizing an adaptive target {\em globally}.
-
-Assume for now that we are dealing with a fixed target~$p$.
-Ideally, we would set $\alpha=1$ to fit~$p$ as quickly as possible.
-However, the fit can only be approximated in practice and may be poor if the target is far off the approximating family.
-A target of the form $q^{1-\alpha}p^\alpha$ is intuitively `easier' to fit if~$q$ lies in the approximating family.
-
-%On the other hand, for a narrow context, VS will pretty much behave like a Taylor expansion.
-So, if we have to choose between the two options, we could say we are on the safe side using a narrow context, but the reality is that we may get a poor fit in both cases.
-
-%We may assume that the best option is a context that is more or less centered at the target peak and just wide enough to capture relevant variations.
-If the target is close to the approximation space, then it does not matter much whether~$w$ is narrow or wide.
-But if it's not, then the question is which scale is relevant to describe it.
-The global KL fit would certainly help.
-Problem is: it's exactly what we are looking for.
-Chicken and egg problem.
-
-The proposed scheme could be interpreted as gradual `non-gaussianity' or some sort of bridge sampling.
-Say we start with a very small~$\alpha$ and some prior guess~$q=\pi$.
-Since~$\alpha$ is small, our target $\pi^{1-\alpha}p^\alpha$ at first iteration is almost Gaussian.
-We get an update~$q$ that is just a little skewed towards~$p$:
-$$
-q \approx \pi^{1-\alpha} p^\alpha
-$$
-
-Now, if we update the target according to $p=q^{1-\alpha}p^\alpha$ and re-fit it, the new fit is a little bit more skewed towards~$p$. Roughly,
-$$
-q^{1-\alpha} p^\alpha 
-\approx \pi^{(1-\alpha)^2} p^{(1-\alpha)\alpha + \alpha}
-= \pi^{1-\beta} p^{\beta},
-$$
-with $\beta=\alpha+(1-\alpha)\alpha$ evaluating the larger weight of~$p$ relatively to the prior~$\pi$ in the new target, $\beta=\alpha+(1-\alpha)\alpha\approx 2\alpha$.
-
-%The updated~$q$ is a natural choice for the context since it is, in principle, a closer fit to~$p$ than the prior.
-
-So the idea is pretty simple: pick a fixed value~$\alpha>0$ and iterate the process of fitting $q^{1-\alpha}p^\alpha$.
-After several (many?) iterations, we hope that the target closely resembles~$p$.
-Is that naive?
-
-What can go wrong? Well, if we use too large a value for~$\alpha$, this clearly won't work... So one first question is: how do we make sure that~$\alpha$ is small enough?
-
-\subsection{Consistency}
-
-Let~$p$ a target distribution and $Q$ an approximating family.
-A sufficient condition for the KL fit (in an exponential family of the type considered in the seminal paper) to have finite integral is obviously that~$p$ have a finite integral because $\int q_\star = \int p$.
-Moreover, if~$p$ has a finite integral, the same goes for $wp$ for any bounded function~$w$, therefore the contextual KL fit will always satisfy $\int q < \infty$.
-This implies that, if the target distribution is proper, the iterative KL minimization considered here yields a proper distribution.
-
-\subsection{Fixed point}
-
-Assume that~$Q$ is an exponential family $\exp(\theta^\top \phi)$. The fixed point~$q_*$ of our algorithm is such that:
-$$
-q_* = \arg\min_q D(q_*^{1-\alpha} p^\alpha \| q),
-$$
-therefore it satisfies the moment condition
-$$
-\int q_*^{1-\alpha}p^\alpha\phi 
-= \int q_* \phi
-$$
-
-Consider the $\alpha$-divergence:
-$$
-D_\alpha(p\|q) = 
-\frac{1}{\alpha(1-\alpha)}
-\int \alpha p + (1-\alpha) q - p^\alpha q^{1-\alpha} 
-$$
-
-Consider minimizing it on~$Q$. Whenever $\alpha\not=0,1$, the first order optimality condition reads:
-\begin{eqnarray*}
-\alpha(1-\alpha) \frac{\partial D_\alpha}{\partial \theta} 
-& = & \int (1-\alpha) q \phi - p^\alpha (1-\alpha) q^{1-\alpha} \phi \\
-& = & (1-\alpha) \int (q - p^\alpha q^{1-\alpha}) \phi \\
-& = & 0
-\end{eqnarray*}
-
-Incredible!!!
-I am realizing now that Minka \cite{Minka-05} noted this result before, but maybe not the implication: it means that the $\alpha$-divergence optimality condition is exactly the same as for the iterative scheme considered here using a {\em fixed target}, call it simple EP or whatever you like.
-Since $D_\alpha$ is convex, this means that, whenever simple EP converges, it converges to the unique minimizer of $D_\alpha(p\|q)$ on~$Q$.
-But how comes Minka or \cite{Dehaene-18} did not include this result in their convergence analysis of EP or AEP?
-Because they were using EP or AEP with variable targets in a data partitioning perspective.
-
-The remaining issue is convergence. Basically, AEP boils down to iterated $M$-projections:
-\begin{equation}
-\label{eq:iter_m_proj}
-q \leftarrow \arg\min_{q'} D(q^{1-\alpha}p^\alpha\|q').
-\end{equation}
-
-Iterated projection is a typical algorithm to find the intersection point between two sets, but AEP is not quite the same thing.
-There is only one projection set here and, at each iteration, the current projection is slightly pushed away before re-projecting.
-
-We know that, if convergence occurs, then AEP minimizes $D_\alpha(p\|q)$, but is AEP guaranteed to converge?
-It would help to have a monotonicity property, {\em i.e.}, the update~(\ref{eq:iter_m_proj}) always decreases $D_\alpha(p\|q)$.
-Is it true?
-Minka \cite{Minka-05} does not seem to think so.
-
-In practice, anyway we can only approximate the $M$-projection, so there might be no theoretical convergence guarantee.
-It may thus be wise to try and modify the iterative scheme so as to force convergence.
-A similar situation occurs in gradient descent schemes, which are not guaranteed to converge unless the step size parameter is tuned on each iteration so as to make the objective decrease.
-
-Can we force convergence? Well, we have a fixed point iteration:
-$$
-\theta_{n+1} = f(\theta_n),
-$$
-with:
-$$
-f(\theta) = \arg\min_{\theta'} \tilde{D}(q_\theta^{1-\alpha}p^\alpha\|q_{\theta'}),
-$$
-where $\tilde{D}$ is some tractable KL divergence proxy.
-The function $f(\theta)$ is fully determined by~$p$ and~$\alpha$.
-The update may be generalized using a damping factor or ``learning rate'' $0<\epsilon_n\leq 1$:
-$$
-\theta_{n+1} = (1-\epsilon_n)\theta_n + \epsilon_n f(\theta_n).
-$$
-
-% We can always tune $\epsilon_n$ so that $\|\theta_{n+1}-\theta_n\|\leq \|\theta_n-\theta_{n-1}\|$, thereby ensuring that $\theta_n$ is a Cauchy sequence.
-
-In practice, a small constant value for~$\epsilon$ might do.
-In a safer version of the algorithm, we may tune $\epsilon_n$ on each iteration so as to guarantee that some $\alpha$-divergence proxy decreases.
-
-\subsection{AEP as relaxed KL~minimization}
-
-The fact that fixed-target AEP is a convex minimization algorithm is certainly an interesting result that qualifies it as a relaxation method, which replaces the KL divergence objective with another $\alpha$-divergence for $0<\alpha<1$, corresponding to an {\em exclusive} and {\em non-zero forcing} divergence.
-
-Why performing such a relaxation?
-Because the original problem is intractable.
-That is the only valid reason.
-By adopting this relaxation, we assume that we are able to minimize the KL divergence at least approximately when the target is `close enough' to the approximating family.
-So the existence of an approximate KL minimization scheme is at the heart of this strategy.
-
-What are the candidates? 
-\begin{enumerate}
-\item Laplace method.
-\item Approximate integration (remarking the equivalence between KL min and moment matching), so anything from the quadrature and Monte Carlo literature.
-\item Variational sampling.
-(We could include in this category approaches such as Bayesian Monte Carlo, i.e.
-VS variants that make use of other costs than the empirical KL divergence, but such methods do not make much sense in the context of approximate KL minimization, do they?).
-\end{enumerate}
-
-We could dismiss approximate integration from the very beginning based on a consistency argument: the fact that they are inexact even if the target lies in the approximating family, unlike both Laplace and VS.
-So, we have two remaining candidates.
-
-What are the compared pros and cons of Laplace and VS? I can think about the following:
-\begin{itemize}
-\item[$-$] Laplace is potentially more efficient if the gradient and hessian have closed-form expressions.
-\item[$+$] Laplace is local: it won't work well if the target is `rugged' or skewed, while VS is arguably more robust to such features. 
-\item[$+$] Laplace does not handle singularities unlike VS (think about a... Laplace distribution, for which the Hessian does not exist at the mode).
-\item[$+$] Laplace can output non-positive hessians, while VS can be constrained not to.
-\end{itemize}
-
-To put it in Christopher Bishop's terms: ``The most serious limitation of the Laplace framework, however, is that it is based purely on the aspects of the true distribution at a specific value of the variable, and so can fail to capture important global properties.'' In short, Laplace is potentially faster but less accurate than VS.
-At least, this is what I could have shown experimentally in my introducing paper, but I decided instead to consider VS as a moment matching method...
-how clever.
-
-%We can't do it when it is far off, but can do it if is close enough -- how weird does that sound?
-I don't know.
-Arguably, it is easier to minimize the $\alpha$-divergence for small~$\alpha$ than for $\alpha\approx 1$.
-I am not even talking about $\alpha\leq 0$ or $\alpha > 1$, fuck it.
-
-\section{Bridge sampling scheme}
-
-Consider the series:
-$$
-\alpha_{n+1}=\alpha_n+(1-\alpha_n)\alpha_n=(2-\alpha_n)\alpha_n,
-\qquad
-0 < \alpha_0 < 1 .
-$$
-
-Clearly, $\alpha_n$ is increasing, upper bounded and therefore converges to a limit satisfying $\ell=(2-\ell)\ell$, hence $\ell=1$ cause $\ell=0$ is impossible if $\alpha_0>0$.
-
-Let $\beta=(2-\alpha)\alpha$. We note that: $1-\beta=1-\alpha - \alpha(1-\alpha)=(1-\alpha)^2$.
-
-\section{Laplace approximation}
-
-Conjecture: if~$w$ approaches a Dirac distribution, the KL fit is the Laplace approximation, {\em i.e.} the fit derived from the Taylor expansion at the cavity center.
-Is that true?
-Sounds obvious.
-What else could it be?
-
-If it's true, then we can claim that gradient descent and Newton method are just limiting cases of local KL minimization. If it's not, then KL minimization might raise some suspicion.
-
-\section{Gradient descent}
-
-One simple and classical way to perform fitting on each iteration is via a second-order Taylor expansion of the target logarithm:
-$$
-F = \log t = (1-\alpha)\log q + \alpha \log p.
-$$
-
-In the `quick Laplace' approach \cite{rr:16}, we perform the expansion at the cavity center~$a$, which verifies $\nabla \log q(a)=0$, hence:
-$$
-\nabla F(a) = \alpha g_p(a),
-\qquad
-\nabla\nabla^\top F(a) = (1-\alpha)H_q + \alpha H_p(a)
-$$
-
-Assuming that~$H$ is positive definite, the value that minimizes the local quadratic approximation is $b = a - H^{-1}g$, hence the update formulas:
-$$
-H \leftarrow (1-\alpha)H + \alpha H_p(a),
-\qquad
-a \leftarrow a - \alpha H ^{-1}g_p(a).
-$$
-
-If $\alpha=1$, this is the famous Newton algorithm.
-Using a learning rate $\alpha < 1$ may be useful even if~$p$ is a fixed function to stabilize the algorithm.
-For instance, if we start far away from the solution, the initial hessian may be off, so it makes sense to regularize it starting, e.g., with a positive scalar matrix for~$H$.
-Weirdly enough, this somewhat natural strategy does not seem to be much used in practice.
-
-If the target is truly Gaussian, this will clearly delay convergence.
-However, if it is not even log-concave, this stabilization scheme may help avoiding convergence to saddle points.
-Instead of seeing~$\alpha$ as a learning rate, we may see it as a `prudence' parameter.
-Prudence is required for complex targets.
-
-\section{Variational sampling}
-
-What is VS?
-It is a numerical method to approximately solve the local KL problem.
-Canonical VS uses~$w$ as a sampling distribution.
-It's possible to distinguish the context from the sampling distributions, as I did in my original presentation \cite{ijasp:13}, but the price to pay is the potential numerical instability of the underlying importance weights.
-Besides, we can always fall back to the canonical version by pre-conditioning the target, therefore we shall restrict ourselves to canonical VS in the sequel.
-
-Some basic results. Surrogate cost:
-$$
-\sum_k 
-w_k \left[
-p(x_k) \log \frac{p(x_k)}{q(x_k)} + q(x_k) - p(x_k)
-\right],
-$$
-where $w_k$ are the weights associated with a quadrature rule for $\int f(x)dw$.
-
-For random independent sampling from $w(x)$, we have $w_k\equiv 1/n$ and the error on the local target integral $z_w(p)= \int wp$, which represents the ideal approximation peak, is asymptotically Gaussian with zero mean and variance:
-$$
-v = \frac{1}{n} \int w(p-q_w)^2 ,
-$$
-where $q_w\equiv\arg\min_q D_w(p,q)$.
-
-Intuitively, the tighter~$w$, the closer~$q_w$ approximates~$p$ on the support of~$w$, therefore the smaller the variance.
-However, the tighter~$w$, the more loosely~$q_w$ approximates~$q_\star$, the global KL fit, hence the larger the bias.
-So, we have a bias/variance tradeoff issue.
-
-To limit the bias, we need a cavity that is wide and centered as much as possible around the actual peak.
-But wide means high variance.
-Moreover, how do we guarantee that we are centered on the actual peak?
-How do we go about that?
-
-Canonical form. Minimize $D(wf\|wg)$ or, equivalently,
-$$
-C(g) = \int w(-f \log g + g),
-$$
-along functions of the form:
-$$
-g_\theta(x) = e^{\theta^\top \phi(x)},
-$$
-yielding:
-$$
-\nabla C (\theta) = \int w (g_\theta-f) \phi,
-\qquad
-\nabla \nabla^\top C(\theta) = \int w g_\theta \phi\phi^\top.
-$$
-
-In practice, we perform approximate minimization by sampling from~$w$ (so the evaluation points $x_k$ depend on~$w$, please keep in mind):
-$$
-\hat{C}(g) = \sum_k w_k \left[
--f(x_k) \log g(x_k) + g(x_k)
-\right]
-$$
-
-Assume:
-$$
-w(x) = e^{\theta_0^\top \phi(x)},
-\qquad
-g(x) = e^{\theta^\top \phi(x)},
-\qquad
-q(x) = e^{(\theta_0+\theta)^\top \phi(x)}
-$$
-
-$$
-\hat{C}(\theta) = \sum_k w_k \left[
--f(x_k) \theta^\top\phi(x_k) + e^{\theta^\top \phi(x_k)}
-\right]
-$$
-
-$$
-\nabla \hat{C}(\theta) 
-= \sum_k w_k 
-[g(x_k) - f(x_k)] \phi(x_k)
-$$
-
-$$
-\nabla\nabla^\top \hat{C}(\theta) 
-= \sum_k w_k 
-g(x_k) \phi(x_k)\phi(x_k)^\top
-$$
-
-We may use a standard optimization algorithm such as L-BFGS to solve this.
-It could be quite costly in memory, though, cause we need to store $\theta_0$, $\theta$, as well as all the sampling points.
-But with big computers, we should be able to do this!
-
-A sensible starting value for~$\theta$ could be derived assuming a constant fit: $g(x)\equiv c$, for which the optimal~$c$ is seen to be:
-$$
-c = \frac{\sum_k w_k f(x_k)}{\sum_k w_k}
-$$
-
-The optimizer outputs a value~$\theta$, that needs to be added to the current~$\theta_0$ to produce the new parameter representation of~$q$.
-If we want to run another iteration, we infer the new~$\theta_0$ by scaling~$q$ by a factor $1-\alpha$ and run the same process.
-
-What I am sketching here is an iterative scheme encompassing VS.
-Instead of trying to minimize $D(p\|q)$ in a single shot, we iteratively minimize divergences of the form $D(\tilde{p}\|q)$ where $\tilde{p}$ itself is an adaptive approximation to~$p$.
-This is a way of solving the chicken and egg problem underlying~VS.
-
-We could call this fractionated VS or bridge VS but, this algorithm turns out to be a special case of average or stochastic expectation propagation (AEP) \cite{Dehaene-18,Li-15}.
-To see this, write down the target distribution as a straightforward product of factors:
-$$
-p = \prod_{i=1}^n f_i,
-\qquad\text{with}\quad
-f_i = p^\frac{1}{n}.
-$$
-
-When factors represent independent opinions as in Minka's original work \cite{Minka-05}, it does make sense to use the fully blown EP algorithm, provided that we have enough memory for that, but when factors are hugely redundant, then it is intuitively more appropriate to use the AEP~variant as we clearly don't need an approximation for each factor.
-
-At the end of the day, VS appears once again as a particular distribution fitting method, competing with Laplace approximation and Monte Carlo methods.
-It is easy to argue that VS has the potential to work around local convergence unlike Laplace, which gets easily trapped in saddle points for `rugged' distributions.
-It is also clear in practice that VS works much better than comparable importance sampling, and the same can be expected for many other MCMC methods since the sampling kernel is not really an issue here -- and we are not mentioning computation time.
-
-So, basically, VS is a strong candidate but, still, my work on VS has attracted little attention, if not zero.
-There are possibly several reasons to this, including the valid one that the computational complexity of VS appears to be discouraging: $O(d^3)$ using Newton's method \cite{rr:16}.
-Of course, it is significantly reduced using limited-memory BFGS.
-I was fairly stupid not to point that out.
-
-In the context where VS is used iteratively to perform $\alpha$-divergence minimization, each iteration approximately minimizes $D(q^{1-\alpha}p^\alpha\|q')$, so we have $w=q^{1-\alpha}$, $f=p^\alpha$ and $q'=q^{1-\alpha}g$.
-It was mentioned above that some damping may be used to make the iterative process converge by tuning the damping factor~$\epsilon_n$ so that the $\alpha$-divergence proxy decreases.
-Since the $\alpha$-divergence is not tractable, a proxy is needed.
-One that is consistent with VS in the sense that it yields the {\em same} first-order condition is:
-$$
-D_\alpha (p\|q) 
-\equiv 
-\int q^{1-\alpha} [(1-\alpha)q^\alpha - p^\alpha]
-\approx 
-(1-\alpha)\langle q^\alpha \rangle_w - \langle p^\alpha \rangle_w
-$$
-
-\section{Coordinate-wise KL minimization}
-
-In very high-dimensional problems, it is unfeasible to run VS in a single shot on the whole approximating family.
-More specifically, VS cannot be run to directly fit $q^{1-\alpha}p^\alpha$ if only because of memory load.
-
-A natural trick to work around this issue is to perform something similar to coordinate block optimization, that is, replace full dimensional factors by coordinate applications.
-Consider a partition $(x_0,x_1)$ of the coordinate space, and let us assume that we want to optimize the parameters associated with~$x_1$ while holding the others fixed.
-
-Due to the fully factorial nature of the approximating family, the factor approximation verifies $q(x)=q_0(x_0)q_1(x_1)$ (this would not be true for a full Gaussian approximation).
-Note that this decomposition is not unique.
-This separation property, however, only allows for dimension reduction if the tilted target, $q^{1-\alpha}p^\alpha$, is also separable, which only happens if~$p$ is separable and it is seldom the case.
-
-To enforce target separation, we may make our contextual target definition a bit more general by including a dimension reduction mechanism.
-Let us pick some point~$a$ depending on~$q$ (for instance the mode) and consider an approximation of the form:
-$$
-t(x) = \gamma q_0(x_0) q_1(x_1)^{1-\alpha}p_1(x_1)^\alpha,
-$$
-where $p_1(x_1)=p(a_0, x_1)$ is the coordinate application of~$p$ along~$x_1$ at fixed $x_0=a_0$.
-
-The intuition behind this is that, if~$q=q_0q_1$ is a good fit to~$p$, then $\tilde{p}\equiv q_0p_1$ should approximate~$p$ even better up to a constant factor.
-How to pick the constant factor?
-To make $\tilde{p}$ look as much as possible like~$p$, it is natural to require $\tilde{p}(a)=p(a)$, therefore:
-$$
-\tilde{p}(x) = \frac{q_0(x_0)}{q_0(a_0)} p_1(x_1),
-$$
-hence normalizing~$q_0$ in some way.
-In other words, for a given coordinate block, we `simplify' the target distribution as $\tilde{p}(x)=q_0(x_0)p_1(x_1)/q_0(a_0)$ and try to fit~$\tilde{p}$ by the same model as before, $q(x)=q_0(x_0)q_1(x_1)$, all this while holding~$q_0$ fixed.
-Because the simplified target is intuitively `closer' to the actual target than the current fit, it may be useful to further improve it.
-
-To achieve this, we need to be careful because there may still be large discrepancies between~$q$ and~$\tilde{p}$ in the early iterations, so we will consider the tilted target:
-\begin{eqnarray*}
-t(x) 
- & = & q(x)^{1-\alpha} \tilde{p}(x)^\alpha \\
- & = & q_0(a_0)^{-\alpha} q_0(x_0) q_1(x_1)^{1-\alpha} p_1(x_1)^\alpha . 
-\end{eqnarray*}
-
-Therefore, the constant factor~$\gamma$ is defined as: 
-$$
-\gamma = q_0(a_0)^{-\alpha}
-= \left[\frac{q_1(a_1)}{q(a)}\right]^{\alpha}.
-$$
-
-We are approximating a form $q_0(x_0)p_1(x_1)$ with a form $q_0(x_0)q_1(x_1)$.
-While it is `obvious' from the factorial structure that the KL minimization problem can be performed in the lower dimensional space~$x_1$, by simply dropping the term~$q_0(x_0)$.
-This is confirmed by the KL divergence decomposition itself.
-Assume~$q_0$ is fixed:
-\begin{eqnarray*}
-D(q_0 p_1 \| q_0 q_1) 
- & \equiv & 
- \int \left[
- q_0(x_0)p_1(x_1) 
- \log q_0(x_0)q_1(x_1) - q_0(x_0)q_1(x_1)
- \right]
- dx_0 dx_1 \\
-& = & 
- c + K_0 D(p_1\|q_1),
-\end{eqnarray*}
-so, if we maximize $D(q_0p_1\|q_0q_1)$ at fixed~$q_0$, we maximize $D(p_1\|q_1)$.
-By the way, we shall easily prove that the maximization at fixed~$q_1$ maintains~$q_0$.
-Here, we exploit a separability property of the KL divergence: another strong argument for choosing this cool divergence measure.
-
-Using coordinate block optimization, we also displace the fix point since we relax the $\alpha$-divergence minimization. The new fixed point is defined by:
-$$
-\int q_b^{1-\alpha}(x_b)p_b^\alpha(x_b) \phi(x_b) dx_b
-= \int q_b(x_b) \phi(x_b) dx_b,
-$$
-for each block~$b$.
-In other words, for each block, $q_b$ minimizes the $\alpha$-divergence from the coordinate application $p_b$, itself implicitly defined by~$q_b$ since it depends on the center of~$q_b$.
-
-What is the limit of this iteration if it exists?
-For any point~$a$, the coordinate applications $p_b(x_b)=p(a_{-b},x_b)$ are well defined, and there is a unique factorial distribution~$q_a=Q(a)$ that minimizes the reduced $\alpha$-divergence for each coordinate block.
-Consider the function~$C$ that maps any~$q$ to its center.
-The limit of the iteration must be a fixed point of $C\circ Q$, that is, we need to have $a = C(Q(a)$.
-We will admit for now that this function has a fixed point.
-
-It is a parametric distribution $q$ such
-
-\section{Constrained KL minimization}
-
-To warrant that the Gaussian fit we are looking for is a proper distribution (has strictly negative second-order coefficients), we may assume that the parameters~$\theta$ associated with second-degree monomials are upper bounded by fixed values.
-These bounds may be interpreted in terms of maximum variance along each coordinate.
-
-Roughly speaking, if we have a prior on~$x$ (which amounts to some kind of regularization), we expect the target to be (much) more sharply peaked than the prior, so we may just set the prior variance as a maximum variance parameter.
-
-\section{Why?}
-
-Why not trying to directly minimize the global KL divergence $D(p\|q)$? The short answer is: because this is intractable.
-
-Expectation propagation and other message passing schemes \cite{Minka-01,Minka-05} provide ways to work around this problem.
-However, the core motivation for EP has remained unclear.
-Minka illustrated how EP provides explicit update rules in a `clutter problem' where direct KL minimization is intractable, but never gave a clear general motivation for EP.
-
-Other authors stressed the computational advantage of distributed inference via data partitioning \cite{Vehtari-14}.
-This would tend to restrict EP to `big data' problems...
-However, direct KL minimization may be intractable in many `small data' problems.
-
-The whole EP framework rests upon the `contextual approximation trick', which can be summarized by the formula: $p\leftarrow q^{1-\alpha}p^\alpha$.
-This can be interpreted as a relaxation: replace in the KL divergence the target $p$ with another target that is presumably much closer to the approximation space, hence it is intuitively `easier' to solve the relaxed KL minimization problem.
-
-Why is it so?
-Minka's clutter example is one in which the relaxed problem enjoys a simple closed-form solution, while the original problem requires a heavy combinatorial computation.
-In many cases, there is not even a closed-form solution to either problem, so that an approximate KL minimization needs be performed.
-This is where some methods, including VS, come into play.
-In such situations, relaxing the problem is a way to reduce the approximation error.
-Nothing to do with data partitioning!
-
-How comes the same idea has too very distinct motivations: memory load, on the one hand, and computational tractability, on the other hand?
-Is it a fortunate coincidence?
-Which one of the two is the stronger?
-
-%The way I see it at the moment is that there are two distinct ideas underlying the EP~algorithm family.
-
-Distributed inference means varying the target factor across iterations by resampling data (randomly or not).
-The most straightforward implementation of this idea is actually the AEP algorithm \cite{Dehaene-18}.
-The EP algorithm itself may be seen as a more sophisticated variant of AEP, one that is also much more memory demanding, that keeps track of all the available data in some sense.
-The question is: do we need this at all?
-I see no reason why, but one remark we shall make is that AEP requires an annealing scheme for the learning rate~$\alpha$ in order to converge, while~$\alpha$ can be kept constant in~EP.
-I will further elaborate on this later on.
-
-If we keep the factor constant, which means that it is actually our target distribution, then AEP becomes a stochastic version of an adaptive $\alpha$-divergence minimization scheme already suggested by Minka \cite{Minka-05} that works by iterative $M$-projections (IMP).
-IMP is not concerned in any way with distributed inference, its only purpose is to implement $\alpha$-divergence minimization, which may be seen as a relaxation of KL divergence minimization.
-In this context, $\alpha$ is no longer a learning rate, it is a numerical parameter that controls the relaxation error: the larger, the better, however this is counterbalanced by the practical KL~approximation error: the larger~$\alpha$, the worse.
-In practice, $\alpha$ should be either fixed, or increased throughout the process to better approximate KL~minimization.
-
-It thus seems that~$\alpha$ is trying to serve two distinct and somewhat contradictory objectives.
-The way I see to solve the dilemma is to introduce two parameters: a relaxation rate~$\alpha$ and a learning rate~$\rho$, and consider the more general scheme whereby the target~$p$ itself is adaptive:
-$$
-p \leftarrow q^{1-\rho} h^\rho,
-\qquad
-q \leftarrow \text{proj}(q^{1-\alpha}p^\alpha),
-$$
-where~$h$ is some empirical innovation (typically, the distribution associated with a mini-batch).
-For the target update to converge, we seem to need the learning rate to decrease to zero across iterations.
-We also note the following:
-$$
-q^{1-\alpha}p^\alpha = q^{1-\alpha\rho}h^{\alpha\rho},
-$$
-so, this is exactly the same as the above algorithm if we substitute~$\alpha$ by $\alpha\rho$.
-In other words, this algorithm cannot distinguish between learning and relaxation.
-If using a small learning rate~$\rho$, which should also be decreased across iterations for convergence, we can afford setting $\alpha=1$ so that $\alpha\rho$ may be interpreted as a pure learning rate.
-
-The original AEP corresponds to a slightly different setting where we have multiple deterministic factors $h_1,\ldots,h_n$ as opposed to a single stochastic one.
-In this case, each iteration picks one of the $h_i$'s and it makes sense to set $\rho=1/n$.
-Hence, AEP corresponds to a fixed $\alpha\rho=1/n$.
-The problem is that this may not converge since~$\rho$ is fixed!
-
-I presume that \cite{Dehaene-18} makes no mention of this little issue because they consider AEP as a theoretical tool to study the asymptotic behavior of EP.
-However, \cite{Li-15} use AEP for practical matters under the name `stochastic EP', and they note that ``SEP converges to the same fixed points as AEP if the learning rates satisfy the Robbins-Monro condition'', thereby implicitly stating that AEP does not converge if the learning rate is kept constant.
-
-Can we have $\alpha\rho$ constant with~$\rho$ decreasing to zero?
-Say, $\rho=1/N$ where~$N$ is the iteration number.
-It would imply that $\alpha = N/n \uparrow \infty$.
-I assume that IMP requires $\alpha \leq 1$ to work, so this is not good.
-Basically, if $\alpha\leq 1$, $\alpha\rho\leq \rho$ so it needs to converge to zero as well.
-
-EP avoids this convergence issue using separate approximations for each fixed factor.
-Using fixed factors is not enough, as AEP proves, therefore one wonders if fixed factors have any advantage in realistic large-scale learning context.
-Since EP can maintain a fixed learning rate, it might converge faster than my simple AEP, but it's not clear what problem EP solves and the memory demand is a real burden.
-My simple AEP has only one fixed point, which is the KL~minimizer.
-
-\section{Stochastic fixed point}
-
-We know that, in the special case where~$h=p$ is fixed, the IMP fixed point for any fixed~$\alpha$ is the $\alpha$-divergence minimizer.
-If we decrease~$\alpha$ to zero according or not to RM, we expect the algorithm to converge to the exclusive KL divergence minimizer $\arg\min_q D(q\|p)$, because the 0-divergence is the exclusive KL divergence.
-
-What happens in the stochastic IMP, that is, in the more general case where~$h$ is a random estimate of~$p$? Assume $\mathbb{E}(h)=p$. I speculate that, in the limit of small~$\alpha$,
-$$
-\mathbb{E}[\text{proj}(q^{1-\alpha}h^\alpha)]
-\approx \text{proj}(q^{1-\alpha}p^\alpha).
-$$
-
-This is, at least, what intuition tells us, since the result is obviously true if~$h$ is deterministic, so it has to hold at first order.
-Therefore, under the condition $\alpha\approx 0$, we are back to the deterministic case: the stochastic iteration will minimize the exclusive KL divergence~$D(q\|p)$ on~$Q$.
-Not the inclusive one, as was our initial objective, but something related...
-not too bad an outcome.
-If~$p$ is unimodal, it does not matter much which divergence we choose.
-If it has multiple modes, then the algorithm will likely pick the strongest (not necessarily the highest).
-Anyway, it is not clear which $\alpha$-divergence is most appropriate to our problem.
-
-\section{Stochastic approximation}
-
-BTW, why is Li referring to Robbins-Monro? Consider a stochastic iteration of the form: $$
-x_{n+1} = f(x_n, \epsilon_n, \omega_n),
-$$
-where $\epsilon_n$ is a deterministic learning rate and $\omega_n$ is random.
-Assuming $\omega_n\perp x_n$, this defines a Markov chain.
-Under the regime of small learning rates, we have by a order-1 Taylor expansion:
-$$
-f(x_n, \epsilon_n, \omega_n) 
-\approx f(x_n, 0, \omega_n) + \epsilon_n \frac{\partial f}{\partial \epsilon} (x_n, 0,\omega_n)
-$$
-
-Let us now assume that $f(x, 0, \omega)=x$ holds whatever~$\omega$ and consider the derivative of~$f$ wrt~$\epsilon$ at $\epsilon=0$. The iteration is approximated by a Robbins-Monro algorithm:
-$$
-x_{n+1} = x_n + \epsilon_n \frac{\partial f}{\partial \epsilon}(x_n,0,\omega_n),
-$$
-
-Therefore, in the small learning rate regime, it should have more or less the same asymptotic behavior as Robbins-Monro under the same conditions.
-In particular, if~$\epsilon_n\sim O(n^{-\alpha})$ with $1/2<\alpha\leq 1$, $x_n$ should converge with probability one to the root of:
-$$F(x)
-\equiv 
-= \frac{\partial \mathbb{E}(f)}{\partial \epsilon} (x,0),
-$$
-(by linearity), therefore:
-$$
-x_n \stackrel{p}{\to} x_\star,
-\qquad 
-\frac{\partial \mathbb{E}(f)}{\partial \epsilon} (x_\star,0)=0.
-$$
-
-In our case:
-$$
-f(q,\epsilon,h) = \text{proj}(q^{1-\epsilon}h^\epsilon)
-$$
-
-\section{Exclusive KL divergence minimization}
-
-I remember being asked by Merlin what is the analytical expression of the Gaussian that minimizes exclusive KL divergence, replying that it is easy to find out from the equations, and then realizing that there is no closed-form solution.
-$$
-D(q\|p) = \int (q \log\frac{q}{p} - q + p)
-$$
-
-Say $q=\exp(\theta^\top \phi)$. 
-$$
-\nabla D(\theta) = 
-\int (\log\frac{q_\theta}{p}) q_\theta\phi.
-$$
-
-Optimality condition:
-$$
-\int q_\theta \phi \phi^\top \theta
-= \int q_\theta (\log p) \phi
-$$
-
-Forget it... What about the hessian?
-$$
-H = \int q_\theta(1 + \log\frac{q_\theta}{p}) \phi\phi^\top 
-$$
-
-\section{Exponentially weighted averages}
-
-How do you compute a moving average in real time? Measure $y_n$ and smooth over time using the recursion:
-$$
-x_0 = 0,
-\qquad
-x_{n+1} = (1-\lambda) x_n + \lambda y_n.
-$$
-
-You can think of this as averaging approximately over the latest $1/\lambda$ samples. More precisely,
-$$
-x_n = \lambda \sum_{i=1}^n (1-\lambda)^i y_{n-i}.
-$$
-
-So the weight assigned to sample~$i$ is $w_i=\lambda (1-\lambda)^{n-i}$. The sum of weights is:
-$$
-\sum_{i=1}^n w_i = 1 - (1-\lambda)^n,
-$$
-so it is closer to one as $n$ increases. For small~$n$, we have bias. To correct $x_n$ for it, we just need to divide it by the sum of weights:
-$$
-x_n^c = \frac{x_n}{1 - (1-\lambda)^n}
-$$
-
-Just a thought: if you want to predict the future from the past, your prediction will always be late on reality, and this is simply because you need to average past data to make accurate predictions.
-
-\section{Stabilization}
-
-The bridge approximation algorithm described in this amazingly well organized document is an iterative process that seeks a fixed point of an $\alpha$-divergence proxy.
-How can we make sure it converges in practice?
-
-There are different parameters that may significantly impact convergence:
-\begin{enumerate}
-\item The learning rate $\epsilon$. A small learning rate will help stabilize the algorithm, and that's what it is intended for. 
-\item The divergence index $\alpha$.
-It looks like $\alpha$ needs to lie in the range $[0,1]$.
-The closer to zero, the smaller the changes on each iteration, hence the more stable the algorithm, although the slower the convergence of course.
-\item The initial fit.
-The closer the initial fit to the target, the more likely convergence.
-As a default rule, the initial fit could be chosen as the prior distribution of the underlying Bayesian problem, hence an annealing procedure on the prior variance could help convergence: sequentially run bridge approximation for increasingly larger prior variances.
-\item Block coordinate size acts on optimization dimensionality: the lower the dimension, the more accurate VS.  condition.
-\end{enumerate}
-
-However, among all these parameters,
-
-\section{One-step VS}
-
-In the bridge approximation scheme, where VS is used repeatedly to improve the fit by decreasing $D(q^{1-\alpha}p^\alpha\|q')$ on each iteration, it might not be needed to use a fully blown optimizer such as L-BFGS.
-Instead, one could do a mere gradient update to try and save computation time and memory load.
-
-Assume we are using VS to try and fit $p=wf$ with $w g_\theta$. Consider first the case where $g_\theta = c$ is simply a constant. We can find the optimal constant explicitly via:
-$$
-c  = \langle f \rangle_w,
-$$
-assuming unit sum of weights, {\em i.e.}, $\langle 1 \rangle_w=1$.
-
-From there, we can simply improve the fit by running a gradient update in a larger parametric space.
-Let us consider $\theta_c = (\log c, 0, 0, \ldots)$ the Gaussian parameter corresponding to the constant fit:
-$$
-\theta = \theta_c + \rho D(\theta_c),
-\qquad \text{with} \quad
-D(\theta_c) = - \nabla \hat{C}(\theta_c)
-= \langle f\phi \rangle_w - c \langle \phi \rangle_w.
-$$
-
-For a sufficiently small damping value~$\rho$, this will decrease the KL divergence, hence improving the fit $wg_\theta$ over~$w$.
-We note that the first component of $D(\theta_c)$ (the one associated with the constant term) is zero since $c\langle 1 \rangle_w - \langle f \rangle_w = 0$ by definition.
-
-$$
-D(\theta_c) = c [E_{wf}(\phi) - E_{w}(\phi)]
-$$
-
-\section{BS}
-
-$$
-F(q,h) = \lim_{\epsilon\to 0} 
-\frac{\text{proj}(q^{1-\epsilon}h^\epsilon)-q}{\epsilon}
-$$
-
-The projection operator is defined implicitly by:
-$$
-f(p,q) = \int (q-p) \phi = 0
-$$
-
-Thus, by the implicit function theorem:
-$$
-\frac{\partial \text{proj}()}{\partial \epsilon}
-= - \left(\frac{\partial F}{\partial \theta}\right)^{-1} \frac{\partial F}{\partial p}
-$$
-
-$
-f^{\epsilon}= \exp [\epsilon\log f]
-\to \log f f^{\epsilon}$
-
-We have:
-$$
-\frac{\partial q^{1-\epsilon}h^\epsilon}{\partial \epsilon}
-=
-q (\log h/q) (h/q)^\epsilon
-$$
-
-The algorithm is of the form:
-$$
-q \leftarrow f(q, \epsilon)
-$$
-where, in our case,
-$$
-f(q,\epsilon) = \text{proj}(q^{1-\epsilon} h^\epsilon)
-$$
-
-$$
-f(q,0) = q
-$$
-
-$$
-f(q,\epsilon) \approx f(q,0) + \epsilon \frac{\partial f}{\partial \epsilon}
-$$
-
-$$
-\int q_{n+1} \phi = \int q_n^{1-\epsilon} h_n^\epsilon \phi
-$$
-
-If $q_n$ converges to a fixed value $q_*$, the expectations need be equal so we need to have: 
-$$
-\int q_*\phi = \int q_*^{1-\epsilon}p^\epsilon \phi
-$$
-
-$$
-x_* = E[f(x_*)] 
-$$
-
-We note that the projection operator is defined implicitly by:
-$$
-F(p,\theta) = \int (q_\theta-p)\phi = 0
-$$
-
-Thus, 
-$$
-\frac{\partial \theta}{\partial p}
-= - \left(\frac{\partial F}{\partial \theta}\right)^{-1} \frac{\partial F}{\partial p}
-$$
-
-$$
-\frac{\partial \theta}{\partial p} = \left(\int q_\theta \phi\phi^\top \right)^{-1} \phi
-$$
-
-My take on this is simply that we should use a single random factor in practice.
-
-Robbins-Monro for KL-min. We want to solve:
-$$
-\theta_\star = \arg\min_\theta C(\theta),
-\qquad
-C(\theta) = -\int p \log q_\theta + \int q_\theta
-$$
-
-Optimality condition:
-$$
-M(\theta)\nabla D(\theta) = 0 = \int (q_\theta-p)\phi
-$$
-
-Robbins-Monro algorithm. Consider a random batch via the associated empirical distribution~$h$, and define:
-$$
-N(\theta) = \int (q_\theta-h) \phi
-$$
-
-Clearly, $E(N)=M$ so we can do:
-$$
-\theta_{n+1} = \theta_n - \epsilon_n N(\theta_n)
-$$
-
-Why is this bad? Because $\int h\phi$ is intractable.
-
-\section{More BS}
-
-MisoSoupe
-Thorstentungstène
-Irina
-Edrot
-LouicheSantouche
-Conovaire
-HamBaLaBa
-Thierry Margarine
-EricDARTELAshburner
-Heartscore
-Ze Gazbites
-Gudrun Eikili
-
-If the context is close enough to the target, VS may improve the fit.
-However, if the context is too `wide', the fit will likely be meaningless.
-On the other hand, for a `narrow' context, VS will pretty much behave like a Taylor expansion.
-So, if we have to choose between the two options, it's clear that we prefer the second one: a narrow $w$.
-
-The choice of context is, however, external to VS and rests upon the context update rule.
-Given a current fit $q$, what is the best context for the next update?
-It would be natural to answer $w=q$, but it's not the only reasonable choice.
-We can think of gradient descent as picking $w(x)=\delta(x-x_q)$ with $x_q=\arg\max q$ as a context function.
-
-But the target update is only part of a wider picture...
-
-The problem I see with this, is that the algorithm may get stuck in a poor local approximation if the actual target is `rugged'.
-
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[scale=1.7]{universe}
-%\caption{The Universe}
-%\label{fig:universe}
-%\end{figure}
+Note that we are not talking about overfitting, at least in a traditional sense.
 

### 2023-05-08 cv/cv.tex
Update CV

@@
 
-\noindent Hands-on technical team leader, engineer at heart, PhD, with a specialist expertise in artificial intelligence and data analysis.
+\noindent Hands-on technical team leader, engineer at heart, PhD, with a specialist expertise in computer vision and artificial intelligence.
 Has operated in public research and private R\&D in various fields (healthcare, electro-domestic, entertainment).
 Strong experience in the development of both open source and commercial software in Python, C, C++.
-{\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, medical image analysis, clinical studies.
+{\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, computer-aided diagnosis, smart appliances, digital avatars.
 
@@
 \ \\
-\noindent \textbf{SCIENTIFIC PROGRAMMING:} 15 years experience programming in Python and~C language in the SciPy ecosystem.
+\textbf{COMPUTATIONAL STATISTICS:} 25 years experience developing inferential statistical models for machine learning, computer vision, computer graphics, \ldots
 
 \ \\
-\textbf{STATISTICS:} Strong experience developing inferential statistical models for machine learning, computer vision, computer graphics, and performing data analysis for cognitive research, clinical studies, software validation, \ldots
+\noindent \textbf{SCIENTIFIC PROGRAMMING:} 15 years experience programming in Python and~C language in the SciPy ecosystem.
 
@@
 
-\bighead{HEAD OF DIDIMO AUTOMATION TEAM}{Didimo, Porto, Portugal}{2019--present}
+\bighead{HEAD OF ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
 \nopagebreak
@@
 \item Leading development of generation pipeline of high-fidelity avatars from photos.
-{\em Keywords:} facial reconstruction, ``deep face'', animation retargetting, image feature extraction, image restoration.
+{\em Keywords:} facial reconstruction, deep face, animation retargeting, image feature extraction, image restoration.
 \end{itemize}
@@
   & \tabitem 116 scholarly publications.\\
-  & \tabitem Google scholar statistics: 7308 citations, h-index: 35, i10-index: 72 (October 2022).\\
+  & \tabitem Google scholar statistics: 7603 citations, h-index: 35, i10-index: 74 (May 2023).\\
   & \tabitem 5 published international patents. 2 ongoing applications.\\

### 2023-05-08 papers/variana/notes.tex
Update CV

@@
 \maketitle
+
+\section{2023 thoughts}
+
+Context: ill posedness (in the sense of Hadamard) of empirical risk minimization, including  maximum likelihood, MAP, etc.
+These methods are at the heart of statistical inference and machine learning, still they are based on ill-posed mathematical problems!
+Shouldn't we worry about that?
+
+The empirical risk is just a proxy for a Bayesian risk that we are not able to compute unless we know the {\em true} distribution of the training data.
+We don't really mean to optimize the empirical risk.
+
+Bayesian inference boils down to a problem of the form:
+$$
+r(\theta) = \int p(x) L(x, \theta) dx 
+$$
+where $p(x)$ is the distribution of some variable of interest and $\theta$ is the parameter to be learned.
+Because we don't know $p(x)$ in agnostic learning context, or we know it but it's intractable, we may resort to an empirical version of the risk:
+$$
+\hat{r}(\theta) = \frac{1}{n} L(x_i, \theta).
+$$
+
+You can use the Laplace approximation at the global optimum, it might work just as well as VS and probably faster.
+However, finding the global optimum is super challenging in high dimension: it usually exists (under mild conditions), but is hopeless to track using differential calculus based optimization, which generally converges to a saddle point.
+This is a fundamental limitation of numerical optimization techniques.
+Needless to say, Laplace is bound to work poorly at a saddle point since the Hessian is not definite positive, hence what you will get is a degenerate Gaussian fit.
+
+Yet, according to the Bernstein-von Mises theorem, we expect Bayesian posteriors to be nearly Gaussian in the limit of large samples.
+How many data points do we need in practice, we don't know, but we should be able to tell how well this approximation holds in the problem at hand.
+
+If it does hold, then the target distribution is peaked around a well-defined global optimum. Finding the global optimum is then akin to fitting a Gaussian distribution.
+
+If it doesn't hold, then finding the global optimum makes little sense because we may have multiple local optima spread across the domain.
+But fitting a Gaussian may still make sense as it can at least tell us that the asymptotic regime is not reached, meaning that more training data is probably needed.
+
+The fact that we get many saddle points in practice suggests that we may lack data, however this restriction is not fundamental if we have an underlying generative model to produce the training data.
 

### 2024-02-06 cv/cv.tex
Work on CV

@@
 
-\noindent Hands-on technical team leader, engineer at heart, PhD, with a specialist expertise in computer vision and artificial intelligence.
+\noindent Hands-on technical team leader, applied researcher and engineer at heart, PhD, with a specialist expertise in computer vision and artificial intelligence.
 Has operated in public research and private R\&D in various fields (healthcare, electro-domestic, entertainment).
-Strong experience in the development of both open source and commercial software in Python, C, C++.
-{\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, computer-aided diagnosis, smart appliances, digital avatars.
+Strong experience in the development of both open source and commercial software.
+{\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, computer-aided diagnosis, smart appliances, digital avatars, Python, C.
 
@@
 \ \\
-\noindent \textbf{SCIENTIFIC REPORTING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 published international patents.
+\noindent \textbf{SCIENTIFIC REPORTING:} Author of more than 100 peer-reviewed scientific articles published in international journals, proceedings and books, and 6 published international patents.
 
@@
 \item Leading development of generation pipeline of high-fidelity avatars from photos.
-{\em Keywords:} facial reconstruction, deep face, animation retargeting, image feature extraction, image restoration.
+{\em Keywords:} facial reconstruction, deep face, animation retargeting, image processing.
+Accepted talk at SIGGRAPH'22.
 \end{itemize}
@@
 \begin{itemize}
-\item Led algorithmic development of a brain morphometry tool (released as part of the {\em AI-Rad Companion} Siemens solutions) to help radiological reading for patients with suspected neurodegeneration.
-{\em Keywords:} magnetic resonance imaging (MRI), brain image segmentation, machine learning for disease classification.
+\item Led algorithmic development of computer-aided diagnosis software (released by Siemens Healthineers as part of the {\em AI-Rad Companion} solution) using brain morphometry for radiological reading.
+{\em Keywords:} magnetic resonance imaging (MRI), brain image segmentation, disease classification, neurodegeneration, Alzheimer's disease.
 \item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
@@
   \textbf{Publications:} &
-  \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}. \\
+  \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}.\\
   & \tabitem 116 scholarly publications.\\
   & \tabitem Google scholar statistics: 7603 citations, h-index: 35, i10-index: 74 (May 2023).\\
-  & \tabitem 5 published international patents. 2 ongoing applications.\\
+  & \tabitem 6 published international patents. 2 ongoing applications.\\
   %\hline

### 2024-02-06 cv2/cv.tex
Work on CV

% !TeX spellcheck = en_GB
% !TeX program = pdflatex
%
% LetzCV-sleek 1.0 LaTeX template
% Author: Andreï V. Kostyrka, University of Luxembourg
%
% This template fills the gap in the available variety of templates
% by proposing something that is not a custom class, not using any
% hard-coded settings deeply hidden in style files, and provides
% a handful of custom command definitions that are as transparent as it gets.
% Developed at the University of Luxembourg.
%
% *NOTHING IS HARCODED, and never should be.*
%
% Target audience: applicants in the IT industry, or business in general
%
% The main strength of this template is, it explicitly showcases how
% to break the flow of text to achieve the most flexible right alignment
% of dates for multiple configurations.

\documentclass[11pt, a4paper]{article}

\usepackage[T1]{fontenc}     % We are using pdfLaTeX,
\usepackage[utf8]{inputenc}  % hence this preparation
\usepackage[british]{babel}  
\usepackage[left = 0mm, right = 0mm, top = 0mm, bottom = 0mm]{geometry}
\usepackage[stretch = 25, shrink = 25]{microtype}  
\usepackage{graphicx}        % To insert pictures
\usepackage{xcolor}          % To add colour to the document
\usepackage{marvosym}        % Provides icons for the contact details

\usepackage{enumitem}        % To redefine spacing in lists
\setlist{parsep = 0pt, topsep = 0pt, partopsep = 1pt, itemsep = 1pt, leftmargin = 6mm}

\usepackage{FiraSans}        % Change this to use any font, but keep it simple
\renewcommand{\familydefault}{\sfdefault}

\definecolor{cvblue}{HTML}{304263}

%%%%%%% USER COMMAND DEFINITIONS %%%%%%%%%%%%%%%%%%%%%%%%%%%
% These are the real workhorses of this template
\newcommand{\dates}[1]{\hfill\mbox{\textbf{#1}}} % Bold stuff that doesn’t got broken into lines
\newcommand{\is}{\par\vskip.5ex plus .4ex} % Item spacing
\newcommand{\smaller}[1]{{\small$\diamond$\ #1}}
\newcommand{\headleft}[1]{\vspace*{3ex}\textsc{\textbf{#1}}\par%
    \vspace*{-1.5ex}\hrulefill\par\vspace*{0.7ex}}
\newcommand{\headright}[1]{\vspace*{2.5ex}\textsc{\Large\color{cvblue}#1}\par%
     \vspace*{-2ex}{\color{cvblue}\hrulefill}\par}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[colorlinks = true, urlcolor = white, linkcolor = white]{hyperref}

\begin{document}

% Style definitions -- killing the unnecessary space and adding the skips explicitly
\setlength{\topskip}{0pt}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}
\setlength{\fboxsep}{0pt}
\pagestyle{empty}
\raggedbottom

\begin{minipage}[t]{0.33\textwidth} %% Left column -- outer definition
%  Left column -- top dark rectangle
\colorbox{cvblue}{\begin{minipage}[t][5mm][t]{\textwidth}\null\hfill\null\end{minipage}}

\vspace{-.2ex} % Eliminates the small gap
\colorbox{cvblue!90}{\color{white}  %% LEFT BOX
\kern0.09\textwidth\relax% Left margin provided explicitly
\begin{minipage}[t][293mm][t]{0.82\textwidth}
\raggedright
\vspace*{2.5ex}

\Large Alexis \textbf{\textsc{Roche}} \normalsize

% Centering without extra vertical spacing
\null\hfill
\begin{center}
\includegraphics[width=0.65\textwidth]{alexis_didimo_cropped.png}\\
\caption{\scriptsize My (hairless) digital twin}
\end{center}
\hfill\null

\vspace*{0.5ex} % Extra space after the picture

\headleft{Profile}
 Hands-on technical team leader, {\bf applied math} researcher and engineer at heart, with a specialist expertise in {\bf computer vision} and {\bf computer graphics}.

\headleft{Contact details}
\small % To fit more content
\MVAt\ {\small alexis.roche@gmail.com} \\[0.4ex]
\Mobilefone\ +351\,910\,327\,756 \\[0.5ex]
\Mundus\ \href{https://www.linkedin.com/in/alexis-roche-3815592}{LinkedIn profile} \\[0.1ex]
\Mundus\ \href{https://orcid.org/0000-0002-4821-6893}{ORCID profile} \\[0.1ex]
\Mundus\ \href{https://scholar.google.com/citations?user=LJJTi1gAAAAJ&hl}{Google Scholar profile} %\\[0.1ex]
%\Letter\ Porto, Portugal
\normalsize

\headleft{Personal information}
\MALE \\[0.5ex]
%Year of birth: \textbf{1861} \\[0.5ex]
\Letter\ Porto, {\bf Portugal} \\[0.5ex]
Citizenship: {\bf France} \\[0.5ex]
%Family: \textbf{Single without children} \\[0.5ex]
Languages: {\bf English} (fluent), {\bf French} (native).

%\headleft{Skills}
%\begin{itemize}
%\item Python, SQL, PySpark
%\item R, Matlab, Azure Databricks
%\item MS Word, Excel, PowerPoint
%\item Communication and team collaboration
%\end{itemize}

\end{minipage}%
\kern0.09\textwidth\relax%%Right margin provided explicitly to stretch the colourbox
}
\end{minipage}% Right column
\hskip2.5em% Left margin for the white area
\begin{minipage}[t]{0.56\textwidth}
\setlength{\parskip}{0.8ex}% Adds spaces between paragraphs; use \\ to add new lines without this space. Shrink this amount to fit more data vertically

\vspace{2ex}

\headright{Experience}

{\sc Head of AI.} {\it Didimo}, Porto, Portugal.  \dates{Sep 2019--present} \\
\smaller{Leading development of automated 3D character creation for {\em Popul8} software: shape reconstruction / generation, rig retargeting, skin texturing, asset fitting, \ldots}

\is % Item spacing -- defined in the preamble
{\sc Senior Computer Vision Scientist.} {\it CoVii (Ar\c celik group)}, Porto, Portugal. \dates{Feb 2017--Jul 2019} \\
\smaller{Developed embedded algorithms for smart domestic appliances: {\em VUXHub} and {\em Artisan oven} prototypes (demonstrated at IFA Berlin, 2017-18).}

\is
{\sc Lead Clinical Research.} {\it Siemens Healthineers / University Hospital (CHUV)}, Lausanne, Switzerland. \dates{Sep 2011--Jan 2017} \\
\smaller{Led development and validation of embedded brain morphometry for radiological reading (released in 2020 as {\em AI-Rad Companion Brain MR} solution).}

\is
{\sc Academic Guest.} {\it Federal Institute of Technology (ETHZ)}, Zurich, Switzerland. \dates{May 2009--Jul 2011} \\
\smaller{Advised the Computer Vision Lab (BIWI) on brain image analysis research.}

\is
{\sc Permanent Researcher.} {\it French Atomic Commission (CEA)}, Paris-Saclay, France. \dates{Dec 2002--Jul 2011} \\
\smaller{Investigated various brain MRI analysis methods at the Neurospin brain imaging center.}

\is
{\sc Research Assistant.} {\it University of Oxford}, UK. \dates{Apr 2001--Oct 2002} \\
\smaller{Developed advanced medical image registration algorithms and consulted for Mirada Solutions Ltd (now Siemens Molecular Imaging).}

\headright{Education}

{\sc PhD in Engineering Science.} {\it INRIA / Universit\'e C\^ote d'Azur}, Sophia Antipolis, France. \dates{Sep 1997--Mar 2001}

\is
{\sc MsC in Cognitive Science.} {\it Sorbonne Universit\'e}, Paris, France. \hphantom{CACACACA} \dates{Sep 1995--Sep 1996}

\is
{\sc Centralien Engineer Degree (equivalent MsC).} {\it CentraleSup\'elec, Universit\'e Paris-Saclay}, France. \dates{Sep 1993--Jun 1996}

\headright{Skills}

{\sc Technical leadership:} Proven ability to coordinate a technical team and manage objectives in multidisciplinary environments.

\is 
{\sc Collaborative programming:} C, C++, Python (including 18 years experience within the NumPy/SciPy ecosystem), Matlab, R, and using Git for version control.

\is 
{\sc Scientific reporting:} Author of 106 peer-reviewed scientific articles and 6 published international patents.

\end{minipage}
\end{document}

### 2024-02-06 papers/variana/notes.tex
Work on CV

@@
 \def\B{\mathbf{B}}
+\def\prec{\mathrm{prec}}
 
@@
 
-\section{2023 thoughts}
-
-Context: ill posedness (in the sense of Hadamard) of empirical risk minimization, including  maximum likelihood, MAP, etc.
-These methods are at the heart of statistical inference and machine learning, still they are based on ill-posed mathematical problems!
-Shouldn't we worry about that?
-
-The empirical risk is just a proxy for a Bayesian risk that we are not able to compute unless we know the {\em true} distribution of the training data.
-We don't really mean to optimize the empirical risk.
-
-Bayesian inference boils down to a problem of the form:
-$$
-r(\theta) = \int p(x) L(x, \theta) dx 
-$$
-where $p(x)$ is the distribution of some variable of interest and $\theta$ is the parameter to be learned.
-Because we don't know $p(x)$ in agnostic learning context, or we know it but it's intractable, we may resort to an empirical version of the risk:
-$$
-\hat{r}(\theta) = \frac{1}{n} L(x_i, \theta).
-$$
-
-You can use the Laplace approximation at the global optimum, it might work just as well as VS and probably faster.
-However, finding the global optimum is super challenging in high dimension: it usually exists (under mild conditions), but is hopeless to track using differential calculus based optimization, which generally converges to a saddle point.
-This is a fundamental limitation of numerical optimization techniques.
-Needless to say, Laplace is bound to work poorly at a saddle point since the Hessian is not definite positive, hence what you will get is a degenerate Gaussian fit.
-
-Yet, according to the Bernstein-von Mises theorem, we expect Bayesian posteriors to be nearly Gaussian in the limit of large samples.
-How many data points do we need in practice, we don't know, but we should be able to tell how well this approximation holds in the problem at hand.
-
-If it does hold, then the target distribution is peaked around a well-defined global optimum. Finding the global optimum is then akin to fitting a Gaussian distribution.
-
-If it doesn't hold, then finding the global optimum makes little sense because we may have multiple local optima spread across the domain.
-But fitting a Gaussian may still make sense as it can at least tell us that the asymptotic regime is not reached, meaning that more training data is probably needed.
-
-The fact that we get many saddle points in practice suggests that we may lack data, however this restriction is not fundamental if we have an underlying generative model to produce the training data.
+\section{History}
+
+It is time to assemble the pieces of the puzzle.
+\begin{itemize}
+\item 2010: discovery of the VS core idea and its intriguing exactness property.
+\item 2011--13: consolidation of the theory behind VS.
+\item 2014--16: realization that VS is useful in practice combined with Expectation Propagation.
+\item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion).
+\item 2022: proven bound on VS asymptotic variance.
+\item 2023: final key insight: VS makes optimization possible in high dimension (at least locally) because it avoids saddle points unlike gradient-based methods.
+\end{itemize}
+
+\section{FAQ}
+
+\paragraph{Is VS better than IS?} No.
+VS is asymptotically equivalent to IS applied to the residual Gaussian fitting error, {\em i.e.}, the difference between the target distribution and its KL optimal Gaussian fit.
+VS will thus be more accurate than IS whenever the residual importance weights vary less than the base importance weights in the sampling distribution support.
+This happens when the target is locally well approximated by a Gaussian.
+But this is not always the case.
+For instance, assume we want to estimate the normalizing constant of a distribution that we are able to sample from; in this case, the base importance weights are constant, and substracting the Gaussian fit doesn't help.
+VS doesn't even converge if the target has shorter tails than a Gaussian, while IS is exact no matter what.
+Generally speaking, VS doesn't appear as the method of choice when we have a good sampler.
+
+%\paragraph{What does ``nearly Gaussian'' mean?} Roughly speaking, it means that the distribution can be decomposed as the product of a Gaussian distribution and a slowly varying function within the Gaussian distribution's support.
+
+\paragraph{Is VS better than Laplace?} The Laplace method has systematic bias, unlike VS.
+This is fundamentally because it works at fixed, infinitesimal scale, while VS works at adjustable scale.
+One can think of simple situations where Laplace fails miserably, such as fitting a Laplace distribution (isn't that ironical?), but it may also work very well provided that the target is nearly Gaussian.
+The problem is that it is impractical in high dimension since it requires maximizing the distribution (which is more or less the problem we are trying to solve using VS).
+In EP context, we can perform a second-order Taylor expansion of the log-factor at the cavity center, yielding a damped Newton iteration.
+The resulting approximate EP is has a tendency to get stuck at saddle points, just like most gradient-based methods.
+VS-based EP is able to escape saddle points by exploring the target distribution in neighborhoods of increasing sizes.
+
+\paragraph{Can I choose any sampling kernel in practice?} Unfortunately, no.
 
@@
 
+\subsection{VS variance inequality}
+
 I have been bumping into this issue for years...
-only to realize now (January 2022) that such an upper bound does exist and is actually pretty simple to exhibit.
+only to realize in January 2022 that such an upper bound does exist and is actually pretty simple to exhibit.
 We will need a straightforward intermediate result:
@@
 $$
-\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g).
-$$
-
-Note that this is only tight if $\max(f,g)$ is constant,
-
-This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes.
-But this holds under the mild condition that both $f$ and $g$ remain bounded.
-This practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.
+\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g),
+$$
+where the infinite norms are taken on the support of~$\pi$, which is infinite if $\pi$ is a Gaussian kernel.
+
+Note that this is only tight if $\max(f,g)$ is constant.
+The inequality tells us, as intuition suggests, that the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes.
+But this holds under the condition that both~$f$ and~$g$ remain bounded on the support of~$\pi$.
+While this may be problematic if~$\pi$ has infinite support, we can just truncate~$\pi$ to guarantee the finiteness of the upper bound at the price of some possibly small bias.
+As noted in the 2013 paper, this boils down to using a localized version of the KL divergence as the target metric.
+It may also be justified in terms of a Bayesian prior, and may be a very reasonable thing to do.
+However, while removing the concern that the VS variance may be infinite, it doesn't prevent it in practice from being very large and yielding
+hopelessly slow convergence.
+
+The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi_i$. The above result implies straighforwardly:
+$$
+(f-g)\phi_i^2
+\leq 2 \max(f\phi_i, g\phi_i) \ell(f\phi_i, g\phi_i)
+= 2 \max(f\phi_i, g\phi_i) \phi_i\ell(f, g)
+= 2 \max(f\phi_i^2, g\phi_i^2) \ell(f, g)
+$$
+where we use the fact that $\ell(f\phi_i,g\phi_i)=\phi_i\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
+$$
+\int \pi (f-g)^2\phi_i^2 \leq 2\max(\|f\phi_i^2\|_\infty, \|g\phi_i^2\|_\infty) D(\pi f\|\pi g).
+$$
+
+This yields the same kind of variance upper bound for any moment estimator we have.
+
+\subsection{Sufficient conditions for VS convergence}
+
+From this point, can we give practical sufficient conditions for VS convergence?
+Convergence in distribution will occur under mild conditions if the integral $\int \pi (f-g_\star)^2\phi\phi^\top$ is finite.
+The above inequality relating the VS variance to the KL~divergence tells us that this holds whenever $\pi$ has finite support.
+
+What if~$\pi$ is stricly Gaussian, hence doesn't have finite support?
+
+Similarly to IS, the condition that $f$ is upper bounded is sufficient for $\pi f^2 \phi\phi^\top$ to be finite, and may be easy to fulfill in practice.
+
+We are left with another sufficient condition (which IS doesn't have): that $\pi g_\star^2 \phi\phi^\top$ be finite too.
+Due to the log-quadratic nature of both $\pi$ and $g_\star$, this holds iff $\pi g_\star$ is Gaussian (i.e.
+its hegated log is a positive definite quadratic form).
+
+In 1D, let us denote $s^2$ and $\sigma^2$ the respective variances of $\pi$ and $\pi g_\star$.
+Clearly, $\sigma^2$ is finite because it is equal to the variance of $\pi f$ by the moment matching constraint, which is finite if $f$ is upper bounded.
+A simple asymptotic analysis then shows that the condition that $\pi g_\star^2$ has finite integral boils down to:
+$$
+\prec(\pi) + 2 [\prec(\pi f) - \prec(\pi)] > 0,
+\quad \Rightarrow \quad
+\prec(\pi) < 2 \prec(\pi f),
+$$
+where $\prec$ denotes the precision, {\em i.e.}, the inverse variance. The condition boils down to:
+$$
+s^2 > \frac{\sigma^2}{2}.
+$$
+
+The same reasoning can be extended in multiple dimension. If we use factorized Gaussians for fitting, the above condition simply applies to each axis.
+
+In short, the kernel size cannot be too small compared to the target distribution. However, we don't know $\sigma$, so how can we guarantee this in practice?
+
+Let now consider a damping factor $0\leq \alpha \leq 1$, and raise our factor~$f$ to the power~$\alpha$.
+
+is practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.
 
 Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.
-
-The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi$. The above result implies straighforwardly:
-$$
-(f-g)\phi^2
-\leq 2 \max(f\phi, g\phi) \ell(f\phi, g\phi)
-= 2 \max(f\phi, g\phi) \phi\ell(f, g)
-= 2 \max(f\phi^2, g\phi^2) \ell(f, g)
-$$
-where we use the fact that $\ell(f\phi,g\phi)=\phi\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
-$$
-\int \pi (f-g)^2\phi^2 \leq 2\max(\|f\phi^2\|_\infty, \|g\phi^2\|_\infty) D(\pi f\|\pi g).
-$$
-
-This yields the same kind of variance upper bound for any moment estimator we have.
 
@@
 
+\section{Some 2023 thoughts}
+
+You can use the Laplace approximation at the global optimum, it might work just as well as VS and probably faster.
+However, finding the global optimum is super challenging in high dimension: it usually exists (under mild conditions), but is hopeless to track using differential calculus based optimization, which generally converges to a saddle point.
+This is a fundamental limitation of numerical optimization techniques.
+Needless to say, Laplace is bound to work poorly at a saddle point since the Hessian is not definite positive, hence what you will get is a degenerate Gaussian fit.
+
+Yet, according to the Bernstein-von Mises theorem, we expect Bayesian posteriors to be nearly Gaussian in the limit of large samples.
+How many data points do we need in practice, we don't know, but we should be able to tell how well this approximation holds in the problem at hand.
+
+If it does hold, then the target distribution is peaked around a well-defined global optimum. Finding the global optimum is then akin to fitting a Gaussian distribution.
+
+If it doesn't hold, then finding the global optimum makes little sense because we may have multiple local optima spread across the domain.
+But fitting a Gaussian may still make sense as it can at least tell us that the asymptotic regime is not reached, meaning that more training data is probably needed.
+
+The fact that we get many saddle points in practice suggests that we may lack data, however this restriction is not fundamental if we have an underlying generative model to produce the training data.
+
 %\begin{figure}[h!]

### 2024-03-04 papers/composite_inference/notes2024.tex
Notes on composite inference

\documentclass[english]{scrartcl}

\usepackage{fullpage}
\usepackage{amstext,amssymb,amsmath,amsthm}
\usepackage{graphicx,subfigure}
\usepackage{multirow}
\usepackage{url}
\usepackage{hyperref}
\usepackage{color}

\graphicspath{{.}{pics/}}

\title{Composite Bayesian inference in hindsight}
%\date{}
\author{Alexis Roche\thanks{\url{alexis.roche@centraliens.net}}}

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\Blambda}{{\boldsymbol{\Lambda}}}
\newcommand{\bell}{{\boldsymbol{\ell}}}
\newcommand{\E}{\mathbb{E}}

\newcommand{\isep}{\mathrel{{.}\,{.}}\nobreak}

\begin{document}

\maketitle

\section{Introduction}

In the years 2015--2019, I worked on an attempt to put forward a non-conventional statistical inference paradigm, which I called composite Bayesian inference.

Composite Bayesian inference is a new viewpoint on the ``naive Bayes'' method that justifies it without relying on implausible independence assumptions. It also comes with a posterior distribution, which is a scaled version of the ``naive Bayes'' posterior, enabling the computation of conservative credibility sets.

Whether composite Bayesian inference is the right way to look at naive Bayes methods is questionable. But the interpretation it provides seems to be useful.

I tried to generalize it it using the concept of ``super composite likelihood'' function, where the weights assigned to the multiple atomic features are dependent on the variable to be inferred, {\em i.e.}, they are variable-of-interest dependent (VOID).

My ultimate goal here was to justify image registration similarity measures in a neat statistical inference framework.

I didn't do a great job at this. The super composite likelihood could more simply be derived as a regular composite likelihood using VOID features, without the need to introduce a more complex definition. The crux of the problem is to justify likelihood functions based on VOID features, but I didn't do that.

\subsection{VOID feature likelihood}

The idea of VOID feature-based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}. It is definitely a relevant topic, but this initial work was not fully successful and didn't seem to convince many.

On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VOID feature-based likelihood function. As noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.

However, it is easy to see how the PDF~projection method can lead to completely absurd likelihood functions, if the VOID features are chosen arbitrarily. Baggenstoss probably saw that but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''. This is nonsense. The choice of features should be driven by explicit assumptions.

The PDF~projection method simply needs an additional constraint to work. I have known the solution for quite some time but have struggled to fully contemplate it.

\subsection{The missing constraint}

The least we can ask to the VOID feature likelihood is to be asymptotically consistent, that is, to be maximal at the true parameter value in the limit of large samples. If this property does not hold, then the method is worthwhile nothing.

A sufficient condition to guarantee consistency is that the VOID feature function $t_\theta$ statisfies:
$$
\forall (\theta, \theta'),
\quad
\int p(t_{\theta'}|\theta) \log \frac{p(t_{\theta'}|\theta)}{\pi(t_{\theta'})} dt_{\theta'}
  \leq
\int p(t_{\theta}|\theta) \log \frac{p(t_{\theta}|\theta)}{\pi(t_{\theta})} dt_{\theta}
.
$$

This condition may look complicated, but it only says that the VOID feature should always be the most informative about the variable of interest, among all feature candidates. In other words, $t_\theta$ should be, for any~$\theta$, the best statistic to test the hypothesis that~$\theta$ is the correct parameter vs the null hypothesis.

Let's take the same example as in my initial paper on composite inference. Assume we want to infer a disease category (infectious / cardiovascular) using VOID likelihood based on two simple clinical measures, body temperature and blood pressure. The above condition tells us that we need to associate body temperature with the infectious category, and blood pressure with the cardiovascular category. If we do it the other way round, the VOID feature likelihood will at best be useless, at worst biased.

It feels totally obvious. Which doctor in his right mind would look at blood pressure rather than body temperature to diagnose an infectious disease?

\subsection{Discarding evidence}

I had this constraint in mind ever since I started working on composite inference, but I didn't bother to incorporate it into the method. Why?

I believe the main reason is because I was (wrongly) looking for a justification of mutual information as a general similarity measure for image registration. I wanted consistency to follow from the formulation rather than imposing it by design.

I have known for a while that mutual information is not always consistent, contrary to a claim made by \cite{Zollei-09}. It's quite easy to find counter-examples suggesting that noisy images with nonlinearly related intensities may break consistency, meaning that mutual information may be maximal far away from the correct alignment transformation; not only conventional mutual information, but basically any similarity measure based on matched intensity pairs.

Also, I spent part of my PhD developing an alternative to mutual information for MR/US registration, arguing that conventional intensity-based similarity measures are not appropriate in this case due to the nature of US~images. This stresses the unsurprising fact that the choice of registration features is key. There is no such thing as a universal intensity-based similarity measure. They all have limited validity domains.

Why did I insist so much in the opposite direction? My thinking was that it was possible to derive image registration similarity measures from some extented likelihood principle. Consistency is a property of standard likelihood. So it was kind of natural to expect consistency to similarly follow from generalized likelihood.

What I overlooked is that the likelihood extension I was considering, basically VOID feature likelihood, stems from a semi-generative statistical model. In other words, it's a partial model that doesn't fully describe images. Now, can consistency follow from partial modeling assumptions? It would be a nice property, but isn't it too much to ask?

It seems it is.

\subsection{Getting VOID feature likelihood right}

At the end of the day, the VOID feature likelihood method is just a restriction of Baggenstoss's 2003 proposal. We say: hey, this is a cool method but you can't choose VOID features randomly.

However, there are other aspects that differ.

First, we can apply MaxEnt in a broader context: instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we can just assume some mean-value constraints and obtain the distributions via game theory. This is conceptually important as it can justify voxel independence assumptions.

Second, nuisance parameters are not correctly handled in Baggenstoss's work. He failed to see that the likelihood ratio involved in the PDF~projection fundamentally arises from a joint KL~minimization objective, hence it reflects a competition between two antagonist hypotheses. Parameter elimination should be consistent with that. This observation naturally leads to maximum likelihood ratio estimates rather than conventional likelihood parameter elimination.

\bibliographystyle{abbrv}
\bibliography{cvis,stat,alexis}

%%\input{draft.biblio}

\end{document}

### 2024-03-11 papers/composite_inference/notes2024.tex
Notes on composite inference

@@
 
-Whether composite Bayesian inference is the right way to look at naive Bayes methods is questionable. But the interpretation it provides seems to be useful.
+I do believe at this point that Bayesian inference is the right way to look at naive Bayes. At the very least, the interpretation it provides is insightful.
 
@@
 
+\subsection{Conservativeness of VOID feature likelihood}
+
+An interesting remark is that the expected ratio $\log p(t_\theta|\theta)/\pi(t_\theta)$ is lower than the true log-likelihood ratio $\log p(x|\theta)/\pi(x)$ by the general fact that the KL~divergence decreases under feature extraction.
+This remains true in the more general context of mean-value constraints as the epistemic distributions involved in the VOID feature likelihood then minimize the KL~divergence by design.
+From there, we can argue that VOID feature likelihood is conservative if used in a Bayesian framework.
+
 \subsection{Getting VOID feature likelihood right}
 
-At the end of the day, the VOID feature likelihood method is just a restriction of Baggenstoss's 2003 proposal. We say: hey, this is a cool method but you can't choose VOID features randomly.
+At the end of the day, the VOID feature likelihood method is just a restriction of Baggenstoss's 2003 proposal.
+We say: hey, this is a cool method but you can't choose VOID features randomly.
+However, there are two other aspects in which my approach significantly deviated from Baggenstoss's.
 
-However, there are other aspects that differ.
+First, the idea of applying MaxEnt in a broader context: instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we may just assume some mean-value constraints and derive epistemic distributions via game theory, which brings connections with minimally informative likelihood, minimum description length and that kind of cool theoretical stuff \cite{Grunwald-04,Grunwald-06}.
+My motivation for this was primarily to justify voxel independence assumptions.
+However, this may be one of the worst ideas I had.
+In this more general framework, it seems that the key weak consistency property would be gone -- think about it, do we want a theory that justifies cross-correlation for multimodal images?
+Moreover, we don't even need that to justify voxel independence for feature registration models: de Finetti's representation theorem does the job, and I had seen it as early as in 2005.
+No need to generalize Baggenstoss's approach here: one point for him.
 
-First, we can apply MaxEnt in a broader context: instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we can just assume some mean-value constraints and obtain the distributions via game theory.
-This is conceptually important as it can justify voxel independence assumptions.
-
-Second, nuisance parameters are not correctly handled in Baggenstoss's work.
+Second, the treatment of nuisance parameters.
 He failed to see that the likelihood ratio involved in the PDF~projection fundamentally arises from a joint KL~minimization objective, hence it reflects a competition between two antagonist hypotheses.
@@
 This observation naturally leads to maximum likelihood ratio estimates rather than conventional likelihood parameter elimination.
+Nuisance parameters are not correctly handled in Baggenstoss's work.That point is for me.
+I think.
 

### 2024-03-11 papers/composite_inference/notes2024.tex
Some corrections in notes 2024

--- 2024-03-04 papers/composite_inference/notes2024.tex
@@
 
-In the years 2015--2019, I worked on an attempt to put forward a non-conventional statistical inference paradigm, which I called composite Bayesian inference.
+In the years 2015--2019, I worked on an attempt to put forward a slightly new statistical inference paradigm, which I called composite Bayesian inference.
 
-Composite Bayesian inference is a new viewpoint on the ``naive Bayes'' method that justifies it without relying on implausible independence assumptions.
+Composite Bayesian inference is a new perspective on the ``naive Bayes'' method that justifies it without relying on implausible independence assumptions.
 It also comes with a posterior distribution, which is a scaled version of the ``naive Bayes'' posterior, enabling the computation of conservative credibility sets.
 
-Whether composite Bayesian inference is the right way to look at naive Bayes methods is questionable. But the interpretation it provides seems to be useful.
+I do believe at this point that Bayesian inference is the right way to look at naive Bayes. At the very least, the interpretation it provides is insightful.
 
-I tried to generalize it it using the concept of ``super composite likelihood'' function, where the weights assigned to the multiple atomic features are dependent on the variable to be inferred, {\em i.e.}, they are variable-of-interest dependent (VOID).
+However, the most general version of composite inference I presented goes well beyond the classical na\"ive Bayes scope, and is based on the concept of ``super composite likelihood'' function, where the weights assigned to the multiple atomic features are dependent on the variable of interest, {\em i.e.}, they are based on variable data representations.
 
-My ultimate goal here was to justify image registration similarity measures in a neat statistical inference framework.
+My hidden goal here was to justify image registration similarity measures in a neat statistical inference framework.
 
 I didn't do a great job at this.
-The super composite likelihood could more simply be derived as a regular composite likelihood using VOID features, without the need to introduce a more complex definition.
-The crux of the problem is to justify likelihood functions based on VOID features, but I didn't do that.
+The super composite likelihood could more simply be derived as a regular composite likelihood using variable features, without the need to introduce a more complex definition.
+The crux of the problem is to justify likelihood functions based on variable features, but I didn't do that.
 
-\subsection{VOID feature likelihood}
+\subsection{Variable feature likelihood}
 
-The idea of VOID feature-based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}.
-It is definitely a relevant topic, but this initial work was not fully successful and didn't seem to convince many.
+The idea of variable feature (VF) based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}.
+It is definitely a relevant idea, and perhaps an important one too, but this initial work was not fully successful.
 
-On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VOID feature-based likelihood function.
-As noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.
+On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VF-based likelihood function.
+As later noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.
 
-However, it is easy to see how the PDF~projection method can lead to completely absurd likelihood functions, if the VOID features are chosen arbitrarily.
-Baggenstoss probably saw that but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''.
+However, it is easy to see how the PDF~projection method can lead to heavily biased, hence completely absurd, likelihood functions, if the variable features are chosen arbitrarily.
+Fixed features can be chosen arbitrarily -- at worst, they will lead to a flat likelihood, but can't bias inference.
+The same cannot be said for variable features.
+
+Baggenstoss probably saw that, but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''.
 This is nonsense.
@@
 
-The least we can ask to the VOID feature likelihood is to be asymptotically consistent, that is, to be maximal at the true parameter value in the limit of large samples.
+The least we can ask to the variable feature likelihood is to be weakly consistent, that is, to be maximal at the true parameter value in the limit of large samples.
 If this property does not hold, then the method is worthwhile nothing.
 
-A sufficient condition to guarantee consistency is that the VOID feature function $t_\theta$ statisfies:
+A sufficient condition to guarantee weak consistency is that the variable feature function $t_\theta$ statisfies:
 $$
@@
 
-This condition may look complicated, but it only says that the VOID feature should always be the most informative about the variable of interest, among all feature candidates.
-In other words, $t_\theta$ should be, for any~$\theta$, the best statistic to test the hypothesis that~$\theta$ is the correct parameter vs the null hypothesis.
+This condition may look complicated, but it only says that the variable feature should always be the most informative about the variable of interest, among all feature candidates.
+Specifically, $t_\theta$ should be, for any~$\theta$, the best statistic to test the hypothesis that~$\theta$ is the correct parameter vs the null hypothesis.
+In plain and simple words, the variable data representation should not discard any critical evidence.
+
+Ideally, $t_\theta$ would be a sufficient statistic for each binary test $H_\theta$ vs $H_0$.
+Baggenstoss did note that, under sufficiency of $t_\theta$, the PDF~projection gives the true data distribution.
+But it is not a necessary condition for the method to work.
+And luckily so, otherwise the VF likelihood method would be pretty difficult to apply in practice.
 
 Let's take the same example as in my initial paper on composite inference.
-Assume we want to infer a disease category (infectious / cardiovascular) using VOID likelihood based on two simple clinical measures, body temperature and blood pressure.
-The above condition tells us that we need to associate body temperature with the infectious category, and blood pressure with the cardiovascular category.
-If we do it the other way round, the VOID feature likelihood will at best be useless, at worst biased.
+Assume we want to infer a disease category using VF likelihood based on basic clinical measures, say, body temperature and blood pressure.
+The above condition tells us that we need to associate body temperature with infectious diseases, and blood pressure with the cardiovascular category.
+If we do it the other way round, the VF likelihood will at best be useless, at worst biased.
 
@@
 
-What I overlooked is that the likelihood extension I was considering, basically VOID feature likelihood, stems from a semi-generative statistical model.
+What I overlooked is that the likelihood extension I was considering, basically VF likelihood, stems from a semi-generative statistical model.
 In other words, it's a partial model that doesn't fully describe images.
@@
 
-\subsection{Getting VOID feature likelihood right}
+\subsection{Conservativeness of VF likelihood}
 
-At the end of the day, the VOID feature likelihood method is just a restriction of Baggenstoss's 2003 proposal. We say: hey, this is a cool method but you can't choose VOID features randomly.
+An interesting remark is that the expected ratio $\log p(t_\theta|\theta)/\pi(t_\theta)$ is lower than the true log-likelihood ratio $\log p(x|\theta)/\pi(x)$ by the general fact that the KL~divergence decreases under feature extraction.
+From there, we can argue that VF likelihood is conservative if used in a Bayesian framework.
 
-However, there are other aspects that differ.
+\subsection{Getting VF likelihood right}
 
-First, we can apply MaxEnt in a broader context: instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we can just assume some mean-value constraints and obtain the distributions via game theory.
-This is conceptually important as it can justify voxel independence assumptions.
+At the end of the day, the VF likelihood method is just a restriction of Baggenstoss's 2003 proposal.
+We say: hey, this is a cool method but you can't choose variable features randomly.
+Yet there are two other aspects in which my approach significantly deviated from Baggenstoss's.
 
-Second, nuisance parameters are not correctly handled in Baggenstoss's work.
-He failed to see that the likelihood ratio involved in the PDF~projection fundamentally arises from a joint KL~minimization objective, hence it reflects a competition between two antagonist hypotheses.
+\paragraph{Applying MaxEnt in a broader context.} Instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we may just assume some mean-value constraints and derive epistemic distributions via KL~minimization, which brings connections with minimally informative likelihood, minimum description length, game theory and that kind of cool publishable stuff.
+
+My motivation for this was primarily to justify voxel independence assumptions.
+However, this may be one of the worst ideas I had.
+In this more general framework, the key weak consistency property is gone -- think about it, do we want a theory that justifies cross-correlation for multimodal images?
+Moreover, we don't even need that to justify voxel independence for feature registration models: de Finetti's representation theorem does the job, as I noticed a while back (in 2005 I think).
+No need to generalize Baggenstoss's approach here: one point for him.
+
+\paragraph{The treatment of nuisance parameters.} Baggenstoss failed to see that the likelihood ratio involved in the PDF~projection fundamentally arises from a joint KL~minimization objective, hence it reflects a competition between two hypotheses.
 Parameter elimination should be consistent with that.
 This observation naturally leads to maximum likelihood ratio estimates rather than conventional likelihood parameter elimination.
+Nuisance parameters are not correctly handled in Baggenstoss's work.
+That point is for me.
+I think.
 

### 2024-03-15 cv2/cv.tex
Update CV

@@
 \usepackage[left = 0mm, right = 0mm, top = 0mm, bottom = 0mm]{geometry}
-\usepackage[stretch = 25, shrink = 25]{microtype}  
+%\usepackage[stretch = 25, shrink = 25]{microtype}  
 \usepackage{graphicx}        % To insert pictures
@@
 \usepackage{FiraSans}        % Change this to use any font, but keep it simple
-\renewcommand{\familydefault}{\sfdefault}
+%\renewcommand{\familydefault}{\sfdefault}
 
@@
 
-% Centering without extra vertical spacing
 \null\hfill
@@
   \includegraphics[width=0.65\textwidth]{alexis_didimo_cropped.png}\\
-\caption{\scriptsize My (hairless) digital twin}
+  {\scriptsize My (hairless) digital twin.}
 \end{center}
@@
 \is
-{\sc Research Assistant.} {\it University of Oxford}, UK. \dates{Apr 2001--Oct 2002} \\
-\smaller{Developed advanced medical image registration algorithms and consulted for Mirada Solutions Ltd (now Siemens Molecular Imaging).}
+{\sc Research Assistant.} {\it University of Oxford}, \hphantom{CACACACA} United Kingdom. \dates{Apr 2001--Oct 2002} \\
+\smaller{Developed medical image registration algorithms at the Wolfson Medical Vision Laboratory and consulted for Mirada Solutions Ltd.}
 

### 2024-03-15 papers/composite_inference/notes2024.tex
Update CV

@@
 
-An interesting remark is that the expected ratio $\log p(t_\theta|\theta)/\pi(t_\theta)$ is lower than the true log-likelihood ratio $\log p(x|\theta)/\pi(x)$ by the general fact that the KL~divergence decreases under feature extraction.
+A key remark is that the expected logarithm ratio $\log p(t_\theta|\theta)/\pi(t_\theta)$ is lower than the true log-likelihood ratio $\log p(x|\theta)/\pi(x)$ by the general fact that the KL~divergence decreases under feature extraction.
+It is also a direct consequence from the fact that VF likelihood minimizes the KL~divergence among all distributions $p(x)$ compatible with $p(t_\theta|\theta)$.
+
+There is not much we can make from this result alone. But if, by any chance, the reference distribution is chosen so that there exists a $\theta_0$ for which $\pi(x)=p(x|\theta_0)$, then...
+
+(the expectation being taken under the {\em true} unknown distribution)
+
 From there, we can argue that VF likelihood is conservative if used in a Bayesian framework.

### 2024-05-03 papers/composite_inference/notes2024.tex
More notes on VS

@@
 \def\y{{\mathbf{y}}}
+\def\z{{\mathbf{z}}}
 \newcommand{\blambda}{{\boldsymbol{\lambda}}}
@@
 
-The idea of variable feature (VF) based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}.
-It is definitely a relevant idea, and perhaps an important one too, but this initial work was not fully successful.
+The idea of variable feature (VF) likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03} under the not very explicit name of ``class-specific method''.
+It is definitely a relevant idea, but this initial work was not fully successful for several reasons.
 
 On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VF-based likelihood function.
-As later noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.
+As later noted by \cite{Minka-04}, it is just an application of the maximum entropy or, more generally, the minimum KL~divergence principle.
 
-However, it is easy to see how the PDF~projection method can lead to heavily biased, hence completely absurd, likelihood functions, if the variable features are chosen arbitrarily.
-Fixed features can be chosen arbitrarily -- at worst, they will lead to a flat likelihood, but can't bias inference.
+However, there are three main flaws in Baggenstoss's presentation in my humble (or not so much) opinion:
+\begin{enumerate}
+\item {\em Bias}.
+It is easy to see how the PDF~projection method can lead to heavily biased likelihood functions if the variable features are chosen arbitrarily.
+The truth is that fixed features can be chosen arbitrarily -- at worst, they will lead to a flat likelihood, but can't bias inference.
 The same cannot be said for variable features.
+Baggenstoss may have seen that, but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''.
+Still, the assumptions under which a particular choice of features works or not should be made explicit.
+\item {\em Arbitrary reference hypothesis}.
+The PDF~projection theorem involves a reference, or null hypothesis distribution that Baggenstoss gives no guideline whatsoever on how to choose.
+Since the likelihood function that arises from PDF~projection clearly depends on the reference distribution, this choice matters.
+He tried to work around this issue by making the reference distribution parametric, which led him to the third flaw:
+\item {\em Inconsistent parameter elimination}.
+Baggenstoss treated the VF likelihood exactly as if it was a true likelihood function, and implicitely proposed to eliminate the nuisance parameters by maximization.
+The deep information-theoretic interpretation of PDF~projection shows that it is wrong.
+It is easy to see that Baggenstoss's approach doesn't work in the simplest instance of VF likelihood where the parameter space is a singleton -- the method should then be equivalent to the well-known maximum likelihood ratio method for binary hypothesis testing.
+This suggests that the nuisance parameters pertaining to the null hypothesis cannot be just mixed with those of the feature distribution; instead, they should be considered as competitive parameter sets.
+\end{enumerate}
 
-Baggenstoss probably saw that, but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''.
-This is nonsense.
-The choice of features should be driven by explicit assumptions.
+\subsection{Solving for the bias: the missing constraint}
 
-The PDF~projection method simply needs an additional constraint to work. I have known the solution for quite some time but have struggled to fully contemplate it.
-
-\subsection{The missing constraint}
+The PDF~projection method simply needs an additional assumption to be unbiased. I have known the solution for quite some time but have struggled to fully contemplate it.
 
@@
 
+\appendix
+
+\section{Baggenstoss 2003 paper: reading notes}
+
+He introduces the class-specific approach by a curse of dimensionality argument:
+\begin{quote}
+Because each feature set can be designed without regard to other classes (or states), it can be of much lower dimension than a common feature set that must account for all classes, effectively avoiding the curse of dimensionality.
+\end{quote}
+
+This is spot on: we want to select features to avoid the curse of dimensionality, however class-independent feature selection may not give us enough flexibility. He further comments on that:
+\begin{quote}
+[Feature selection] leads to a fundamental tradeoff: whether to discard features in an attempt to reduce the dimension to something manageable or to include them and suffer the problems associated with estimating a PDF at high dimension.
+\end{quote}
+
+He mentions that he did publish a previous work on the topic in 1999, which was based on sufficient statistics and used a fixed reference hypothesis.
+
+His exposition of the PDF~projection theorem starts with the remark that, if $X$ is a random vector and $Z = f(X)$ is a one-to-one function of~$X$, then:
+$$
+p_x(\x)
+=
+|J_f(\x)| p_z(\z)
+.
+$$
+
+In the more general case of a many-to-one function, there are multiple distributions~$p_x(\x)$ that are compatible with a given~$p_z(\z)$.
+To solve for this ambiguity, Baggenstoss postulates that a reference distribution is needed:
+\begin{quote}
+In order to apply the constraint, it is necessary to make use of a reference hypothesis $H_0$ for which we know the PDF of both $\x$ and $\z$.
+\end{quote}
+
+In his derivation of the PDF~projection theorem, he postulates the
+
 %%\input{draft.biblio}

### 2024-05-03 papers/variana/2024.tex
More notes on VS

--- 2024-02-06 papers/variana/notes.tex
@@
 
-\title{Variana Reloaded}
+\title{Variational sampling reloaded}
 \author{Alexis Roche}
@@
 
+\section{Introduction}
+
+Variational sampling is a simple and general method for computational statistics that extends in a rather natural way the basic importance sampling method, aka Monte Carlo integration, which dates back (at least) to World War II.
+
+I discovered it in 2010 while trying to interpret importance sampling in a variational framework, and coined the name ``variational sampling'' as a result of that.
+Despite its relative simplicity, it did not seem to be known under any name in the literature, which puzzled me back then, and keeps puzzling me.
+
+There could only be two reasons for that. Either it was a bad idea, or researchers had overlooked it. I was crazy enough to believe in the latter explanation.
+
+Hence, I tried hard to publish it in NIPS, but failed four times in as many attempts from 2010 to 2016.
+I did publish it on arXiv in May 2011 on the advice of Pr Nanny Wermuth, who had come across my work and encouraged me to push it forward, and later in a low cost e-journal nicely called {\em International Journal of Advanced Statistics and Probability} in a desperate, yet successful, attempt to make at least one peer-reviewed publication on the topic.
+
+As of May 2024, these papers have a total of 4 citations in Google Scholar, including 2 self-citations.
+
+Variational sampling is, unquestionably, the worst failure of my whole researcher career in terms of scientific impact, that being said while said career operates on reasonably modest standards.
+
+%%%, but much simpler than many advanced sampling and approximate inference methods found in the literature.
+%thought it should be part of any standard computational statistics toolbox
+
 \section{History}
 
-It is time to assemble the pieces of the puzzle.
+How the pieces of the puzzle came together.
 \begin{itemize}
@@
 \item 2014--16: realization that VS is useful in practice combined with Expectation Propagation.
-\item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion).
+\item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion); unveiling connection between distribution scaling and $\alpha$-divergence minimization.
 \item 2022: proven bound on VS asymptotic variance.
-\item 2023: final key insight: VS makes optimization possible in high dimension (at least locally) because it avoids saddle points unlike gradient-based methods.
+\item 2023: revisited custom optimizer.
 \end{itemize}
-
-\section{FAQ}
-
-\paragraph{Is VS better than IS?} No.
-VS is asymptotically equivalent to IS applied to the residual Gaussian fitting error, {\em i.e.}, the difference between the target distribution and its KL optimal Gaussian fit.
-VS will thus be more accurate than IS whenever the residual importance weights vary less than the base importance weights in the sampling distribution support.
-This happens when the target is locally well approximated by a Gaussian.
-But this is not always the case.
-For instance, assume we want to estimate the normalizing constant of a distribution that we are able to sample from; in this case, the base importance weights are constant, and substracting the Gaussian fit doesn't help.
-VS doesn't even converge if the target has shorter tails than a Gaussian, while IS is exact no matter what.
-Generally speaking, VS doesn't appear as the method of choice when we have a good sampler.
-
-%\paragraph{What does ``nearly Gaussian'' mean?} Roughly speaking, it means that the distribution can be decomposed as the product of a Gaussian distribution and a slowly varying function within the Gaussian distribution's support.
-
-\paragraph{Is VS better than Laplace?} The Laplace method has systematic bias, unlike VS.
-This is fundamentally because it works at fixed, infinitesimal scale, while VS works at adjustable scale.
-One can think of simple situations where Laplace fails miserably, such as fitting a Laplace distribution (isn't that ironical?), but it may also work very well provided that the target is nearly Gaussian.
-The problem is that it is impractical in high dimension since it requires maximizing the distribution (which is more or less the problem we are trying to solve using VS).
-In EP context, we can perform a second-order Taylor expansion of the log-factor at the cavity center, yielding a damped Newton iteration.
-The resulting approximate EP is has a tendency to get stuck at saddle points, just like most gradient-based methods.
-VS-based EP is able to escape saddle points by exploring the target distribution in neighborhoods of increasing sizes.
-
-\paragraph{Can I choose any sampling kernel in practice?} Unfortunately, no.
 
@@
 
+Note, however, that IS can also be seen as a variational sampling instance corresponding to another surrogate criterion:
+$$
+\tilde{D}_{IS}(\theta)
+= - \frac{1}{n} \sum_i f(x_i) \log g_\theta(x_i) 
+  + \int \pi g_\theta
+.
+$$
+
+The key difference with $\tilde{D}(\theta)$ is that the sample mean of $g_\theta(x_i)$ is replaced by its asymptotical value.
+Hence, $\tilde{D}_{IS}$ may be seen as a ``partial Rao-Blackwellization'' of $\tilde{D}$, partial because we can't do it for the term in $\log g_\theta(x_i)$; and, because it's partial, it may not improve the estimator accuracy (our main claim is that it doesn't in many situations of practical interest).
+
 \section{Variance upper bound}
@@
 $$
-implying that the loss is not only convex in~$g$ but also staisfies the conditions of Alexis's theorem with $a=f$. Hence,
+implying that the loss is not only convex in~$g$ but also satisfies the conditions of Alexis's theorem with $a=f$. Hence,
 $$
@@
 Integrating this over~$\pi$, we get this long-waited inequality:
-$$
+\begin{equation}
+\label{eq:variance_inequality}
 \int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g),
-$$
+\end{equation}
 where the infinite norms are taken on the support of~$\pi$, which is infinite if $\pi$ is a Gaussian kernel.
@@
 
+Substituing~$g$ with the minimizer~$g_\star$ of $D(\pi f\|\pi g)$ in the inequality~(\ref{eq:variance_inequality}), we get an upper bound for the VS 0-order moment estimate asymptotic variance:
+$$
+\int \pi (f-g_\star)^2
+\leq
+2\max(\|f\|_\infty, \|g_\star\|_\infty)
+D(\pi f\|\pi g_\star)
+.
+$$
+
+Obviously, a sufficient condition for the left-hand side to be finite is that the right-hand side be finite.
+
+We note that, if $\|f\|_\infty$ is finite, then $D(\pi f\|\pi g_\star)$ is finite.
+The reason is that, by construction, $g_\star$ minimizes the KL~divergence on a function space that contains the constant function $g\equiv 1$, therefore:
+$$
+D(\pi f\|\pi g_\star)
+\leq
+D(\pi f\|\pi 1)
+=
+\int \pi (f\log f - f + 1)
+.
+$$
+
+Now, the function $x\log x - x + 1$ is positive-valued on $\mathbb{R}_+$, decreasing from 1 to 0 between 0 and 1, and increasing to infinity from 1.
+It follows that $(f\log f - f + 1)\leq \max(1, \|f\|_\infty \log \|f\|_\infty - \|f\|_\infty + 1)$, hence the KL~divergence term is finite since it is upper bounded by a finite constant:
+$$
+D(\pi f\|\pi g_\star) \leq
+\max(1, \|f\|_\infty \log \|f\|_\infty - \|f\|_\infty + 1)
+.
+$$
+
+Is the condition $\|f\|_\infty$ finite sufficient?
+
+For the IS estimate, it is -- because the asymptotic variance is related to the variance of $f(x)$ under $\pi(x)$:
+$$
+{\rm Var}[f(X)]
+= \min_{c\in\mathbb{R}} \int \pi (f - c)^2,
+$$
+which is easily seen to be finite whenever $f$ is upper bounded.
+
+For VS, however, we need $\|g_\star\|_\infty$ to be finite too.
+Intuitively, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be much smaller than wrt a constant baseline, so we expect $\int \pi(f-g_\star)^2$ to be quite smaller than ${\rm Var}[f(X)]$.
+We can make this a little bit more precise using :
+$$
+\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
+$$
+
+However, $g_\star$ does not minimize $\pi (f-g)^2$, it minimizes the KL divergence $D(\pi f\|\pi g)$, therefore the above reasoning doesn't work.
+
+So let's go back to the KL~minimization.
+One thing we can note is that the variances of $\pi f$ and $\pi g_\star$ are the same owing to the moment matching property of KL~minimization, and so are the precisions:
+$$
+\prec(\pi g_\star) = \prec(\pi f).
+$$
+
+Now, both $\pi$ and $g_\star$ are log-quadratic functions so the precisions add up:
+$$
+\prec(\pi g_\star) = \prec(\pi) + \prec(g_\star).
+$$
+
+Clearly, $\|g_\star\|_\infty$ is finite iff $\prec(g_\star)$ is positive (not necessarily positive definite). This is the case iff: 
+$$
+\prec(\pi f) \succeq \prec(\pi)
+.
+$$
+
+In other words, sufficient conditions for VS to converge are:
+\begin{enumerate}
+\item The factor $f$ is upper bounded.
+\item The ``posterior'' variance is smaller than the ``prior'' variance.
+\end{enumerate}
+
+Is the second condition implied by the first? No, it isn't.
+
+\paragraph{Damning counter-example.}
+
+Here is a simple one-dimensional counter-example where $\|f\|_\infty$ is finite and $\|g_\star\|_\infty$ is not.
+Consider a normalized Gaussian reference $\pi = N(0,1)$, and a uniform factor defined by $f = \chi(-a,a)/\pi$, where $a$ is a positive constant, so that the target distribution $\pi f$ is the uniform distribution $\chi(-a,a)$ and its variance is therefore $a^2/ 3$.
+Following the same reasoning as above, we see that the precision of the fitted factor is:
+$$
+\prec(g_\star) = \frac{3}{a^2} - 1,
+$$
+which, for $a$ sufficiently large, namely $a>\sqrt{3}$, is strictly negative, implying that $g_\star$ is not upper bounded.
+Even worse than that, $g_\star$ goes very fast to infinity as $x$ departs from $0$, so $\int \pi g_\star^2$ is infinite, implying that $\int \pi (f-g_\star)^2$ is also infinite, implying that VS does not converge!
+
+In this case, the optimal fit $g_\star$ goes to infinity while $f$ vanishes.
+This is weird.
+But it happens because, on the one hand, the factor looks like a U-shaped function in the context defined by $\pi$, and, on the other hands, the loss incurred in regions where the factor vanihes, $\ell(0,g)=g$, is only linear in~$g$, so it's ``mild'' even for large values of~$g$ when weighted by~$\pi$.
+This is consistent with the fact that the KL~divergence is an inclusive  measure (``zero avoiding'' for $g_\star$).
+In contrast, minimizing $\pi(f-g)^2$ would yield a well-behaved (bell-shaped) factor approximation -- to see this, just realize that $\inf_g\int\pi(f-g)^2 \leq \int\pi f^2$, which is finite.
+
+\paragraph{Wrapping up.}
+
+We learn several things from the above counter-example.
+
+First off, VS is not necessarily better than IS (it may not converge at all while IS converges).
+
+Second, the intuition that $g_\star$ follows $f$ closely is not necessarily correct.
+
+Third, this intuition may not be correct even though the factor is ``almost'' perfectly log-quadratic on the domain defined by $\pi$.
+In the counter-example, VS will do a good job in practice (much better than IS).
+Why?
+Just think about it: because on a finite sample drawn from $\pi$, chances are that we don't get any point outside $[-a,a]$, where $f$ is perfectly log-quadratic.
+
+%For instance, drawing 100000 points from a normal distribution, we have almost no chance to have any point more than 5 times standard deviations away from the mean.
+With 1000000 points, the same goes for 6 standard deviations.
+
+Hence, and that's a 4th take-away, the asymptotic variance does not tell us the whole truth about how VS performs in practice.
+The sample size required to reach asymptotic regime may heavily depend on the factor at hand.
+Besides, the fact that VS may actually worsen by increasing the sample size is both unexpected and concerning.
+
+As already pointed out, truncating the reference distribution doesn't really solve the problem.
+The asymptotic variance would become finite, but still possibly way too big.
+In the counter-example, how would we know beforehand that we should truncate outside $[-a,a]$?
+
+So, how can we work around this problem?
+
+%we can put the blame on both the factor for being far off the approximating family and the context $\pi$ as for being too narrow to ``see'' $f$ globally. But we may not know that beforehand.
+
+%Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
+
+\paragraph{Bullshite.}
+
 From this point, can we give practical sufficient conditions for VS convergence?
@@
 Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.
-
-\section{VS vs IS}
-
-The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$.
-This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor.
-The IS estimator is just a sample mean:
-$$
-\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
-\qquad x_i \stackrel{iid}{\sim} \pi(x),
-$$
-which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
-$$
-\mathbb{E}[f(X)] = \int\pi f,
-$$
-we also have that:
-$$
-{\rm Var}[f(X)]
-= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
-$$
-
-Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.
-
-As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$.
-Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.
-
-Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
-Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''.
-We can make this a little bit more precise using :
-$$
-\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
-$$
 
@@
 
+\section{FAQ}
+
+\paragraph{Is VS better than IS?} No.
+VS is asymptotically equivalent to IS applied to the residual Gaussian fitting error, {\em i.e.}, the difference between the target distribution and its KL optimal Gaussian fit.
+VS will thus be more accurate than IS whenever the residual importance weights vary less than the base importance weights in the sampling distribution support.
+This happens when the target is locally well approximated by a Gaussian.
+But this is not always the case.
+For instance, assume we want to estimate the normalizing constant of a distribution that we are able to sample from; in this case, the base importance weights are constant, and substracting the Gaussian fit doesn't help.
+VS doesn't even converge if the target has shorter tails than a Gaussian, while IS is exact no matter what.
+Generally speaking, VS doesn't appear as the method of choice when we have a good sampler.
+
+%\paragraph{What does ``nearly Gaussian'' mean?} Roughly speaking, it means that the distribution can be decomposed as the product of a Gaussian distribution and a slowly varying function within the Gaussian distribution's support.
+
+\paragraph{Is VS better than Laplace?} The Laplace method has systematic bias, unlike VS.
+This is fundamentally because it works at fixed, infinitesimal scale, while VS works at adjustable scale.
+One can think of simple situations where Laplace fails miserably, such as fitting a Laplace distribution (isn't that ironical?), but it may also work very well provided that the target is nearly Gaussian.
+The problem is that it is impractical in high dimension since it requires maximizing the distribution (which is more or less the problem we are trying to solve using VS).
+In EP context, we can perform a second-order Taylor expansion of the log-factor at the cavity center, yielding a damped Newton iteration.
+The resulting approximate EP is has a tendency to get stuck at saddle points, just like most gradient-based methods.
+VS-based EP is able to escape saddle points by exploring the target distribution in neighborhoods of increasing sizes.
+
+\paragraph{Can I choose any sampling kernel in practice?} Unfortunately, no.
+
+\section{VS vs IS}
+
+The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$.
+This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor.
+The IS estimator is just a sample mean:
+$$
+\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
+\qquad x_i \stackrel{iid}{\sim} \pi(x),
+$$
+which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
+$$
+\mathbb{E}[f(X)] = \int\pi f,
+$$
+we also have that:
+$$
+{\rm Var}[f(X)]
+= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
+$$
+
+Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.
+
+As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$.
+Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.
+
+Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
+Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''.
+We can make this a little bit more precise using :
+$$
+\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
+$$
+
 \section{Probability updating}

### 2024-08-16 cv2/cv.tex
CV updates

@@
 \newcommand{\is}{\par\vskip.5ex plus .4ex} % Item spacing
-\newcommand{\smaller}[1]{{\small$\diamond$\ #1}}
+\newcommand{\smaller}[1]{{\small #1}} %{{\small$\diamond$\ #1}}
 \newcommand{\headleft}[1]{\vspace*{3ex}\textsc{\textbf{#1}}\par%
@@
 \is
-{\sc Research Assistant.} {\it University of Oxford}, \hphantom{CACACACA} United Kingdom. \dates{Apr 2001--Oct 2002} \\
+{\sc Post-doctoral research Assistant.} {\it University of Oxford}, United Kingdom. \dates{Apr 2001--Oct 2002} \\
 \smaller{Developed medical image registration algorithms at the Wolfson Medical Vision Laboratory and consulted for Mirada Solutions Ltd.}
@@
 \is 
-{\sc Collaborative programming:} C, C++, Python (including 18 years experience within the NumPy/SciPy ecosystem), Matlab, R, and using Git for version control.
+{\sc Collaborative programming:} Python (have used NumPy/SciPy since 2006), C, C++, Matlab, R, and using Git for version control.
 

### 2024-08-16 papers/variana/2024.tex
CV updates

@@
 \def\B{\mathbf{B}}
-\def\prec{\mathrm{prec}}
+\def\prec{\mathrm{Prec}}
+\def\var{\mathrm{Var}}
 
@@
 \item 2011--13: consolidation of the theory behind VS.
-\item 2014--16: realization that VS is useful in practice combined with Expectation Propagation.
+\item 2014--16: realization that VS is useful combined with Expectation Propagation.
 \item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion); unveiling connection between distribution scaling and $\alpha$-divergence minimization.
 \item 2022: proven bound on VS asymptotic variance.
-\item 2023: revisited custom optimizer.
+\item 2023: key insight: VS leverages more information from a sample than stochastic gradient descent on KL~divergence.
 \end{itemize}
@@
 
+The gradient and Hessian of the KL~proxy are given by:
+$$
+\nabla \tilde{D} = \frac{1}{n}\sum_i [g_\theta(x_i)-f(x_i)]\phi(x_i),
+\qquad
+\nabla \nabla^T \tilde{D} = \frac{1}{n}\sum_i g_\theta(x_i)\phi(x_i)\phi(x_i)^\top
+,
+$$
+which can be interpreted as IS estimates.
+
 We have proved in our initial paper from 2013 that the resulting estimator of the integral, or moment, $I(\phi) \equiv \int \pi f \phi$, is asymptotically unbiased with variance matrix:
@@
 Hence, $\tilde{D}_{IS}$ may be seen as a ``partial Rao-Blackwellization'' of $\tilde{D}$, partial because we can't do it for the term in $\log g_\theta(x_i)$; and, because it's partial, it may not improve the estimator accuracy (our main claim is that it doesn't in many situations of practical interest).
+
+The gradient and Hessian of this alternative KL~proxy are seen to be:
+$$
+\nabla \tilde{D}_{IS} = \int \pi g_\theta \phi - \frac{1}{n}\sum_i f(x_i)\phi(x_i),
+\qquad
+\nabla \nabla^T \tilde{D}_{IS} = \int \pi g_\theta \phi \phi^\top
+.
+$$
+
+\section{Stochastic convergence}
+
+We have seen that the VS moment estimate converges in distribution to the true moment if the following matrix is finite:
+$$
+\Sigma = \int \pi (f - g_\star)^2 \phi \phi^\top
+.
+$$
+
+Can we give simple conditions for it to be the case?
+
+Let us focus for now on the first diagonal element:
+$$
+\sigma_0^2 = \int \pi (f - g_\star)^2.
+$$
+
+We note that $(f-g_\star)^2 = f^2 - 2fg_\star + g_\star^2\leq f^2 + g_\star^2$.
+So, if both $\int \pi f^2$ and $\int \pi g_\star^2$ exist, then $\sigma_0^2$ is finite.
+Both conditions depend on the distribution tails of $f$ and $g_\star$, respectively.
+
+In practice, we might be able to ascertain that the target $f$ is bounded, so that $\|f\|_\infty$ is finite and $\int \pi f^2$ is therefore finite.
+
+What about $g_\star$? Let's go back to the KL~minimization. The variances of $\pi f$ and $\pi g_\star$ are the same owing to the moment matching property of KL~minimization, and so are the precisions:
+$$
+\prec(\pi g_\star) = \prec(\pi f).
+$$
+
+Now, both $\pi$ and $g_\star$ are log-quadratic functions so the precisions satisfy some algebra rules:
+$$
+\prec(\pi g_\star^2)
+= 2 \prec(\pi g_\star) - \prec(\pi)
+= 2 \prec(\pi f) - \prec(\pi).
+$$
+
+Clearly, $\int \pi g_\star^2$ is finite iff $\prec(\pi g_\star^2)$ is positive (not necessarily positive definite), which brings the condition:
+$$
+\prec(\pi f) \succeq \frac{1}{2}\prec(\pi)
+.
+$$
+
+In other words, the ``posterior'' variance should not be more than twice as large as the ``prior'' variance:
+$$
+\var(\pi f) \preceq 2 \var(\pi)
+.
+$$
+
+It feels like a pretty mild condition, but it's not implied by the condition that $f$ is bounded.
+
+\subsection{Damning counter-example}
+
+Here is a simple one-dimensional counter-example where $\|f\|_\infty$ is finite and $\|g_\star\|_\infty$ is not.
+Consider a normalized Gaussian reference $\pi = N(0,1)$, and a uniform factor defined by $f = \chi(-a,a)/\pi$, where $a$ is a positive constant, so that the target distribution $\pi f$ is the uniform distribution $\chi(-a,a)$ and its variance is therefore $a^2/ 3$.
+Following the same reasoning as above, we see that the precision of the fitted factor is:
+$$
+\prec(g_\star) = \frac{3}{a^2} - 1,
+$$
+which, for $a$ sufficiently large, namely $a>\sqrt{3}$, is strictly negative, implying that $g_\star$ is not upper bounded.
+Even worse than that, $g_\star$ goes very fast to infinity as $x$ departs from $0$, so $\int \pi g_\star^2$ is infinite, implying that $\int \pi (f-g_\star)^2$ is also infinite, implying that VS does not converge!
+
+In this case, the optimal fit $g_\star$ goes to infinity while $f$ vanishes.
+This is weird.
+But it happens because, on the one hand, the factor looks like a U-shaped function in the context defined by $\pi$, and, on the other hand, the loss incurred in regions where the factor vanihes, $\ell(0,g)=g$, is only linear in~$g$, so it's ``mild'' even for large values of~$g$ when weighted by~$\pi$.
+This is consistent with the fact that the KL~divergence is an inclusive  measure (``zero avoiding'' for $g_\star$).
+In contrast, minimizing $\pi(f-g)^2$ would yield a well-behaved (bell-shaped) factor approximation -- to see this, just realize that $\inf_g\int\pi(f-g)^2 \leq \int\pi f^2$, which is finite.
+
+\subsection{Sufficient conditions for VS convergence}
+
+In summary, sufficient conditions for $\sigma_0^2$ to be finite are:
+\begin{enumerate}
+\item The factor $f$ is upper bounded.
+\item The ``posterior'' variance is smaller than twice the ``prior'' variance.
+\end{enumerate}
+
+It is easy to see that these conditions imply that each element of $\Sigma$ is finite, so they are sufficient for VS convergence.
+The key reason is that $\pi g_\star^2$ is a Gaussian distribution if condition~2 holds (as opposed to a U-shaped log-quadratic function), so $\int \pi g_\star^2 \phi \phi^\top$ is finite.
+
+The condition on the posterior variance feels mild because VS is meant to process Bayesian posteriors that may be sharply peaked around their mode.
+But such situation might be questionable in very large dimension, where the variance may be flat in some directions.
+Besides, it is difficult to test beforehand (the posterior variance is what we aim to estimate).
+
+As already pointed out, truncating the reference distribution wouldn't really solve the problem.
+The asymptotic variance would become finite, but still possibly way too big.
+In the counter-example, how would we know beforehand that we should truncate outside $[-a,a]$?
+
+\section{VS vs IS}
+
+The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$.
+This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor.
+The IS estimator is just a sample mean:
+$$
+\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
+\qquad x_i \stackrel{iid}{\sim} \pi(x),
+$$
+which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
+$$
+\mathbb{E}[f(X)] = \int\pi f,
+$$
+we also have that:
+$$
+\var[f(X)]
+= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
+$$
+
+Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on.
+
+\subsection{Asymptotic variance}
+
+Quite clearly, the condition that $\|f\|_\infty$ is finite is sufficient to guarantee that $\var[f(X)]$ is finite, and similarly for $\var[f(X)\phi(X)]$ for all $\phi$, which guarantees in turn the convergence of IS in the context of sampling from a Gaussian distribution $\pi$.
+
+As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$.
+Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.
+
+For VS, however, we have an extra condition on $g_\star$.
+Intuitively, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be smaller than wrt a constant baseline, so we expect $\int \pi(f-g_\star)^2$ to be smaller than $\var[f(X)]$.
+We could try and make this a little bit more precise using the following argument:
+$$
+\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = \var[f(X)].
+$$
+
+However, $g_\star$ does not minimize $\pi (f-g)^2$, it minimizes the KL divergence $D(\pi f\|\pi g)$, therefore the above reasoning doesn't work, as shown by the above counter-example.
+
+We learn several things from that.
+\begin{enumerate}
+\item VS may not converge while IS converges, even very slowly.
+\item Therefore, VS is not necessarily better than IS.
+\item The intuition that $g_\star$ follows $f$ closely is not always correct. 
+\item This intuition may not be correct even though the factor is ``almost'' perfectly log-quadratic on the domain covered by $\pi$.
+In the counter-example, VS will actually be much more stable than IS in practice.
+Why?
+Just think about it: because on a finite sample drawn from $\pi$, chances are that we don't get any point outside $[-a,a]$, where $f$ is perfectly log-quadratic.
+\item The sample size required to reach asymptotic regime may heavily depend on the factor at hand. 
+\end{enumerate}
 
@@
 
-\subsection{Sufficient conditions for VS convergence}
+\subsection{Convergence analysis}
+
+Does this upper bound help undestanding when VS converges?
 
@@
 
-Is the condition $\|f\|_\infty$ finite sufficient?
-
-For the IS estimate, it is -- because the asymptotic variance is related to the variance of $f(x)$ under $\pi(x)$:
-$$
-{\rm Var}[f(X)]
-= \min_{c\in\mathbb{R}} \int \pi (f - c)^2,
-$$
-which is easily seen to be finite whenever $f$ is upper bounded.
-
-For VS, however, we need $\|g_\star\|_\infty$ to be finite too.
-Intuitively, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be much smaller than wrt a constant baseline, so we expect $\int \pi(f-g_\star)^2$ to be quite smaller than ${\rm Var}[f(X)]$.
-We can make this a little bit more precise using :
-$$
-\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
-$$
-
-However, $g_\star$ does not minimize $\pi (f-g)^2$, it minimizes the KL divergence $D(\pi f\|\pi g)$, therefore the above reasoning doesn't work.
-
-So let's go back to the KL~minimization.
-One thing we can note is that the variances of $\pi f$ and $\pi g_\star$ are the same owing to the moment matching property of KL~minimization, and so are the precisions:
-$$
-\prec(\pi g_\star) = \prec(\pi f).
-$$
-
-Now, both $\pi$ and $g_\star$ are log-quadratic functions so the precisions add up:
-$$
-\prec(\pi g_\star) = \prec(\pi) + \prec(g_\star).
-$$
-
-Clearly, $\|g_\star\|_\infty$ is finite iff $\prec(g_\star)$ is positive (not necessarily positive definite). This is the case iff: 
-$$
-\prec(\pi f) \succeq \prec(\pi)
+At this point, we see that a sufficient condition for VS to converge is that both $f$ and $g_\star$ are upper bounded.
+This condition on $g_\star$ is equivalent to assuming that $g_\star$ is bell-shaped, that is, $\prec(g_\star)\succeq 0$.
+Because $\prec(\pi f)=\prec(\pi g_\star)=\prec(\pi) + \prec(g_\star)$, this boils down to the assumption:
+$$
+\prec(\pi f) \succeq \prec(\pi),
+$$
+or,
+$$
+\var(\pi f) \preceq \var(\pi)
 .
@@
 
-In other words, sufficient conditions for VS to converge are:
-\begin{enumerate}
-\item The factor $f$ is upper bounded.
-\item The ``posterior'' variance is smaller than the ``prior'' variance.
-\end{enumerate}
-
-Is the second condition implied by the first? No, it isn't.
-
-\paragraph{Damning counter-example.}
-
-Here is a simple one-dimensional counter-example where $\|f\|_\infty$ is finite and $\|g_\star\|_\infty$ is not.
-Consider a normalized Gaussian reference $\pi = N(0,1)$, and a uniform factor defined by $f = \chi(-a,a)/\pi$, where $a$ is a positive constant, so that the target distribution $\pi f$ is the uniform distribution $\chi(-a,a)$ and its variance is therefore $a^2/ 3$.
-Following the same reasoning as above, we see that the precision of the fitted factor is:
-$$
-\prec(g_\star) = \frac{3}{a^2} - 1,
-$$
-which, for $a$ sufficiently large, namely $a>\sqrt{3}$, is strictly negative, implying that $g_\star$ is not upper bounded.
-Even worse than that, $g_\star$ goes very fast to infinity as $x$ departs from $0$, so $\int \pi g_\star^2$ is infinite, implying that $\int \pi (f-g_\star)^2$ is also infinite, implying that VS does not converge!
-
-In this case, the optimal fit $g_\star$ goes to infinity while $f$ vanishes.
-This is weird.
-But it happens because, on the one hand, the factor looks like a U-shaped function in the context defined by $\pi$, and, on the other hands, the loss incurred in regions where the factor vanihes, $\ell(0,g)=g$, is only linear in~$g$, so it's ``mild'' even for large values of~$g$ when weighted by~$\pi$.
-This is consistent with the fact that the KL~divergence is an inclusive  measure (``zero avoiding'' for $g_\star$).
-In contrast, minimizing $\pi(f-g)^2$ would yield a well-behaved (bell-shaped) factor approximation -- to see this, just realize that $\inf_g\int\pi(f-g)^2 \leq \int\pi f^2$, which is finite.
-
-\paragraph{Wrapping up.}
-
-We learn several things from the above counter-example.
-
-First off, VS is not necessarily better than IS (it may not converge at all while IS converges).
-
-Second, the intuition that $g_\star$ follows $f$ closely is not necessarily correct.
-
-Third, this intuition may not be correct even though the factor is ``almost'' perfectly log-quadratic on the domain defined by $\pi$.
-In the counter-example, VS will do a good job in practice (much better than IS).
-Why?
-Just think about it: because on a finite sample drawn from $\pi$, chances are that we don't get any point outside $[-a,a]$, where $f$ is perfectly log-quadratic.
-
-%For instance, drawing 100000 points from a normal distribution, we have almost no chance to have any point more than 5 times standard deviations away from the mean.
-With 1000000 points, the same goes for 6 standard deviations.
-
-Hence, and that's a 4th take-away, the asymptotic variance does not tell us the whole truth about how VS performs in practice.
-The sample size required to reach asymptotic regime may heavily depend on the factor at hand.
-Besides, the fact that VS may actually worsen by increasing the sample size is both unexpected and concerning.
-
-As already pointed out, truncating the reference distribution doesn't really solve the problem.
-The asymptotic variance would become finite, but still possibly way too big.
-In the counter-example, how would we know beforehand that we should truncate outside $[-a,a]$?
-
-So, how can we work around this problem?
-
-%we can put the blame on both the factor for being far off the approximating family and the context $\pi$ as for being too narrow to ``see'' $f$ globally. But we may not know that beforehand.
-
-%Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
-
-\paragraph{Bullshite.}
-
-From this point, can we give practical sufficient conditions for VS convergence?
-Convergence in distribution will occur under mild conditions if the integral $\int \pi (f-g_\star)^2\phi\phi^\top$ is finite.
-The above inequality relating the VS variance to the KL~divergence tells us that this holds whenever $\pi$ has finite support.
-
-What if~$\pi$ is stricly Gaussian, hence doesn't have finite support?
-
-Similarly to IS, the condition that $f$ is upper bounded is sufficient for $\pi f^2 \phi\phi^\top$ to be finite, and may be easy to fulfill in practice.
-
-We are left with another sufficient condition (which IS doesn't have): that $\pi g_\star^2 \phi\phi^\top$ be finite too.
-Due to the log-quadratic nature of both $\pi$ and $g_\star$, this holds iff $\pi g_\star$ is Gaussian (i.e.
-its hegated log is a positive definite quadratic form).
-
-In 1D, let us denote $s^2$ and $\sigma^2$ the respective variances of $\pi$ and $\pi g_\star$.
-Clearly, $\sigma^2$ is finite because it is equal to the variance of $\pi f$ by the moment matching constraint, which is finite if $f$ is upper bounded.
-A simple asymptotic analysis then shows that the condition that $\pi g_\star^2$ has finite integral boils down to:
-$$
-\prec(\pi) + 2 [\prec(\pi f) - \prec(\pi)] > 0,
-\quad \Rightarrow \quad
-\prec(\pi) < 2 \prec(\pi f),
-$$
-where $\prec$ denotes the precision, {\em i.e.}, the inverse variance. The condition boils down to:
-$$
-s^2 > \frac{\sigma^2}{2}.
-$$
-
-The same reasoning can be extended in multiple dimension. If we use factorized Gaussians for fitting, the above condition simply applies to each axis.
-
-In short, the kernel size cannot be too small compared to the target distribution. However, we don't know $\sigma$, so how can we guarantee this in practice?
-
-Let now consider a damping factor $0\leq \alpha \leq 1$, and raise our factor~$f$ to the power~$\alpha$.
-
-is practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.
-
-Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.
+This condition is just more stringent than the one we identified before, without being easier to verify, therefore it's not super useful.
+But it does tell us the variance inequality yields sufficient conditions for convergence that are quite similar to the direct analysis, so the inequality gives some useful insight into what is going on.
+
+\section{Scaling}
+
+We may interpret the KL divergence $D(\pi f\|\pi g)$ as a divergence between the distributions $\pi f$ and $\pi g$, but also as a divergence between the factors $f$ and $g$ in the context defined by~$\pi$.
+This second interpretation opens a new perspective on~VS, and enables generalizations.
+For instance, we may consider the more general divergence:
+$$
+\delta_\alpha (f, g) \equiv
+D(\pi f^\alpha \|\pi g^\alpha),
+$$
+where $\alpha$ is a positive scaling factor.
+The intuition behind this generalization is that the scaled target factor $f^\alpha$ gets flatter as $\alpha$ approaches zero, hence the VS estimator should have lower variance considering that the approximating family contains constant functions.
+This assumes that the VS variance is continuous in~$\alpha$ but I can't see any reason to think otherwise.
+On the other hand, applying a scaling factor~$\alpha< 1$ amounts to ``shrinking'' the context with respect to the factor, hence biasing the target distribution  towards the particular context.
+
+This is a typical bias/variance tradeoff. Could the Expectation Propagation (EP) solve it? More exactly, its ``averaged'' variant (AEP).
+
+Consider rewriting the target distribution as follows:
+$$
+p = \pi \prod_{k=1}^n f^\alpha,
+\quad
+\alpha = \frac{1}{k}.
+$$
+
+Assume that we have a current approximation $g^\alpha$ for $f^\alpha$. Initially, this approximation could be just $g^\alpha\equiv 1$.
+
+The AEP update consists of fitting the factor $f^\alpha$ in the cavity $c=\pi g^{1-\alpha}$ by minimizing $D(cf^\alpha\|cg^\alpha)$ over a new $g$.
+This can be approximated using VS, with $c$ as the context, and $f^\alpha$ as the target factor.
+This is arguably a better conditioned problem than the original one in the sense that VS may converge faster.
+
+We see that, in the first iteration, since $g^\alpha=1$, the AEP update boils down to the scaling approach we have just proposed before.
+
+In the second iteration, the context would incorporate some information from the target distribution as the $g^\alpha$ output by VS would approach $f^\alpha$, therefore the new context would resemble a bit more $\pi f^{1-\alpha}$.
+By iterating this process, we might get a fix point for $g$ that is actually the minimizer of the $\alpha$-divergence $D_\alpha(\pi f\|\pi g)$.
+
+This could solve the chicken-and-egg problem I mentioned in my early work on VS: $\pi$ would be ideally chosen depending on a Gaussian approximation to the target distribution $p$, but we need $\pi$ to compute such an approximation via VS...
+
+The drawback is that we are no longer minimizing the KL divergence $D(\pi f\|\pi g)$, but (hopefully) the $\alpha$-divergence $D_\alpha(\pi f\|\pi g)$.
+This might be the price to pay to work around slow VS convergence.
 
@@
 
-\section{VS vs IS}
-
-The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$.
-This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor.
-The IS estimator is just a sample mean:
-$$
-\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
-\qquad x_i \stackrel{iid}{\sim} \pi(x),
-$$
-which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
-$$
-\mathbb{E}[f(X)] = \int\pi f,
-$$
-we also have that:
-$$
-{\rm Var}[f(X)]
-= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
-$$
-
-Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.
-
-As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$.
-Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.
-
-Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?
-Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''.
-We can make this a little bit more precise using :
-$$
-\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
-$$
-
 \section{Probability updating}
@@
 But, for the time being, let us see why this is interesting for VS.
-
-\section{Scaling}
-
-We may interpret the KL divergence $D(\pi f\|\pi g)$ as a divergence between the distributions $\pi f$ and $\pi g$, but also as a divergence between the factors $f$ and $g$ in the context defined by~$\pi$.
-This second interpretation opens a new perspective on~VS, and enables generalizations.
-For instance, we may consider the more general divergence:
-$$
-\delta_\varepsilon (f, g) \equiv
-D(\pi f^\varepsilon \|\pi g^\varepsilon),
-$$
-where $\varepsilon$ is a positive scaling factor.
-The intuition behind this generalization is that the scaled target factor $f^\varepsilon$ gets flatter as $\varepsilon$ approaches zero, hence the VS estimator should have lower variance considering that the approximating family contains constant functions.
-On the other hand, applying a scaling factor~$\varepsilon$ smaller than one amounts to ``shrinking'' the context with respect to the factor, hence biasing the divergence minimizer (the VS target) towards the particular context.
-This yields a typical bias/variance tradeoff.
-
-Could the Expectation Propagation (EP) solve this?
-
-$$
-p = \pi \prod_{k=1}^n f^\varepsilon,
-\quad
-\varepsilon = \frac{1}{k}.
-$$
-
-Say current approx is $g^\epsilon$ for all factors.
-An EP update is to fit the factor $f^\varepsilon$ in the cavity $c=\pi g^{1-\varepsilon}$ by minimizing $D(cf^\varepsilon\|cg^\varepsilon)$ over a new $g$.
-This can be approximated using VS, with $c$ as the context, and $f^\varepsilon$ as the target factor.
-This is arguably an easier problem than the original one.
 

### 2024-11-20 cv2/cv.tex
CV updates

@@
 
-\usepackage{FiraSans}        % Change this to use any font, but keep it simple
+%\usepackage{FiraSans}        % Change this to use any font, but keep it simple
 %\renewcommand{\familydefault}{\sfdefault}
@@
 \headleft{Profile}
- Hands-on technical team leader, {\bf applied math} researcher and engineer at heart, with a specialist expertise in {\bf computer vision} and {\bf computer graphics}.
+ Hands-on technical team leader, {\bf applied math} research engineer at heart, with a specialist expertise in {\bf computer vision} and {\bf computer graphics}.
 
@@
 
-{\sc Head of AI.} {\it Didimo}, Porto, Portugal.  \dates{Sep 2019--present} \\
-\smaller{Leading development of automated 3D character creation for {\em Popul8} software: shape reconstruction / generation, rig retargeting, skin texturing, asset fitting, \ldots}
+{\sc Head of Artificial Intelligence.} {\it Didimo}, Porto, Portugal. \hphantom{CACACACA} \dates{Sep 2019--present} \\
+\smaller{Leading development of automated 3D character creation for {\em Popul8} software: shape reconstruction / generation, rig retargeting, skin texturing, asset fitting, mesh wrapping.}
 
@@
 \is 
-{\sc Scientific reporting:} Author of 106 peer-reviewed scientific articles and 6 published international patents.
+{\sc Scientific writing:} Author of 106 peer-reviewed scientific articles and 6 published international patents. 7847 citations, {\em h}-index: 35.
 

### 2024-11-29 cv2/cv.tex
Update CV

@@
 \headleft{Personal information}
-\MALE \\[0.5ex]
+%\MALE \\[0.5ex]
 %Year of birth: \textbf{1861} \\[0.5ex]
@@
 {\sc Head of Artificial Intelligence.} {\it Didimo}, Porto, Portugal. \hphantom{CACACACA} \dates{Sep 2019--present} \\
-\smaller{Leading development of automated 3D character creation for {\em Popul8} software: shape reconstruction / generation, rig retargeting, skin texturing, asset fitting, mesh wrapping.}
+\smaller{Leading development of automated 3D character creation for {\em Popul8} software: shape reconstruction / generation, skin texturing, asset fitting, autorigging, autoskinning, mesh shrink-wrapping.}
 
 \is % Item spacing -- defined in the preamble
-{\sc Senior Computer Vision Scientist.} {\it CoVii (Ar\c celik group)}, Porto, Portugal. \dates{Feb 2017--Jul 2019} \\
-\smaller{Developed embedded algorithms for smart domestic appliances: {\em VUXHub} and {\em Artisan oven} prototypes (demonstrated at IFA Berlin, 2017-18).}
+{\sc Senior Computer Vision Scientist.} {\it CoVii (Ar\c celik/Beko R\& D center)}, Porto, Portugal. \dates{Feb 2017--Jul 2019} \\
+\smaller{Developed embedded ML algorithms for smart domestic appliances: {\em VUXHub} and {\em Artisan oven} prototypes (demonstrated at IFA Berlin, 2017-18).}
 
@@
 \is 
-{\sc Scientific writing:} Author of 106 peer-reviewed scientific articles and 6 published international patents. 7847 citations, {\em h}-index: 35.
+{\sc Scientific writing:} Author of 106 peer-reviewed scientific articles and 6 published international patents. 8086 citations, {\em h}-index: 38.
 
